---
subtitle: "Importance Sampling"
author: "Longhai Li"
format: banner_slides-revealjs
number-sections: false
code-fold: true
---

```{r}
#| include: false
set.seed(812)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center")
library(latex2exp)
library(ars)
ess <- function(w) sum(w)^2 / sum(w^2)
log_sum_exp <- function(lx) { m <- max(lx); m + log(sum(exp(lx - m))) }
log_ess <- function(lw) exp(2 * log_sum_exp(lw) - log_sum_exp(2 * lw))
```

# 1. Goal and Idea

## The Goal

Rejection sampling turns draws from an envelope $g\ge f$ into exact draws from $f$, but every rejected proposal is wasted — and in the tails of a poor envelope, almost every proposal is rejected.

Two integrals with respect to an **unnormalized** function $f(\theta)$, e.g. $f(\theta) = \pi(\theta)L(\theta)$:

1. A posterior expectation.
$$
A = \frac{\int a(\theta)\,f(\theta)\,d\theta}{\int f(\theta)\,d\theta} = E_f\big[a(\theta)\big], \qquad \theta\sim\frac{f(\theta)}{C_f}
$$

2. The normalizing constant (marginal likelihood).
$$
C_f = \int f(\theta)\,d\theta.
$$

Plain Monte Carlo needs draws from $f/C_f$, which we cannot produce. Rejection sampling needs an envelope $g\ge f$, which may be hard to find or very wasteful.

## The Idea: Reweighting {.smaller}

**Importance sampling** keeps every proposal instead of discarding most of them: draw from an easier function $g$ and attach a *weight* that corrects for sampling from the wrong distribution. Nothing is ever thrown away, and no envelope condition $g\ge f$ is required — the price is that the sample is no longer an exact iid sample from $f$, only a *weighted* one.

- Draw $\theta\sim g(\theta)/C_g$, where $g$ is another (possibly unnormalized) function we can sample from

- Attach the **importance weight**,
$$
w(\theta) = \frac{f(\theta)}{g(\theta)}
$$

- Hope that $w(\theta)\approx$ constant, i.e. $g$ has roughly the shape of $f$

**Requirement on $g$**: wherever $f$ is positive, $g$ must be positive too,
$$
\{\theta: g(\theta)>0\}\ \supseteq\ \{\theta: f(\theta)>0\}.
$$

Otherwise part of $f$ is never visited and the estimate is biased (Example 2).

# 2. The Importance Reweighting Formula

## Derivation {.smaller}

Let $C_g = \int g(\theta)\,d\theta$. For any $a$,
$$
\begin{aligned}
E_f\big[a(\theta)\big] &= \int a(\theta)\,\frac{f(\theta)}{C_f}\,d\theta
= \frac{C_g}{C_f}\int a(\theta)\,\frac{f(\theta)}{g(\theta)}\cdot\frac{g(\theta)}{C_g}\,d\theta \\[4pt]
&= \frac{C_g}{C_f}\,E_g\big[a(\theta)\,w(\theta)\big], \qquad w(\theta) = \frac{f(\theta)}{g(\theta)}.
\end{aligned}
$$

The division by $g$ is legitimate only on $\{g>0\}$, hence the support requirement.

Taking $a\equiv1$,
$$
\frac{C_f}{C_g} = \int\frac{f(\theta)}{g(\theta)}\cdot\frac{g(\theta)}{C_g}\,d\theta = E_g\big[w(\theta)\big]
= \frac{\int f(\theta)\,d\theta}{\int g(\theta)\,d\theta}.
$$

Combining,
$$
E_f\big[a(\theta)\big] = \frac{E_g\big[a(\theta)\,w(\theta)\big]}{E_g\big[w(\theta)\big]}.
$$

## The Estimators

Draw $\theta_1,\ldots,\theta_{\mathrm{S}} \overset{iid}{\sim} g/C_g$ and compute $w_s = w(\theta_s) = f(\theta_s)/g(\theta_s)$.

**Ratio of normalizing constants**
$$
\widehat{\frac{C_f}{C_g}} = \frac1S\sum_{s=1}^S w_s.
$$

Unbiased for $C_f/C_g$; if $g$ is a normalized density ($C_g=1$) this estimates $C_f$ itself — e.g. the marginal likelihood.

**Self-normalized expectation**
$$
\hat A = \frac{\frac1S\sum_s a(\theta_s)\,w_s}{\frac1S\sum_s w_s}
= \sum_{s=1}^S a(\theta_s)\,\tilde w_s, \qquad
\tilde w_s = \frac{w_s}{\sum_{j=1}^S w_j}, \quad \sum_s\tilde w_s = 1
$$

$\tilde w_s$ turns the proposal sample into a **weighted sample from $f$**; unknown constants in $f$ and $g$ cancel. It is a ratio of two averages, so it is biased with bias $O(1/S)$, but consistent.

## Effective Sample Size

How many "equally informative" draws is a weighted sample worth?.
$$
S_{\text{eff}} = \frac{1}{\sum_{s=1}^S \tilde w_s^2}.
$$

If $w_s\approx$ constant (proposal shaped like the target), $\tilde w_s = 1/S$ and $S_{\text{eff}} = S$: full efficiency. If one weight dominates, $S_{\text{eff}}\approx1$: only one draw is doing any work.

**Variance heuristic**: $\operatorname{Var}_g(\hat A)\approx\operatorname{Var}_f(a)/S_{\text{eff}}$, the same form as for an iid sample of size $S_{\text{eff}}$ from $f$ itself.

The catch, central to this lecture: $S_{\text{eff}}$ is computed **only from the weights actually observed**. A proposal that never visits the part of $\theta$-space where $f$ and $g$ disagree most will report a reassuring $S_{\text{eff}}$ while being badly wrong.

## Where the ESS Formula Comes From {.smaller}

Write $\hat A = \overline N/\overline D$ with $N_s = w_s a(\theta_s)$, $D_s = w_s$. The delta method gives
$$
\operatorname{Var}_g(\hat A) \approx \frac{1}{S}E_g\!\bigl[w^{2}(a-A)^{2}\bigr] = \frac{1}{S}E_f\!\bigl[w(a-A)^{2}\bigr],
$$
using $E_g[w\,h] = E_f[h]$. If $w$ varies little across the support of $(a-A)^2$, this factors as $E_g[w^2]\operatorname{Var}_f(a)/S$, giving
$$
\operatorname{Var}_g(\hat A) \approx \frac{\operatorname{Var}_f(a)}{S_{\text{eff}}^{\star}}, \qquad S_{\text{eff}}^{\star} := \frac{S}{E_g[w^{2}]}
$$

$S_{\text{eff}}^{\star}$ is a property of $f$ and $g$ alone — independent of $a$ or the sample — and it is the metric that reveals pathologies (missed modes, heavy tails) even when the *observed* $S_{\text{eff}}$ looks fine.

Plugging in sample moments, $\frac{1}{S}\sum w_s$ for $E_g[w]$ and $\frac{1}{S}\sum w_s^2$ for $E_g[w^2]$, turns $S_{\text{eff}}^\star$ into **Kish's formula** $(\sum w_s)^2/\sum w_s^2 = 1/\sum\tilde w_s^2$ — a principled plug-in estimator, not an ad hoc heuristic.

## Limitations of the ESS Diagnostic {.smaller}

Kish's approximation is not uniformly precise. With $f=N(0,1)$, $g=N(0,2^2)$, $a(\theta)=\theta^2$, $S=4000$:

- empirical $\operatorname{Var}(\hat A)\approx 3.15\times10^{-4}$, matching exact delta-method calculations

- but $\operatorname{Var}_f(a)/S_{\text{eff}}^\star \approx 7.56\times10^{-4}$ — over twice as large, because $w$ and $(a-A)^2$ are negatively correlated here

More critically: $\widehat{S_{\text{eff}}}$ is computed **only from the observed weights**. If the proposal never explores where $f$ and $g$ disagree most, the estimator reports a falsely reassuring $\widehat{S_{\text{eff}}}$ and masks a severe sampling inadequacy — the subject of Examples 1 and 2 below.

# 3. Example 1: Same Support, but a Mode Is Missed

## Example 1: Same Support, but a Mode Is Missed,
$$
f(\theta) = \tfrac12\varphi(\theta+d) + \tfrac12\varphi(\theta-d), \qquad
g_{\sigma}(\theta) = \tfrac{1}{\sigma}\varphi\!\left(\tfrac{\theta+d}{\sigma}\right),
\qquad d = 5.
$$

Both are positive on all of $\mathbb{R}$, so the support requirement holds. But for $\sigma$ of order one the proposal concentrates around $-d$ and never produces draws near the second mode at $+d$.

For $\theta$ near $-d$ and $\sigma=1$,
$$
w(\theta) = \tfrac12 + \tfrac12\,\frac{\varphi(\theta-d)}{\varphi(\theta+d)}
          = \tfrac12 + \tfrac12 e^{2d\theta} \;\approx\; \tfrac12,
$$

so every observed weight is nearly equal, $\tilde w_i\approx1/S$ and $S_{\text{eff}}\approx S$, while $\widehat{C_f/C_g}\approx\tfrac12$ (truth $1$) and $\hat E_f(\theta)\approx-d$ (truth $0$). The draws that would expose the error carry $w\approx\tfrac12e^{2d\theta}\gg1$ near $\theta=+d$ and have probability $\Phi(-2d)\approx10^{-23}$ under $g_1$: they never appear.

## Example 1 (Continued): The Figure

```{r}
#| label: fig-missed-mode
#| fig-width: 9
#| fig-height: 7
d <- 5; sg <- 1; S <- 2000
set.seed(11)

log_f <- function(th, d) {
    a <- dnorm(th, -d, log = TRUE); b <- dnorm(th, d, log = TRUE)
    m <- pmax(a, b)
    m + log(exp(a - m) + exp(b - m)) - log(2)
}
log_Ew2 <- function(d, sg) {
    A <- 2 - 1 / sg^2
    if (A <= 0) return(Inf)
    lterm <- function(a, b) {
        m <- (a + b) / 2
        B <- 2 * m + d / sg^2
        C <- 2 * m^2 - d^2 / sg^2
        dnorm(a, b, sqrt(2), log = TRUE) + log(sg) + 0.5 * log(2) +
            0.5 * log(2 * pi / A) - 0.5 * (C - B^2 / A)
    }
    lv <- c(lterm(-d, -d), log(2) + lterm(-d, d), lterm(d, d))
    m  <- max(lv)
    m + log(sum(exp(lv - m))) - log(4)
}

th   <- rnorm(S, -d, sg)
logw <- log_f(th, d) - dnorm(th, -d, sg, log = TRUE)
mx   <- max(logw); v <- exp(logw - mx)
Chat <- mean(v) * exp(mx); se <- sd(v) / sqrt(S) * exp(mx)
wn   <- v / sum(v); ess_obs <- 1 / sum(wn^2)
Ehat <- sum(wn * th); ess_x <- 100 * exp(-log_Ew2(d, sg))

xr <- range(c(th, -d - 4, d + 4)); xs <- seq(xr[1], xr[2], length.out = 1000)
fx <- exp(log_f(xs, d)); gx <- dnorm(xs, -d, sg)
lwc <- log_f(xs, d) - dnorm(xs, -d, sg, log = TRUE) - (mx + log(sum(v)))

layout(matrix(1:2, 2), heights = c(3, 2))
par(mar = c(0.4, 5, 3, 1))
yl <- c(1e-10, max(fx, gx) * 2)
plot(xs, fx, type = "n", log = "y", ylim = yl, xlim = xr, xaxt = "n",
     xlab = "", ylab = "density (log scale)",
     main = "bimodal target and a proposal centred on one mode")
lines(xs, fx, lwd = 4, col = "red"); lines(xs, gx, lwd = 3, col = "grey60")
rug(th, col = adjustcolor("grey45", 0.35))
legend("bottomright", bty = "n", cex = 1.0, text.col = "navy",
       lwd = c(4, 3, NA, NA, NA), col = c("grey25", "grey60", NA, NA, NA),
       legend = TeX(c(
           r"(target $f$)",
           sprintf(r"(proposal $g$, sd $= %.2f$)", sg),
           sprintf(r"($C_f/C_g$: true 1, est. %.3f)", Chat),
           sprintf(r"(95%% CI $(%.3f, %.3f)$)", Chat - 1.96 * se, Chat + 1.96 * se),
           sprintf(r"($E_f(\theta)$: true 0, est. %.2f)", Ehat))))

par(mar = c(4.5, 5, 0.4, 1))
wl <- c(min(wn) / 5, max(wn) * 5)
plot(th, wn, type = "n", log = "y", ylim = wl, xlim = xr,
     xlab = expression(theta), ylab = "normalized weight (log scale)")
lines(xs, exp(pmin(lwc, 700)), lwd = 2, col = "grey25")
segments(th, wl[1], th, wn, col = adjustcolor("grey45", 0.5))
abline(h = 1 / S, lty = 2, col = "grey55", lwd = 2)
legend("bottomright", bty = "n", cex = 1.0, text.col = "navy",
       lty = c(2, 1, NA, NA), lwd = c(2, 2, NA, NA), col = c("grey55", "grey25", NA, NA),
       legend = TeX(c(
           r"($1/S$)", r"(normalized weight function)",
           sprintf(r"(observed $S_{eff}$ = %.1f%% of S)", 100 * ess_obs / S),
           sprintf(r"(exact $S_{eff}$ = %.3g%% of S)", ess_x))))
```

The observed effective sample size reports near-perfect efficiency. The **exact** effective fraction $1/E_g(w^2)$ is smaller by forty orders of magnitude — a property of $f,g$ alone, invisible to any single sample. The variance is finite at $\sigma=1$ (it diverges only for $\sigma\le1/\sqrt2$), so the estimator is consistent — it would just need $S\sim10^{43}$ to look normal.


# 4. Example 2: Tail Probability of a Student-$t$ Distribution

## Example 2: Tail Probability of a Student-$t$ Distribution

The target is a standard Student-$t_\nu$: $\nu=1$ is the Cauchy, $\nu\to\infty$ recovers the normal — $\nu$ is a **tuning parameter of the target's tail weight**. Goal: $P(c_1<X<c_2)$ for an interval $(c_1,c_2)$, possibly one-sided ($c_2=\infty$), from a truncated-normal proposal sampled by adaptive rejection sampling (ARS).

$\log\varphi$ is concave, and restricting to $(c_1,c_2)$ preserves concavity, so the `ars` package applies directly to the *proposal*.

```{r}
#| label: rtnorm-ars
#| echo: true
rtnorm_ars <- function(n, mu = 0, sigma = 1, lo = 0, hi = Inf) {
    logf   <- function(x) dnorm(x, mu, sigma, log = TRUE)
    fprima <- function(x) -(x - mu) / sigma^2
    if (is.finite(hi)) {
        x0 <- lo + (hi - lo) * c(0.1, 0.5, 0.9)
    } else {
        right <- max(mu, lo) + 3 * sigma
        x0 <- sort(unique(c(lo + 0.1 * sigma, max(lo + 0.2 * sigma, mu + 1e-6), right)))
    }
    ars(n, f = logf, fprima = fprima, x = x0,
        lb = TRUE, xlb = lo, ub = is.finite(hi), xub = if (is.finite(hi)) hi else 0)
}
```

The target $\log f_{t_\nu}$ is **not concave** at any finite $\nu$ — convex out in the tails, concave everywhere only in the $\nu\to\infty$ limit — so ARS cannot sample it directly. The truncated-normal proposal is a substitute, which is exactly why the sample needs importance weights.

## Example 2 (Continued): The Estimator

Let $g$ be the normalized truncated normal on $(c_1,c_2)$, $f(x)=f_{t_\nu}(x)I(c_1<x<c_2)$. Draw $X_1,\ldots,X_{\mathrm{S}}\sim g$, weight by $w=f/g$, average:
$$
\widehat C = \frac1S\sum_{i=1}^S w(X_i),
$$
unbiased for $C_f=P(c_1<X<c_2)$ for *any* $g$ positive wherever $f$ is — regardless of $c_1,c_2,\mu,\sigma,\nu$.

## Example 2 (Continued): When Its Variance Is Infinite {.smaller}

*Precision* is a different question:
$$
\operatorname{Var}_g\bigl(w(X)\bigr) = \int_{c_1}^{c_2}\frac{f_{t_\nu}(x)^2}{g(x)}\,dx - C_f^2
$$

- **$c_2$ finite**: $f,g$ share the same bounded interval, $f/g$ is bounded on it, the integral is finite — well-behaved at *any* $\nu$.

- **$c_2\to\infty$**: $f_{t_\nu}(x)^2$ decays only polynomially ($\propto x^{-2(\nu+1)}$), $g$'s Gaussian tail decays much faster — the integrand grows without bound and the integral **diverges**, for every $\mu,\sigma$ and every finite $\nu$.

$S_{\text{eff}}$ need not warn you which regime you are in.

## Example 2 (Continued): `is_t_interval()`

```{r}
#| label: is-t-interval
#| echo: true
is_t_interval <- function(S, df, c1, c2, mu, sigma, seed = NULL) {
    if (!is.null(seed)) set.seed(seed)
    X <- rtnorm_ars(S, mu, sigma, c1, c2)

    lZ   <- log(pnorm(c2, mu, sigma) - pnorm(c1, mu, sigma))
    logw <- dt(X, df, log = TRUE) - (dnorm(X, mu, sigma, log = TRUE) - lZ)

    m    <- max(logw); v <- exp(logw - m)
    Chat <- mean(v) * exp(m); se <- sd(v) / sqrt(S) * exp(m)
    wn   <- v / sum(v)

    list(X = X, wn = wn, Chat = Chat, se = se, ess = 1 / sum(wn^2),
         truth = pt(c2, df) - pt(c1, df))
}
```

```{r}
#| label: tbl-is-t-examples
#| echo: true
settings <- data.frame(df = c(1, 10, 1, 30), c1 = c(1, 1, 1, 1), c2 = c(3, 3, Inf, Inf))
res <- lapply(seq_len(nrow(settings)), function(i) with(settings[i, ],
    is_t_interval(S = 5000, df = df, c1 = c1, c2 = c2, mu = 0, sigma = 1, seed = 1)))

knitr::kable(data.frame(
    df = settings$df, c1 = settings$c1,
    c2 = ifelse(is.finite(settings$c2), as.character(settings$c2), "Inf"),
    truth = sapply(res, `[[`, "truth"), estimate = sapply(res, `[[`, "Chat"),
    se = sapply(res, `[[`, "se"), ess_pct = 100 * sapply(res, `[[`, "ess") / 5000
), digits = 4, col.names = c("df", "c1", "c2", "truth", "estimate", "s.e.", "ESS %"))
```

A finite interval is well-behaved at any $\nu$; $c_2=\infty$ reproduces the divergence — badly at $\nu=1$, only *deceptively* well at $\nu=30$.


## Example 2 (Continued):Observations {.smaller}

- **$c_2$ finite (default)**: weights have finite variance; raising $S$ improves the estimate at the usual $1/\sqrt S$ rate, and the 95% interval covers about as often as claimed

- **Check "$c_2=\infty$"**: reproduces the divergent case. Raising $S$ no longer shrinks a variance that doesn't exist; repeated *New sample* draws swing far more than the SE admits. Large $\nu$ can still look fine by chance; small $\nu$ (Cauchy) bites often — but $S_{\text{eff}}$ warns you either way only sometimes
- **Small $\sigma$, or $\mu$ far outside $(c_1,c_2)$**: draws pile up wherever the proposal has mass, weights nearly equal, $S_{\text{eff}}$ stays high, estimate can still miss badly

- **Large $\sigma$**: one draw's weight can dominate, $S_{\text{eff}}$ collapses toward $1$ — visibly unreliable, an improvement on invisibly unreliable

# 5. Example 3: Marginal Likelihood via Importance Sampling

## Example 3: Marginal Likelihood via Importance Sampling

Model: unknown mean $\mu$ and $w=\log\sigma^2$, independent normal priors. Importance sampling estimates $P(y)=\int L(\theta)\pi(\theta)\,d\theta$ with a built-in diagnostic ($S_{\text{eff}}$).

Take $g(\mu,w)$ to be the **Gaussian approximation to the posterior** from a Laplace fit: mode $(\hat\mu,\hat w)$, $g=N_2\big((\hat\mu,\hat w), H^{-1}\big)$ for $H$ the Hessian of $-\log[L\pi]$ at the mode:
$$
\widehat{P(y)} = \frac1S\sum_{s=1}^S w_s, \qquad
w_s = \frac{L(\theta_s)\pi(\theta_s)}{g(\theta_s)}, \qquad \theta_s\sim g.
$$

Contrast: sampling directly from the (wide, uninformative) **prior** is importance sampling with $g=\pi$ and weights $L(\theta_s)$ — a "good story" versus a proposal built with no information about where the likelihood actually lives.

## Example 3 (Continued): Implementation {.smaller}

```{r}
#| label: marlik-functions
#| echo: true
#| code-fold: false
log_lik   <- function(x, mu, w) sum(dnorm(x, mu, exp(w / 2), log = TRUE))
log_prior <- function(mu, w, mu_0, sigma_mu, w_0, sigma_w) {
    dnorm(mu, mu_0, sigma_mu, log = TRUE) + dnorm(w, w_0, sigma_w, log = TRUE)
}
neg_log_post <- function(theta, x, mu_0, sigma_mu, w_0, sigma_w) {
    -log_lik(x, theta[1], theta[2]) -
        log_prior(theta[1], theta[2], mu_0, sigma_mu, w_0, sigma_w)
}
log_dmvnorm_batch <- function(Theta, mu, A) {           # log N(mu, A^-1) at every column at once
    Tc <- Theta - mu
    quad <- colSums((A %*% Tc) * Tc)
    0.5 * (-nrow(Theta) * log(2 * pi) + sum(log(svd(A)$d)) - quad)
}
log_marlik_imps <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    fit <- nlm(neg_log_post, c(mean(x), log(stats::var(x))), hessian = TRUE,
               x = x, mu_0 = mu_0, sigma_mu = sigma_mu, w_0 = w_0, sigma_w = sigma_w)
    mu_hat <- fit$estimate; H <- fit$hessian
    L <- t(chol(solve(H)))
    Theta <- L %*% matrix(rnorm(2 * S), 2, S) + mu_hat
    log_g <- log_dmvnorm_batch(Theta, mu_hat, H)
    log_f <- -apply(Theta, 2, neg_log_post, x = x, mu_0 = mu_0, sigma_mu = sigma_mu,
                     w_0 = w_0, sigma_w = sigma_w)
    lw <- log_f - log_g
    m  <- max(lw); v <- exp(lw - m)
    structure(log_sum_exp(lw) - log(S), ess = log_ess(lw), se = sd(v) / (sqrt(S) * mean(v)))
}
log_marlik_prior <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    mus <- rnorm(S, mu_0, sigma_mu); ws <- rnorm(S, w_0, sigma_w)
    lw <- vapply(seq_len(S), function(i) log_lik(x, mus[i], ws[i]), numeric(1))
    m  <- max(lw); v <- exp(lw - m)
    structure(log_sum_exp(lw) - log(S), ess = log_ess(lw), se = sd(v) / (sqrt(S) * mean(v)))
}
log_marlik_grid <- function(x, mu_0, sigma_mu, w_0, sigma_w, n_grid = 400, width = 8) {
    n   <- length(x)
    fit <- nlm(neg_log_post, c(mean(x), log(stats::var(x))), hessian = TRUE,
               x = x, mu_0 = mu_0, sigma_mu = sigma_mu, w_0 = w_0, sigma_w = sigma_w)
    se  <- sqrt(diag(solve(fit$hessian)))
    mu_grid <- fit$estimate[1] + seq(-width, width, length.out = n_grid) * se[1]
    w_grid  <- fit$estimate[2] + seq(-width, width, length.out = n_grid) * se[2]
    SS <- sum(x^2) - 2 * mu_grid * sum(x) + n * mu_grid^2
    log_lik_grid   <- outer(SS, w_grid,
                             function(ss, w) -0.5 * n * log(2 * pi) - n * w / 2 - 0.5 * exp(-w) * ss)
    log_prior_grid <- outer(dnorm(mu_grid, mu_0, sigma_mu, log = TRUE),
                             dnorm(w_grid, w_0, sigma_w, log = TRUE), "+")
    h_mu <- diff(mu_grid[1:2]); h_w <- diff(w_grid[1:2])
    log_sum_exp(as.vector(log_lik_grid + log_prior_grid)) + log(h_mu) + log(h_w)
}
```

`log_marlik_grid()` supplies ground truth by quadrature — a midpoint grid centred and scaled by the Laplace fit, vectorized via the sufficient statistic $\sum(x-\mu)^2$.

## Example 3 (Continued): Results Across Sample Sizes {.smaller}

```{r}
#| label: tbl-marlik-merged
#| echo: true
set.seed(4)
ns <- c(5, 20, 100, 400); S <- 2000

summarize_is <- function(fn, x, S) {
    est <- fn(x, 0, 5, 0, 5, S); se <- attr(est, "se")
    c(estimate = sprintf("%.2f", as.numeric(est)),
      ci       = sprintf("(%.2f, %.2f)", as.numeric(est) - 1.96 * se, as.numeric(est) + 1.96 * se),
      ess      = sprintf("%.1f", 100 * attr(est, "ess") / S))
}
merged_tbl <- do.call(rbind, lapply(ns, function(n) {
    x   <- rnorm(n)
    lap <- summarize_is(log_marlik_imps,  x, S)
    pri <- summarize_is(log_marlik_prior, x, S)
    data.frame(n = n, truth = sprintf("%.2f", log_marlik_grid(x, 0, 5, 0, 5)),
               lap_estimate = lap[["estimate"]], lap_ci = lap[["ci"]], lap_ess = lap[["ess"]],
               pri_estimate = pri[["estimate"]], pri_ci = pri[["ci"]], pri_ess = pri[["ess"]],
               row.names = NULL)
}))
names(merged_tbl) <- c("n", "truth", "Laplace: est.", "Laplace: 95% CI", "Laplace: ESS %",
                       "prior: est.", "prior: 95% CI", "prior: ESS %")
knitr::kable(merged_tbl, align = "r")
```

Laplace-shaped: tight interval centred on truth at every $n$, $S_{\text{eff}}$ near nominal. Prior sampling degrades monotonically as the posterior concentrates at rate $n^{-1/2}$ while the prior stays put — $S_{\text{eff}}$ falls to a fraction of one percent, and its (still sometimes-covering) interval is only as trustworthy as those few dominant draws.

## Example 3 (Continued): What to Look For

Sorted, normalized weights, largest first, one panel per proposal (their scales differ by orders of magnitude) — a good proposal gives a flat curve near $1/S$; a bad one falls off a cliff.

# 6. Advanced Applications of Importance Sampling

## Free Energy and the Marginal Likelihood

A system with energy $U(x)$ at inverse temperature $\beta$ has.
$$
Z(\beta) = \int e^{-\beta U(x)}\,dx, \qquad F(\beta) = -\frac1\beta\log Z(\beta).
$$

Importance sampling from $\beta_0$ to $\beta_1$ with $g=e^{-\beta_0U}$, $f=e^{-\beta_1U}$, $w(x)=e^{-(\beta_1-\beta_0)U(x)}$:
$$
\frac{Z(\beta_1)}{Z(\beta_0)} = E_{\beta_0}\Big[e^{-(\beta_1-\beta_0)U(x)}\Big].
$$

## The Bayesian Analogue {.smaller}

$U(\theta)=-\log L(\theta)$, $g=\pi(\theta)$ (prior, $\beta_0=0$), $f=L(\theta)\pi(\theta)$ (posterior, $\beta_1=1$):
$$
P(y) = \frac{C_f}{C_g} = E_\pi\big[L(\theta)\big] \approx \frac1S\sum_{s=1}^S L(\theta_s), \qquad \theta_s\sim\pi
$$

$-\log P(y)$ is the free energy of the posterior. "Prior sampling" is exactly Example 1's failure in disguise: the prior is diffuse, the likelihood a narrow spike, weights are $L(\theta_s)$, and almost all prior draws carry negligible weight — jumping straight from $\beta=0$ to $\beta=1$ asks one importance-sampling step to bridge the *entire* temperature range at once.

## Stepping Stones: Many Small Jumps Instead of One Big One {.smaller}

Chain intermediate temperatures $\beta_0=0<\beta_1<\cdots<\beta_K=1$:
$$
\frac{Z(1)}{Z(0)} = \prod_{k=1}^K \frac{Z(\beta_k)}{Z(\beta_{k-1})}, \qquad
\log Z(1)-\log Z(0) = \sum_{k=1}^K \log\frac{Z(\beta_k)}{Z(\beta_{k-1})}.
$$

Each factor is its own small importance-sampling problem: sample from the tempered distribution at $\beta_{k-1}$, weight by $e^{-(\beta_k-\beta_{k-1})U(\theta)}$. If temperatures are close together, $g_{\beta_{k-1}}$ and $f_{\beta_k}$ nearly coincide and every step is a good-story proposal with near-constant weights — the same principle as building an envelope from many well-fitting pieces rather than one loose one.

This chain is a **stepping-stones** estimator — a discretization of thermodynamic integration, and the schedule idea behind simulated annealing and parallel tempering. The prior-vs-posterior mismatch has not gone away; spreading it over many small, well-matched steps rather than absorbing it in one large jump is what makes tempering-based samplers practical where a direct jump is not.

## Bayesian Leave-One-Out Cross-Validation

The Bayesian analogue of a cross-validation fold is the **leave-one-out predictive density**,
$$
p(y_i\mid y_{-i}) = \int p(y_i\mid\theta)\,p(\theta\mid y_{-i})\,d\theta,
$$

the posterior predictive density for $y_i$ under the posterior that never saw $y_i$. Refitting the model $n$ times — one full posterior per left-out observation — is exactly the cost importance sampling is suited to avoid.

## Reusing the Full Posterior as the Proposal {.smaller}

Take $g(\theta)=p(\theta\mid y)$, the **full** posterior (already available from one fit), as the proposal for the leave-one-out posterior $f(\theta)=p(\theta\mid y_{-i})$. Since $p(\theta\mid y)\propto p(y_i\mid\theta)\,p(\theta\mid y_{-i})$,
$$
f(\theta) = p(\theta\mid y_{-i}) \propto \frac{p(\theta\mid y)}{p(y_i\mid\theta)}, \qquad
w_i(\theta) = \frac{f(\theta)}{g(\theta)} \propto \frac{1}{p(y_i\mid\theta)}.
$$

The self-normalized estimator with $a(\theta)=p(y_i\mid\theta)$ collapses to a **harmonic mean**:
$$
\hat p(y_i\mid y_{-i}) = \frac{\sum_s p(y_i\mid\theta_s)\,w_i(\theta_s)}{\sum_s w_i(\theta_s)}
= \frac{S}{\sum_{s=1}^S 1/p(y_i\mid\theta_s)},
$$

over draws from the *full* posterior (Gelfand, Dey & Chang, 1992) — one weighted average per left-out observation, all reusing the same $S$ draws from a single fit.

## When It Works, and When It Breaks {.smaller}

The full posterior is **narrower** than any leave-one-out posterior (it has one more observation), so $g$ has lighter tails than $f$: the *heavy-tailed-weight* failure from Example 2, not the missed-mode failure from Example 1. Typical points: the two posteriors nearly coincide, weights stable. Influential points (outliers, high leverage): $1/p(y_i\mid\theta_s)$ can have infinite variance, and the estimate is dominated by a few draws.

**Diagnostics and fixes**: report $S_{\text{eff}}$ *per observation*; **Pareto-smoothed importance sampling** (Vehtari, Gelman & Gabry, 2017) fits a generalized Pareto distribution to the largest weights and uses the shape $\hat k$ as a diagnostic ($\hat k>0.7$ unreliable); fall back to exact refitting only for the flagged observations.

## Summary (1/2)

- Importance sampling reweights draws from a proposal $g$ by $w(\theta)=f(\theta)/g(\theta)$ instead of accepting/rejecting them:
$$
\widehat{C_f/C_g} = \frac1S\sum_s w_s, \qquad \hat A = \sum_s a(\theta_s)\,\tilde w_s, \quad \tilde w_s = \frac{w_s}{\sum_j w_j},
$$
only the support condition $\{g>0\}\supseteq\{f>0\}$ is required, not an envelope $g\ge f$

- A good proposal shaped like the target on its support gives near-constant weights and $S_{\text{eff}}\approx S$ (Examples 2 and 3), and a single weighted sample can answer several questions about the same target

## Summary (2/2)

- A proposal that misses a mode or part of the support fails **silently** (Example 1): observed weights look healthy while the estimate is badly wrong, and no amount of extra sampling fixes it — $S_{\text{eff}}$ diagnoses only the *noisy* failure, never the silent one

- The marginal likelihood: a Laplace-shaped proposal stays efficient as data accumulate and the posterior sharpens; the prior does not
- Advanced uses: normalizing-constant ratios (free energy, marginal likelihood) via tempering/stepping-stones; leave-one-out cross-validation from full-posterior draws with $w_i=1/p(y_i\mid\theta)$, diagnosed per observation by Pareto-smoothed importance sampling
