---
subtitle: "Laplace Approximation"
pagetitle: "laplace-slides"
author: "Longhai Li"
format: banner_slides-revealjs
number-sections: false
---

```{r}
#| include: false
set.seed(812)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center")
library(latex2exp)
```

# 1. Goal and Idea

## The Goal

Compute posterior expectations
$$E_f\big[a(\theta)\big] = \frac{\int a(\theta)\,f(\theta)\,d\theta}{\int f(\theta)\,d\theta}$$

where

- $a(\theta)$ is a function of $\theta\in\mathbb R^p$ (a parameter, a prediction, an indicator)
- $f(\theta)$ is the **unnormalized posterior**, e.g. $f(\theta) = L(\theta)\,\pi(\theta)$, or a penalized likelihood
- both numerator and denominator are $p$-dimensional integrals

Quadrature works only for small $p$; Monte Carlo needs a sampler. The **Laplace approximation** replaces the integrals by a closed-form Gaussian integral, exploiting the fact that for large $n$ the posterior is nearly normal.

## The Idea

Write the negative log unnormalized posterior
$$h(\theta) = -\log f(\theta), \qquad f(\theta) = e^{-h(\theta)}$$

- $h$ is (up to a constant) the negative log-likelihood plus the negative log-prior: a sum of $n$ terms, so it grows like $n$ and becomes sharply curved at its minimum
- When $n$ is large,
$$\theta\mid\text{data} \ \approx\ N\big(\hat\theta_0,\ \hat\Sigma_0\big)$$
with $\hat\theta_0$ the posterior mode and $\hat\Sigma_0$ determined by the curvature of $h$ there

The approximation is obtained by expanding $h$ to second order around its minimum and integrating the resulting Gaussian exactly.

# 2. Derivation

## Second-Order Taylor Expansion {.smaller}

Let $\theta_0 = \arg\min_\theta h(\theta)$ (the posterior mode), found by optimization. Expand $h$ around $\theta_0$:
$$
h(\theta) = h(\theta_0) + \underbrace{\nabla h(\theta_0)^\top}_{=\,0 \text{ at the minimum}}(\theta-\theta_0)
+ \tfrac12(\theta-\theta_0)^\top\nabla^2 h(\theta_0)(\theta-\theta_0) + O\big(\|\theta-\theta_0\|^3\big)
$$

Keeping terms up to second order,
$$h(\theta)\ \approx\ \tilde h(\theta) = h(\theta_0) + \tfrac12(\theta-\theta_0)^\top H\,(\theta-\theta_0),
 H = \nabla^2 h(\theta_0)$$

- $H$ is the Hessian of $-\log f$ at the mode; it is positive definite at a strict minimum
- For $f = L\pi$, $H$ is the **observed information** plus the prior curvature

## From $\tilde h$ to a Gaussian {.smaller}

$$
\begin{aligned}
f(\theta)\ \approx\ \tilde f(\theta) &= e^{-\tilde h(\theta)}
= e^{-h(\theta_0)}\cdot\exp\Big\{-\tfrac12(\theta-\theta_0)^\top H(\theta-\theta_0)\Big\} \\[4pt]
&= \underbrace{f(\theta_0)\,(2\pi)^{p/2}\,|H|^{-1/2}}_{\text{constant}}
\cdot\underbrace{(2\pi)^{-p/2}\,|H|^{1/2}\exp\Big\{-\tfrac12(\theta-\theta_0)^\top H(\theta-\theta_0)\Big\}}_{N_p(\theta_0,\,H^{-1})\text{ density, integrates to }1}
\end{aligned}
$$

**Review**: if $X\sim N_p(\mu,\Sigma)$, then $p(x) = (2\pi)^{-p/2}|\Sigma|^{-1/2}\exp\{-\tfrac12(x-\mu)^\top\Sigma^{-1}(x-\mu)\}$. Here $\Sigma = H^{-1}$, so $|\Sigma|^{-1/2} = |H|^{1/2}$.

## The Two Results

1. **Posterior approximation**
$$\theta\mid\text{data}\ \approx\ N_p\Big(\theta_0,\ \big[\nabla^2 h(\theta_0)\big]^{-1}\Big)$$

2. **Normalizing constant (marginal likelihood)**
$$\int f(\theta)\,d\theta\ \approx\ f(\theta_0)\,(2\pi)^{p/2}\,\big|\nabla^2 h(\theta_0)\big|^{-1/2}$$

Only two ingredients are needed: the mode $\theta_0$ and the Hessian of $-\log f$ at the mode, both available from a numerical optimizer (`optim(..., hessian = TRUE)` in R).

The relative error of (2) is $O(1/n)$ for regular models; the error of (1) is driven by the skewness of the true posterior.

## The Two Results (Continued)

Red: unnormalized posterior $f(\theta)$. Green: the Gaussian $\tilde f(\theta)$ matching height and curvature at the mode. The shaded rectangle-like region has area $f(\theta_0)(2\pi)^{p/2}|H|^{-1/2}$, the Laplace estimate of $\int f(\theta)\,d\theta$.

```{r}
#| fig-width: 9
#| fig-height: 4.6
par(mar = c(4, 4, 1, 1))
a <- 6; b <- 2                                  # unnormalized Gamma(6, 2) kernel as "posterior"
f <- function(t) t^(a - 1) * exp(-b * t)
t0 <- (a - 1) / b; H <- (a - 1) / t0^2          # mode and -d2/dt2 log f at mode
ft <- function(t) f(t0) * exp(-0.5 * H * (t - t0)^2)
ts <- seq(0, 8, length.out = 400)
plot(ts, f(ts), type = "n", xlab = TeX(r"($\theta$)"), ylab = "", ylim = c(0, f(t0) * 1.15))
polygon(c(ts, rev(ts)), c(ft(ts), rep(0, 400)), col = adjustcolor("forestgreen", 0.25), border = NA)
lines(ts, f(ts), lwd = 3, col = "firebrick"); lines(ts, ft(ts), lwd = 3, col = "forestgreen")
abline(v = t0, lty = 3); text(t0, 0, TeX(r"($\theta_0$)"), pos = 1, xpd = NA)
legend("topright", bty = "n", lwd = 3, col = c("firebrick", "forestgreen"),
       legend = TeX(c(r"($f(\theta) = L(\theta)\pi(\theta)$)", r"($\tilde{f}(\theta) \propto N(\theta_0, H^{-1})$)")))
mtext(sprintf("exact integral = %.3f,  Laplace = %.3f", integrate(f, 0, Inf)$value, f(t0) * sqrt(2 * pi / H)),
      side = 3, line = -1.5, adj = 0.02, cex = 0.95)
```

# 3. Applications

## Application 1: Marginal Likelihood and BIC {.smaller}

With $f(\theta) = L(\theta;D)\,\pi(\theta)$, the marginal likelihood $P(D) = \int L(\theta;D)\pi(\theta)\,d\theta$ satisfies
$$\log P(D)\ \approx\ \log\big[P(D\mid\theta_0)\,\pi(\theta_0)\big] + \frac p2\log(2\pi) - \frac12\log\big|\nabla^2 h(\theta_0)\big|$$

**Simplification for large $n$.** The Hessian is dominated by the likelihood: $\nabla^2 h(\theta_0)\approx n\,I(\theta_0)$ with $I$ the per-observation Fisher information, so $|\nabla^2 h(\theta_0)| \approx n^p\,|I(\theta_0)|$ and
$$\log P(D) \approx \log P(D\mid\theta_0) - \frac p2\log n + \underbrace{\log\pi(\theta_0) + \frac p2\log(2\pi) - \frac12\log|I(\theta_0)|}_{O(1),\ \text{dropped}}$$

Multiplying by $-2$ and replacing the posterior mode by the MLE gives the **Bayesian information criterion**:
$$\text{BIC} = -2\log L(\hat\theta_{\text{MLE}}) + p\log n$$



## Application 2: Linear Functions of $\theta$

If $a(\theta) = A^\top\theta + a_0$ is linear, replace $f$ by its Gaussian approximation:
$$E_f\big[a(\theta)\big]\ \approx\ E_{\tilde f}\big[a(\theta)\big] = A^\top E_{\tilde f}(\theta) + a_0 = A^\top\theta_0 + a_0$$

- The posterior mean is approximated by the posterior mode
- Posterior variances and covariances come from $H^{-1}$: $\text{Var}_{\tilde f}(A^\top\theta) = A^\top H^{-1}A$
- Credible intervals: $\theta_{0j}\pm z_{\alpha/2}\sqrt{(H^{-1})_{jj}}$

This is exactly the Bayesian counterpart of MLE plus Wald standard errors.

## Example: Poisson Rate with a Gamma Prior

$y_1,\ldots,y_n\mid\lambda\sim\text{Pois}(\lambda)$, $\lambda\sim\text{Gamma}(\alpha,\beta)$. The posterior is $\text{Gamma}(\alpha+\sum y_i,\ \beta+n)$ and $P(y)$ is available exactly, so we can measure the Laplace error.

```{r}
#| echo: true
alpha <- 2; beta <- 1
laplace_ml <- function(y) { n <- length(y); s <- sum(y)
  h  <- function(l) -(sum(dpois(y, l, log = TRUE)) + dgamma(l, alpha, beta, log = TRUE))
  op <- optim(mean(y) + 0.5, h, method = "L-BFGS-B", lower = 1e-6, hessian = TRUE)
  c(mode = op$par, log_ml = -op$value + 0.5 * log(2 * pi) - 0.5 * log(op$hessian[1, 1])) }
exact_log_ml <- function(y) { n <- length(y); s <- sum(y)
  alpha * log(beta) - lgamma(alpha) + lgamma(alpha + s) - (alpha + s) * log(beta + n) - sum(lfactorial(y)) }
res <- t(sapply(c(5, 20, 100, 500), function(n) { y <- rpois(n, 3)
  c(n = n, laplace_ml(y), exact = exact_log_ml(y), post_mean = (alpha + sum(y)) / (beta + n)) }))
round(cbind(res, error = res[, "log_ml"] - res[, "exact"]), 4)
```

The error in $\log P(y)$ shrinks roughly like $1/n$, and the mode approaches the posterior mean as the Gamma posterior becomes symmetric.

## Example (Continued)

Exact Gamma posterior (red) and the Laplace normal approximation (green) for $n = 5$ and $n = 100$.

```{r}
#| fig-width: 10
#| fig-height: 4.4
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
for (n in c(5, 100)) { y <- rpois(n, 3); a1 <- alpha + sum(y); b1 <- beta + n
  mode <- (a1 - 1) / b1; H <- (a1 - 1) / mode^2
  curve(dgamma(x, a1, b1), max(0, mode - 4 / sqrt(H)), mode + 4 / sqrt(H), lwd = 3, col = "firebrick",
        xlab = TeX(r"($\lambda$)"), ylab = "density", main = sprintf("n = %d", n))
  curve(dnorm(x, mode, 1 / sqrt(H)), add = TRUE, lwd = 3, col = "forestgreen")
  if (n == 5) legend("topright", bty = "n", lwd = 3, col = c("firebrick", "forestgreen"),
                     legend = c("exact posterior", "Laplace approximation")) }
```

## When Laplace Works Well, and When It Does Not

**Works well**

- Large $n$ relative to $p$, unimodal and roughly symmetric posterior
- Parameters on an unbounded scale: transform first ($\log\sigma$, $\text{logit}\,p$) so the Gaussian is not truncated at a boundary; the Jacobian must then be included in $f$

**Problems**

- Skewed posteriors (small $n$, parameters near a boundary): the mode is a poor centre and tails are wrong
- Multimodal posteriors: only the mode found is represented
- Hierarchical models with many parameters: $p$ grows with $n$, so the $O(1/n)$ error does not vanish; nested Laplace (INLA) integrates out the high-dimensional part analytically

## Summary

- Laplace approximation: expand $h = -\log f$ to second order at its minimum $\theta_0$ and integrate the Gaussian exactly
$$\theta\mid\text{data}\approx N_p\big(\theta_0, [\nabla^2h(\theta_0)]^{-1}\big), 
\int f(\theta)\,d\theta \approx f(\theta_0)(2\pi)^{p/2}|\nabla^2 h(\theta_0)|^{-1/2}$$
- Needs only an optimizer and a Hessian; cost does not grow with the number of integrals
- Gives BIC as a large-$n$ simplification of the marginal likelihood
- Linear $a(\theta)$: mode and $H^{-1}$ suffice; general $a(\theta)$: simulate from the Gaussian or use two Laplace approximations
- Deterministic, fast, and accurate to $O(1/n)$, but blind to skewness and multimodality; Monte Carlo methods (next lectures) remove these limitations at higher computational cost
