---
subtitle: "Random Numbers and Elementary Monte Carlo"
pagetitle: "montecarlo-slides"
author: "Longhai Li"
date: last-modified
format: 
  banner_slides-revealjs:
    slide-level: 3
execute:
  echo: true
  fig-width: 7
  fig-height: 4
filters:
  - shinylive
---

```{r}
#| echo: false
set.seed(812)
```

# Generating Random Numbers

## Pseudo Random Numbers

A computer is deterministic, so it cannot produce truly random numbers. It produces **pseudo random numbers**: a deterministic sequence that *behaves* like iid $\mathrm{Unif}(0,1)$ draws.

A classic generator, the **linear congruential generator (LCG)**:
$$
x_{i+1} = (a\,x_i + c) \bmod m, \qquad u_{i+1} = x_{i+1}/m
$$

- The starting value $x_0$ is the **seed**. The same seed gives the same sequence, which makes simulations reproducible.

- The sequence eventually repeats; a good generator has a very long period and passes statistical tests of uniformity and independence.
- R's default is the Mersenne Twister (period $2^{19937}-1$). All other distributions in R are built from $\mathrm{Unif}(0,1)$ draws.

```{r}
#| code-fold: true
lcg <- function(n, seed, a = 69069, c = 1, m = 2^32) {
  x <- numeric(n); x[1] <- seed
  for (i in 2:n) x[i] <- (a * x[i - 1] + c) %% m
  x / m
}
round(lcg(5, seed = 812), 4)
```
## Pseudo Random Numbers: A Trace

```{r}
#| code-fold: true
#| fig-width: 8
#| fig-height: 3
u <- lcg(50, seed = 812)
plot(u, type = "b", pch = 16, cex = 0.7, ylim = c(0, 1),
     xlab = "i", ylab = expression(u[i]),
     main = "Trace of 50 LCG draws")
abline(h = c(0, 0.5, 1), lty = 3, col = "grey60")
```

The trace shows no visible trend, drift, or periodicity, and the points fill $(0,1)$ without clustering. This is what "behaves like iid uniform" means in practice, though a convincing assessment requires formal tests of uniformity and serial correlation rather than a single plot of 50 values.

## Seeds and Reproducibility

```{r}
set.seed(1); runif(3)
set.seed(1); runif(3)      # identical
runif(3)                   # continues the sequence
```

Always `set.seed()` at the top of a simulation study so results can be reproduced and debugged.

## Example: Generating $\mathrm{Unif}(a,b)$

If $U \sim \mathrm{Unif}(0,1)$ then $X = a + (b-a)\,U \sim \mathrm{Unif}(a,b)$:
$$
P(X \le x) = P\!\left(U \le \frac{x-a}{b-a}\right) = \frac{x-a}{b-a}, \qquad a \le x \le b .
$$

```{r}
u <- runif(10000); x <- 2 + 3 * u          # Unif(2, 5)
hist(x, breaks = 30, freq = FALSE, main = "a + (b - a) U with a = 2, b = 5")
abline(h = 1/3, col = 2, lwd = 2)
```

## Inverting the CDF (Continuous Case)

Let $F_X$ be a continuous CDF with inverse (quantile function) $F_X^{-1}$.

**Theorem.** If $U \sim \mathrm{Unif}(0,1)$, then $X = F_X^{-1}(U)$ has CDF $F_X$.

**Proof.**.
$$
P(X \le x) = P\big(F_X^{-1}(U) \le x\big) = P\big(U \le F_X(x)\big) = F_X(x).
$$

Equivalently, by the change-of-variable formula with $u = F_X(x)$:
$$
f_X(x) = f_U\big(F_X(x)\big)\left\lvert\frac{du}{dx}\right
\rvert = 1 \times f(x).
$$

The converse also holds: if $X \sim F_X$ then $F_X(X) \sim \mathrm{Unif}(0,1)$ (the **probability integral transform**), which we will use for p-values.

## Inverting the CDF: Picture

```{r}
#| echo: false
curve(pexp(x, 1), 0, 5, lwd = 2, xlab = "x", ylab = "F(x)", main = "X = F^{-1}(U)")
u <- c(0.2, 0.6, 0.9); x <- qexp(u, 1)
segments(0, u, x, u, col = 2, lty = 2); segments(x, 0, x, u, col = 4, lty = 2)
points(x, u, pch = 19); axis(2, u, labels = paste0("u=", u), las = 1, cex.axis = 0.8)
axis(1, x, labels = round(x, 2), cex.axis = 0.8)
```

Draw $u$ on the vertical axis, read $x = F^{-1}(u)$ off the horizontal axis. Regions where $F$ is steep (high density) receive more draws.

## Example: Exponential Distribution

$X \sim \mathrm{Exp}(\lambda)$ has $F_X(x) = 1 - e^{-\lambda x}$ for $x > 0$. Solving $u = 1 - e^{-\lambda x}$:
$$
x = F_X^{-1}(u) = -\frac{1}{\lambda}\log(1-u).
$$

Since $1 - U \sim \mathrm{Unif}(0,1)$ as well, we may use $X = -\log(U)/\lambda$.

```{r}
#| code-fold: false
lambda <- 2
u <- runif(10000)
x <- -log(u) / lambda
c(sample_mean = mean(x), true_mean = 1 / lambda)
```

## Example: Exponential Distribution (cont.)

```{r}
#| code-fold: true
#| fig-width: 8
#| fig-height: 4
hist(x, breaks = 40, freq = FALSE, main = "Exp(2) via inverse CDF")
curve(dexp(x, lambda), add = TRUE, col = 2, lwd = 2)
```


## A Transformation for Generating Normal Samples {.smaller}

$\Phi^{-1}$ has no closed form, so the inverse-CDF method is not directly usable. Instead, look at a pair $(Z_1, Z_2) \overset{iid}{\sim} N(0,1)$ in **polar coordinates**:
$$
Z_1 = R\cos\Theta, \qquad Z_2 = R\sin\Theta, \qquad R^2 = Z_1^2 + Z_2^2, \quad \Theta = \operatorname{atan2}(Z_2, Z_1).
$$

The joint density $\frac{1}{2\pi}e^{-(z_1^2+z_2^2)/2}$ depends on $(z_1, z_2)$ only through $r^2 = z_1^2 + z_2^2$, so it is **rotationally symmetric** about the origin. Consequently:

- The angle $\Theta \sim \mathrm{Unif}(0, 2\pi)$, independent of $R$.

- The squared radius $R^2 = Z_1^2 + Z_2^2 \sim \chi^2_2$, and $\chi^2_2$ **is** the exponential distribution with rate $1/2$:
$$
f_{\chi^2_2}(w) = \frac{1}{2}e^{-w/2}, \quad w > 0 \qquad\Longrightarrow\qquad R^2 \sim \mathrm{Exp}(\text{rate} = 1/2).
$$

Both $R^2$ and $\Theta$ **can** be generated by inverting a CDF.

## Illustration: Two Standard Normals in Polar Coordinates

```{r}
#| echo: false
library(latex2exp)
z1 <- rnorm(3000); z2 <- rnorm(3000); r2 <- z1^2 + z2^2; th <- atan2(z2, z1) %% (2 * pi)
layout(matrix(c(1, 1, 2, 3), 2, 2), widths = c(1.1, 1)); par(pty = "s", mar = c(4, 4, 2, 1))
plot(z1, z2, pch = 20, cex = 0.4, col = "grey40", asp = 1, xlab = TeX("$Z_1$"), ylab = TeX("$Z_2$"),
     main = TeX("$(Z_1, Z_2)$ iid N(0,1)"))
q <- sqrt(qchisq(c(0.5, 0.9, 0.99), 2))
for (r in q) symbols(0, 0, circles = r, inches = FALSE, add = TRUE, fg = 2, lwd = 2)
text(q + 0.25, 0.25, c("50%", "90%", "99%"), col = 2, cex = 0.8)
par(pty = "m")
hist(r2, breaks = 40, freq = FALSE, main = TeX("$R^2 = Z_1^2 + Z_2^2$"), xlab = "")
curve(dexp(x, 1/2), add = TRUE, col = 2, lwd = 2); legend("topright", "Exp(rate 1/2)", col = 2, lwd = 2, bty = "n")
hist(th, breaks = 24, freq = FALSE, main = TeX("$\\Theta$"), xlab = ""); abline(h = 1 / (2 * pi), col = 2, lwd = 2)
```

The point cloud is rotationally symmetric: the red circles enclose 50%, 90%, 99% of the points ($R^2 \le \chi^2_2$ quantiles); the angle is uniform; $R^2$ follows $\mathrm{Exp}(1/2)$.

## The Box–Muller Transformation

Reverse the picture: **generate** $R^2$ and $\Theta$, then rotate back to $(Z_1, Z_2)$.

1. $R^2 \sim \mathrm{Exp}(1/2)$ by inverse CDF: $F(w) = 1 - e^{-w/2}$, so $R^2 = -2\log U_1$, i.e. $R = \sqrt{-2\log U_1}$.

2. $\Theta \sim \mathrm{Unif}(0, 2\pi)$: $\Theta = 2\pi U_2$.
3. $Z_1 = R\cos\Theta$, $Z_2 = R\sin\Theta$.

**Theorem (Box–Muller).** If $U_1, U_2 \overset{iid}{\sim} \mathrm{Unif}(0,1)$, then.
$$
Z_1 = \sqrt{-2\log U_1}\,\cos(2\pi U_2), \qquad Z_2 = \sqrt{-2\log U_1}\,\sin(2\pi U_2),
$$
are independent $N(0,1)$. Two uniforms produce two normals. Then $\mu + \sigma Z \sim N(\mu, \sigma^2)$.

## Box–Muller in R

```{r}
box_muller <- function(n) {
  u1 <- runif(n); u2 <- runif(n)
  r  <- sqrt(-2 * log(u1))            # R^2 ~ Exp(1/2), i.e. chi-square(2)
  th <- 2 * pi * u2                   # Theta ~ Unif(0, 2 pi)
  cbind(z1 = r * cos(th), z2 = r * sin(th))
}
z <- box_muller(3000)
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
par(pty = "s"); plot(z, pch = 20, cex = 0.4, col = "grey40", asp = 1, main = "Box-Muller output")
par(pty = "m"); qqnorm(c(z), pch = ".", main = "Normal Q-Q plot of both coordinates"); qqline(c(z), col = 2)
c(mean = mean(z), sd = sd(z), corr = cor(z)[1, 2])
```

# Monte Carlo Integration

## Law of Large Numbers and Central Limit Theorem {.smaller}

Monte Carlo estimates an expectation $\mu = E(g(X))$ by an average of simulated values,
$$
\hat\mu_n = \frac{1}{n}\sum_{i=1}^n g(X_i), \qquad X_i \overset{iid}{\sim} f .
$$

**Theorem (Strong Law of Large Numbers).** If $X_1, X_2, \ldots$ are iid with finite mean $\mu$, then.
$$
P\left(\frac{X_1 + \cdots + X_n}{n} \to \mu\right) = 1 .
$$

The LLN guarantees that the average **converges** to the truth. It does not say how fast.

**Theorem (Central Limit Theorem).** If in addition the variance $\sigma^2$ is finite, then
$$
\frac{\overline X - \mu}{\sigma/\sqrt n} \;\xrightarrow{d}\; N(0,1).
$$

The CLT gives the **error bound**: the Monte Carlo error is of order $\sigma/\sqrt n$.

## Monte Carlo Error Bound from the CLT

From (5.2), with confidence level 95%,
$$
P\!\left(\lvert\overline X - \mu
\rvert \le 1.96\,\frac{\sigma}{\sqrt n}\right) \approx 0.95,
$$
so a 95% interval for $\mu$ is
$$
\overline X \pm 1.96\,\frac{\sigma}{\sqrt n}, \qquad \sigma \text{ replaced by the sample SD } s .
$$

- Halving the error requires **four times** as many samples; each extra decimal digit of accuracy costs a **100-fold** increase in $n$.

- The rate $n^{-1/2}$ does not depend on the dimension of $X$, which is why Monte Carlo beats numerical quadrature in high dimensions.

## Demonstration of the LLN

Running averages of $X_i \sim \mathrm{Exp}(1)$ ($\mu = 1$) for three independent sequences:

```{r}
n <- 5000
plot(NULL, xlim = c(1, n), ylim = c(0.5, 1.6), log = "x",
     xlab = "n (log scale)", ylab = "running average", main = "LLN: Exp(1)")
for (k in 1:3) lines(cumsum(rexp(n)) / (1:n), col = k + 1)
abline(h = 1, lty = 2)
```

## Demonstration of the CLT

For each of $N = 5000$ repetitions, draw $n$ values from $\mathrm{Exp}(1)$ and standardize the mean, $(\overline X - 1)/(1/\sqrt n)$:

```{r}
clt_demo <- function(n, N = 5000) replicate(N, (mean(rexp(n)) - 1) * sqrt(n))
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
for (n in c(2, 10, 100)) {
  hist(clt_demo(n), breaks = 40, freq = FALSE, xlim = c(-4, 4), main = paste("n =", n), xlab = "")
  curve(dnorm, add = TRUE, col = 2, lwd = 2)
}
```

The skewed exponential becomes approximately normal as $n$ grows.

## An Example: Estimating $\pi$ by Monte Carlo

Let $(X, Y)$ be uniform on the square $S = (-1,1)\times(-1,1)$, and let $C = \{(x,y): x^2 + y^2 \le 1\}$ be the unit disk. Then.
$$
P\big((X,Y) \in C\big) = \frac{\text{area}(C)}{\text{area}(S)} = \frac{\pi}{4}.
$$

Define $Z = 4\, I\big((X,Y)\in C\big)$, so $E(Z) = \pi$ and
$$
\mathrm{Var}(Z) = 16 \cdot \frac{\pi}{4}\left(1 - \frac{\pi}{4}\right) \approx 2.70, \qquad \mathrm{SD}(Z) \approx 1.64 .
$$

Drawing $(X_i, Y_i)$ and averaging $Z_i$:
$$
\hat\pi_n = \frac{1}{n}\sum_{i=1}^n Z_i = 4 \times \frac{\# \{i: X_i^2 + Y_i^2 \le 1\}}{n}.
$$

## Visualizing the Method 

```{r}
n <- 2000; x <- runif(n, -1, 1); y <- runif(n, -1, 1); inside <- x^2 + y^2 <= 1
par(pty = "s")
plot(x, y, col = ifelse(inside, "steelblue", "grey60"), pch = 20, cex = 0.6,
     main = sprintf("n = %d, proportion inside = %.3f, pi-hat = %.3f", n, mean(inside), 4 * mean(inside)))
curve(sqrt(1 - x^2), -1, 1, add = TRUE, lwd = 2); curve(-sqrt(1 - x^2), -1, 1, add = TRUE, lwd = 2)
```

## The Monte Carlo Estimate and Its Error

```{r}
est_pi <- function(n) {
  z <- 4 * (runif(n, -1, 1)^2 + runif(n, -1, 1)^2 <= 1)
  c(estimate = mean(z), se = sd(z) / sqrt(n))
}
r <- est_pi(10000)
r
r["estimate"] + c(-1.96, 1.96) * r["se"]    # 95% interval; contains pi = 3.14159?
```

The standard error $s/\sqrt n$ is computed from the **same** simulated values — no extra cost. Always report it with a Monte Carlo estimate.

## Results by Sample Size

```{r}
ns <- 10^(2:6)
res <- t(sapply(ns, est_pi))
data.frame(n = ns, estimate = round(res[, 1], 5), error = round(res[, 1] - pi, 5),
           se = signif(res[, 2], 3), theory_se = signif(1.64 / sqrt(ns), 3))
```

- The observed error is typically within about $2$ standard errors.

- SE shrinks by $\sqrt{10} \approx 3.2$ for each 10-fold increase in $n$, as the CLT predicts.

## Convergence of $\hat\pi_n$

```{r}
#| label: pi-convergence
#| fig-width: 10
#| fig-height: 4
set.seed(1)
n <- 1e5; z <- 4 * (runif(n, -1, 1)^2 + runif(n, -1, 1)^2 <= 1)
k <- 1:n; pihat <- cumsum(z) / k
par(mar = c(4, 4.5, 0.5, 0.5), mgp = c(2.5, 0.8, 0))
plot(k, pihat, type = "l", log = "x", ylim = c(2, 4),
     xlab = "n (log scale)", ylab = expression(hat(pi)[n]))
lines(k, pi + 1.96 * 1.64 / sqrt(k), lty = 2, col = 2)
lines(k, pi - 1.96 * 1.64 / sqrt(k), lty = 2, col = 2)
abline(h = pi, col = 4)
```

The red band is $\pi \pm 1.96\,\sigma/\sqrt n$; the running estimate stays inside it most of the time.

# Evaluating a Point Estimator by Simulation

## Point Estimators

A point estimator $\hat\theta = \hat\theta(X_1, \ldots, X_n)$ is judged by its **mean squared error**
$$
\mathrm{MSE}(\hat\theta) = E\big[(\hat\theta - \theta)^2\big] = \mathrm{Var}(\hat\theta) + \big[\mathrm{Bias}(\hat\theta)\big]^2 .
$$

MSE is an expectation, so it is estimated by Monte Carlo: simulate $N$ independent data sets from a model with **known** $\theta$,
$$
\hat\theta^{(j)} = \hat\theta\big(X_1^{(j)}, \ldots, X_n^{(j)}\big), \qquad j = 1, \ldots, N,
$$
and compute
$$
\widehat{\mathrm{MSE}} = \frac{1}{N}\sum_{j=1}^N \big(\hat\theta^{(j)} - \theta\big)^2, \qquad
\widehat{\mathrm{Bias}} = \overline{\hat\theta} - \theta, \qquad
\widehat{\mathrm{Var}} = \frac{1}{N}\sum_{j=1}^N \big(\hat\theta^{(j)} - \overline{\hat\theta}\big)^2 .
$$

Two sample sizes appear: $n$ (data per replicate, fixed by the problem) and $N$ (replicates, chosen large enough that the Monte Carlo error is negligible).

## Example: Mean vs. Median vs. Trimmed Mean

Estimating the centre $\theta = 0$ of a symmetric distribution, $n = 20$, $N = 5000$ replicates.

```{r}
eval_est <- function(rdist, n = 20, N = 5000) {
  est <- replicate(N, { x <- rdist(n); c(mean = mean(x), median = median(x), trim10 = mean(x, trim = 0.1)) })
  rbind(bias = rowMeans(est), var = apply(est, 1, var), mse = rowMeans(est^2))
}
round(eval_est(rnorm), 4)                          # normal data
round(eval_est(function(n) rt(n, df = 3)), 4)      # heavy-tailed t3 data
```

No estimator is best everywhere: the mean wins under normality, the trimmed mean and median win under heavy tails.


# Evaluating and Conducting Hypothesis Tests

## Hypothesis Tests and the Two Types of Error

A test rejects $H_0$ when a statistic $T$ exceeds a critical value $c_\alpha$.

|          | reject: $T > c_\alpha$      | do not reject: $T \le c_\alpha$      |
|----------|---------------------|-----------------------|
| $H_0$ true    | Type I error (false positive)   | true negative                    |
| $H_1$ true    | true positive                  | Type II error (false negative)    |
$$
\text{size} = P(T > c_\alpha \mid H_0) = \alpha, \qquad \text{power} = P(T > c_\alpha \mid H_1) = 1 - \beta .
$$

Both are probabilities — hence expectations of indicators — so both can be **estimated by Monte Carlo** once we can simulate data under $H_0$ and $H_1$.

## Estimating False and True Positive Rate of $T$ {.tiny}

For a test that rejects when $T > c$, simulate $N$ data sets **under $H_0$** and $N$ data sets **under $H_1$**, and compute $T$ on each:

| replicate | $T^{(j)}_{H_0}$ | $I(T^{(j)}_{H_0} > c)$ | $T^{(j)}_{H_1}$ | $I(T^{(j)}_{H_1} > c)$ |
|:---:|:---:|:---:|:---:|:---:|
| 1 | $T^{(1)}_{H_0}$ | 0 | $T^{(1)}_{H_1}$ | 1 |
| 2 | $T^{(2)}_{H_0}$ | 1 | $T^{(2)}_{H_1}$ | 1 |
| $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ |
| $N$ | $T^{(N)}_{H_0}$ | 0 | $T^{(N)}_{H_1}$ | 0 |
$$
\widehat{\text{FPR}}(c) = \frac{1}{N}\sum_{j=1}^N I\big(T^{(j)}_{H_0} > c\big), \qquad
\widehat{\text{TPR}}(c) = \frac{1}{N}\sum_{j=1}^N I\big(T^{(j)}_{H_1} > c\big).
$$

- $\widehat{\text{FPR}}$ estimates the size (type I error rate); $\widehat{\text{TPR}}$ estimates the power, $1 - \beta$.

- To get size $\alpha$, take $\hat c_\alpha$ = the $(1-\alpha)$ sample quantile of $T^{(1)}_{H_0}, \ldots, T^{(N)}_{H_0}$, which replaces the (often unknown) null distribution of $T$ by its **empirical** distribution.

## P-value: A PIT transfromation of $T$ under $H_0$

The **p-value** is the probability, under $H_0$, of a statistic at least as extreme as the observed one:
$$
\text{p-value}(t) = P(T \ge t \mid H_0) = 1 - F_{H_0}(t),
$$
the **survival function** of $T$ under $H_0$ evaluated at $t^{obs}$. Rejecting when $\text{p-value}(t^{obs}) < \alpha$ is the same as rejecting when $t^{obs} > c_\alpha$.

```{r}
#| echo: false
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
t0 <- 1.8; curve(dchisq(x, 3), 0, 12, lwd = 2, main = "density of T | H0", xlab = "t", ylab = "")
xx <- seq(t0, 12, length = 100); polygon(c(t0, xx, 12), c(0, dchisq(xx, 3), 0), col = "lightblue", border = NA)
abline(v = t0, col = 2); text(t0 + 3, 0.15, "pv = shaded area")
curve(1 - pchisq(x, 3), 0, 12, lwd = 2, main = "survival function 1 - F(t)", xlab = "t", ylab = "")
abline(v = t0, col = 2); abline(h = 1 - pchisq(t0, 3), lty = 2)
```

## Uniformity of $\text{p-value}(T)$ under $H_0$

**Theorem.** If $T \mid H_0$ has continuous CDF $F_{H_0}$, then $\text{p-value}(T) = 1 - F_{H_0}(T) \sim \mathrm{Unif}(0,1)$ under $H_0$.

**Proof.** For $t \in (0,1)$,
$$
\begin{aligned}
P\big(\text{p-value}(T) < t\big) &= P\big(1 - F_{H_0}(T) < t\big) = P\big(F_{H_0}(T) > 1 - t\big) \\
&= P\big(T > F_{H_0}^{-1}(1-t)\big) = 1 - F_{H_0}\big(F_{H_0}^{-1}(1-t)\big) = t .
\end{aligned}
$$

**Consequences**:

- $P(\text{p-value} < \alpha \mid H_0) = \alpha$ **exactly**: rejecting when $\text{p-value} < \alpha$ has size $\alpha$.

- A histogram of simulated p-values under $H_0$ should be flat — the standard **test of a test**.

## Example: P-value Distribution of `t.test` 

One-sample $t$-test of $H_0: \mu = 0$, $n = 20$, $N = 5000$:

```{r}
pv_sim <- function(mu, n = 20, N = 5000) replicate(N, t.test(rnorm(n, mu))$p.value)
pv0 <- pv_sim(0); pv1 <- pv_sim(0.5)
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
hist(pv0, breaks = 20, main = expression(H[0]: mu == 0), xlab = "p-value"); abline(h = 250, col = 2, lty = 2)
hist(pv1, breaks = 20, main = expression(H[1]: mu == 0.5), xlab = "p-value")
```

Under $H_0$ the p-values are uniform; under $H_1$ they pile up near 0.

## Monte Carlo Estimates of Size and Power
$$
\widehat{\text{size}} = \hat P(\text{p-value} < \alpha \mid H_0) = \frac{1}{N}\sum_{j=1}^N I\big(\text{p-value}^{(j)}_{H_0} < \alpha\big),
$$
$$
\widehat{\text{power}} = \hat P(\text{p-value} < \alpha \mid H_1) = \frac{1}{N}\sum_{j=1}^N I\big(\text{p-value}^{(j)}_{H_1} < \alpha\big)
$$

```{r}
c(size = mean(pv0 < 0.05), power = mean(pv1 < 0.05),
  exact_power = power.t.test(n = 20, delta = 0.5, sd = 1, type = "one.sample")$power)
```

Each estimate is a proportion, so its Monte Carlo standard error is $\sqrt{\hat p(1-\hat p)/N}$ — about $0.003$ for the size with $N = 5000$.

## Monte Carlo for Calculating (Not Evaluating) a P-value

When $F_{H_0}$ is unknown, estimate $\text{p-value}(t^{obs})$ by simulating $T$ under $H_0$:
$$
\widehat{\text{p-value}}(t^{obs}) = \frac{1 + \sum_{j=1}^N I\big(T^{(j)} \ge t^{obs}\big)}{N + 1}.
$$

(The "+1" counts the observed statistic as one draw from $H_0$; it keeps the estimate above 0 and the test exactly valid.)

Example: test $H_0: \mu = 0$ for $n = 15$ observations using $T = \lvert\overline X
\rvert/(s/\sqrt n)$, but suppose we did not know the $t$-distribution.

```{r}
x <- rnorm(15, mean = 0.6); tobs <- abs(mean(x)) / (sd(x) / sqrt(15))
Tnull <- replicate(5000, { y <- rnorm(15); abs(mean(y)) / (sd(y) / sqrt(15)) })
c(mc_pv = (1 + sum(Tnull >= tobs)) / 5001, exact_pv = 2 * pt(-tobs, df = 14))
```

## Example 1: A Permutation Test for the Correlation Coefficient

### Setting and Procedure

Data $(X_i, Y_i)$, $i = 1, \ldots, n$. Test $H_0: X \perp Y$ (which implies $\mathrm{corr}(X,Y) = 0$) with $T = \lvert\hat\rho
\rvert$.

Under $H_0$, **every pairing of the $X$ values with the $Y$ values is equally likely**, so the null distribution of $T$ is obtained by re-computing $T$ after randomly permuting $Y$:

1. Compute $\rho^{obs} = \hat\rho(X, Y)$.

2. For $j = 1, \ldots, m$: permute $Y$ to get $Y^{(j)}$, compute $\rho^{(j)} = \hat\rho(X, Y^{(j)})$.
3. $\widehat{\text{p-value}} = \dfrac{1 + \# \{j : \lvert\rho^{(j)}
\rvert \ge \lvert\rho^{obs}
\rvert\}}{m+1}$.

No normality assumption is needed; the null distribution is generated **from the data itself**.

### Replicated Datasets by Permutation

$X$ is held fixed; each of the $m$ permutations re-shuffles the $Y$ column (shown for $n = 5$):

|  | $X$ | observed $Y$ | $Y^{(1)}$ | $Y^{(2)}$ | $\cdots$ | $Y^{(m)}$ |
|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| $i = 1$ | $x_1$ | $y_1$ | $y_3$ | $y_5$ | $\cdots$ | $y_2$ |
| $i = 2$ | $x_2$ | $y_2$ | $y_1$ | $y_4$ | $\cdots$ | $y_5$ |
| $i = 3$ | $x_3$ | $y_3$ | $y_5$ | $y_1$ | $\cdots$ | $y_4$ |
| $i = 4$ | $x_4$ | $y_4$ | $y_2$ | $y_3$ | $\cdots$ | $y_1$ |
| $i = 5$ | $x_5$ | $y_5$ | $y_4$ | $y_2$ | $\cdots$ | $y_3$ |
| **statistic** |  | $\rho^{obs}$ | $\rho^{(1)}$ | $\rho^{(2)}$ | $\cdots$ | $\rho^{(m)}$ |
| **as extreme?** |  |  | $I(\lvert\rho^{(1)}\rvert \ge \lvert\rho^{obs}\rvert)$ | $I(\lvert\rho^{(2)}\rvert \ge \lvert\rho^{obs}\rvert)$ | $\cdots$ | $I(\lvert\rho^{(m)}\rvert \ge \lvert\rho^{obs}\rvert)$ |

$\rho^{(1)}, \ldots, \rho^{(m)}$ are draws from the null distribution of $\hat\rho$. The p-value is $1$ plus the sum of the last row, divided by $m + 1$.

### Implementation in R

```{r}
#| label: perm-cor
#| code-fold: false
perm_cor_test <- function(x, y, m = 2000) {
  robs <- cor(x, y)
  rperm <- replicate(m, cor(x, sample(y)))
  list(rho = robs, pv = (1 + sum(abs(rperm) >= abs(robs))) / (m + 1), rperm = rperm)
}
```

### Evaluation Under $H_0$ and $H_1$ (1) {.smaller}

Generate non-normal $(X, Y)$: independent under $H_0$, and $Y = 0.4X + \epsilon$ under $H_1$. Check that the permutation p-values are uniform with size $\alpha$ under $H_0$, and estimate the power under $H_1$.

| dataset | data under $H_0$ (or $H_1$) | $\rho^{obs}$ | permutation test | p-value | reject? |
|:---:|:---:|:---:|:-----:|:---:|:-----:|
| $k = 1$ | $(X^{[1]}, Y^{[1]})$ | $\rho^{obs}_{1}$ | $m$ permutations of $Y^{[1]}$ | $pv_1$ | $I(pv_1 < \alpha)$ |
| $k = 2$ | $(X^{[2]}, Y^{[2]})$ | $\rho^{obs}_{2}$ | $m$ permutations of $Y^{[2]}$ | $pv_2$ | $I(pv_2 < \alpha)$ |
| $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ |
| $k = N$ | $(X^{[N]}, Y^{[N]})$ | $\rho^{obs}_{N}$ | $m$ permutations of $Y^{[N]}$ | $pv_N$ | $I(pv_N < \alpha)$ |
| **summary** |  |  |  | histogram | $\widehat{\text{size}}$ (or $\widehat{\text{power}}$) $= \frac{1}{N}\sum_k I(pv_k < \alpha)$ |
### Evaluation Under $H_0$ and $H_1$ (2)



Below, $N = 500$ and $m = 500$.

```{r}
#| label: cal-perm-cor-h0
#| cache: true
#| code-fold: true
sim_cor <- function(n = 30, b = 0) { x <- rexp(n); list(x = x, y = b * x + rexp(n)) }
eval_perm_cor <- function(b, N = 500) replicate(N, {
  d <- sim_cor(b = b)
  perm_cor_test(d$x, d$y, m = 500)$pv
})
pv_cor0 <- eval_perm_cor(b = 0)      # H0: independent
pv_cor1 <- eval_perm_cor(b = 0.4)    # H1: dependent

hist_reject <- function(pv, main, alpha = 0.05) {
  br <- seq(0, 1, by = alpha)                  # first bar = rejection region
  ymax <- max(hist(pv, breaks = br, plot = FALSE)$counts)
  hist(pv, breaks = br, main = main, xlab = "p-value",
       ylim = c(0, 1.3 * ymax),                # headroom for the legend
       col = ifelse(br[-1] <= alpha, "red", "grey80"))
  abline(v = alpha, col = "red", lwd = 2, lty = 2)
  legend("topright", bty = "n", fill = c("red", NA), border = c("black", NA),
         lty = c(NA, 2), col = c(NA, "red"), lwd = c(NA, 2),
         legend = c(sprintf("rejected: %.3f", mean(pv < alpha)),
                    as.expression(bquote(alpha == .(alpha)))))
}
par(mfrow = c(1, 2))
hist_reject(pv_cor0, "permutation p-values under H0")
hist_reject(pv_cor1, "permutation p-values under H1")
```

Under $H_0$ the permutation p-values are uniform on $[0,1]$ and the size is close to 0.05, with no distributional assumption on $(X, Y)$. Under $H_1$ they pile up near 0; the proportion below 0.05 is the power.



## Example 2: Testing for Post-selection Regression

### Setting

Response $Y$ and $p = 20$ candidate predictors $X_1, \ldots, X_{20}$, $n = 50$. We want to test $H_0$: $Y$ is unrelated to **all** of the $X$'s.

A common practice:

1. Pick the predictor $X_{k^*}$ most correlated with $Y$.

2. Fit $Y \sim X_{k^*}$ and report the usual $t$-test p-value for its slope.

**Question.** Is this a valid test of $H_0$? Does it have size $\alpha$?

### A Test With Post-Selection Fitting

```{r}
#| label: test-lm-sel
test_lm_sel <- function(y, X) {
  k <- which.max(abs(cor(X, y)))
  summary(lm(y ~ X[, k]))$coefficients[2, 4]    # t-test p-value of selected slope
}
```

The $t$-test p-value assumes $X_{k^*}$ was **chosen in advance**. Selecting the best of 20 predictors *after* looking at the data means that, even under $H_0$, $\max_k \lvert\hat\rho_k
\rvert$ tends to be large. The reported p-value ignores this selection.

### Simulation for Evaluating `test_lm_sel`

Simulate $N = 2000$ datasets with $Y = \beta X_1 + \epsilon$: $\beta = 0$ under $H_0$, and $\beta = 0.5$ under $H_1$.

```{r}
#| label: test-lm-sel-h0
sim_lm <- function(n = 50, p = 20, beta = 0) {
  X <- matrix(rnorm(n * p), n, p); list(y = beta * X[, 1] + rnorm(n), X = X)
}
pv_sel0 <- replicate(2000, { d <- sim_lm(beta = 0);   test_lm_sel(d$y, d$X) })
pv_sel1 <- replicate(2000, { d <- sim_lm(beta = 0.5); test_lm_sel(d$y, d$X) })
par(mfrow = c(1, 2))
hist_reject(pv_sel0, "test_lm_sel p-values under H0")
hist_reject(pv_sel1, "test_lm_sel p-values under H1")
```

The true size is far above 0.05: `test_lm_sel` rejects a true $H_0$ far too often. Under $H_0$ its p-value is **not** uniform, which is a general diagnostic for an invalid test. Its high rejection rate under $H_1$ is not real power, since the test does not have size $\alpha$.

### Permutation Test for Post-selection Regression

Fix: repeat the **whole procedure** — select the most correlated variable, fit the linear model, report its p-value — on permuted responses, and compare the observed p-value with the permutation distribution of p-values.

1. Compute $p^{obs}$ = p-value of `test_lm_sel` on $(Y, X)$.

2. For $j = 1, \ldots, m$: permute $Y$, compute $p^{(j)}$ = p-value of `test_lm_sel` on $(Y^{(j)}, X)$ (**re-doing the selection** each time).
3. $\widehat{\text{p-value}} = \dfrac{1 + \# \{j : p^{(j)} \le p^{obs}\}}{m+1}$, the proportion of permuted p-values at least as small as the observed one.


### Tree Diagram: Re-selecting the Predictor with Permuated $Y$

$k^*(Y) = \arg\max_k \lvert\hat\rho(X_k, Y)\rvert$ is the index of the column of $X$ most correlated with the response. It is a function of the response, so each permuted $Y^{(j)}$ gets its **own** selected predictor $X_{k^*(Y^{(j)})}$:

```{r}
#| label: perm-lm-sel-tree
#| echo: false
#| fig-width: 16
#| fig-height: 7.5
par(mar = c(0, 0, 0, 0))
plot(NA, xlim = c(0, 13.2), ylim = c(0.2, 6.3), axes = FALSE, xlab = "", ylab = "")
box_at <- function(x, y, lab, col = "grey95", w = 1, h = 0.36) {
  rect(x - w, y - h, x + w, y + h, col = col, border = "grey40", lwd = 1.5)
  text(x, y, lab, cex = 1.25)
}
arr <- function(x0, y0, x1, y1) arrows(x0, y0, x1, y1, length = 0.1, lwd = 1.5, col = "grey30")
xs <- c(1.05, 3.6, 6.2, 8.8, 11.9)              # data, response, selection, p-value, compare
ys <- c(5, 3.8, 2.6, 1.55, 0.6)                 # branches: observed, 1, 2, ..., m
text(xs, 6.15, c("data", "response", "selected predictor", "p-value of slope", "as extreme?"),
     font = 2, cex = 1.15)
text(mean(xs[2:3]), 5.5, "max |cor|", cex = 1, col = "grey30")
text(mean(xs[3:4]), 5.5, "lm t-test", cex = 1, col = "grey30")
box_at(xs[1], 2.8, expression(atop(X ~ "fixed", (n %*% 20))), col = "lightyellow", h = 0.55)
resp <- list(quote(Y), quote(Y^(1)), quote(Y^(2)), NULL, quote(Y^(m)))
sel  <- list(quote(X[k^"*"*(Y)]), quote(X[k^"*"*(Y^(1))]), quote(X[k^"*"*(Y^(2))]), NULL,
             quote(X[k^"*"*(Y^(m))]))
pv   <- list(quote(p^obs), quote(p^(1)), quote(p^(2)), NULL, quote(p^(m)))
cmp  <- list(NULL, quote(I(p^(1) <= p^obs)), quote(I(p^(2) <= p^obs)), NULL, quote(I(p^(m) <= p^obs)))
for (b in c(1, 2, 3, 5)) {
  fill <- if (b == 1) "lightblue" else "grey95"
  arr(xs[1] + 1, 2.8, xs[2] - 1, ys[b])
  box_at(xs[2], ys[b], as.expression(resp[[b]]), col = fill)
  arr(xs[2] + 1, ys[b], xs[3] - 1, ys[b])
  box_at(xs[3], ys[b], as.expression(sel[[b]]), col = fill)
  arr(xs[3] + 1, ys[b], xs[4] - 1, ys[b])
  box_at(xs[4], ys[b], as.expression(pv[[b]]), col = fill)
  if (b > 1) {
    arr(xs[4] + 1, ys[b], xs[5] - 1.2, ys[b])
    box_at(xs[5], ys[b], as.expression(cmp[[b]]), col = "mistyrose", w = 1.2)
  }
}
text(xs[2:5], ys[4] + 0.1, ":", cex = 2.2, font = 2)
text(1.05, 4.3, "observed", col = "steelblue", cex = 1.05)
text(1.05, 1.3, "permute Y", col = "grey30", cex = 1.05)
```

Keeping $X_{k^*(Y)}$ fixed and only permuting $Y$ would **not** reproduce the selection step, and the resulting test would still be invalid.

### Implementation in R

```{r}
#| label: perm-lm-sel
#| code-fold: false
perm_lm_sel <- function(y, X, m = 500) {
  pobs <- test_lm_sel(y, X)
  pperm <- replicate(m, test_lm_sel(sample(y), X))
  (1 + sum(pperm <= pobs)) / (m + 1)
}
```

### Evaluation Under $H_0$ and $H_1$ (1) {.smaller}

Simulate $(Y, X)$ from `sim_lm`: $\beta = 0$ under $H_0$, and $\beta = 0.5$ under $H_1$. Check that the permutation p-values are uniform with size $\alpha$ under $H_0$, and estimate the power under $H_1$.

| dataset | data under $H_0$ (or $H_1$) | $p^{obs}$ | permutation test | p-value | reject? |
|:---:|:---:|:---:|:-----:|:---:|:-----:|
| $k = 1$ | $(Y^{[1]}, X^{[1]})$ | $p^{obs}_{1}$ | $m$ permutations of $Y^{[1]}$ | $pv_1$ | $I(pv_1 < \alpha)$ |
| $k = 2$ | $(Y^{[2]}, X^{[2]})$ | $p^{obs}_{2}$ | $m$ permutations of $Y^{[2]}$ | $pv_2$ | $I(pv_2 < \alpha)$ |
| $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ |
| $k = N$ | $(Y^{[N]}, X^{[N]})$ | $p^{obs}_{N}$ | $m$ permutations of $Y^{[N]}$ | $pv_N$ | $I(pv_N < \alpha)$ |
| **summary** |  |  |  | histogram | $\widehat{\text{size}}$ (or $\widehat{\text{power}}$) $= \frac{1}{N}\sum_k I(pv_k < \alpha)$ |

### Evaluation Under $H_0$ and $H_1$ (2)

Below, $N = 300$ and $m = 500$.

```{r}
#| label: perm-lm-sel-size-power
#| cache: true
#| code-fold: true
eval_perm_lm_sel <- function(beta, N = 300) replicate(N, {
  d <- sim_lm(beta = beta)
  perm_lm_sel(d$y, d$X, m = 500)
})
pv_perm0 <- eval_perm_lm_sel(beta = 0)      # H0: Y unrelated to X
pv_perm1 <- eval_perm_lm_sel(beta = 0.5)    # H1: Y depends on X1
par(mfrow = c(1, 2))
hist_reject(pv_perm0, "perm_lm_sel p-values under H0")
hist_reject(pv_perm1, "perm_lm_sel p-values under H1")
```

Under $H_0$ the permutation p-values are uniform and the size is close to 0.05; under $H_1$ they pile up near 0, and the proportion below 0.05 is an honest power.


## Summary

- All simulation starts from pseudo-random $\mathrm{Unif}(0,1)$ draws; other distributions come from **inverse-CDF** or **transformations** (Box–Muller).

- The LLN justifies Monte Carlo averages; the CLT gives their error, $\sigma/\sqrt n$, which is reported as a standard error.
- Bias, variance and MSE of estimators, and size and power of tests, are expectations — simulate with known truth and average.

- P-values are uniform under $H_0$: a flat histogram of simulated p-values is the basic test of a test.
- When the null distribution is unknown, generate it by simulation or by **permutation**; the permuted statistic must reproduce the **entire** data-analysis procedure.
