---
subtitle: "Optimization for Maximum Likelihood Estimation"
pagetitle: "optimization-slides"
author: "Longhai Li"
format: banner_slides-revealjs
code-fold: true
execute:
  echo: true
  fig-width: 7
  fig-height: 4
---

```{r}
#| echo: false
set.seed(812)
```

# Review: Likelihood Theory and MLE

## Likelihood Function

A statistical model specifies the probability (density) of data $D$ given a parameter $\theta$: $f(D \mid \theta)$.

- **As a function of $D$** for fixed $\theta$: the sampling distribution.

- **As a function of $\theta$** for fixed (observed) $D$: the **likelihood function**.
$$
L(\theta; D) = f(D \mid \theta).
$$

The likelihood measures how well each value of $\theta$ *explains* the observed data. For iid data $D = (x_1, \ldots, x_n)$,
$$
L(\theta; D) = \prod_{i=1}^n f(x_i \mid \theta), \qquad \ell(\theta) = \log L(\theta; D) = \sum_{i=1}^n \log f(x_i \mid \theta).
$$

We work with the **log-likelihood** $\ell(\theta)$ to avoid underflow of the product and because sums are easier to differentiate.

## Maximum Likelihood (Minimum Loss) Estimator
$$
\hat\theta_{\mathrm{MLE}} = \arg\max_\theta L(\theta; D) = \arg\max_\theta \ell(\theta)
= \arg\min_\theta \mathcal{L}(\theta), \qquad \mathcal{L}(\theta) = -\ell(\theta).
$$

Every algorithm in this lecture **minimizes** the negative log-likelihood
$\mathcal{L}$ (`nll` in code). Three quantities recur:

- the **gradient** $g(\theta) = \mathcal{L}'(\theta) = -\ell'(\theta)$, whose root locates the minimum;
- the **observed information** $J(\theta) = g'(\theta) = \mathcal{L}''(\theta) = -\ell''(\theta)$;
- the **expected information** $nI(\theta) = E_D[\,g'(\theta; D)\,]$.

The classical **score** is $\ell'(\theta) = -g(\theta)$; we work with $g$ throughout so that
every method is a minimization.

## Example: Bernoulli Likelihood

$Y_1, Y_2, Y_3 \overset{iid}{\sim} \mathrm{Bern}(\theta)$, observed $y = (1, 1, 0)$.

:::: {.columns}

::: {.column width="50%"}
| $\theta$ | $P(1, 1, 0 \mid \theta) = \theta^2(1-\theta)$ |
|:---:|:---|
| $0$ | $0$ |
| $1/2$ | $1/8 = 0.1250$ |
| $2/3$ | $4/27 = 0.1481$ **(max)** |
| $3/4$ | $9/64 = 0.1406$ |
| $1$ | $0$ |
:::

::: {.column width="50%"}
```{r}
#| echo: false
curve(x^2 * (1 - x), 0, 1, xlab = expression(theta), ylab = "likelihood", lwd = 2); abline(v = 2/3, col = 2, lty = 2)
```
:::

::::

## Example: Bernoulli Likelihood {.sub}


$P(y_i \mid \theta) = \theta^{y_i}(1-\theta)^{1-y_i}$, so with $n_1 = \sum y_i$ and $n_0 = n - n_1$,
$$
L(\theta) = \theta^{n_1}(1-\theta)^{n_0}, \qquad \ell(\theta) = n_1\log\theta + n_0\log(1-\theta).
$$

The negative log-likelihood and its gradient are
$$
\mathcal{L}(\theta) = -n_1\log\theta - n_0\log(1-\theta), \qquad
g(\theta) = \mathcal{L}'(\theta) = \frac{n_0}{1-\theta} - \frac{n_1}{\theta} .
$$

The MLE solves $g(\theta) = 0 \iff \hat\theta = \dfrac{n_1}{n} = \overline y$, and the curvature
$J(\theta) = g'(\theta) = \dfrac{n_1}{\theta^2} + \dfrac{n_0}{(1-\theta)^2} > 0$ confirms a **minimum** of $\mathcal{L}$.

## Example: Normal Mean with Known Variance

$X_1, \ldots, X_n \overset{iid}{\sim} N(\mu, \sigma_0^2)$:
$$
\mathcal{L}(\mu) = \frac{n}{2}\log(2\pi\sigma_0^2) + \frac{1}{2\sigma_0^2}\sum_{i=1}^n (x_i - \mu)^2
= C + \frac{n}{2\sigma_0^2}(\mu - \overline x)^2,
$$
using $\sum(x_i - \mu)^2 = \sum(x_i - \overline x)^2 + n(\overline x - \mu)^2$.
$$
g(\mu) = \frac{n}{\sigma_0^2}(\mu - \overline x) = 0 \implies \hat\mu = \overline x, \qquad J(\mu) = g'(\mu) = \frac{n}{\sigma_0^2}.
$$

- $\mathcal{L}$ is exactly **quadratic** in $\mu$; the gradient $g$ is **linear**; the curvature $J$ is **constant** and grows with $n$.

- $J = \mathcal{L}'' = -\ell''$ measures how sharply the data pin down $\mu$: it is the **Fisher information**.

## Negative Log-likelihood, Gradient and Curvature

```{r}
#| echo: false
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
x20 <- rnorm(20, 1); x50 <- rnorm(50, 1); mu <- seq(0, 2, length = 200)
nll <- function(mu, x) sapply(mu, function(m) -sum(dnorm(x, m, log = TRUE)))
plot(mu, nll(mu, x20) - min(nll(mu, x20)), type = "l", lwd = 2, ylab = expression(L(mu) - min), main = "negative log-likelihood")
lines(mu, nll(mu, x50) - min(nll(mu, x50)), lwd = 2, col = 2); legend("bottomright", c("n = 20", "n = 50"), col = 1:2, lwd = 2, bty = "n")
plot(mu, 20 * (mu - mean(x20)), type = "l", lwd = 2, ylab = expression(g(mu)), main = "gradient"); lines(mu, 50 * (mu - mean(x50)), lwd = 2, col = 2); abline(h = 0, lty = 3)
plot(mu, rep(20, 200), type = "l", lwd = 2, ylim = c(0, 60), ylab = expression(J(mu)), main = "curvature"); lines(mu, rep(50, 200), lwd = 2, col = 2)
```

The gradient crosses zero where $\mathcal{L}$ is minimized; larger $n$ makes $\mathcal{L}$ more sharply curved, $g$ steeper, and $J$ larger.

## Example: Cauchy Location

$X_1, \ldots, X_n \overset{iid}{\sim} \mathrm{Cauchy}(\theta)$, $f(x \mid \theta) = \dfrac{1}{\pi\,[1 + (x-\theta)^2]}$:
$$
\mathcal{L}(\theta) = n\log\pi + \sum_{i=1}^n \log\!\big(1 + (x_i - \theta)^2\big), \qquad
g(\theta) = \sum_{i=1}^n \frac{2(\theta - x_i)}{1 + (x_i - \theta)^2}.
$$

The equation $g(\theta) = 0$ has **no closed-form solution** and may have **several roots**:

```{r}
xc <- rcauchy(8, location = 3)
nllc <- function(th, dat) sapply(th, function(t) -sum(dcauchy(dat, t, log = TRUE)))
curve(nllc(x, xc), -10, 15, lwd = 2, xlab = expression(theta), ylab = expression(L(theta)), main = "Cauchy negative log-likelihood, n = 8")
```

This is why we need **numerical optimization**.

## Properties of the MLE

Under regularity conditions (satisfied by most models in practice), for iid data with true parameter $\theta_*$:

1. **Consistency**: $\hat\theta_{\mathrm{MLE}} \to \theta_*$ as $n \to \infty$.

2. **Asymptotic normality**:
$$
\hat\theta_{\mathrm{MLE}} \;\dot\sim\; N\!\big(\theta_*,\ [n I(\theta_*)]^{-1}\big), \qquad
I(\theta) = -E_X\!\left[\frac{\partial^2}{\partial\theta\,\partial\theta'}\log f(X \mid \theta)\right],
$$
the **expected Fisher information** per observation ($p \times p$ for a $p$-dimensional $\theta$).

3. **Asymptotic efficiency**: no consistent estimator has smaller asymptotic variance (Cramér–Rao bound).

Note $n I(\theta) = E_D\big[\mathcal{L}''(\theta; D)\big] = -E_D\big[\ell''(\theta; D)\big]$: the expected curvature of $\mathcal{L}$.

## Observed and Expected Fisher Information

$\theta_*$ is unknown, so the variance $[nI(\theta_*)]^{-1}$ is estimated by plugging in $\hat\theta$. Two versions:

- **Expected information** $n I(\hat\theta)$: requires computing $E_X[\cdot]$ analytically.

- **Observed information** $J(\hat\theta) = \mathcal{L}''(\hat\theta; D) = g'(\hat\theta) = -\sum_{i=1}^n \frac{\partial^2}{\partial\theta\partial\theta'}\log f(x_i \mid \hat\theta)$: the actual curvature at the MLE, no expectation needed.

By the LLN, $J(\hat\theta)/n \approx I(\hat\theta)$. Both give
$$
\widehat{\mathrm{Var}}(\hat\theta) \approx J(\hat\theta)^{-1} \quad\text{or}\quad [nI(\hat\theta)]^{-1}, \qquad \widehat{\mathrm{SE}}(\hat\theta_j) = \sqrt{\big[\widehat{\mathrm{Var}}(\hat\theta)\big]_{jj}} .
$$

**Computational consequence**: the Hessian of $\mathcal{L}$ at the optimum, which Newton-type optimizers compute anyway, gives the standard errors for free.

# Univariate Optimization

## The Problem

Minimize $\mathcal{L}(\theta) = -\ell(\theta)$, $\theta \in \mathbb{R}$; equivalently, find a root of
$g(\theta) = \mathcal{L}'(\theta)$, or rearrange that root equation into a fixed point $\theta = \varphi(\theta)$.

Methods differ in **what they need** and **how fast they converge**:

| method | needs | convergence |
|:--|:--|:--|
| fixed-point iteration | a rearrangement $\theta = \varphi(\theta)$ | linear, if at all |
| bisection | $g(\theta)$ and a sign-changing bracket | linear (halves bracket) |
| golden section | $\mathcal{L}(\theta)$ only, and a bracket | linear |
| **Newton–Raphson** | $g(\theta)$ and $g'(\theta) = \mathcal{L}''(\theta)$ | **quadratic** near root |
| Fisher scoring | $g(\theta)$ and $I(\theta)$ | linear–quadratic |

Every method below is illustrated on the same zero-truncated Poisson likelihood, so the pictures are directly comparable.

## Running Example: Zero-Truncated Poisson {.smaller}

We observe only the **nonzero** counts of a Poisson process ($X = Y \mid Y > 0$, $Y \sim \mathrm{Pois}(\lambda)$):
$$
P(x \mid \lambda) = \frac{e^{-\lambda}\lambda^x}{x!}\cdot\frac{1}{1 - e^{-\lambda}}, \qquad x = 1, 2, \ldots
$$
$$
\mathcal{L}(\lambda) = n\lambda - \sum x_i \log\lambda + n\log(1 - e^{-\lambda}) + C
$$
$$
g(\lambda) = \frac{n}{1 - e^{-\lambda}} - \frac{\sum x_i}{\lambda}
$$
$$
g'(\lambda) = \frac{\sum x_i}{\lambda^2} - \frac{n e^{-\lambda}}{(1 - e^{-\lambda})^2}.
$$

Using $E(X) = \dfrac{\lambda}{1 - e^{-\lambda}}$,
$$
nI(\lambda) = E[g'(\lambda)] = \frac{n}{\lambda(1 - e^{-\lambda})} - \frac{n e^{-\lambda}}{(1 - e^{-\lambda})^2}.
$$

## Zero-Truncated Poisson in R

```{r}
#| label: ztp-setup
library(latex2exp)
set.seed(812)
y <- rpois(200, 1.5); x <- y[y > 0]; n <- length(x); sx <- sum(x); xbar <- mean(x)
nll <- function(l) n * l - sx * log(l) + n * log(1 - exp(-l))   # the objective
g   <- function(l) n / (1 - exp(-l)) - sx / l                   # gradient  L'(lambda)
dg  <- function(l) sx / l^2 - n * exp(-l) / (1 - exp(-l))^2     # curvature L''(lambda)
nI  <- function(l) n / (l * (1 - exp(-l))) - n * exp(-l) / (1 - exp(-l))^2
lhat <- uniroot(g, c(1e-6, 20), tol = 1e-12)$root   # reference value for the pictures
lam  <- seq(0.4, 4, length.out = 400)               # plotting grid used throughout
c(naive_mean = xbar, mle = lhat, true_lambda = 1.5)
```

```{r}
#| label: ztp-curves
#| fig-width: 9
#| fig-height: 3.6
par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.2, 1))
plot(lam, nll(lam) - nll(lhat), type = "l", lwd = 2, main = "negative log-likelihood",
     xlab = TeX("$\\lambda$"), ylab = TeX("$L(\\lambda) - L(\\hat{\\lambda})$"))
abline(v = lhat, lty = 3); abline(v = xbar, col = 4, lty = 5)
legend("bottomright", bty = "n", lty = c(3, 5), col = c(1, 4), legend = c("MLE", "sample mean"))
plot(lam, g(lam), type = "l", lwd = 2, main = "gradient",
     xlab = TeX("$\\lambda$"), ylab = TeX("$g(\\lambda)$"))
abline(h = 0, lty = 3); points(lhat, 0, pch = 19, cex = 1.2)
```

The plain sample mean **overestimates** $\lambda$ because the zeros are missing; the MLE sits where the gradient crosses zero.

<!-- Replaces "## Fixed-Point (Simple) Iteration" and "## Fixed-Point: Two Rearrangements".
     Depends on the ztp-setup chunk: nll, g, lam, lhat, xbar. -->

## Fixed-Point Iteration: Two Rearrangements {.smaller}

The stationarity equation $g(\lambda) = 0$ can be written as
$$
\frac{\overline x}{\lambda} = \frac{1}{1 - e^{-\lambda}} .
$$

Solving for the $\lambda$ on the left gives one rearrangement; solving for the exponential gives another:
$$
\lambda = \overline x\,(1 - e^{-\lambda}) =: \varphi_1(\lambda),
\qquad
\lambda = -\log\!\Big(1 - \frac{\lambda}{\overline x}\Big) =: \varphi_2(\lambda) .
$$

Both have exactly the same fixed points, $\hat\lambda$ and the spurious root $\lambda = 0$, and indeed $\varphi_2 = \varphi_1^{-1}$.
(The map is written $\varphi$, not $g$, because $g$ is reserved for the gradient.)

## Fixed-Point Iteration: The Method {.smaller}

Pick a rearrangement $\lambda = \varphi(\lambda)$ and iterate
$$
\lambda_{k+1} = \varphi(\lambda_k), \qquad k = 0, 1, 2, \ldots
$$

Writing $e_k = \lambda_k - \lambda^*$ and expanding $\varphi$ about the fixed point, $e_{k+1} \approx \varphi'(\lambda^*) e_k$:

- the iteration converges to $\lambda^*$ when $|\varphi'(\lambda^*)| < 1$, and the error then shrinks by that **constant factor** each step — **linear** convergence;
- it is repelled from $\lambda^*$ when $|\varphi'(\lambda^*)| > 1$.

Here $\varphi_1'(\lambda) = \overline x\, e^{-\lambda}$ and $\varphi_2'(\lambda) = (\overline x - \lambda)^{-1}$. Since $\varphi_2 = \varphi_1^{-1}$, the slopes are reciprocal at every fixed point, so **at most one** of the two rearrangements can converge to $\hat\lambda$.

```{r}
#| label: fixed-point
fixed_point <- function(phi, x0, tol = 1e-8, maxit = 500) {
  for (k in 1:maxit) { x1 <- phi(x0); if (abs(x1 - x0) < tol) return(c(root = x1, iter = k)); x0 <- x1 }
  stop("not converged")
}
phi1 <- function(l) xbar * (1 - exp(-l))
phi2 <- function(l) -log(1 - l / xbar)
fixed_point(phi1, x0 = xbar)
```

Advantage: no derivatives. Disadvantage: a suitable $\varphi$ must be found, and the rearrangement alone decides the outcome.

## Fixed-Point: The Cobweb {.smaller}

```{r}
#| label: fixed-point-cobweb
#| fig-width: 9
#| fig-height: 4
cobweb <- function(phi, x0, nstep, main) {
  gg <- seq(0.05, 1.9, length.out = 300)
  plot(gg, phi(gg), type = "l", lwd = 2, xlim = c(0, 1.9), ylim = c(0, 1.9),
       xlab = TeX("$\\lambda_k$"), ylab = TeX("$\\varphi(\\lambda_k)$"), main = main)
  abline(0, 1, lty = 2); points(lhat, lhat, pch = 19, cex = 1.3)
  for (k in 1:nstep) {
    x1 <- phi(x0); if (!is.finite(x1) || x1 < 0) break
    segments(x0, x0, x0, x1, col = 2, lwd = 2)     # up (or down) to the curve
    segments(x0, x1, x1, x1, col = 2, lwd = 2)     # across to the diagonal
    points(x0, x1, pch = 19, col = 2, cex = 0.8)
    x0 <- x1
  }
}
par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.4, 1))
cobweb(phi1, x0 = 1.85, nstep = 6, main = TeX("$\\varphi_1$: converges, $|\\varphi_1'| < 1$"))
cobweb(phi2, x0 = 1.40, nstep = 6, main = TeX("$\\varphi_2$: repelled, $|\\varphi_2'| > 1$"))
```

The staircase walks to the intersection of $\varphi$ with the $45^\circ$ line. For $\varphi_1$ the crossing is shallow and each step strides toward it; $\varphi_2$ is the inverse map, its slope there is the reciprocal, and the iterates are pushed away from $\hat\lambda$ toward the fixed point at $\lambda = 0$.

## Fixed-Point: Traces of the Iterates {.smaller}

```{r}
#| label: fixed-point-trace
#| fig-width: 9.5
#| fig-height: 4
trace_fp <- function(phi, x0, nstep) {
  p <- x0
  for (k in 1:nstep) {
    xn <- phi(p[k])
    if (!is.finite(xn) || xn <= 0) break
    p <- c(p, xn)
  }
  p
}
t1 <- trace_fp(phi1, 1.85, 14); t2 <- trace_fp(phi2, 1.40, 14)
par(mfrow = c(1, 2), mar = c(4.2, 4.4, 2.4, 1))

## (a) the iterates themselves
plot(seq_along(t1) - 1, t1, type = "b", pch = 19, lwd = 2, ylim = c(0, 2),
     xlim = c(0, 14), xlab = "iteration $k$", ylab = TeX("$\\lambda_k$"), main = "traces")
lines(seq_along(t2) - 1, t2, type = "b", pch = 19, lwd = 2, col = 2)
abline(h = lhat, lty = 3); abline(h = 0, lty = 3)
mtext(TeX("$\\hat{\\lambda}$"), side = 4, at = lhat, las = 1, line = 0.3, cex = 0.9)
legend("topright", bty = "n", lwd = 2, pch = 19, col = c(1, 2),
       legend = c(TeX("$\\varphi_1$"), TeX("$\\varphi_2$")))

## (b) the errors, on a log scale
e1 <- pmax(abs(t1 - lhat), 1e-16); e2 <- pmax(abs(t2 - lhat), 1e-16)
plot(seq_along(e1) - 1, e1, type = "b", pch = 19, lwd = 2, log = "y",
     ylim = c(1e-10, 5), xlim = c(0, 14),
     xlab = "iteration $k$", ylab = TeX("$|\\lambda_k - \\hat{\\lambda}|$"), main = "errors")
lines(seq_along(e2) - 1, e2, type = "b", pch = 19, lwd = 2, col = 2)
lines(0:14, e1[1] * (xbar * exp(-lhat))^(0:14), lty = 5, col = 4, lwd = 2)
legend("bottomleft", bty = "n", lwd = 2, lty = c(1, 1, 5), pch = c(19, 19, NA),
       col = c(1, 2, 4), legend = c(TeX("$\\varphi_1$"), TeX("$\\varphi_2$"), TeX("rate $|\\varphi_1'(\\hat{\\lambda})|^k$")))
```

Left: $\varphi_1$ settles onto $\hat\lambda$ within a few steps, while $\varphi_2$ slides past it and decays toward the spurious fixed point at zero. Right: the $\varphi_1$ error falls along a straight line on the log scale, with the slope predicted by $|\varphi_1'(\hat\lambda)|$ — one fixed number of digits gained per iteration, against Newton's doubling. The $\varphi_2$ error rises and then flattens at $\hat\lambda$, the distance from the limit $0$ to the root it failed to find.


## Bisection

If $g$ is continuous and $g(a)$, $g(b)$ have opposite signs, a root lies in $[a, b]$. Evaluate $g$ at the midpoint and keep the half that still brackets the root.

```{r}
#| label: bisection
bisection <- function(g, a, b, tol = 1e-8) {
  k <- 0
  while (b - a > tol) { m <- (a + b) / 2; if (g(a) * g(m) <= 0) b <- m else a <- m; k <- k + 1 }
  c(root = (a + b) / 2, iter = k)
}
bisection(g, 0.1, 10)
```

- Guaranteed to converge; the bracket halves each step, so $\log_2\big((b-a)/\text{tol}\big) \approx 30$ iterations here.

- Needs only $g$, not $g'$. R's `uniroot()` uses a refined version (Brent's method).

## Bisection, Visualized

```{r}
#| label: bisection-plot
#| fig-width: 8
#| fig-height: 4
gv <- g(lam); yr <- range(gv); pad <- 0.35 * diff(yr)
plot(lam, gv, type = "l", lwd = 2, ylim = c(yr[1], yr[2] + pad),
     xlab = TeX("$\\lambda$"), ylab = TeX("$g(\\lambda)$"))
abline(h = 0, lty = 3); points(lhat, 0, pch = 19, cex = 1.3)

a <- 0.5; b <- 3.6
ylev <- seq(yr[2] + 0.08 * pad, yr[2] + 0.95 * pad, length.out = 4)
for (k in 1:4) {
  m <- (a + b) / 2
  segments(a, ylev[k], b, ylev[k], col = 4, lwd = 3)      # the current bracket
  points(c(a, b), rep(ylev[k], 2), pch = "|", col = 4)
  segments(m, ylev[k], m, g(m), lty = 3, col = 2)         # midpoint carried down to g
  points(m, ylev[k], pch = 19, col = 2)
  if (g(a) * g(m) <= 0) b <- m else a <- m
}
```

Each bar is the bracket at one iteration; the dot is its midpoint, mapped onto $g$ to decide which half survives. The bracket halves every step regardless of the shape of $g$ — no faster near the root than far from it.

## Golden Section Search

Works on $\mathcal{L}(\theta) = -\ell(\theta)$ directly, **without any derivative**: keep a bracket $[a, b]$ containing the minimum of a unimodal $\mathcal{L}$ and two interior points $a' < b'$ placed at the golden ratio ($\phi \approx 0.618$).

- If $\mathcal{L}(a') < \mathcal{L}(b')$ the minimum is in $[a, b']$; otherwise in $[a', b]$.
- The golden placement lets the surviving interior point be **reused**, so each step costs one new evaluation and shrinks the bracket by $0.618$.

```{r}
#| label: golden
golden <- function(f, a, b, tol = 1e-8) {
  phi <- (sqrt(5) - 1) / 2; k <- 0
  x1 <- b - phi * (b - a); x2 <- a + phi * (b - a); f1 <- f(x1); f2 <- f(x2)
  while (b - a > tol) {
    if (f1 < f2) { b <- x2; x2 <- x1; f2 <- f1; x1 <- b - phi * (b - a); f1 <- f(x1) }
    else         { a <- x1; x1 <- x2; f1 <- f2; x2 <- a + phi * (b - a); f2 <- f(x2) }
    k <- k + 1 }
  c(root = (a + b) / 2, iter = k)
}
golden(nll, 0.1, 10)                     # R equivalent: optimize(nll, c(0.1, 10))
```

## Golden Section, Visualized

```{r}
#| label: golden-plot
#| fig-width: 8
#| fig-height: 4
fv <- nll(lam); yr <- range(fv); pad <- 0.3 * diff(yr)
plot(lam, fv, type = "l", lwd = 2, ylim = c(yr[1], yr[2] + pad),
     xlab = TeX("$\\lambda$"), ylab = TeX("$L(\\lambda)$"))
points(lhat, nll(lhat), pch = 19, cex = 1.3)

gr <- (3 - sqrt(5)) / 2; xl <- 0.5; xu <- 3.6
ylev <- seq(yr[2] + 0.08 * pad, yr[2] + 0.95 * pad, length.out = 3)
for (k in 1:3) {
  xml <- xl + gr * (xu - xl); xmu <- xu - gr * (xu - xl)
  segments(xl, ylev[k], xu, ylev[k], col = 4, lwd = 3)
  points(c(xl, xu), rep(ylev[k], 2), pch = "|", col = 4)
  keep_lower <- nll(xml) < nll(xmu)
  points(xml, ylev[k], pch = 19, cex = 0.9, col = if (keep_lower) 4 else 2)
  points(xmu, ylev[k], pch = 19, cex = 0.9, col = if (keep_lower) 2 else 4)
  if (keep_lower) xu <- xmu else xl <- xml
}
```

The point that survives (blue) is already correctly placed for the next iteration, so only the red point costs a fresh evaluation.


## Newton–Raphson: Derivation

Approximate the gradient by its tangent line at the current point $\theta_k$ and solve for where the tangent hits zero:
$$
g(\theta) \approx g(\theta_k) + g'(\theta_k)(\theta - \theta_k) = 0
\quad\Longrightarrow\quad
\theta_{k+1} = \theta_k - \frac{g(\theta_k)}{g'(\theta_k)} = \theta_k - \frac{\mathcal{L}'(\theta_k)}{\mathcal{L}''(\theta_k)} .
$$

```{r}
#| label: nr-one-step
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
par(mar = c(4.2, 4.2, 2.2, 1))
l0 <- 3.0; l1 <- l0 - g(l0) / dg(l0)
plot(lam, g(lam), type = "l", lwd = 2, main = "one Newton step",
     xlab = TeX("$\\lambda$"), ylab = TeX("$g(\\lambda)$"))
abline(h = 0, lty = 3); points(lhat, 0, pch = 19, cex = 1.2)
xt <- c(l1 - 0.3, l0 + 0.3)                      # draw the whole tangent, not just the step
lines(xt, g(l0) + dg(l0) * (xt - l0), col = 2, lwd = 2)
points(c(l0, l1), c(g(l0), 0), pch = 19, col = 2)
segments(l1, 0, l1, g(l1), lty = 3, col = "grey40")
text(l0, g(l0), TeX("$\\lambda_k$"), pos = 4)
text(l1, 0, TeX("$\\lambda_{k+1}$"), pos = 1)
```

## Newton–Raphson: The Tangent Steps {.smaller}

```{r}
#| label: nr-tangent-steps
#| fig-width: 10
#| fig-height: 4.2
it <- 0.5                                        # starting value
for (k in 1:3) it <- c(it, it[k] - g(it[k]) / dg(it[k]))
par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.4, 1))

## (a) tangents to the gradient
plot(lam, g(lam), type = "l", lwd = 2, main = "tangents to the gradient",
     xlab = TeX("$\\lambda$"), ylab = TeX("$g(\\lambda)$"))
abline(h = 0, lty = 3); points(lhat, 0, pch = 19, cex = 1.2)
for (k in 1:3) {
  xt <- range(c(it[k], it[k + 1])) + c(-0.15, 0.15)
  lines(xt, g(it[k]) + dg(it[k]) * (xt - it[k]), col = 2, lwd = 2)
  segments(it[k + 1], 0, it[k + 1], g(it[k + 1]), lty = 3, col = "grey40")
  points(it[k], g(it[k]), pch = 19, col = 2, cex = 0.9)
  mtext(TeX(sprintf("$\\lambda_%d$", k - 1)), side = 1, at = it[k], line = -1, col = 2, cex = 0.8)
}

## (b) the quadratic model that each step minimizes
nllc <- function(l) nll(l) - nll(lhat)
plot(lam, nllc(lam), type = "l", lwd = 2, main = "the quadratic model behind each step",
     xlab = TeX("$\\lambda$"), ylab = TeX("$L(\\lambda) - L(\\hat{\\lambda})$"))
for (k in 1:3) {
  w <- seq(it[k] - 1, it[k] + 1, length.out = 200)
  lines(w, nllc(it[k]) + g(it[k]) * (w - it[k]) + 0.5 * dg(it[k]) * (w - it[k])^2,
        col = 2, lwd = 1.5, lty = 5)
  segments(it[k + 1], par("usr")[3], it[k + 1], nllc(it[k + 1]), lty = 3, col = "grey40")
  points(it[k], nllc(it[k]), pch = 19, col = 2, cex = 0.9)
}
points(lhat, 0, pch = 19, cex = 1.2)
```

Left: each red line is the tangent to $g$ at the current iterate, followed down to the axis to give the next one. Right: the same steps seen on $\mathcal{L}$ — the dashed parabola is the second-order Taylor expansion at $\lambda_k$, and $\lambda_{k+1}$ is its vertex. As the iterates approach $\hat\lambda$ the parabola hugs $\mathcal{L}$ and the vertex lands essentially on the MLE.


## Newton–Raphson in R

```{r}
#| label: nr-code
newton <- function(g, dg, x0, tol = 1e-8, maxit = 100, trace = FALSE) {
  for (k in 1:maxit) {
    x1 <- x0 - g(x0) / dg(x0)
    if (trace) cat(sprintf("iter %2d: %.10f\n", k, x1))
    if (!is.finite(x1)) stop("diverged")
    if (abs(x1 - x0) < tol) return(c(root = x1, iter = k))
    x0 <- x1 }
  stop("not converged")
}
newton(g, dg, x0 = xbar, trace = TRUE)
```

Iteration counts on the same problem: fixed point `r fixed_point(phi1, xbar)["iter"]`, bisection 30, golden 40, Newton `r newton(g, dg, xbar)["iter"]`.

## Convergence Rate of Newton–Raphson

Let $e_k = \theta_k - \hat\theta$. Expanding $g(\hat\theta) = 0$ around $\theta_k$:
$$
0 = g(\theta_k) - g'(\theta_k)e_k + \tfrac12 g''(\xi)e_k^2
\quad\Longrightarrow\quad
e_{k+1} = e_k - \frac{g(\theta_k)}{g'(\theta_k)} = \frac{g''(\xi)}{2g'(\theta_k)}\,e_k^2 .
$$

**Quadratic convergence**: the error is squared each step, so the number of correct digits roughly **doubles**. Linear methods gain a fixed number of digits per step.

```{r}
#| label: nr-error
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
par(mar = c(4.2, 4.2, 1.5, 1))
err_newton <- {e <- c(); x0 <- 3; for (k in 1:6) { x0 <- x0 - g(x0)/dg(x0); e <- c(e, abs(x0 - lhat)) }; e}
err_bisec  <- {a <- 0.1; b <- 10; e <- c(); for (k in 1:6) { m <- (a+b)/2; if (g(a)*g(m) <= 0) b <- m else a <- m; e <- c(e, abs((a+b)/2 - lhat)) }; e}
plot(1:6, pmax(err_newton, 1e-16), log = "y", type = "b", pch = 19, ylim = c(1e-16, 10),
     xlab = "iteration", ylab = "|error| (log scale)")
lines(1:6, err_bisec, type = "b", col = 2, pch = 19)
legend("bottomleft", c("Newton", "bisection"), col = 1:2, pch = 19, bty = "n")
```

## When Newton–Raphson Fails {.smaller}

Newton uses only **local** information, so it can misbehave:

- **Wrong curvature**: if $g'(\theta_k) = \mathcal{L}''(\theta_k) < 0$, the step moves toward a *maximum* of $\mathcal{L}$ or diverges.
- **Overshooting**: a flat gradient ($g'(\theta_k) \approx 0$) gives a huge step.
- **Multiple roots**: it converges to whichever root is nearest the start, possibly a local maximum.

The truncated Poisson overshoots from a poor start: at $\lambda_0 = 3.5$ the gradient is already flat, so the tangent crosses zero outside the parameter space.

```{r}
#| label: nr-overshoot
#| fig-width: 8
#| fig-height: 3.8
l0 <- 3.5; l_nr <- l0 - g(l0) / dg(l0)
par(mar = c(4.2, 4.2, 1.5, 1))
plot(NA, xlim = c(-0.8, 4), ylim = range(g(lam)),
     xlab = TeX("$\\lambda$"), ylab = TeX("$g(\\lambda)$"))
rect(-0.8, par("usr")[3], 0, par("usr")[4], col = "grey90", border = NA)
text(-0.4, 0.55 * par("usr")[4], TeX("$\\lambda < 0$"), col = "grey30", cex = 0.9)
lines(lam, g(lam), lwd = 2); abline(h = 0, lty = 3); points(lhat, 0, pch = 19, cex = 1.2)
xt <- c(l_nr - 0.2, l0 + 0.3)
lines(xt, g(l0) + dg(l0) * (xt - l0), col = 2, lwd = 2)
points(c(l0, l_nr), c(g(l0), 0), pch = 19, col = 2)
mtext(TeX("$\\lambda_0$"), side = 1, at = l0, line = -1, col = 2, cex = 0.9)
mtext(TeX("$\\lambda_1$"), side = 1, at = l_nr, line = -1, col = 2, cex = 0.9)
```

```{r}
#| label: nr-overshoot-num
c(lambda_1 = l_nr)
tryCatch(newton(g, dg, l0), error = function(e) conditionMessage(e))
```

Safeguards: start from a robust estimate, **halve the step** until $\mathcal{L}$ decreases, or fall back to bisection/golden section, which only need a bracket.

## Flat Tails: The Cauchy Location Model {.smaller}

A single-observation Cauchy likelihood shows the same failure in pure form: $\mathcal{L}(\theta) = -\ell(\theta) = \log(1+\theta^2)$ has gradient $g(\theta) = 2\theta/(1+\theta^2)$, which rises, turns over at $\theta = 1$, then decays — the Newton step blows up wherever $g$ is flat.

```{r}
#| label: cauchy-flat
#| fig-width: 9
#| fig-height: 3.4
nll_c <- function(t) log(1 + t^2)
g_c   <- function(t) 2 * t / (1 + t^2)
par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.2, 1))
curve(nll_c, -5, 5, n = 400, lwd = 2, xlab = TeX("$\\theta$"), ylab = TeX("$L(\\theta)$"), main = "negative log-likelihood")
curve(g_c, -5, 5, n = 400, lwd = 2, xlab = TeX("$\\theta$"), ylab = TeX("$g(\\theta)$"), main = "gradient: flat tails")
abline(h = 0, lty = 3); abline(v = c(-1, 1), col = 2, lty = 2)
```

From $\theta_0 = 0.5$ Newton converges quadratically; from $\theta_0 = 1.2$, beyond the inflection point, it runs away — each step landing where the gradient is flatter still.

```{r}
#| label: cauchy-starts
set.seed(2024)
x_c <- rcauchy(8, 3)
g_cau  <- function(t) sum(2 * (t - x_c) / (1 + (x_c - t)^2))       # gradient of the Cauchy nll
dg_cau <- function(t) sum(2 * (1 - (x_c - t)^2) / (1 + (x_c - t)^2)^2)
starts <- c(-5, 0, median(x_c), 8)
sapply(starts, function(s) tryCatch(newton(g_cau, dg_cau, s)["root"], error = function(e) NA))
```

## Univariate Fisher Scoring

Replace the observed curvature $J(\theta_k) = g'(\theta_k)$ by its expectation $nI(\theta_k)$:
$$
\theta_{k+1} = \theta_k - \frac{g(\theta_k)}{n I(\theta_k)} .
$$

- $nI(\theta) > 0$ always, so the step is always in a descent direction for $\mathcal{L}$ — more robust than Newton far from the optimum.
- Near the MLE $nI(\hat\theta) \approx g'(\hat\theta)$, so the two methods behave alike.
- Requires the expectation in closed form.

```{r}
#| label: scoring-code
fisher_scoring <- function(g, nI, x0, tol = 1e-8, maxit = 100) {
  for (k in 1:maxit) { x1 <- x0 - g(x0) / nI(x0); if (abs(x1 - x0) < tol) return(c(root = x1, iter = k)); x0 <- x1 }
  stop("not converged")
}
rbind(newton = newton(g, dg, 3), scoring = fisher_scoring(g, nI, 3))
```

## Newton vs. Scoring from the Same Point {.smaller}

```{r}
#| label: scoring-vs-newton
#| fig-width: 10
#| fig-height: 4.2
par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.4, 1))

## (a) the two lines from the bad start
l_fs <- l0 - g(l0) / nI(l0)
plot(NA, xlim = c(-0.8, 4), ylim = range(g(lam)), main = "observed vs. expected curvature",
     xlab = TeX("$\\lambda$"), ylab = TeX("$g(\\lambda)$"))
rect(-0.8, par("usr")[3], 0, par("usr")[4], col = "grey90", border = NA)
lines(lam, g(lam), lwd = 2); abline(h = 0, lty = 3); points(lhat, 0, pch = 19, cex = 1.2)
xn <- c(l_nr - 0.2, l0 + 0.3); xf <- c(l_fs - 0.3, l0 + 0.3)
lines(xn, g(l0) + dg(l0) * (xn - l0), col = 2, lwd = 2)
lines(xf, g(l0) + nI(l0) * (xf - l0), col = 4, lwd = 2, lty = 5)
points(c(l0, l_nr, l_fs), c(g(l0), 0, 0), pch = 19, col = c(1, 2, 4))
legend("bottomright", bty = "n", lwd = 2, lty = c(1, 5), col = c(2, 4),
       legend = c(TeX("Newton: slope $g'$"), TeX("scoring: slope $nI$")))

## (b) the scoring path from the same start
plot(lam, g(lam), type = "l", lwd = 2, main = "scoring converges from the bad start",
     xlab = TeX("$\\lambda$"), ylab = TeX("$g(\\lambda)$"))
abline(h = 0, lty = 3); points(lhat, 0, pch = 19, cex = 1.2)
fs <- l0
for (k in 1:4) fs <- c(fs, fs[k] - g(fs[k]) / nI(fs[k]))
for (k in 1:4) {
  xt <- range(c(fs[k], fs[k + 1])) + c(-0.1, 0.1)
  lines(xt, g(fs[k]) + nI(fs[k]) * (xt - fs[k]), col = 4, lwd = 1.8, lty = 5)
  segments(fs[k + 1], 0, fs[k + 1], g(fs[k + 1]), lty = 3, col = "grey40")
  points(fs[k], g(fs[k]), pch = 19, col = 4, cex = 0.9)
}
```

```{r}
#| label: scoring-vs-newton-num
c(newton_step = l_nr, scoring_step = l_fs)
```

The two lines pass through the same point with different slopes. Newton uses the actual slope $g'(\lambda_0)$, which is nearly flat here and throws the iterate out of the parameter space; scoring uses $nI(\lambda_0)$, an average over hypothetical data, which stays steeply positive and keeps every iterate positive. The slopes agree as $\lambda_k \to \hat\lambda$, so the two methods finish at the same speed.

## Standard Error from the Curvature

At the MLE $\hat\lambda$, the observed information $J(\hat\lambda) = \mathcal{L}''(\hat\lambda) = g'(\hat\lambda)$ and the expected information $nI(\hat\lambda)$ both estimate the precision:

```{r}
#| label: standard-error
c(mle = lhat, se_observed = 1 / sqrt(dg(lhat)), se_expected = 1 / sqrt(nI(lhat)))
lhat + c(-1.96, 1.96) / sqrt(dg(lhat))    # 95% Wald interval
```

## Numerical Derivatives

When $g$ or $g'$ is tedious to derive, approximate by finite differences:
$$
g(\theta) = \mathcal{L}'(\theta) \approx \frac{\mathcal{L}(\theta + h) - \mathcal{L}(\theta - h)}{2h}, 
$$
$$
g'(\theta) = \mathcal{L}''(\theta) \approx \frac{\mathcal{L}(\theta + h) - 2\mathcal{L}(\theta) + \mathcal{L}(\theta - h)}{h^2}.
$$

- $h$ too large: truncation error; $h$ too small: roundoff error (large − large). Typical choice $h \approx \sqrt{\epsilon}\,(1 + \lvert\theta\rvert) \approx 10^{-8}$ for the first derivative, $\epsilon^{1/3}$ for the second.

- R's `nlm()` and `optim(method = "BFGS")` use numerical gradients unless you supply analytic ones; `numDeriv::grad()` and `hessian()` are more accurate.

```{r}
#| label: numerical-derivative
h <- 1e-5
c(analytic = dg(lhat), numeric = (nll(lhat + h) - 2 * nll(lhat) + nll(lhat - h)) / h^2)
```

## Does the Effort Pay Off? MLE vs. the Naive Estimator

Simulating many truncated samples shows the naive estimator's bias never vanishes — it is the wrong model, not too little data — while the MLE is centred on the truth.

```{r}
#| label: mle-vs-naive
#| fig-width: 9
#| fig-height: 3.6
mle_nzp <- function(n_obs)
  uniroot(function(l) length(n_obs) / (1 - exp(-l)) - sum(n_obs) / l,
          c(1e-6, 50), tol = 1e-10)$root

sim_nzp <- function(lambda, n_gen = 200, B = 500)
  vapply(1:B, function(b) {
    yb <- rpois(n_gen, lambda); nb <- yb[yb > 0]
    c(naive = mean(nb), mle = mle_nzp(nb))
  }, c(naive = 0, mle = 0))

est <- sim_nzp(1.5)
par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.2, 1))
for (m in c("naive", "mle"))
  { hist(est[m, ], breaks = 25, col = "grey90", xlim = range(est),
         xlab = TeX("$\\hat{\\lambda}$"),
         main = sprintf("%s, bias = %.3f", m, mean(est[m, ]) - 1.5))
    abline(v = 1.5, col = 2, lwd = 2, lty = 2) }
```

The naive estimator's bias grows as $\lambda$ shrinks, since more mass sits on the discarded zero; the MLE's spread matches the information-based standard error from two slides ago.

# Multivariate Newton–Raphson and Fisher Scoring

## Gradient and Hessian

For $\theta = (\theta_1, \ldots, \theta_p)'$, the **gradient** and **Hessian** of $\mathcal{L}$ are
$$
g(\theta) = \nabla\mathcal{L}(\theta) = \begin{pmatrix}\partial\mathcal{L}/\partial\theta_1 \\ \vdots \\ \partial\mathcal{L}/\partial\theta_p\end{pmatrix}, \qquad
H(\theta) = \nabla^2\mathcal{L}(\theta) = \left(\frac{\partial^2\mathcal{L}}{\partial\theta_i\,\partial\theta_j}\right)_{p\times p} .
$$


- $g$ points in the direction of **steepest ascent** of $\mathcal{L}$ and is perpendicular to the contour through $\theta$; every method below moves along $-g$.

- $H$ describes the local curvature. At a local minimum it is positive semidefinite; if it is positive definite, the minimum is strict and locally unique. Near the minimum, $\mathcal{L}$ is well approximated by a quadratic, so $H$ determines the shape of the contours. At the MLE, $H(\hat\theta) = J(\hat\theta)$, the observed information.


## Gradient and Hessian: Illustration

Quadratic: $\mathcal{L}(\theta) = \tfrac12(\theta - u)'A(\theta - u)$, $A$ positive definite, $g(\theta) = A(\theta - u)$, $H = A$ (constant). Contours are ellipses centred at $u$. Write $A = Q\Lambda Q'$:

:::: {.columns}

::: {.column width="55%"}

- axes point along the eigenvectors of $Q$;
- half-length along axis $j$ is $\propto 1/\sqrt{\lambda_j}$, so a large eigenvalue means high curvature and a short, well-determined direction;
- $\lambda_{\max}/\lambda_{\min}$ measures elongation; tilted ellipses indicate correlated parameters.
- Gradient arrows (green) are perpendicular to contours and point uphill, away from $u$; a descent method steps along $-g$, which need not point at $u$ unless the ellipse is a circle.
- Near its minimum any smooth $\mathcal{L}$ looks like this, with $A = \nabla^2\mathcal{L}(\hat\theta) = J(\hat\theta)$.
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 7
A <- matrix(c(2, 1.2, 1.2, 1), 2)
s <- seq(-2, 2, length.out = 80); gridpts <- expand.grid(x = s, y = s)
z <- matrix(apply(gridpts, 1, function(v) 0.5 * drop(t(v) %*% A %*% v)), 80)
par(pty = "s", mar = c(4, 4, 1, 1))
contour(s, s, z, nlevels = 12, xlab = expression(theta[1]), ylab = expression(theta[2]))
# gradient arrows
pts <- rbind(c(1, 1), c(-1.5, 0.5), c(0.5, -1.5))
for (i in 1:3) { gr <- A %*% pts[i, ]
  arrows(pts[i,1], pts[i,2], pts[i,1] + 0.3*gr[1], pts[i,2] + 0.3*gr[2],
         col = 3, lwd = 2, length = 0.1) }
# principal axes with eigenvalue labels: length 1/sqrt(lambda) on each side
e <- eigen(A); k <- 1.6
for (j in 1:2) { v <- e$vectors[, j] * k / sqrt(e$values[j])
  segments(-v[1], -v[2], v[1], v[2], col = "firebrick", lwd = 2, lty = 2)
  text(v[1], v[2], bquote(lambda[.(j)] == .(round(e$values[j], 2))),
       col = "firebrick", pos = if (j == 1) 4 else 2, cex = 0.9) }
points(0, 0, pch = 19)
```
:::

::::

## Multivariate Newton–Raphson

Second-order Taylor expansion at $\theta_k$, minimized exactly:
$$
\mathcal{L}(\theta) \approx \mathcal{L}(\theta_k) + g(\theta_k)'(\theta - \theta_k) + \tfrac12(\theta - \theta_k)'H(\theta_k)(\theta - \theta_k)
$$
$$
\Longrightarrow\quad
\underbrace{\theta_{k+1}}_{p\times1} = \theta_k - \underbrace{\big[H(\theta_k)\big]^{-1}}_{p\times p}\underbrace{g(\theta_k)}_{p\times1}.
$$

Compare with the univariate $\theta_{k+1} = \theta_k - g(\theta_k)/g'(\theta_k)$: division becomes multiplication by the inverse Hessian.

**Fisher scoring** replaces $H(\theta_k) = \nabla^2\mathcal{L}(\theta_k)$ by the expected information $nI(\theta_k)$:
$$
\theta_{k+1} = \theta_k - \big[nI(\theta_k)\big]^{-1}g(\theta_k).
$$

Each iteration costs one $p \times p$ linear solve — $O(p^3)$ — fine for $p$ in the hundreds, expensive for $p$ in the millions.

## Example: Logistic Regression {.smaller}

$y_i \mid x_i \sim \mathrm{Bern}(\pi_i)$ with $\pi_i = \dfrac{e^{\eta_i}}{1 + e^{\eta_i}} = \dfrac{1}{1 + e^{-\eta_i}}$, $\eta_i = x_i'\beta$, $x_i = (1, x_{i1}, \ldots, x_{ip})'$.
$$
L(\beta) = \prod_{i=1}^n \pi_i^{y_i}(1-\pi_i)^{1-y_i}
= \prod_{i=1}^n \frac{e^{y_i\eta_i}}{1 + e^{\eta_i}}
$$
$$
\mathcal{L}(\beta) = \sum_{i=1}^n \big[\log(1 + e^{x_i'\beta}) - y_i\,x_i'\beta\big].
$$

Using $\dfrac{\partial}{\partial\beta}\log(1 + e^{x_i'\beta}) = \pi_i x_i$ and $\dfrac{\partial\pi_i}{\partial\beta} = \pi_i(1-\pi_i)x_i$:
$$
g(\beta) = \sum_{i=1}^n (\pi_i - y_i)\,x_i = X'(\pi - y), \qquad
$$
$$
H(\beta) = \sum_{i=1}^n \pi_i(1-\pi_i)\,x_i x_i' = X'WX, \text{with}~W = \mathrm{diag}\big(\pi_i(1-\pi_i)\big)
$$
.

## Logistic Regression: Newton = Fisher Scoring = IRLS

- $H = X'WX$ is positive semi-definite, so $\mathcal{L}$ is **convex**: any stationary point is the global minimum.

- The Hessian does not involve $y$, so its expectation equals itself: **Newton–Raphson and Fisher scoring coincide**.
- The update
$$
\beta_{k+1} = \beta_k - (X'W_kX)^{-1}X'(\pi_k - y) = (X'W_kX)^{-1}X'W_k z_k,
$$ 
$$
z_k = X\beta_k + W_k^{-1}(y - \pi_k),
$$
is a weighted least squares fit of the *working response* $z_k$: **iteratively reweighted least squares (IRLS)**, which is how `glm()` fits all generalized linear models.

- Caveat: if the classes are perfectly separable, $\mathcal{L}$ decreases forever and the MLE does not exist ($\lvert\hat\beta\rvert \to \infty$).

## Logistic Regression in R

```{r}
n <- 500; X <- cbind(1, rnorm(n), rnorm(n)); beta_true <- c(-0.5, 1, -2)
y <- rbinom(n, 1, plogis(X %*% beta_true))
nll_lr <- function(b) sum(log1p(exp(X %*% b)) - y * (X %*% b))
g_lr   <- function(b) crossprod(X, plogis(X %*% b) - y)
H_lr   <- function(b) { p <- plogis(X %*% b); crossprod(X, X * drop(p * (1 - p))) }

newton_mv <- function(g, H, b0, tol = 1e-8, maxit = 50) {
  path <- drop(b0)
  for (k in 1:maxit) {
    b1 <- b0 - solve(H(b0), g(b0)); path <- rbind(path, drop(b1))
    if (max(abs(b1 - b0)) < tol)
      return(list(est = drop(b1), iter = k, se = sqrt(diag(solve(H(b1)))), path = path))
    b0 <- b1 }
}
fit <- newton_mv(g_lr, H_lr, rep(0, 3)); fit$iter
rbind(newton = fit$est, glm = coef(glm(y ~ X[, -1], family = binomial)))
rbind(newton = fit$se, glm = sqrt(diag(vcov(glm(y ~ X[, -1], family = binomial)))))
```

## Variance of the MLE $\hat\beta$
$$
\widehat{\mathrm{Var}}(\hat\beta) = \big[\nabla^2\mathcal{L}(\hat\beta)\big]^{-1} = J(\hat\beta)^{-1} = (X'\hat WX)^{-1}
= \begin{pmatrix}\hat\sigma_{00} & \cdots & \hat\sigma_{0p} \\ \vdots & \ddots & \vdots \\ \hat\sigma_{p0} & \cdots & \hat\sigma_{pp}\end{pmatrix},
$$
$$
\widehat{\mathrm{SE}}(\hat\beta_j) = \sqrt{\hat\sigma_{jj}} .
$$

Wald tests and confidence intervals follow directly:
$$
z_j = \frac{\hat\beta_j}{\widehat{\mathrm{SE}}(\hat\beta_j)} \;\dot\sim\; N(0,1) \text{ under } H_0: \beta_j = 0, \qquad \hat\beta_j \pm 1.96\,\widehat{\mathrm{SE}}(\hat\beta_j).
$$

This is the `summary(glm(...))` table. The variance comes from the same Hessian that drove the Newton iterations.

## The Need to Robustify Newton–Raphson

The Newton step $d_k = -H_k^{-1} g_k$, with $g_k = \nabla\mathcal{L}(\theta_k)$ and
$H_k = \nabla^2\mathcal{L}(\theta_k)$, solves the local quadratic model exactly. Two things
can go wrong away from the solution:

- **The direction may not descend.** $g_k^\top d_k = -g_k^\top H_k^{-1} g_k < 0$
  requires $H_k \succ 0$. Where $\mathcal{L}$ is not convex, $H_k$ may be indefinite or
  singular and $d_k$ can point uphill.
- **The unit step may overshoot.** Even with $H_k \succ 0$, the quadratic model
  is trustworthy only near $\theta_k$; $\alpha_k = 1$ can increase $\mathcal{L}$.

The two failures are independent and need separate remedies: one fixes the
**direction**, the other the **length**.

## Robustification of the Hessian

Replace $H_k$ by a positive definite $\tilde H_k$ and solve
$\tilde H_k d_k = -g_k$.

**Shift by a multiple of the identity.** With $\lambda_{\min}$ the smallest
eigenvalue of $H_k$,
$$
\tilde H_k = H_k + \tau I, \qquad
\tau = \max\left(0,\; \delta - \lambda_{\min}\right), \qquad
\delta = 10^{-3}\lambda_{\max} .
$$
Eigenvectors are unchanged and every eigenvalue is shifted by $\tau$. As
$\tau \to \infty$ the direction rotates toward $-g_k$, so the method
interpolates between Newton and steepest descent.

**Spectral modification.** With $H_k = V \Lambda V^\top$, replace
$\lambda_i$ by $\max(|\lambda_i|, \delta)$. This reverses negative-curvature
directions rather than damping them, so the step moves *away* from a saddle.

Where $H_k \succ 0$ already, $\tau = 0$ and $\tilde H_k = H_k$: the refinement
is inactive near a well-behaved minimum.

## Step length: backtracking and line search

Given a descent direction, choose $\alpha_k$ by an inexact search along
$\phi(\alpha) = \mathcal{L}(\theta_k + \alpha d_k)$.

**Armijo (sufficient decrease).**
$$
\mathcal{L}(\theta_k + \alpha d_k) \le \mathcal{L}(\theta_k) + c_1 \alpha\, g_k^\top d_k,
\qquad c_1 = 10^{-4}.
$$

**Backtracking.** Start at $\alpha = 1$ and set $\alpha \leftarrow \rho\alpha$
with $\rho = \tfrac12$ until Armijo holds. Simple, and it accepts $\alpha_k = 1$
whenever the full Newton step is acceptable.

**Curvature (Wolfe) condition.**
$$
g(\theta_k + \alpha d_k)^\top d_k \ \ge\ c_2\, g_k^\top d_k,
\qquad c_2 = 0.9,
$$
which rules out steps that are too short. Backtracking alone cannot enforce it,
since it never lengthens the trial step. The condition matters for conjugate
gradient and is what keeps $s_k^\top y_k > 0$, hence the BFGS update well
defined.

## The safeguarded algorithm

::: {.callout-note appearance="minimal"}
At $\theta_k$: compute $g_k$, $H_k$; form $\tilde H_k \succ 0$; solve
$\tilde H_k d_k = -g_k$; find $\alpha_k$ from $\alpha_0 = 1$ by line search;
set $\theta_{k+1} = \theta_k + \alpha_k d_k$.
:::

Both safeguards are needed, and both switch themselves off:

| | near a minimum | far away / indefinite |
|---|---|---|
| $\tau$ | $0$ | $> 0$, direction rotates toward $-g_k$ |
| $\alpha_k$ | $1$ | $< 1$ (or $> 1$ under Wolfe) |

Quadratic convergence is therefore retained locally, while sufficient decrease
at every iteration gives convergence from remote starting points. This is
**safeguarded Newton**, not "Newton–Raphson" in the textbook sense — and not
what `nlm()` runs, which globalizes with a trust region.

# Gradient-based Methods

## What If the Hessian Is Unavailable or Too Big?

| method | uses | per-iteration cost | typical use |
|:--|:--|:--|:--|
| Nelder–Mead | $\mathcal{L}$ only | $O(p)$ evaluations | small $p$, non-smooth or noisy $\mathcal{L}$ |
| gradient descent + line search | $g$ | $O(p)$ + line search | large $p$ |
| conjugate gradient | $g$ | $O(p)$ + line search | large $p$, quadratic-like $\mathcal{L}$ |
| quasi-Newton (BFGS, L-BFGS) | $g$, builds $\approx$ Hessian | $O(p^2)$ / $O(mp)$ | default general-purpose |
| Newton / Fisher scoring | $g$, $H$ | $O(p^3)$ | moderate $p$, gives SEs |

All are available through `optim(par, fn, gr, method = ...)` in R. Convention below: **minimize** $\mathcal{L}(\theta) = -\ell(\theta)$, with $g = \nabla\mathcal{L}$.


## Gradient Descent with Line Search

Move in the direction of steepest descent $d_k = -g(\theta_k)$, choosing the step length $\alpha_k$ by a **line search**:
$$
\theta_{k+1} = \theta_k + \alpha_k d_k, \qquad \alpha_k = \arg\min_{\alpha > 0} \mathcal{L}(\theta_k + \alpha d_k) .
$$

- Exact line search: a univariate minimization (golden section / `optimize`) along the ray.

- Inexact (**backtracking, Armijo**): start with $\alpha = 1$ and halve until $\mathcal{L}(\theta_k + \alpha d_k) \le \mathcal{L}(\theta_k) + c\,\alpha\,g(\theta_k)'d_k$ (a "sufficient decrease").

```{r}
grad_descent <- function(nll, g, x0, maxit = 1000, tol = 1e-8) {
  path <- x0
  for (k in 1:maxit) {
    d <- -g(x0); a <- optimize(function(a) nll(x0 + a * d), c(0, 10))$minimum   # exact line search
    x1 <- x0 + a * d; path <- rbind(path, x1)
    if (sqrt(sum((x1 - x0)^2)) < tol) break
    x0 <- x1 }
  list(par = x1, iter = k, path = path)
}
```

## Why Gradient Descent Zig-Zags

With exact line search, consecutive directions are **orthogonal** ($g(\theta_{k+1})'d_k = 0$), so on elongated contours the path zig-zags across the valley:

```{r}
#| echo: false
nllq <- function(v) 0.5 * drop(t(v) %*% A %*% v); gq <- function(v) drop(A %*% v)
gd <- grad_descent(nllq, gq, c(2, -1.8), maxit = 30)
par(pty = "s", mar = c(4, 4, 2, 1)); contour(seq(-2, 2, l = 80), seq(-2, 2, l = 80), z, nlevels = 15, main = sprintf("gradient descent, %d iterations", gd$iter))
lines(gd$path, type = "b", col = 2, pch = 19, cex = 0.6)
```

The number of iterations grows with the **condition number** of the Hessian (ratio of largest to smallest eigenvalue).

## Conjugate Gradient

Fix the zig-zag by making each new direction **conjugate** to the previous ones with respect to the Hessian $Q$: $d_i'Qd_j = 0$ for $i \ne j$. For a quadratic $\mathcal{L}(\theta) = \tfrac12\theta'Q\theta - b'\theta$, minimizing along $p$ such directions reaches the exact minimum in **at most $p$ steps**.

The nonlinear version builds directions from gradients only:
$$
d_0 = -g_0, \qquad d_{k+1} = -g_{k+1} + \beta_k d_k, \qquad
$$
$$
\beta_k^{FR} = \frac{g_{k+1}'g_{k+1}}{g_k'g_k}\ \text{(Fletcher–Reeves)}, \quad
\beta_k^{PR} = \frac{g_{k+1}'(g_{k+1} - g_k)}{g_k'g_k}\ \text{(Polak–Ribière)},
$$
with $\theta_{k+1} = \theta_k + \alpha_k d_k$ from a line search and $g_k = g(\theta_k) = \nabla\mathcal{L}(\theta_k)$.

- Memory: only two vectors. No matrix ever formed. Ideal for very large $p$.

- Restart with $d = -g_k$ every $p$ iterations, since conjugacy degrades on non-quadratic $\mathcal{L}$.

## Conjugate Gradient in R

```{r}
conj_grad <- function(nll, g, x0, maxit = 1000, tol = 1e-8, restart = length(x0)) {
  g0 <- g(x0); d <- -g0; path <- x0
  for (k in 1:maxit) {
    a <- optimize(function(a) nll(x0 + a * d), c(0, 10))$minimum
    x1 <- x0 + a * d; g1 <- g(x1); path <- rbind(path, x1)
    if (sqrt(sum(g1^2)) < tol) break
    beta <- if (k %% restart == 0) 0 else max(0, sum(g1 * (g1 - g0)) / sum(g0^2))   # Polak-Ribiere+
    d <- -g1 + beta * d; x0 <- x1; g0 <- g1 }
  list(par = x1, iter = k, path = path)
}
cg <- conj_grad(nllq, gq, c(2, -1.8))
par(pty = "s", mar = c(4, 4, 2, 1)); contour(seq(-2, 2, l = 80), seq(-2, 2, l = 80), z, nlevels = 15, main = sprintf("conjugate gradient, %d iterations", cg$iter))
lines(gd$path, type = "b", col = 2, pch = 19, cex = 0.6); lines(cg$path, type = "b", col = 4, pch = 19, lwd = 2)
legend("topleft", c("gradient descent", "conjugate gradient"), col = c(2, 4), lwd = 2, bty = "n")
```

## Quasi-Newton: BFGS

Between gradient methods and Newton: build an approximation $B_k \approx \nabla^2\mathcal{L}$ from successive gradient differences $s_k = \theta_{k+1} - \theta_k$, $y_k = g_{k+1} - g_k$, via a rank-two update that keeps $B_k$ positive definite. Steps are $d_k = -B_k^{-1}g_k$ with a line search.

```{r}
g_lr_vec <- function(b) drop(g_lr(b))
fits <- list(NM = optim(rep(0, 3), nll_lr, method = "Nelder-Mead"),
             CG = optim(rep(0, 3), nll_lr, g_lr_vec, method = "CG"),
             BFGS = optim(rep(0, 3), nll_lr, g_lr_vec, method = "BFGS", hessian = TRUE))
t(sapply(fits, function(f) c(f$par, f_evals = f$counts[1], g_evals = f$counts[2])))
sqrt(diag(solve(fits$BFGS$hessian)))     # SEs from the numerically computed Hessian at the optimum
```

`optim(..., hessian = TRUE)` returns a finite-difference Hessian at the solution, so standard errors are available even for gradient-only methods.

# Stochastic Gradient Descent

## Motivation: Large $n$

For iid data the negative log-likelihood and its gradient are **sums over observations**:
$$
\mathcal{L}(\theta) = \sum_{i=1}^n \mathcal{L}_i(\theta), \qquad g(\theta) = \sum_{i=1}^n g_i(\theta), \qquad g_i = \nabla\mathcal{L}_i .
$$

Every method so far needs the **full** gradient, an $O(np)$ pass through all data, per iteration (and Newton needs $O(np^2)$ for the Hessian). With $n$ in the millions and $p$ in the millions (neural networks), even one full gradient is expensive.

**Idea.** $g_i(\theta)$ for a randomly chosen $i$ is an **unbiased estimate** of the average gradient:
$$
E_i\big[g_i(\theta)\big] = \frac{1}{n}\,g(\theta).
$$

A noisy but cheap gradient estimate is enough to make progress — a Monte Carlo idea applied to optimization.


## The SGD Algorithm {.smaller}

For $t = 1, 2, \ldots$:

1. Sample an index $i_t$ uniformly from $\{1, \ldots, n\}$ (or shuffle the data and cycle through it; one pass is an **epoch**).

2. Update in the descent direction of that single observation's negative log-likelihood:
$$
\theta_{t+1} = \theta_t - \gamma_t\,g_{i_t}(\theta_t).
$$

**Mini-batch SGD** averages the gradient over a batch $B_t$ of $m$ observations ($m = 32$–$512$ typical):
$$
\theta_{t+1} = \theta_t - \gamma_t\,\frac{1}{m}\sum_{i\in B_t}g_i(\theta_t),
$$
reducing the variance by $1/m$ at $m$ times the cost, and using vectorized/GPU arithmetic.

The **learning rate** $\gamma_t$ replaces the line search (a line search would need full-data evaluations).

## Learning Rate and Convergence

Because the gradient is noisy, a constant $\gamma$ makes $\theta_t$ **bounce around** the optimum with variance $\propto \gamma$. Robbins–Monro (1951): convergence to the optimum is guaranteed if
$$
\sum_t \gamma_t = \infty \quad\text{and}\quad \sum_t \gamma_t^2 < \infty, \qquad\text{e.g. } \gamma_t = \frac{\gamma_0}{1 + t/t_0} .
$$

- $\gamma$ too large: divergence or persistent noise. $\gamma$ too small: painfully slow.

- Convergence is only **sublinear**, far slower than Newton *per iteration* — but each iteration costs $O(p)$ instead of $O(np)$, and early progress is fast.
- Common refinements: **momentum** (average past gradients), **Adam** (per-coordinate adaptive rates), **averaging** the iterates $\overline\theta_T = \frac1T\sum_t\theta_t$ (Polyak–Ruppert), which restores the $\sqrt n$-efficiency of the MLE.


## A Smaller Logistic Regression, for SGD {.smaller}

Let's apply SGD to logistic regrression.

```{r}
log1pexp <- function(u) pmax(u, 0) + log1p(exp(-abs(u)))
nll_g_info <- function(b, r, z) {
  u <- b[1] + b[2] * z; p <- plogis(u); w <- p * (1 - p)
  list(nll  = -sum(u * r - log1pexp(u)),                   # L(beta)
       g    = c(sum(p - r), sum(z * (p - r))),             # g(beta) = grad L
       info = matrix(c(sum(w), sum(z * w), sum(z * w), sum(z * z * w)), 2, 2))
}
nll_logistic <- function(b, z, r) { u <- b[1] + b[2] * z; -sum(u * r - log1pexp(u)) }
mle_logistic_nr <- function(b0, no_iter, r, z) {
  out <- matrix(NA_real_, no_iter + 1, 3, dimnames = list(NULL, c("beta0", "beta1", "nll")))
  b <- b0
  for (i in seq_len(no_iter + 1)) {
    q <- nll_g_info(b, r, z); out[i, ] <- c(b, q$nll)
    if (i > no_iter) break
    step <- tryCatch(solve(q$info, q$g), error = function(e) rep(NA_real_, 2))
    if (any(!is.finite(step))) break
    b <- b - step
  }
  out
}

set.seed(1)
n2 <- 200; z2 <- sort(runif(n2, -2, 2)); r2 <- rbinom(n2, 1, plogis(1.5 * z2))
logit_nlm <- nlm(nll_logistic, c(0, 0), z = z2, r = r2, hessian = TRUE)
logit_nlm$estimate
```

## Example: SGD for Logistic Regression in R

```{r}
sgd_logistic <- function(z, r, b0, gamma0 = 0.5, t0 = 100, epochs = 30, batch = 10) {
  n <- length(z); b <- b0; t <- 0; path <- b0
  for (e in 1:epochs) {
    idx <- sample(n)
    for (s in seq(1, n, by = batch)) {
      i <- idx[s:min(s + batch - 1, n)]
      t <- t + 1
      u <- b[1] + b[2] * z[i]
      p <- plogis(u)
      gb <- c(sum(p - r[i]), sum(z[i] * (p - r[i]))) / length(i)   # batch gradient of L
      b <- b - gamma0 / (1 + t / t0) * gb
      path <- rbind(path, b)
    }
  }
  list(est = b, path = path)
}

set.seed(2)
beta_true2 <- c(0, 1.5)
sg <- sgd_logistic(z2, r2, c(0, 0))
rbind(sgd = sg$est, nlm = logit_nlm$estimate, truth = beta_true2)
```

## SGD Trajectory vs. Newton–Raphson

```{r}
#| echo: false
nr_path <- mle_logistic_nr(c(0, 0), 8, r2, z2)     # beta0, beta1, nll per Newton iteration
m <- 10; per_epoch <- length(z2) / m
xs <- 0:(nrow(sg$path) - 1)
xn <- (0:(nrow(nr_path) - 1)) * per_epoch

matplot(xs, sg$path, type = "l", lty = 1, col = c(2, 4),
        ylim = range(sg$path, nr_path[, 1:2]),
        xlab = "mini-batch updates (one epoch = one full pass through the data)",
        ylab = "coefficient",
        main = expression(paste("SGD vs. Newton-Raphson: ", beta[0], " (red), ", beta[1], " (blue)")))
matlines(xn, nr_path[, 1:2], type = "b", lty = 2, pch = 19, col = c(2, 4), lwd = 2)
abline(h = logit_nlm$estimate, col = c(2, 4), lty = 3)
abline(v = seq(0, max(xs), per_epoch), col = "grey85")
legend("right", c("SGD (mini-batch)", "Newton (per full pass)", "MLE"),
       lty = c(1, 2, 3), pch = c(NA, 19, NA), bty = "n", cex = 0.85)
```

- Newton–Raphson needed `r fit$iter` full-data passes; SGD made `r nrow(sg$path) - 1` cheap updates over 5 epochs and lands near the MLE with residual noise.

- Per data pass, Newton–Raphson is far more efficient here ($n = 500$, $p = 3$); SGD earns its keep when one full gradient — or the $O(p^3)$ Hessian solve — is prohibitively expensive.

# Nelder–Mead (Simplex Method)

## Overview

Maintains a **simplex** of $p+1$ points (a triangle in 2D) and moves it downhill using only function values:

1. Order the vertices $f(\theta_{(1)}) \le \cdots \le f(\theta_{(p+1)})$; let $\overline\theta$ be the centroid of the best $p$.

2. **Reflect** the worst vertex through $\overline\theta$; if the reflected point is the new best, try **expanding** further.
3. If reflection is poor, **contract** toward $\overline\theta$; if that also fails, **shrink** all vertices toward the best.

```{r}
rosen <- function(v) (1 - v[1])^2 + 100 * (v[2] - v[1]^2)^2       # Rosenbrock "banana", min at (1, 1)
nm <- optim(c(-1.2, 1), rosen, method = "Nelder-Mead"); c(nm$par, evals = nm$counts["function"])
```

Robust and derivative-free (R's `optim` default), but slow in high dimensions and offers no standard errors.

## Nelder–Mead: The Simplex Moves {.smaller}

Golden section needs a bracket, which exists only in one dimension. Nelder–Mead generalizes the idea: keep $d+1$ points (a **simplex**) in $\mathbb{R}^d$ and replace the worst one at each iteration, using **function values only** — no derivative, no bracket.

Order the vertices $f(v_1) \le \cdots \le f(v_{d+1})$, let $c$ be the centroid of all but the worst, and try in turn
$$
x_r = c + \alpha(c - v_{d+1}), \quad
x_e = c + \gamma(x_r - c), \quad
x_c = c + \rho(v_{d+1} - c),
$$
with the R defaults $\alpha = 1$, $\gamma = 2$, $\rho = \sigma = 1/2$. If even $x_c$ fails, **shrink** every vertex halfway toward $v_1$.

```{r}
#| label: nm-moves
#| echo: false
#| fig-width: 7
#| fig-height: 3.6
par(mar = c(0, 0, 0, 0))
B <- c(0.2, 0.3); G <- c(2.2, 0.6); W <- c(1.0, 2.0)
cen <- (B + G) / 2
xr <- cen + (cen - W); xe <- cen + 2 * (xr - cen); xc <- cen + 0.5 * (W - cen)
plot(NA, xlim = c(-0.4, 2.7), ylim = c(-2.9, 2.3), axes = FALSE, ann = FALSE, asp = 1)
polygon(rbind(B, G, W), lwd = 2, border = "grey30")
segments(W[1], W[2], xe[1], xe[2], lty = 3, col = "grey50")
points(rbind(B, G, W), pch = 19)
points(rbind(xr, xe, xc), pch = 19, col = c(2, 4, "orange3"))
points(cen[1], cen[2], pch = 4, lwd = 2)
text(B[1], B[2], TeX("$v_1$"), pos = 2); text(G[1], G[2], TeX("$v_2$"), pos = 4)
text(W[1], W[2], TeX("$v_3$ (worst)"), pos = 3); text(cen[1], cen[2], TeX("$c$"), pos = 4)
text(xr[1], xr[2], TeX("$x_r$ reflect"), pos = 4, col = 2)
text(xe[1], xe[2], TeX("$x_e$ expand"), pos = 4, col = 4)
text(xc[1], xc[2], TeX("$x_c$ contract"), pos = 2, col = "orange3")
```


## Summary {.smaller}

| | needs | strength | weakness |
|:--|:--|:--|:--|
| bisection / golden section | bracket | always converges | 1-D only, linear |
| Newton–Raphson | $g$, $H = \nabla^2\mathcal{L}$ | quadratic convergence, SEs from Hessian | can diverge; $O(p^3)$ |
| Fisher scoring / IRLS | $g$, $I(\theta)$ | positive-definite steps; GLMs | needs expectations |
| Nelder–Mead | $\mathcal{L}$ | derivative-free | slow, small $p$ |
| gradient descent / CG / BFGS | $g$ | scales to large $p$ | line search; ill-conditioning |
| SGD | $g_i$ | scales to large $n$ and $p$ | noisy, tuning of $\gamma_t$ |

$\mathcal{L}$ is approximately quadratic near the MLE (asymptotic normality); every method above exploits that, and its curvature $J(\hat\theta)$ at the optimum delivers the standard errors.
