---
title: "Introduction to Statistical Computation"
code-fold: true
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false
set.seed(1)
```

Statistics is a science of inference from data, and every inference eventually
becomes a calculation. This opening chapter reviews the three basic tasks of
statistical inference — point estimation, hypothesis testing, and Bayesian
inference — and shows that each of them, stated honestly, reduces to a
computational problem: calculating a quantity reliably, simulating from a
distribution, optimizing a function, or integrating over a parameter. The rest
of the book is organized around these four computational primitives, and this
chapter is a map of where each one is needed and why closed-form answers run
out so quickly outside of textbook examples.

## Review of Basic Concepts

### Statistical Inference: The Setup

A statistical model starts from data $x_1, \ldots, x_n$, assumed to be observed
values of independent and identically distributed random variables with
density or probability mass function $f(x \mid \theta)$, where $\theta$ is an
unknown parameter (possibly a vector). Almost everything a statistician does
with such a model falls into one of three tasks:

1. **Point estimation**: produce a single value $\hat\theta$ that is close to
   the unknown $\theta$;
2. **Hypothesis testing**: decide whether the data are compatible with a
   specific hypothesis $H_0$ about $\theta$;
3. **Bayesian inference**: describe the full state of uncertainty about
   $\theta$ after seeing the data, as a probability distribution rather than a
   single number.

The three tasks ask different questions of the same model, and, as this
chapter will show, they differ mainly in *what has to be computed* to answer
them.

### Point Estimation

An **estimator** is simply a function of the data,
$$
\hat\theta = \hat\theta(x_1, \ldots, x_n),
$$
computed before any particular sample is observed; once the data are in hand,
evaluating the function gives an **estimate**. Three estimators are familiar
from an introductory course:

- the sample mean $\hat\mu = \bar x = \frac{1}{n}\sum_{i=1}^n x_i$, estimating
  a population mean;
- the sample variance $\hat\sigma^2 = S^2 = \frac{1}{n-1}\sum_{i=1}^n (x_i -
  \bar x)^2$, estimating a population variance;
- the least-squares estimator in linear regression, $\hat\beta = (X'X)^{-1}
  X'y$, estimating a vector of regression coefficients.

Each is an explicit formula that a computer evaluates in one pass over the
data. The interesting statistical question is not how to compute these three
particular estimators — that part is trivial — but how to *evaluate* an
estimator once it has been proposed, and how to construct one when no such
tidy formula exists.

### Evaluating an Estimator

Because $\hat\theta$ is a function of random data, it is itself a random
variable, with its own **sampling distribution**: the distribution of the
values $\hat\theta$ would take if the whole data-collection process were
repeated over and over. Two estimators for the same $\theta$ can have sampling
distributions that differ in location and in spread, as in the schematic
below.

```{r}
#| label: sampling-distributions
#| fig-height: 3.6
#| fig-width: 9
par(mar = c(4, 1, 1, 1))
curve(dnorm(x, 0.4, 1), -3.5, 4, lwd = 3, col = "steelblue", axes = FALSE,
      xlab = expression(hat(theta)), ylab = "", ylim = c(0, 0.75))
curve(dnorm(x, 0, 0.55), add = TRUE, lwd = 3, col = "firebrick")
axis(1, at = c(0, 0.4), labels = c(expression(theta), expression(E(hat(theta)^(1)))))
abline(v = 0, lty = 2); abline(v = 0.4, lty = 3, col = "steelblue")
arrows(0, 0.72, 0.4, 0.72, code = 3, length = 0.08); text(0.2, 0.74, "bias", pos = 3)
legend("topright", bty = "n", lwd = 3, col = c("steelblue", "firebrick"),
       legend = c(expression(hat(theta)^(1)*": biased, large variance"),
                  expression(hat(theta)^(2)*": unbiased, small variance")))
```

Three numbers summarize a sampling distribution for the purpose of comparing
estimators:

- the **bias**, $E(\hat\theta) - \theta$, how far the sampling distribution's
  center sits from the truth;
- the **variance**, $V(\hat\theta)$, how spread out the sampling distribution
  is around its own center;
- the **mean squared error**, $\text{MSE}(\hat\theta) = E(\hat\theta -
  \theta)^2 = \text{Bias}(\hat\theta)^2 + V(\hat\theta)$, which combines both
  into a single criterion.

Deciding whether $\hat\theta^{(1)}$ or $\hat\theta^{(2)}$ is the better
estimator therefore requires knowing — or, when no closed form is available,
approximating — both sampling distributions. This is the first place
computation enters: when $\hat\theta$ is a complicated function of the data,
its exact sampling distribution is rarely available in closed form, and the
only way to see it is to simulate many datasets from the assumed model,
compute $\hat\theta$ on each one, and look at the resulting histogram. That
technique, Monte Carlo simulation, is the subject of the next unit.

### Maximum Likelihood Estimation

Among all conceivable estimators, one construction recurs throughout applied
statistics: the **maximum likelihood estimator**. Writing the joint density of
the data as a function of $\theta$ with $x_1, \ldots, x_n$ held fixed gives the
**likelihood**, and its logarithm the **log-likelihood**,
$$
\ell(\theta) = \log f(x_1, \ldots, x_n \mid \theta) = \log \prod_{i=1}^n f(x_i \mid \theta) = \sum_{i=1}^n \log f(x_i \mid \theta).
$$
The maximum likelihood estimator (MLE) is the value of $\theta$ that makes the
observed data most probable under the model,
$$
\hat\theta_{\text{MLE}} = \arg\max_\theta \ell(\theta).
$$
A handful of textbook models — the normal mean with known variance, the
binomial proportion — admit a closed-form maximizer found by setting a
derivative to zero and solving algebraically. The moment a model becomes even
mildly realistic — a generalized linear model, a mixed-effects model, a
survival model with censoring — the score equation $\ell'(\theta) = 0$ no
longer has an algebraic solution, and $\hat\theta_{\text{MLE}}$ must be located
by a numerical search. Locating the maximum of a function without a formula
for where it is, is **optimization**, the third computational primitive of the
course.

### Hypothesis Testing

Testing $H_0: \mu = \mu_0$ against $H_1: \mu \ne \mu_0$ proceeds by forming a
test statistic, such as
$$
T = \frac{\bar x - \mu_0}{S/\sqrt{n}},
$$
and asking how extreme the observed value of $T$ is if $H_0$ were true. That
question can only be answered by knowing the **null distribution** of $T$,
the distribution $T$ would follow under repeated sampling if $H_0$ held
exactly. For data that really are normal, $T$ follows a $t_{n-1}$ distribution
exactly, a fact discovered algebraically a century ago. For almost any other
kind of data — counts, proportions, skewed measurements, a statistic more
complicated than a studentized mean — the null distribution of $T$ is known
only approximately, through large-sample theory, or not at all in closed form.

### $p$-value, Type I Error, and Power

```{r}
#| label: null-alternative-distributions
#| fig-height: 4.2
#| fig-width: 10
par(mar = c(4, 1, 1, 1))
tobs <- 1.5; tcrit <- qnorm(0.95); mu1 <- 2.5
curve(dnorm(x), -3.5, 6.5, lwd = 3, col = "steelblue", axes = FALSE,
      xlab = "T", ylab = "", ylim = c(0, 0.45))
curve(dnorm(x, mu1), add = TRUE, lwd = 3, col = "firebrick")
shade <- function(from, to, mean, col) {
  xs <- seq(from, to, length.out = 200)
  polygon(c(xs, rev(xs)), c(dnorm(xs, mean), rep(0, 200)), col = col, border = NA)
}
shade(tcrit, 6.5, mu1, adjustcolor("firebrick", 0.25))   # power
shade(tcrit, 6.5, 0,   adjustcolor("steelblue", 0.45))   # alpha
shade(tobs,  6.5, 0,   adjustcolor("gray30", 0.35))      # p-value
axis(1, at = c(0, tobs, tcrit, mu1),
     labels = c("0", expression(t[obs]), expression(t[alpha]), expression(mu[1])))
abline(v = tcrit, lty = 2); abline(v = tobs, lty = 3)
text(-1.6, 0.30, expression(H[0]*": null distribution"), col = "steelblue")
text(4.3, 0.30, expression(H[1]*": alternative"), col = "firebrick")
text(2.0, 0.055, "p-value", pos = 4, cex = 0.9)
text(tcrit, 0.42, expression(alpha*" = type I error (blue tail)"), pos = 4, cex = 0.9)
text(3.3, 0.16, "power (red tail)", pos = 4, cex = 0.9)
```

Every quantity used to judge a hypothesis test is a tail probability computed
from one of these two curves:

| Quantity | Definition | Distribution used |
|:--|:--|:--|
| $p$-value | $P_{H_0}(T \ge t_{\text{obs}})$: probability, under $H_0$, of a statistic at least as extreme as observed | null |
| Type I error $\alpha$ | $P_{H_0}(T \ge t_\alpha)$: probability of rejecting $H_0$ when it is true | null |
| Power | $P_{H_1}(T \ge t_\alpha)$: probability of correctly rejecting $H_0$ when $H_1$ holds | alternative |

Computing any of these three numbers requires the same thing: the ability to
evaluate a tail area of a distribution that, outside of the normal-theory
special cases, is not available in closed form. Simulating the null and
alternative distributions by repeated sampling — the same Monte Carlo idea
used to study sampling distributions above — is often the most direct route
when an exact formula does not exist.

### Bayesian Inference

The Bayesian approach treats $\theta$ itself as random, assigns it a **prior**
distribution $\pi(\theta)$ encoding what is known before the data arrive, and
updates that belief with the likelihood via Bayes' theorem to obtain the
**posterior**,
$$
p(\theta \mid x) = \frac{f(x \mid \theta)\,\pi(\theta)}{\int f(x \mid \theta)\,\pi(\theta)\, d\theta}.
$$
The numerator is easy: it is just the likelihood times the prior, both of
which are specified. The denominator — the **marginal likelihood** or
**normalizing constant** — is an integral over the entire parameter space, and
for anything beyond a handful of conjugate textbook models, that integral has
no closed form, especially once $\theta$ has more than one or two components.

### Prior, Likelihood, and Posterior

```{r}
#| label: prior-likelihood-posterior
#| fig-height: 4.2
#| fig-width: 10
par(mar = c(4, 1, 1, 1))
# normal mean with known sigma: vague prior N(0, 3^2), n = 5, xbar = 2, sigma = 2
s2p <- 9; xbar <- 2; n <- 5; s2 <- 4
v_post <- 1 / (1 / s2p + n / s2); m_post <- v_post * (n * xbar / s2)
curve(dnorm(x, 0, sqrt(s2p)), -8, 8, lwd = 3, col = "gray40", lty = 2,
      axes = FALSE, xlab = expression(theta), ylab = "", ylim = c(0, 0.5))
curve(dnorm(x, xbar, sqrt(s2 / n)), add = TRUE, lwd = 3, col = "firebrick")
curve(dnorm(x, m_post, sqrt(v_post)), add = TRUE, lwd = 3, col = "steelblue")
axis(1)
legend("topleft", bty = "n", lwd = 3, lty = c(2, 1, 1),
       col = c("gray40", "firebrick", "steelblue"),
       legend = c(expression("vague prior "*pi(theta)),
                  expression("likelihood "*f(x*"|"*theta)),
                  expression("posterior "*p(theta*"|"*x))))
```

In this normal-mean example the posterior can be written down exactly, and the
picture makes the mechanism visible: a vague (wide, nearly flat) prior lets
the likelihood dominate, so the posterior sits close to the likelihood, just
slightly narrower and pulled a little toward the prior mean. Once the answer
is available, inference reports posterior summaries — the posterior mean
$E(\theta \mid x)$, the posterior variance $V(\theta \mid x)$, a credible
interval, or a posterior probability $P(\theta \in H_0 \mid x)$ — and every one
of these summaries is, again, an integral against $p(\theta \mid x)$. When the
posterior itself has no closed form, none of its summaries do either.

### Summary of the Three Tasks

| Task | Object needed | Typical difficulty |
|:--|:--|:--|
| Point estimation | Sampling distribution of $\hat\theta$; maximizer of $\ell(\theta)$ | No closed form |
| Hypothesis testing | Null distribution of $T$ | Unknown for general data |
| Bayesian inference | $\int f(x\mid\theta)\pi(\theta)\,d\theta$; $E(\theta \mid x)$ | High-dimensional integral |

Read across the rows, and a pattern emerges: each of the three classical tasks
of statistical inference reduces to a computational problem — **calculating**
a quantity accurately, **simulating** from a distribution that cannot be
sampled directly, **optimizing** a function with no algebraic maximizer, or
**integrating** over a parameter space with no closed-form antiderivative.
None of these problems is solved by more statistical theory alone; all four
are solved by numerical methods, which is what the rest of this course builds.

## The Use of Computers in Statistics

The four computational tasks identified above — calculating, simulating,
optimizing, integrating — organize the whole course. This section previews
each one with a small example, and points ahead to the unit where it is
developed properly.

### 1. Manipulating Numbers

Even simple calculations are not as trivial on a computer as they look on
paper:

- $1.1 + 2.1 + \cdots + 100.1$
- $S^2 = \sum_{i=1}^n (x_i - \bar x)^2 / (n-1)$
- $\prod_{i=1}^n f(x_i \mid \theta)$ for large $n$

A computer stores numbers with finite precision, so arithmetic that is exact
on paper can go wrong in three specific ways: **overflow**, when a result is
too large to represent and becomes `Inf`; **underflow**, when a result is too
small and is rounded to zero; and **rounding error**, the gradual loss of
significant digits as many operations accumulate. All three are silent — R
will not raise an error — so a computation can return a plausible-looking but
wrong number without any warning.

```{r}
#| echo: true
x <- c(1e10 + 1, 1e10 + 2, 1e10 + 3)
var(x)                                    # correct answer: 1
sum(x^2)/2 - 3 * mean(x)^2 / 2           # textbook formula
```

The two lines above compute the same variance by two algebraically equivalent
formulas; they disagree because the second formula first squares numbers near
$10^{10}$ and then subtracts two similarly huge quantities, destroying most of
the precision in the answer — a textbook illustration of **catastrophic
cancellation**.

```{r}
#| echo: true
prod(dnorm(rnorm(1000)))                  # likelihood underflows to 0
sum(dnorm(rnorm(1000), log = TRUE))       # log-likelihood is fine
```

Multiplying a thousand densities together underflows to exactly zero, even
though the true likelihood is a small but perfectly well-defined positive
number; summing the log-densities instead avoids the problem entirely, which
is exactly why $\ell(\theta) = \sum_i \log f(x_i \mid \theta)$, and not
$L(\theta) = \prod_i f(x_i \mid \theta)$, is the quantity every optimizer in
this course actually works with. The unit on computer arithmetic develops
these failure modes systematically — how floating-point numbers are stored,
when each failure occurs, and the standard tricks (working on the log scale,
the log-sum-exp identity, centering and scaling) that avoid them.

### 2. Simulation with Random Numbers

**Monte Carlo** methods replace a probability calculation that has no closed
form with a large number of repeated random experiments, summarized
numerically. The simplest possible illustration: if $X_1, X_2$ are independent
$\text{Unif}(0,1)$ random variables, what is the distribution of $X_1 + X_2$?
It can be worked out by hand (it is triangular on $[0, 2]$), but it can also
just be *seen*, by generating many pairs and plotting a histogram of their
sums.

```{r}
#| echo: true
#| fig-height: 3.5
z <- runif(10000) + runif(10000)
hist(z, breaks = 40, freq = FALSE, main = "", xlab = "X1 + X2")
```

The same idea scales up directly to the sampling-distribution question from
earlier in this chapter. Repeating "draw $x_1, \ldots, x_n$ from the assumed
model, then compute the statistic" many times over gives an empirical
approximation to the sampling distribution of any estimator or test
statistic — $\hat\mu^{(1)}$ versus $\hat\mu^{(2)}$, the null distribution of
$T$ for non-normal data, the actual coverage of a nominal 95% confidence
interval — without ever deriving a formula for it. The unit on random numbers
and Monte Carlo develops this systematically: how pseudo-random numbers are
generated, how draws from one distribution are transformed into draws from
another, the Law of Large Numbers and Central Limit Theorem that justify
treating a simulation average as an estimate, and how to use simulation to
evaluate estimators and tests.

### 3. Optimization

The MLE problem above is one instance of a completely general task: find
$x_0 = \arg\max_x f(x)$.

```{r}
#| fig-height: 3.5
curve(dgamma(x, 3, 1), 0, 10, ylab = "f(x)", lwd = 2, main = "")
abline(v = 2, lty = 2)
text(2.1, 0.05, expression(x[0]), adj = 0)
```

For a gamma density plotted over its whole domain, as above, the maximizer is
obvious by eye. For a log-likelihood $\ell(\theta)$ with many parameters,
plotting is impossible and "obvious by eye" is not an option: the maximum has
to be found by an iterative numerical search that starts somewhere and moves
uphill. The unit on MLE and optimization develops these searches in order of
increasing generality — bisection and golden-section search in one dimension,
Newton–Raphson and Fisher scoring using derivative information, and
gradient-based methods (BFGS, stochastic gradient descent) for
high-dimensional problems where forming a Hessian is too expensive.

### 4. Advanced Simulation and Integration

The quantities that summarize a distribution are integrals:
$$
E(X) = \int x f(x)\,dx, \qquad V(X), \qquad \int f(x\mid\theta)\pi(\theta)\,d\theta.
$$
Two broad strategies approximate an integral that has no closed form.
**Numerical quadrature** approximates the area under the integrand by a sum of
rectangles, trapezoids, or higher-order pieces; it is accurate and efficient
in one or two dimensions, but its cost grows exponentially with the number of
dimensions, which makes it impractical beyond a handful of parameters.
**Monte Carlo integration** instead draws random points from (or related to)
$f$ and averages, trading the deterministic accuracy of quadrature for a
convergence rate that does not depend on dimension — the only practical route
once $\theta$ has more than a few components. For a Bayesian posterior
$p(\theta \mid x)$, direct sampling is typically impossible because the
normalizing constant is unknown, which is precisely the setting where
**Markov chain Monte Carlo** (MCMC) is used: a Markov chain is constructed
whose stationary distribution is the posterior itself, so that simulating the
chain for long enough produces (dependent) draws from $p(\theta \mid x)$
without ever having to evaluate the integral in its denominator.

```{r}
#| fig-height: 4
f <- function(x) dgamma(x, 3, 1)
curve(f, 0, 8, lwd = 2, ylab = "f(x)", main = "")
k <- 12; xs <- seq(0, 8, length.out = k + 1)
rect(xs[-(k+1)], 0, xs[-1], f((xs[-(k+1)] + xs[-1]) / 2),
     col = adjustcolor("red", 0.3), border = "red")
```

The rectangles above illustrate the quadrature idea directly:
$$
\int_a^b f(x)\,dx \approx \sum_{j=1}^k f(m_j)\,\Delta x,
$$
where $m_j$ is the midpoint of the $j$th rectangle of width $\Delta x$. The
later units on numerical quadrature, Laplace approximation, rejection and
importance sampling, and Markov chain Monte Carlo develop this strategy and
its Monte Carlo counterpart from the ground up, culminating in the algorithms
that make modern Bayesian computation possible.

### Course Roadmap

| Topic | Statistical task |
|:--|:--|
| Computer arithmetic | Reliable calculation of likelihoods and variances |
| Random numbers and Monte Carlo | Sampling distributions, evaluating estimators and tests |
| MLE and optimization | Point estimation without closed forms |
| Numerical integration and MCMC | Bayesian inference |

Throughout the course, three themes recur alongside the statistical content:
programming in R, writing vectorized code instead of explicit loops, and
paying attention to the efficiency and numerical stability of the code itself
— since, as the arithmetic examples above already show, a correct formula
implemented carelessly can still give a wrong answer.
