---
title: "Importance Sampling"
format:
  html:
    toc: true
    tbl-colwidths: true
filters:
  - shinylive
editor_options:
  chunk_output_type: console
code-fold: true
---

```{r}
#| label: setup
#| message: false
library(ars)
```

Rejection sampling turns draws from an envelope $g \ge f$ into exact draws from
$f$, but every rejected proposal is wasted, and in the tails of a poor
envelope almost every proposal is rejected. **Importance sampling** keeps
every proposal instead of discarding most of them: draw from an easier
function $g$ and attach a *weight* to each draw that corrects for sampling
from the wrong distribution. The price is that the resulting sample is no
longer an exact iid sample from $f$ — it is a *weighted* sample — but nothing
is ever thrown away, and no envelope condition $g \ge f$ is required.

## The Reweighting Idea

### The goal

As in the last two chapters, $f(\theta)$ is an **unnormalized** function of
interest — typically $f(\theta) = L(\theta)\pi(\theta)$ for a likelihood $L$
and prior $\pi$ — and we want two kinds of integral:
$$
A = E_f\big[a(\theta)\big] = \frac{\int a(\theta)\,f(\theta)\,d\theta}{\int f(\theta)\,d\theta},
\qquad\qquad
C_f = \int f(\theta)\,d\theta,
$$
a posterior expectation and a normalizing constant (the marginal likelihood
when $f = L\pi$). Plain Monte Carlo needs draws from $f/C_f$, which we
generally cannot produce.

Suppose instead we can draw from another function $g(\theta)$ (also possibly
unnormalized, with $C_g = \int g(\theta)\,d\theta$). Wherever $f$ puts mass,
$g$ must too:
$$
\{\theta : g(\theta) > 0\} \ \supseteq\ \{\theta : f(\theta) > 0\}.
$$
Unlike rejection sampling, $g$ need **not** dominate $f$ everywhere — there is
no requirement $g \ge f$. This is what makes importance sampling applicable
in situations where no usable envelope exists.

### Two estimators

Define the **importance weight** $w(\theta) = f(\theta)/g(\theta)$. For any
$a$,
$$
E_f\big[a(\theta)\big]
= \int a(\theta)\,\frac{f(\theta)}{C_f}\,d\theta
= \frac{C_g}{C_f}\, E_g\big[a(\theta)\,w(\theta)\big],
$$
and taking $a \equiv 1$ gives $C_f/C_g = E_g[w(\theta)]$. Combining the two,
$$
E_f\big[a(\theta)\big] = \frac{E_g\big[a(\theta)\,w(\theta)\big]}{E_g\big[w(\theta)\big]}.
$$
Draw $\theta_1,\ldots,\theta_S \overset{iid}{\sim} g/C_g$ and compute
$w_s = w(\theta_s)$. Two estimators follow directly:

- **Ratio of normalizing constants**: $\widehat{C_f/C_g} = \frac1S\sum_s w_s$,
  unbiased for $C_f/C_g$; if $g$ is already a normalized density ($C_g=1$)
  this estimates $C_f$ itself.
- **Self-normalized expectation**: $\hat A = \sum_{s=1}^S a(\theta_s)\,\tilde w_s$,
  with normalized weights $\tilde w_s = w_s / \sum_j w_j$ summing to $1$. This
  turns the proposal sample into a *weighted sample from $f$*: any unknown
  constants in $f$ and $g$ cancel in the ratio, so $\hat A$ needs only $f$ and
  $g$ up to a proportionality constant. It is a ratio of two averages and so
  is biased with bias $O(1/S)$, but consistent.

### Effective sample size

$\hat A$ is a weighted average of $a(\theta_s)$: draws where $f \gg g$ count
more, draws where $f \ll g$ count less. How many *equally weighted* draws is
the weighted sample worth?
$$
S_{\text{eff}} = \frac{1}{\displaystyle\sum_{s=1}^S \tilde w_s^2}.
$$
If the weights are all equal, $\tilde w_s = 1/S$ and $S_{\text{eff}} =
1/(S \cdot 1/S^2) = S$: the proposal is shaped just like the target, and every
draw counts fully. If one weight dominates all the others, $\tilde w_s
\approx 1$ for that draw and $S_{\text{eff}} \approx 1$: only one draw is
doing any work. A variance heuristic makes this precise:
$\text{Var}_g(\hat A) \approx \text{Var}_f(a)/S_{\text{eff}}$, the same form as
for an iid sample of size $S_{\text{eff}}$ from $f$ itself.

The catch, central to this chapter, is that $S_{\text{eff}}$ is computed
**only from the weights actually observed**. A proposal that never visits the
part of $\theta$-space where $f$ and $g$ disagree most will report a
reassuring $S_{\text{eff}}$ while being badly wrong — this is exactly the
subject of the "bad stories" below.

```{r}
#| label: ess-helpers
### effective sample size from raw (possibly unnormalized) weights
ess <- function(w) sum(w)^2 / sum(w^2)
### numerically stable log(sum(exp(lx)))
log_sum_exp <- function(lx) { m <- max(lx); m + log(sum(exp(lx - m))) }
### effective sample size computed from LOG weights (avoids under/overflow)
log_ess <- function(lw) exp(2 * log_sum_exp(lw) - log_sum_exp(2 * lw))
```

## Example 1: Tail Probability of a Student-$t$ Distribution {#sec-ars-tnorm}

The target is a standard Student-$t_{\nu}$ distribution, with the degrees of
freedom $\nu$ a **tuning parameter of the target** that controls how heavy
its tail is: $\nu=1$ is the standard Cauchy, and $\nu\to\infty$ recovers the
standard normal. The goal is $P(c_{1}<X<c_{2})$ for an interval $(c_{1},c_{2})$
— possibly one-sided, $c_{2}=\infty$ — estimated from draws of a
truncated-normal proposal generated by adaptive rejection sampling (ARS).

As in the ARS section of the previous chapter, $\log\varphi$ is concave, and
restricting to $(c_{1},c_{2})$ (or one-sided, $(c_{1},\infty)$) preserves
concavity, so the `ars` package applies directly: `rtnorm_ars()` below just
hands it the log-density and its derivative, plus the truncation bounds.

```{r}
#| label: rtnorm-ars
## ---------------------------------------------------------------------
##  rtnorm_ars(): draw from N(mu, sigma^2) truncated to (lo, hi) (hi may
##  be Inf) via the ars package. log-phi is concave, so ars() applies
##  directly; x0 gives it three starting abscissae inside (lo, hi), with
##  the rightmost past mu when hi = Inf so the upper piece has negative
##  slope and stays integrable.
## ---------------------------------------------------------------------
rtnorm_ars <- function(n, mu = 0, sigma = 1, lo = 0, hi = Inf) {
    logf   <- function(x) dnorm(x, mu, sigma, log = TRUE)
    fprima <- function(x) -(x - mu) / sigma^2
    if (is.finite(hi)) {
        x0 <- lo + (hi - lo) * c(0.1, 0.5, 0.9)
    } else {
        right <- max(mu, lo) + 3 * sigma
        x0 <- sort(unique(c(lo + 0.1 * sigma, max(lo + 0.2 * sigma, mu + 1e-6), right)))
    }
    ars(n, f = logf, fprima = fprima, x = x0,
        lb = TRUE, xlb = lo, ub = is.finite(hi), xub = if (is.finite(hi)) hi else 0)
}
```

The `ars` package is itself available inside the browser — webR fetches it
from CRAN's WebAssembly mirror the first time `library(ars)` runs — so the
interactive app below calls this exact function, not a hand-rolled
substitute.

The contrast with the target is the point of the exercise: ARS needs
log-concavity, and $\log f_{t_{\nu}}$ is simply **not concave**, for any
finite $\nu$ — it is convex out in the tails, and only the limit
$\nu\to\infty$ (the normal) is concave everywhere. So ARS cannot sample the
target directly at any finite $\nu$; the truncated-normal proposal above is a
substitute, not a target-side match, which is exactly why the sample needs
importance weights at all.

### The estimator, and when its variance is infinite

A one-sided tail $(c,\infty)$ is a bad target for this proposal, so consider
instead a general interval $(c_{1},c_{2})$, with $c_{2}=\infty$ allowed as a
special case. Let $g$ be the normalized truncated normal on $(c_{1},c_{2})$
and $f(x)=f_{t_{\nu}}(x)I(c_{1}<x<c_{2})$. Draw $X_{1},\dots,X_{S}\sim g$,
weight each by $w(x)=f(x)/g(x)$, and average:
$$
\widehat{C} \;=\; \frac{1}{S}\sum_{i=1}^{S} w(X_{i}).
$$
This is unbiased for $C_{f}=P(c_{1}<X<c_{2})$ for *any* $g$ that is positive
wherever $f$ is — that is the whole mechanism of importance sampling, and it
does not care what $c_{1},c_{2},\mu,\sigma,\nu$ are.

Whether $\widehat C$ is *precise* is a different question, governed by
$$
\operatorname{Var}_{g}\bigl(w(X)\bigr)
 \;=\; \int_{c_{1}}^{c_{2}}\frac{f_{t_{\nu}}(x)^{2}}{g(x)}\,dx \;-\; C_{f}^{2} .
$$
When $c_{2}$ is finite, $f$ and $g$ share the same bounded interval, the
ratio $f/g$ is bounded on it, and the integral is finite — a well-behaved
estimator, whatever $\nu$ is. Send $c_{2}\to\infty$, though, and the tails
stop matching: $f_{t_{\nu}}(x)^{2}$ decays only polynomially
($\propto x^{-2(\nu+1)}$) while $g$'s Gaussian tail decays much faster, so the
integrand grows without bound and the integral diverges — for *every*
$\mu,\sigma$ and *every* finite $\nu$. The app below lets you flip between
the two regimes with a single checkbox, and, as before, $S_{\text{eff}}$ need
not warn you when you have landed in the bad one.

`is_t_interval()` turns the estimator above into code: draw $X$ from
`rtnorm_ars()`, form the log-weight $\log f_{t_\nu}(X) - \log g(X)$, and
report $\widehat C$, its standard error, and $S_{\text{eff}}$.

```{r}
#| label: is-t-interval
## ---------------------------------------------------------------------
##  is_t_interval(): importance-sampling estimate of P(c1 < X < c2) for a
##  standard t_df target, using X ~ N(mu, sigma^2) truncated to (c1, c2)
##  as the proposal.
## ---------------------------------------------------------------------
is_t_interval <- function(S, df, c1, c2, mu, sigma, seed = NULL) {
    if (!is.null(seed)) set.seed(seed)
    X <- rtnorm_ars(S, mu, sigma, c1, c2)

    ## log importance weight  log f_t(x; df) - log g(x),  with g the
    ## normalized truncated normal on (c1, c2)
    lZ   <- log(pnorm(c2, mu, sigma) - pnorm(c1, mu, sigma))
    logw <- dt(X, df, log = TRUE) - (dnorm(X, mu, sigma, log = TRUE) - lZ)

    m    <- max(logw)
    v    <- exp(logw - m)                       # rescaled weights
    Chat <- mean(v) * exp(m)                    # unbiased estimate of P(c1 < X < c2)
    se   <- sd(v) / sqrt(S) * exp(m)            # nominal standard error
    wn   <- v / sum(v)                          # normalized weights

    list(X = X, wn = wn, Chat = Chat, se = se, ess = 1 / sum(wn^2),
         truth = pt(c2, df) - pt(c1, df))
}
```

Four settings make the point concretely: a finite interval is well-behaved
at any $\nu$, while $c_{2}=\infty$ reproduces the divergence — badly at
$\nu=1$, but deceptively well at $\nu=30$, exactly as @tbl-is-t-examples
shows.

```{r}
#| label: tbl-is-t-examples
#| tbl-cap: "Four calls to `is_t_interval()` with S = 5000: a finite c2 is
#|   well-behaved at every df, c2 = Inf reproduces the divergent case,
#|   badly at df = 1 and only deceptively well at df = 30."
settings <- data.frame(df = c(1, 10, 1, 30), c1 = c(1, 1, 1, 1), c2 = c(3, 3, Inf, Inf))
res <- lapply(seq_len(nrow(settings)), function(i) with(settings[i, ],
    is_t_interval(S = 5000, df = df, c1 = c1, c2 = c2, mu = 0, sigma = 1, seed = 1)))

knitr::kable(data.frame(
    df       = settings$df,
    c1       = settings$c1,
    c2       = ifelse(is.finite(settings$c2), as.character(settings$c2), "Inf"),
    truth    = sapply(res, `[[`, "truth"),
    estimate = sapply(res, `[[`, "Chat"),
    se       = sapply(res, `[[`, "se"),
    ess_pct  = 100 * sapply(res, `[[`, "ess") / 5000
), digits = 4, col.names = c("df", "c1", "c2", "truth", "estimate", "s.e.", "ESS %"))
```

### An interactive view {#sec-t-is-app}

The app below puts $c_{1}$, $c_{2}$, $\mu$, $\sigma$ and the target's degrees
of freedom $\nu$ under slider control (a checkbox sends $c_{2}$ to $\infty$),
and displays, on a common $x$-axis: the two truncated densities on a log
scale with a rug of the draws, and the normalized importance weights
$\tilde w_{i}=w_{i}/\sum_{j}w_{j}$ at the sampled points. The legend reports
the true $P(c_{1}<X<c_{2})$, the estimate, and its nominal 95% interval,
colored green when the interval covers the truth and red when it does not.
Leave $c_{2}$ finite and the estimator behaves; check the box and watch it
misbehave exactly as the variance calculation predicts, while
$S_{\text{eff}}$ in the bottom panel need not warn you either way. This app
runs live only in the HTML edition of this book; readers of other formats can
follow the link below it.

::: {#fig-t-is-app .content-visible when-format="html"}

```{shinylive-r}
#| label: is-t-app
#| standalone: true
#| viewerHeight: 850

library(shiny)
library(ars)

## rtnorm_ars() and is_t_interval(): identical to the versions defined
## earlier in the chapter (a {shinylive-r} chunk cannot see objects
## defined elsewhere in the document, so they are pasted in verbatim).
rtnorm_ars <- function(n, mu = 0, sigma = 1, lo = 0, hi = Inf) {
    logf   <- function(x) dnorm(x, mu, sigma, log = TRUE)
    fprima <- function(x) -(x - mu) / sigma^2
    if (is.finite(hi)) {
        x0 <- lo + (hi - lo) * c(0.1, 0.5, 0.9)
    } else {
        right <- max(mu, lo) + 3 * sigma
        x0 <- sort(unique(c(lo + 0.1 * sigma, max(lo + 0.2 * sigma, mu + 1e-6), right)))
    }
    ars(n, f = logf, fprima = fprima, x = x0,
        lb = TRUE, xlb = lo, ub = is.finite(hi), xub = if (is.finite(hi)) hi else 0)
}

is_t_interval <- function(S, df, c1, c2, mu, sigma, seed = NULL) {
    if (!is.null(seed)) set.seed(seed)
    X <- rtnorm_ars(S, mu, sigma, c1, c2)

    lZ   <- log(pnorm(c2, mu, sigma) - pnorm(c1, mu, sigma))
    logw <- dt(X, df, log = TRUE) - (dnorm(X, mu, sigma, log = TRUE) - lZ)

    m    <- max(logw)
    v    <- exp(logw - m)
    Chat <- mean(v) * exp(m)
    se   <- sd(v) / sqrt(S) * exp(m)
    wn   <- v / sum(v)

    list(X = X, wn = wn, Chat = Chat, se = se, ess = 1 / sum(wn^2),
         truth = pt(c2, df) - pt(c1, df))
}

ui <- fluidPage(
    titlePanel("Importance sampling for Estimating a Student-t Probabilities with a truncated-normal proposal"),
    sidebarLayout(
        sidebarPanel(
            width = 3,
            sliderInput("df", HTML("target df &nu;"), min = 1, max = 30, value = 3, step = 1),
            sliderInput("c1", HTML("c<sub>1</sub>"), min = -5, max = 10, value = 1, step = 0.25),
            sliderInput("c2", HTML("c<sub>2</sub>"), min = -5, max = 15, value = 3, step = 0.25),
            checkboxInput("c2inf", HTML("c<sub>2</sub> = &infin;  (infinite-variance case)"), FALSE),
            sliderInput("mu", HTML("proposal &mu;"), min = -5, max = 20, value = 0, step = 0.25),
            sliderInput("sg", HTML("proposal &sigma;"), min = 0.25, max = 10, value = 1, step = 0.25),
            selectInput("S", "sample size  S",
                        choices = c(200, 1000, 5000, 20000), selected = 1000),
            actionButton("new", "New sample"),
            checkboxInput("logw", "log scale for weights", FALSE)
        ),
        mainPanel(width = 9, plotOutput("panels", height = "680px"))
    )
)

server <- function(input, output) {

    sim <- reactive({
        S  <- as.integer(input$S)
        c1 <- input$c1
        c2 <- if (input$c2inf) Inf else input$c2
        validate(need(c2 > c1, "c2 must be greater than c1"))
        mu <- input$mu; sg <- input$sg; df <- input$df

        r <- is_t_interval(S, df, c1, c2, mu, sg, seed = input$new + 1)
        c(r, list(S = S, c1 = c1, c2 = c2, mu = mu, sg = sg, df = df))
    })

    output$panels <- renderPlot({
        s  <- sim()
        c1 <- s$c1; c2 <- s$c2; X <- s$X
        lo <- c1
        hi <- if (is.finite(c2)) c2 else max(max(X) + 0.5, c1 + 5)
        xs <- seq(lo, hi, length.out = 800)

        ## both densities normalized on (c1, c2)
        ft <- dt(xs, s$df) / s$truth
        gt <- dnorm(xs, s$mu, s$sg) / (pnorm(c2, s$mu, s$sg) - pnorm(c1, s$mu, s$sg))
        yl <- range(c(ft, gt)[c(ft, gt) > 0])
        yl[1] <- max(yl[1], 1e-12)

        lwr <- s$Chat - 1.96 * s$se
        upr <- s$Chat + 1.96 * s$se
        covered <- lwr <= s$truth && s$truth <= upr
        ci_col  <- if (covered) "darkgreen" else "firebrick"

        layout(matrix(1:2, 2), heights = c(3, 2))

        ## ---- top panel: the two densities ----------------------------
        par(mar = c(0.4, 5, 3, 1))
        plot(xs, ft, type = "n", log = "y", ylim = yl, xlim = c(lo, hi),
             xaxt = "n", xlab = "", ylab = "density (log scale)",
             main = "target and proposal, both normalized on (c1, c2)")
        if (!is.finite(c2)) {
            rect(max(X), yl[1], hi, yl[2] * 10,
                 col = adjustcolor("grey60", 0.18), border = NA)
            text(max(X), yl[2], "  no draw ever reached here", adj = c(0, 1.4),
                 col = "grey35", cex = 0.95)
        }
        lines(xs, ft, lwd = 4, col = "firebrick")
        lines(xs, gt, lwd = 3, col = "steelblue")
        rug(X, col = adjustcolor("steelblue", 0.25))
        legend("bottomleft", inset = c(0.02, 0.05), bty = "n", cex = 1.05,
               lwd = c(4, 3, NA, NA, NA),
               col = c("firebrick", "steelblue", NA, NA, NA),
               text.col = c("black", "black", "black", "black", ci_col),
               legend = c(
                   sprintf("target: t(df=%g) on (c1, c2)", s$df),
                   sprintf("proposal: N(%.2f, %.2f^2) on (c1, c2)", s$mu, s$sg),
                   sprintf("true  P = %.5g", s$truth),
                   sprintf("est.   P = %.5g", s$Chat),
                   sprintf("95%% CI = (%.5g, %.5g)", lwr, upr)))

        ## ---- bottom panel: normalized weights ------------------------
        par(mar = c(4.5, 5, 0.4, 1))
        wn <- s$wn
        if (input$logw) {
            wl <- c(max(min(wn[wn > 0]), 1e-12), max(wn))
            plot(X, wn, type = "n", log = "y", ylim = wl, xlim = c(lo, hi),
                 xlab = "x", ylab = "normalized weight (log scale)")
            segments(X, wl[1], X, wn, col = adjustcolor("grey50", 0.6))
        } else {
            plot(X, wn, type = "n", ylim = c(0, max(wn) * 1.08),
                 xlim = c(lo, hi), xlab = "x", ylab = "normalized weight")
            segments(X, 0, X, wn, col = adjustcolor("grey50", 0.6))
        }
        abline(h = 1 / s$S, lty = 2, col = "darkgreen", lwd = 2)
        legend("topright", inset = c(0.02, 0.05), bty = "n", cex = 1.05,
               lty = c(2, NA, NA), lwd = c(2, NA, NA),
               col = c("darkgreen", NA, NA),
               text.col = "blue",
               legend = c("1 / S  (perfectly balanced weights)",
                          sprintf("ESS = %.1f   (%.2f%% of S = %d)",
                                  s$ess, 100 * s$ess / s$S, s$S),
                          sprintf("largest weight = %.1f%% of the total",
                                  100 * max(wn))))
    })
}

shinyApp(ui, server)
```

**Shinylive App for Importance Sampling of a Student-$t$ Interval.** Sliders
control $c_{1}$, $c_{2}$ (or $c_{2}=\infty$ via the checkbox), the target's
degrees of freedom $\nu$, and the proposal's $\mu,\sigma$; the top panel
compares the (normalized) target and proposal densities on a log scale, with
the CI in the legend colored by whether it covers the truth, and the bottom
panel shows the normalized importance weights at the sampled points.

:::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit12_importance_sampling.html#sec-t-is-app>.
:::

::: {.callout-note}
### Reproducing this figure
The chunk above uses the [shinylive](https://github.com/quarto-ext/shinylive)
extension, which compiles the app to WebAssembly so that it runs in the
reader's browser with no Shiny server (`quarto add quarto-ext/shinylive`, then
`filters: [shinylive]` in the document or project YAML, already set for this
book). A `{shinylive-r}` chunk cannot see objects defined elsewhere in the
document, so `rtnorm_ars()` and `is_t_interval()` are pasted inside the app
chunk verbatim rather than reused from the chunks above. `library(ars)` works
inside the app too: webR fetches CRAN packages compiled to WebAssembly on
demand, and `ars` is one of them. To develop the app interactively instead,
an ordinary `{r}` chunk with `shinyApp(ui, server)` at the end can be run
directly in RStudio or Positron, which opens it in the viewer pane.
:::

Several settings repay attention.

- **$c_{2}$ finite (the default).** Target and proposal share the same
  bounded support, so the weights have finite variance: raising $S$ improves
  the estimate at the usual $1/\sqrt S$ rate, and the 95% interval covers
  about as often as it claims to.
- **Check "$c_{2}=\infty$".** This reproduces the divergent case derived
  above. Raising $S$ no longer shrinks a variance that does not exist, and
  repeated *New sample* draws swing far more than the reported standard error
  admits. With large $\nu$ the run-away region is rarely reached, so the app
  can still look fine by chance; with small $\nu$ (Cauchy, $\nu=1$) it bites
  often — but $S_{\text{eff}}$ need not warn you either way.
- **Small $\sigma$, or $\mu$ far outside $(c_{1},c_{2})$.** Draws pile up
  wherever the proposal happens to put mass inside the interval, the weights
  are nearly equal, $S_{\text{eff}}$ stays high, and the estimate can still
  miss badly: uniform weights certify only that the proposal is internally
  consistent, not that it matches the target's shape there.
- **Large $\sigma$.** One draw's weight can dominate the sum,
  $S_{\text{eff}}$ collapses toward $1$ — visibly unreliable, which is an
  improvement on being invisibly unreliable.


## Example 2: Marginal Likelihood via Importance Sampling

### Setup

The [quadrature](unit09_integral.qmd) and Laplace-approximation chapters
compute the marginal likelihood $P(y) = \int L(\theta)\pi(\theta)\,d\theta$ of
the same conjugate-free Gaussian model — unknown mean $\mu$ and
$w=\log\sigma^2$, independent normal priors — by a product grid and by a
mode-and-Hessian approximation, respectively. Importance sampling offers a
third route, and one that comes with a built-in accuracy diagnostic
($S_{\text{eff}}$) that neither of the other two methods provides directly.

Take $g(\mu,w)$ to be the **Gaussian approximation to the posterior**
obtained by a Laplace fit: locate the posterior mode $(\hat\mu,\hat w)$ by
optimization, take $g = N_2\big((\hat\mu,\hat w),\, H^{-1}\big)$ for $H$ the
Hessian of $-\log[L\pi]$ at the mode, and importance-sample from $g$ with
$f = L\pi$:
$$
\widehat{P(y)} = \frac{1}{S}\sum_{s=1}^S w_s, \qquad
w_s = \frac{L(\theta_s)\pi(\theta_s)}{g(\theta_s)}, \qquad \theta_s \sim g.
$$
Because $g$ is shaped like the posterior near its mode, this is a "good
story" proposal by construction — in sharp contrast to sampling directly from
the (typically much wider, uninformative) prior, which is exactly the
`log_marlik_mc` function used elsewhere in this book as a naive-Monte-Carlo
check: it is importance sampling with $g = \pi$, and its weights are just the
likelihood values $L(\theta_s)$.

### An efficient implementation

```{r}
#| label: marlik-functions
log_lik   <- function(x, mu, w) sum(dnorm(x, mu, exp(w / 2), log = TRUE))
log_prior <- function(mu, w, mu_0, sigma_mu, w_0, sigma_w) {
    dnorm(mu, mu_0, sigma_mu, log = TRUE) + dnorm(w, w_0, sigma_w, log = TRUE)
}
neg_log_post <- function(theta, x, mu_0, sigma_mu, w_0, sigma_w) {
    -log_lik(x, theta[1], theta[2]) -
        log_prior(theta[1], theta[2], mu_0, sigma_mu, w_0, sigma_w)
}

### log density of N(mu, A^{-1}) at every column of a 2 x S matrix of points at once,
### replacing a per-point apply() call with one vectorized quadratic form
log_dmvnorm_batch <- function(Theta, mu, A) {
    Tc <- Theta - mu
    quad <- colSums((A %*% Tc) * Tc)
    0.5 * (-nrow(Theta) * log(2 * pi) + sum(log(svd(A)$d)) - quad)
}

### log marginal likelihood by importance sampling from a Laplace-shaped Gaussian.
### Besides S_eff, attaches "se": the standard error of the *log* estimate,
### sd(v) / (sqrt(S) * mean(v)) for v = exp(lw - max(lw)) -- the delta-method
### transfer of the raw estimator's SE onto the log scale, exactly as used for
### the log estimate in the t-interval app above.
log_marlik_imps <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    fit <- nlm(neg_log_post, c(mean(x), log(stats::var(x))), hessian = TRUE,
               x = x, mu_0 = mu_0, sigma_mu = sigma_mu, w_0 = w_0, sigma_w = sigma_w)
    mu_hat <- fit$estimate; H <- fit$hessian
    L <- t(chol(solve(H)))                        # Sigma = H^{-1} = L L'
    Theta <- L %*% matrix(rnorm(2 * S), 2, S) + mu_hat

    log_g <- log_dmvnorm_batch(Theta, mu_hat, H)
    log_f <- -apply(Theta, 2, neg_log_post, x = x, mu_0 = mu_0, sigma_mu = sigma_mu,
                     w_0 = w_0, sigma_w = sigma_w)
    lw <- log_f - log_g
    m  <- max(lw); v <- exp(lw - m)
    structure(log_sum_exp(lw) - log(S),
              ess = log_ess(lw), se = sd(v) / (sqrt(S) * mean(v)))
}

### log marginal likelihood by naive Monte Carlo from the prior (g = pi): a
### second, independent importance-sampling estimator with an unfavourable g
log_marlik_prior <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    mus <- rnorm(S, mu_0, sigma_mu); ws <- rnorm(S, w_0, sigma_w)
    lw <- vapply(seq_len(S), function(i) log_lik(x, mus[i], ws[i]), numeric(1))
    m  <- max(lw); v <- exp(lw - m)
    structure(log_sum_exp(lw) - log(S),
              ess = log_ess(lw), se = sd(v) / (sqrt(S) * mean(v)))
}

### "truth": log marginal likelihood by a midpoint grid over (mu, w), centred
### and scaled by the Laplace fit. The likelihood term is vectorized over the
### whole mu-grid at once via the sufficient statistic sum((x - mu)^2), so --
### like log_dmvnorm_batch above -- no R-level loop over grid points, or over
### data points beyond the one sum(), is needed.
log_marlik_grid <- function(x, mu_0, sigma_mu, w_0, sigma_w, n_grid = 400, width = 8) {
    n   <- length(x)
    fit <- nlm(neg_log_post, c(mean(x), log(stats::var(x))), hessian = TRUE,
               x = x, mu_0 = mu_0, sigma_mu = sigma_mu, w_0 = w_0, sigma_w = sigma_w)
    se  <- sqrt(diag(solve(fit$hessian)))
    mu_grid <- fit$estimate[1] + seq(-width, width, length.out = n_grid) * se[1]
    w_grid  <- fit$estimate[2] + seq(-width, width, length.out = n_grid) * se[2]

    SS <- sum(x^2) - 2 * mu_grid * sum(x) + n * mu_grid^2   # sum((x - mu)^2)
    log_lik_grid   <- outer(SS, w_grid,
                             function(ss, w) -0.5 * n * log(2 * pi) - n * w / 2 - 0.5 * exp(-w) * ss)
    log_prior_grid <- outer(dnorm(mu_grid, mu_0, sigma_mu, log = TRUE),
                             dnorm(w_grid, w_0, sigma_w, log = TRUE), "+")

    h_mu <- diff(mu_grid[1:2]); h_w <- diff(w_grid[1:2])
    log_sum_exp(as.vector(log_lik_grid + log_prior_grid)) + log(h_mu) + log(h_w)
}
```

Replacing the two `apply()` calls used in earlier drafts of this code — one
for the proposal log-density, one for the log-posterior — with a single
vectorized quadratic form (`log_dmvnorm_batch`) removes $S$ separate R
function calls for the proposal evaluation; only the data-dependent
log-likelihood still needs one call per draw, since it sums over the $n$
observations for each $\theta_s$ separately.

### Checking accuracy

`log_marlik_grid()` supplies ground truth by quadrature — a midpoint grid
over $(\mu,w)$, centred and scaled by the same Laplace fit used for the
importance proposal, wide enough (8 posterior SDs each way, on a $400\times
400$ grid) that refining it further does not move the answer. Comparing both
proposals against it in a single run, with a 95% CI built from the standard
error of that run, makes the difference between them concrete rather than
just asserted.

```{r}
#| label: tbl-marlik-check
#| tbl-cap: "Log marginal likelihood: truth by grid quadrature versus a single
#|   run of importance sampling (S = 2000) from a Laplace-shaped proposal and
#|   from the prior, for two simulated data sets."
set.seed(4)
x_small <- rnorm(10);  x_large <- rnorm(200)

marlik_row <- function(x, dataset, proposal_fn, proposal, S = 2000) {
    truth <- log_marlik_grid(x, 0, 5, 0, 5)
    est   <- proposal_fn(x, 0, 5, 0, 5, S)
    se    <- attr(est, "se"); ess <- attr(est, "ess")
    data.frame(dataset = dataset, proposal = proposal,
               truth = truth, estimate = as.numeric(est),
               ci_lower = as.numeric(est) - 1.96 * se,
               ci_upper = as.numeric(est) + 1.96 * se,
               ess_pct = 100 * ess / S)
}

check_tbl <- rbind(
    marlik_row(x_small, "n = 10",  log_marlik_imps,  "Laplace"),
    marlik_row(x_small, "n = 10",  log_marlik_prior, "prior"),
    marlik_row(x_large, "n = 200", log_marlik_imps,  "Laplace"),
    marlik_row(x_large, "n = 200", log_marlik_prior, "prior")
)
knitr::kable(check_tbl, digits = 3,
             col.names = c("dataset", "proposal", "truth", "estimate",
                           "95% CI lower", "95% CI upper", "ESS %"))
```

At both sample sizes the Laplace-shaped proposal's CI is tight and centred on
the truth, with $S_{\text{eff}}$ close to the nominal $S=2000$ — a second-order
approximation to the posterior is evidently a good match to it. The prior's
CI can still cover the truth (it does, barely, at $n=200$ above), but its
$S_{\text{eff}}$ collapses to a fraction of a percent as $n$ grows: a handful
of draws are carrying the entire estimate, so that CI is only as trustworthy
as those few draws happen to be, run to run — exactly the "noisy failure"
this chapter's diagnostic is built to catch, in contrast to the *silent*
failures earlier where $S_{\text{eff}}$ looked fine and the estimate was
wrong anyway.

### Why the proposal matters: Laplace-shaped versus prior sampling

The prior is a valid proposal for the same target — every posterior draw is
also a prior draw with positive density — so `log_marlik_prior` is a
legitimate importance-sampling estimator too. But the prior is far more
diffuse than the posterior, especially once $n$ is not tiny, so it is a much
poorer match to $f = L\pi$.

```{r}
#| label: tbl-marlik-ess-vs-n
#| tbl-cap: "Effective sample size (of S = 500 draws) for the Laplace-shaped proposal versus prior sampling, as the sample size n of the simulated data grows. Both target the same posterior for (mu, w)."
set.seed(5)
ns <- c(5, 20, 100, 400)
ess_tbl <- do.call(rbind, lapply(ns, function(n) {
    x <- rnorm(n)
    data.frame(n = n,
               ess_laplace_proposal = attr(log_marlik_imps(x, 0, 5, 0, 5, 500), "ess"),
               ess_prior_sampling   = attr(log_marlik_prior(x, 0, 5, 0, 5, 500), "ess"))
}))
knitr::kable(ess_tbl, digits = c(0, 1, 1))
```

As $n$ grows the posterior sharpens around the truth while the prior does
not, so prior sampling's effective sample size deteriorates steadily; the
Laplace-shaped proposal, which re-centers and re-scales itself around the
posterior mode for every data set, keeps pace. This is the marginal-likelihood
instance of the same lesson as the missed-mode example above: a fixed,
data-independent proposal becomes an increasingly bad match to a target that
concentrates as more data arrive, and only a proposal that adapts to the
target avoids the resulting collapse in efficiency.

### Seeing it: sorted weights {#sec-marlik-app}

A single number like $S_{\text{eff}}$ summarizes the weight distribution;
looking at the weights themselves is more direct. The app below draws $S$
importance samples from each proposal and plots their *normalized weights,
sorted from largest to smallest* against rank$/S$, on a log scale. A proposal
shaped like the target gives a nearly flat curve close to $1/S$ (every draw
contributes about equally); a poorly matched proposal gives a curve that
falls off a cliff — a handful of draws near rank $0$ carry nearly all the
weight, and everything past them is numerically negligible. The legend
reports the grid-quadrature truth and, for each proposal, its estimate and
95% CI, colored green when the interval covers the truth and red when it does
not. This app runs live only in the HTML edition of this book; readers of
other formats can follow the link below it.

::: {#fig-marlik-app .content-visible when-format="html"}

```{shinylive-r}
#| label: marlik-weights-app
#| standalone: true
#| viewerHeight: 650

library(shiny)

log_sum_exp <- function(lx) { m <- max(lx); m + log(sum(exp(lx - m))) }

log_lik   <- function(x, mu, w) sum(dnorm(x, mu, exp(w / 2), log = TRUE))
log_prior <- function(mu, w, mu_0, sigma_mu, w_0, sigma_w) {
    dnorm(mu, mu_0, sigma_mu, log = TRUE) + dnorm(w, w_0, sigma_w, log = TRUE)
}
neg_log_post <- function(theta, x, mu_0, sigma_mu, w_0, sigma_w) {
    -log_lik(x, theta[1], theta[2]) -
        log_prior(theta[1], theta[2], mu_0, sigma_mu, w_0, sigma_w)
}
log_dmvnorm_batch <- function(Theta, mu, A) {
    Tc <- Theta - mu
    quad <- colSums((A %*% Tc) * Tc)
    0.5 * (-nrow(Theta) * log(2 * pi) + sum(log(svd(A)$d)) - quad)
}

## same grid-quadrature "truth" as log_marlik_grid() above
marlik_truth <- function(x, mu_0, sigma_mu, w_0, sigma_w, n_grid = 400, width = 8) {
    n   <- length(x)
    fit <- nlm(neg_log_post, c(mean(x), log(stats::var(x))), hessian = TRUE,
               x = x, mu_0 = mu_0, sigma_mu = sigma_mu, w_0 = w_0, sigma_w = sigma_w)
    se  <- sqrt(diag(solve(fit$hessian)))
    mu_grid <- fit$estimate[1] + seq(-width, width, length.out = n_grid) * se[1]
    w_grid  <- fit$estimate[2] + seq(-width, width, length.out = n_grid) * se[2]
    SS <- sum(x^2) - 2 * mu_grid * sum(x) + n * mu_grid^2
    log_lik_grid   <- outer(SS, w_grid,
                             function(ss, w) -0.5 * n * log(2 * pi) - n * w / 2 - 0.5 * exp(-w) * ss)
    log_prior_grid <- outer(dnorm(mu_grid, mu_0, sigma_mu, log = TRUE),
                             dnorm(w_grid, w_0, sigma_w, log = TRUE), "+")
    h_mu <- diff(mu_grid[1:2]); h_w <- diff(w_grid[1:2])
    log_sum_exp(as.vector(log_lik_grid + log_prior_grid)) + log(h_mu) + log(h_w)
}

## Laplace-shaped proposal: draw, weight, and return the log estimate, its
## standard error (delta method), and the sorted normalized weights
sim_laplace <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    fit <- nlm(neg_log_post, c(mean(x), log(stats::var(x))), hessian = TRUE,
               x = x, mu_0 = mu_0, sigma_mu = sigma_mu, w_0 = w_0, sigma_w = sigma_w)
    mu_hat <- fit$estimate; H <- fit$hessian
    L <- t(chol(solve(H)))
    Theta <- L %*% matrix(rnorm(2 * S), 2, S) + mu_hat
    log_g <- log_dmvnorm_batch(Theta, mu_hat, H)
    log_f <- -apply(Theta, 2, neg_log_post, x = x, mu_0 = mu_0, sigma_mu = sigma_mu,
                     w_0 = w_0, sigma_w = sigma_w)
    lw <- log_f - log_g
    m  <- max(lw); v <- exp(lw - m); wn <- v / sum(v)
    list(est = log_sum_exp(lw) - log(S), se = sd(v) / (sqrt(S) * mean(v)),
         ess = 1 / sum(wn^2), wn = sort(wn, decreasing = TRUE))
}

## prior proposal (g = pi): same bookkeeping, unfavourable g
sim_prior <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    mus <- rnorm(S, mu_0, sigma_mu); ws <- rnorm(S, w_0, sigma_w)
    lw  <- vapply(seq_len(S), function(i) log_lik(x, mus[i], ws[i]), numeric(1))
    m   <- max(lw); v <- exp(lw - m); wn <- v / sum(v)
    list(est = log_sum_exp(lw) - log(S), se = sd(v) / (sqrt(S) * mean(v)),
         ess = 1 / sum(wn^2), wn = sort(wn, decreasing = TRUE))
}

ui <- fluidPage(
    titlePanel("Sorted importance weights: Laplace-shaped proposal vs. prior"),
    sidebarLayout(
        sidebarPanel(
            width = 3,
            sliderInput("n", "data size  n", min = 5, max = 200, value = 20, step = 5),
            sliderInput("sig_mu", HTML("prior sd on &mu;"), min = 1, max = 10, value = 5, step = 0.5),
            sliderInput("sig_w", HTML("prior sd on w"), min = 1, max = 10, value = 5, step = 0.5),
            selectInput("S", "importance sample size  S",
                        choices = c(500, 2000, 5000, 20000), selected = 2000),
            actionButton("new", "New sample")
        ),
        mainPanel(width = 9, plotOutput("panel", height = "560px"))
    )
)

server <- function(input, output) {

    sim <- reactive({
        n <- input$n; S <- as.integer(input$S)
        sig_mu <- input$sig_mu; sig_w <- input$sig_w
        set.seed(input$new + 1)
        x <- rnorm(n)

        list(S = S,
             truth = marlik_truth(x, 0, sig_mu, 0, sig_w),
             lap   = sim_laplace(x, 0, sig_mu, 0, sig_w, S),
             pri   = sim_prior(x, 0, sig_mu, 0, sig_w, S))
    })

    output$panel <- renderPlot({
        s <- sim(); S <- s$S
        rank_frac <- seq_len(S) / S

        floor  <- max(min(c(s$lap$wn[s$lap$wn > 0], s$pri$wn[s$pri$wn > 0])), 1e-300)
        wn_lap <- pmax(s$lap$wn, floor)
        wn_pri <- pmax(s$pri$wn, floor)

        par(mar = c(4.5, 5, 2.5, 1))
        plot(rank_frac, wn_lap, type = "n", log = "y", ylim = c(floor, 1),
             xlab = "rank / S  (draws sorted by weight, largest first)",
             ylab = "normalized weight (log scale)",
             main = "sorted importance weights")
        lines(rank_frac, wn_lap, lwd = 3, col = "steelblue")
        lines(rank_frac, wn_pri, lwd = 3, col = "firebrick")
        abline(h = 1 / S, lty = 2, col = "darkgreen", lwd = 2)

        lwr_lap <- s$lap$est - 1.96 * s$lap$se; upr_lap <- s$lap$est + 1.96 * s$lap$se
        lwr_pri <- s$pri$est - 1.96 * s$pri$se; upr_pri <- s$pri$est + 1.96 * s$pri$se
        cov_lap <- if (lwr_lap <= s$truth && s$truth <= upr_lap) "darkgreen" else "firebrick"
        cov_pri <- if (lwr_pri <= s$truth && s$truth <= upr_pri) "darkgreen" else "firebrick"

        legend("topright", inset = c(0.02, 0.05), bty = "n", cex = 1.0,
               lwd = c(NA, 3, 3, NA, NA),
               col  = c(NA, "steelblue", "firebrick", NA, NA),
               text.col = c("black", "black", "black", cov_lap, cov_pri),
               legend = c(
                   sprintf("truth (grid quadrature): log P(y) = %.3f", s$truth),
                   "Laplace proposal",
                   "prior proposal",
                   sprintf("Laplace: est=%.3f  CI=(%.3f,%.3f)  ESS=%.1f%%",
                           s$lap$est, lwr_lap, upr_lap, 100 * s$lap$ess / S),
                   sprintf("prior:   est=%.3f  CI=(%.3f,%.3f)  ESS=%.2f%%",
                           s$pri$est, lwr_pri, upr_pri, 100 * s$pri$ess / S)))
    })
}

shinyApp(ui, server)
```

**Shinylive App for Sorted Importance Weights.** Sliders control the
simulated data size $n$ and the prior spreads on $\mu$ and $w$; the plot
compares the sorted, normalized importance weights from the Laplace-shaped
proposal against those from the prior, with the dashed line at $1/S$ marking
perfectly balanced weights.

:::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit12_importance_sampling.html#sec-marlik-app>.
:::

Raise $n$ with the sliders at their defaults and watch the prior's curve
rotate from a gentle slope into a cliff at the very first few ranks, while the
Laplace curve barely moves — the same collapse as
@tbl-marlik-ess-vs-n, now visible in the shape of the curve rather than
only in a single summary number.

## Summary

- Importance sampling reweights draws from a proposal $g$ by
  $w(\theta) = f(\theta)/g(\theta)$ instead of accepting or rejecting them,
  using $\widehat{C_f/C_g} = \frac1S\sum_s w_s$ and the self-normalized
  $\hat A = \sum_s a(\theta_s)\tilde w_s$; only the support condition
  $\{g>0\}\supseteq\{f>0\}$ is required, not an envelope $g \ge f$.
- A good proposal shaped like the target on its support gives near-constant
  weights and $S_{\text{eff}} \approx S$ (the truncated-normal and
  Laplace-proposal examples), and a single weighted sample can answer several
  questions about the same target — a probability and an expectation alike.
- A proposal that misses a mode or part of the support fails **silently**: the
  observed weights can look perfectly healthy while both $C_f/C_g$ and
  $E_f[a]$ are badly wrong, and no amount of extra sampling fixes it.
  $S_{\text{eff}}$ diagnoses only the noisy failure (wildly varying observed
  weights), never the silent one.
- For the marginal likelihood, a proposal built from a Laplace approximation
  at the posterior mode remains efficient as the data accumulate and the
  posterior sharpens, while a fixed, data-independent proposal such as the
  prior does not — reinforcing why the mode-finding methods from earlier
  chapters remain useful even once we turn to sampling-based methods.
