---
title: Random Number Generator and Monte Carlo Elements
code-fold: true
---

## Pseudo random numbers

Computers generate "random" numbers using deterministic algorithms. One of the most common methods is the **Linear Congruential Generator (LCG)**. It produces a sequence of pseudo-random integers $X_n$ defined by the recurrence relation:

$$
X_n = (a X_{n-1} + c) \pmod M
$$

where $X_0$ is the seed, $a$ is the multiplier, $c$ is the increment, and $M$ is the modulus. To get numbers in the interval $[0,1]$, we divide by $M-1$. In this example, we use the parameters $a = 7^5$, $c = 0$, and $M = 2^{31}-1$ (a minimal standard LCG).

```{r echo=TRUE}
A <- 7^5
M <-  2^31-1

N <- 500
rn <- rep (0, N)
rn[1] <- 10 # Initial seed
for (i in 2:length (rn))
{
    rn[i] <- (A * rn[i-1] ) %% M
}

## Scale the numbers to fall between 0 and 1
nrn <- rn/(M-1)

par(mfrow=c(1,3),mar=c(4,4,3,1))
plot(nrn[1:100], main="Trace Plot", ylab="Value", xlab="Index")
acf(nrn, main="Autocorrelation")
hist(nrn, main="Histogram")

## Compare with R's built-in runif()
n <- 500
a <- runif(n)
par(mfrow=c(1,2),mar=c(4,4,3,1))
hist(a,xlab="Random Numbers",main="Built-in runif()")
acf(a,main="Built-in acf")
```

## Inverting CDF 

If we can generate uniform random variables $U \sim \text{Unif}(0,1)$, we can generate a random variable $X$ from any continuous distribution using the **Inverse Transform Method**. If $F(x)$ is the cumulative distribution function (CDF) of $X$, then $X = F^{-1}(U)$ has the desired distribution.

For an Exponential distribution with rate $\lambda = 1$, the CDF is $F(x) = 1 - e^{-x}$. 
Setting $U = 1 - e^{-x}$ and solving for $x$ gives:

$$
x = -\ln(1-U)
$$

```{r}
## generate exponential random numbers
## use method of inverse cdf to generate iid sample from exp(1)
gen_exp <- function(n)
{
    #generate unif(0,1) random numbers
    u <- runif(n)
    #transform the random numbers
    -log(1-u)
}

b <- gen_exp(10000)
par(mfrow=c(1,3),mar=c(4,4,3,1))
hist(b, main="Custom gen_exp")
acf(b, main="Custom acf")

### r built-in generator for comparison
hist(rexp (10000), main="Built-in rexp")
```

## A Special Transformation for Generating Normal Sample

The **Box-Muller Transform** is an elegant method for generating pairs of independent Standard Normal random variables from independent Uniform random variables. It relies on converting cartesian coordinates $(X,Y)$ into polar coordinates $(R, \theta)$. 

If $X, Y \sim N(0,1)$ are independent, their squared radius $R^2 = X^2 + Y^2$ follows an Exponential distribution with rate $\lambda = 1/2$, and their angle $\theta$ is uniformly distributed over $[0, 2\pi]$. By generating $R$ and $\theta$, we can map them back to Normal variables:

$$
X = R\cos(\theta) \quad \text{and} \quad Y = R\sin(\theta)
$$

```{r echo=TRUE}
gen_normal <- function(n)
{
    # calculates size of random samples (we generate pairs, so we need half)
    size_sample <- ceiling(n/2)
    
    # R^2 ~ Exp(1/2), so we take the square root to get R
    R <- sqrt(2*rexp(size_sample))
    theta <- runif(size_sample,0,2*pi)
    
    X <- R*cos(theta)  
    Y <- R*sin(theta)
    
    c(X,Y)[1:n]  
}

normal_sample <- gen_normal(1000)

par(mfrow=c(1,2),mar=c(4,4,1,1))
hist(normal_sample,main="Custom Normal")
qqnorm(normal_sample, main="Custom QQ")
qqline(normal_sample)

## R's built-in generator for comparison
par(mfrow=c(1,2),mar=c(4,4,1,1))
nsample2 <- rnorm (1000) 
hist(nsample2, main="Built-in Normal")
qqnorm(nsample2, main="Built-in QQ")
qqline(nsample2)
```

## Demonstration of CLT and LLN

The **Law of Large Numbers (LLN)** states that as the sample size $n$ grows, the sample mean $\bar{X}_n$ converges to the expected value $\mu$:
$$
\bar{X}_n \to \mu \quad \text{as} \quad n \to \infty
$$

The **Central Limit Theorem (CLT)** states that the standardized sample mean converges in distribution to a Standard Normal $N(0,1)$:
$$
Z = \frac{\bar{X}_n - \mu}{\sigma/\sqrt{n}} \xrightarrow{d} N(0,1) \quad \text{as} \quad n \to \infty
$$

Here we draw from a Gamma distribution with shape parameter $\alpha = 2$ and rate $\beta = 1$. The theoretical mean is $\mu = \alpha/\beta = 2$, and the variance is $\sigma^2 = \alpha/\beta^2 = 2$.

```{r echo=TRUE}
n <- 100
rn <- rgamma(n, shape = 2)

par(mfrow = c(2,2))
plot(rn[1:100], main="Raw Gamma Draws", ylab="Value")
hist(rn, main="Histogram of Draws")

## Calculate running averages (LLN)
xbar.rep <- replicate(5, cumsum(rgamma(n, shape = 2))/(1:n))
## Calculate running standardized means (CLT)
se.rep <- replicate(5, (cumsum(rgamma(n, shape = 2))/(1:n) - 2) / sqrt(2/(1:n)))

## Plot LLN: should converge to mu = 2
plot(xbar.rep[,1], main = "Demonstrating LLN", log = "x", type = "b", ylim = c(0, 6), ylab="Sample Mean")
abline(h = 2, lwd=2)
for (i in 2:ncol(xbar.rep)) points(xbar.rep[,i], col = i, pch = i, type = "b")

## Plot CLT: should stabilize within roughly [-3, 3] standard deviations
plot(se.rep[,1], ylim = c(-3,3), type="b", log = "x", main="Demonstrating CLT", ylab="Z-Score")
for (i in 2:ncol(se.rep)) points(se.rep[,i], col = i, pch = i, type = "b")
abline(h = c(0, -2, 2), lty = c(1,2,2), lwd=2)
```

## An Example of Monte Carlo for Estimating $\pi$

We can estimate the value of $\pi$ by throwing random darts at a square with side length 2 (from $-1$ to $1$) and counting how many land inside the inscribed circle of radius $1$.

The area of the square is $(2)^2 = 4$, and the area of the circle is $\pi(1)^2 = \pi$. Therefore, the probability that a uniform random point $(X,Y)$ falls inside the circle is:
$$
P(X^2 + Y^2 \le 1) = \frac{\text{Area of Circle}}{\text{Area of Square}} = \frac{\pi}{4}
$$

#### Visualizing the Method

Before we estimate $\pi$, let's visually simulate throwing 1,000 darts:

```{r echo=TRUE, fig.height=5, fig.width=5}
set.seed(123)
n_visual <- 1000
x_pts <- runif(n_visual, -1, 1)
y_pts <- runif(n_visual, -1, 1)
inside_circle <- (x_pts^2 + y_pts^2) <= 1

## Plot points
plot(x_pts, y_pts, col = ifelse(inside_circle, "steelblue", "tomato"), 
     pch = 20, asp = 1, xlab = "X", ylab = "Y", 
     main = paste(sum(inside_circle), "out of", n_visual, "points in circle"))

## Draw the boundaries
rect(-1, -1, 1, 1, border = "black", lwd = 2)
theta_seq <- seq(0, 2*pi, length.out = 100)
lines(cos(theta_seq), sin(theta_seq), col = "black", lwd = 2)
```

#### The Monte Carlo Estimate

If we define an indicator variable $I_i$ that equals 1 if the $i$-th dart lands in the circle and 0 otherwise, then $E[I_i] = \pi/4$. 
We define $Z_i = 4 \times I_i$, which gives $E[Z_i] = \pi$. 

Our Monte Carlo estimate $\hat{\pi}$ is the sample average of the $Z_i$ values:
$$
\hat{\pi} = \bar{Z} = \frac{1}{n}\sum_{i=1}^n Z_i
$$

The Standard Error (SE) measures the uncertainty of our estimate. Using the sample standard deviation $S_Z$, the 95% margin of error is calculated as:
$$
\text{Margin of Error} = 1.96 \times \text{SE} = 1.96 \times \frac{S_Z}{\sqrt{n}}
$$

#### Results by Sample Size

We encapsulate the logic in a function and track how precision improves as $n$ increases. The table below outlines our findings.

```{r echo=TRUE, warning=FALSE}
library(knitr)

## n is the number of samples drawn uniformly from the rectangle (-1,1)  * (-1,1)
## an estimate of pi is returned
pi_est_mc <- function(n)
{
    X <- runif(n,-1,1)
    Y <- runif(n,-1,1)
    
    # 4 times the indicator function
    Z <- 4 * (X^2 + Y^2 <= 1)
    
    mu <- mean(Z)
    error <- 1.96 * sd(Z) / sqrt(n)
    
    list(n = n, pi.est = mu, error.95perc = error, ci.lower = mu - error, ci.upper = mu + error)
}  

## Run the estimation for varying sample sizes
sample_sizes <- c(100, 10000, 100000, 10000000)
results_list <- lapply(sample_sizes, pi_est_mc)

## Combine results into a dataframe for tabular display
results_df <- data.frame(
  N = sapply(results_list, function(x) x$n),
  Estimate = sapply(results_list, function(x) x$pi.est),
  Error_Margin = sapply(results_list, function(x) x$error.95perc),
  CI_Lower = sapply(results_list, function(x) x$ci.lower),
  CI_Upper = sapply(results_list, function(x) x$ci.upper)
)

## Render the formatted table
kable(results_df, 
      col.names = c("Sample Size (N)", "Estimated Pi", "95% Margin of Error", "Lower CI", "Upper CI"), 
      digits = 6, 
      caption = "Monte Carlo Estimation of Pi across different sample sizes")
```