---
title: "Optimization for Maximum Likelihood Estimation"
format:
  html:
    toc: true
    tbl-colwidths: true
editor_options:
  chunk_output_type: console
code-fold: true
---

This chapter develops maximum likelihood estimation and the numerical methods
used to find it, first for a scalar parameter and then for a parameter vector.
The univariate case, where the negative log-likelihood can be plotted and a root
of its derivative can be bracketed, motivates the two ideas that carry over
unchanged to higher dimensions: replacing that function locally by a quadratic
and solving for its minimum (Newton–Raphson), and using the curvature of that
quadratic to report a standard error. The multivariate case, where neither plotting nor
bracketing is available, shows what survives of that idea and what has to be
added — safeguards against divergence, and a wider menu of derivative-free and
gradient-only methods (Nelder–Mead, gradient descent, conjugate gradient,
BFGS, and stochastic gradient descent) for when the Hessian is unavailable or
too expensive. Several of these methods are illustrated below with live,
in-browser animations that let you change the objective function, the
starting point, and the algorithm before watching the iterates respond.


```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false
library(latex2exp)   # TeX() renders LaTeX in plot titles, axis labels and legends
set.seed(812)
```

## Maximum Likelihood: Theory

### Likelihood, Negative Log-Likelihood, and Information

A model specifies the density or probability of the data $D$ given a parameter $\theta$, written $f(D \mid \theta)$. As a function of $D$ with $\theta$ fixed it is the sampling distribution; as a function of $\theta$ with $D$ observed it is the likelihood. For independent observations $x_1, \ldots, x_n$,
$$
L(\theta) = \prod_{i=1}^{n} f(x_i \mid \theta), \qquad \ell(\theta) = \log L(\theta) = \sum_{i=1}^{n} \log f(x_i \mid \theta).
$$

We never work with $L$ itself: the product underflows to zero for even moderate $n$, and sums are easier to differentiate. Maximizing $L$ is the same as maximizing $\ell$, which in turn is the same as **minimizing** the negative log-likelihood
$$
\mathcal{L}(\theta) = -\ell(\theta) = -\sum_{i=1}^{n}\log f(x_i \mid \theta) = \sum_{i=1}^{n} \mathcal{L}_i(\theta) ,
$$
and it is the minimization form that the rest of this chapter — and every optimizer in R — actually uses. Where $\mathcal{L}$ is smooth with an interior minimum, $\hat\theta$ solves,
$$
g(\theta) = \mathcal{L}'(\theta) = 0, \qquad g'(\hat\theta) > 0 .
$$

Three quantities recur throughout the chapter:

- the **gradient** $g(\theta) = \mathcal{L}'(\theta) = -\ell'(\theta)$, whose root locates the minimum;

- the **observed information** $J(\theta) = g'(\theta) = \mathcal{L}''(\theta) = -\ell''(\theta)$, the curvature of $\mathcal{L}$ at $\theta$;

- the **expected information** $n I(\theta) = E_D\left[g'(\theta; D)\right]$, its expectation under the model.

The derivative of the *log*-likelihood, $\ell'(\theta) = -g(\theta)$, is what the statistical literature calls the **score**, and $g(\theta) = 0$ is the **score equation** written with the opposite sign. That terminology survives in two standard names used later — the *method of scoring*, and Fisher information defined as the variance of the score — but the algorithms below are all stated in terms of $\mathcal{L}$ and $g$, so the score does not appear as a separate object.

The second derivative plays a double role. It determines how a Newton-type algorithm steps toward the minimum, and it determines how precisely the data pin the parameter down. An optimizer that computes the curvature has computed the standard error at the same time.

### Example: iid Bernoulli

Before turning to a model with a continuous parameter and an exactly quadratic
negative log-likelihood, it is worth solving $g(\theta) = 0$ by hand in the
simplest possible case. Let $Y_1, \ldots, Y_n \overset{iid}{\sim} \mathrm{Bernoulli}(\theta)$,
so that $P(y_i \mid \theta) = \theta^{y_i}(1 - \theta)^{1 - y_i}$. Writing
$n_1 = \sum_i y_i$ for the number of successes and $n_0 = n - n_1$ for the
number of failures,
$$
L(\theta) = \theta^{n_1}(1 - \theta)^{n_0}, \qquad
\mathcal{L}(\theta) = -n_1 \log\theta - n_0 \log(1 - \theta).
$$

Its derivative is,
$$
g(\theta) = -\frac{n_1}{\theta} + \frac{n_0}{1 - \theta},
$$
and setting $g(\theta) = 0$ gives $n_1(1-\theta) = n_0\theta$, i.e.
$n_1 = (n_1 + n_0)\theta = n\theta$, so
$$
\hat\theta = \frac{n_1}{n} = \overline y.
$$

The second derivative $g'(\theta) = n_1/\theta^{2} + n_0/(1-\theta)^{2}$ is
positive wherever it is defined, confirming that $\overline y$ is a minimum of
$\mathcal{L}$ and not a saddle point. With only three observations, $y = (1, 1, 0)$,
both the likelihood and $\mathcal{L}$ can be tabulated and plotted directly, which
makes $\hat\theta = 2/3 = \overline y$ visible without any calculus at all — the
value that makes the data most probable is exactly the value that makes
$\mathcal{L}$ smallest.

| $\theta$ | $L(\theta) = \theta^{2}(1-\theta)$ | $\mathcal{L}(\theta) = -\log L(\theta)$ |
|:---:|:---|:---|
| $0$ | $0$ | $\infty$ |
| $1/2$ | $1/8 = 0.1250$ | $2.079$ |
| $2/3$ | $4/27 = 0.1481$ **(max)** | $1.910$ **(min)** |
| $3/4$ | $9/64 = 0.1406$ | $1.962$ |
| $1$ | $0$ | $\infty$ |

```{r}
#| label: fig-bernoulli-toy-likelihood
#| fig-width: 9
#| fig-height: 4
#| fig-cap: "Introduces the likelihood and the negative log-likelihood with the simplest possible example, showing that the parameter value making the data most probable is the one that minimizes the function every optimizer in this chapter works with.
#|   Likelihood and negative log-likelihood of $\\theta$ for three Bernoulli observations $y = (1,1,0)$: the maximum of $L$ and the minimum of $\\mathcal{L}$ both sit at $\\hat\\theta = \\bar y = 2/3$."
par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
curve(x^2 * (1 - x), 0, 1, lwd = 2, xlab = TeX(r"($\theta$)"),
      ylab = TeX(r"($L(\theta)$)"), main = TeX(r"(likelihood)"))
abline(v = 2 / 3, col = 2, lty = 2)

curve(-log(x^2 * (1 - x)), 0.05, 0.98, n = 400, lwd = 2, ylim = c(1.8, 4),
      xlab = TeX(r"($\theta$)"), ylab = TeX(r"($-\log L(\theta)$)"),
      main = TeX(r"(negative log-likelihood)"))
abline(v = 2 / 3, col = 2, lty = 2)
```

### Closed Form: the Normal Mean

Let $X_1, \ldots, X_n$ be iid $N(\mu, \sigma_0^{2})$ with $\sigma_0^{2}$ known. Using the identity $\sum_i (x_i - \mu)^2 = \sum_i (x_i - \overline x)^2 + n(\overline x - \mu)^2$,
$$
\mathcal{L}(\mu) = \frac{n}{2}\log\left(2\pi\sigma_0^{2}\right) + \frac{1}{2\sigma_0^{2}}\sum_{i=1}^{n}(x_i - \mu)^{2} = C + \frac{n}{2\sigma_0^{2}}(\mu - \overline x)^{2} .
$$

Differentiating,
$$
g(\mu) = \frac{n}{\sigma_0^{2}}\left(\mu - \overline x\right), \qquad g'(\mu) = \frac{n}{\sigma_0^{2}}, \qquad \hat\mu = \overline x .
$$

Every feature of the general theory is visible here in exact form. $\mathcal{L}$ is an upward parabola, so it has a unique minimum. Its derivative $g$ is linear, so $g(\mu) = 0$ is solved in one line. The curvature $J(\mu) = g'(\mu) = n/\sigma_0^{2}$ does not depend on $\mu$ or on the data, so observed and expected information coincide, $J = nI$, and both grow linearly with $n$: doubling the sample doubles the sharpness of the well and halves the variance of $\hat\mu$.

```{r}
#| label: fig-normal-nll-gradient-curvature
#| fig-width: 10
#| fig-height: 4
#| fig-cap: "Relates the three central objects of maximum likelihood, the negative log-likelihood, its gradient, and its curvature, and shows how each changes with sample size.
#|   Negative log-likelihood, its gradient $g$, and its curvature for a normal mean at $n = 20$ and $n = 50$: the larger sample gives a sharper well, a steeper $g$, and larger curvature."
x20 <- rnorm(20, mean = 1); x50 <- rnorm(50, mean = 1)
mu <- seq(0, 2, length.out = 200)

nll_norm <- function(mu, x) vapply(mu, function(m) -sum(dnorm(x, m, log = TRUE)), numeric(1))

par(mfrow = c(1, 3), mar = c(4, 4.5, 2.5, 1))

matplot(mu, cbind(nll_norm(mu, x20) - min(nll_norm(mu, x20)),
                  nll_norm(mu, x50) - min(nll_norm(mu, x50))),
        type = "l", lty = 1, lwd = 2, col = c(1, 2),
        xlab = TeX(r"($\mu$)"), ylab = TeX(r"($L(\mu) - L(\hat{\mu})$)"),
        main = TeX(r"(negative log-likelihood)"))
legend("top", bty = "n", lty = 1, lwd = 2, col = c(1, 2),
       legend = TeX(c(r"($n = 20$)", r"($n = 50$)")))

matplot(mu, cbind(20 * (mu - mean(x20)), 50 * (mu - mean(x50))),
        type = "l", lty = 1, lwd = 2, col = c(1, 2),
        xlab = TeX(r"($\mu$)"), ylab = TeX(r"($g(\mu)$)"),
        main = TeX(r"(gradient)"))
abline(h = 0, lty = 3)

matplot(mu, cbind(rep(20, 200), rep(50, 200)), type = "l", lty = 1, lwd = 2, col = c(1, 2),
        ylim = c(0, 60), xlab = TeX(r"($\mu$)"), ylab = TeX(r"($g'(\mu) = J(\mu)$)"),
        main = TeX(r"(curvature)"))
```

The derivative $g$ crosses zero exactly where $\mathcal{L}$ bottoms out, and the larger sample gives a sharper well, a steeper $g$, and a larger curvature.

### Consistency, Asymptotic Normality, and Fisher Information

Under regularity conditions — the support of $f$ not depending on $\theta$, the true value interior to the parameter space, and two derivatives passing under the integral sign — the MLE has three properties as $n \to \infty$, with $\theta_*$ the true parameter.

**Consistency.** $\hat\theta \to \theta_*$ in probability. The reason is that $\mathcal{L}(\theta)/n$ converges to $-E_{\theta_*}[\log f(X \mid \theta)]$, which by the information inequality is minimized at $\theta = \theta_*$: the surface being minimized converges to one whose lowest point is at the truth.

#### Asymptotic normality. {-}
$$
\hat\theta \;\dot\sim\; N\left(\theta_*,\ \left[n I(\theta_*)\right]^{-1}\right), \qquad I(\theta) = -E_X\left[\frac{\partial^{2}}{\partial\theta^{2}}\log f(X \mid \theta)\right] = E_X\left[\left(\frac{\partial}{\partial\theta}\log f(X\mid\theta)\right)^{2}\right].
$$

A sketch: expand $g(\hat\theta) = 0$ about $\theta_*$ to get $\hat\theta - \theta_* \approx -g(\theta_*)/g'(\theta_*)$. The numerator is a sum of iid mean-zero terms — this is the step in which $I(\theta)$ appears as the variance of the score — hence asymptotically normal with variance $nI(\theta_*)$; the denominator converges to $nI(\theta_*)$ by the law of large numbers. Dividing gives variance $[nI(\theta_*)]^{-1}$.

**Asymptotic efficiency.** No consistent estimator has smaller asymptotic variance — the Cramér–Rao bound is attained.

Since $\theta_*$ is unknown, the variance is estimated by substituting $\hat\theta$. Two versions are available: the expected information $nI(\hat\theta)$, which requires the expectation analytically, and the observed information $J(\hat\theta) = g'(\hat\theta)$, which requires only the curvature that a Newton step already computes. By the law of large numbers $J(\hat\theta)/n \approx I(\hat\theta)$, and both give
$$
\widehat{\mathrm{SE}}(\hat\theta) = J(\hat\theta)^{-1/2} \quad\text{or}\quad \left[nI(\hat\theta)\right]^{-1/2}, \qquad \hat\theta \pm 1.96\,\widehat{\mathrm{SE}}(\hat\theta).
$$

These are asymptotic statements. A flat $\mathcal{L}$ — small curvature — gives a large standard error, and an $\mathcal{L}$ that is far from quadratic gives a standard error that should not be trusted at all; the last section of this chapter shows a case of exactly that.

### Demonstration: the Cauchy Location Model

The Cauchy density $f(x \mid \theta) = 1/\{\pi[1 + (x - \theta)^2]\}$ has no closed-form MLE:
$$
\mathcal{L}(\theta) = n\log\pi + \sum_{i=1}^{n}\log\left(1 + (x_i - \theta)^{2}\right), \qquad
g(\theta) = -\sum_{i=1}^{n}\frac{2(x_i - \theta)}{1 + (x_i - \theta)^{2}} .
$$

Setting $g(\theta) = 0$ gives a polynomial equation of degree $2n - 1$. Plotting $\mathcal{L}$ for samples of increasing size shows the well sharpening around the truth, which is consistency and growing information in one picture.

```{r}
#| label: fig-cauchy-nll-by-n
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Shows how the Cauchy location negative log-likelihood, which is harder to minimize than the normal one, concentrates around the true value as the sample size grows.
#|   Cauchy negative log-likelihood for samples of increasing size ($n = 50, 100, 200$) from a true location $\\theta_* = 20$: the well sharpens around the truth as $n$ grows."
nll_cauchy_fn <- function(theta, x) -sum(dcauchy(x, theta, log = TRUE))

thetas <- seq(15, 25, length.out = 200)
n_set <- c(50, 100, 200)

nll_cauchy <- sapply(n_set, function(n) {
  d <- rcauchy(n, location = 20)
  sapply(thetas, nll_cauchy_fn, x = d)
})
nll_cauchy <- sweep(nll_cauchy, 2, apply(nll_cauchy, 2, min))   # align minima at 0

par(mar = c(4, 4.5, 2.5, 1))
matplot(thetas, nll_cauchy, type = "l", lty = 1, lwd = 2, col = c(1, 2, 4), ylim = c(0, 20),
        xlab = TeX(r"($\theta$)"), ylab = TeX(r"($L(\theta) - L(\hat{\theta})$)"),
        main = TeX(r"(Cauchy negative log-likelihood, $\theta_* = 20$)"))
abline(v = 20, lty = 3)
legend("topright", bty = "n", lty = 1, lwd = 2, col = c(1, 2, 4),
       legend = TeX(sprintf(r"($n = %d$)", n_set)))
```

The variability of the MLE across repeated samples is the sampling distribution the theory describes. Repeating the experiment 50 times at two sample sizes shows the curves clustering more tightly, and their minima concentrating, as $n$ grows.

```{r}
#| label: fig-cauchy-nll-variability
#| fig-width: 10
#| fig-height: 4.5
#| cache: true
#| fig-cap: "Shows the sampling variability of the negative log-likelihood itself: different data sets give different curves, and the minima of those curves, the MLEs, become less variable as $n$ grows.
#|   50 repeated Cauchy negative log-likelihood curves at $n = 20$ versus $n = 200$: the curves cluster more tightly, and their minima concentrate, as $n$ grows."
nll_curves <- function(n, nrep = 50)
  replicate(nrep, {
    d <- rcauchy(n, location = 20)
    lc <- sapply(thetas, nll_cauchy_fn, x = d)
    lc - min(lc)
  })

curves_small <- nll_curves(20)
curves_large <- nll_curves(200)

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
for (obj in list(list(m = curves_small, n = 20), list(m = curves_large, n = 200)))
  matplot(thetas, obj$m, type = "l", lty = 1, col = adjustcolor(1, 0.35), ylim = c(0, 10),
          xlab = TeX(r"($\theta$)"), ylab = TeX(r"($L(\theta) - L(\hat{\theta})$)"),
          main = TeX(sprintf(r"(sample size $n = %d$)", obj$n)))

### sampling distribution of the MLE at the two sample sizes
mle_at <- function(m) thetas[apply(m, 2, which.min)]
rbind(`n = 20`  = c(mean = mean(mle_at(curves_small)), sd = sd(mle_at(curves_small))),
      `n = 200` = c(mean = mean(mle_at(curves_large)), sd = sd(mle_at(curves_large))))
```

The standard deviation falls by roughly the factor $\sqrt{10}$ predicted by $[nI(\theta_*)]^{-1/2}$. Two features of these curves also warn us about computation: for small $n$ the negative log-likelihood has several local minima, and its shape away from the well is far from quadratic. Numerical optimization is required, and it must be used with care.

## Univariate Optimization Algorithms

Every algorithm in this section **minimizes** the negative log-likelihood
$$
\mathcal{L}(\theta) = -\ell(\theta),
$$
so that the quantity driving each step is its derivative,
$$
g(\theta) = \mathcal{L}'(\theta) = -\ell'(\theta),
$$
and the quantity measuring curvature is $g'(\theta) = \mathcal{L}''(\theta) = -\ell''(\theta)$.
For iid data $\mathcal{L}$ is a sum of per-observation terms, $\mathcal{L}(\theta) = \sum_i \mathcal{L}_i(\theta)$,
a decomposition that matters only in the section on stochastic gradients. Maximizing $\ell$
and minimizing $-\ell$ are the same problem, but stating it as a minimization
matches what `optim`, `nlm` and `optimize` expect, and it is the form the
multivariate and gradient-based methods later in the chapter use as well. Note
that $g'$ is exactly the observed information $J(\theta)$ of the previous
section, so the curvature an optimizer computes on its way to the minimum is
the same curvature that supplies the standard error.

### The Running Example: Zero-Truncated Poisson Counts

All the methods below are illustrated on one model, so that their pictures and
their iterate sequences can be compared directly.

Counts are often recorded only when positive: a study of hospital stays sees admitted patients, a survey of litters sees litters that exist. If $Y \sim \mathrm{Poisson}(\lambda)$ and we observe $X = Y \mid Y > 0$,
$$
P(x \mid \lambda) = \frac{e^{-\lambda}\lambda^{x}}{x!}\cdot\frac{1}{1 - e^{-\lambda}}, \qquad x = 1, 2, \ldots.
$$

Dropping the factorial terms, which do not involve $\lambda$, and dividing by $n$
gives the negative log-likelihood **per observation**,
$$
\mathcal{L}(\lambda) = -\overline x \log\lambda + \lambda + \log\left(1 - e^{-\lambda}\right),
$$
whose derivatives are,
$$
g(\lambda) = -\ell'(\lambda) = -\frac{\overline x}{\lambda} + \frac{1}{1 - e^{-\lambda}}, \qquad
g'(\lambda) = -\ell''(\lambda) = \frac{\overline x}{\lambda^{2}} - \frac{e^{-\lambda}}{\left(1 - e^{-\lambda}\right)^{2}} .
$$

Two features of this model make it a convenient running example. First, the data
enter only through the sample mean $\overline x$, so the total negative
log-likelihood is exactly $n$ times the per-observation one and the two have the
same minimizer; the figures below therefore plot the per-observation version,
which does not grow with $n$, while the standard error at the end uses the total.
Second, since $E(X) = \lambda/(1 - e^{-\lambda})$, substituting $E(X)$ for
$\overline x$ in $g'$ gives the expected information per observation in closed
form,
$$
I(\lambda) = \frac{1}{\lambda\left(1 - e^{-\lambda}\right)} - \frac{e^{-\lambda}}{\left(1 - e^{-\lambda}\right)^{2}} ,
$$
which is what Fisher scoring needs. The equation $g(\lambda) = 0$ has no
closed-form root, but $\mathcal{L}$ is unimodal — an ideal problem on which to compare
all the methods against a single known answer.

```{r}
#| label: nzp-functions
set.seed(2024)          # the same data underlies every figure in this section

## Negative log-likelihood per observation and its derivatives: nll is L(lambda),
## g is the gradient L'(lambda), dg is the curvature L''(lambda). The data enter
## only through xbar, so the total L is length(xp) times these.
nll <- function(l, xb = xbar) -xb * log(l) + l + log(1 - exp(-l))    # L(lambda)
g   <- function(l, xb = xbar) -xb / l + 1 / (1 - exp(-l))            # gradient  L'
dg  <- function(l, xb = xbar)  xb / l^2 - exp(-l) / (1 - exp(-l))^2  # curvature L''

## expected information per observation, and the fixed-point map used below
info <- function(l) 1 / (l * (1 - exp(-l))) - exp(-l) / (1 - exp(-l))^2
fp_map <- function(l, xb = xbar) xb * (1 - exp(-l))

## simulated data: Poisson counts with the zeros discarded
x  <- rpois(100, lambda = 1.5)
xp <- x[x > 0]
xbar <- mean(xp)

lhat <- uniroot(g, c(0.2, 20))$root       # the answer every method must find

c(n_observed = length(xp), sample_mean = xbar, true_lambda = 1.5, mle = lhat)
```

The sample mean overshoots 1.5, because the zeros that would have pulled it
down were never recorded; the MLE corrects for exactly that.

```{r}
#| label: fig-nzp-nll-gradient
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: "Sets up the running example of a zero-truncated Poisson model by showing the negative log-likelihood that every method in this section minimizes, together with its gradient, whose root is the minimizer.
#|   Negative log-likelihood per observation and its gradient $g(\\lambda)$ for the zero-truncated Poisson model: the minimum of $\\mathcal{L}$ sits exactly where $g$ crosses zero."
lam <- seq(0.3, 3, length.out = 300)

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
plot(lam, nll(lam), type = "l", lwd = 2,
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($L(\lambda)$)"),
     main = TeX(r"(negative log-likelihood)"))
abline(v = lhat, lty = 3)
points(lhat, nll(lhat), pch = 19, cex = 1.2)

plot(lam, g(lam), type = "l", lwd = 2,
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($g(\lambda)$)"),
     main = TeX(r"(gradient)"))
abline(h = 0, lty = 3); abline(v = lhat, lty = 3)
points(lhat, 0, pch = 19, cex = 1.2)
```

The minimum of $\mathcal{L}$ and the zero crossing of $g$ are the same point,
$\hat\lambda \approx `r round(lhat, 3)`$. Since $g$ is *increasing* through
that crossing — the mirror image of the decreasing score $\ell' = -g$ — the curvature
$g'(\hat\lambda)$ is positive, as it must be at a minimum.

### Three Equivalent Formulations

The same task can be posed three ways, and the algorithms below attack different versions of it:

- minimize $\mathcal{L}(\theta) = -\ell(\theta)$ directly, using function values only;

- find a root of $g(\theta) = -\ell'(\theta)$;

- rearrange that root equation into a fixed point $\theta = \varphi(\theta)$.

Root-finders are usually faster because they use derivative information, but they locate any stationary point, including maxima and saddle points. Direct minimizers need only function values and always move downhill, but converge more slowly. Which to prefer depends on whether derivatives are available and whether the function is smooth.

Each method below is illustrated with a schematic showing its first few iterations on the truncated-Poisson `nll` just defined. The minimizer $\hat\lambda$ is marked in every figure.

### Fixed-Point (Simple) Iteration

Rearrange the stationarity equation $g(\theta) = -\ell'(\theta) = 0$ algebraically into the form $\theta = \varphi(\theta)$ and iterate $\theta_{k+1} = \varphi(\theta_k)$. For the truncated Poisson, $-\overline x/\lambda + 1/(1 - e^{-\lambda}) = 0$ rearranges to
$$
\lambda = \overline x\left(1 - e^{-\lambda}\right) = \varphi(\lambda), \qquad \varphi'(\lambda) = \overline x e^{-\lambda}.
$$

If $\varphi$ maps an interval into itself and $\lvert\varphi'(\theta^{*})\rvert < 1$ at the fixed point, the mean value theorem gives
$$
\lvert\theta_{k+1} - \theta^{*}\rvert = \lvert\varphi(\theta_k) - \varphi(\theta^{*})\rvert \le \lvert\varphi'(\xi)\rvert \lvert\theta_k - \theta^{*}\rvert ,
$$
so the error contracts by roughly the constant factor $\lvert\varphi'(\theta^{*})\rvert$ at each step. This is **linear convergence**: a fixed number of digits is gained per iteration, and the closer $|g'|$ is to 1 the slower the progress.

```{r}
#| label: fixed-point
fixed_point <- function(g, x0, tol = 1e-10, maxit = 500) {
  for (k in 1:maxit) {
    x1 <- g(x0)
    if (abs(x1 - x0) < tol) return(c(root = x1, iter = k))
    x0 <- x1
  }
  stop("not converged")
}
```

The mechanism is easiest to see as a **cobweb**: from $\lambda_k$ on the diagonal, move vertically to the curve $\varphi$ to obtain $\lambda_{k+1}$, then horizontally back to the diagonal to start the next step. The iterates walk toward the intersection of $g$ with the $45^{\circ}$ line, and the shallower $\varphi$ is at that crossing — the smaller $\lvert\varphi'\rvert$ — the larger each stride toward it.

```{r}
#| label: fig-scheme-fixed-point
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Illustrates fixed-point iteration, the simplest scheme for solving the stationarity equation, and how its iterates converge to the solution.
#|   Fixed-point iteration as a cobweb diagram: iterates walk toward the intersection of $\\varphi$ with the diagonal."
par(mar = c(4, 4.5, 2.5, 1))
plot(lam, fp_map(lam), type = "l", lwd = 2, xlim = c(0.3, 3), ylim = c(0.3, 3),
     xlab = TeX(r"($\lambda_k$)"), ylab = TeX(r"($\varphi(\lambda_k) = \lambda_{k+1}$)"),
     main = TeX(r"(fixed-point iteration: the cobweb)"))
abline(0, 1, lty = 2)

x0 <- 2.8
for (k in 1:5) {
  x1 <- fp_map(x0)
  segments(x0, x0, x0, x1, col = 2, lwd = 2)          # up or down to the curve
  segments(x0, x1, x1, x1, col = 2, lwd = 2)          # across to the diagonal
  points(x0, x1, pch = 19, col = 2, cex = 0.7)
  if (k <= 3) text(x0, 0.35, TeX(sprintf(r"($\lambda_%d$)", k - 1)), col = 2, cex = 0.9)
  x0 <- x1
}
points(lhat, lhat, pch = 19, cex = 1.2)
text(lhat, lhat, TeX(r"($\hat{\lambda}$)"), pos = 2, offset = 0.6)
```

The method needs no derivatives, but the rearrangement must be found by hand, and different rearrangements of the same equation can converge at different speeds or diverge outright. Geometrically, a rearrangement with $\lvert\varphi'\rvert > 1$ at the crossing produces a cobweb that spirals outward instead of inward.

### Bisection

If $g$ is continuous and $g(a)$ and $g(b)$ have opposite signs, a root lies in $[a, b]$. Evaluate $g$ at the midpoint and retain whichever half still changes sign. For the truncated Poisson $-\ell'$ is negative at small $\lambda$ and positive at large $\lambda$, so any bracket straddling $\hat\lambda$ works.

```{r}
#| label: bisection
bisection <- function(d1, a, b, tol = 1e-10) {          # d1 is nll'
  if (d1(a) * d1(b) > 0) stop("endpoints do not bracket a root")
  k <- 0
  while (b - a > tol) {
    m <- (a + b) / 2
    if (d1(a) * d1(m) <= 0) b <- m else a <- m
    k <- k + 1
  }
  c(root = (a + b) / 2, iter = k)
}
```

The schematic below draws the bracket at each iteration as a bar beneath the curve of $g$. The dot is the midpoint, and the dotted line carries it up to the curve, where its sign decides which half survives.

```{r}
#| label: fig-scheme-bisection
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Illustrates bisection, a slow but robust root-finding scheme that only needs an interval known to contain the root.
#|   Bisection on the gradient $g$ of the negative log-likelihood: each iteration halves the bracket that contains the root."
par(mar = c(4, 4.5, 2.5, 1))
plot(lam, g(lam), type = "l", lwd = 2, ylim = c(-4.0, 0.9), xlim = c(0.12, 3),
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($g(\lambda)$)"),
     main = TeX(r"(bisection: each step halves the bracket)"))
abline(h = 0, lty = 3)
points(lhat, 0, pch = 19, cex = 1.2)

a <- 0.3; b <- 3
ylev <- seq(-2.9, -3.8, length.out = 4)
for (k in 1:4) {
  m <- (a + b) / 2
  segments(a, ylev[k], b, ylev[k], col = 4, lwd = 3)
  points(c(a, b), rep(ylev[k], 2), pch = "|", col = 4)
  segments(m, ylev[k], m, g(m), lty = 3, col = 2)
  points(m, ylev[k], pch = 19, col = 2, cex = 0.8)
  text(0.19, ylev[k], k, cex = 0.8, col = 4)
  if (g(a) * g(m) <= 0) b <- m else a <- m
}
```

The bracket halves at every step, so after $k$ iterations the error is at most $(b - a)/2^{k}$ and the iteration count is known in advance: about $\log_2\{(b-a)/\text{tol}\}$, regardless of the shape of $g$. Bisection cannot diverge, which makes it the natural fallback when a faster method misbehaves. R's `uniroot` uses Brent's method, which retains the bracketing guarantee while interpolating for speed.

### Newton–Raphson

#### Derivation

Approximate $g$ by its tangent line at $\theta_k$ and solve for where the tangent crosses zero:
$$
g(\theta) \approx g(\theta_k) + g'(\theta_k)(\theta - \theta_k) = 0
\implies \theta_{k+1} = \theta_k - \frac{g(\theta_k)}{g'(\theta_k)}
= \theta_k - \frac{-\ell'(\theta_k)}{-\ell''(\theta_k)} .
$$

The two minus signs cancel, so this is numerically the same step a score-based
derivation would give; writing it this way keeps the whole section in
minimization form. Equivalently, replace $\mathcal{L}$ by its second-order Taylor
expansion at $\theta_k$ and jump to the minimum of that parabola. For an exactly
quadratic $\mathcal{L}(\theta) = a(\theta - u)^2 + C$ with $a > 0$ we have
$g = 2a(\theta - u)$ and $g' = 2a$, so
$$
\theta_{k+1} = \theta_k - \frac{2a(\theta_k - u)}{2a} = u,
$$
in one step from any starting value. The normal-mean example of the previous section is exactly of this form: Newton converges immediately.

The left panel below traces three steps. At each iterate, the tangent to $g$ is followed to the horizontal axis, and where it lands is the next iterate. As the curve straightens near the root, the tangent becomes an excellent approximation and the steps land almost exactly on the answer.

The right panel contrasts one Newton step with one step of the method of scoring from the same point. Newton uses the actual slope $g'(\lambda_k)$, the observed information; scoring uses $I(\lambda_k)$, the same curvature averaged over hypothetical datasets. The two lines differ away from the root and coincide as $\lambda_k$ approaches it.

```{r}
#| label: fig-scheme-newton
#| fig-width: 10
#| fig-height: 4.8
#| fig-cap: "Illustrates Newton-Raphson, which replaces the gradient of the negative log-likelihood by its tangent line at each step, and contrasts it with Fisher scoring, which uses the expected rather than the observed curvature.
#|   Newton-Raphson following the tangent to $-\\ell'$ (left), and one Newton step compared with one Fisher-scoring step from the same point (right)."
par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))

### three Newton steps
plot(lam, g(lam), type = "l", lwd = 2, ylim = c(-2.6, 0.9),
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($g(\lambda)$)"),
     main = TeX(r"(Newton-Raphson: follow the tangent)"))
abline(h = 0, lty = 3)
points(lhat, 0, pch = 19, cex = 1.2)

x0 <- 0.45
for (k in 1:3) {
  x1 <- x0 - g(x0) / dg(x0)
  segments(x0, g(x0), x1, 0, col = 2, lwd = 2)
  segments(x1, 0, x1, g(x1), lty = 3, col = "grey40")
  points(c(x0, x1), c(g(x0), 0), pch = 19, col = 2, cex = 0.7)
  text(x0, g(x0), TeX(sprintf(r"($\lambda_%d$)", k - 1)),
       pos = if (k == 1) 4 else 1, col = 2, cex = 0.9)
  x0 <- x1
}

## Newton step versus scoring step from the same point
plot(lam, g(lam), type = "l", lwd = 2, ylim = c(-2.6, 0.9),
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($g(\lambda)$)"),
     main = TeX(r"(observed versus expected curvature)"))
abline(h = 0, lty = 3)
points(lhat, 0, pch = 19, cex = 1.2)

x0 <- 0.45
x_nr <- x0 - g(x0) / dg(x0)     # slope  nll''(lambda) = -l''(lambda)
x_fs <- x0 - g(x0) / info(x0)       # slope  I(lambda)
segments(x0, g(x0), x_nr, 0, col = 2, lwd = 2)
segments(x0, g(x0), x_fs, 0, col = 4, lwd = 2, lty = 5)
points(c(x0, x_nr, x_fs), c(g(x0), 0, 0), pch = 19, cex = 0.7, col = c(1, 2, 4))
legend("bottomright", bty = "n", lwd = 2, lty = c(1, 5), col = c(2, 4),
       legend = TeX(c(r"(Newton: slope $g'(\lambda_k)$)", r"(scoring: slope $I(\lambda_k)$)")))
```

### Advantage: quadratic convergence

Let $e_k = \theta_k - \hat\theta$ and expand $g(\hat\theta) = 0$ about $\theta_k$:
$$
0 = g(\theta_k) - g'(\theta_k)e_k + \tfrac12 g''(\xi)e_k^{2}
\implies e_{k+1} = e_k - \frac{g(\theta_k)}{g'(\theta_k)} = \frac{g''(\xi)}{2\,g'(\theta_k)}\,e_k^{2} .
$$

The error is squared at each step, so the number of correct digits roughly doubles. Two conditions appear in the constant: $g'(\hat\theta) \ne 0$, meaning the minimum is a genuine bowl and not a flat spot, and $g''$ bounded near the root. Asymptotic normality says that negative log-likelihoods are approximately quadratic near the MLE, which is why Newton is so effective there.

```{r}
#| label: newton
newton <- function(d1, d2, x0, tol = 1e-10, maxit = 100, trace = FALSE) {
  for (k in 1:maxit) {                                 # d1 = nll', d2 = nll''
    x1 <- x0 - d1(x0) / d2(x0)
    if (trace) cat(sprintf("iter %2d: %.10f\n", k, x1))
    if (!is.finite(x1)) stop("diverged")
    if (abs(x1 - x0) < tol) return(c(root = x1, iter = k))
    x0 <- x1
  }
  stop("not converged")
}
```

### A Classic Illustration: Newton's Method for $\sqrt{a}$

**Derivation.** To find $\sqrt{a}$, take $f(x) = x^2 - a$, whose positive root is $\sqrt a$. Then $f'(x) = 2x$, and the Newton step is,
$$
x_{k+1} = x_k - \frac{f(x_k)}{f'(x_k)} = x_k - \frac{x_k^{2} - a}{2x_k} = \frac{1}{2}\left(x_k + \frac{a}{x_k}\right).
$$

The final form is the Babylonian rule: if $x_k$ is too small then $a/x_k$ is too large, and the root lies between them, so average the pair. Newton's tangent line lands exactly on the midpoint.

The error follows the general quadratic law. Writing $e_k = x_k - \sqrt a$,
$$
e_{k+1} = \frac{1}{2}\left(x_k + \frac{a}{x_k}\right) - \sqrt a = \frac{\left(x_k - \sqrt a\right)^{2}}{2x_k} = \frac{e_k^{2}}{2x_k},
$$

so the number of correct digits roughly doubles each step.

```{r}
#| label: babylonian-sqrt
babylonian <- function(a, x0 = 1, iters = 6)
  Reduce(function(x, k) (x + a / x) / 2, 1:iters, init = x0, accumulate = TRUE)

traces <- sapply(c(10, 20, 50), babylonian)
dimnames(traces) <- list(paste("iter", 0:6), sprintf("sqrt(%d)", c(10, 20, 50)))

print(traces, digits = 12)
sqrt(c(10, 20, 50))                       # for comparison
signif(abs(traces - rep(sqrt(c(10, 20, 50)), each = 7)), 3)    # errors
```

Starting from $x_0 = 1$ in every case, the first two steps do the coarse work — the larger $a$ is, the further the opening leap — and from the third step on the error squares: roughly $10^{-1}$, then $10^{-3}$, then $10^{-6}$, then machine precision. Six iterations suffice for all three, though $\sqrt{50}$ spends one more step than $\sqrt{10}$ recovering from the poor start.

### Problem: divergence

Newton uses only local information, so the quadratic model can be badly wrong away from the maximum. Three failures occur:

- **wrong curvature** — where $g' < 0$ the local model is an upside-down bowl, and the step heads toward a maximum or off to infinity;

- **overshooting** — where $g$ is nearly flat, $g'(\theta_k) \approx 0$ produces an enormous step, possibly outside the parameter space;

- **multiple roots** — the iteration converges to whichever stationary point it happens to reach, which need not be the global minimum.

The single-observation Cauchy likelihood is the standard illustration. With one observation at zero, $\mathcal{L}(\theta) = -\ell(\theta) = \log(1 + \theta^{2})$ and $g(\theta) = -\ell'(\theta) = 2\theta/(1 + \theta^{2})$, whose only root is $\theta = 0$. That derivative rises, turns over at $\theta = 1$, and decays to zero in the tails — exactly the flat region that defeats Newton.

```{r}
#| label: fig-cauchy-single-obs
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: "Explains why Newton-Raphson can fail on the Cauchy likelihood by showing the shape of the contribution of a single observation.
#|   Single-observation Cauchy negative log-likelihood and its gradient: the gradient flattens in the tails, the region that defeats Newton-Raphson."
nll_c <- function(t) log(1 + t^2)                  # negative log-likelihood
g_c   <- function(t) 2 * t / (1 + t^2)             # gradient  L'(theta)
dg_c  <- function(t) 2 * (1 - t^2) / (1 + t^2)^2   # curvature L''(theta)

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
curve(nll_c, -5, 5, n = 400, lwd = 2, xlab = TeX(r"($\theta$)"), ylab = TeX(r"($L(\theta)$)"),
      main = TeX(r"(negative log-likelihood)"))
curve(g_c, -5, 5, n = 400, lwd = 2, xlab = TeX(r"($\theta$)"), ylab = TeX(r"($g(\theta)$)"),
      main = TeX(r"(gradient, with the flat tails)"))
abline(h = 0, lty = 3); abline(v = c(-1, 1), col = 2, lty = 2)
```

Tracing the iterates from three starting values shows the boundary at $\lvert\theta_0\rvert = 1$, where $g$ turns over:

```{r}
#| label: newton-divergence-trace
newton_trace <- function(d1, d2, x0, iters = 8) {
  path <- numeric(iters + 1); path[1] <- x0
  for (k in 1:iters) {
    x0 <- x0 - d1(x0) / d2(x0)
    path[k + 1] <- if (is.finite(x0)) x0 else NA
  }
  path
}

starts <- c(0.5, 0.8, 1.2)
traces <- vapply(starts, function(v) newton_trace(g_c, dg_c, v), numeric(9))
colnames(traces) <- sprintf("start = %.1f", starts)
rownames(traces) <- paste("iter", 0:8)
signif(traces, 5)
```

From $\theta_0 = 0.5$, inside the region where the quadratic model is adequate, Newton converges quadratically to zero. From $\theta_0 = 0.8$ the first step already overshoots into the tail, and from $\theta_0 = 1.2$ — beyond the inflection point of $-\ell'$ — the iterates run away, each step landing further out where the derivative is flatter still.

```{r}
#| label: newton-animation
#| eval: false
## Interactive versions of the same three trajectories (requires the animation
## package and a graphics device that can write animations)
library(animation)
ani.options(interval = 1, nmax = 50)
newton.method(FUN = g_c, init = 0.5, rg = c(-1.5, 1.5))
newton.method(FUN = g_c, init = 0.8, rg = c(-8, 1.5))
newton.method(FUN = g_c, init = 1.2, rg = c(-2, 20))
```

### Univariate Fisher Scoring

Newton's step divides by the *observed* curvature $g'(\theta_k) = -\ell''(\theta_k)$, computed from the data at hand, which can be negative — or nearly zero — far from the minimum. Replacing it by the *expected* curvature $nI(\theta_k)$ gives the **method of scoring**:
$$
\theta_{k+1} = \theta_k - \frac{g(\theta_k)}{nI(\theta_k)}
= \theta_k + \frac{\ell'(\theta_k)}{nI(\theta_k)} .
$$

Because $nI(\theta) > 0$ for every $\theta$ by construction, the denominator never changes sign: every step of scoring is a descent step on $\mathcal{L}$, even at a point where Newton's step would point uphill or blow up. Near the minimum $nI(\hat\theta) \approx g'(\hat\theta)$, so scoring and Newton coincide and both converge quickly; the two methods differ only in how they behave away from the root.

```{r}
#| label: fisher-scoring-univariate
fisher_scoring <- function(d1, nI, x0, tol = 1e-10, maxit = 100) {
  for (k in 1:maxit) {                                 # d1 = nll', nI = expected info
    x1 <- x0 - d1(x0) / nI(x0)
    if (abs(x1 - x0) < tol) return(c(root = x1, iter = k))
    x0 <- x1
  }
  stop("not converged")
}

## a poor start: at lambda = 3.5 the curvature nll'' is nearly flat
c(newton_step_from_3.5 = 3.5 - g(3.5) / dg(3.5))  # overshoots to an infeasible negative lambda
rbind(newton  = tryCatch(newton(g, dg, 3.5), error = function(e) c(root = NA, iter = NA)),
      scoring = fisher_scoring(g, info, 3.5))
```

From this poor start, Newton's step lands on a negative — and hence infeasible — value of $\lambda$ and never recovers, while scoring's positive denominator lands its very first step near $\lambda \approx 1.64$, already close to the root, and converges normally from there. The price of that safety is that $nI(\theta)$ must be available in closed form; where it is not, Newton with the safeguards below is the fallback.

### Safeguards

Three remedies cover most cases. Start from a robust estimate — the sample median for a location model. Apply **step halving**: accept the full Newton step only if it decreases $\mathcal{L}$, otherwise halve it until it does. Or replace the observed curvature by the expected information, giving the **method of scoring**,
$$
\theta_{k+1} = \theta_k - \frac{g(\theta_k)}{n I(\theta_k)} ,
$$
in which the divisor is positive by construction, so every step descends. Near the MLE $nI(\hat\theta) \approx g'(\hat\theta)$, so scoring inherits Newton's speed while being far more stable away from the optimum. Its cost is that the expectation must be available in closed form.

```{r}
#| label: safe-newton
safe_newton <- function(f, d1, d2, x0, lower = -Inf, upper = Inf,
                        tol = 1e-10, maxit = 100, max_halve = 40) {
  f0 <- f(x0)                                    # f = nll, d1 = nll', d2 = nll''
  for (k in 1:maxit) {
    step <- -d1(x0) / d2(x0)
    if (d2(x0) < 0) step <- -sign(d1(x0))        # wrong curvature: just move downhill
    for (h in 0:max_halve) {                     # halve until feasible and improving
      x1 <- x0 + step / 2^h
      if (x1 > lower && x1 < upper && is.finite(f(x1)) && f(x1) <= f0) break
    }
    if (abs(x1 - x0) < tol) return(c(root = x1, iter = k, halvings = h))
    x0 <- x1; f0 <- f(x1)
  }
  stop("not converged")
}
```

### Golden Section Search

Golden section minimizes $\mathcal{L}$ directly, using **no derivative at all**, which makes it the method of choice when $\ell$ is not differentiable or its derivative is unreliable. Keep a bracket $[x_l, x_u]$ containing the minimum of a unimodal $\mathcal{L}$, with two interior points placed at the golden fraction $\kappa = (3 - \sqrt5)/2 \approx 0.382$ of the way in from each end. Comparing the two interior values discards one end, and the ratio is chosen precisely so that the surviving interior point is correctly positioned for the next iteration: one new evaluation per step, with the bracket shrinking by the factor $0.618$.

```{r}
#| label: golden
golden <- function(f, brack.int, eps = 1e-4, ...) {
  kap <- (3 - sqrt(5)) / 2
  xl <- min(brack.int); xu <- max(brack.int)
  tmp <- kap * (xu - xl)
  xmu <- xu - tmp; xml <- xl + tmp
  fl <- f(xml, ...); fu <- f(xmu, ...)
  while (abs(xu - xl) > (1e-5 + abs(xl)) * eps) {
    if (fl < fu) {
      xu <- xmu; xmu <- xml; fu <- fl
      xml <- xl + kap * (xu - xl); fl <- f(xml, ...)
    } else {
      xl <- xml; xml <- xmu; fl <- fu
      xmu <- xu - kap * (xu - xl); fu <- f(xmu, ...)
    }
  }
  if (fl < fu) xml else xmu
}
```

The schematic makes the reuse visible. Each bar is the current bracket with its two interior points; the point that survives into the next iteration is drawn in blue and sits at exactly the right position there, so only the red point has to be evaluated afresh.

```{r}
#| label: fig-scheme-golden
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Illustrates golden section search, a derivative-free method for optimizing a one-dimensional function, and why it needs only one new function evaluation per step.
#|   Golden section search: the interior point that survives (blue) is already correctly positioned for the next iteration, so only one new point (red) is evaluated per step."
kap <- (3 - sqrt(5)) / 2

par(mar = c(4, 4.5, 2.5, 1))
plot(lam, nll(lam), type = "l", lwd = 2, ylim = c(0.30, 1.30), xlim = c(0.12, 3),
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($L(\lambda)$)"),
     main = TeX(r"(golden section: one new evaluation per step)"))
points(lhat, nll(lhat), pch = 19, cex = 1.2)

xl <- 0.3; xu <- 3
ylev <- seq(0.42, 0.34, length.out = 3)
for (k in 1:3) {
  xml <- xl + kap * (xu - xl); xmu <- xu - kap * (xu - xl)
  segments(xl, ylev[k], xu, ylev[k], col = 4, lwd = 3)
  points(c(xl, xu), rep(ylev[k], 2), pch = "|", col = 4)
  keep_lower <- nll(xml) < nll(xmu)
  points(xml, ylev[k], pch = 19, cex = 0.9, col = if (keep_lower) 4 else 2)
  points(xmu, ylev[k], pch = 19, cex = 0.9, col = if (keep_lower) 2 else 4)
  text(0.19, ylev[k], k, cex = 0.8, col = 4)
  if (keep_lower) xu <- xmu else xl <- xml
}
```

The first test is the negative log-likelihood of the classical genetic linkage model, with cell counts $(1997, 906, 904, 32)$ and probabilities $\left((2+\theta)/4, (1-\theta)/4, (1-\theta)/4, \theta/4\right)$; it is smooth, so any method would work. The second, $f(x) = |x - 10|$, is not differentiable at its minimum, so no root-finder applies — golden section handles it without modification.

```{r}
#| label: fig-golden-tests
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: "Tests golden section search on a smooth likelihood and on a non-differentiable function to show that the method needs no derivatives.
#|   Golden section search applied to a smooth genetic-linkage log-likelihood (left) and to a non-differentiable function $|x-10|$ (right), with the located minimizer marked in red."
f1 <- function(theta) -1997 * log(2 + theta) - 1810 * log(1 - theta) - 32 * log(theta)
f2 <- function(x) abs(x - 10)

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
curve(f1, 0.001, 0.999, n = 400, lwd = 2, xlab = TeX(r"($\theta$)"), ylab = TeX(r"($L(\theta)$)"),
      main = TeX(r"(genetic linkage model)"))
abline(v = golden(f1, c(0.001, 0.999)), col = 2, lty = 2, lwd = 2)

curve(f2, 0, 20, n = 400, lwd = 2, xlab = TeX(r"($x$)"), ylab = TeX(r"($|x - 10|$)"),
      main = TeX(r"(not differentiable at the minimum)"))
abline(v = golden(f2, c(0, 20)), col = 2, lty = 2, lwd = 2)

c(linkage = golden(f1, c(0.001, 0.999)), abs_value = golden(f2, c(0, 20)))
```

### Numerical Derivatives

When $g$ or $g'$ is tedious or error-prone to derive by hand, both can be approximated by finite differences:
$$
g(\theta) \approx \frac{\mathcal{L}(\theta + h) - \mathcal{L}(\theta - h)}{2h}, \qquad
g'(\theta) \approx \frac{\mathcal{L}(\theta + h) - 2\,\mathcal{L}(\theta) + \mathcal{L}(\theta - h)}{h^{2}}.
$$

The step $h$ trades off two sources of error: too large and the approximation suffers **truncation error** from the curvature of $\mathcal{L}$ itself; too small and it suffers **rounding error**, because the numerator subtracts two nearly equal floating-point numbers. A step of $h \approx \sqrt{\varepsilon}\,(1 + \lvert\theta\rvert) \approx 10^{-8}$ balances the two for the first derivative, and $h \approx \varepsilon^{1/3}$ for the second, where $\varepsilon \approx 2 \times 10^{-16}$ is double-precision machine epsilon.

```{r}
#| label: numerical-derivative-check
h <- 1e-5
c(analytic = dg(lhat),
  numeric  = (nll(lhat + h) - 2 * nll(lhat) + nll(lhat - h)) / h^2)
```

The two agree to several digits, which is reassuring but also the point: `nlm()` and `optim(method = "BFGS")` fall back on exactly this kind of finite differencing whenever an analytic gradient is not supplied, and `numDeriv::grad()` / `numDeriv::hessian()` implement more accurate variants (Richardson extrapolation) for when a single finite difference is not precise enough.

### Summary of the Methods

| method | needs | convergence | fails when |
|:--|:--|:--|:--|
| fixed point | a rearrangement $\theta = \varphi(\theta)$ | linear, factor $\lvert \varphi'\rvert$ | $\lvert \varphi'\rvert \ge 1$ |
| bisection | $g$ and a sign-changing bracket | linear, factor $1/2$ | no bracket available |
| Newton–Raphson | $g$ and $g'$ | quadratic near the root | flat or wrong curvature |
| method of scoring | $g$ and $I(\theta)$ | linear to quadratic | $E[\cdot]$ unavailable |
| golden section | $\mathcal{L}$ only, and a bracket | linear, factor $0.618$ | $\mathcal{L}$ not unimodal |

### Putting the Methods Together on the Truncated Poisson

Each method above was drawn on this model one step at a time. Running them to
convergence side by side shows how the iterate sequences themselves differ.

#### The Three Iterations

Each method below returns the whole sequence of iterates together with the
value of $\mathcal{L}$ attained, so the trajectories can be compared directly. All
three use the `nll`, `g`, `dg` and `info` — that is, $\mathcal{L}$, $g$, $g'$ and
$I$ — defined at the start of this
section, with **simple iteration** using the rearrangement
$\lambda = \overline x(1 - e^{-\lambda}) = \varphi(\lambda)$ derived earlier.

```{r}
#| label: nzp-iterations
## Each returns the iterate sequence together with the nll attained.
iterate_path <- function(step, n_iter, lambda0) {
  lambda <- numeric(n_iter + 1); lambda[1] <- lambda0
  for (i in 1:n_iter) lambda[i + 1] <- step(lambda[i])
  data.frame(lambda = lambda, nll = nll(lambda))
}

nzp.simple.iteration <- function(n_iter, lambda0)
  iterate_path(function(l) fp_map(l), n_iter, lambda0)

nzp.newton.raphson <- function(n_iter, lambda0)
  iterate_path(function(l) l - g(l) / dg(l), n_iter, lambda0)

nzp.method.of.scoring <- function(n_iter, lambda0)
  iterate_path(function(l) l - g(l) / info(l), n_iter, lambda0)
```

#### Comparison From Several Starting Values

The three methods are run for 15 iterations from a sensible start, a start far below, and a start far above.

```{r}
#| label: nzp-compare-starts
starts <- c(xbar, 0.1, 100)

trace_tab <- function(fun, n_iter = 15)
  vapply(starts, function(s) fun(n_iter, s)$lambda, numeric(n_iter + 1))

paths <- list(simple = trace_tab(nzp.simple.iteration),
              newton = trace_tab(nzp.newton.raphson),
              scoring = trace_tab(nzp.method.of.scoring))

### first six iterates from each starting value
for (m in names(paths)) {
  cat("\n", m, "\n")
  tab <- paths[[m]][1:6, ]
  dimnames(tab) <- list(paste("iter", 0:5), sprintf("start = %.4g", starts))
  print(signif(tab, 6))
}
```

The pattern is the one the theory predicts. Newton–Raphson takes the fewest iterations from a good start, and its digits lock in from the left as the error is squared each step. Simple iteration converges from every start, including the extreme ones, but slowly and at a steady rate. The method of scoring is intermediate: always descending on $\mathcal{L}$, and nearly as fast as Newton once close.

```{r}
#| label: fig-nzp-convergence-plot
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Compares the convergence speed of three iterative algorithms on the same problem by showing how quickly each reduces its error.
#|   Error decay (log scale) of simple iteration, Newton-Raphson, and the method of scoring for the zero-truncated Poisson MLE: Newton's error curve bends downward, reflecting quadratic convergence."
err <- sapply(paths, function(p) abs(p[, 1] - lhat))   # from the xbar start

par(mar = c(4, 4.5, 2.5, 1))
matplot(0:15, pmax(err, 1e-17), log = "y", type = "b", pch = 19, lty = 1, col = c(4, 1, 2),
        xlab = TeX(r"(iteration $k$)"), ylab = TeX(r"($|\lambda_k - \hat{\lambda}|$)"),
        main = TeX(r"(error decay from $\lambda_0 = \bar{x}$)"))
legend("bottomleft", bty = "n", pch = 19, lty = 1, col = c(4, 1, 2),
       legend = c("simple iteration", "Newton-Raphson", "method of scoring"))
```

On a log scale a linear method is a straight line; Newton's curve bends downward until it reaches the floor set by machine precision.

```{r}
#| label: nzp-builtin-check
c(newton      = unname(newton(g, dg, xbar)["root"]),
  fixed_point = unname(fixed_point(fp_map, xbar)["root"]),
  bisection   = unname(bisection(g, 0.1, 20)["root"]),
  golden      = golden(nll, c(0.1, 20), eps = 1e-8),
  uniroot     = uniroot(g, c(0.1, 20), tol = 1e-10)$root,
  optimize    = optimize(nll, c(0.1, 20), tol = 1e-10)$minimum)
```

All six agree, which is the point of running them together: the methods differ in cost and robustness, not in the answer.

#### Standard Error From the Curvature

The optimizer has already produced everything inference needs: the curvature
$g'(\hat\lambda)$ that Newton divided by *is* the observed information per
observation, so the total is $J(\hat\lambda) = n\,g'(\hat\lambda)$.

```{r}
#| label: nzp-standard-error
n_obs  <- length(xp)
se_obs <- 1 / sqrt(n_obs * dg(lhat))      # observed information
se_exp <- 1 / sqrt(n_obs * info(lhat))        # expected information

c(mle = lhat, se_observed = se_obs, se_expected = se_exp)
lhat + c(-1.96, 1.96) * se_obs                     # 95% Wald interval
```

```{r}
#| label: fig-nzp-quadratic-approx
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Shows how the curvature of the negative log-likelihood at the MLE yields a standard error and a Wald confidence interval, by approximating it with a parabola.
#|   Zero-truncated Poisson negative log-likelihood versus its quadratic approximation at $\\hat\\lambda$, with the resulting Wald 95% interval marked."
lam_se <- seq(lhat - 4 * se_obs, lhat + 4 * se_obs, length.out = 300)
quad <- 0.5 * n_obs * dg(lhat) * (lam_se - lhat)^2

par(mar = c(4, 4.5, 2.5, 1))
plot(lam_se, n_obs * (nll(lam_se) - nll(lhat)), type = "l", lwd = 2,
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($L(\lambda) - L(\hat{\lambda})$)"),
     main = TeX(r"(negative log-likelihood and its quadratic approximation at $\hat{\lambda}$)"))
lines(lam_se, quad, col = 2, lwd = 2, lty = 2)
abline(h = 1.92, col = 4, lty = 3)
abline(v = lhat + c(-1.96, 1.96) * se_obs, col = 4, lty = 3)
legend("top", bty = "n", lwd = 2, lty = c(1, 2), col = c(1, 2),
       legend = TeX(c(r"($L(\lambda) - L(\hat{\lambda})$)", r"(quadratic from $J(\hat{\lambda})$)")))
```

The parabola implied by the observed information tracks the negative log-likelihood closely over the width of the interval, so the Wald interval is trustworthy here.

#### Evaluating the Naive Estimator

All of this machinery is worth the effort only if the MLE beats the easy alternative. The naive estimator ignores the truncation and reports $\overline x$. Simulating many truncated samples gives the sampling distributions of both estimators.

```{r}
#| label: fig-nzp-mle-vs-naive
#| cache: true
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: "Uses simulation to show why the MLE is worth computing: ignoring the truncation gives a biased estimator, whereas the MLE is centred on the truth.
#|   Sampling distributions of the naive estimator $\\bar x$ versus the MLE for a zero-truncated Poisson mean, $\\lambda_{true} = 1.5$: the naive estimator is badly biased upward, the MLE is centred on the truth."
mle_nzp <- function(n)
  uniroot(function(l) g(l, xb = mean(n)), c(1e-6, 50), tol = 1e-10)$root

sim_nzp <- function(lambda, n_gen = 100, B = 500)
  vapply(1:B, function(b) {
    y <- rpois(n_gen, lambda); n <- y[y > 0]
    c(naive = mean(n), mle = mle_nzp(n))
  }, c(naive = 0, mle = 0))

for (lambda_true in c(0.5, 1.5)) {
  est <- sim_nzp(lambda_true)
  tab <- t(apply(est, 1, function(e)
    c(bias = mean(e) - lambda_true, sd = sd(e), rmse = sqrt(mean((e - lambda_true)^2)))))
  cat("\nlambda =", lambda_true, "\n"); print(round(tab, 4))
}

est <- sim_nzp(1.5)
par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
for (m in c("naive", "mle")) {
  hist(est[m, ], breaks = 25, col = "grey90",
       xlim = range(est), xlab = TeX(r"($\hat{\lambda}$)"),
       main = TeX(sprintf(r"(%s, bias $= %.3f$)", m, mean(est[m, ]) - 1.5)))
  abline(v = 1.5, col = 2, lwd = 2, lty = 2)
}
```

The naive estimator is biased upward, and severely so when $\lambda$ is small, since that is when most of the probability mass sits on the discarded zero. Its bias does not shrink with sample size: it reflects the wrong model, not too little data. The MLE is centred on the truth with a standard deviation matching the information-based standard error of the previous section. The optimization was the means; the correct estimator was the end.

## Multivariate Newton-Raphson Optimzation Methods


### Gradient and Hessian: A Geometric Illustration

The two objects entering the multivariate Newton step have a direct geometric reading. The negative gradient $-g(\theta)$ is the direction of steepest descent and is perpendicular to the contour of $\mathcal{L}$ through $\theta$; the Hessian describes how that direction curves, and at a strict local minimum $J(\theta) = \nabla g(\theta)$ is positive definite. Near $\hat\theta$, $\mathcal{L}$ is well approximated by the quadratic
$$
\mathcal{L}(\theta) \approx \mathcal{L}(\hat\theta) + \tfrac12(\theta - \hat\theta)^\top J(\hat\theta)(\theta - \hat\theta),
$$
whose contours are ellipses centred at $\hat\theta$. Writing the eigendecomposition $J(\hat\theta) = Q\Lambda Q^\top$, the axes of these ellipses point along the eigenvectors in $Q$, and the half-length along axis $j$ is proportional to $1/\sqrt{\lambda_j}$: a large eigenvalue means sharp curvature and a short, well-determined direction, while $\lambda_{\max}/\lambda_{\min}$ measures how elongated — and, once the axes are tilted, how correlated the parameters are.

```{r}
#| label: fig-hessian-ellipse
#| echo: false
#| fig-width: 7
#| fig-height: 7
#| fig-cap: "Gives a geometric picture of the gradient and the Hessian in two dimensions, explaining why the direction of steepest descent generally does not point at the minimum.
#|   Contours of a quadratic negative log-likelihood: steepest-descent arrows (green) run perpendicular to the contours but do not point at the minimum unless the ellipses are circular; the dashed axes are the eigenvectors of $J(\\hat\\theta)$, each with half-length $1/\\sqrt{\\lambda_j}$."
A <- matrix(c(2, 1.2, 1.2, 1), 2)
s <- seq(-2, 2, length.out = 80); grid_pts <- expand.grid(x = s, y = s)
z <- matrix(apply(grid_pts, 1, function(v) 0.5 * drop(t(v) %*% A %*% v)), 80)   # nll contours
par(pty = "s", mar = c(4, 4, 1, 1))
contour(s, s, z, nlevels = 12, xlab = expression(theta[1]), ylab = expression(theta[2]))
## gradient arrows
pts <- rbind(c(1, 1), c(-1.5, 0.5), c(0.5, -1.5))
for (i in 1:3) {
  desc <- -A %*% pts[i, ]                       # steepest-descent direction
  arrows(pts[i, 1], pts[i, 2], pts[i, 1] + 0.3 * desc[1], pts[i, 2] + 0.3 * desc[2],
         col = 3, lwd = 2, length = 0.1)
}
## principal axes with eigenvalue labels: length 1/sqrt(lambda) on each side
e <- eigen(A); k <- 1.6
for (j in 1:2) {
  v <- e$vectors[, j] * k / sqrt(e$values[j])
  segments(-v[1], -v[2], v[1], v[2], col = "firebrick", lwd = 2, lty = 2)
  text(v[1], v[2], bquote(lambda[.(j)] == .(round(e$values[j], 2))),
       col = "firebrick", pos = if (j == 1) 4 else 2, cex = 0.9)
}
points(0, 0, pch = 19)
```

Every arrow in the figure is perpendicular to the contour it starts on and points generally downhill toward the minimum at the origin, but only an arrow started exactly on a principal axis points straight at it. This is why Newton's step, which multiplies the gradient by $J(\hat\theta)^{-1}$ rather than simply following it, converges directly to $\hat\theta$: it rescales and rotates the descent direction to undo exactly this distortion. The plain gradient-descent methods of a later section have no such correction, and zig-zag across elongated contours like these.

### Naive Newton-Raphson Iterations

In the previous section the parameter was a scalar, and the log-likelihood could
be inspected by plotting it. Once $\theta \in \mathbb{R}^p$ neither is true: we
cannot see the surface, and the notion of "bracketing a root" that underlies
bisection and golden section has no direct multivariate analogue. What survives
is the local quadratic approximation, and the Newton–Raphson algorithm is its
direct consequence.

This section develops the multivariate Newton–Raphson method and its
statistical variant, Fisher scoring, applies both to logistic regression,
examines interactively how and why the iteration can diverge, and closes with a
survey of the derivative-free and gradient-only alternatives.

Write the log-likelihood as $\ell(\theta)$ for $\theta \in \mathbb{R}^p$, and
define.
$$
g(\theta) = \nabla \mathcal{L}(\theta) = -\frac{\partial \ell(\theta)}{\partial \theta}
\quad (p \times 1, \text{ the gradient}),
\qquad
\nabla g(\theta) = -\frac{\partial^2 \ell(\theta)}{\partial \theta \, \partial \theta^{\top}}
\quad (p \times p, \text{ the Hessian of } \mathcal{L}).
$$

The univariate $g = \mathcal{L}'$ becomes the gradient $g = \nabla\mathcal{L}$ with no change of
meaning. As before, the Hessian of $\mathcal{L}$ *is* the **observed information
matrix**, so we write $J(\theta) = \nabla g(\theta)$ throughout.

Expand $\mathcal{L}$ to second order about the current iterate $\theta^{(k)}$:
$$
\mathcal{L}(\theta) \;\approx\; \mathcal{L}(\theta^{(k)})

 + g(\theta^{(k)})^{\top}(\theta - \theta^{(k)})
 + \tfrac{1}{2}\,(\theta - \theta^{(k)})^{\top} J(\theta^{(k)}) \,(\theta - \theta^{(k)}).
$$

This is a quadratic in $\theta$. If $J(\theta^{(k)})$ is positive definite the
quadratic has a unique minimum, found by setting its derivative to zero:
$$
g(\theta^{(k)}) + J(\theta^{(k)})(\theta - \theta^{(k)}) = 0 .
$$

Solving for $\theta$ gives the **Newton–Raphson update**:
$$
\boxed{\;\theta^{(k+1)} = \theta^{(k)} - J(\theta^{(k)})^{-1} g(\theta^{(k)}) \;}.
$$

The algorithm therefore replaces the negative log-likelihood by the quadratic that
matches its value, slope and curvature at the current point, and jumps to the
minimum of that quadratic. Everything that follows — the fast convergence, and
the modes of failure — comes from how good or bad that replacement is.

Three remarks are worth making explicit.

*The inverse is never formed.* Write the update as
$\theta^{(k+1)} = \theta^{(k)} + \delta^{(k)}$ where $\delta^{(k)}$ solves the
linear system $J(\theta^{(k)})\,\delta = -g(\theta^{(k)})$. In R this is
`solve(J, -g)`, not `solve(J) %*% (-g)`. The former is faster and numerically more
stable; the latter computes $p^2$ numbers in order to use $p$ of them.

*The step is not the steepest-descent direction.* $\delta^{(k)} = -J^{-1}g$
points along $-g$ only when $J$ is a multiple of the identity. The information matrix
rescales and rotates the descent direction to account for differing curvature and
correlation among the parameters, which is exactly why Newton–Raphson is
insensitive to the scaling of the parameters while gradient descent is not.

*Positive definiteness is an assumption, not a guarantee.* Away from the minimum
$J(\theta^{(k)})$ can be indefinite or singular, in which case the "minimum of
the quadratic" is a saddle point or does not exist, and $\delta^{(k)}$ may point
uphill. @sec-divergence examines this.

#### Fisher scoring

The observed information $J(\theta)$ depends on the data. Its expectation under
the model,
$$
I(\theta) = E_{\theta}\{J(\theta)\} = -E_{\theta}\left\{ \frac{\partial^2 \ell(\theta)}{\partial \theta \, \partial \theta^{\top}} \right\},
$$

is the **expected (Fisher) information matrix**. Replacing $J$ by $I$ in the
update gives the **method of scoring**:
$$
\theta^{(k+1)} = \theta^{(k)} - I(\theta^{(k)})^{-1} g(\theta^{(k)}).
$$

The two algorithms have different iterates but the same fixed point, since both
stop where $g(\theta) = 0$. The practical differences:

| | Newton–Raphson | Fisher scoring |
|:---|:---|:---|
| Curvature matrix | Observed information $J(\theta)$ | Expected information $I(\theta)$ |
| Derivation | Second derivative of $\mathcal{L}$ | Expectation of that second derivative |
| Positive definite? | Only guaranteed near the minimum | Always, wherever the model is identified |
| Algebra | Can be messy | Often simpler, as data-dependent terms average out |
| Local convergence | Quadratic | Linear, but typically with a very small rate constant |
| Stability from poor starts | Less stable | More stable |

: Contrasts Newton-Raphson with Fisher scoring, the two choices of curvature matrix in a second-order update, to help decide between them. Newton–Raphson compared with Fisher scoring {#tbl-nr-scoring}

The guaranteed positive definiteness of $I(\theta)$ is the main practical
argument for scoring: the update direction is always a descent direction on
$\mathcal{L}$, so the iteration cannot be sent uphill by an ill-behaved Hessian.

For generalised linear models with the **canonical link** the two coincide
exactly, because the second derivative of the log-likelihood does not involve
the response. Logistic regression uses the canonical logit link, so the example
below is simultaneously Newton–Raphson, Fisher scoring, and — as we note at the
end of the section — iteratively reweighted least squares.


### Convergence properties

Near a minimum $\hat\theta$ of $\mathcal{L}$ at which $J(\hat\theta)$ is positive definite,
Newton–Raphson converges **quadratically**: the error satisfies
$$
\lVert\theta^{(k+1)} - \hat\theta\rVert \;\le\; C \,\lVert\theta^{(k)} - \hat\theta\rVert^{2},
$$

for some constant $C$ depending on the third derivatives. Informally, the number
of correct digits doubles at each iteration, which is why five or six iterations
are usually enough and why a well-started run appears to converge instantly.

The qualification "near" is essential. The theorem is local: it guarantees
nothing about a start far from $\hat\theta$, and nothing at all if $J$ fails to
be positive definite along the way.

### Standard errors as a by-product

At convergence the same matrix that drove the iteration supplies the asymptotic
variance. Under the usual regularity conditions,
$$
\hat\theta \;\dot\sim\; N\!\left(\theta, \; J(\hat\theta)^{-1}\right),
$$

so $\operatorname{\mathrm{SE}}(\hat\theta_j) = \sqrt{[J(\hat\theta)^{-1}]_{jj}}$. This is
a genuine practical advantage of Newton-type methods over derivative-free ones:
the curvature information needed for inference has already been computed. Methods
that never form a Hessian require a separate numerical differentiation step, which
is what `hessian = TRUE` requests from `nlm` and `optim`.

### Example: Multivariate Newton-Raphson for Logistic Regression

#### The model and its derivatives

With a single covariate $x_i$ and binary response $y_i$, let
$u_i = \beta_0 + \beta_1 x_i$ and $p_i = e^{u_i}/(1 + e^{u_i})$. The
log-likelihood is
$$
\ell(\beta) = \sum_{i=1}^{n}\left\{ y_i u_i - \log\left(1 + e^{u_i}\right) \right\}.
$$

Writing $X$ for the $n \times 2$ design matrix with rows $(1, x_i)$, the gradient
of $\mathcal{L}(\beta) = -\ell(\beta)$ and the information are,
$$
g(\beta) = -X^{\top}(y - p), \qquad J(\beta) = X^{\top} W X, \qquad W = \operatorname{diag}\{p_i(1 - p_i)\}.
$$

Note that $J$ does not involve $y$. This is the canonical-link property
mentioned above, and it means $J(\beta) = I(\beta)$ identically.

Two numerical points, both carried over from the chapter on computer arithmetic.
First, $\log(1 + e^{u})$ overflows for $u \gtrsim 710$; the stable form is
$\log(1+e^u) = \max(u,0) + \log(1 + e^{-\vert{}u\vert{}})$. Second, $p_i(1-p_i) \to 0$ when
$\vert{}u_i\vert{}$ is large, so $W$ — and hence $J$ — degenerates toward singularity when
the fitted probabilities are extreme. That observation is the whole explanation
of the divergence we are about to see.

#### The model as a computation graph {#sec-computation-graph}

Before differentiating anything it helps to fix the object being differentiated.
Write the model as a **computation graph**: one node for every intermediate
quantity — the inputs, each linear unit, each activation, the linear predictor,
the loss — and one directed edge $u \to z$ whenever $z$ is computed directly from
$u$. The graph is acyclic, values flow forward along its edges, and the whole of
backpropagation is two rules stated on it.

The first is the **node rule**. Write
$$\delta_z = \frac{\partial \mathcal{L}_i}{\partial z}$$
for the sensitivity of the loss to node $z$. Since $z$ can influence the loss
only through the nodes it feeds, the multivariate chain rule gives one term per
outgoing edge,
$$\delta_z \;=\; \sum_{p \,\in\, \mathrm{consumers}(z)} \delta_p \, \frac{\partial p}{\partial z} ,$$
and the recursion starts at the loss itself with $\delta_{\mathcal{L}_i} = 1$.

The second is the **edge rule**. Every parameter sits on an edge that forms a
linear combination, and it reaches $z$ only through the single term $\omega\, u$ it
contributes, so $\partial z/\partial \omega = u$ and
$$\frac{\partial \mathcal{L}_i}{\partial \omega_{u \to z}} \;=\; \delta_z \cdot u :$$
the sensitivity at the arrow's head times the value at its tail.

That is the entire algorithm. Everything below — logistic regression here, a
hidden layer later in this chapter, a hundred layers in a modern network — is
these two rules applied to a larger graph. Note that the graph is neither a chain
nor a tree: a unit typically feeds *several* others, so most nodes have more than
one consumer, and the sum in the node rule is what gathers those contributions.
@sec-backprop-dag writes the two rules out for a network of arbitrary depth.

#### The gradient by the chain rule

The gradient above can be written down directly, but it is worth deriving it the
way an automatic-differentiation system would, because the same mechanics carry
over unchanged to models with many more layers. Logistic regression *is* a
neural network — the smallest interesting one: two inputs, one linear unit, a
sigmoid activation, and a cross-entropy loss.

Take a single observation and follow it forward through the model:
$$
u_i = \beta_0 \cdot 1 + \beta_1 x_i, \qquad p_i = \sigma(u_i) = \frac{1}{1 + e^{-u_i}}, \qquad \mathcal{L}_i = -\left\{ y_i \log p_i + (1 - y_i)\log(1 - p_i) \right\} .
$$

Each stage is a simple function of the one before it, so $\partial \mathcal{L}_i / \partial \beta$ is a product of three local derivatives. Working backwards:

**Loss with respect to the probability.** Differentiating the cross-entropy,
$$
\frac{\partial \mathcal{L}_i}{\partial p_i} = -\frac{y_i}{p_i} + \frac{1 - y_i}{1 - p_i} = \frac{p_i - y_i}{p_i(1 - p_i)} .
$$

**Probability with respect to the linear unit.** The sigmoid has the convenient
derivative
$$
\frac{\partial p_i}{\partial u_i} = \sigma'(u_i) = p_i(1 - p_i) .
$$

**Linear unit with respect to each parameter.** Every weight sees only the
input it multiplies, so there is one local derivative per edge entering $u_i$:
$$
\frac{\partial u_i}{\partial \beta_0} = 1 \quad\text{(the edge from the constant input)}, \qquad \frac{\partial u_i}{\partial \beta_1} = x_i \quad\text{(the edge from } x_i \text{)} .
$$

Multiplying the first two gives the quantity that backpropagation carries
between layers — the sensitivity of the loss to the linear unit, written
$\delta_i$:
$$
\delta_i \;=\; \frac{\partial \mathcal{L}_i}{\partial u_i} = \frac{\partial \mathcal{L}_i}{\partial p_i}\cdot\frac{\partial p_i}{\partial u_i} = \frac{p_i - y_i}{p_i(1 - p_i)} \cdot p_i(1 - p_i) = p_i - y_i .
$$

The factor $p_i(1-p_i)$ cancels exactly. This cancellation is the reason
cross-entropy is paired with a sigmoid rather than, say, squared error: the
awkward $\sigma'$ term disappears and the signal sent backwards is simply the
**prediction error** $p_i - y_i$. Multiplying by the last local derivative and
summing over observations,
$$
g(\beta) = \nabla \mathcal{L}(\beta) = \sum_{i=1}^{n} \delta_i   \begin{pmatrix} \partial u_i/\partial\beta_0 \\ \partial u_i/\partial\beta_1 \end{pmatrix} = \sum_{i=1}^{n} (p_i - y_i)\begin{pmatrix} 1 \\ x_i \end{pmatrix} = X^{\top}(p - y) = -X^{\top}(y - p),
$$

which is the expression used above and coded below. Note that $\delta_i$ is
also exactly the per-observation gradient contribution that stochastic gradient
descent samples later in this chapter.

The diagram below lays the same computation out as a network, drawn the same way
as the two-layer picture later in this chapter: values flow left to right in the
top panel (the **forward pass**), and derivatives accumulate right to left in the
bottom panel (the **backward pass**), each arrow carrying exactly one local
derivative from the list above. The intercept enters as an ordinary input $x_0
\equiv 1$, lifted to the top of its layer so that the space beneath the row is
free: there the sensitivity $\delta$ of each node sits directly under that node,
and the two parameter gradients are collected in the box beneath the fan of
$\beta$ arrows.

```{tikz}
%| label: fig-logistic-chain-rule
%| echo: false
%| fig-width: 10
%| fig-height: 10
%| engine.opts:
%|   extra.preamble: "\\usepackage{amsmath,amssymb,eucal,mathrsfs}"
%| fig-cap: "Shows how the gradient of the negative log-likelihood is assembled by the chain rule, by drawing logistic regression as the one-unit neural network it is. Logistic regression as a one-unit network: the forward pass (top) maps inputs to a linear unit, a sigmoid and a loss; the backward pass (bottom) multiplies the local derivatives, and the sigmoid factor cancels so that the signal reaching the linear unit is the prediction error $p_i - y_i$. As in the two-layer picture, each arrow carries only the local derivative of its head with respect to its tail, the sensitivity $\\delta$ of each node sits beneath that node, and the parameter gradients are collected in the box beneath the fan of weight arrows."
\usetikzlibrary{arrows.meta, backgrounds, positioning}
\begin{tikzpicture}[
  >={Stealth[length=2.5mm]},
  var/.style={circle, draw, minimum size=1.1cm, fill=gray!5, inner sep=0pt},
  loss/.style={circle, draw, minimum size=1.1cm, fill=gray!15, inner sep=0pt},
  data/.style={circle, draw=gray!80, minimum size=0.9cm, fill=white, inner sep=0pt},
  lbl/.style={font=\small},
  dlt/.style={red, font=\small, anchor=north, fill=white, inner sep=1pt, yshift=-2pt},
  grad/.style={blue, font=\small, align=left, anchor=north, fill=white,
               draw=blue!30, rounded corners, inner sep=5pt},
  arr/.style={thick, ->},
  backarr/.style={thick, ->, red},
  bbackarr/.style={thick, ->, blue}
]

% ---------------- Forward pass ----------------
% x_i, u_i, p_i and L_i sit on one straight line; the constant input x_0 is
% lifted to the top, leaving the space below the row free for the deltas.
\node (f-title) at (6, 4.0) {\textbf{forward pass: values}};
\node[var]  (fx0) at (0,    2.2) {$x_0$};
\node[var]  (fxi) at (0,    0)   {$x_i$};
\node[var]  (fu)  at (4,    0)   {$u_i$};
\node[var]  (fp)  at (8,    0)   {$p_i$};
\node[loss] (fL)  at (12,   0)   {$\mathcal{L}_i$};
\node[data] (fy)  at (12,   2.2) {$y_i$};

\node[lbl, gray, anchor=east, xshift=-3pt] at (fx0.west) {$\equiv 1$};

\draw[arr, blue] (fx0) -- (fu) node[pos=0.45, sloped, above, lbl, fill=white, inner sep=1pt] {$\beta_0$};
\draw[arr, blue] (fxi) -- (fu) node[pos=0.62, above, lbl] {$\beta_1$};
\draw[arr] (fu) -- (fp) node[midway, above, lbl] {$\sigma$};
\draw[arr] (fp) -- (fL) node[midway, above, lbl] {cross-entropy};
\draw[arr, gray] (fy) -- (fL);

% ---------------- Backward pass ----------------
\begin{scope}[yshift=-6.5cm]
\node (b-title) at (6, 4.0) {\textbf{backward pass: local derivatives}};
\node[var]  (bx0) at (0,    2.2) {$x_0$};
\node[var]  (bxi) at (0,    0)   {$x_i$};
\node[var]  (bu)  at (4,    0)   {$u_i$};
\node[var]  (bp)  at (8,    0)   {$p_i$};
\node[loss] (bL)  at (12,   0)   {$\mathcal{L}_i$};
\node[data, draw=gray!40, text=gray!40] (by) at (12,   2.2) {$y_i$};

% every arrow carries only the local derivative of its head w.r.t. its tail
\draw[bbackarr] (bu) -- (bx0) node[pos=0.45, sloped, above, lbl, fill=white, inner sep=1pt] {$\frac{\partial u_i}{\partial x_0} = \beta_0$};
\draw[bbackarr] (bu) -- (bxi) node[pos=0.38, above, lbl, fill=white, inner sep=1pt] {$\frac{\partial u_i}{\partial x_i} = \beta_1$};
\draw[backarr] (bp) -- (bu) node[midway, above=2pt, lbl, fill=white, inner sep=1pt] {$\frac{\partial p_i}{\partial u_i} = p_i(1-p_i)$};
\draw[backarr] (bL) -- (bp) node[midway, above=2pt, lbl, fill=white, inner sep=1pt] {$\frac{\partial \mathcal{L}_i}{\partial p_i} = \frac{p_i - y_i}{p_i(1-p_i)}$};

% Sensitivities (delta), beneath the bottom node of each layer
\node[dlt] at (bp.south) {$\delta_{p_i} = \frac{p_i - y_i}{p_i(1-p_i)}$};
% the red box marks the cancellation of the sigmoid factor
\node[red, font=\small, anchor=north, yshift=-2pt, draw, fill=red!5, thick,
      inner sep=5pt] at (bu.south) {$\delta_{u_i} = p_i - y_i$};
% delta at the input layer: well defined, but no parameter hangs off it
\node[dlt, align=center] at (bxi.south)
  {$\delta_{x_i} = \delta_{u_i} \, \beta_1$ \\ \textcolor{gray}{(not needed here)}};

% Parameter gradients, beneath the fan of weight arrows they belong to
\node[grad] at (2, -2.6) {
  $\dfrac{\partial \mathcal{L}_i}{\partial \beta_1} = \delta_{u_i} \, x_i$ \\[6pt]
  $\dfrac{\partial \mathcal{L}_i}{\partial \beta_0} = \delta_{u_i} \, x_0 = \delta_{u_i}$
};
\end{scope}

\end{tikzpicture}
```

The red box marks the cancellation: whatever the sigmoid contributes on the way
forward is undone on the way back, leaving $\delta_i = p_i - y_i$. In a deeper
network the same $\delta$ would be passed further left and multiplied by the
next layer's local derivative; here there is only one layer, so the recursion
stops immediately and $g(\beta) = \sum_i \delta_i\,(1, x_i)^{\top}$.

#### Implementation

```{r}
#| label: logistic-fns
#| message: false

### numerically stable log(1 + exp(u))
log1pexp <- function(u) pmax(u, 0) + log1p(exp(-abs(u)))

### negative log-likelihood, its gradient g, and the information matrix
nll_grad_info <- function(b, y, x) {
    u <- b[1] + b[2] * x
    p <- plogis(u)
    w <- p * (1 - p)
    list(
        neg_loglik = -sum(u * y - log1pexp(u)),
        grad       = -c(sum(y - p), sum(x * (y - p))),   # g(beta)
        inf        = matrix(c(sum(w), sum(x * w),
                              sum(x * w), sum(x * x * w)), 2, 2)
    )
}

neg_logp_logistic <- function(b, x, y) {
    u <- b[1] + b[2] * x
    -sum(u * y - log1pexp(u))
}
```

Because we minimise the *negative* log-likelihood, the update
$\beta^{(k+1)} = \beta^{(k)} + J^{-1}S$ becomes
$\beta^{(k+1)} = \beta^{(k)} - J^{-1} g(\beta^{(k)})$, which is
the form coded below.

```{r}
#| label: logistic-nr

mle_logistic_nr <- function(b0, no_iter, y, x, debug = FALSE) {
    out <- matrix(NA_real_, no_iter + 1, 3,
                  dimnames = list(NULL, c("beta0", "beta1", "neg_loglike")))
    b <- b0
    for (i in seq_len(no_iter + 1)) {
        q <- nll_grad_info(b, y, x)
        out[i, ] <- c(b, q$neg_loglik)
        if (debug) print(out[i, ])
        if (i > no_iter) break
        step <- tryCatch(solve(q$inf, q$grad),
                         error = function(e) rep(NA_real_, 2))
        if (any(!is.finite(step))) break
        b <- b - step
    }
    out
}
```

```{r}
#| label: fig-logistic-data
#| fig-cap: "Shows the simulated logistic-regression data used as a running example for multivariate optimization.
#|   Simulated logistic data: true success probability and observed responses."

gen_logistic_data <- function(b, n) {
    x <- sort(runif(n, -2, 2))
    p <- plogis(b[1] + b[2] * x)
    y <- rbinom(n, 1, p)
    plot(x, p, type = "l", ylim = c(0, 1), ylab = "P(y = 1)")
    points(x, y, col = y + 1)
    list(x = x, y = y)
}

set.seed(1)
data <- gen_logistic_data(c(0, 1.5), 200)
```

A start close to the truth converges in a handful of iterations:

```{r}
#| label: logistic-nr-good
round(mle_logistic_nr(c(0, 3), 8, data$y, data$x), 6)
```

The doubling of correct digits predicted by the quadratic-convergence result is
visible in the final rows.

#### Comparison with `nlm` and `glm`

```{r}
#| label: logistic-nlm
logit_nlm <- nlm(neg_logp_logistic, c(0, 0),
                 x = data$x, y = data$y, hessian = TRUE)
logit_nlm$estimate
sqrt(diag(solve(logit_nlm$hessian)))   # asymptotic standard errors
```

```{r}
#| label: logistic-glm
summary(glm(y ~ x, family = binomial(), data = data))$coefficients
```

The three routes agree to printing accuracy. `glm` is doing the same arithmetic
as `mle_logistic_nr`: for a GLM the Newton–Raphson step can be rearranged as a
weighted least-squares fit of the working response
$u_i + (y_i - p_i)/\{p_i(1-p_i)\}$ on $X$ with weights $p_i(1-p_i)$, giving
**iteratively reweighted least squares (IRLS)**. This is not a different
algorithm but a different arrangement of the same one, and it explains why
`glm` reports "Fisher Scoring iterations" in its output.

### Enhancements to Multivariate Newton-Raphson Methods {#sec-divergence}

Newton–Raphson can diverge even on a well-behaved problem. Far from
$\hat\theta$ the quadratic approximation is unreliable, and $J_k = J(\theta^{(k)})$
may be nearly singular or not positive definite, so the raw step
$\delta^{(k)} = -J_k^{-1}g_k$, with $g_k = g(\theta^{(k)})$, can be enormous or
fail to point downhill at all. The
standard enhancements all keep the Newton idea and add a guard.

* **Step halving (line search).** Move $\theta^{(k+1)} = \theta^{(k)} + s\,\delta^{(k)}$,
starting at $s = 1$ and halving until $\mathcal{L}$ decreases by enough —
the **Armijo condition**
$\mathcal{L}(\theta^{(k)} + s\,\delta^{(k)}) \le \mathcal{L}(\theta^{(k)}) + c_1 s\,g_k^\top\delta^{(k)}$
with a tiny $c_1$ (say $10^{-4}$). Near the minimum $s = 1$ passes on the first
try, so fast local convergence is untouched.

  The condition asks for more than a lower value of $\mathcal{L}$. Written as
  $$
  \underbrace{\mathcal{L}(\theta^{(k)} + s\,\delta^{(k)}) - \mathcal{L}(\theta^{(k)})}_{\text{actual decrease}}
  \;\le\; c_1 \cdot \underbrace{s\,g_k^\top\delta^{(k)}}_{\text{decrease predicted by the tangent plane}},
  $$
  it says that the decrease must be at least a fixed fraction of what the slope
  promises for a step of that length. (When $J_k$ is positive definite,
  $g_k^\top\delta^{(k)} = -g_k^\top J_k^{-1} g_k < 0$, so the right-hand side is a
  genuine decrease.) Asking only for *some* decrease is not enough, because the
  gains can shrink so fast that the iterates stall at a point that is not a
  minimum. For $\mathcal{L}(\theta) = \theta^2$, the sequence
  $\theta^{(k)} = (-1)^k(1 + 2^{-k})$ lowers $\mathcal{L}$ at every step and moves
  downhill each time, yet it bounces towards $\pm 1$, where the gradient is
  $\pm 2$, never $0$. With $\delta = -g = -2\theta$ the Armijo condition reduces to
  $s \le 1 - c_1$, and the overshooting steps need $s \to 1$, so they are
  rejected. Armijo only forbids steps that are too long for the gain they
  deliver. Starting at $s = 1$ and halving only when the test fails is what
  keeps the accepted steps from being needlessly short.

* **Ridging (Levenberg–Marquardt).** Replace $J_k$ by $J_k + \tau\mathbb{I}$, with
$\tau \ge 0$ just large enough to make it positive definite. This repairs
indefiniteness and shortens the step; as $\tau$ grows the direction rotates from
Newton toward plain gradient descent, and $\tau = 0$ wherever $J_k$ is already
well behaved. This is roughly what `nlm` does internally.

  A simple rule takes $\tau = \max\{0,\ \epsilon - \lambda_{\min}(J_k)\}$,
  which lifts the smallest eigenvalue of $J_k$ to a small positive $\epsilon$.
  A modified Cholesky factorisation achieves the same without an
  eigendecomposition. Since $\delta^{(k)}(\tau) = -(J_k + \tau\mathbb{I})^{-1}g_k$
  has length decreasing in $\tau$, ridging also acts as a step-length control,
  and with a positive definite matrix the direction satisfies
  $g_k^\top\delta^{(k)} < 0$, so a line search along it is guaranteed to find a
  decrease. A related guard for GLMs is **Fisher scoring**: the expected
  information $I(\theta) = X^\top W X$ is positive semi-definite at every
  $\theta$, and positive definite whenever $X$ has full column rank and all
  weights are positive, so the ridge is rarely needed at all.

* **Trust regions.** Minimise the quadratic approximation only within
$\lVert\theta - \theta^{(k)}\rVert \le \Delta_k$, growing or shrinking $\Delta_k$
according to how well the quadratic predicted the realised change. This is
`nlminb` and the `PORT` routines.

  The quality of the prediction is measured by the ratio
  $$
  \rho_k = \frac{\mathcal{L}(\theta^{(k)} + \delta) - \mathcal{L}(\theta^{(k)})}
                {g_k^\top\delta + \tfrac12\,\delta^\top J_k\,\delta},
  $$
  actual decrease over the decrease predicted by the quadratic model. A typical rule
  rejects the step if $\rho_k$ is small or negative and shrinks $\Delta_k$
  (say by a factor of four); it accepts the step and keeps $\Delta_k$ when
  $\rho_k$ is moderate; and it doubles $\Delta_k$ when $\rho_k \ge 3/4$ and the
  step reached the boundary. The constrained subproblem has the solution
  $\delta = -(J_k + \tau\mathbb{I})^{-1}g_k$ for some $\tau \ge 0$, so a trust
  region is ridging with $\tau$ chosen by the step length rather than the other
  way round. Unlike a line search, it first fixes how far to go and then
  chooses the direction. That makes it robust when $J_k$ is indefinite:
  at a saddle point, where the Newton step points the wrong way, it still
  moves downhill.

* **Better parameterisation and starting values.** Often the cheapest fix.
Centring and scaling the covariates improves the conditioning of $J$.
Parameters with natural bounds are better optimised on an unconstrained scale
— $\log\sigma$ rather than $\sigma$, $\operatorname{logit} p$ rather than $p$ —
which removes steps that leave the parameter space and usually makes $\ell$
closer to quadratic. Good starting values come from a simpler fit: method of
moments estimates, a least-squares fit, or the null model with an intercept
only. `glm` sidesteps the choice altogether: it starts from fitted means rather
than coefficients, taking $\mu_i^{(0)} = (y_i + 0.5)/2$ for binary data, which
are always inside $(0, 1)$, and it therefore never begins in the flat region
where the weights $p_i(1-p_i)$ vanish.

Combining the first two gives **safeguarded Newton**: at each iteration form a
positive definite $\tilde J_k$, solve $\tilde J_k\,\delta^{(k)} = -g_k$, then
backtrack on $s$. Both guards switch themselves off near a well-behaved
maximum, so the method keeps Newton's fast local convergence while still making
progress from remote starting values.

#### Shinylive App for Safeguarded Newton-Raphson {#sec-nr-divergence-app}

There is a [shinylive app](../shinyliveapps_compstat/app_nr-divergence.html) to show how the full Newton-Raphson step diverges on a logistic-regression likelihood started far from the maximum, and how backtracking on the step length rescues the same starting value.

## Example: Poisson Regression with Neural Networks

Logistic regression was a one-unit network, and its Hessian was small enough to
invert. Adding a single hidden layer breaks that comfort, and shows why the
methods in this section exist.

Let each observation carry $p = 3$ covariates $x_i = (x_{i1}, x_{i2}, x_{i3})$ and a
count response $y_i$. Put $J = 3$ hidden units $h_{i1}, h_{i2}, h_{i3}$ between the
covariates and the mean, and use a log link so that the mean stays positive:
$$a_{ij} = \sum_{k=0}^{p} w_{jk}\, x_{ik}, \qquad h_{ij} = \tanh(a_{ij}), \qquad \eta_i = \sum_{j=0}^{J} v_j\, h_{ij}, \qquad \lambda_i = e^{\eta_i},$$
where $x_{i0} \equiv 1$ and $h_{i0} \equiv 1$ are constant inputs, so that the
intercept of each hidden unit is the weight $w_{j0}$ and the intercept of the
linear predictor is the weight $v_0$,
with $y_i \mid x_i \sim \mathrm{Poisson}(\lambda_i)$. Dropping the $\log y_i!$ term,
which carries no parameters, the negative log-likelihood is
$$\mathcal{L}(\theta) = \sum_{i=1}^{n} \mathcal{L}_i(\theta), \qquad \mathcal{L}_i(\theta) = \lambda_i - y_i \eta_i ,$$
where $\theta$ collects the $J(p+1) = 12$ input weights $w_{jk}$, $k = 0, \ldots, p$,
and the $J + 1 = 4$ output weights $v_j$, $j = 0, \ldots, J$ — sixteen parameters
in all.

The intercepts matter: without $w_{j0}$ every hidden unit would pass through the
origin, and without $v_0$ the level of $\lambda$ could not be shifted freely.
Carrying them as ordinary weights on constant inputs means they need no separate
symbol and no separate rule: $w_{j0}$ and $v_0$ are differentiated exactly like
every other weight, and their local derivative happens to be the tail value $1$.

Three features make this a natural home for gradient-only methods. The
score equations have no closed-form solution. $\mathcal{L}$ is **not convex** —
permuting the hidden units leaves it unchanged, so every minimum is one of
$J!$ equivalent copies, and the Hessian is indefinite over much of the space,
which is exactly where a raw Newton step misbehaves. And while a $12 \times 12$
Hessian is still cheap here, its size grows as $(Jp + J)^2$: widen the layer or
the covariate set and the $O(p^3)$ solve becomes the bottleneck long before the
gradient does.

### What the model can represent

With a single covariate the model is easy to see. Each `tanh` unit switches on
at a location set by its intercept $w_{j0}$ and at a rate set by its weight
$w_{j1}$, and enters the linear predictor scaled by its output weight $v_j$. It
is those *scaled* units $v_j h_j(x)$ that are worth plotting: they add up, with
$v_0$, to $\eta(x)$, and exponentiating $\eta$ gives a $\lambda(x)$ with a hump —
a shape a log-linear Poisson regression, whose $\log\lambda$ is a straight line,
cannot produce at all.

```{r}
#| label: fig-nn-poisson-model
#| fig-width: 10
#| fig-height: 4.4
#| fig-cap: "Shows what a hidden layer buys in a Poisson regression, by displaying the three scaled hidden units and the non-monotone mean curve they combine into, together with counts simulated from it.
#|   A one-covariate Poisson network with three tanh units: the scaled activations $v_j h_j(x)$ and the linear predictor $\\eta(x)$ they sum to (left), and the resulting mean $\\lambda(x) = e^{\\eta(x)}$ with simulated counts (right), a shape no log-linear model could produce."
set.seed(7)

w_true  <- c(3.0, 3.0, 1.0)     # input weight of each hidden unit
w0_true <- c(1.5, -1.5, 0.0)    # hidden intercepts: where each unit switches on
v_true  <- c(1.2, -1.2, 0.3)    # output weights
v0_true <- -0.3                 # output intercept

lambda_nn <- function(x, w, w0, v, v0)
  as.vector(exp(v0 + tanh(sweep(outer(x, w), 2, w0, "+")) %*% v))

xg <- seq(-3, 3, length.out = 400)
H  <- tanh(sweep(outer(xg, w_true), 2, w0_true, "+"))
VH <- sweep(H, 2, v_true, "*")             # the scaled units v_j h_j(x)
eg <- v0_true + rowSums(VH)                # eta(x), what they add up to

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))

matplot(xg, VH, type = "l", lty = 1, lwd = 2, col = c(1, 2, 4),
        ylim = range(VH, eg) + c(0, 1.3),   # headroom for the legend
        xlab = TeX(r"($x$)"), ylab = TeX(r"($v_j h_j(x)$)"),
        main = TeX(r"(the three scaled hidden units)"))
lines(xg, eg, lwd = 2.5, lty = 2, col = "grey30")
abline(h = 0, lty = 3)
legend("topleft", bty = "n", lty = c(1, 1, 1, 2), lwd = 2,
       col = c(1, 2, 4, "grey30"), cex = 0.8,
       legend = TeX(c(sprintf(r"($v_%d h_%d$: $w = %.1f$, $w_0 = %.1f$, $v = %.1f$)",
                              1:3, 1:3, w_true, w0_true, v_true),
                      r"($\eta(x) = v_0 + \sum_j v_j h_j(x)$)")))

## simulate counts from the curve
n  <- 150
xi <- runif(n, -3, 3)
yi <- rpois(n, lambda_nn(xi, w_true, w0_true, v_true, v0_true))

plot(xg, lambda_nn(xg, w_true, w0_true, v_true, v0_true), type = "l", lwd = 2,
     ylim = c(0, max(yi) + 0.5),
     xlab = TeX(r"($x$)"), ylab = TeX(r"($\lambda(x)$ and $y$)"),
     main = TeX(r"(mean curve and simulated counts)"))
points(xi, yi, pch = 19, col = adjustcolor(2, 0.45), cex = 0.7)
legend("topright", bty = "n", lwd = c(2, NA), pch = c(NA, 19), col = c(1, adjustcolor(2, 0.6)),
       legend = TeX(c(r"($\lambda(x)$)", r"($y_i \sim Poisson(\lambda(x_i))$)")))
```

The counts scatter around the curve with variance equal to the mean, so they
bunch near zero in the flat regions and spread out under the peak.

### The gradient by backpropagation

The chain rule applies exactly as in the logistic case, only with one more layer
to traverse. Working backwards from the loss for a single observation:

**Loss to the linear predictor.** The two local derivatives at the output are
$$\frac{\partial \mathcal{L}_i}{\partial \lambda_i} = 1 - \frac{y_i}{\lambda_i}, \qquad \frac{\partial \lambda_i}{\partial \eta_i} = \lambda_i ,$$
whose product is again a bare prediction error, with the $\lambda_i$ cancelling
just as $p_i(1-p_i)$ did for the sigmoid:
$$\delta_i \;=\; \frac{\partial \mathcal{L}_i}{\partial \eta_i} = \left(1 - \frac{y_i}{\lambda_i}\right)\lambda_i = \lambda_i - y_i .$$

**Output layer.** Since $\partial \eta_i / \partial v_j = h_{ij}$ and
$\partial \eta_i / \partial v_0 = h_{i0} = 1$,
$$\frac{\partial \mathcal{L}_i}{\partial v_j} = \delta_i\, h_{ij}, \qquad \frac{\partial \mathcal{L}_i}{\partial v_0} = \delta_i .$$

**Into the hidden layer.** The path from $a_{ij}$ to the loss runs through
$h_{ij}$ and then $\eta_i$, contributing $\partial \eta_i/\partial h_{ij} = v_j$ and
$\partial h_{ij}/\partial a_{ij} = 1 - h_{ij}^2$ (the `tanh` derivative). Taking them
one at a time gives a sensitivity at each node on the way,
$$\delta_{ij}^{h} \;=\; \frac{\partial \mathcal{L}_i}{\partial h_{ij}} = \delta_i\, v_j , \qquad \delta_{ij}^{a} \;=\; \frac{\partial \mathcal{L}_i}{\partial a_{ij}} = \delta_{ij}^{h}\left(1 - h_{ij}^{2}\right) .$$

**Input layer.** Finally $\partial a_{ij}/\partial w_{jk} = x_{ik}$ for every
$k$, including $\partial a_{ij}/\partial w_{j0} = x_{i0} = 1$, so
$$\frac{\partial \mathcal{L}_i}{\partial w_{jk}} = \delta_{ij}^{a}\, x_{ik}, \qquad \frac{\partial \mathcal{L}_i}{\partial w_{j0}} = \delta_{ij}^{a} .$$

Each intercept derivative is the sensitivity at its own unit, unmultiplied. Splitting
at $a_{ij}$ is also what makes the computation efficient: $\delta_{ij}^{a}$ is
formed once and reused for all $p$ weights entering unit $j$.

### Two kinds of arrow

Every arrow in the network is one of two kinds, and the distinction organises the
whole computation.

**Arrows that carry a weight**, drawn blue below, are the ones forming a linear
combination: $x_{ik} \to a_{ij}$ carries $w_{jk}$, the constant input
$x_{i0} \to a_{ij}$ carries $w_{j0}$, $h_{ij} \to \eta_i$ carries $v_j$, and the
constant hidden unit $h_{i0} \to \eta_i$ carries $v_0$. *Every
parameter of the model sits on one of these arrows*, and these are the only
derivatives we ultimately want.

**Arrows that carry no weight** are the fixed transformations: $a_{ij} \to h_{ij}$
is `tanh`, $\eta_i \to \lambda_i$ is $\exp$, $\lambda_i \to \mathcal{L}_i$ is the Poisson
loss. Nothing is estimated on them, so they contribute no gradient component of
their own; they only pass sensitivity along.

Backpropagation computes one number per node — the sensitivity of the loss to
that node,
$$\delta_z = \frac{\partial \mathcal{L}_i}{\partial z},$$
obtained by multiplying the local derivatives along the arrows from the loss back
to $z$. Once $\delta$ is known at a node, the edge rule of @sec-computation-graph
makes the weights feeding it free:

> **The gradient of the weight on a blue arrow $u \to z$ is $\delta_z$ times the
> value at the arrow's tail:**
>.
$$> \frac{\partial \mathcal{L}_i}{\partial \omega_{u \to z}} = \delta_z \cdot u . >.$$

The reason is immediate: $z$ is a linear combination, so it depends on $\omega$
only through the single term $\omega\, u$, giving $\partial z/\partial \omega = u$.
For this network, with $\delta_{\eta} = \lambda_i - y_i$,
$\delta_{h_j} = \delta_{\eta}\, v_j$ and $\delta_{a_j} = \delta_{h_j}(1 - h_{ij}^2)$,
$$\frac{\partial \mathcal{L}_i}{\partial v_j} = \delta_{\eta}\, h_{ij}, \qquad \frac{\partial \mathcal{L}_i}{\partial v_0} = \delta_{\eta} \cdot 1, \qquad \frac{\partial \mathcal{L}_i}{\partial w_{jk}} = \delta_{a_j}\, x_{ik}, \qquad \frac{\partial \mathcal{L}_i}{\partial w_{j0}} = \delta_{a_j} \cdot 1 .$$

The intercept gradients are the same expression with a tail value of $1$, which
is the whole content of "an intercept is a weight on a constant input".

This also explains the one label that might look out of place. Along the blue
arrow $h_{ij} \to \eta_i$ the local derivative is
$$\frac{\partial \eta_i}{\partial h_{ij}} = v_j ,$$
and it plays **no part in $v_j$'s own gradient** — a weight's gradient never
involves that weight. Its job is to carry $\delta$ from $\eta_i$ back to $h_{ij}$,
so $v_j$ appears in the gradient of every weight *behind* unit $j$ and in none at
or after it. In a deeper network this is the only kind of factor that
accumulates: a path from the loss to an early weight is a long product of such
node-to-node derivatives, closed off at the end by one tail value.

Summing over observations gives the full gradient $g(\theta)$. This is
backpropagation in miniature: one forward sweep stores $h$, $\eta$ and $\lambda$;
one backward sweep turns $\delta_i$ into $\delta_{ij}^{h}$, then $\delta_{ij}^{a}$,
reading off every partial derivative on the way. The cost is a small multiple of one function
evaluation, no matter how many parameters there are — which is precisely why
first-order methods scale where Newton does not.

The diagram below is the logistic picture with a layer inserted, and it is drawn
twice: once for the forward pass and once for the backward pass, so that neither
set of arrows has to share an edge with the other. The constant inputs appear as
nodes of their own — $x_0$ at the input layer and $h_0$ at the hidden layer —
drawn at the *top* of their layer so that the space below each layer stays free.
Every parameter is then a $w$ or a $v$ sitting on an ordinary edge, the
intercepts $w_{j0}$ and $v_0$ included.

The pre-activation $a_j$ is drawn as a node of its own, so that every arrow
joins two quantities that actually appear in the picture and the `tanh` step gets
an edge to sit on. The backward panel traces one **continuous path** from the
loss to a single weight, $L \to \lambda \to \eta \to h_j \to a_j \to w_{jk}$, in bold. Multiplying the four
local derivatives along it gives that weight's partial derivative; every other
weight is reached by a path of the same shape, drawn faint. Only the highlighted
edges are labelled — the rest carry the same formulas with different indices.

Each arrow in the backward panel carries **only** its local derivative. The
sensitivity $\delta$ of each node is written directly beneath that node (including
$\delta_{x_k}$ at the input layer, which is well defined but never needed, since no
parameter hangs off an input). The parameter gradients — each one a $\delta$ at an
arrow's head times the value at its tail — are collected in the boxes beneath the
fan of arrows whose weights they belong to.


```{tikz}
%| label: fig-nn-poisson-backprop
%| echo: false
%| fig-width: 10
%| fig-height: 11.5
%| engine.opts:
%|   extra.preamble: '\usepackage{amsmath,amssymb,eucal,mathrsfs}'
%| fig-cap: "This figure shows the way that backpropagation assembles the gradient of a two-layer Poisson network, by labelling each edge with its local derivative. The network is oriented left-to-right, with the nodes of each layer stacked vertically: the forward pass (top) carries values rightward, and the backward pass (bottom) carries local derivatives leftward. Each arrow carries only the local derivative of its head with respect to its tail; the sensitivity $\\delta$ of each node sits directly beneath that node, and the parameter gradients — obtained by multiplying the $\\delta$ at the arrow's head by the value at its tail — are collected in the boxes beneath the corresponding fan of weight arrows."

\usetikzlibrary{arrows.meta, backgrounds, positioning}
\begin{tikzpicture}[
  >={Stealth[length=2.5mm]},
  var/.style={circle, draw, minimum size=1.1cm, fill=gray!5, inner sep=0pt},
  loss/.style={circle, draw, minimum size=1.1cm, fill=gray!15, inner sep=0pt},
  data/.style={circle, draw=gray!80, minimum size=0.9cm, fill=white, inner sep=0pt},
  lbl/.style={font=\small},
  dlt/.style={red, font=\small, anchor=north, fill=white, inner sep=1pt, yshift=-2pt},
  grad/.style={blue, font=\small, align=left, anchor=north, fill=white,
               draw=blue!30, rounded corners, inner sep=5pt},
  arr/.style={thick, ->},
  backarr/.style={thick, ->, red},
  bbackarr/.style={thick, ->, blue},
  faintw/.style={->, blue, opacity=0.25, line width=0.5pt},
  faintt/.style={->, black, opacity=0.25, line width=0.5pt},
  faintbw/.style={->, blue, opacity=0.25, line width=0.5pt},
  faintbt/.style={->, red, opacity=0.25, line width=0.5pt}
]

% ==========================================
% Forward pass (Values flowing RIGHT)
% ==========================================
\begin{scope}
\node (f-title) at (8.25, 6.4) {\textbf{forward pass: values}};

% Layers (Left to Right; nodes stacked top to bottom).
% The constant inputs x_0 and h_0 sit on TOP of their layers, so the space
% below each layer is free for the sensitivities and the gradient boxes.
\node[var] (fx0) at (0,  4.4) {$x_0$};
\node[var] (fx1) at (0,  2.2) {$x_1$};
\node[var] (fx2) at (0,  0)   {$x_2$};
\node[var] (fx3) at (0, -2.2) {$x_3$};

\node[var] (fa1) at (4,  2.2) {$a_1$};
\node[var] (fa2) at (4,  0)   {$a_2$};
\node[var] (fa3) at (4, -2.2) {$a_3$};

\node[var] (fh0) at (8,  4.4) {$h_0$};
\node[var] (fh1) at (8,  2.2) {$h_1$};
\node[var] (fh2) at (8,  0)   {$h_2$};
\node[var] (fh3) at (8, -2.2) {$h_3$};

\node[var]  (feta) at (11.5, 0)    {$\eta$};
\node[var]  (flam) at (14,   0)    {$\lambda$};
\node[loss] (fL)   at (16.5, 0)    {$\mathcal{L}_i$};
\node[data] (fy)   at (16.5, -2.2) {$y_i$};

\node[lbl, gray, anchor=east, xshift=-3pt] at (fx0.west) {$\equiv 1$};
\node[lbl, gray, anchor=east, xshift=-3pt] at (fh0.west) {$\equiv 1$};

% Faint connections (right)
\foreach \i in {1,2,3} {
  \foreach \j in {1,2,3} {
    \ifnum\i=1 \ifnum\j=2 \else \draw[faintw] (fx\i) -- (fa\j); \fi \else \draw[faintw] (fx\i) -- (fa\j); \fi
  }
  \ifnum\i=2 \else \draw[faintw] (fx0) -- (fa\i); \fi
}
\foreach \i in {1,2,3} {
  \ifnum\i=2 \else \draw[faintt] (fa\i) -- (fh\i); \fi
}
\foreach \i in {1,2,3} {
  \ifnum\i=2 \else \draw[faintw] (fh\i) -- (feta); \fi
}
\draw[faintw] (fh0) -- (feta);

% Bold connections (right)
\draw[arr, blue] (fx1) -- (fa2) node[pos=0.35, above=2pt, lbl] {$w_{jk}$};
\draw[arr, blue] (fx0) -- (fa2) node[pos=0.3, sloped, above, lbl, fill=white, inner sep=1pt] {$w_{j0}$};
\draw[arr] (fa2) -- (fh2) node[midway, above, lbl] {$\tanh$};
\draw[arr, blue] (fh2) -- (feta) node[midway, above, lbl] {$v_j$};
\draw[arr, blue] (fh0) -- (feta) node[pos=0.3, sloped, above, lbl, fill=white, inner sep=1pt] {$v_0$};
\draw[arr] (feta) -- (flam) node[midway, above, lbl] {$e^\eta$};
\draw[arr] (flam) -- (fL) node[midway, above, lbl] {Poisson};
\draw[arr, gray] (fy) -- (fL);
\end{scope}

% ==========================================
% Backward pass (Derivatives flowing LEFT)
% ==========================================
\begin{scope}[yshift=-11cm]
\node (b-title) at (8.25, 6.4) {\textbf{backward pass: local derivatives}};

% Layers (Left to Right; nodes stacked top to bottom)
\node[var] (bx0) at (0,  4.4) {$x_0$};
\node[var] (bx1) at (0,  2.2) {$x_1$};
\node[var] (bx2) at (0,  0)   {$x_2$};
\node[var] (bx3) at (0, -2.2) {$x_3$};

\node[var] (ba1) at (4,  2.2) {$a_1$};
\node[var] (ba2) at (4,  0)   {$a_2$};
\node[var] (ba3) at (4, -2.2) {$a_3$};

\node[var] (bh0) at (8,  4.4) {$h_0$};
\node[var] (bh1) at (8,  2.2) {$h_1$};
\node[var] (bh2) at (8,  0)   {$h_2$};
\node[var] (bh3) at (8, -2.2) {$h_3$};

\node[var]  (beta) at (11.5, 0)    {$\eta$};
\node[var]  (blam) at (14,   0)    {$\lambda$};
\node[loss] (bL)   at (16.5, 0)    {$\mathcal{L}_i$};
\node[data, text=gray!40, draw=gray!40] (by) at (16.5, -2.2) {$y_i$};

% Faint connections (left)
\foreach \i in {1,2,3} {
  \foreach \j in {1,2,3} {
    \ifnum\i=1 \ifnum\j=2 \else \draw[faintbw] (ba\j) -- (bx\i); \fi \else \draw[faintbw] (ba\j) -- (bx\i); \fi
  }
  \ifnum\i=2 \else \draw[faintbw] (ba\i) -- (bx0); \fi
}
\foreach \i in {1,2,3} {
  \ifnum\i=2 \else \draw[faintbt] (bh\i) -- (ba\i); \fi
}
\foreach \i in {1,2,3} {
  \ifnum\i=2 \else \draw[faintbw] (beta) -- (bh\i); \fi
}
\draw[faintbw] (beta) -- (bh0);

% Bold connections (left): each arrow carries ONLY the local derivative
\draw[bbackarr] (ba2) -- (bx1) node[pos=0.32, sloped, above, lbl, fill=white, inner sep=1pt] {$\frac{\partial a_j}{\partial x_k} = w_{jk}$};
\draw[bbackarr] (ba2) -- (bx0) node[pos=0.70, sloped, above, lbl, fill=white, inner sep=1pt] {$\frac{\partial a_j}{\partial x_0} = w_{j0}$};
\draw[backarr] (bh2) -- (ba2) node[midway, above, lbl] {$\frac{\partial h_j}{\partial a_j} = 1 - h_j^2$};
\draw[bbackarr] (beta) -- (bh2) node[midway, above, lbl] {$\frac{\partial \eta}{\partial h_j} = v_j$};
\draw[bbackarr] (beta) -- (bh0) node[pos=0.62, sloped, above, lbl, fill=white, inner sep=1pt] {$\frac{\partial \eta}{\partial h_0} = v_0$};
\draw[backarr] (blam) -- (beta) node[midway, above, lbl] {$\frac{\partial \lambda}{\partial \eta} = \lambda$};
\draw[backarr] (bL) -- (blam) node[midway, above=4pt, lbl, fill=white, inner sep=1pt] {$\frac{\partial \mathcal{L}_i}{\partial \lambda} = 1 - \frac{y_i}{\lambda}$};

% Sensitivities (delta), one per layer, beneath the BOTTOM node of that layer
\node[dlt] at (blam.south) {$\delta_\lambda = 1 - \frac{y_i}{\lambda}$};
\node[dlt] at (beta.south) {$\delta_\eta = \lambda - y_i$};
\node[dlt] at (bh3.south)  {$\delta_{h_j} = \delta_\eta \, v_j$};
\node[dlt] at (ba3.south)  {$\delta_{a_j} = \delta_{h_j}(1 - h_j^2)$};
% delta at the input layer: well defined, but no parameter hangs off it
\node[dlt, align=center] at (bx3.south)
  {$\delta_{x_k} = \sum_j \delta_{a_j} w_{jk}$ \\ \textcolor{gray}{(not needed here)}};

% Parameter gradients, beneath the fan of weight arrows they belong to
\node[grad] at (2, -4.5) {
  $\dfrac{\partial \mathcal{L}_i}{\partial w_{jk}} = \delta_{a_j} \, x_k$ \\[6pt]
  $\dfrac{\partial \mathcal{L}_i}{\partial w_{j0}} = \delta_{a_j} \, x_0 = \delta_{a_j}$
};
\node[grad] at (9.75, -4.5) {
  $\dfrac{\partial \mathcal{L}_i}{\partial v_j} = \delta_\eta \, h_j$ \\[6pt]
  $\dfrac{\partial \mathcal{L}_i}{\partial v_0} = \delta_\eta \, h_0 = \delta_\eta$
};
\end{scope}

\end{tikzpicture}

```

### A General Description of Gradient of Neural Networks {#sec-backprop-dag}

Neither rule of @sec-computation-graph mentions depth, so the same pair describes
a network of any size. Write a feed-forward network of $L$ layers as
$$h^{(0)} = x, \qquad a^{(l)} = W^{(l)} h^{(l-1)}, \qquad h^{(l)} = f\!\left(a^{(l)}\right), \qquad l = 1, \ldots, L ,$$
with a constant entry folded into each $h^{(l-1)}$ so that the intercepts are
ordinary columns of $W^{(l)}$, exactly as $x_0$ and $h_0$ are in
@fig-nn-poisson-backprop.

**The backward pass has the same shape as the forward pass.** Applying the node
rule to a whole layer at once turns the sum over consumers into a matrix–vector
product with the transposed weight matrix:
$$\delta_{h^{(l-1)}} = W^{(l)\top} \delta_{a^{(l)}}, \qquad \delta_{a^{(l)}} = \delta_{h^{(l)}} \odot f'\!\left(a^{(l)}\right) ,$$
where $\odot$ is the elementwise product. Set the two sweeps side by side and
they are the same two operations:

|             | forward                                  | backward                                                       |
|-------------|------------------------------------------|----------------------------------------------------------------|
| combine     | $a^{(l)} = W^{(l)} h^{(l-1)}$            | $\delta_{h^{(l-1)}} = W^{(l)\top} \delta_{a^{(l)}}$             |
| transform   | $h^{(l)} = f\!\left(a^{(l)}\right)$      | $\delta_{a^{(l)}} = \delta_{h^{(l)}} \odot f'\!\left(a^{(l)}\right)$ |

A layer of the backward pass takes a linear combination of the quantities in the
layer just processed and then rescales each entry — the same two steps the
forward pass performs on values. Three things differ. The matrix is transposed,
so the linear combination runs across the arrows *entering* a unit rather than
those leaving it. The elementwise step multiplies by $f'(a^{(l)})$ instead of
applying $f$, which makes the backward pass **linear** in $\delta$ once the
activations are fixed. And the sweep runs from $l = L$ down to $l = 1$.

**Starting the recursion.** The sweep needs one number to begin with, the
sensitivity at the output. For the Poisson network of this section that is
$$\delta_{\eta_i} = \frac{\partial \mathcal{L}_i}{\partial \eta_i} = \lambda_i - y_i ,$$
and for the logistic model of @sec-computation-graph it is $p_i - y_i$. In both
cases the derivative of the output activation cancels against the derivative of
the loss, which is why the seed is a bare prediction error.

**Reading off the gradients.** Once $\delta_{a^{(l)}}$ is known, the edge rule
delivers the whole weight matrix at once as an outer product,
$$\frac{\partial \mathcal{L}_i}{\partial W^{(l)}} = \delta_{a^{(l)}} \, h^{(l-1)\top} ,$$
one entry $\delta_{a^{(l)}_j} h^{(l-1)}_k$ per weight $w^{(l)}_{jk}$: the
sensitivity at the arrow's head times the value at its tail. The label
$\delta_{x_k} = \sum_j \delta_{a_j} w_{jk}$ in @fig-nn-poisson-backprop is the node
rule at $l = 1$. It is marked unneeded there because no parameter hangs off an
input, but in a deeper network it is precisely the $\delta$ handed to the layer
below.

The complete gradient for one observation is therefore four steps: sweep forward
storing $a^{(l)}$ and $h^{(l)}$; seed $\delta$ at the output; sweep backward
through the pair of recursions above; and form one outer product per layer along
the way. Summing over observations gives $g(\theta)$.

**Cost.** The two sweeps consist of the same $L$ matrix–vector products, one per
layer, with the same matrices — $W^{(l)}$ going up and $W^{(l)\top}$ coming down.
A gradient therefore costs a small constant multiple of one evaluation of
$\mathcal{L}_i$, about $O(L W^{2})$ for width $W$, or one multiply–add per
parameter per sweep, *independent of how many parameters there are*. This is
worth contrasting with the form in which the chain rule is usually written.
Expanding $\partial \mathcal{L}_i / \partial w$ as a sum over paths from the loss to
$w$ produces about $W^{L}$ terms, exponential in depth. The recursion computes
the same quantity by visiting each *edge* once: a node's $\delta$ is formed once
and reused by every node that feeds it, so shared sub-paths are never
recomputed. In the language of algorithms the graph is a DAG and the backward
sweep is dynamic programming over it in reverse topological order, which stands
to path enumeration as a memoized recursion stands to a naive one.

**Memory.** The backward sweep consumes what the forward sweep produced —
$h^{(l-1)}$ for the outer product, $f'(a^{(l)})$ for the elementwise step — so the
activations have to be stored, at a cost proportional to the total number of
nodes. That is why the diagrams above have a forward panel at all, and it is the
cost that gradient checkpointing trades back for recomputation in very deep
models.

**Two extensions.** Weight sharing needs no new machinery: when one parameter
sits on many edges, as in a convolutional or recurrent network, its gradient is
the sum of the edge rule over all of them, which is the node rule again. And the
direction of the sweep is a choice. Propagating $\partial z / \partial \theta_k$
*forward* is equally correct but costs one sweep per parameter, whereas sweeping
backward costs one sweep per output and there is a single scalar output here.
That asymmetry is what makes a full gradient affordable for models with millions
of parameters, and it is the property the rest of this chapter relies on.

Backpropagation is easy to get subtly wrong, so it is worth testing against
finite differences before handing the gradient to an optimizer — the same
comparison made for the univariate case earlier in this chapter.

```{r}
#| label: nn-poisson-grad
## pack/unpack the 16 parameters: W (J x p), w0 (J), v (J), v0 (1)
J <- 3

nn_pois <- function(theta, X, y) {
  p <- ncol(X)
  W  <- matrix(theta[1:(J * p)], J, p)
  w0 <- theta[J * p + 1:J]
  v  <- theta[J * p + J + 1:J]
  v0 <- theta[J * p + 2 * J + 1]

  A   <- sweep(X %*% t(W), 2, w0, "+")     # n x J  linear units
  H   <- tanh(A)                           # n x J  activations
  eta <- as.vector(v0 + H %*% v)
  lam <- exp(eta)

  nll <- sum(lam - y * eta)                # dropping log(y!)

  d   <- lam - y                           # n      dL/deta
  dv  <- as.vector(t(H) %*% d)             # J      dL/dv_j
  dv0 <- sum(d)                            # 1      dL/dv_0
  dh  <- outer(d, v) * (1 - H^2)           # n x J  dL/da_ij
  dW  <- t(dh) %*% X                       # J x p  dL/dw_jk
  dw0 <- colSums(dh)                       # J      dL/dw_j0
  list(nll = nll, grad = c(as.vector(dW), dw0, dv, dv0))
}

set.seed(3)
n <- 40; p <- 3
X <- matrix(rnorm(n * p), n, p)
theta0 <- rnorm(J * p + 2 * J + 1, sd = 0.5)
## counts simulated from the network at theta0
ytrue <- rpois(n, local({
  W <- matrix(theta0[1:(J * p)], J, p); w0 <- theta0[J * p + 1:J]
  v <- theta0[J * p + J + 1:J];        v0 <- theta0[J * p + 2 * J + 1]
  exp(v0 + tanh(sweep(X %*% t(W), 2, w0, "+")) %*% v)
}))

back <- nn_pois(theta0, X, ytrue)$grad

## central differences, one coordinate at a time
h <- 1e-6
numeric_grad <- vapply(seq_along(theta0), function(k) {
  e <- numeric(length(theta0)); e[k] <- h
  (nn_pois(theta0 + e, X, ytrue)$nll - nn_pois(theta0 - e, X, ytrue)$nll) / (2 * h)
}, numeric(1))

cbind(backprop = back, numeric = numeric_grad)[1:6, ]
max(abs(back - numeric_grad))              # agreement across all 16 coordinates
```

The two agree to the accuracy of the difference quotient. With a gradient this
cheap and this reliable, BFGS or conjugate gradient can fit the model without
ever forming a Hessian — which is what the rest of this section is about.


## Gradient-Based Methods

Newton–Raphson is the method of choice when the second derivatives are available
and $p$ is modest. Outside that regime, three considerations push toward
alternatives: derivatives may be unavailable or painful to code; the $O(p^3)$
cost of solving the Newton system and the $O(p^2)$ storage of $J$ become
prohibitive for large $p$; and the objective may be non-smooth, noisy or
multimodal.

As everywhere in this chapter, the objective is the negative log-likelihood $\mathcal{L}(\theta) = -\ell(\theta)$, whose minimum is exactly the maximum of $\ell$. This is also the convention that R's `optim()`, `nlm()`, and `nlminb()` use internally, so the code below matches what those functions do under the hood. The methods here apply to any smooth objective, not only a likelihood, so this section writes $f$ for the function being minimized and keeps $g = \nabla f$ for its gradient — the same $g$ used everywhere else in the chapter, where $f = \mathcal{L}$ and $g = \mathcal{L}'(\theta)$ in the univariate case.

### Classification by derivative information

Methods are most usefully organised by what they require of the objective.

| Order | Requires | Methods | Typical use |
|---:|:---|:---|:---|
| 0 | Function values only | Nelder–Mead, simulated annealing, genetic algorithms | Non-smooth, discontinuous, or awkward-to-differentiate objectives; small $p$ |
| 1 | Gradient | Steepest descent, conjugate gradient, quasi-Newton (BFGS, L-BFGS), stochastic gradient descent | The workhorse class; large $p$; gradients cheap by automatic differentiation |
| 2 | Gradient and Hessian | Newton–Raphson, Fisher scoring, IRLS, trust-region Newton | Statistical models of modest dimension where the Hessian is also wanted for inference |

: Organizes the optimization methods of this chapter by how much derivative information each one requires. Optimization methods classified by the derivative information required {#tbl-orders}


### Steepest Descent with Line Search

Move along $-g(\theta_k)$, the direction of steepest descent, choosing the distance by a one-dimensional search:
$$
\theta_{k+1} = \theta_k - \alpha_k\, g(\theta_k), \qquad
\alpha_k = \arg\min_{\alpha > 0} f\!\left(\theta_k - \alpha\, g(\theta_k)\right) .
$$

Writing $g_k$ for $g(\theta_k)$, the step is simply $\theta_{k+1} = \theta_k - \alpha_k g_k$: the
search direction *is* the negative gradient, so it needs no name of its own.

The step length $\alpha_k$ can come from an **exact** line search — a univariate minimization by golden section or `optimize()` along the ray — or an **inexact** one: start at $\alpha = 1$ and halve until the Armijo sufficient-decrease condition $f(\theta_k - \alpha g_k) \le f(\theta_k) - c\,\alpha\,g_k'g_k$ holds. This is conceptually the simplest gradient method, and the one that best illustrates the role of a line search, but its convergence is only linear, and the rate deteriorates with the condition number of the Hessian: on an elongated valley the iterates zig-zag across it rather than travelling along it, as the next figure shows. It is rarely used unmodified.

```{r}
#| label: grad-descent-fn
grad_descent <- function(f, g, x0, maxit = 1000, tol = 1e-8) {
  path <- x0
  for (k in 1:maxit) {
    gk <- g(x0)
    a <- optimize(function(a) f(x0 - a * gk), c(0, 10))$minimum   # exact line search
    x1 <- x0 - a * gk; path <- rbind(path, x1)
    if (sqrt(sum((x1 - x0)^2)) < tol) break
    x0 <- x1 }
  list(par = x1, iter = k, path = path)
}
```

### Why Gradient Descent Zig-Zags

With an exact line search, consecutive gradients are **orthogonal** — $g_{k+1}'g_k = 0$, since the search stops exactly where the directional derivative along $-g_k$ vanishes. On the elongated contours of the quadratic from @fig-hessian-ellipse, that orthogonality forces the path to zig-zag back and forth across the valley instead of running down its length:

```{r}
#| label: fig-gd-zigzag
#| echo: false
#| fig-cap: "Illustrates the main weakness of gradient descent: on an elongated objective it makes slow zig-zag progress.
#|   Gradient descent with an exact line search zig-zags across an elongated valley: each step is orthogonal to the next."
fq <- function(v) 0.5 * drop(t(v) %*% A %*% v); gq <- function(v) drop(A %*% v)
gd <- grad_descent(fq, gq, c(2, -1.8), maxit = 30)
par(pty = "s", mar = c(4, 4, 2, 1))
contour(s, s, -z, nlevels = 15, main = sprintf("gradient descent, %d iterations", gd$iter))
lines(gd$path, type = "b", col = 2, pch = 19, cex = 0.6)
```

The number of iterations needed grows with the **condition number** $\lambda_{\max}/\lambda_{\min}$ of the Hessian — precisely the elongation of the ellipses in the earlier figure.

### Conjugate Gradient

The zig-zag is fixed by choosing each new direction to be **conjugate** to the previous ones with respect to the Hessian $Q$: $d_i'Qd_j = 0$ for $i \ne j$. For a quadratic $f(\theta) = \tfrac12\theta'Q\theta - b'\theta$, minimizing along $p$ such directions in succession reaches the exact minimum in **at most $p$ steps**, without ever forming or storing $Q$ — storage is $O(p)$, its defining advantage over Newton.

Here a separate symbol is warranted: the search direction $d_k$ is no longer the
negative gradient, but a correction of it. The nonlinear version builds these
directions from gradients only:
$$
d_0 = -g_0, \qquad d_{k+1} = -g_{k+1} + \beta_k d_k,
$$
$$
\beta_k^{\mathrm{FR}} = \frac{g_{k+1}'g_{k+1}}{g_k'g_k}\ \text{(Fletcher–Reeves)}, \qquad
\beta_k^{\mathrm{PR}} = \frac{g_{k+1}'(g_{k+1} - g_k)}{g_k'g_k}\ \text{(Polak–Ribière)},
$$
with $\theta_{k+1} = \theta_k + \alpha_k d_k$ from a line search and $g_k = g(\theta_k)$ as before. Fletcher–Reeves and Polak–Ribière differ only in this correction coefficient. The method is sensitive to the accuracy of the line search, and conjugacy degrades on non-quadratic $f$, so it is usually restarted with $d = -g$ every $p$ iterations.

```{r}
#| label: fig-conjugate-gradient
#| fig-cap: "Shows how conjugate gradient repairs the zig-zag behaviour of gradient descent by choosing search directions that do not undo earlier progress.
#|   Conjugate gradient (blue) reaches the minimum of the same elongated quadratic in far fewer iterations than gradient descent (red)."
conj_grad <- function(f, g, x0, maxit = 1000, tol = 1e-8, restart = length(x0)) {
  g0 <- g(x0); d <- -g0; path <- x0            # d: a conjugate direction, not -g
  for (k in 1:maxit) {
    a <- optimize(function(a) f(x0 + a * d), c(0, 10))$minimum
    x1 <- x0 + a * d; g1 <- g(x1); path <- rbind(path, x1)
    if (sqrt(sum(g1^2)) < tol) break
    beta <- if (k %% restart == 0) 0 else max(0, sum(g1 * (g1 - g0)) / sum(g0^2))   # Polak-Ribiere+
    d <- -g1 + beta * d; x0 <- x1; g0 <- g1 }
  list(par = x1, iter = k, path = path)
}
cg <- conj_grad(fq, gq, c(2, -1.8))
par(pty = "s", mar = c(4, 4, 2, 1))
contour(s, s, -z, nlevels = 15, main = sprintf("conjugate gradient, %d iterations", cg$iter))
lines(gd$path, type = "b", col = 2, pch = 19, cex = 0.6); lines(cg$path, type = "b", col = 4, pch = 19, lwd = 2)
legend("topleft", c("gradient descent", "conjugate gradient"), col = c(2, 4), lwd = 2, bty = "n")
```

### Quasi-Newton: BFGS

BFGS sits between the gradient methods above and Newton's method: it builds an approximation $B_k \approx J(\theta_k)$ from successive gradient differences $s_k = \theta_{k+1} - \theta_k$, $y_k = g_{k+1} - g_k$, via a rank-two update that keeps $B_k$ positive definite by construction (which is exactly why the Wolfe curvature condition of an earlier section matters: it is what guarantees $s_k'y_k > 0$, and hence a well-defined update). Steps are $\theta_{k+1} = \theta_k - \alpha_k B_k^{-1} g_k$ with a line search, giving **superlinear** convergence from only first derivatives — again a rescaling of $g_k$ rather than $g_k$ itself. BFGS itself stores a full $p \times p$ approximation; **L-BFGS** keeps only the last few update vectors, reducing storage to $O(mp)$ and making it the default choice for high-dimensional smooth problems. For most likelihood problems where the Hessian is awkward to derive, BFGS with an analytic gradient is the best practical compromise.

Returning to the logistic-regression example, `optim()` can fit it with no Hessian at all — Nelder–Mead needs only the log-likelihood, CG and BFGS also take the analytic gradient already coded for Newton–Raphson:

```{r}
#| label: bfgs-vs-nm-cg
negll_fn  <- function(b) neg_logp_logistic(b, data$x, data$y)
neggr_fn  <- function(b) nll_grad_info(b, data$y, data$x)$grad

fits <- list(NM   = optim(c(0, 0), negll_fn, method = "Nelder-Mead"),
             CG   = optim(c(0, 0), negll_fn, neggr_fn, method = "CG"),
             BFGS = optim(c(0, 0), negll_fn, neggr_fn, method = "BFGS", hessian = TRUE))
t(sapply(fits, function(f) c(f$par, f_evals = f$counts[1], g_evals = f$counts[2])))
sqrt(diag(solve(fits$BFGS$hessian)))     # SEs from the numerically computed Hessian at the optimum
```

All three land on the same estimate as `mle_logistic_nr` and `glm` above, and `optim(..., hessian = TRUE)` returns a finite-difference Hessian at the solution, so standard errors are available even for a method that never formed one during the iterations.

### Shinylive App for Comparing the Optimizers {#sec-gradient-methods-app}

There is a [shinylive app](../shinyliveapps_compstat/app_gradient-methods.html) to let you compare the paths that gradient descent, conjugate gradient, BFGS and safeguarded Newton take on the same objective, to see how curvature information affects speed and stability.

## Nelder–Mead (Simplex Method)

Nelder–Mead maintains a **simplex** of $p+1$ points (a triangle in 2D) and moves it downhill using only function values, with no derivatives at all:

1. Order the vertices $f(\theta_{(1)}) \le \cdots \le f(\theta_{(p+1)})$; let $\overline\theta$ be the centroid of the best $p$.

2. **Reflect** the worst vertex through $\overline\theta$; if the reflected point is the new best, try **expanding** further.

3. If reflection is poor, **contract** toward $\overline\theta$; if that also fails, **shrink** all vertices toward the best.

```{r}
#| label: nelder-mead-rosenbrock
rosen <- function(v) (1 - v[1])^2 + 100 * (v[2] - v[1]^2)^2       # Rosenbrock "banana", min at (1, 1)
nm <- optim(c(-1.2, 1), rosen, method = "Nelder-Mead"); c(nm$par, evals = nm$counts["function"])
```

The method needs no derivatives and tolerates mild non-smoothness, but it converges slowly, degrades badly as $p$ grows past roughly 10–20, and can stall at non-stationary points. It is `optim`'s default, which makes it by far the most-used — and most-overused — method in R: a reasonable first look at an unfamiliar objective, but a poor choice for a likelihood you will fit thousands of times.

### Shinylive App for the Nelder–Mead Algorithm {#sec-nelder-mead-app}

There is a [shinylive app](../shinyliveapps_compstat/app_nelder-mead.html) to show how the Nelder-Mead method finds a minimum without derivatives, by repeatedly reflecting, expanding, contracting, and shrinking a simplex.

## Stochastic Gradient Descent

### Motivation: Large $n$

For iid data the negative log-likelihood and its gradient are **sums over observations**:
$$
\mathcal{L}(\theta) = \sum_{i=1}^n \mathcal{L}_i(\theta), \qquad g(\theta) = \sum_{i=1}^n g_i(\theta).
$$

Every method in the two previous sections needs the **full** gradient — an $O(np)$ pass through all the data — at every single iteration, and Newton needs $O(np^2)$ for the Hessian on top of that. With $n$ in the millions and $p$ in the millions, as in a large neural network, even one full gradient evaluation is expensive.

The idea behind stochastic gradient descent is that $g_i(\theta)$ for a randomly chosen $i$ is an **unbiased estimate** of the average gradient,
$$
E_i\big[g_i(\theta)\big] = \frac{1}{n}g(\theta),
$$
so a noisy but cheap gradient estimate is often enough to make progress — the same Monte Carlo idea used elsewhere in this book, now applied to optimization rather than integration.

### The SGD Algorithm

For $t = 1, 2, \ldots$:

1. Sample an index $i_t$ uniformly from $\{1, \ldots, n\}$ — or shuffle the data once and cycle through it, so that one full pass is called an **epoch**.

2. Update in the descent direction of that single observation's negative log-likelihood:
$$
\theta_{t+1} = \theta_t - \gamma_t\,g_{i_t}(\theta_t).
$$

**Mini-batch SGD** averages the gradient over a batch $B_t$ of $m$ observations (typically $m = 32$–$512$ in machine-learning practice, though far smaller batches suffice for the simple models in this book):
$$
\theta_{t+1} = \theta_t - \gamma_t\,\frac{1}{m}\sum_{i\in B_t}g_i(\theta_t),
$$
which reduces the variance of the gradient estimate by a factor of $1/m$ at $m$ times the per-step cost, and is what lets the computation vectorize efficiently or run on a GPU. The **learning rate** $\gamma_t$ plays the role that a line search played for the deterministic methods above — a line search itself would need a full pass over the data to evaluate $\mathcal{L}$, defeating the purpose.

### Learning Rate and Convergence

Because each update uses a noisy gradient, a constant learning rate $\gamma$ makes $\theta_t$ **bounce around** the optimum indefinitely, with a variance proportional to $\gamma$. The classical result of Robbins and Monro (1951) is that $\theta_t$ converges to the optimum provided
$$
\sum_t \gamma_t = \infty \quad\text{and}\quad \sum_t \gamma_t^2 < \infty, \qquad\text{e.g. } \gamma_t = \frac{\gamma_0}{1 + t/t_0},
$$
the first condition allowing the iterate to travel arbitrarily far from its start, and the second forcing the noise to die out fast enough for it to settle.

A learning rate that is too large causes divergence or persistent noise; one that is too small is painfully slow. Convergence is only **sublinear** — far slower than Newton's quadratic rate *per iteration* — but each SGD iteration costs $O(p)$ or $O(mp)$ rather than $O(np)$ for a full gradient or $O(np^2)$ for a Hessian, and early progress is often fast even when final convergence is not. Common refinements include **momentum** (averaging past gradients to damp the noise), **Adam** (per-coordinate adaptive learning rates), and **Polyak–Ruppert averaging** of the iterates, $\overline\theta_T = \tfrac1T\sum_{t\le T}\theta_t$, which restores the $\sqrt n$-efficiency of the MLE that the noisy path alone does not have.

### Shinylive App for Stochastic Gradient Descent {#sec-sgd-app}

There is a [shinylive app](../shinyliveapps_compstat/app_sgd.html) to illustrate how stochastic gradient descent trades the accuracy of each step for lower cost, by comparing mini-batch with full-batch updates on the same problem.

### Example: SGD for Logistic Regression in R

Returning one last time to the running logistic-regression example, mini-batch SGD needs only the per-observation gradient already coded for Newton–Raphson, applied to a small random subset at each update:

```{r}
#| label: sgd-logistic
sgd_logistic <- function(x, y, b0, gamma0 = 0.5, t0 = 100, epochs = 30, batch = 10) {
  n <- length(x); b <- b0; t <- 0; path <- b0
  for (e in 1:epochs) {
    idx <- sample(n)
    for (s in seq(1, n, by = batch)) {
      i <- idx[s:min(s + batch - 1, n)]
      t <- t + 1
      u <- b[1] + b[2] * x[i]
      p <- plogis(u)
      grad <- c(sum(y[i] - p), sum(x[i] * (y[i] - p))) / length(i)
      b <- b + gamma0 / (1 + t / t0) * grad
      path <- rbind(path, b)
    }
  }
  list(est = b, path = path)
}

set.seed(2)
beta_true <- c(0, 1.5)                    # the parameters used to simulate data$x, data$y above
sg <- sgd_logistic(data$x, data$y, c(0, 0))
rbind(sgd = sg$est, nlm = logit_nlm$estimate, truth = beta_true)
```

### SGD Trajectory vs. Newton–Raphson

```{r}
#| label: fig-sgd-vs-newton
#| fig-cap: "Compares stochastic gradient descent, which is cheap per step but noisy, with Newton-Raphson, which is expensive per step but takes large steps, on the same logistic regression problem.
#|   SGD (fine, noisy trajectory) versus Newton-Raphson (large steps, one per full pass through the data) converging to the same logistic-regression MLE."
nr_path <- mle_logistic_nr(c(0, 0), 8, data$y, data$x)     # beta0, beta1, neg_loglike per Newton iteration
m <- 10; per_epoch <- length(data$x) / m
xs <- 0:(nrow(sg$path) - 1)
xn <- (0:(nrow(nr_path) - 1)) * per_epoch

matplot(xs, sg$path, type = "l", lty = 1, col = c(2, 4),
        ylim = range(sg$path, nr_path[, 1:2]),
        xlab = "mini-batch updates (one epoch = one full pass through the data)",
        ylab = "coefficient",
        main = expression(paste("SGD vs. Newton-Raphson: ", beta[0], " (red), ", beta[1], " (blue)")))
matlines(xn, nr_path[, 1:2], type = "b", lty = 2, pch = 19, col = c(2, 4), lwd = 2)
abline(h = logit_nlm$estimate, col = c(2, 4), lty = 3)
abline(v = seq(0, max(xs), per_epoch), col = "grey85")
legend("right", c("SGD (mini-batch)", "Newton (per full pass)", "MLE"),
       lty = c(1, 2, 3), pch = c(NA, 19, NA), bty = "n", cex = 0.85)
```

Newton needs only a handful of full-data passes to converge on this small problem ($n = 200$, $p = 2$); SGD makes many cheap updates and lands near the MLE with residual noise from its stochastic gradient. Per data pass, Newton is far more efficient here — its advantage narrows, and eventually reverses, only once a full gradient, or the $O(p^3)$ Hessian solve, becomes too expensive to compute at every iteration, which is exactly the regime where SGD earns its keep.

## Comparing All the Methods

The chapter has covered a lot of ground, from bisection on a scalar gradient $g$ to stochastic gradient descent on millions of parameters. The three tables below collect it into a single reference: which method needs what, which R function to reach for, and how to choose among them for a given problem.

### Comparison

| Method | Derivatives | Storage | Cost per iteration | Local rate | Robust start | Hessian for SEs |
|:---|:---|:---|:---|:---|:---|:---|
| Nelder–Mead | none | $O(p^2)$ | $O(p)$ evaluations | sublinear | good | no |
| Steepest descent | gradient | $O(p)$ | 1 gradient + line search | linear (slow) | good | no |
| Conjugate gradient | gradient | $O(p)$ | 1 gradient + line search | linear (fast) | moderate | no |
| BFGS | gradient | $O(p^2)$ | 1 gradient + $O(p^2)$ | superlinear | moderate | approximate |
| L-BFGS | gradient | $O(mp)$ | 1 gradient + $O(mp)$ | superlinear | moderate | no |
| Newton–Raphson | gradient, Hessian | $O(p^2)$ | 1 Hessian + $O(p^3)$ | quadratic | poor | exact |
| Fisher scoring / IRLS | gradient, expected information | $O(p^2)$ | 1 weighted LS fit | near-quadratic | fair | exact |
| Stochastic gradient | minibatch gradient | $O(p)$ | $O(\text{batch})$ | sublinear | good | no |

: Summarizes the trade-offs among the multivariate optimization methods of this chapter: what each needs, what it costs, and how fast and robust it is. Comparison of multivariate optimization methods. $p$ is the number of parameters and $m$ the L-BFGS memory length (typically 5–20). {#tbl-comparison}

### Implementations in R

| Function | Package | Methods available | Notes |
|:---|:---|:---|:---|
| `optim` | stats | Nelder–Mead, BFGS, CG, L-BFGS-B, SANN | General workhorse; supply `gr` or the gradient is differenced; `hessian = TRUE` for inference; L-BFGS-B allows box constraints |
| `nlm` | stats | Newton-type with ridging | Uses analytic gradient/Hessian if attached as attributes, otherwise numerical; returns `$hessian` |
| `nlminb` | stats | Trust-region (PORT) | Reliable on badly scaled problems; supports box constraints |
| `optimize`, `uniroot` | stats | Golden section, Brent | One dimension only |
| `glm` | stats | IRLS (Fisher scoring) | The specialised case; use `start` when separation is suspected |
| `nls` | stats | Gauss–Newton, Golub–Pereyra | Nonlinear least squares; `port` algorithm adds constraints |
| `optimx` | optimx | Unified front end to a dozen optimizers | Useful for comparing methods on the same objective |
| `nloptr` | nloptr | Large NLopt collection, with constraints | Global and derivative-free options |
| `Rsolnp`, `alabama` | — | Nonlinear equality and inequality constraints | For genuinely constrained problems |

: Maps the optimization methods to the R functions and packages that implement them. R interfaces to multivariate optimizers {#tbl-r-functions}

### Choosing a method

| Situation | Suggested approach |
|:---|:---|
| Hessian derivable, $p$ small, standard errors wanted | Newton–Raphson or Fisher scoring |
| GLM or a model expressible as one | `glm` (IRLS) |
| Gradient available, $p$ moderate | BFGS via `optim(method = "BFGS")` |
| Gradient available, $p$ large | L-BFGS |
| Derivatives unavailable, $p < 10$ | Nelder–Mead as a first pass, then refine |
| Objective noisy or non-smooth | Nelder–Mead or simulated annealing |
| Divergence or poor conditioning suspected | `nlm` or `nlminb`; rescale the parameters |
| Objective a sum over very many observations | Stochastic gradient descent and variants |
| Multiple local maxima suspected | Multiple random starts with any of the above |

: Gives practical rules for choosing an optimizer according to the features of the problem. Practical guidance on method selection {#tbl-choosing}

The last row deserves emphasis. Every method in this chapter is a *local*
method: each converges to a stationary point near where it started, and none
can certify that the point found is the global maximum. Restarting from a
dispersed set of initial values and comparing the attained log-likelihoods is
the only general safeguard, and it should be routine for any likelihood that is
not known to be concave.
