---
title: Introduction to R Programming
name: Longhai Li
date: last-modified
code-fold: false
warning: false
---

## A toy example

Throughout this unit we will use one running example to introduce the basic building blocks of R: vectors, strings, special values, matrices, data frames, lists, functions, file I/O, and graphics.

**The mission:** a class of 100 students has taken graded work in three *categories*: five quizzes, two tests, and one final exam, each scored out of 100 points. A student's **course score** is a weighted average of the category means -- quizzes count 30%, tests count 30%, and the final counts 40%. To make the simulated data realistic, a student's test and final scores are generated from their quiz performance by a simple linear regression, with a different intercept for Program A and Program B students, and a few quizzes are missing entirely (a student who did not sit that quiz). We want to

1. work out the calculation step by step for one student,
2. package that calculation into a reusable function,
3. apply the function to every student to get a full set of results, and
4. present those results as a table and as plots.

This mirrors how real data-analysis code is usually built: explore on one case, generalize into a function, then apply and report.

## Generating an artificial score dataset

### Vectors: recording quiz scores

Each quiz, across all students, is naturally stored as a vector. We simulate a class of 100 students, five quizzes scored out of 100, using `rnorm` and clipping the result to a valid `0`-`100` range with `pmin`/`pmax`:

```{r}
set.seed(123)
n <- 100

gen.quiz <- function(mean, sd)
{
    x <- round(rnorm(n, mean, sd))
    pmin(100, pmax(0, x))
}

Quiz1 <- gen.quiz(75, 20)
Quiz2 <- gen.quiz(78, 20)
Quiz3 <- gen.quiz(72, 20)
Quiz4 <- gen.quiz(80, 20)
Quiz5 <- gen.quiz(74, 20)

Quiz1

```

### Strings: naming the categories and students

We could type out `"Quiz1"`, `"Quiz2"`, ... by hand, but `paste0` builds such labels programmatically -- a habit that pays off the moment there are 40 quizzes instead of 5:

```{r}
quiz.names <- paste0("Quiz", 1:5)
test.names <- paste0("Test", 1:2)

quiz.names
test.names

```

With 100 students, typing out individual names is not realistic either. `sprintf` generates zero-padded IDs like `"S001"`, `"S002"`, ... in one line -- exactly the kind of job it is built for:

```{r}
students <- sprintf("S%03d", 1:n)

head(students)

```

### Simulating the tests and the final using vectorized math

In a real class, a student's test and final scores are not independent of how they did on the quizzes. We simulate that by treating each student's **mean quiz score** as a predictor in a simple linear regression.

This is a great place to demonstrate **vectorized calculations** in R. Instead of writing a `for` loop to calculate each student's score one by one, we can simply add and multiply whole vectors of length 100 together. R aligns the elements automatically:

```{r}
program <- factor(sample(c("Program A", "Program B"), n, replace = TRUE))

quiz.mat <- cbind(Quiz1, Quiz2, Quiz3, Quiz4, Quiz5)
quiz.mean.true <- rowMeans(quiz.mat)

b0.test  <- ifelse(program == "Program A", 12, 6)

b1.test  <- 0.80

b0.final <- ifelse(program == "Program A", 15, 8)
b1.final <- 0.85

clip.round <- function(x) round(pmin(100, pmax(0, x)))

b0.test
b1.test
quiz.mean.true
jitter <- rnorm(n, 0, 8)
Test1.raw <- b0.test  + b1.test*quiz.mean.true  + rnorm(n, 0, 8)
Test1 <- clip.round(Test1.raw)

data.frame(program, b0.test, b1.test, quiz.mean.true, jitter, Test1.raw)
```

Using the above method, we generate two tests and a final:

```{r}
# Vectorized math: b0.test + b1.test * quiz.mean.true + rnorm(...)
Test1 <- clip.round(b0.test  + b1.test*quiz.mean.true  + rnorm(n, 0, 8))
Test2 <- clip.round(b0.test  + b1.test*quiz.mean.true  + rnorm(n, 0, 8))
Final <- clip.round(b0.final + b1.final*quiz.mean.true + rnorm(n, 0, 6))
```

### Matrices: combining all the scores (Complete Data)

Before we simulate absences (missing values), let's bind the quiz, test, and final columns together into one complete matrix. It will have one row per student and one column per graded item.

```{r}
scores <- cbind(quiz.mat, Test1, Test2, Final)
colnames(scores) <- c(quiz.names, test.names, "Final")
rownames(scores) <- students

# Here is our complete dataset before any missing values
head(scores)

```

### Adding and handling missing values

In reality, students occasionally miss a quiz. Let's randomly drop about 5% of the quiz scores by assigning them the special value `NA` (Not Available).

```{r}
miss.rate <- 0.05

# We only want to introduce NAs into the first 5 columns (the quizzes)
for(j in 1:ncol(scores)) {
    scores[sample(n, round(n*miss.rate)), j] <- NA
}

head(scores, 10)

```

`is.na` locates missing values anywhere in an object. Only the quiz columns have any, since we generated `Test1`, `Test2`, and `Final` without missingness:

```{r}
colSums(is.na(scores))

```

We will rely on `na.rm = TRUE` later so that a missing quiz simply drops out of a student's quiz average, rather than returning `NA` for their whole course score.

### Data frames: adding names and program

A data frame lets us combine columns of different types -- here, the students' names and program alongside their (numeric) scores -- into a single table:

```{r}
roster <- data.frame(name = students, program = program, scores, check.names = FALSE)

roster

```

## Calculating a Single Student's Score

Before writing a function, it often helps to work through the logic step by step on a single example. Let's find a student who missed a quiz so we can see how missing values are handled, and extract their record:

```{r}
# Find the index of the first student who has at least one NA in columns 1 to 5
example.ix <- which(rowSums(is.na(scores[, 1:5])) > 0)[1]
student_record <- roster[example.ix, ]

student_record

```

To avoid hardcoding column positions, we can extract the column names from this record and use `grep` to dynamically search for columns starting with "Quiz", "Test", or "Final":

```{r}
# Extract all the column names from the student record
col_names <- names(student_record)

# Use grep to find the numeric indices of specific categories
quiz.cols <- grep("^Quiz", col_names)
test.cols <- grep("^Test", col_names)
final.col <- grep("^Final", col_names)

# Check which columns grep found for the quizzes
quiz.cols

```

Next, we extract the scores using those indices, convert them to numeric values, and calculate the mean for each category. Using `na.rm = TRUE` ensures that the missing quiz is simply ignored rather than turning the whole average into `NA`:

```{r}
# Calculate the mean for each category, ignoring NAs
quiz.mean <- mean(as.numeric(student_record[quiz.cols]), na.rm = TRUE)
test.mean <- mean(as.numeric(student_record[test.cols]), na.rm = TRUE)
final.score <- as.numeric(student_record[final.col])

c(Quiz = quiz.mean, Test = test.mean, Final = final.score)

```

Finally, we apply the syllabus weights to calculate the overall course score:

```{r}
# Calculate the weighted course score
course.score <- (0.3 * quiz.mean) + (0.3 * test.mean) + (0.4 * final.score)

course.score

```

### Aggregating into a Reusable Function

Now that we know the logic works perfectly for one student, we can package those exact steps into a reusable function. This function will take any `student_record` as its input and return a named vector containing the category means and the final score. 

To make the grading policy more forgiving, we are adding a conditional rule: if a student's final exam score is strictly higher than their test average, the final exam score will replace the test average. Because `compute.course.score` processes exactly one student at a time, we evaluate a single logical condition using a standard `if` statement rather than a vectorized alternative like `ifelse`. We also use `!is.na()` checks to ensure the `if` statement doesn't crash by attempting to evaluate a missing value (which could happen if a student had no test or final scores).

```{r}
compute.course.score <- function(student_record, w_quiz = 0.3, w_test = 0.3, w_final = 0.4) {
    
    col_names <- names(student_record)
    quiz.cols <- grep("^Quiz", col_names)
    test.cols <- grep("^Test", col_names)
    final.col <- grep("^Final", col_names)
    
    quiz.mean <- mean(as.numeric(student_record[quiz.cols]), na.rm = TRUE)
    test.mean <- mean(as.numeric(student_record[test.cols]), na.rm = TRUE)
    final.score <- as.numeric(student_record[final.col])
    
    # Explicit conditional check: replace test mean with final score if it is higher
    if (!is.na(final.score) && !is.na(test.mean) && final.score > test.mean) {
        test.mean <- final.score
    }
    
    # The weights are now dynamic variables rather than hardcoded numbers
    course.score <- (w_quiz * quiz.mean) + (w_test * test.mean) + (w_final * final.score)
    
    c(Quiz = quiz.mean, Test = test.mean, Final = final.score, course.score = course.score)
}

```

Let's test our updated function on two specific scenarios to verify the conditional logic. We will format these inputs as **named vectors** rather than data frames. This mimics exactly what the `apply()` function will do later when it passes rows of our dataset to this function.

**Student A** struggled on the tests (average: 70) but aced the final (95). The function should detect that `95 > 70` and replace their test mean with 95 before calculating the final course score.
**Student B** did great on the tests (average: 90) but dropped on the final (75). The function should keep their original test mean.

```{r}
# Create the mock record for Student A as a named vector
student_A <- c(
    name = "Student_A", program = "Program A",
    Quiz1 = "80", Quiz2 = "80", Quiz3 = "80", Quiz4 = "80", Quiz5 = "80", # Quiz mean = 80
    Test1 = "65", Test2 = "75",                                           # Test mean = 70
    Final = "95"
)

# Create the mock record for Student B as a named vector
student_B <- c(
    name = "Student_B", program = "Program B",
    Quiz1 = "80", Quiz2 = "80", Quiz3 = "80", Quiz4 = "80", Quiz5 = "80", # Quiz mean = 80
    Test1 = "90", Test2 = "90",                                           # Test mean = 90
    Final = "75"
)

# Test Student A: The Test mean should be updated to 95 in the output
cat("Student A Results:\n")
compute.course.score(student_A)
# Math check: (0.3 * 80) + (0.3 * 95) + (0.4 * 95) = 90.5

# Test Student B: The Test mean should remain 90 in the output
cat("\nStudent B Results:\n")
compute.course.score(student_B)
# Math check: (0.3 * 80) + (0.3 * 90) + (0.4 * 75) = 81
```

## Applying the Function to Every Student

Now that we have a function working for one student, we can apply it across the entire class. The `apply` function is perfect for this: it takes our `roster` data frame, loops over either rows (`1`) or columns (`2`), and executes our custom function.

```{r}
# apply over margin 1 (rows)
all.results <- t(apply(roster, 1, compute.course.score))

# Combine the original demographic info with the calculated scores
results <- cbind(roster[, c("name", "program")], all.results)

results

```

**Looping over the dataset: `for` loop vs. `apply`**

While `apply()` is idiomatic R and very concise, we can also use a traditional `for` loop to process each student. To do this efficiently, we first pre-allocate an empty matrix of the correct size to store our output. Then, we iterate through the row indices one by one.

Inside the loop, we use `unlist()` to convert the single row of our data frame into a named vector. This guarantees we are feeding our `compute.course.score` function the exact same data structure that `apply()` uses under the hood.

```{r}
# 1. Pre-allocate an empty matrix to hold the results
num_students <- nrow(roster)
loop_results <- matrix(NA, nrow = num_students, ncol = 4)

# Set the column and row names to match our expected output
colnames(loop_results) <- c("Quiz", "Test", "Final", "course.score")
rownames(loop_results) <- roster$name

# 2. Iterate over each row index
for (i in 1:num_students) {
    # Extract the single row as a named character vector
    student_record <- unlist(roster[i, ]) 
    
    # Compute the score and store it in the corresponding row of our matrix
    loop_results[i, ] <- compute.course.score(student_record)
}

# Peek at the first few rows of the loop results
head(loop_results)
```

Now, let's run the exact same calculation using `apply()` and use the `all.equal()` function to verify that both methods yield the exact same mathematical results. 

```{r}

# 3. Check if they are mathematically identical
# (We set check.attributes = FALSE to ignore minor metadata differences, 
# like how apply() inherits row names vs. how we manually assigned them)
is_same <- all.equal(loop_results, results, check.attributes = FALSE)

print(paste("Do the for loop and apply() give the exact same results?", is_same))
```

### Saving and reloading the results

The `results` table is worth keeping. A data frame can be written as a plain csv file for spreadsheets, or saved as an `.RDS` file to preserve it exactly (including the `program` factor) for future R sessions:

```{r}
write.csv(results, file = "results.csv", row.names = FALSE)

saveRDS(results, file = "results.RDS")

results.reloaded <- readRDS("results.RDS")

identical(results, results.reloaded)

```


## Writing a Function for the Whole Process with List Output

Sometimes a function needs to return multiple distinct pieces of information that don't fit neatly into a single table. A **list** in R is a flexible container that can bundle objects of completely different types and sizes together.

Let's write a wrapper function that takes our raw `roster` data and our weighting scheme. It will calculate the scores, but instead of just returning the table, it will return a list containing four very different items:

1. The full processed data frame.
2. A numeric vector of just the final course scores.
3. A frequency table of scores binned into 10-point intervals.
4. A character vector of the names of students who failed (scored below 60).

```{r}
analyze_class_performance <- function(roster_data, w_quiz = 0.3, w_test = 0.3, w_final = 0.4) {
    
    # 1. Pass the weights down into compute.course.score
    all.results <- t(apply(roster_data, 1, compute.course.score, 
                           w_quiz = w_quiz, w_test = w_test, w_final = w_final))
                           
    processed_table <- cbind(roster_data[, c("name", "program")], all.results)
    
    # 2. Extract final course scores 
    final_scores <- processed_table$course.score
    
    # 3. Create bins (handling NAs)
    bins <- seq(0, 100, by = 10)
    score_bins <- table(cut(final_scores, breaks = bins, include.lowest = TRUE), useNA = "ifany")
    
    # 4. Extract names of failing students (safely ignoring NAs using which)
    failing_students <- processed_table$name[which(final_scores < 60)]
    
    # 5. Extract names of students with missing overall scores
    incomplete_students <- processed_table$name[is.na(final_scores)]
    
    # Bundle everything into a named list
    list(
        ProcessedTable = processed_table,
        CourseScores = final_scores,
        ScoreDistribution = score_bins,
        Failing_Students = failing_students,
        incomplete_students = incomplete_students
    )
}
```

When we run this function, it returns a single list object. We can use the `str()` function to quickly inspect the structure of our new list and see all the different data types nested inside it:

```{r}
# Run the analysis
class_report <- analyze_class_performance(roster)

# Inspect the structure of the returned list
str(class_report, max.level = 1)

```

To access individual elements inside a list, you use the `$` operator (just like with data frames) or double brackets `[[ ]]`. Let's look at the score distribution and the failing students:

```{r}
# Access the 10-point bin distribution
class_report$ScoreDistribution

# Access the names of the failing students
class_report$Failing_Students
class_report$incomplete_students 
```

## Presenting the Results

### Presenting the Processed Results with `gt`

Finally, **gt** turns the raw data frame of results into a polished, presentation-ready report. We can extract the `ProcessedTable` directly from our `class_report` list and enhance it. 

Here we use `tab_spanner()` to group the category scores and the final score under distinct visual headers, and `fmt_number()` to ensure all grades display exactly two decimal places.

```{r}
library(gt)

# Extract the full processed table from our list output
# (We use head() to display just the first 10 rows in the report)
full_results <- head(class_report$ProcessedTable, 10)

gt(full_results) |>
    tab_header(
        title = "Class Performance Report",
        subtitle = "Individual category means and final calculated course scores"
    ) |>
    # Student Information
    tab_spanner(
        label = "Student Information",
        columns = c(name, program)
    )|>
    # Group the individual graded categories
    tab_spanner(
        label = "Category Breakdown",
        columns = c(Quiz, Test, Final)
    ) |>
    # Group the final processed output
    tab_spanner(
        label = "Submitted Score",
        columns = c(course.score)
    ) |>
    # Enforce two decimal places for all numeric score columns
    fmt_number(
        columns = c(Quiz, Test, Final),
        decimals = 2
    ) |>
    fmt_number(
        columns = c(course.score),
        decimals = 0
    ) |>
    cols_label(
        name = "Name",
        program = "Program",
        Quiz = "Quizzes",
        Test = "Tests",
        Final = "Final Exam",
        course.score = "Course Score"
    ) |>
    tab_stubhead(label = "Student ID")

```

### Base R graphics

A quick histogram shows the distribution of course scores, with a vertical line marking the class average:

```{r}
hist(class_report$ProcessedTable$course.score, breaks = 15, col = "lightblue",
     main = "Distribution of Course Scores", xlab = "Course Score")
abline(v = mean(class_report$ProcessedTable$course.score, na.rm = TRUE), col = "red", lwd = 2)

```

### Plotting with ggplot2

**ggplot2** builds a comparison across programs declaratively, by mapping columns of `class_report$ProcessedTable` to visual properties and layering geometries with `+`. Here the boxplot is enhanced with the individual course scores jittered on top:

```{r}
library(ggplot2)

ggplot(class_report$ProcessedTable, aes(x = program, y = course.score, fill = program)) +
    geom_boxplot(show.legend = FALSE, alpha = 0.6) +
    geom_jitter(width = 0.05, size = 1.5, alpha = 0.6) +
    labs(
        title = "Course Score by Academic Program",
        x = "Program", y = "Course Score"
    ) +
    theme_minimal()

```

We built a program-specific intercept into the simulation when generating the final exam scores; a scatter plot of the final versus each student's mean quiz score, with a fitted line per program, should reveal it:

```{r}
ggplot(class_report$ProcessedTable, aes(x = Quiz, y = Final, color = program)) +
    geom_point(size = 2, alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE) +
    labs(
        title = "Final Exam vs. Mean Quiz Score, by Program",
        x = "Mean Quiz Score", y = "Final Exam Score", color = "Program"
    ) +
    theme_minimal()

```

The two fitted lines are roughly parallel but vertically offset -- exactly the intercept difference we built into `b0.final` when simulating the data.