---
title: Introduction to R Programming
code-fold: false
---

## A toy example

Throughout this unit we will use one running example to introduce the basic building blocks of R: vectors, strings, special values, matrices, data frames, lists, functions, file I/O, and graphics.

**The mission:** a class of 100 students has taken graded work in three *categories*: five quizzes, two tests, and one final exam, each scored out of 100 points. A student's **course score** is a weighted average of the category means -- quizzes count 30%, tests count 30%, and the final counts 40%. To make the simulated data realistic, a student's test and final scores are generated from their quiz performance by a simple linear regression, with a different intercept for female and male students, and a few quizzes are missing entirely (a student who did not sit that quiz). We want to

1. work out the calculation step by step for one student,
2. package that calculation into a reusable function,
3. apply the function to every student to get a full set of results, and
4. present those results as a table and as plots.

This mirrors how real data-analysis code is usually built: explore on one case, generalize into a function, then apply and report.

## Generating an artificial score dataset

### Vectors: recording quiz scores

Each quiz, across all students, is naturally stored as a vector. We simulate a class of 100 students, five quizzes scored out of 100, using `rnorm` and clipping the result to a valid `0`-`100` range with `pmin`/`pmax`:

```{r}
set.seed(123)
n <- 100

gen.quiz <- function(mean, sd)
{
    x <- round(rnorm(n, mean, sd))
    pmin(100, pmax(0, x))
}

Quiz1 <- gen.quiz(75, 20)
Quiz2 <- gen.quiz(78, 20)
Quiz3 <- gen.quiz(72, 20)
Quiz4 <- gen.quiz(80, 20)
Quiz5 <- gen.quiz(74, 20)

Quiz1

```

### Strings: naming the categories and students

We could type out `"Quiz1"`, `"Quiz2"`, ... by hand, but `paste0` builds such labels programmatically -- a habit that pays off the moment there are 40 quizzes instead of 5:

```{r}
quiz.names <- paste0("Quiz", 1:5)
test.names <- paste0("Test", 1:2)

quiz.names
test.names

```

With 100 students, typing out individual names is not realistic either. `sprintf` generates zero-padded IDs like `"S001"`, `"S002"`, ... in one line -- exactly the kind of job it is built for:

```{r}
students <- sprintf("S%03d", 1:n)

head(students)

```

### Simulating the tests and the final using vectorized math

In a real class, a student's test and final scores are not independent of how they did on the quizzes. We simulate that by treating each student's **mean quiz score** as a predictor in a simple linear regression.

This is a great place to demonstrate **vectorized calculations** in R. Instead of writing a `for` loop to calculate each student's score one by one, we can simply add and multiply whole vectors of length 100 together. R aligns the elements automatically:

```{r}
gender <- factor(sample(c("f","m"), n, replace = TRUE))

quiz.mat <- cbind(Quiz1, Quiz2, Quiz3, Quiz4, Quiz5)
quiz.mean.true <- rowMeans(quiz.mat)

b0.test  <- ifelse(gender == "f", 12, 6)

b1.test  <- 0.80

b0.final <- ifelse(gender == "f", 15, 8)
b1.final <- 0.85

clip.round <- function(x) round(pmin(100, pmax(0, x)))

b0.test
b1.test
quiz.mean.true
jitter <- rnorm(n, 0, 8)
Test1.raw <- b0.test  + b1.test*quiz.mean.true  + rnorm(n, 0, 8)
Test1 <- clip.round(Test1.raw)

data.frame(gender,b0.test, b1.test, quiz.mean.true, jitter, Test1.raw)
```

Using the above method, we generate two tests and a final
```{r}
# Vectorized math: b0.test + b1.test * quiz.mean.true + rnorm(...)
Test1 <- clip.round(b0.test  + b1.test*quiz.mean.true  + rnorm(n, 0, 8))
Test2 <- clip.round(b0.test  + b1.test*quiz.mean.true  + rnorm(n, 0, 8))
Final <- clip.round(b0.final + b1.final*quiz.mean.true + rnorm(n, 0, 6))
```

### Matrices: combining all the scores (Complete Data)

Before we simulate absences (missing values), let's bind the quiz, test, and final columns together into one complete matrix. It will have one row per student and one column per graded item.

```{r}
scores <- cbind(quiz.mat, Test1, Test2, Final)
colnames(scores) <- c(quiz.names, test.names, "Final")
rownames(scores) <- students

# Here is our complete dataset before any missing values
head(scores)

```

### Adding and handling missing values

In reality, students occasionally miss a quiz. Let's randomly drop about 5% of the quiz scores by assigning them the special value `NA` (Not Available).

```{r}
miss.rate <- 0.05

# We only want to introduce NAs into the first 5 columns (the quizzes)
for(j in 1:ncol(scores)) {
    scores[sample(n, round(n*miss.rate)), j] <- NA
}

head(scores,10)

```

`is.na` locates missing values anywhere in an object. Only the quiz columns have any, since we generated `Test1`, `Test2`, and `Final` without missingness:

```{r}
colSums(is.na(scores))

```

We will rely on `na.rm = TRUE` later so that a missing quiz simply drops out of a student's quiz average, rather than returning `NA` for their whole course score.

### Data frames: adding names and gender

A data frame lets us combine columns of different types -- here, the students' names and gender alongside their (numeric) scores -- into a single table:

```{r}
roster <- data.frame(name = students, gender = gender, scores, check.names = FALSE)

roster

```

## Calculating a Single Student's Score

Before writing a function, it often helps to work through the logic step by step on a single example. Let's find a student who missed a quiz so we can see how missing values are handled, and extract their record:

```{r}
# Find the index of the first student who has at least one NA in columns 1 to 5
example.ix <- which(rowSums(is.na(scores[, 1:5])) > 0)[1]
student_record <- roster[example.ix, ]

student_record

```

To avoid hardcoding column positions, we can extract the column names from this record and use `grep` to dynamically search for columns starting with "Quiz", "Test", or "Final":

```{r}
# Extract all the column names from the student record
col_names <- names(student_record)

# Use grep to find the numeric indices of specific categories
quiz.cols <- grep("^Quiz", col_names)
test.cols <- grep("^Test", col_names)
final.col <- grep("^Final", col_names)

# Check which columns grep found for the quizzes
quiz.cols

```

Next, we extract the scores using those indices, convert them to numeric values, and calculate the mean for each category. Using `na.rm = TRUE` ensures that the missing quiz is simply ignored rather than turning the whole average into `NA`:

```{r}
# Calculate the mean for each category, ignoring NAs
quiz.mean <- mean(as.numeric(student_record[quiz.cols]), na.rm = TRUE)
test.mean <- mean(as.numeric(student_record[test.cols]), na.rm = TRUE)
final.score <- as.numeric(student_record[final.col])

c(Quiz = quiz.mean, Test = test.mean, Final = final.score)

```

Finally, we apply the syllabus weights to calculate the overall course score:

```{r}
# Calculate the weighted course score
course.score <- (0.3 * quiz.mean) + (0.3 * test.mean) + (0.4 * final.score)

course.score

```

### Aggregating into a Reusable Function

Now that we know the logic works perfectly for one student, we can package those exact steps into a reusable function. This function will take any `student_record` as its input and return a named vector containing the category means and the final score.

```{r}
compute.course.score <- function(student_record, w_quiz = 0.3, w_test = 0.3, w_final = 0.4) {
    
    col_names <- names(student_record)
    quiz.cols <- grep("^Quiz", col_names)
    test.cols <- grep("^Test", col_names)
    final.col <- grep("^Final", col_names)
    
    quiz.mean <- mean(as.numeric(student_record[quiz.cols]), na.rm = TRUE)
    test.mean <- mean(as.numeric(student_record[test.cols]), na.rm = TRUE)
    final.score <- as.numeric(student_record[final.col])
    
    # The weights are now dynamic variables rather than hardcoded numbers
    course.score <- (w_quiz * quiz.mean) + (w_test * test.mean) + (w_final * final.score)
    
    c(Quiz = quiz.mean, Test = test.mean, Final = final.score, course.score = course.score)
}

```

Let's test our new function on that same student to verify it produces the exact same results we just calculated by hand:

```{r}
compute.course.score(roster[example.ix, ])

```

## Applying the Function to Every Student

Now that we have a function working for one student, we can apply it across the entire class. The `apply` function is perfect for this: it takes our `roster` data frame, loops over either rows (`1`) or columns (`2`), and executes our custom function.

```{r}
# apply over margin 1 (rows)
all.results <- t(apply(roster, 1, compute.course.score))

# Combine the original demographic info with the calculated scores
results <- cbind(roster[, c("name","gender")], all.results)

results

```

### Saving and reloading the results

The `results` table is worth keeping. A data frame can be written as a plain csv file for spreadsheets, or saved as an `.RDS` file to preserve it exactly (including the `gender` factor) for future R sessions:

```{r}
write.csv(results, file = "results.csv", row.names = FALSE)

saveRDS(results, file = "results.RDS")

results.reloaded <- readRDS("results.RDS")

identical(results, results.reloaded)

```


## Write a Function to Handle a Raw Score (Demonstrating list)

Sometimes a function needs to return multiple distinct pieces of information that don't fit neatly into a single table. A **list** in R is a flexible container that can bundle objects of completely different types and sizes together.

Let's write a wrapper function that takes our raw `roster` data and our weighting scheme. It will calculate the scores, but instead of just returning the table, it will return a list containing four very different items:
1. The full processed data frame.
2. A numeric vector of just the final course scores.
3. A frequency table of scores binned into 10-point intervals.
4. A character vector of the names of students who failed (scored below 60).

```{r}
analyze_class_performance <- function(roster_data, w_quiz = 0.3, w_test = 0.3, w_final = 0.4) {
    
    # 1. Pass the weights down into compute.course.score
    all.results <- t(apply(roster_data, 1, compute.course.score, 
                           w_quiz = w_quiz, w_test = w_test, w_final = w_final))
                           
    processed_table <- cbind(roster_data[, c("name", "gender")], all.results)
    
    # 2. Extract final course scores 
    final_scores <- processed_table$course.score
    
    # 3. Create bins (handling NAs)
    bins <- seq(0, 100, by = 10)
    score_bins <- table(cut(final_scores, breaks = bins, include.lowest = TRUE), useNA = "ifany")
    
    # 4. Extract names of failing students (safely ignoring NAs using which)
    failing_students <- processed_table$name[which(final_scores < 60)]
    
    # 5. Extract names of students with missing overall scores
    incomplete_students <- processed_table$name[is.na(final_scores)]
    
    # Bundle everything into a named list
    list(
        ProcessedTable = processed_table,
        CourseScores = final_scores,
        ScoreDistribution = score_bins,
        Failing_Students = failing_students,
        incomplete_students = incomplete_students
    )
}
```

When we run this function, it returns a single list object. We can use the `str()` function to quickly inspect the structure of our new list and see all the different data types nested inside it:

```{r}
# Run the analysis
class_report <- analyze_class_performance(roster)

# Inspect the structure of the returned list
str(class_report, max.level = 1)

```

To access individual elements inside a list, you use the `$` operator (just like with data frames) or double brackets `[[ ]]`. Let's look at the score distribution and the failing students:

```{r}
# Access the 10-point bin distribution
class_report$ScoreDistribution

# Access the names of the failing students
class_report$Failing_Students
class_report$incomplete_students 
```


## Presenting the Results

### Base R graphics

A quick histogram shows the distribution of course scores, with a vertical line marking the class average:

```{r}
hist(results$course.score, breaks = 15, col = "lightblue",
     main = "Distribution of Course Scores", xlab = "Course Score")
abline(v = mean(results$course.score), col = "red", lwd = 2)

```

### ggplot2

**ggplot2** builds a comparison across gender declaratively, by mapping columns of `results` to visual properties and layering geometries with `+`. Here the boxplot is enhanced with the individual course scores jittered on top:

```{r}
library(ggplot2)

ggplot(results, aes(x = gender, y = course.score, fill = gender)) +
    geom_boxplot(show.legend = FALSE, alpha = 0.6) +
    geom_jitter(width = 0.15, size = 1.5, alpha = 0.6) +
    labs(
        title = "Course Score by Gender",
        x = "Gender", y = "Course Score"
    ) +
    theme_minimal()

```

We built a gender-specific intercept into the simulation when generating the final exam scores; a scatter plot of the final versus each student's mean quiz score, with a fitted line per gender, should reveal it:

```{r}
ggplot(results, aes(x = Quiz, y = Final, color = gender)) +
    geom_point(size = 2, alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE) +
    labs(
        title = "Final Exam vs. Mean Quiz Score, by Gender",
        x = "Mean Quiz Score", y = "Final Exam Score", color = "Gender"
    ) +
    theme_minimal()

```

The two fitted lines are roughly parallel but vertically offset -- exactly the intercept difference we built into `b0.final` when simulating the data.

### gt

Finally, **gt** turns a group summary of `results` into a polished report table: the average of each category mean and the average course score, by gender.

```{r}
library(gt)

summary.by.gender <- aggregate(cbind(Quiz, Test, Final, course.score) ~ gender, data = results, FUN = mean)

summary.by.gender |>
    gt(rowname_col = "gender") |>
    tab_header(
        title = "Category Means and Course Score by Gender",
        subtitle = "Averaged across all students in each group"
    ) |>
    fmt_number(columns = -gender, decimals = 1) |>
    cols_label(
        Quiz = "Quiz Mean",
        Test = "Test Mean",
        Final = "Final Exam",
        course.score = "Course Score"
    ) |>
    tab_stubhead(label = "Gender")

```
