---
title: "Random Numbers and Monte Carlo Methods"
code-fold: true
filters:
   - shinylive
---

```{r}
#| label: setup
#| echo: false
library(latex2exp)   # TeX() renders LaTeX in plot titles, axis labels and legends
```

This chapter introduces Monte Carlo methods in two stages. The first half asks how a computer produces "random" numbers at all, and how those numbers are transformed into draws from any distribution we like. The second half puts those draws to work: once we can simulate data, we can estimate almost any quantity — the risk of an estimator, the size or power of a hypothesis test, a p-value with no closed form — simply by repeating an experiment many times and averaging.

## Pseudo Random Numbers

Computers generate "random" numbers using deterministic algorithms. One of the most common methods is the **Linear Congruential Generator (LCG)**. It produces a sequence of pseudo-random integers $X_n$ defined by the recurrence relation:

$$
X_n = (a X_{n-1} + c) \pmod M
$$

where $X_0$ is the seed, $a$ is the multiplier, $c$ is the increment, and $M$ is the modulus. To get numbers in the interval $[0,1]$, we divide by $M-1$. In this example, we use the parameters $a = 7^5$, $c = 0$, and $M = 2^{31}-1$ (a minimal standard LCG).

```{r}
#| label: fig-lcg-vs-runif
#| echo: true
#| fig-cap: "A hand-rolled linear congruential generator versus R's built-in `runif()`: trace plot, autocorrelation, and histogram for each."
A <- 7^5
M <-  2^31-1

N <- 500
rn <- rep (0, N)
rn[1] <- 10 # Initial seed
for (i in 2:length (rn))
{
    rn[i] <- (A * rn[i-1] ) %% M
}

## Scale the numbers to fall between 0 and 1
nrn <- rn/(M-1)

## Compare with R's built-in runif()
n <- 500
a <- runif(n)

par(mfrow=c(2,3),mar=c(4,4,3,1))
plot(nrn[1:100], main="Trace Plot", ylab="Value", xlab="Index")
acf(nrn, main="Autocorrelation")
hist(nrn, main="Histogram")
plot.new()
hist(a,xlab="Random Numbers",main="Built-in runif()")
acf(a,main="Built-in acf")
```

## Inverting the CDF

If we can generate uniform random variables $U \sim \text{Unif}(0,1)$, we can generate a random variable $X$ from any continuous distribution using the **Inverse Transform Method**. If $F(x)$ is the cumulative distribution function (CDF) of $X$, then $X = F^{-1}(U)$ has the desired distribution.

For an Exponential distribution with rate $\lambda = 1$, the CDF is $F(x) = 1 - e^{-x}$.
Setting $U = 1 - e^{-x}$ and solving for $x$ gives:

$$
x = -\ln(1-U)
$$

## Shinylive App: Inverse CDF Sampling {#sec-inverse-cdf-app}

::: {#fig-inverse-cdf-app}

```{shinylive-r}
#| standalone: true
#| viewerHeight: 620
#| code-fold: true

library(shiny)

lambda <- 1
xmax   <- qexp(0.995, lambda)
grid   <- seq(0, xmax, length.out = 400)

ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      width = 3,
      sliderInput("n", "Points per step (n)", min = 1, max = 50, value = 10, step = 1),
      sliderInput("speed", "Play interval (ms)", min = 200, max = 2000, value = 800, step = 100),
      actionButton("draw", "Draw more samples", class = "btn-primary", width = "100%"),
      br(), br(),
      actionButton("play", "Play", width = "100%"),
      br(), br(),
      actionButton("reset", "Reset", width = "100%"),
      hr(),
      strong(textOutput("count")),
      textOutput("moments")
    ),
    mainPanel(
      width = 9,
      plotOutput("cdfPlot",  height = "280px"),
      plotOutput("histPlot", height = "260px")
    )
  )
)

server <- function(input, output, session) {

  rv <- reactiveValues(x = numeric(0), u_last = numeric(0), x_last = numeric(0))
  playing <- reactiveVal(FALSE)

  draw_step <- function() {
    u  <- runif(input$n)
    xs <- -log(1 - u) / lambda
    rv$u_last <- u
    rv$x_last <- xs
    rv$x      <- c(rv$x, xs)
  }

  observeEvent(input$draw, draw_step())

  observeEvent(input$play, {
    playing(!playing())
    updateActionButton(session, "play", label = if (playing()) "Pause" else "Play")
  })

  observeEvent(input$reset, {
    playing(FALSE)
    updateActionButton(session, "play", label = "Play")
    rv$x <- numeric(0); rv$u_last <- numeric(0); rv$x_last <- numeric(0)
  })

  observe({
    if (playing()) {
      invalidateLater(input$speed, session)
      isolate(draw_step())
    }
  })

  output$count <- renderText(paste0("Total draws: ", length(rv$x)))

  output$moments <- renderText({
    if (length(rv$x) < 2) return("")
    sprintf("sample mean = %.3f  (true = %.3f)", mean(rv$x), 1 / lambda)
  })

  output$cdfPlot <- renderPlot({
    par(mar = c(4, 4, 2.5, 1))
    plot(grid, pexp(grid, lambda), type = "n", ylim = c(0, 1),
         xlab = "x", ylab = "u = F(x)",
         main = "Draw u on the vertical axis, read x off the CDF")
    u <- rv$u_last; x <- rv$x_last
    if (length(u) > 0) {
      segments(0, u, pmin(x, xmax), u, col = "grey65")
      inside <- x <= xmax
      segments(x[inside], u[inside], x[inside], 0, col = "grey65")
    }
    lines(grid, pexp(grid, lambda), col = "steelblue", lwd = 2.5)
    if (length(u) > 0) {
      points(pmin(x, xmax), u, pch = 16, col = "firebrick", cex = 1.1)
      rug(x[x <= xmax], side = 1, col = "firebrick", lwd = 2)
      points(rep(0, length(u)), u, pch = 16, col = "grey35", cex = 0.8)
    }
  })

  output$histPlot <- renderPlot({
    par(mar = c(4, 4, 2.5, 1))
    if (length(rv$x) < 2) {
      plot.new()
      title(main = "Draw some samples to see the histogram")
      return(invisible())
    }
    hist(rv$x, breaks = 30, freq = FALSE, col = "grey85", border = "white",
         xlim = c(0, xmax), xlab = "x", ylab = "density",
         main = "Accumulated draws vs. the true Exp(1) density")
    curve(dexp(x, lambda), add = TRUE, col = "steelblue", lwd = 2.5)
    abline(v = mean(rv$x), col = "firebrick", lwd = 2, lty = 2)
    legend("topright", bty = "n", lwd = 2, lty = c(1, 2),
           col = c("steelblue", "firebrick"),
           legend = c("true density", "sample mean"))
  })
}

shinyApp(ui, server)
```

**Shinylive App Illustrating the Inverse CDF Method.** The app below draws samples one step at a time (or continuously via **Play**), showing on the top panel how each uniform draw $u$ maps through the CDF to a sample $x = F^{-1}(u)$, and on the bottom panel how those draws accumulate into a histogram against the true $\text{Exp}(1)$ density.

:::

## A Special Transformation for Generating Normal Samples

The **Box-Muller Transform** is an elegant method for generating pairs of independent Standard Normal random variables from independent Uniform random variables. It relies on converting cartesian coordinates $(X,Y)$ into polar coordinates $(R, \theta)$.

If $X, Y \sim N(0,1)$ are independent, their squared radius $R^2 = X^2 + Y^2$ follows an Exponential distribution with rate $\lambda = 1/2$, and their angle $\theta$ is uniformly distributed over $[0, 2\pi]$. By generating $R$ and $\theta$, we can map them back to Normal variables:

$$
X = R\cos(\theta) \quad \text{and} \quad Y = R\sin(\theta)
$$

```{r}
#| label: fig-box-muller
#| echo: true
#| fig-cap: "Normal samples generated by the Box-Muller transform, compared with R's built-in `rnorm()`: histogram and QQ plot for each."
gen_normal <- function(n)
{
    # calculates size of random samples (we generate pairs, so we need half)
    size_sample <- ceiling(n/2)

    # R^2 ~ Exp(1/2), so we take the square root to get R
    R <- sqrt(2*rexp(size_sample))
    theta <- runif(size_sample,0,2*pi)

    X <- R*cos(theta)
    Y <- R*sin(theta)

    c(X,Y)[1:n]
}

normal_sample <- gen_normal(1000)

## R's built-in generator for comparison
nsample2 <- rnorm (1000)

par(mfrow=c(2,2),mar=c(4,4,2,1))
hist(normal_sample,main="Custom Normal")
qqnorm(normal_sample, main="Custom QQ")
qqline(normal_sample)
hist(nsample2, main="Built-in Normal")
qqnorm(nsample2, main="Built-in QQ")
qqline(nsample2)
```

## Demonstration of CLT and LLN

The **Law of Large Numbers (LLN)** states that as the sample size $n$ grows, the sample mean $\bar{X}_n$ converges to the expected value $\mu$:
$$
\bar{X}_n \to \mu \quad \text{as} \quad n \to \infty
$$

The **Central Limit Theorem (CLT)** states that the standardized sample mean converges in distribution to a Standard Normal $N(0,1)$:
$$
Z = \frac{\bar{X}_n - \mu}{\sigma/\sqrt{n}} \xrightarrow{d} N(0,1) \quad \text{as} \quad n \to \infty
$$

Here we draw from a Gamma distribution with shape parameter $\alpha = 2$ and rate $\beta = 1$. The theoretical mean is $\mu = \alpha/\beta = 2$, and the variance is $\sigma^2 = \alpha/\beta^2 = 2$.

```{r}
#| label: fig-lln-clt-gamma
#| echo: true
#| fig-cap: "Law of Large Numbers and Central Limit Theorem for the sample mean of Gamma(2,1) draws: raw draws, their histogram, running averages converging to $\\mu=2$, and standardized running means stabilizing to $N(0,1)$."
n <- 100
rn <- rgamma(n, shape = 2)

par(mfrow = c(2,2))
plot(rn[1:100], main="Raw Gamma Draws", ylab="Value")
hist(rn, main="Histogram of Draws")

## Calculate running averages (LLN)
xbar.rep <- replicate(5, cumsum(rgamma(n, shape = 2))/(1:n))
## Calculate running standardized means (CLT)
se.rep <- replicate(5, (cumsum(rgamma(n, shape = 2))/(1:n) - 2) / sqrt(2/(1:n)))

## Plot LLN: should converge to mu = 2
plot(xbar.rep[,1], main = "Demonstrating LLN", log = "x", type = "b", ylim = c(0, 6), ylab="Sample Mean")
abline(h = 2, lwd=2)
for (i in 2:ncol(xbar.rep)) points(xbar.rep[,i], col = i, pch = i, type = "b")

## Plot CLT: should stabilize within roughly [-3, 3] standard deviations
plot(se.rep[,1], ylim = c(-3,3), type="b", log = "x", main="Demonstrating CLT", ylab="Z-Score")
for (i in 2:ncol(se.rep)) points(se.rep[,i], col = i, pch = i, type = "b")
abline(h = c(0, -2, 2), lty = c(1,2,2), lwd=2)
```

## An Example of Monte Carlo for Estimating $\pi$

We can estimate the value of $\pi$ by throwing random darts at a square with side length 2 (from $-1$ to $1$) and counting how many land inside the inscribed circle of radius $1$.

The area of the square is $(2)^2 = 4$, and the area of the circle is $\pi(1)^2 = \pi$. Therefore, the probability that a uniform random point $(X,Y)$ falls inside the circle is:
$$
P(X^2 + Y^2 \le 1) = \frac{\text{Area of Circle}}{\text{Area of Square}} = \frac{\pi}{4}
$$

#### Visualizing the Method

Before we estimate $\pi$, let's visually simulate throwing 1,000 darts:

```{r}
#| label: fig-mc-pi-darts
#| echo: true
#| fig-height: 5
#| fig-width: 5
#| fig-cap: "1,000 uniform points ('darts') thrown at the $[-1,1]^2$ square; the fraction landing inside the inscribed circle gives a Monte Carlo estimate of $\\pi$."
set.seed(123)
n_visual <- 1000
x_pts <- runif(n_visual, -1, 1)
y_pts <- runif(n_visual, -1, 1)
inside_circle <- (x_pts^2 + y_pts^2) <= 1

## Plot points
plot(x_pts, y_pts, col = ifelse(inside_circle, "steelblue", "tomato"),
     pch = 20, asp = 1, xlab = "X", ylab = "Y",
     main = paste(sum(inside_circle), "out of", n_visual, "points in circle"))

## Draw the boundaries
rect(-1, -1, 1, 1, border = "black", lwd = 2)
theta_seq <- seq(0, 2*pi, length.out = 100)
lines(cos(theta_seq), sin(theta_seq), col = "black", lwd = 2)
```

#### The Monte Carlo Estimate

If we define an indicator variable $I_i$ that equals 1 if the $i$-th dart lands in the circle and 0 otherwise, then $E[I_i] = \pi/4$.
We define $Z_i = 4 \times I_i$, which gives $E[Z_i] = \pi$.

Our Monte Carlo estimate $\hat{\pi}$ is the sample average of the $Z_i$ values:
$$
\hat{\pi} = \bar{Z} = \frac{1}{n}\sum_{i=1}^n Z_i
$$

The Standard Error (SE) measures the uncertainty of our estimate. Using the sample standard deviation $S_Z$, the 95% margin of error is calculated as:
$$
\text{Margin of Error} = 1.96 \times \text{SE} = 1.96 \times \frac{S_Z}{\sqrt{n}}
$$

#### Results by Sample Size

We encapsulate the logic in a function and track how precision improves as $n$ increases. The table below outlines our findings.

```{r}
#| label: tbl-mc-pi-by-n
#| echo: true
#| warning: false
#| tbl-cap: "Monte Carlo estimation of pi across different sample sizes."
library(knitr)

## n is the number of samples drawn uniformly from the rectangle (-1,1)  * (-1,1)
## an estimate of pi is returned
pi_est_mc <- function(n)
{
    X <- runif(n,-1,1)
    Y <- runif(n,-1,1)

    # 4 times the indicator function
    Z <- 4 * (X^2 + Y^2 <= 1)

    mu <- mean(Z)
    error <- 1.96 * sd(Z) / sqrt(n)

    list(n = n, pi.est = mu, error.95perc = error, ci.lower = mu - error, ci.upper = mu + error)
}

## Run the estimation for varying sample sizes
sample_sizes <- c(100, 10000, 100000, 10000000)
results_list <- lapply(sample_sizes, pi_est_mc)

## Combine results into a dataframe for tabular display
results_df <- data.frame(
  N = sapply(results_list, function(x) x$n),
  Estimate = sapply(results_list, function(x) x$pi.est),
  Error_Margin = sapply(results_list, function(x) x$error.95perc),
  CI_Lower = sapply(results_list, function(x) x$ci.lower),
  CI_Upper = sapply(results_list, function(x) x$ci.upper)
)

## Render the formatted table
kable(results_df,
      col.names = c("Sample Size (N)", "Estimated Pi", "95% Margin of Error", "Lower CI", "Upper CI"),
      digits = 6,
      caption = "Monte Carlo Estimation of Pi across different sample sizes")
```

## Point Estimation

Assume data $X = (X_1, \ldots, X_n)$ is drawn from a distribution $f(x; \theta)$ indexed by an unknown parameter $\theta \in \Theta$. An estimator $\hat\theta = T(X)$ is a function derived exclusively from the data. Because it relies on random samples, $\hat\theta$ is itself a random variable. Its behavior is defined by its sampling distribution.

The two fundamental summaries of an estimator's performance are bias and variance:


$$\mathrm{bias}(\hat\theta) = E_{\theta}(\hat\theta) - \theta, \qquad \mathrm{Var}_{\theta}(\hat\theta) = E_{\theta}\left[\left(\hat\theta - E_{\theta}(\hat\theta)\right)^{2}\right].$$

The most comprehensive single-number summary is the Mean Squared Error (MSE), defined as $\mathrm{MSE}_{\theta}(\hat\theta) = E_{\theta}[(\hat\theta - \theta)^{2}]$. By adding and subtracting $E_{\theta}(\hat\theta)$ inside the square, the cross term vanishes, yielding the MSE decomposition:


$$\mathrm{MSE}_{\theta}(\hat\theta) = \mathrm{Var}_{\theta}(\hat\theta) + \left[\mathrm{bias}_{\theta}(\hat\theta)\right]^{2}.$$

**Key Insight:** Unbiasedness is not strictly necessary. An estimator that introduces a small bias in exchange for a massive reduction in variance will frequently achieve a lower MSE than a strictly unbiased competitor.

Because MSE is a *function* of $\theta$, it acts as the risk function under squared-error loss. Two estimators rarely have one absolute winner; their risk curves often cross. To choose a single estimator globally, statisticians summarize the risk curve via two main criteria:

* **Minimax:** Minimize the worst-case scenario: $\sup_{\theta \in \Theta} \mathrm{MSE}_{\theta}(\hat\theta)$
* **Bayes Risk:** Average the error against a prior distribution $\pi$: $\int_{\Theta} \mathrm{MSE}_{\theta}(\hat\theta)\, \pi(\theta)\, d\theta$

Asymptotically, $\hat\theta_n$ is consistent if it converges to $\theta$ in probability. Under regularity conditions, Maximum Likelihood Estimators (MLEs) are consistent and asymptotically normal, achieving the Cramér–Rao lower variance bound $1/I(\theta)$. However, in small samples, the MLE can often be outperformed.

### Monte Carlo Evaluation of an Estimator

Risk functions are rarely available in closed form, but because they are expectations, we can estimate them efficiently using Monte Carlo simulation.

**The Monte Carlo Recipe:**

1. Define a grid of $\theta$ values.
2. For each $\theta$, generate $M$ independent datasets.
3. Compute the estimate $\hat\theta^{(m)}$ for each dataset.
4. Average the squared errors:

$$\widehat{\mathrm{MSE}}_{\theta} = \frac{1}{M} \sum_{m=1}^{M} \left(\hat\theta^{(m)} - \theta\right)^{2}$$



Because this represents an average of independent, identically distributed (i.i.d.) quantities, its standard error is $s/\sqrt{M}$, where $s^{2}$ is the sample variance of the squared errors. Always report the Monte Carlo error boundaries ($\widehat{\mathrm{MSE}}_{\theta} \pm 1.96\, s/\sqrt{M}$) to distinguish true differences in risk curves from mere simulation noise.

### Two Estimators of Proportions

Consider estimating a binomial proportion $p$ from $n$ independent trials.

* **Maximum Likelihood Estimator ($\hat{p}_1$):** The sample mean, $\hat{p}_1 = \bar{X}$. This is unbiased but exhibits high variance in small samples, especially near boundaries ($p \approx 0$ or $1$).
* **Smoothed Estimator ($\hat{p}_2$):** Often called the Laplace or "add-one-in" estimator, $\hat{p}_2 = (\sum_i X_i + 1)/(n + 2)$. This introduces bias by shrinking the estimate toward $1/2$ (the posterior mean under a uniform prior), but reduces variance by a factor of $[n/(n+2)]^2$.

We can estimate the risk curves on a grid of $p$ values using the Monte Carlo method. Both estimators rely strictly on the sufficient statistic $S = \sum_i X_i \sim \mathrm{Binomial}(n, p)$, allowing us to vectorize the simulation without looping over replicates.

```{r}
#| label: prop-estimators
###Estimators, written in terms of the sufficient statistic S = sum(X_i)
p_est1 <- function(s, n) s / n
p_est2 <- function(s, n) (s + 1) / (n + 2)

### Monte Carlo estimate of the MSE and its standard error.  All no_sim
### replicates are generated in a single vectorized draw.
mse_est_mc <- function(n, p, no_sim, p_est)
{
    sq_error <- (p_est(rbinom(no_sim, size = n, prob = p), n) - p)^2
    c(mse = mean(sq_error), sd = sqrt(var(sq_error) / no_sim))
}

### MSE over a grid of p values, returned as a 2 by length(p_set) matrix
mse_curve <- function(n, p_set, no_sim, p_est)
    vapply(p_set, function(p) mse_est_mc(n, p, no_sim, p_est), c(mse = 0, sd = 0))

```

```{r}
#| label: fig-prop-mse-plot
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Simulated MSE (with 95% Monte Carlo error bands) of $\\hat p_1 = S/n$ versus the smoothed $\\hat p_2 = (S+1)/(n+2)$, across $n = 2, 10, 50, 100$."
no_sim <- 10000
p_set <- seq(0, 1, by = 0.05)
n_set <- c(2, 10, 50, 100)

par(mfrow = c(2, 2), mar = c(4, 4.5, 2.5, 1))

for (n in n_set)
{
    r1 <- mse_curve(n, p_set, no_sim, p_est1)
    r2 <- mse_curve(n, p_set, no_sim, p_est2)

    bands <- cbind(r1["mse", ], r1["mse", ] + 1.96 * r1["sd", ], r1["mse", ] - 1.96 * r1["sd", ],
                   r2["mse", ], r2["mse", ] + 1.96 * r2["sd", ], r2["mse", ] - 1.96 * r2["sd", ])

    matplot(p_set, bands, type = "l", lty = c(1, 2, 2, 1, 2, 2), col = rep(c(1, 2), each = 3),
            xlab = TeX(r"($p$)"),
            ylab = TeX(r"($E[(\hat{p} - p)^2]$)"),
            main = TeX(sprintf(r"($n = %d$)", n)))

    legend("top", bty = "n", lty = 1, col = c(1, 2), cex = 0.9,
           legend = TeX(c(r"($\hat{p}_1 = S/n$)", r"($\hat{p}_2 = (S+1)/(n+2)$)")))
}

```

When $n$ is small, the smoothed estimator significantly outperforms the MLE across most of the parameter space. As $n$ grows, the curves converge, reflecting the asymptotic efficiency of the MLE. The exception occurs strictly at the extremes ($p \approx 0$ or $1$), where the shrinkage penalty overrides the variance reduction, making $\hat{p}_1$ superior. The confidence bands confirm this crossing is a true mathematical property, not simulation variance.

## Hypothesis Testing

A hypothesis test maps data to a binary decision regarding a null hypothesis ($H_0$) versus an alternative ($H_1$). It relies on a test statistic $T(X)$, where extreme values serve as evidence against $H_0$.

* **Type I Error:** Rejecting a true $H_0$.
* **Type II Error:** Failing to reject a false $H_0$.
* **Power Function $\beta(\theta)$:** The probability of rejecting $H_0$. It represents the Type I error rate when $\theta \in H_0$, and the statistical power when $\theta \in H_1$.
* **Size & Level:** The maximum Type I error probability across all parameters in $H_0$. A test has *level $\alpha$* if its size is at most $\alpha$.

The **p-value** repackages the test rule without locking into a fixed $\alpha$ in advance. It represents the probability of observing a test statistic at least as extreme as the one observed, assuming $H_0$ is true:


$$pv = \Pr_{H_0}\left(T(X^{*}) \ge T(x^{obs})\right)$$

A functional p-value must maintain **validity**: $\Pr_{H_0}(pv \le \alpha) \le \alpha$ for every $\alpha \in (0, 1)$. If $T$ possesses a continuous null distribution and is applied exactly, this inequality becomes an equality, meaning the p-values will follow a uniform distribution on $[0,1]$ under $H_0$.

**Diagnostic:** Simulate multiple datasets under $H_0$, compute the p-value for each, and plot the histogram.

* **Flat histogram:** The test is valid.
* **Left-skewed (mass near 0):** The test is invalid and liberal; true size exceeds nominal level.
* **Right-skewed (mass near 1):** The test is excessively conservative; sacrifices statistical power.

P-values generally fail uniformity for two reasons:

1. **Distributional Misspecification:** A foundational assumption (e.g., normality) fails, skewing the null distribution.
2. **Optimization Bias (Data-dependent procedures):** The null distribution is computed assuming a fixed procedure, but the actual procedure was chosen by inspecting the data (e.g., feature selection).

### Permutation Tests

Permutation tests offer a non-parametric alternative when analytical assumptions are dubious. If the null hypothesis states two variables are independent, the exact null distribution can be mapped without assuming normality.

Because the variables are independent under $H_0$, their labels can be scrambled without altering the joint distribution (exchangeability). The true p-value is simply the rank of the observed statistic among all permuted statistics:


$$\widehat{pv} = \frac{1 + \#\{j : T^{(j)} \ge T^{obs}\}}{N + 1}$$

The $+1$ correction counts the observed dataset alongside the $N$ permutations, ensuring $\widehat{pv}$ remains mathematically valid for any finite $N$.

### Example: The t-test on Non-Normal Observations

The standard Student's t-test assumes normally distributed data. Here, we test its robustness when exposed to heavy-tailed distributions. This highlights the first failure mode: applying the wrong null distribution to the test statistic.

We simulate data using a t-distribution across varying degrees of freedom (df). Lower df creates heavier tails, with df=1 representing the Cauchy distribution.

```{r}
#| label: t-test-functions
t.test.pvalue <- function (x, mu = 0)
{
    if (!is.numeric(x) || !is.numeric(mu) || length(mu) != 1) { stop("Invalid argument") }
    n <- length(x)
    if (n < 2) { stop("Can't do a t test with less than two data points") }

    t <- (mean(x) - mu) / sqrt(var(x) / n)
    2 * pt (-abs(t), n - 1)
}

t.test.sim <- function (N, n, df, mu.true = 0, mu = 0)
{
    X <- matrix(rt(n * N, df), n, N) + mu.true
    m <- colMeans(X)
    v <- colSums((X - rep(m, each = n))^2) / (n - 1)
    2 * pt(-abs((m - mu) / sqrt(v / n)), n - 1)
}

hist.pv <- function (pv, n, df, bins = 20)
{
    hist(pv, nclass = bins, prob = TRUE, col = "grey90",
         xlab = TeX(sprintf(r"($%d$ p-values, $n = %d$, $\nu = %d$)", length(pv), n, df)),
         main = TeX(sprintf(r"(size at $\alpha = 0.05$: $%.3f$)", mean(pv < 0.05))))
    abline(h = 1, col = 4, lwd = 2)
}

```

#### Null Distribution Diagnostics

We expect these histograms to sit perfectly flat against the blue uniform density line.

```{r}
#| label: t-test-settings
N <- 2000
bins <- 20

```

**100 Degrees of Freedom (Nearly Normal)**
With 100 df, the distribution is practically Normal. The p-values form a perfect uniform line, confirming validity.

```{r}
#| label: fig-t-test-df100
#| fig-cap: "Null distribution of t-test p-values at df = 100 (nearly normal), for $n = 2$ and $n = 20$: uniform, as expected."
par (mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
hist.pv(t.test.sim(N, 2, 100), n = 2, df = 100, bins)
hist.pv(t.test.sim(N, 20, 100), n = 20, df = 100, bins)

```

**3 Degrees of Freedom (Heavy Tails)**
Heavier tails warp the p-value distribution. The nominal Type I error rate is no longer perfectly calibrated, leading to false confidence.

```{r}
#| label: fig-t-test-df3
#| fig-cap: "Null distribution of t-test p-values at df = 3 (heavy tails), for $n = 2$ and $n = 20$: the tails distort the p-value distribution away from uniform."
par (mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
hist.pv(t.test.sim(N, 2, 3), n = 2, df = 3, bins)
hist.pv(t.test.sim(N, 20, 3), n = 20, df = 3, bins)

```

**1 Degree of Freedom (Cauchy Distribution)**
With an undefined mean and variance, extreme outliers dominate. The t-test entirely breaks down, clustering p-values near 1.0 and rendering the test virtually useless.

```{r}
#| label: fig-t-test-df1
#| fig-cap: "Null distribution of t-test p-values at df = 1 (Cauchy): the test breaks down completely, with p-values clustering near 1."
par (mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
hist.pv(t.test.sim(N, 2, 1), n = 2, df = 1, bins)
hist.pv(t.test.sim(N, 20, 1), n = 20, df = 1, bins)

```

**Increasing Sample Size ($n=30$)**
As sample size scales, the Central Limit Theorem rescues the underlying math. The p-value distributions for df=5 and df=10 flatten back into uniformity.

```{r}
#| label: fig-t-test-n30
#| fig-cap: "Null distribution of t-test p-values at $n = 30$, df = 10 and df = 5: the Central Limit Theorem restores uniformity even with heavy-tailed data."
par (mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
hist.pv(t.test.sim(N, 30, 10), n = 30, df = 10, bins)
hist.pv(t.test.sim(N, 30, 5), n = 30, df = 5, bins)

```

#### Power of the t-test

Statistical power represents the ability to successfully detect a true effect. Here, we inject a true mean of $\mu = 0.5$ while testing against $H_0: \mu = 0$.

```{r}
#| label: hist-power-fn
hist.power <- function (pv, n, df, mu.true)
    hist(pv, nclass = 20, col = "grey90",
         xlab = TeX(sprintf(r"($%d$ p-values, $n = %d$, $\nu = %d$, $\mu = %.1f$)",
                            length(pv), n, df, mu.true)),
         main = TeX(sprintf(r"(power at $\alpha = 0.05$: $%.3f$)", mean(pv < 0.05))))

```

```{r}
#| label: fig-t-test-power-df5
#| fig-cap: "Power of the t-test at df = 5 against a true mean $\\mu = 0.5$, for $n = 30$ and $n = 60$."
par (mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
hist.power(t.test.sim(N, 30, 5, mu.true = 0.5), n = 30, df = 5, mu.true = 0.5)
hist.power(t.test.sim(N, 60, 5, mu.true = 0.5), n = 60, df = 5, mu.true = 0.5)

```

```{r}
#| label: fig-t-test-power-df3
#| fig-cap: "Power of the t-test at df = 3 against a true mean $\\mu = 0.5$, for $n = 30$ and $n = 60$."
par (mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
hist.power(t.test.sim(N, 30, 3, mu.true = 0.5), n = 30, df = 3, mu.true = 0.5)
hist.power(t.test.sim(N, 60, 3, mu.true = 0.5), n = 60, df = 3, mu.true = 0.5)

```

### Example: A Permutation Test for Correlation

When evaluating the correlation between variables $X$ and $Y$, dropping the analytical normality assumption in favor of a permutation test avoids the first failure mode entirely.

**The Procedure:**

1. Calculate the observed correlation $\rho^{obs} = \hat\rho(X, Y)$.
2. For $j = 1, \ldots, N$: randomly shuffle $Y$ to create $Y^{(j)}$, then calculate $\rho^{(j)} = \hat\rho(X, Y^{(j)})$.
3. Calculate the permutation p-value as the proportion of shuffled correlations whose absolute value meets or exceeds the observed correlation.

```{r}
#| label: perm-cor
perm.cor.test <- function (x, y, no.perm = 100)
{
    Yp <- replicate(no.perm, sample(y))          # n by no.perm permuted responses
    permuted.abs.cor <- abs(as.vector(cor(x, Yp)))
    real.cor <- cor(x, y)

    list (rho = real.cor,
          pv = (1 + sum(permuted.abs.cor >= abs(real.cor))) / (no.perm + 1),
          perm.abs.cor = permuted.abs.cor)
}

```

#### Performance on Non-Normal Dependent Data

We generate non-normal exponential variables with a direct dependency. The observed $\vert{}\hat\rho\vert{}$ rests squarely in the far right tail of the randomized permutation distribution.

```{r}
#| label: fig-perm-cor-illustration
#| cache: true
#| fig-width: 8
#| fig-height: 4.5
#| fig-cap: "Permutation distribution of $|\\hat\\rho|$ for non-normal, dependent data, with the observed correlation marked in red: it sits far in the right tail."
par (mfrow = c(1, 1), mar = c(4, 4, 3, 1))

x <- rexp(30); y <- 0.4 * x + rexp(30)     # non-normal, dependent
out <- perm.cor.test(x, y, no.perm = 2000)

hist(out$perm.abs.cor, breaks = 40, col = "grey90",
     xlab = TeX(r"($|\hat{\rho}|$ from permuted data)"),
     main = TeX(sprintf(r"(Permutation distribution of $|\hat{\rho}|$; $\hat{\rho}^{obs} = %.3f$)",
                        out$rho)))
abline(v = abs(out$rho), col = 2, lwd = 2)

c(rho_obs = out$rho, perm_pv = out$pv, pearson_pv = cor.test(x, y)$p.value)

```

#### Evaluating Null Validity

Under an identical, normal distribution, the permutation test perfectly mirrors a standard analytical test, maintaining true uniformity under $H_0$.

```{r}
#| label: fig-perm-cor-normal-h0
#| cache: true
#| fig-width: 8
#| fig-height: 4.5
#| fig-cap: "Null distribution of the permutation-test p-value for correlation, normal data: perfectly uniform, as with a standard analytical test."
test.perm.norm <- function (no.sim, no.perm, n = 20)
    replicate(no.sim, perm.cor.test(rnorm(n), rnorm(n), no.perm)$pv)

pv <- test.perm.norm(2000, 100)

par (mfrow = c(1, 1), mar = c(4, 4, 3, 1))
hist(pv, 20, prob = TRUE, col = "grey90",
     xlab = TeX(r"($2000$ permutation p-values, normal data)"),
     main = TeX(sprintf(r"(size at $\alpha = 0.05$: $%.3f$)", mean(pv < 0.05))))
abline(h = 1, col = 4, lwd = 2)

```

Crucially, when testing non-normal, entirely independent data, the permutation test remains flawless. The standard Pearson test (relying on a normal approximation) struggles, while the permutation test yields a perfect uniform distribution without needing normality.

```{r}
#| label: fig-perm-cor-nonnormal-h0
#| cache: true
#| fig-cap: "Null distribution of p-values for independent, non-normal data: the permutation test (left) stays uniform while the Pearson test (right) struggles."
res <- replicate(500, {
  x <- rexp(30); y <- rexp(30)                    # independent, non-normal
  c(perm    = perm.cor.test(x, y, no.perm = 500)$pv,
    pearson = cor.test(x, y)$p.value)
})

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

for (m in c("perm", "pearson")) {
  hist(res[m, ], breaks = 20, prob = TRUE, col = "grey90",
       xlab = TeX(sprintf(r"($500$ %s p-values under $H_0$)", m)),
       main = TeX(sprintf(r"(size at $\alpha = 0.05$: $%.3f$)", mean(res[m, ] < 0.05))))
  abline(h = 1, col = 4, lwd = 2)
}

```

### Example: Variable Selection and Optimization Bias

This section addresses the second failure mode: calculating a p-value as if the model was fixed in advance, when the model was actually chosen by actively querying the data (p-hacking).

**The Scenario:** You have a response $Y$ and 20 candidate predictors, $X_1 \dots X_{20}$, with a sample size of $n=50$. You test the null hypothesis that $Y$ has no relationship with any $X$.

A deeply flawed but common practice is to:

1. Scan for the single predictor $X_{k^*}$ that happens to correlate highest with $Y$.
2. Run a standard linear regression $Y \sim X_{k^*}$.
3. Report the standard t-test p-value for the slope.

```{r}
#| label: test-lm-sel
test_lm_sel <- function(y, X) {
  k <- which.max(abs(cor(X, y)))
  summary(lm(y ~ X[, k]))$coefficients[2, 4]    # t-test p-value of selected slope
}

```

```{r}
#| label: test-lm-sel-fast
### two-sided p-value of the slope in a simple linear regression, vectorized in r
cor_pvalue <- function(r, n) 2 * pt(-abs(r * sqrt((n - 2) / (1 - r^2))), n - 2)
test_lm_sel_fast <- function(y, X) cor_pvalue(max(abs(cor(X, y))), length(y))
sim_null <- function(n = 50, p = 20) list(y = rnorm(n), X = matrix(rnorm(n * p), n, p))

d <- sim_null()
c(lm = test_lm_sel(d$y, d$X), fast = test_lm_sel_fast(d$y, d$X))   # identical

```

By intentionally fishing for the maximum correlation first, the naive test is overwhelmingly likely to discover "significant" correlations strictly out of random noise.

```{r}
#| label: fig-test-lm-sel-h0
#| cache: true
#| fig-width: 8
#| fig-height: 4.5
#| fig-cap: "Null distribution of the naive p-value after picking the best-correlated of 20 predictors: massively inflated near 0, showing the false-positive cascade from optimization bias."
pv_sel <- replicate(2000, { d <- sim_null(); test_lm_sel_fast(d$y, d$X) })

par (mfrow = c(1, 1), mar = c(4, 4, 3, 1))
hist(pv_sel, breaks = 20, prob = TRUE, col = "grey90",
     xlab = TeX(r"($2000$ naive p-values after selection, $H_0$ true)"),
     main = TeX(sprintf(r"(size at $\alpha = 0.05$: $%.3f$)", mean(pv_sel < 0.05))))
abline(h = 1, col = 4, lwd = 2)

```

As the histogram proves, the false-positive rate under this naive approach explodes far above the accepted 0.05 threshold.

#### The Permutation Fix

To repair the p-value, we must embed the *entire selection process* inside the permutation loop.

1. Record the naive p-value from the observed data.
2. Shuffle $Y$, re-select the best predictor, and compute a new p-value. Repeat $N$ times.
3. The true valid p-value is the proportion of permuted runs that beat the original observed run.

```{r}
#| label: perm-lm-sel
perm_lm_sel <- function(y, X, N = 500) {
  n <- length(y)
  pobs <- test_lm_sel_fast(y, X)

  Yp <- replicate(N, sample(y))            # n by N permuted responses
  rmax <- apply(abs(cor(X, Yp)), 2, max)   # selection redone in each column
  pperm <- cor_pvalue(rmax, n)

  (1 + sum(pperm <= pobs)) / (N + 1)
}

```

```{r}
#| label: fig-perm-lm-sel-size-power
#| cache: true
#| fig-width: 8
#| fig-height: 4.5
#| fig-cap: "Null distribution of the permutation-corrected p-value after variable selection: uniform again, restoring valid size."
sim_alt <- function(n = 50, p = 20, beta = 0.5) {
  X <- matrix(rnorm(n * p), n, p); list(y = beta * X[, 1] + rnorm(n), X = X)
}

pv_perm0 <- replicate(300, { d <- sim_null(); perm_lm_sel(d$y, d$X) })
pv_perm1 <- replicate(300, { d <- sim_alt();  perm_lm_sel(d$y, d$X) })
pv_sel1  <- replicate(300, { d <- sim_alt();  test_lm_sel_fast(d$y, d$X) })

par (mfrow = c(1, 1), mar = c(4, 4, 3, 1))
hist(pv_perm0, breaks = 20, prob = TRUE, col = "grey90",
     xlab = TeX(r"($300$ permutation p-values, $H_0$ true)"),
     main = TeX(sprintf(r"(size at $\alpha = 0.05$: $%.3f$)", mean(pv_perm0 < 0.05))))
abline(h = 1, col = 4, lwd = 2)

rbind(test_lm_sel = c(size = mean(pv_sel < 0.05),   power = mean(pv_sel1 < 0.05)),
      permutation = c(size = mean(pv_perm0 < 0.05), power = mean(pv_perm1 < 0.05)))

```

The permutation distribution restores the flat, uniform distribution. Because power can only be honestly compared between tests of the same size, the permutation test provides the mathematically sound baseline.

### Example: A Power Regression Test

The same optimization bias occurs when iterating through continuous power transformations (e.g., $X^{0.5}, X^2$) to maximize correlation between $X^q$ and $Y$. Because the selection rule adapts to the data, applying standard analytical p-values directly to the final "optimized" model will yield artificially inflated significance.

We employ the same fix: encapsulate the entire search process inside the permutation randomization.

```{r}
#| label: power-regression-functions
pow.matrix <- function (x, powers = seq(0.1, 4.0, by = 0.1))
{
    if (!is.vector(x, "numeric")) stop("Argument should be a numeric vector")
    if (any(x < 0)) stop("Predictors must be non-negative")
    outer(x, powers, "^")
}

pvalue.lm.pow <- function (Y, Xp)
{
    Y <- as.matrix(Y)
    rmax <- apply(abs(cor(Xp, Y)), 2, max)
    cor_pvalue(rmax, nrow(Y))
}

test.lm.power <- function (Xp, N)
    pvalue.lm.pow(matrix(rnorm(nrow(Xp) * N), nrow(Xp), N), Xp)

pvalue.perm.lm.pow <- function (y, Xp, no.perm = 100)
{
    actual <- pvalue.lm.pow(y, Xp)
    permuted <- pvalue.lm.pow(replicate(no.perm, sample(y)), Xp)
    (1 + sum(permuted <= actual)) / (no.perm + 1)
}

test.perm.power <- function (x, Xp, power.sim, N = 100, resid.std.dev = 1, no.perm = 99)
{
    n <- length(x)
    vapply(1:N,
           function(k) pvalue.perm.lm.pow(x^power.sim + rnorm(n, 0, resid.std.dev),
                                          Xp, no.perm),
           numeric(1))
}

```

#### Diagnostic Uniformity and Statistical Power

As observed previously, the naive approach causes a massive false-positive cascade (left). The permutation algorithm absorbs the optimization phase, correcting the bias and yielding perfectly valid, uniform p-values (right).

```{r}
#| label: fig-power-regression-h0
#| cache: true
#| fig-cap: "Null distribution of p-values for the best power transformation of $X$: naive selection (left) inflates the false-positive rate, while the permutation fix (right) restores uniformity."
x <- exp(rnorm(20))
Xp <- pow.matrix(x)

naive.pv <- test.lm.power(Xp, 1000)
perm.pv <- test.perm.power(x, Xp, power.sim = 0, N = 1000, no.perm = 99)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

hist(naive.pv, nclass = 20, prob = TRUE, col = "grey90",
     xlab = TeX(r"($1000$ naive p-values, $H_0$ true)"),
     main = TeX(sprintf(r"(naive test, size $= %.3f$)", mean(naive.pv < 0.05))))
abline(h = 1, col = 4, lwd = 2)

hist(perm.pv, nclass = 20, prob = TRUE, col = "grey90",
     xlab = TeX(r"($1000$ permutation p-values, $H_0$ true)"),
     main = TeX(sprintf(r"(permutation test, size $= %.3f$)", mean(perm.pv < 0.05))))
abline(h = 1, col = 4, lwd = 2)

```

With the size strictly regulated, we inject mathematical signals ($Y = X^{1.5} + \epsilon$ and $Y = X^{0.5} + \epsilon$) to observe statistical power dynamically capturing the non-linear true signals.

```{r}
#| label: fig-power-regression-power
#| cache: true
#| fig-cap: "Power of the permutation test to detect $Y = X^{1.5} + \\epsilon$ (left) and $Y = X^{0.5} + \\epsilon$ (right) against the null of no relationship."
perm.pv.alt1 <- test.perm.power(x, Xp, power.sim = 1.5, N = 1000, no.perm = 99)
perm.pv.alt2 <- test.perm.power(x, Xp, power.sim = 0.5, N = 1000, no.perm = 99)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

hist(perm.pv.alt1, nclass = 20, col = "grey90",
     xlab = TeX(r"($1000$ permutation p-values)"),
     main = TeX(sprintf(r"($H_1: y = x^{1.5} + \epsilon$, power $= %.3f$)",
                        mean(perm.pv.alt1 < 0.05))))

hist(perm.pv.alt2, nclass = 20, col = "grey90",
     xlab = TeX(r"($1000$ permutation p-values)"),
     main = TeX(sprintf(r"($H_1: y = x^{0.5} + \epsilon$, power $= %.3f$)",
                        mean(perm.pv.alt2 < 0.05))))

```
