---
title: Introduction to R Programming
---

# The Mission

Throughout this unit we will use one running example to introduce the basic building blocks of R: vectors, strings, special values, matrices, data frames, lists, functions, file I/O, and graphics.

**The mission:** a class of students has taken four quizzes. Each student's final grade is the average of their *best three* quiz scores -- that is, we drop each student's single lowest quiz score before averaging. To make things realistic, one student missed a quiz entirely, so we also need to handle a missing value correctly. We want to

1. work out the calculation step by step for one student,
2. package that calculation into a reusable function,
3. apply the function to every student to get a full set of results, and
4. present those results as a table and as plots.

This mirrors how real data-analysis code is usually built: explore on one case, generalize into a function, then apply and report.

# Step by Step, for One Student

## Vectors: recording quiz scores

The score on each quiz, across all students, is naturally stored as a vector. Here are the scores for four quizzes and eight students (in the same order):

```{r}
quiz1 <- c(7,8,6,9,5,8,7,6)
quiz2 <- c(8,7,5,9,6,7,8,5)
quiz3 <- c(6,9,7,8,NA,9,6,7)
quiz4 <- c(9,8,8,7,7,8,5,8)
```

Notice the `NA` in `quiz3`: the fifth student missed that quiz. Ordinary arithmetic propagates a missing value rather than silently ignoring it, which is exactly the safe default we want:

```{r}
quiz3

mean(quiz3)

mean(quiz3, na.rm = TRUE)
```

## Strings: naming the quizzes and students

We could type out `"Quiz1"`, `"Quiz2"`, ... by hand, but `paste0` builds such labels programmatically -- a habit that pays off the moment there are 40 quizzes instead of 4:

```{r}
quiz.names <- paste0("Quiz", 1:4)

quiz.names
```

The students' names are just another character vector:

```{r}
students <- c("Ava","Ben","Cleo","Dae","Ellis","Faye","Gus","Hana")

students
```

`sprintf` is useful for turning numbers back into readable text, for example a one-line summary for a single student:

```{r}
sprintf("%s's scores were: %s", students[5], paste(c(quiz1[5],quiz2[5],quiz3[5],quiz4[5]), collapse = ", "))
```

## Matrices: all students, all quizzes, at once

Binding the four quiz vectors together as columns gives a matrix with one row per student and one column per quiz. We then attach the row and column names we just built:

```{r}
scores <- cbind(quiz1,quiz2,quiz3,quiz4)

colnames(scores) <- quiz.names
rownames(scores) <- students

scores
```

A matrix is indexed as `[row, column]`. This lets us pull out one student's record, one quiz's results, or a logical selection such as "every score above 8":

```{r}
scores["Ellis", ]

scores[, "Quiz1"]

scores[scores > 8]
```

## Special values: a missing quiz

`is.na` locates missing values anywhere in an object, including a whole matrix at once, and `which(..., arr.ind = TRUE)` reports their row/column position:

```{r}
is.na(scores)

which(is.na(scores), arr.ind = TRUE)
```

It is worth knowing what happens in the extreme case where a student has *no* recorded quiz at all: taking the mean of zero numbers is not zero, it is undefined, and R represents that with `NaN` ("not a number") rather than `NA`:

```{r}
mean(numeric(0))

is.nan(mean(numeric(0)))
```

We will rely on this distinction later, so that our function can automatically flag a student with no data at all.

## Data frames: adding names and gender

A data frame lets us combine columns of different types -- here, the students' names and gender alongside their (numeric) quiz scores -- into a single table:

```{r}
gender <- factor(c("f","m","f","m","f","f","m","f"))

roster <- data.frame(name = students, gender = gender, scores)

roster
```

Because the row names of `scores` were dropped when it became columns of `roster`, we now identify students by the `name` column instead. `grep` can find, by pattern, which columns hold quiz scores -- handy since we built those names programmatically earlier:

```{r}
quiz.cols <- grep("^Quiz", names(roster))

quiz.cols

names(roster)[quiz.cols]
```

The usual data-frame subsetting tools -- `[ , ]`, `$`, and `subset` -- work as expected, and `tapply` lets us summarize a column by group, e.g. the average `Quiz1` score by gender:

```{r}
roster[roster$gender == "f", ]

subset(roster, Quiz1 > 7)

tapply(roster$Quiz1, roster$gender, mean)
```

## Lists: packaging one student's full result

We now have everything needed to work out Ellis's final grade by hand. A **list** is the natural container for the result, because it bundles pieces of different shapes -- the raw scores, the scores kept, the score dropped, and a single final number -- under one name:

```{r}
ellis <- as.numeric(roster[roster$name == "Ellis", quiz.cols])

ellis

valid <- ellis[!is.na(ellis)]

valid

i.min <- which.min(valid)

kept <- valid[-i.min]

final <- mean(kept)

result <- list(raw = ellis, kept = kept, dropped = valid[i.min], final = final)

result
```

This confirms the calculation for one student. The rest of the unit turns this scratch work into something we can reuse and apply to everyone.

# Writing the Core Function

## A loop-based implementation

Before relying on `which.min`, it is worth seeing explicitly how a computer finds a minimum: by walking through the values one at a time and remembering the smallest one seen so far. This is the same control-flow pattern (a `for` loop with an `if` inside) used throughout numerical computing:

```{r}
find.lowest <- function(x)
{
    n <- length(x)
    i.min <- 1
    x.min <- x[1]

    if(n > 1)
    {
        for(i in 2:n)
        {
            if(x[i] < x.min)
            {
                x.min <- x[i]
                i.min <- i
            }
        }
    }

    list(value = x.min, index = i.min)
}

find.lowest(valid)
```

## A vectorized implementation

R's built-in `which.min` (and, for dropping more than one value, `order`) does the same job in one call, and is what we will use going forward:

```{r}
which.min(valid)

order(valid)
```

We now wrap the full calculation -- discard missing quizzes, drop the lowest remaining score(s), average what's left -- into one function. Passing `n.drop` as an argument (rather than hard-coding "drop 1") makes the function reusable for a different grading policy later. Following the special-value discussion above, a student with no valid quiz at all correctly returns `NaN` instead of causing an error:

```{r}
compute.final.grade <- function(scores, n.drop = 1)
{
    valid <- scores[!is.na(scores)]

    if(length(valid) == 0)
    {
        return(list(raw = scores, kept = numeric(0), dropped = numeric(0), final = NaN))
    }

    #never drop every remaining score
    n.drop <- min(n.drop, length(valid) - 1)

    if(n.drop > 0)
    {
        drop.ix <- order(valid)[1:n.drop]
        dropped <- valid[drop.ix]
        kept <- valid[-drop.ix]
    }
    else
    {
        dropped <- numeric(0)
        kept <- valid
    }

    list(raw = scores, kept = kept, dropped = dropped, final = mean(kept))
}
```

Checking the function against the by-hand result for Ellis, and against the all-missing edge case:

```{r}
compute.final.grade(ellis)

compute.final.grade(c(NA,NA,NA,NA))
```

# Applying the Function to Every Student

`apply(X, 1, FUN)` calls `FUN` once per **row** of `X` -- exactly what we need to run `compute.final.grade` on every student's four quiz scores:

```{r}
final.scores <- apply(roster[, quiz.cols], 1, function(x) compute.final.grade(x)$final)

final.scores

roster$final <- final.scores
```

The dropped score for each student is extracted the same way, and both new columns are added to `roster`:

```{r}
roster$dropped <- sapply(1:nrow(roster), function(i) {
    r <- compute.final.grade(as.numeric(roster[i, quiz.cols]))
    if(length(r$dropped) == 0) NA else r$dropped
})

roster
```

It is instructive to compare this with a *fully vectorized* alternative that needs no per-student function call at all: if we did **not** need to drop a quiz, the weighted total for every student could be computed in one line via matrix multiplication of the score matrix by a vector of weights:

```{r}
w <- rep(1/4, 4)

as.matrix(roster[, quiz.cols]) %*% w
```

Notice that Ellis's entry comes back as `NA` -- matrix multiplication has no way to skip a missing value on its own. This is precisely why we wrote `compute.final.grade`: it encodes the extra logic (ignore missing quizzes, then drop the lowest) that a single matrix operation cannot express.

## Saving and reloading the results

The final `roster` is worth keeping. A data frame can be written as a plain csv file for spreadsheets, or saved as an `.RDS` file to preserve it exactly (including the `gender` factor) for future R sessions:

```{r}
write.csv(roster, file = "roster.csv", row.names = FALSE)

saveRDS(roster, file = "roster.RDS")

roster.reloaded <- readRDS("roster.RDS")

identical(roster, roster.reloaded)
```

# Presenting the Results

## Base R graphics

A quick histogram shows the distribution of final grades, with a vertical line marking the class average:

```{r}
hist(roster$final, breaks = 5, col = "lightblue",
     main = "Distribution of Final Grades", xlab = "Final Grade")
abline(v = mean(roster$final), col = "red", lwd = 2)
```

A side-by-side boxplot compares the two gender groups:

```{r}
boxplot(final ~ gender, data = roster,
        main = "Final Grade by Gender", xlab = "Gender", ylab = "Final Grade")
```

## ggplot2

**ggplot2** builds the same kind of comparison declaratively, by mapping columns of `roster` to visual properties and layering geometries with `+`. Here the boxplot is enhanced with the individual student scores jittered on top:

```{r}
library(ggplot2)

ggplot(roster, aes(x = gender, y = final, fill = gender)) +
    geom_boxplot(show.legend = FALSE, alpha = 0.6) +
    geom_jitter(width = 0.1, size = 2) +
    labs(
        title = "Final Grade by Gender",
        x = "Gender", y = "Final Grade"
    ) +
    theme_minimal()
```

We can also ask a natural follow-up question -- does a strong first quiz predict a strong final grade? -- with a scatter plot colored by gender:

```{r}
ggplot(roster, aes(x = Quiz1, y = final, color = gender)) +
    geom_point(size = 2) +
    labs(
        title = "Final Grade vs. Quiz 1 Score",
        x = "Quiz 1 Score", y = "Final Grade", color = "Gender"
    ) +
    theme_minimal()
```

## gt

Finally, **gt** turns `roster` into a polished report table instead of a plain printout: a title, grouped "Quizzes" columns, one decimal place on the computed grade, and a readable symbol for Ellis's missing quiz.

```{r}
library(gt)

roster |>
    gt(rowname_col = "name") |>
    tab_header(
        title = "Course Grade Report",
        subtitle = "Each student's lowest quiz score is dropped"
    ) |>
    sub_missing(columns = everything(), missing_text = "—") |>
    fmt_number(columns = final, decimals = 1) |>
    cols_label(
        gender = "Gender",
        dropped = "Dropped Score",
        final = "Final Grade"
    ) |>
    tab_spanner(label = "Quizzes", columns = c(Quiz1, Quiz2, Quiz3, Quiz4)) |>
    tab_stubhead(label = "Student")
```
