---
title: "EM Algorithms"
format:
  html:
    toc: true
    tbl-colwidths: true
editor_options:
  chunk_output_type: console
code-fold: true
---

```{r}
#| label: setup
#| include: false
set.seed(812)

### censored Poisson example used for the illustrations in Section 1 and
### developed fully in Section 3: m exact observations, ncen observations
### known only to satisfy y < 2
yobs_d <- c(3, 2, 4, 1, 3, 2, 5, 3)
m_d    <- length(yobs_d)
ncen_d <- 12
n_d    <- m_d + ncen_d

lobs  <- function(l) sum(dpois(yobs_d, l, log = TRUE)) + ncen_d * log(ppois(1, l))
lcomp <- function(l, z) -n_d * l + log(l) * (sum(yobs_d) + sum(z)) - sum(lfactorial(yobs_d))
Qfun  <- function(l, lt) lcomp(l, rep(lt / (1 + lt), ncen_d))
lam   <- seq(0.4, 3.2, length.out = 300)
```

## Overview of EM

### Setup and motivation

Let $Y$ denote the observed data and $\theta$ the parameter, with likelihood
$L(\theta) = P(Y \mid \theta)$. In many models this quantity is itself an
integral (or sum) over latent or missing data $Z$:

$$
P(Y \mid \theta) = \int P(Y, Z \mid \theta) \, dZ .
$$

The integral is what makes direct maximization awkward: it destroys the product
structure that would otherwise make the log-likelihood a sum of simple terms.
The EM algorithm avoids it by working with the complete-data likelihood, which
retains that structure, and correcting for the fact that $Z$ is unobserved.

| | Likelihood | Log-likelihood |
|:---|:---|:---|
| Observed | $L_{\text{obs}}(\theta) = P(Y \mid \theta)$ | $\ell_{\text{obs}}(\theta) = \log P(Y \mid \theta)$ |
| Complete | $L_{\text{comp}}(\theta) = P(Y, Z \mid \theta)$ | $\ell_{\text{comp}}(\theta; Y, Z) = \log P(Y, Z \mid \theta)$ |

: Fixes the notation for the observed-data and complete-data likelihoods used throughout the chapter. Observed and complete likelihoods {#tbl-terminology}

### The algorithm

Start from $\hat\theta^{(0)}$. For $t = 0, 1, 2, \ldots$:

**E-step.** Replace $\ell_{\text{obs}}(\theta)$ by the expected complete
log-likelihood,

$$
Q\!\left(\theta \mid \hat\theta^{(t)}\right)
= E_{Z \mid Y, \hat\theta^{(t)}}\!\left[\ell_{\text{comp}}(\theta; Y, Z)\right].
$$

**M-step.** Maximize it,

$$
\hat\theta^{(t+1)} = \arg\max_{\theta} \; Q\!\left(\theta \mid \hat\theta^{(t)}\right).
$$

Iterate until $\hat\theta^{(t)}$ converges.

::: {.callout-warning}
### A common misreading
$Q$ is an *average of* $\ell_{\text{comp}}(\theta; Y, Z)$ over
$P(Z \mid Y, \hat\theta^{(t)})$. It is **not**
$\ell_{\text{comp}}\!\left(\theta; Y, E(Z \mid Y, \hat\theta^{(t)})\right)$ —
that is, EM is not "fill in the missing values with their expectations and
maximize". The two coincide only when $\ell_{\text{comp}}$ happens to be linear
in $Z$, which is true in both examples of this chapter and false in general.
:::

### $Q$ as an average of complete log-likelihoods

Fix $\hat\theta^{(t)}$. Each draw $Z^{(s)} \sim P(Z \mid Y, \hat\theta^{(t)})$
gives one curve $\ell_{\text{comp}}(\theta; Y, Z^{(s)})$, and
$Q(\theta \mid \hat\theta^{(t)})$ is their expectation. The figure uses the
censored Poisson model of @sec-poisson with $\lambda^{(t)} = 1.5$.

```{r}
#| label: fig-q-average
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Illustrates the $Q$ function of the EM algorithm as the expectation of the complete-data log-likelihood over the unobserved data, approximated here by averaging simulated completions.
#|   Ten complete-data log-likelihood curves, their average, and the exact $Q$ function."
lt <- 1.5; S <- 10
Z <- replicate(S, rbinom(ncen_d, 1, lt / (1 + lt)))
curves <- sapply(1:S, function(s) sapply(lam, lcomp, z = Z[, s]))
Qexact <- sapply(lam, Qfun, lt = lt)
par(mar = c(4, 4, 1, 1))
matplot(lam, curves, type = "l", lty = 1, lwd = 1, col = "gray60",
        xlab = expression(theta), ylab = "log-likelihood",
        ylim = range(curves) + c(0, 2))
lines(lam, rowMeans(curves), lwd = 3, col = "firebrick")
lines(lam, Qexact, lwd = 3, col = "black", lty = 2)
abline(v = lt, lty = 3)
legend("bottomright", bty = "n", lwd = c(1, 3, 3), lty = c(1, 1, 2),
       col = c("gray60", "firebrick", "black"),
       legend = c(expression(l[comp](theta*";"*Y*","*Z^(s)) ~ "for 10 draws"),
                  "average of the 10 curves",
                  expression(Q(theta*"|"*hat(theta)^(t)) ~ "(exact expectation)")))
```

### $Q$ functions as tangent lower bounds

The geometric picture that explains why EM works: each
$Q(\cdot \mid \hat\theta^{(t)})$, shifted by a constant, lies **below**
$\ell_{\text{obs}}$ and **touches** it at $\hat\theta^{(t)}$. Maximizing the
bound therefore moves uphill on $\ell_{\text{obs}}$ itself.

```{r}
#| label: fig-tangent-bound
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Illustrates why EM increases the likelihood: each iteration maximizes a lower bound that touches the observed log-likelihood at the current estimate.
#|   Three EM iterations viewed as successive tangent minorants of the observed log-likelihood."
lo <- sapply(lam, lobs); top <- max(lo)
par(mar = c(4, 4, 1, 1))
plot(lam, lo, type = "l", lwd = 3, xlab = expression(theta),
     ylab = "log-likelihood", ylim = c(top - 16, top + 2.5))
lt <- 0.5; cols <- c("firebrick", "darkorange", "forestgreen")
for (k in 1:3) {
  shift <- lobs(lt) - Qfun(lt, lt)            # make Q touch l_obs at theta^(t)
  lines(lam, sapply(lam, Qfun, lt = lt) + shift, col = cols[k], lwd = 2, lty = 2)
  points(lt, lobs(lt), pch = 19, col = cols[k])
  lt_new <- (sum(yobs_d) + ncen_d * lt / (1 + lt)) / n_d     # M-step
  segments(lt_new, top - 16, lt_new, lobs(lt_new), col = cols[k], lty = 3)
  text(lt, lobs(lt), bquote(hat(theta)^(.(k - 1))), pos = 3, col = cols[k], xpd = NA)
  lt <- lt_new
}
legend("bottomright", bty = "n", lwd = c(3, 2), lty = c(1, 2),
       legend = c(expression(l[obs](theta)),
                  expression(Q(theta*"|"*hat(theta)^(t)) + "const")))
```

### The ascending property

Write $h(Y, Z \mid \theta) = P(Y, Z \mid \theta)$,
$g(Y \mid \theta) = \int h(Y, Z \mid \theta) \, dZ$, and
$K(Z \mid Y, \theta) = h / g$.

::: {.callout-note}
### Theorem
$\ell_{\text{obs}}\!\left(\hat\theta^{(t+1)}\right) \ \ge\ \ell_{\text{obs}}\!\left(\hat\theta^{(t)}\right)$
at every iteration.
:::

**Key identity.** Since
$\log g(Y \mid \theta) = \log h(Y, Z \mid \theta) - \log K(Z \mid Y, \theta)$
holds for every $Z$, taking $E_{Z \mid Y, \theta_0}$ of both sides gives, for any
$\theta_0$,

$$
\ell_{\text{obs}}(\theta)
= \underbrace{E_{Z \mid Y, \theta_0}\!\left[\log h(Y, Z \mid \theta)\right]}_{Q(\theta \mid \theta_0)}
- \underbrace{E_{Z \mid Y, \theta_0}\!\left[\log K(Z \mid Y, \theta)\right]}_{H(\theta \mid \theta_0)} .
$$

The left side does not depend on $Z$, so the expectation leaves it unchanged;
this is the whole content of the step.

**Proof.** Set $\theta_0 = \hat\theta^{(t)}$ and subtract the identity evaluated
at $\theta = \hat\theta^{(t)}$ from the one evaluated at
$\theta = \hat\theta^{(t+1)}$:

$$
\begin{aligned}
\ell_{\text{obs}}\!\left(\hat\theta^{(t+1)}\right) - \ell_{\text{obs}}\!\left(\hat\theta^{(t)}\right)
&= \underbrace{\left[Q\!\left(\hat\theta^{(t+1)} \mid \hat\theta^{(t)}\right) - Q\!\left(\hat\theta^{(t)} \mid \hat\theta^{(t)}\right)\right]}_{\ge\, 0 \text{ by the M-step}} \\
&\quad - \underbrace{\left[H\!\left(\hat\theta^{(t+1)} \mid \hat\theta^{(t)}\right) - H\!\left(\hat\theta^{(t)} \mid \hat\theta^{(t)}\right)\right]}_{\le\, 0 \text{ by Jensen, below}} .
\end{aligned}
$$

**The $H$ term.** We need
$H(\theta \mid \hat\theta^{(t)}) \le H(\hat\theta^{(t)} \mid \hat\theta^{(t)})$
for every $\theta$, that is

$$
E_{Z \mid Y, \hat\theta^{(t)}}\!\left[\log \frac{K(Z \mid Y, \theta)}{K(Z \mid Y, \hat\theta^{(t)})}\right] \le 0 .
$$

Because $\log$ is concave, Jensen's inequality gives $E[\log W] \le \log E[W]$,
so

$$
\begin{aligned}
E_{Z \mid Y, \hat\theta^{(t)}}\!\left[\log \frac{K(Z \mid Y, \theta)}{K(Z \mid Y, \hat\theta^{(t)})}\right]
&\le \log \int \frac{K(Z \mid Y, \theta)}{K(Z \mid Y, \hat\theta^{(t)})} \, K(Z \mid Y, \hat\theta^{(t)}) \, dZ \\
&= \log \int K(Z \mid Y, \theta) \, dZ = \log 1 = 0 . \qquad \blacksquare
\end{aligned}
$$

The quantity on the left is
$-\mathrm{KL}\!\left(K(\cdot \mid Y, \hat\theta^{(t)}) \,\|\, K(\cdot \mid Y, \theta)\right)$.
Note carefully what the theorem does and does not say: the observed likelihood
never decreases, but that guarantees convergence only to a stationary point,
which may be a local maximum or even a saddle point.

### EM as alternating optimization

There is a second formulation that makes the algorithm's structure clearer and
generalizes better. Let $\tilde P(\cdot)$ be any distribution on the hidden data
$Z$, and define

$$
F\!\left(\tilde P, \theta\right)
= \int \tilde P(z) \log P(Y, z \mid \theta) \, dz
- \int \tilde P(z) \log \tilde P(z) \, dz .
$$

Using $P(Y, z \mid \theta) = P(Y \mid \theta) \, P(z \mid Y, \theta)$,

$$
F\!\left(\tilde P, \theta\right)
= \ell_{\text{obs}}(\theta) - \mathrm{KL}\!\left(\tilde P \,\|\, P(\cdot \mid Y, \theta)\right),
\qquad
\mathrm{KL}(P \| q) = \int P(z) \log \frac{P(z)}{q(z)} \, dz \; \ge\; 0 .
$$

Both steps of EM are now coordinate ascent on the single function $F$:

- **E-step** maximizes $F$ over $\tilde P$ with $\theta = \hat\theta^{(t)}$
  fixed. Since $\mathrm{KL} \ge 0$ with equality if and only if
  $\tilde P = P(\cdot \mid Y, \hat\theta^{(t)})$, the E-step sets $\tilde P$ to
  the conditional distribution of $Z$.
- **M-step** maximizes $F$ over $\theta$ with $\tilde P$ fixed. The entropy term
  is free of $\theta$, so this is $\arg\max_\theta Q(\theta \mid \hat\theta^{(t)})$.

The ascending property is then immediate: alternating maximization of a single
objective cannot decrease it. This view also explains the variational methods
used when the E-step is intractable — restrict $\tilde P$ to a manageable family
and accept a bound rather than an equality.

For completeness, $\mathrm{KL} \ge 0$ follows from $\log x \le x - 1$:

$$
\mathrm{KL}(P \| q) = -\int P(z) \log \frac{q(z)}{P(z)} \, dz
\;\ge\; -\int P(z) \left(\frac{q(z)}{P(z)} - 1\right) dz
= -\int \left[q(z) - P(z)\right] dz = 0 .
$$

## Example 1: Two-Component Normal Mixture {#sec-mixture}

### Model and likelihoods

$$
Z_i \sim \text{Bern}(p), \qquad
Y_i \mid Z_i = 1 \sim N(\mu_1, 1), \qquad
Y_i \mid Z_i = 0 \sim N(\mu_0, 1) .
$$

Observed: $y_1, \ldots, y_n$. Hidden: $z_1, \ldots, z_n$. Parameter
$\theta = (\mu_0, \mu_1, p)$.

**Complete likelihood**

$$
P(y, z \mid \theta) = \prod_{i=1}^{n} P(y_i \mid z_i, \theta) \, P(z_i \mid \theta)
= \prod_{i=1}^{n} \varphi(y_i \mid \mu_{z_i}, 1) \, p^{z_i} (1-p)^{1-z_i} .
$$

**Observed likelihood**, summing out each $z_i$,

$$
P(y \mid \theta) = \prod_{i=1}^{n}\left[\varphi(y_i \mid \mu_1, 1)\, p + \varphi(y_i \mid \mu_0, 1)(1 - p)\right] .
$$

The sum inside the product is what makes direct maximization awkward.

```{r}
#| label: fig-mixture-components
#| fig-width: 9
#| fig-height: 4.6
#| fig-cap: "Introduces the two-component normal mixture model by showing that its density is the weighted sum of the two component densities.
#|   The observed density is a sum of two weighted normal components."
par(mar = c(4, 4, 1, 1)); mu0 <- 0; mu1 <- 4; p <- 0.3
curve((1 - p) * dnorm(x, mu0), -3.5, 7.5, lwd = 2, col = "steelblue",
      ylab = "density", xlab = "y", ylim = c(0, 0.3))
curve(p * dnorm(x, mu1), add = TRUE, lwd = 2, col = "firebrick")
curve((1 - p) * dnorm(x, mu0) + p * dnorm(x, mu1), add = TRUE, lwd = 3)
legend("topright", bty = "n", lwd = c(2, 2, 3),
       col = c("steelblue", "firebrick", "black"),
       legend = c(expression((1-p)*phi(y*"|"*mu[0])),
                  expression(p*phi(y*"|"*mu[1])),
                  expression(f(y*"|"*theta) == "sum")))
```

The marginal density may be bimodal or unimodal depending on $p$ and
$|\mu_1 - \mu_0|$ — a point worth keeping in mind, since the unimodal case is
where EM struggles most.

```{r}
#| label: fig-mixture-shapes
#| fig-width: 10
#| fig-height: 4.4
#| fig-cap: "Shows that a mixture density can look unimodal or bimodal depending on its parameters, which affects how easily those parameters can be estimated.
#|   Separation and mixing proportion determine whether the mixture is visibly bimodal."
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
mix <- function(x, mu0, mu1, p) (1 - p) * dnorm(x, mu0) + p * dnorm(x, mu1)
for (s in list(c(0, 4, 0.5), c(0, 4, 0.15), c(0, 1.5, 0.5))) {
  curve(mix(x, s[1], s[2], s[3]), -4, 7, lwd = 3, xlab = "y", ylab = "density",
        main = bquote(mu[0] == .(s[1]) ~ "," ~ mu[1] == .(s[2]) ~ "," ~ p == .(s[3])))
  y <- ifelse(rbinom(200, 1, s[3]) == 1, rnorm(200, s[2]), rnorm(200, s[1]))
  rug(y, col = "gray40")
}
```

### Complete log-likelihood

With $\log \varphi(y \mid \mu, 1) = -\tfrac{1}{2}(y - \mu)^2 - \log\sqrt{2\pi}$,

$$
\begin{aligned}
\ell_{\text{comp}}(\theta; y, z)
&= \sum_{i=1}^{n}\left[z_i \log p + (1 - z_i)\log(1-p) - \tfrac{1}{2}(y_i - \mu_{z_i})^2\right] + C \\
&= \left(\sum_{i=1}^{n} z_i\right) \log \frac{p}{1-p} + n \log(1-p) \\
&\quad - \frac{1}{2}\sum_{i=1}^{n}\left[z_i (y_i - \mu_1)^2 + (1 - z_i)(y_i - \mu_0)^2\right] + C .
\end{aligned}
$$

Two features drive everything that follows. The expression is **linear in each
$z_i$**, so the E-step needs only $E(Z_i \mid y_i, \hat\theta^{(t)})$ rather than
the full conditional distribution. And given $z$, the parameters **separate**:
$p$ is a Bernoulli proportion while $\mu_1$ and $\mu_0$ are group means.

### E-step: soft classification

Given $\hat\theta^{(t)} = (\hat\mu_0^{(t)}, \hat\mu_1^{(t)}, \hat p^{(t)})$,
Bayes' rule gives

$$
\hat p_i^{(t)} \equiv P\!\left(Z_i = 1 \mid y_i, \hat\theta^{(t)}\right)
= \frac{\varphi\!\left(y_i \mid \hat\mu_1^{(t)}, 1\right) \hat p^{(t)}}
       {\varphi\!\left(y_i \mid \hat\mu_0^{(t)}, 1\right)\!\left(1 - \hat p^{(t)}\right)
        + \varphi\!\left(y_i \mid \hat\mu_1^{(t)}, 1\right) \hat p^{(t)}} ,
$$

and therefore

$$
\begin{aligned}
Q\!\left(\theta \mid \hat\theta^{(t)}\right)
&= \left(\sum_i \hat p_i^{(t)}\right) \log \frac{p}{1-p} + n \log(1-p) \\
&\quad - \frac{1}{2}\sum_i \left[\hat p_i^{(t)} (y_i - \mu_1)^2 + \left(1 - \hat p_i^{(t)}\right)(y_i - \mu_0)^2\right] .
\end{aligned}
$$

Computationally, $\hat p_i^{(t)}$ must be formed on the log scale. Writing
$a_i = \log \varphi(y_i \mid \hat\mu_1^{(t)}, 1) + \log \hat p^{(t)}$ and
$b_i = \log \varphi(y_i \mid \hat\mu_0^{(t)}, 1) + \log(1 - \hat p^{(t)})$, we
have $\hat p_i^{(t)} = \exp\{a_i - \operatorname{lse}(a_i, b_i)\}$ where
$\operatorname{lse}$ is the log-sum-exp function of the chapter on computer
arithmetic. Forming the ratio directly underflows as soon as an observation sits
far from both means.

```{r}
#| label: fig-estep
#| fig-width: 10
#| fig-height: 4.6
#| fig-cap: "Illustrates the E-step of EM for a mixture: the responsibility of a component is the posterior probability that an observation came from it.
#|   Left: the responsibility at each $y_i$ is the red share of the total height. Right: responsibility as a function of $y$."
par(mfrow = c(1, 2), mar = c(4, 4, 1, 1)); mu0 <- 0; mu1 <- 4; p <- 0.3
f0 <- function(x) (1 - p) * dnorm(x, mu0)
f1 <- function(x) p * dnorm(x, mu1)
curve(f0, -3.5, 7.5, lwd = 2, col = "steelblue", ylab = "weighted density",
      xlab = "y", ylim = c(0, 0.3))
curve(f1, add = TRUE, lwd = 2, col = "firebrick")
yi <- c(-0.5, 1.8, 3.2)
for (y in yi) {
  segments(y, 0, y, f0(y), col = "steelblue", lwd = 4)
  segments(y, f0(y), y, f0(y) + f1(y), col = "firebrick", lwd = 4)
  text(y, f0(y) + f1(y), sprintf("%.2f", f1(y) / (f0(y) + f1(y))), pos = 3, cex = 0.9)
}
legend("topright", bty = "n", lwd = 2, col = c("steelblue", "firebrick"),
       legend = c(expression((1-hat(p))*phi(y*"|"*hat(mu)[0])),
                  expression(hat(p)*phi(y*"|"*hat(mu)[1]))))
curve(f1(x) / (f0(x) + f1(x)), -3.5, 7.5, lwd = 3, xlab = "y",
      ylab = expression(P(Z[i] == 1*"|"*y[i]*","*hat(theta)^(t))))
abline(h = c(0, 1), lty = 3); points(yi, f1(yi) / (f0(yi) + f1(yi)), pch = 19)
```

### M-step: weighted proportions and means

$Q$ separates into one term in $p$ and two in $\mu_1$, $\mu_0$, each maximized
in closed form.

| Update | Stationary equation $\Longrightarrow$ solution |
|:---|:---|
| $p$ | $\dfrac{\partial Q}{\partial p} = \dfrac{\sum_i \hat p_i^{(t)}}{p(1-p)} - \dfrac{n}{1-p} = 0 \;\Longrightarrow\; \hat p^{(t+1)} = \dfrac{1}{n}\sum_{i=1}^{n} \hat p_i^{(t)}$ |
| $\mu_1$ | $\dfrac{\partial Q}{\partial \mu_1} = \sum_i \hat p_i^{(t)}(y_i - \mu_1) = 0 \;\Longrightarrow\; \hat\mu_1^{(t+1)} = \dfrac{\sum_i \hat p_i^{(t)} y_i}{\sum_i \hat p_i^{(t)}}$ |
| $\mu_0$ | weights $1 - \hat p_i^{(t)}$: $\;\hat\mu_0^{(t+1)} = \dfrac{\sum_i \left(1 - \hat p_i^{(t)}\right) y_i}{\sum_i \left(1 - \hat p_i^{(t)}\right)}$ |

: Derives the M-step updates for the mixture model, showing that each is a weighted version of the complete-data MLE. M-step updates for the two-component mixture {#tbl-mstep-mixture}

In words: the M-step is the complete-data MLE with each observation split
between the two groups in proportion to its soft classification. Return to the
E-step with $\hat\theta^{(t+1)}$ and recompute the responsibilities.

### Implementation

```{r}
#| label: mix-functions

log_sum_exp <- function(log_x) {
    max_log_x <- max(log_x)
    max_log_x + log(sum(exp(log_x - max_log_x)))
}

### observed log-likelihood, theta = (p, mu1, mu0)
log_like_obs <- function(theta, Y) {
    log_joint <- cbind(log(theta[1])     + dnorm(Y, theta[2], 1, log = TRUE),
                       log(1 - theta[1]) + dnorm(Y, theta[3], 1, log = TRUE))
    sum(apply(log_joint, 1, log_sum_exp))
}

### E-step: responsibilities P(Z_i = 1 | y_i, theta)
e_step <- function(theta, Y) {
    l1 <- dnorm(Y, theta[2], 1, log = TRUE) + log(theta[1])
    l0 <- dnorm(Y, theta[3], 1, log = TRUE) + log(1 - theta[1])
    exp(l1 - apply(cbind(l1, l0), 1, log_sum_exp))
}

### M-step: closed-form weighted updates
m_step <- function(w, Y) {
    c(mean(w),
      sum(Y * w) / sum(w),
      sum(Y * (1 - w)) / sum(1 - w))
}

em_mixnorm <- function(theta0, Y, no_iters) {
    out <- matrix(NA_real_, no_iters + 1, 4,
                  dimnames = list(NULL, c("p", "mu1", "mu0", "log_lik")))
    theta <- theta0
    out[1, ] <- c(theta, log_like_obs(theta, Y))
    for (i in seq_len(no_iters)) {
        theta <- m_step(e_step(theta, Y), Y)
        out[i + 1, ] <- c(theta, log_like_obs(theta, Y))
    }
    out
}
```

```{r}
#| label: fig-mix-data
#| fig-width: 6
#| fig-height: 4
#| fig-cap: "Shows the simulated data used to demonstrate EM for a mixture; the colours reveal the component labels, which are unobserved in practice and must be inferred.
#|   Simulated mixture data, coloured by the (normally unobserved) component label."

gen_mixnorm <- function(theta, n) {
    Z <- rbinom(n, 1, theta[1])
    Y <- rnorm(n, mean = ifelse(Z == 1, theta[2], theta[3]), sd = 1)
    plot(Y, Z, col = cm.colors(2)[Z + 1], pch = 19,
         ylab = "component label Z")
    Y
}

set.seed(812)
data <- gen_mixnorm(c(0.3, 0, 3), 200)
```

A well-chosen start converges quickly:

```{r}
#| label: mix-run-good
round(em_mixnorm(c(0.5, 0, 3), data, 10), 4)
```

The monotone increase of the last column is the ascending property in action;
it is the single most useful diagnostic when debugging an EM implementation,
since any decrease indicates an error in the E- or M-step.

### Shinylive App for the EM Algorithm for a Normal Mixture {#sec-mixture-app}

There is a [shinylive app](../shinyliveapps_compstat/app_mixture.html) to let you watch the EM algorithm fit a two-component normal mixture, showing how the estimates and the likelihood evolve from iteration to iteration.

### Comparison with general-purpose optimizers

EM is not the only way to maximize the observed likelihood. Applying a
general-purpose optimizer directly requires attention to the parameter space:
$p$ must lie in $(0,1)$, so we reparameterize with the logit.

```{r}
#| label: mix-compare
#| warning: false

neg_loglike_obs <- function(theta, Y) -log_like_obs(theta, Y)

### transformed parameterization: ttheta[1] = logit(p)
neg_loglike_obs_transf <- function(ttheta, Y) {
    theta <- ttheta
    theta[1] <- 1 / (1 + exp(-ttheta[1]))
    neg_loglike_obs(theta, Y)
}

lp0 <- log(0.3 / 0.7)

### Newton-type, untransformed: p can leave (0, 1)
try(nlm(neg_loglike_obs, p = c(0.3, 0, 3), Y = data)$estimate)
try(nlm(neg_loglike_obs, p = c(0.5, -10, 5), Y = data)$estimate)

### Newton-type, transformed
try(nlm(neg_loglike_obs_transf, p = c(lp0, 0, 3), Y = data)$estimate)
try(nlm(neg_loglike_obs_transf, p = c(0, -10, 5), Y = data)$estimate)

### Nelder-Mead
try(optim(c(lp0, 0, 3), neg_loglike_obs_transf, Y = data, method = "Nelder-Mead")$par)
try(optim(c(0, -10, 5), neg_loglike_obs_transf, Y = data, method = "Nelder-Mead")$par)

### Conjugate gradient
try(optim(c(0, 0, 3), neg_loglike_obs_transf, Y = data, method = "CG")$par)
try(optim(c(0, -10, 5), neg_loglike_obs_transf, Y = data, method = "CG")$par)
```

Applied to the observed log-likelihood, these algorithms are more fragile than
EM when the initial values are poorly chosen: the untransformed runs can step
outside $p \in (0,1)$ and produce `NaN`, and the gradient-based methods can stall
in the flat region that a distant start creates. Nelder–Mead is the most stable
of the three here, at the cost of many more function evaluations.

| | EM | General optimizer on $\ell_{\text{obs}}$ |
|:---|:---|:---|
| Constraints | Respected automatically by the M-step | Require reparameterization |
| Monotonicity | Guaranteed; a decrease signals a coding error | None; a bad step can be accepted |
| Derivatives | None needed | Gradient helpful, Hessian better |
| Convergence rate | Linear, rate set by the missing information | Superlinear or quadratic near the optimum |
| Standard errors | Not produced; need Louis' method or a bootstrap | Hessian available directly |
| Behaviour from poor starts | Slow but stable | Can fail outright |

: Helps decide between EM and a general-purpose optimizer by listing the strengths and weaknesses of each. EM compared with direct numerical maximization {#tbl-em-vs-optim}

The last row is the practical trade. EM rarely fails and rarely needs tuning,
which is why it dominates in mixture and latent-variable modelling; it is also
slow near the optimum and gives no standard errors, which is why a hybrid — EM
to get close, then a Newton step to polish and to obtain the Hessian — is common
in production code.

## Example 2: Censored Poisson Data {#sec-poisson}

### Truncated distributions in brief

Let $X \sim f(x)$ and $Y = I(X > c)$. Then $Y \sim \text{Bern}\{P(X > c)\}$, and
the conditional distribution of $X$ given $Y$ is $f$ restricted to one side of
$c$ and renormalized:

$$
f(x \mid Y = 1) = \frac{f(x)\, I(x > c)}{\int_c^{\infty} f(t)\,dt},
\qquad
f(x \mid Y = 0) = \frac{f(x)\, I(x \le c)}{\int_{-\infty}^{c} f(t)\,dt} .
$$

For discrete $X$ the integrals are sums. This is exactly the distribution the
E-step needs when an observation is known only to lie on one side of $c$.

```{r}
#| label: fig-truncation
#| fig-width: 10
#| fig-height: 4.8
#| fig-cap: "Illustrates the conditional distributions needed in the E-step when an observation is only known to lie on one side of a cut point.
#|   A density split at a cut point, and the two renormalized conditional densities."
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1)); cc <- 1
curve(dnorm, -3.5, 3.5, lwd = 3, xlab = "x", ylab = "density",
      main = "f(x) with cut point c")
xs <- seq(cc, 3.5, length.out = 100)
polygon(c(xs, rev(xs)), c(dnorm(xs), rep(0, 100)),
        col = adjustcolor("firebrick", 0.3), border = NA)
xs <- seq(-3.5, cc, length.out = 100)
polygon(c(xs, rev(xs)), c(dnorm(xs), rep(0, 100)),
        col = adjustcolor("steelblue", 0.3), border = NA)
abline(v = cc, lty = 2); text(cc, 0.38, "c", pos = 4)
text(1.8, 0.05, "P(X > c)", col = "firebrick")
text(-1, 0.1, expression(P(X <= c)), col = "steelblue")
curve(dnorm(x) * (x > cc) / pnorm(cc, lower.tail = FALSE), -3.5, 3.5, lwd = 3,
      col = "firebrick", n = 1001, xlab = "x", ylab = "density",
      main = "conditional densities", ylim = c(0, 1.6))
curve(dnorm(x) * (x <= cc) / pnorm(cc), add = TRUE, lwd = 3, col = "steelblue", n = 1001)
abline(v = cc, lty = 2)
legend("topleft", bty = "n", lwd = 3, col = c("steelblue", "firebrick"),
       legend = c("f(x | Y = 0)", "f(x | Y = 1)"))
```

### Data and likelihoods

Let $y_i \overset{iid}{\sim} \text{Pois}(\lambda)$ for $i = 1, \ldots, n$. The
first $m$ values are observed exactly; for $i > m$ we know only that $y_i < 2$,
recorded as $x_i = I(y_i < 2) = 1$.

**Observed data**: $y_1, \ldots, y_m$ together with
$x_{m+1} = \cdots = x_n = 1$. **Missing**: $y_{m+1}, \ldots, y_n$.

**Complete likelihood**

$$
\prod_{i=1}^{n} f^{\text{Pois}}(y_i \mid \lambda) \cdot \prod_{i=m+1}^{n} I(y_i < 2) .
$$

**Observed likelihood**

$$
\prod_{i=1}^{m} f^{\text{Pois}}(y_i \mid \lambda) \cdot \prod_{i=m+1}^{n} P(y_i < 2 \mid \lambda),
\qquad P(y_i < 2 \mid \lambda) = e^{-\lambda}(1 + \lambda) .
$$

Taking logarithms and dropping constants,

$$
\ell_{\text{obs}}(\lambda) = m \bar y_{1:m} \log \lambda - n \lambda + (n - m)\log(1 + \lambda) .
$$

The second factor makes the MLE nonlinear in $\lambda$; EM sidesteps it.

### E-step

$$
\ell_{\text{comp}}(\lambda) = -n\lambda + (\log \lambda) \sum_{i=1}^{m} y_i
+ (\log \lambda) \underbrace{\sum_{i=m+1}^{n} y_i}_{\text{missing}}
- \sum_{i=1}^{n} \log(y_i!) .
$$

For $i > m$ we have $y_i \in \{0, 1\}$, and the Poisson pmf truncated to that set
given $\lambda^{(t)}$ is

$$
P\!\left(y_i = 0 \mid x_i = 1\right) = \frac{1}{1 + \lambda^{(t)}},
\qquad
P\!\left(y_i = 1 \mid x_i = 1\right) = \frac{\lambda^{(t)}}{1 + \lambda^{(t)}} ,
$$

so that

$$
\hat y^{(t)} \equiv E\!\left(y_i \mid x_i = 1, \lambda^{(t)}\right) = \frac{\lambda^{(t)}}{1 + \lambda^{(t)}} .
$$

Since $y_i! = 1$ for $y_i \in \{0, 1\}$, the $\log(y_i!)$ term is a known
constant and no further expectation is needed.

```{r}
#| label: fig-poisson-estep
#| fig-width: 10
#| fig-height: 4.4
#| fig-cap: "Illustrates the E-step for censored Poisson counts: the missing count follows the Poisson distribution restricted to the values consistent with the censoring information.
#|   The conditional distribution of a censored count is the Poisson pmf restricted to $\\{0,1\\}$ and renormalized."
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1)); lt <- 1.5; k <- 0:6
barplot(dpois(k, lt), names.arg = k,
        col = c("firebrick", "firebrick", rep("gray80", 5)),
        xlab = "y", ylab = "probability",
        main = bquote("Pois(" * lambda^(t) == .(lt) * ")"))
barplot(c(1, lt) / (1 + lt), names.arg = 0:1, col = "firebrick",
        xlab = "y", ylab = "probability",
        main = expression("truncated to {0, 1}: " * P(y[i]*"|"*x[i] == 1)))
```

### M-step and the EM update

$$
Q\!\left(\lambda \mid \lambda^{(t)}\right)
= -n\lambda + (\log \lambda)\left[\sum_{i=1}^{m} y_i + (n - m)\,\hat y^{(t)}\right] + C^{(t)} ,
$$

$$
\frac{\partial Q}{\partial \lambda}
= -n + \frac{1}{\lambda}\left[\sum_{i=1}^{m} y_i + (n - m)\,\hat y^{(t)}\right] = 0 ,
$$

$$
\Longrightarrow \quad
\lambda^{(t+1)} = \frac{m\, \bar y_{1:m} + (n - m)\dfrac{\lambda^{(t)}}{1 + \lambda^{(t)}}}{n} .
$$

The update is the ordinary Poisson MLE — the sample mean — with each censored
value replaced by its conditional expectation. As warned in the overview, this
works *only* because $\ell_{\text{comp}}$ is linear in the missing $y_i$.

### Implementation

```{r}
#| label: poisson-em

### observed log-likelihood, up to an additive constant
##   m    : number of exactly observed counts
##   ybar : their mean
##   ncen : number of counts known only to be < 2
loglik_obs_pois <- function(m, ybar, ncen, lambda) {
    m * ybar * log(lambda) - (m + ncen) * lambda + ncen * log(1 + lambda)
}

em_censored_poisson <- function(m, ybar, ncen,
                                lambda0 = (m * ybar + ncen) / (m + ncen),
                                iterations = 20) {
    out <- matrix(NA_real_, iterations + 1, 2,
                  dimnames = list(NULL, c("lambda", "log_lik")))
    lambda <- lambda0
    out[1, ] <- c(lambda, loglik_obs_pois(m, ybar, ncen, lambda))
    for (i in seq_len(iterations)) {
        y_mis  <- lambda / (1 + lambda)                       # E-step
        lambda <- (m * ybar + ncen * y_mis) / (m + ncen)       # M-step
        out[i + 1, ] <- c(lambda, loglik_obs_pois(m, ybar, ncen, lambda))
    }
    out
}
```

```{r}
#| label: poisson-data
set.seed(812)
y    <- rpois(200, 3)
ncen <- sum(y < 2)
m    <- length(y) - ncen
ybar <- mean(y[y >= 2])
c(m = m, ncen = ncen, ybar_observed = ybar)
```

Discarding the censored observations and averaging the rest gives an estimate
clearly above the true mean of $3$: conditioning on $y \ge 2$ removes the small
counts and biases the mean upward. EM recovers them.

```{r}
#| label: poisson-run
round(em_censored_poisson(m, ybar, ncen, lambda0 = 100, iterations = 10), 5)
round(em_censored_poisson(m, ybar, ncen, lambda0 = ybar, iterations = 10), 5)
```

Both starting values reach the same limit, and the log-likelihood increases
monotonically from each. Note how slowly the run from $\lambda^{(0)} = 100$
descends at first — a single-parameter illustration of the same slow-drift
behaviour seen in the mixture app.

### Comparison with general-purpose optimizers

```{r}
#| label: poisson-compare
#| warning: false
neg_loglik_pois <- function(lambda, m, ybar, ncen) {
    -loglik_obs_pois(m, ybar, ncen, lambda)
}

try(nlm(neg_loglik_pois, p = ybar, m = m, ybar = ybar, ncen = ncen)$estimate)
try(nlm(neg_loglik_pois, p = 100,  m = m, ybar = ybar, ncen = ncen)$estimate)
try(optim(ybar, neg_loglik_pois, method = "CG",
          m = m, ybar = ybar, ncen = ncen)$par)
try(optim(100,  neg_loglik_pois, method = "CG",
          m = m, ybar = ybar, ncen = ncen)$par)
```

With one parameter and a smooth concave objective this is an easy problem for
any method; the contrast with the mixture example is instructive, since there
the multimodality and the constraint on $p$ were what made the general
optimizers fragile.

## More Complex Applications

### Model-based clustering

$$
Z_i \sim \text{Categorical}(\pi_1, \ldots, \pi_K),
\qquad Y_i \mid Z_i = k \sim N_d(\mu_k, \Sigma_k) .
$$

This is the multivariate, $K$-component generalization of @sec-mixture. The
E-step computes soft cluster memberships
$P(Z_i = k \mid y_i, \hat\theta^{(t)})$; the M-step returns weighted means,
covariances and proportions. The `mclust` package implements this with a family
of constrained covariance structures.

```{r}
#| label: fig-clustering
#| fig-width: 6
#| fig-height: 4.6
#| fig-cap: "Illustrates model-based clustering, the multivariate extension of the normal mixture model, in which EM assigns observations to clusters of different shapes.
#|   Three-component multivariate normal mixture with unequal covariances."
par(mar = c(4, 4, 1, 1))
ctr <- rbind(c(0, 0), c(4, 3), c(1, 5))
Sig <- list(diag(2) * 0.8, matrix(c(1, 0.7, 0.7, 1), 2), diag(c(1.2, 0.4)))
cols <- c("steelblue", "firebrick", "forestgreen")
ang <- seq(0, 2 * pi, length.out = 100)
plot(NA, xlim = c(-3, 7), ylim = c(-3, 8),
     xlab = expression(y[1]), ylab = expression(y[2]))
for (k in 1:3) {
  R <- chol(Sig[[k]])
  pts <- matrix(rnorm(120), 60) %*% R + rep(ctr[k, ], each = 60)
  points(pts, col = adjustcolor(cols[k], 0.6), pch = 19, cex = 0.8)
  ell <- (cbind(cos(ang), sin(ang)) * 2.1) %*% R + rep(ctr[k, ], each = 100)
  lines(ell, col = cols[k], lwd = 2)
  points(ctr[k, 1], ctr[k, 2], pch = 4, lwd = 3, col = cols[k])
}
```

### Hidden Markov models

Hidden states $h_1 \to h_2 \to \cdots \to h_n$ form a Markov chain, and each
observation depends only on its own state:

$$
P(y_{1:n} \mid h_{1:n}, \theta) = \prod_{i=1}^{n} P(y_i \mid h_i, \theta) .
$$

The E-step is the forward–backward algorithm, which computes
$P(h_i \mid y_{1:n}, \hat\theta^{(t)})$ and
$P(h_i, h_{i+1} \mid y_{1:n}, \hat\theta^{(t)})$ without enumerating all $K^n$
state sequences; the M-step produces weighted transition and emission estimates.
The combination is known as the Baum–Welch algorithm, and it is the clearest
example of an E-step that is itself a nontrivial computation.

```{r}
#| label: fig-hmm
#| fig-width: 10
#| fig-height: 3.4
#| fig-cap: "Illustrates a hidden Markov model, whose unobserved state sequence generates the data, as a further example to which EM applies.
#|   A two-state hidden Markov chain with normal emissions; background shading shows the hidden state."
par(mar = c(4, 4, 1, 1)); N <- 120
Pm <- matrix(c(0.95, 0.05, 0.1, 0.9), 2, byrow = TRUE)
h <- integer(N); h[1] <- 1
for (i in 2:N) h[i] <- sample(1:2, 1, prob = Pm[h[i - 1], ])
yy <- rnorm(N, mean = c(0, 3)[h])
plot(yy, type = "n", xlab = "i", ylab = expression(y[i]))
r <- rle(h); e <- cumsum(r$lengths); s <- c(1, head(e, -1) + 1)
rect(s - 0.5, par("usr")[3], e + 0.5, par("usr")[4],
     col = adjustcolor(c("steelblue", "firebrick")[r$values], 0.15), border = NA)
lines(yy, col = "gray30")
points(yy, pch = 19, col = c("steelblue", "firebrick")[h], cex = 0.7)
legend("topright", bty = "n", pch = 19, col = c("steelblue", "firebrick"),
       legend = c(expression(h[i] == 1), expression(h[i] == 2)), horiz = TRUE)
```

### Factor analysis and the common pattern

$$
y_i = \Gamma x_i + \varepsilon_i,
\qquad x_i \sim N_q(0, I) \text{ hidden},
\qquad \varepsilon_i \sim N_d(0, \Psi) .
$$

The E-step uses the fact that $x_i \mid y_i, \hat\theta^{(t)}$ is normal with
closed-form mean and covariance; the M-step is a regression of $y$ on the
expected factors.

Across all these applications the recipe is the same:

1. Introduce a hidden variable that makes the complete-data model tractable.
2. **E-step**: compute the conditional distribution of the hidden data, or just
   the expected sufficient statistics if $\ell_{\text{comp}}$ is linear in them.
3. **M-step**: compute the complete-data MLE with those expected sufficient
   statistics in place of the unobserved ones.

The art is entirely in step 1. The same observed-data model often admits several
choices of $Z$, and they lead to EM algorithms with quite different convergence
rates — faster when the augmentation adds little information, slower when the
missing information is a large fraction of the total.

## Summary

- EM maximizes $\ell_{\text{obs}}$ by alternating between forming
  $Q(\theta \mid \hat\theta^{(t)}) = E_{Z \mid Y, \hat\theta^{(t)}}[\ell_{\text{comp}}]$
  and maximizing it.
- Every iteration weakly increases $\ell_{\text{obs}}$. Equivalently, EM is
  coordinate ascent on
  $F(\tilde P, \theta) = \ell_{\text{obs}}(\theta) - \mathrm{KL}(\tilde P \,\|\, P(Z \mid Y, \theta))$.
- When $\ell_{\text{comp}}$ is linear in sufficient statistics of $Z$, the E-step
  reduces to computing their conditional expectations. Otherwise the full
  conditional distribution is required, and "plug in the expectation" is wrong.
- Convergence is linear, often slow, and only to a stationary point. Run from
  several starting values and compare the attained log-likelihoods.
- Standard errors are not a by-product. Use Louis' identity, the SEM algorithm,
  a bootstrap, or a final Newton step.

