---
title: "Importance Sampling"
format:
  html:
    toc: true
    tbl-colwidths: true
filters:
  - shinylive
editor_options:
  chunk_output_type: console
code-fold: true
---

```{r}
#| label: setup
#| message: false
library(ars)
```

Rejection sampling turns draws from an envelope $g \ge f$ into exact draws from
$f$, but every rejected proposal is wasted, and in the tails of a poor
envelope almost every proposal is rejected. **Importance sampling** keeps
every proposal instead of discarding most of them: draw from an easier
function $g$ and attach a *weight* to each draw that corrects for sampling
from the wrong distribution. The price is that the resulting sample is no
longer an exact iid sample from $f$ — it is a *weighted* sample — but nothing
is ever thrown away, and no envelope condition $g \ge f$ is required.

## The Reweighting Idea

### The goal

As in the last two chapters, $f(\theta)$ is an **unnormalized** function of
interest — typically $f(\theta) = L(\theta)\pi(\theta)$ for a likelihood $L$
and prior $\pi$ — and we want two kinds of integral:
$$
A = E_f\big[a(\theta)\big] = \frac{\int a(\theta)\,f(\theta)\,d\theta}{\int f(\theta)\,d\theta},
\qquad\qquad
C_f = \int f(\theta)\,d\theta,
$$
a posterior expectation and a normalizing constant (the marginal likelihood
when $f = L\pi$). Plain Monte Carlo needs draws from $f/C_f$, which we
generally cannot produce.

Suppose instead we can draw from another function $g(\theta)$ (also possibly
unnormalized, with $C_g = \int g(\theta)\,d\theta$). Wherever $f$ puts mass,
$g$ must too:
$$
\{\theta : g(\theta) > 0\} \ \supseteq\ \{\theta : f(\theta) > 0\}.
$$
Unlike rejection sampling, $g$ need **not** dominate $f$ everywhere — there is
no requirement $g \ge f$. This is what makes importance sampling applicable
in situations where no usable envelope exists.

### Two estimators

Define the **importance weight** $w(\theta) = f(\theta)/g(\theta)$. For any
$a$,
$$
E_f\big[a(\theta)\big]
= \int a(\theta)\,\frac{f(\theta)}{C_f}\,d\theta
= \frac{C_g}{C_f}\, E_g\big[a(\theta)\,w(\theta)\big],
$$
and taking $a \equiv 1$ gives $C_f/C_g = E_g[w(\theta)]$. Combining the two,
$$
E_f\big[a(\theta)\big] = \frac{E_g\big[a(\theta)\,w(\theta)\big]}{E_g\big[w(\theta)\big]}.
$$
Draw $\theta_1,\ldots,\theta_S \overset{iid}{\sim} g/C_g$ and compute
$w_s = w(\theta_s)$. Two estimators follow directly:

- **Ratio of normalizing constants**: $\widehat{C_f/C_g} = \frac1S\sum_s w_s$,
  unbiased for $C_f/C_g$; if $g$ is already a normalized density ($C_g=1$)
  this estimates $C_f$ itself.
- **Self-normalized expectation**: $\hat A = \sum_{s=1}^S a(\theta_s)\,\tilde w_s$,
  with normalized weights $\tilde w_s = w_s / \sum_j w_j$ summing to $1$. This
  turns the proposal sample into a *weighted sample from $f$*: any unknown
  constants in $f$ and $g$ cancel in the ratio, so $\hat A$ needs only $f$ and
  $g$ up to a proportionality constant. It is a ratio of two averages and so
  is biased with bias $O(1/S)$, but consistent.

### Effective sample size

$\hat A$ is a weighted average of $a(\theta_s)$: draws where $f \gg g$ count
more, draws where $f \ll g$ count less. How many *equally weighted* draws is
the weighted sample worth?
$$
S_{\text{eff}} = \frac{1}{\displaystyle\sum_{s=1}^S \tilde w_s^2}.
$$
If the weights are all equal, $\tilde w_s = 1/S$ and $S_{\text{eff}} =
1/(S \cdot 1/S^2) = S$: the proposal is shaped just like the target, and every
draw counts fully. If one weight dominates all the others, $\tilde w_s
\approx 1$ for that draw and $S_{\text{eff}} \approx 1$: only one draw is
doing any work. A variance heuristic makes this precise:
$\text{Var}_g(\hat A) \approx \text{Var}_f(a)/S_{\text{eff}}$, the same form as
for an iid sample of size $S_{\text{eff}}$ from $f$ itself.

The catch, central to this chapter, is that $S_{\text{eff}}$ is computed
**only from the weights actually observed**. A proposal that never visits the
part of $\theta$-space where $f$ and $g$ disagree most will report a
reassuring $S_{\text{eff}}$ while being badly wrong — this is exactly the
subject of the "bad stories" below.

```{r}
#| label: ess-helpers
### effective sample size from raw (possibly unnormalized) weights
ess <- function(w) sum(w)^2 / sum(w^2)
### numerically stable log(sum(exp(lx)))
log_sum_exp <- function(lx) { m <- max(lx); m + log(sum(exp(lx - m))) }
### effective sample size computed from LOG weights (avoids under/overflow)
log_ess <- function(lw) exp(2 * log_sum_exp(lw) - log_sum_exp(2 * lw))
```

## A Good Story: Estimating a Normal Tail Probability

### From plain Monte Carlo to importance sampling

For $X \sim N(0,1)$, take $f(x) = \varphi(x)\,I(a<x<b)$, the standard normal
density truncated to $[a,b]$, so that $C_f = P(a<X<b)$. Plain Monte Carlo
draws $X_i \sim N(0,1)$ and averages the indicator $I(a<X_i<b)$; this wastes
every draw that lands outside $[a,b]$; for a proposal $g$ supported on $[a,b]$
itself, every draw contributes.

The simplest such $g$ is uniform on $[a,b]$: $g(x) = I(a\le x\le b)/(b-a)$, so
$w(x) = (b-a)\varphi(x)$ and $\widehat{C_f} = (b-a)\cdot\frac1S\sum_s
\varphi(x_s)$ — the *Monte Carlo* version of the midpoint rule from the
quadrature chapter: a random grid instead of an equally spaced one, evaluating
the same weighted sum.

```{r}
#| label: tbl-tail-good
#| tbl-cap: "Mean squared error of the truncated-probability estimate over 1000 replications, for a moderate and a rare event, S = 100 draws per replication."
est_normprob_mc   <- function(A, S) { X <- rnorm(S); mean((X > A[1]) & (X < A[2])) }
est_normprob_imps <- function(A, S) { X <- runif(S, A[1], A[2]); mean(dnorm(X)) * (A[2] - A[1]) }

set.seed(1)
cases <- list(c(-2, 2), c(5, 6))
mse_tbl <- do.call(rbind, lapply(cases, function(A) {
    tp <- pnorm(A[2]) - pnorm(A[1])
    mc  <- replicate(1000, est_normprob_mc(A, 100))
    imp <- replicate(1000, est_normprob_imps(A, 100))
    data.frame(interval = sprintf("(%g, %g)", A[1], A[2]),
               true_prob = formatC(tp, format = "e", digits = 3),
               mse_mc = formatC(mean((mc - tp)^2), format = "e", digits = 3),
               mse_importance = formatC(mean((imp - tp)^2), format = "e", digits = 3))
}))
knitr::kable(mse_tbl, align = c("l", "r", "r", "r"))
```

For the moderate interval $(-2,2)$ both methods work; for the rarer event
$(5,6)$, where plain Monte Carlo almost never lands an observation inside the
interval, the importance-sampling error is many orders of magnitude smaller
— the uniform proposal *guarantees* every draw falls in $[a,b]$.

### An underflowing tail probability

Push the interval further out, to $A = (50,100)$. Now even $\varphi(50)$
itself underflows to exactly $0$ in double precision (its log is about
$-1251$, far below the smallest representable log), so every calculation must
be carried out on the log scale.

```{r}
#| label: exact-far-tail
a <- 50; b <- 100
log_minus_exp <- function(la, lb) la + log1p(-exp(lb - la))
log_exact <- log_minus_exp(pnorm(-a, log = TRUE), pnorm(-b, log = TRUE))
log_exact
```

### Matching the proposal to the tail: a better envelope

A uniform proposal on $(50,100)$ still works in principle, but $\varphi$ decays
so fast across the interval that the weights vary enormously: nearly all of
the probability mass under $\varphi$ restricted to $(50,100)$ sits within a
sliver near $x=50$, so a uniform draw almost never lands where the weight is
large. A **shifted exponential** proposal with rate $\lambda \approx a$
matches that decay and keeps the weights nearly constant:
$$
g(x) = \frac{\lambda e^{-\lambda(x-a)}}{1-e^{-\lambda(b-a)}}\,I(a\le x\le b).
$$

```{r}
#| label: tbl-tail-proposals
#| tbl-cap: "Two proposals for the same rare event, S = 10,000 draws: log-scale weights avoid the underflow that would occur if densities were multiplied directly."
S <- 1e4; lam <- a
x_u <- runif(S, a, b);  lw_u <- log(b - a) + dnorm(x_u, log = TRUE)
u_e <- runif(S); x_e <- a - log(1 - u_e * (1 - exp(-lam * (b - a)))) / lam        # inverse cdf
lg_e <- log(lam) - lam * (x_e - a) - log1p(-exp(-lam * (b - a)))
lw_e <- dnorm(x_e, log = TRUE) - lg_e

tail_tbl <- data.frame(
    proposal   = c("uniform on (a, b)", "shifted exponential"),
    log_C_f    = c(log_sum_exp(lw_u) - log(S), log_sum_exp(lw_e) - log(S)),
    S_eff      = c(log_ess(lw_u), log_ess(lw_e))
)
knitr::kable(tail_tbl, digits = c(0, 4, 1))
```

Both proposals give a log-probability close to `r round(log_exact, 3)`, but
their effective sample sizes are worlds apart: the exponential proposal keeps
essentially all `r S` draws informative, while the uniform proposal's
effective sample size collapses because almost all of its weight concentrates
on the handful of draws nearest $a$. Shaping $g$ like $f$ is not just
elegant — it is what makes the computation efficient.

## Two Bad Stories: When Reweighting Fails Silently

$S_{\text{eff}}$ diagnoses only *one* kind of failure: weights that vary
wildly among the draws actually observed. It says nothing about mass that $g$
never visits at all, and a badly wrong estimate can carry a perfectly healthy
$S_{\text{eff}}$.

### A proposal that misses a mode

Let $f(\theta) = \tfrac12\varphi(\theta+5) + \tfrac12\varphi(\theta-5)$ (a
50/50 mixture, so $C_f=1$ and $E_f(\theta)=0$) and $g(\theta) = \varphi(\theta+5)$.
Both are positive everywhere, so the support requirement holds. But $g$ places
essentially all of its mass near $\theta=-5$ and, for $\theta$ near $-5$,
$$
w(\theta) = \tfrac12 + \tfrac12\,\frac{\varphi(\theta-5)}{\varphi(\theta+5)}
= \tfrac12 + \tfrac12\, e^{10\theta} \ \approx\ \tfrac12,
$$
so every weight the sampler actually sees looks almost identical.

```{r}
#| label: fig-missed-mode
#| fig-width: 10
#| fig-height: 4.2
#| fig-cap: "Left: target f (mixture of two normals) and proposal g (only the left component). Right: normalized weights look almost flat — S_eff is close to S — even though the estimate is badly wrong."
set.seed(2)
S <- 5000
f_mix <- function(t) 0.5 * dnorm(t + 5) + 0.5 * dnorm(t - 5)
g_mix <- function(t) dnorm(t + 5)
th <- rnorm(S, -5)
w  <- f_mix(th) / g_mix(th)
wt <- w / sum(w)

par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
curve(f_mix, -9, 9, lwd = 3, col = "firebrick", xlab = expression(theta), ylab = "",
      main = "target f and proposal g", ylim = c(0, 0.42))
curve(g_mix, add = TRUE, lwd = 3, col = "steelblue")
rug(th[1:300], col = adjustcolor("steelblue", 0.5))
legend("topright", bty = "n", lwd = 3, col = c("firebrick", "steelblue"), legend = c("f", "g"))
hist(wt * S, breaks = 40, col = "gray85", border = "white", xlab = expression(S * tilde(w)[s]),
     main = sprintf("weights: S_eff = %.0f of %d", ess(w), S))
```

```{r}
#| label: tbl-missed-mode
#| tbl-cap: "Missed-mode failure: the effective sample size looks fine, but both quantities of interest are far from the truth."
missed_mode_tbl <- data.frame(
    quantity = c("C_f / C_g (true 1)", "E_f[theta] (true 0)", "S_eff (of S = 5000)"),
    estimate = c(mean(w), sum(th * wt), ess(w))
)
knitr::kable(missed_mode_tbl, digits = 3)
```

The estimate of $C_f/C_g$ comes out near `r round(mean(w), 2)` instead of $1$,
and $E_f[\theta]$ near `r round(sum(th * wt), 2)` instead of $0$ — both wrong
by a wide margin — while $S_{\text{eff}}$ reports essentially the full sample
size. The draws that would reveal the problem, near $\theta=+5$, have
probability under $g$ of order $\Phi(-10) \approx 10^{-23}$: they simply never
appear, at any practical $S$.

### A proposal narrower than the target's support

The same failure in its starkest form: $f(\theta) = I(0\le\theta\le2)$,
$g(\theta) = I(0\le\theta\le1)$. The true ratio is $C_f/C_g = 2$, but every
draw from $g$ lands in $(0,1)$, where $w(\theta)=f(\theta)/g(\theta)=1$
identically.

```{r}
#| label: tbl-support-violation
#| tbl-cap: "Estimated C_f / C_g (true value 2) as the sample size grows: no amount of additional sampling helps, because g never visits (1, 2]."
set.seed(3)
f_indicator <- function(x) as.numeric(x >= 0 & x <= 2)   # f = I(0 <= theta <= 2)
g_indicator <- function(x) as.numeric(x >= 0 & x <= 1)   # g = I(0 <= theta <= 1)
support_tbl <- data.frame(
    S = c(1e2, 1e3, 1e4, 1e5),
    estimate = sapply(c(1e2, 1e3, 1e4, 1e5), function(S) {
        th <- runif(S, 0, 1)                              # draws from g
        mean(f_indicator(th) / g_indicator(th))            # weight is 1 for every draw
    })
)
knitr::kable(support_tbl, digits = c(0, 3))
```

The estimate sits at $1$ regardless of $S$: this is a bias that more sampling
cannot fix, because it comes from where $g$ looks, not from how much it looks
there. The reverse situation — $g$ wider than $f$ — is harmless: draws outside
$\text{supp}(f)$ simply receive weight $0$ and are correctly down-weighted.
Together with the missed-mode example, the lesson is that $S_{\text{eff}}$
alone cannot certify a good proposal; the proposal must be checked against
what is actually known about where $f$ puts its mass.

### Explore it interactively {#sec-importance-app}

This app runs live
only in the HTML edition of this book; readers of other formats can follow
the link below it.

::: {#fig-importance-app .content-visible when-format="html"}

```{shinylive-r}
#| label: missed-mode-app
#| standalone: true
#| viewerHeight: 800

library(shiny)

f_mix <- function(t) 0.5 * dnorm(t + 5) + 0.5 * dnorm(t - 5)
ess   <- function(w) sum(w)^2 / sum(w^2)

ui <- fluidPage(
    titlePanel(NULL),
    sidebarLayout(
        sidebarPanel(
            width = 4,
            sliderInput("mean_g", "proposal mean", min = -9, max = 9, value = -5, step = 0.5),
            sliderInput("sd_g", "proposal sd", min = 0.5, max = 8, value = 1, step = 0.5),
            helpText("Target f is a 50/50 mixture of N(-5, 1) and N(5, 1), so the ",
                     "truth is C_f/C_g = 1 and E_f(theta) = 0 when g integrates to 1. ",
                     "A narrow proposal centred on one mode gives a deceptively healthy ",
                     "S_eff while both estimates are wrong; only a proposal wide enough ",
                     "or centred between the modes recovers the truth."),
            tableOutput("tab")
        ),
        mainPanel(
            width = 8,
            plotOutput("curves", height = "260px"),
            plotOutput("weights", height = "260px")
        )
    )
)

server <- function(input, output, session) {

    draws <- reactive({
        set.seed(1)
        th <- rnorm(5000, input$mean_g, input$sd_g)
        w  <- f_mix(th) / dnorm(th, input$mean_g, input$sd_g)
        list(th = th, w = w)
    })

    output$curves <- renderPlot({
        d <- draws()
        par(mar = c(4, 4, 2, 1))
        curve(f_mix, -12, 12, lwd = 3, col = "firebrick", xlab = expression(theta), ylab = "",
              main = "target f and proposal g", ylim = c(0, 0.45))
        curve(dnorm(x, input$mean_g, input$sd_g), add = TRUE, lwd = 3, col = "steelblue")
        rug(d$th[1:300], col = adjustcolor("steelblue", 0.5))
    })

    output$weights <- renderPlot({
        d <- draws(); wt <- d$w / sum(d$w)
        par(mar = c(4, 4, 2, 1))
        hist(wt * length(d$w), breaks = 40, col = "gray85", border = "white",
             xlab = expression(S * tilde(w)[s]),
             main = sprintf("S_eff = %.0f of %d", ess(d$w), length(d$w)))
    })

    output$tab <- renderTable({
        d <- draws(); wt <- d$w / sum(d$w)
        data.frame(quantity = c("C_f / C_g (true 1)", "E_f[theta] (true 0)", "S_eff (of 5000)"),
                   estimate = c(mean(d$w), sum(d$th * wt), ess(d$w)))
    }, digits = 3, rownames = FALSE)
}

shinyApp(ui, server)
```

**Shinylive App Illustrating Importance Sampling with a Missed Mode.** The app below reproduces the missed-mode example with an adjustable proposal:
move its mean and spread and watch the effective sample size stay high while
the estimated mean drifts away from the true value of $0$ whenever the
proposal fails to reach both components of the target.

:::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit11_importance_sampling.html#sec-importance-app>.
:::

::: {.callout-note}
### Reproducing this figure
See the callout in [Numerical Quadrature](unit09_integral.qmd) for how to
install the `shinylive` extension used above.
:::

## Estimating a Truncated-Normal Expectation

Return to a truncated $N(0,1)$ on $[l,u]$, and estimate $E\big[a(X)\mid l<X<u\big]$
for $a(x)=x^2$. Two competing samplers:

- **Importance sampling** with $g(x) = I(l\le x\le u)/(u-l)$ (uniform), giving
  $w(x) = \varphi(x)$ up to the constant $(u-l)$ that cancels in the
  self-normalized estimator.
- **Rejection sampling** via adaptive rejection sampling (`ars` package,
  applicable because $\log\varphi$ is concave), producing exact iid draws.

```{r}
#| label: tnorm-functions
### importance-sampling estimate of E[a(X) | l < X < u], X ~ N(0,1)
est_tnorm_imps <- function(a_fn, l, u, S) {
    X <- runif(S, l, u)
    W <- dnorm(X)
    ahat <- sum(a_fn(X) * W) / sum(W)
    attr(ahat, "ess") <- ess(W)
    ahat
}

### exact draws via adaptive rejection sampling (log phi is concave)
sample_tnorm_ars <- function(n, l, u) {
    logf   <- function(x) dnorm(x, log = TRUE)
    fprima <- function(x) -x
    ars(n, f = logf, fprima = fprima, x = c(l, (l + u) / 2, u),
        lb = TRUE, ub = TRUE, xlb = l, xub = u)
}
```

The original version of `sample_tnorm_ars` always drew exactly $10{,}000$
points regardless of the requested `n`; the version above respects `n`, which
matters for the timing comparison next.

```{r}
#| label: tbl-tnorm-compare
#| tbl-cap: "Importance sampling versus adaptive rejection sampling for E[X^2 | l < X < u], averaged over 100 replications at S = 1000 draws each. Both recover the midpoint-rule reference to the precision shown; timings are milliseconds per replication."
a_fn <- function(x) x^2
midpoint_ref <- function(l, u, n = 100000) {
    h <- (u - l) / n; xm <- l + (seq_len(n) - 0.5) * h
    sum(a_fn(xm) * dnorm(xm)) * h / (pnorm(u) - pnorm(l))
}

intervals <- list(c(1, 2), c(-1, 1))
tnorm_tbl <- do.call(rbind, lapply(intervals, function(iv) {
    l <- iv[1]; u <- iv[2]; ref <- midpoint_ref(l, u)
    t_imps <- system.time(EA_imps <- replicate(100, est_tnorm_imps(a_fn, l, u, 1000)))
    t_rej  <- system.time(EA_rej  <- replicate(100, mean(a_fn(sample_tnorm_ars(1000, l, u)))))
    data.frame(interval = sprintf("(%g, %g)", l, u), reference = ref,
               mean_imps = mean(EA_imps), mse_imps = mean((EA_imps - ref)^2),
               ms_imps = 1000 * t_imps[["elapsed"]] / 100,
               mean_rej = mean(EA_rej), mse_rej = mean((EA_rej - ref)^2),
               ms_rej = 1000 * t_rej[["elapsed"]] / 100)
}))
knitr::kable(tnorm_tbl, digits = c(0, 4, 4, 6, 2, 4, 6, 2))
```

Both estimators are unbiased for this smooth, bounded target and agree with
the midpoint-rule reference; importance sampling is markedly faster here
because it needs no accept/reject loop or envelope construction, at the cost
of returning a *weighted* rather than an exact sample.

## Application: Marginal Likelihood via Importance Sampling

### Setup

The [quadrature](unit09_integral.qmd) and Laplace-approximation chapters
compute the marginal likelihood $P(y) = \int L(\theta)\pi(\theta)\,d\theta$ of
the same conjugate-free Gaussian model — unknown mean $\mu$ and
$w=\log\sigma^2$, independent normal priors — by a product grid and by a
mode-and-Hessian approximation, respectively. Importance sampling offers a
third route, and one that comes with a built-in accuracy diagnostic
($S_{\text{eff}}$) that neither of the other two methods provides directly.

Take $g(\mu,w)$ to be the **Gaussian approximation to the posterior**
obtained by a Laplace fit: locate the posterior mode $(\hat\mu,\hat w)$ by
optimization, take $g = N_2\big((\hat\mu,\hat w),\, H^{-1}\big)$ for $H$ the
Hessian of $-\log[L\pi]$ at the mode, and importance-sample from $g$ with
$f = L\pi$:
$$
\widehat{P(y)} = \frac{1}{S}\sum_{s=1}^S w_s, \qquad
w_s = \frac{L(\theta_s)\pi(\theta_s)}{g(\theta_s)}, \qquad \theta_s \sim g.
$$
Because $g$ is shaped like the posterior near its mode, this is a "good
story" proposal by construction — in sharp contrast to sampling directly from
the (typically much wider, uninformative) prior, which is exactly the
`log_marlik_mc` function used elsewhere in this book as a naive-Monte-Carlo
check: it is importance sampling with $g = \pi$, and its weights are just the
likelihood values $L(\theta_s)$.

### An efficient implementation

```{r}
#| label: marlik-functions
log_lik   <- function(x, mu, w) sum(dnorm(x, mu, exp(w / 2), log = TRUE))
log_prior <- function(mu, w, mu_0, sigma_mu, w_0, sigma_w) {
    dnorm(mu, mu_0, sigma_mu, log = TRUE) + dnorm(w, w_0, sigma_w, log = TRUE)
}
neg_log_post <- function(theta, x, mu_0, sigma_mu, w_0, sigma_w) {
    -log_lik(x, theta[1], theta[2]) -
        log_prior(theta[1], theta[2], mu_0, sigma_mu, w_0, sigma_w)
}

### log density of N(mu, A^{-1}) at every column of a 2 x S matrix of points at once,
### replacing a per-point apply() call with one vectorized quadratic form
log_dmvnorm_batch <- function(Theta, mu, A) {
    Tc <- Theta - mu
    quad <- colSums((A %*% Tc) * Tc)
    0.5 * (-nrow(Theta) * log(2 * pi) + sum(log(svd(A)$d)) - quad)
}

### log marginal likelihood by importance sampling from a Laplace-shaped Gaussian
log_marlik_imps <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    fit <- nlm(neg_log_post, c(mean(x), log(stats::var(x))), hessian = TRUE,
               x = x, mu_0 = mu_0, sigma_mu = sigma_mu, w_0 = w_0, sigma_w = sigma_w)
    mu_hat <- fit$estimate; H <- fit$hessian
    L <- t(chol(solve(H)))                        # Sigma = H^{-1} = L L'
    Theta <- L %*% matrix(rnorm(2 * S), 2, S) + mu_hat

    log_g <- log_dmvnorm_batch(Theta, mu_hat, H)
    log_f <- -apply(Theta, 2, neg_log_post, x = x, mu_0 = mu_0, sigma_mu = sigma_mu,
                     w_0 = w_0, sigma_w = sigma_w)
    lw <- log_f - log_g
    structure(log_sum_exp(lw) - log(S), ess = log_ess(lw))
}

### log marginal likelihood by naive Monte Carlo from the prior (g = pi): a
### second, independent importance-sampling estimator with an unfavourable g
log_marlik_prior <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    mus <- rnorm(S, mu_0, sigma_mu); ws <- rnorm(S, w_0, sigma_w)
    lw <- vapply(seq_len(S), function(i) log_lik(x, mus[i], ws[i]), numeric(1))
    structure(log_sum_exp(lw) - log(S), ess = log_ess(lw))
}
```

Replacing the two `apply()` calls used in earlier drafts of this code — one
for the proposal log-density, one for the log-posterior — with a single
vectorized quadratic form (`log_dmvnorm_batch`) removes $S$ separate R
function calls for the proposal evaluation; only the data-dependent
log-likelihood still needs one call per draw, since it sums over the $n$
observations for each $\theta_s$ separately.

### Checking accuracy

```{r}
#| label: tbl-marlik-check
#| tbl-cap: "Log marginal likelihood by importance sampling (Laplace-shaped proposal, S = 200) across 500 replications, for two simulated data sets."
set.seed(4)
x_small <- rnorm(10);  x_large <- rnorm(200)

check_tbl <- do.call(rbind, lapply(list(n10 = x_small, n200 = x_large), function(x) {
    reps <- replicate(500, log_marlik_imps(x, 0, 5, 0, 5, 200), simplify = FALSE)
    vals <- vapply(reps, as.numeric, numeric(1))
    esss <- vapply(reps, attr, numeric(1), which = "ess")
    data.frame(mean_log_ml = mean(vals), sd_log_ml = sd(vals), mean_ess = mean(esss))
}))
check_tbl <- cbind(dataset = c("n = 10", "n = 200"), check_tbl)
knitr::kable(check_tbl, digits = 4)
```

The Monte Carlo standard deviation across replications is small relative to
the estimate itself, and the effective sample size stays close to the nominal
$S=200$ in both cases — evidence that the Laplace-shaped proposal tracks the
posterior well even though it is, after all, only a second-order
approximation to it.

### Why the proposal matters: Laplace-shaped versus prior sampling

The prior is a valid proposal for the same target — every posterior draw is
also a prior draw with positive density — so `log_marlik_prior` is a
legitimate importance-sampling estimator too. But the prior is far more
diffuse than the posterior, especially once $n$ is not tiny, so it is a much
poorer match to $f = L\pi$.

```{r}
#| label: tbl-marlik-ess-vs-n
#| tbl-cap: "Effective sample size (of S = 500 draws) for the Laplace-shaped proposal versus prior sampling, as the sample size n of the simulated data grows. Both target the same posterior for (mu, w)."
set.seed(5)
ns <- c(5, 20, 100, 400)
ess_tbl <- do.call(rbind, lapply(ns, function(n) {
    x <- rnorm(n)
    data.frame(n = n,
               ess_laplace_proposal = attr(log_marlik_imps(x, 0, 5, 0, 5, 500), "ess"),
               ess_prior_sampling   = attr(log_marlik_prior(x, 0, 5, 0, 5, 500), "ess"))
}))
knitr::kable(ess_tbl, digits = c(0, 1, 1))
```

As $n$ grows the posterior sharpens around the truth while the prior does
not, so prior sampling's effective sample size deteriorates steadily; the
Laplace-shaped proposal, which re-centers and re-scales itself around the
posterior mode for every data set, keeps pace. This is the marginal-likelihood
instance of the same lesson as the missed-mode example above: a fixed,
data-independent proposal becomes an increasingly bad match to a target that
concentrates as more data arrive, and only a proposal that adapts to the
target avoids the resulting collapse in efficiency.

## Summary

- Importance sampling reweights draws from a proposal $g$ by
  $w(\theta) = f(\theta)/g(\theta)$ instead of accepting or rejecting them,
  using $\widehat{C_f/C_g} = \frac1S\sum_s w_s$ and the self-normalized
  $\hat A = \sum_s a(\theta_s)\tilde w_s$; only the support condition
  $\{g>0\}\supseteq\{f>0\}$ is required, not an envelope $g \ge f$.
- A good proposal shaped like the target on its support gives near-constant
  weights and $S_{\text{eff}} \approx S$ (the tail-probability and
  Laplace-proposal examples).
- A proposal that misses a mode or part of the support fails **silently**: the
  observed weights can look perfectly healthy while both $C_f/C_g$ and
  $E_f[a]$ are badly wrong, and no amount of extra sampling fixes it.
  $S_{\text{eff}}$ diagnoses only the noisy failure (wildly varying observed
  weights), never the silent one.
- For the marginal likelihood, a proposal built from a Laplace approximation
  at the posterior mode remains efficient as the data accumulate and the
  posterior sharpens, while a fixed, data-independent proposal such as the
  prior does not — reinforcing why the mode-finding methods from earlier
  chapters remain useful even once we turn to sampling-based methods.
