---
title: "Rejection Sampling"
format:
  html:
    toc: true
    tbl-colwidths: true
filters:
  - shinylive
editor_options:
  chunk_output_type: console
code-fold: true
---

```{r}
#| label: setup
#| message: false
library(ars)
```

Monte Carlo integration estimates $E_f[a(\theta)] = \int a(\theta) f(\theta)\,d\theta$ from
iid draws $\theta_1,\ldots,\theta_n \sim f$. The catch is the sampling step
itself: R has built-in samplers for standard families, but a posterior
$f(\theta) \propto L(\theta)\pi(\theta)$ is typically known only up to a
constant, and there is no generic way to draw from it directly. **Rejection
sampling** solves this by producing *exact* iid draws from $f$ using draws
from an easier distribution $g$, following the recipe *reject, reweight,
repeat*.

## The Envelope and the Algorithm

### The envelope condition

Suppose we can find a function $g(x)$ such that

- $g(x) \ge f(x)$ for every $x$ — an **envelope** of $f$, and
- we can sample from the normalized density $g(x)/\!\int g(x)\,dx$.

Neither $f$ nor $g$ needs to be normalized: $f$ may be an unnormalized
posterior and $g$ a scaled density $M\,g_0(x)$ for a sampler $g_0$ we already
have. The overall acceptance rate of the algorithm below turns out to be the
ratio of the two areas,
$$
P(\text{accept}) = \frac{\int f(x)\,dx}{\int g(x)\,dx},
$$
so a tight envelope wastes few draws and a loose one wastes many — the entire
design problem is finding a $g$ that is both easy to sample and close to $f$.

### The algorithm

**Input**: target $f$, envelope $g \ge f$ with a sampler for
$g(x)/\!\int g(x)\,dx$. For each draw needed, repeat until acceptance:

1. Draw $x \sim g(x)/\!\int g(x)\,dx$.
2. Compute $r = f(x)/g(x)$.
3. Draw $U \sim \text{Unif}(0,1)$.
4. If $U < r$, accept $x$; otherwise return to step 1.

Only the ratio $f/g$ is ever evaluated, so unknown normalizing constants in
$f$ (and in $g$) cancel automatically as long as the *versions actually used*
satisfy $g \ge f$. The number of proposals needed per accepted draw is
geometric with mean $1/P(\text{accept})$.

### An equivalent geometric picture

An illuminating restatement: draw $X \sim g$, then draw a height
$V \sim \text{Unif}(0, g(X))$ (equivalently $V = U\,g(X)$), and accept $X$ if
$V < f(X)$. This is the same algorithm, since
$V < f(X) \iff U < f(X)/g(X)$, but it exposes *why* rejection sampling works:

- Steps "draw $X\sim g$, then $V\sim\text{Unif}(0,g(X))$" place $(X,V)$
  **uniformly** in the region under $g$, $G = \{(x,v): 0<v<g(x)\}$.
- Keeping only points with $v < f(x)$ restricts to the sub-region under $f$,
  $F = \{(x,v): 0<v<f(x)\} \subset G$.
- A uniform point on $G$, conditioned on landing in $F$, is uniform on $F$.
  The $x$-coordinate of a uniform point on $F$ has density proportional to
  the *height* of $F$ at $x$, which is $f(x)$. That is exactly a draw from
  $f/\!\int f$.

```{r}
#| label: fig-envelope-scatter
#| fig-width: 10
#| fig-height: 4.4
#| fig-cap: "Left: target $f(x)=e^{-x^2/2}$ (unnormalized standard normal) under
#|   its Laplace envelope $g(x) = e^{1/2-|x|}$, with the two areas shaded.
#|   Right: 600 candidate points drawn uniformly under $g$; the accepted ones
#|   (red, under $f$ too) are uniformly distributed under $f$, so their
#|   $x$-coordinates are draws from the normalized $f$."
f_demo <- function(x) exp(-x^2 / 2)
g_demo <- function(x) exp(0.5 - abs(x))
xs <- seq(-6, 6, length.out = 500)

par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(xs, g_demo(xs), type = "n", ylim = c(0, 1.75), xlab = "x", ylab = "")
polygon(c(xs, rev(xs)), c(g_demo(xs), rep(0, 500)), col = adjustcolor("steelblue", 0.3), border = NA)
polygon(c(xs, rev(xs)), c(f_demo(xs), rep(0, 500)), col = adjustcolor("firebrick", 0.4), border = NA)
lines(xs, g_demo(xs), col = "steelblue", lwd = 3); lines(xs, f_demo(xs), col = "firebrick", lwd = 3)
legend("topright", bty = "n", fill = adjustcolor(c("steelblue", "firebrick"), 0.35), border = NA,
       legend = c("area under envelope g", "area under target f"))

set.seed(1)
N <- 600
rlaplace <- function(n) { u <- runif(n); ifelse(u < 0.5, log(2 * u), -log(2 * (1 - u))) }
xp <- rlaplace(N); vp <- runif(N) * g_demo(xp); acc <- vp < f_demo(xp)
plot(xs, g_demo(xs), type = "l", lwd = 3, col = "steelblue", ylim = c(0, 1.75), xlab = "x", ylab = "v")
lines(xs, f_demo(xs), lwd = 3, col = "firebrick")
points(xp[!acc], vp[!acc], pch = 4, col = adjustcolor("gray40", 0.7), cex = 0.7)
points(xp[acc], vp[acc], pch = 19, col = adjustcolor("firebrick", 0.7), cex = 0.7)
legend("topright", bty = "n", pch = c(19, 4), col = c("firebrick", "gray40"),
       legend = c(sprintf("accepted (%d)", sum(acc)), sprintf("rejected (%d)", sum(!acc))))
```

## Example 1: Sampling a Normal from a Laplace Envelope

The figure above already uses a concrete, closed-form example worth stating
in full, because it is simple enough to check every claim exactly. The target
is the unnormalized standard normal density $f(x) = e^{-x^2/2}$. Take
$g_0(x) = \tfrac12 e^{-|x|}$, the standard Laplace (double-exponential)
density, which is trivial to sample by inversion. Since
$e^{-x^2/2} \le e^{1/2-|x|}$ for all $x$ (equality exactly at $|x| = 1$, where
the two curves in the figure touch), the envelope
$$
g(x) = 2e^{1/2}\,g_0(x) = e^{1/2 - |x|} \ \ge\ f(x)
$$
is valid, and the exact acceptance rate is
$$
P(\text{accept}) = \frac{\int f(x)\,dx}{\int g(x)\,dx}
 = \frac{\sqrt{2\pi}}{2e^{1/2}} \approx 0.760.
$$

```{r}
#| label: rnorm-reject
#| tbl-cap: "Theoretical vs. simulated acceptance rate and sample moments for the Laplace-envelope normal sampler."
rnorm_reject <- function(n) {
    out <- numeric(0); tries <- 0
    while (length(out) < n) {
        m <- n - length(out)
        x <- rlaplace(m); tries <- tries + m
        out <- c(out, x[runif(m) < f_demo(x) / g_demo(x)])
    }
    list(x = out, acceptance = n / tries)
}
set.seed(2)
res <- rnorm_reject(2e5)
knitr::kable(data.frame(
    quantity = c("acceptance rate", "sample mean", "sample sd"),
    theoretical = c(sqrt(2 * pi) / (2 * sqrt(exp(1))), 0, 1),
    simulated = c(res$acceptance, mean(res$x), sd(res$x))
), digits = 4)
```

Roughly 3 in 4 proposals are kept — a tight, efficient envelope, and the
benchmark against which the next example's Cauchy envelope should be judged.

### Trying it interactively: tight vs. loose envelopes {#sec-rejection-app}

The app below scales the envelope by a factor $c \ge 1$,
$g_c(x) = c\, e^{1/2-|x|}$: larger $c$ keeps the envelope valid but wastes
more draws (acceptance rate $= 0.760/c$). Dragging $c$ below $1$ shows what
an *invalid* envelope looks like — the target pokes through it, and rejection
sampling would silently under-represent that region. This app runs live only
in the HTML edition of this book; readers of other formats can follow the
link below it.

::: {.content-visible when-format="html"}

```{shinylive-r}
#| label: envelope-tightness-app
#| standalone: true
#| viewerHeight: 600

library(shiny)

f_demo <- function(x) exp(-x^2 / 2)
g_demo <- function(x, c) c * exp(0.5 - abs(x))
rlaplace <- function(n) { u <- runif(n); ifelse(u < 0.5, log(2 * u), -log(2 * (1 - u))) }

ui <- fluidPage(
    titlePanel(NULL),
    sidebarLayout(
        sidebarPanel(
            width = 4,
            sliderInput("c", "envelope scale c (g = c * e^(1/2 - |x|))",
                        min = 0.6, max = 3, value = 1.2, step = 0.05),
            helpText("c >= 1 keeps g above f everywhere: a valid, if looser, ",
                     "envelope. c < 1 breaks the envelope condition near ",
                     "|x| = 1, where the target actually exceeds it."),
            actionButton("resample", "draw new points"),
            tableOutput("tab")
        ),
        mainPanel(
            width = 8,
            plotOutput("scatter", height = "440px")
        )
    )
)

server <- function(input, output, session) {

    pts <- eventReactive(list(input$c, input$resample), {
        set.seed(input$resample + 1)
        N <- 500
        x <- rlaplace(N)
        v <- runif(N) * g_demo(x, input$c)
        list(x = x, v = v, acc = v < f_demo(x))
    })

    output$scatter <- renderPlot({
        p <- pts()
        xs <- seq(-6, 6, length.out = 500)
        par(mar = c(4, 4, 2, 1))
        plot(xs, g_demo(xs, input$c), type = "l", lwd = 3, col = "steelblue",
             ylim = c(0, max(2.2, 1.05 * input$c * exp(0.5))),
             xlab = "x", ylab = "v", main = sprintf("c = %.2f", input$c))
        lines(xs, f_demo(xs), lwd = 3, col = "firebrick")
        bad <- f_demo(xs) > g_demo(xs, input$c)
        if (any(bad)) lines(xs[bad], f_demo(xs)[bad], col = "black", lwd = 5)
        points(p$x[!p$acc], p$v[!p$acc], pch = 4, col = adjustcolor("gray40", 0.6), cex = 0.7)
        points(p$x[p$acc], p$v[p$acc], pch = 19, col = adjustcolor("firebrick", 0.7), cex = 0.7)
        legend("topright", bty = "n", lwd = c(3, 3, 5), col = c("firebrick", "steelblue", "black"),
               legend = c("target f", "envelope g", "f exceeds g here: envelope invalid"))
    })

    output$tab <- renderTable({
        p <- pts()
        data.frame(quantity = c("theoretical acceptance", "observed acceptance", "envelope valid?"),
                   value = c(sprintf("%.3f", 0.7602 / input$c), sprintf("%.3f", mean(p$acc)),
                             ifelse(input$c >= 1, "yes", "NO (c < 1)")))
    }, colnames = FALSE)
}

shinyApp(ui, server)
```

:::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit10_rejection_sampling.html#sec-rejection-app>.
:::

::: {.callout-note}
### Reproducing this figure
The chunk above uses the [shinylive](https://github.com/quarto-ext/shinylive)
extension, compiled to WebAssembly so it runs in the reader's browser with no
Shiny server (`quarto add quarto-ext/shinylive`, then `filters: [shinylive]`
in the document or project YAML). Only base R and `shiny` are available
inside the app.
:::

## Example 2: Sampling a Gamma Distribution

The Laplace envelope above was designed to match a normal target. A
different target needs a different envelope. Here is one built for
$\text{Gamma}(\alpha, 1)$, $\alpha > 1$, whose unnormalized log-density is
$(\alpha-1)\log x - x$.

```{r}
#| label: gamma-envelope-fns
### log of a function that is (for alpha > 2) always above the Gamma(alpha, 1) log-density
log_g_gamma <- function(x, alpha) {
    (alpha - 1) * (log(alpha - 1) - 1) - log(1 + (x - (alpha - 1))^2 / (2 * alpha - 1))
}
```

### Where the envelope comes from

`log_g_gamma` is not an arbitrary formula. Up to an additive constant that
depends only on $\alpha$, it is the log-density of a **rescaled, shifted
Cauchy** distribution: writing $Y = (\alpha-1) + \sqrt{2\alpha-1}\,C$ for a
standard Cauchy $C$, a change of variables gives
$\log f_Y(y) = -\log\pi - \tfrac12\log(2\alpha-1) - \log\!\big(1 + (y-(\alpha-1))^2/(2\alpha-1)\big)$,
which differs from `log_g_gamma(y, alpha)` only by a term that does not
depend on $y$. We can check this numerically rather than take it on faith:

```{r}
#| label: verify-cauchy-link
x_check <- seq(0.1, 10, by = 0.37)
alpha_check <- 3.3
log_f_Y <- dcauchy((x_check - (alpha_check - 1)) / sqrt(2 * alpha_check - 1), log = TRUE) -
    log(sqrt(2 * alpha_check - 1))
range(log_g_gamma(x_check, alpha_check) - log_f_Y)   # constant across x confirms the claim
```

So sampling the envelope is just sampling a (rescaled) Cauchy, and the
constant in `log_g_gamma` was chosen so that $g$ touches the Gamma density
exactly at the shared mode $x = \alpha - 1$. Whether $g$ stays *above* the
Gamma density everywhere else is a separate question, and the docstring's
claim "alpha must be > 2" is a claim about exactly that:

```{r}
#| label: tbl-envelope-threshold
#| tbl-cap: "Minimum of log g(x) - log f(x) over x > 0, for several alpha. Negative means the envelope is violated somewhere."
alphas_check <- c(1.1, 1.5, 1.9, 1.99, 2.0, 2.01, 2.1, 2.5, 3, 4.5)
xv <- seq(1e-6, 60, length.out = 2e4)
gaps <- vapply(alphas_check, function(a) min(log_g_gamma(xv, a) - dgamma(xv, a, log = TRUE)), numeric(1))
knitr::kable(data.frame(alpha = alphas_check, min_log_gap = gaps, envelope_valid = gaps >= 0), digits = 5)
```

The gap is exactly $0$ at $\alpha = 2$ and negative below it: the Cauchy-based
envelope is tangent to the Gamma density at $\alpha = 2$ and crosses below it
for smaller $\alpha$, so $\alpha > 2$ is precisely the validity condition, not
a conservative rule of thumb.

### Sampling function

A batch-and-refill implementation is both clearer and much faster than
drawing one candidate at a time in an R loop: generate a vector of
candidates, vectorize the accept/reject test, keep the survivors, and top up
only the shortfall.

```{r}
#| label: gamma-sampler
sample_gamma_rej <- function(n, alpha, max_tries = 2e6) {
    out <- numeric(0); tries <- 0
    while (length(out) < n) {
        if (tries > max_tries) stop("acceptance rate too low for this alpha: envelope is too loose")
        m <- n - length(out)
        x <- rcauchy(m) * sqrt(2 * alpha - 1) + (alpha - 1)
        tries <- tries + m
        keep <- log(runif(m)) < dgamma(x, shape = alpha, log = TRUE) - log_g_gamma(x, alpha)
        out <- c(out, x[keep])
    }
    structure(out[seq_len(n)], accept.rate = n / tries)
}
```

### Test cases and efficiency

```{r}
#| label: tbl-gamma-tests
#| tbl-cap: "Acceptance rate falls sharply as alpha grows: the Cauchy envelope
#|   is tuned to match curvature at alpha = 2 and gets looser as the Gamma
#|   density becomes more Gaussian-shaped further from that point."
set.seed(3)
alphas_test <- c(2.1, 2.5, 4.5)
gamma_tests <- lapply(alphas_test, function(a) sample_gamma_rej(2000, a))
knitr::kable(data.frame(
    alpha = alphas_test,
    accept_rate = vapply(gamma_tests, attr, numeric(1), "accept.rate"),
    ks_stat = mapply(function(x, a) suppressWarnings(ks.test(x, "pgamma", a))$statistic,
                      gamma_tests, alphas_test)
), digits = 4, col.names = c("alpha", "acceptance rate", "KS statistic vs true Gamma"))
```

```{r}
#| label: fig-gamma-tests
#| fig-width: 10
#| fig-height: 3.6
#| fig-cap: "Q-Q plots of rejection-sampled draws against the true Gamma(alpha,1)
#|   quantiles, for three of the test cases: the envelope produces exact draws
#|   regardless of alpha, only the cost per draw changes."
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
for (i in 1:3) {
    a <- alphas_test[i]
    qqplot(gamma_tests[[i]], rgamma(2000, a), xlab = "rejection sample", ylab = "rgamma sample",
           main = sprintf("alpha = %g (accept %.1f%%)", a, 100 * attr(gamma_tests[[i]], "accept.rate")))
    abline(0, 1, col = "firebrick", lwd = 2)
}
```

The Q-Q plots confirm exactness at every $\alpha$: rejection sampling never
trades accuracy for speed, only speed changes. Efficient random-number
generators for the Gamma distribution (as used inside `rgamma` itself) use
envelopes that stay tight across a much wider range of $\alpha$; see Marsaglia
& Tsang (2000, *ACM TOMS*) for the algorithm R actually uses.

### Trying it interactively: the Cauchy envelope {#sec-gamma-rejection-app}

The app below shows the target $f(x)=\text{Gamma}(\alpha,1)$ density against
the Cauchy-based envelope $g(x)=\exp(\texttt{log\_g\_gamma}(x,\alpha))$ as
$\alpha$ is dragged (left), and a histogram of 2000 fresh rejection-sampled
draws against the true density (right) — an exact match whenever the
envelope is valid. Drag $\alpha$ below $2$ and the target pokes through the
envelope, as in @tbl-envelope-threshold; the crossing is a modest one for
this particular envelope, so the histogram only dips slightly short of the
curve in the affected region rather than breaking outright, but the sampler
is no longer drawing from the exact target. This app runs live only in the
HTML edition of this book; readers of other formats can follow the link
below it.

::: {.content-visible when-format="html"}

```{shinylive-r}
#| label: gamma-envelope-app
#| standalone: true
#| viewerHeight: 550

library(shiny)

log_g_gamma <- function(x, alpha) {
    (alpha - 1) * (log(alpha - 1) - 1) - log(1 + (x - (alpha - 1))^2 / (2 * alpha - 1))
}
g_gamma <- function(x, alpha) exp(log_g_gamma(x, alpha))

ui <- fluidPage(
    titlePanel(NULL),
    sidebarLayout(
        sidebarPanel(
            width = 3,
            sliderInput("alpha", "shape alpha (target: Gamma(alpha, 1))",
                        min = 1.1, max = 8, value = 3, step = 0.05),
            helpText("The envelope is only valid (g >= f everywhere) for ",
                     "alpha > 2. Below alpha = 2 the target pokes through ",
                     "the envelope near the mode, and the histogram falls ",
                     "slightly short of the true density there."),
            actionButton("resample", "draw new sample")
        ),
        mainPanel(
            width = 9,
            plotOutput("panels", height = "420px")
        )
    )
)

server <- function(input, output, session) {

    draws <- eventReactive(list(input$alpha, input$resample), {
        set.seed(input$resample + 1)
        a <- input$alpha; n <- 2000
        out <- numeric(0); tries <- 0; max_tries <- 2e6
        while (length(out) < n && tries < max_tries) {
            m <- n - length(out)
            x <- rcauchy(m) * sqrt(2 * a - 1) + (a - 1)
            tries <- tries + m
            keep <- log(runif(m)) < dgamma(x, a, log = TRUE) - log_g_gamma(x, a)
            out <- c(out, x[keep])
        }
        list(x = out[seq_len(min(n, length(out)))], accept = length(out) / tries)
    })

    output$panels <- renderPlot({
        d <- draws(); a <- input$alpha
        hi <- max(15, qgamma(0.999, a) * 1.3)
        xs <- seq(1e-3, hi, length.out = 500)
        ymax <- max(dgamma(xs, a), na.rm = TRUE) * 1.6   # scale to the target's
        # peak, not the envelope's: g touches f at the mode but can be much
        # looser far from it (especially for alpha far above 2), which would
        # otherwise blow up the y-axis and hide the interesting region

        par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

        plot(xs, g_gamma(xs, a), type = "l", lwd = 3, col = "steelblue",
             ylim = c(0, ymax), xlab = "x", ylab = "density",
             main = sprintf("target vs. envelope (alpha = %.2f)", a))
        lines(xs, dgamma(xs, a), lwd = 3, col = "firebrick")
        bad <- dgamma(xs, a) > g_gamma(xs, a)
        if (any(bad)) lines(xs[bad], dgamma(xs, a)[bad], col = "black", lwd = 5)
        legend("topright", bty = "n", lwd = c(3, 3, 5), col = c("firebrick", "steelblue", "black"),
               legend = c("target f", "envelope g", "f > g: invalid"))

        hist(d$x, breaks = 30, freq = FALSE, col = "steelblue", border = "white",
             xlim = c(0, hi), xlab = "x",
             main = sprintf("%d draws (accept rate %.1f%%)", length(d$x), 100 * d$accept))
        curve(dgamma(x, a), add = TRUE, lwd = 3, col = "firebrick")
        legend("topright", bty = "n", lwd = 3, col = "firebrick", legend = "true Gamma density")
    })
}

shinyApp(ui, server)
```

:::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit10_rejection_sampling.html#sec-gamma-rejection-app>.
:::

## Adaptive Rejection Sampling for Log-Concave Targets

Designing an envelope by hand, as above, is the hard part of rejection
sampling. For a broad and important class of targets it can be automated.

### Log-concavity and the tangent hull

A density $f$ is **log-concave** if $\log f(x)$ is concave. Many posteriors
are: the normal, the gamma with shape $\ge 1$, the beta with both parameters
$\ge 1$, logistic-regression posteriors under normal priors, and any product
of log-concave factors (so most conjugate and many non-conjugate exponential-
family posteriors qualify).

A concave function lies below every one of its tangent lines. So for tangent
lines $\ell_1,\ldots,\ell_k$ to $\log f$ at points $x_1 < \cdots < x_k$, the
piecewise-linear **upper hull**
$$
\log g(x) = \min_j \ell_j(x)
$$
satisfies $\log g \ge \log f$ everywhere, hence $g = e^{\log g} \ge f$: a
valid envelope, built automatically from a handful of tangent points and
their slopes (the log-density's derivative), with no hand-crafted formula.
Because $g$ is piecewise-exponential, both its normalizing integral and its
inverse CDF are closed-form on each piece, so sampling from $g$ is cheap.

### The adaptive algorithm

**Adaptive rejection sampling** (Gilks & Wild, 1992) refines the hull as it
goes:

1. Start with a few tangent points, at least one on each side of the mode.
2. Draw $x \sim g$, $U \sim \text{Unif}(0,1)$; accept if $U < f(x)/g(x)$.
3. Whenever a point is **rejected**, add it to the tangent set and rebuild
   the hull.

Each rejection adds a tangent exactly where the envelope was loose, so the
acceptance rate rises toward $1$ as sampling proceeds — the envelope adapts
itself to the target instead of being fixed in advance. Only $\log f$ and its
derivative are required (a derivative-free variant uses secants in place of
tangents). The `ars` package implements this for any user-supplied
log-concave `f`.

### Example 3: Truncated Normal

The standard normal is log-concave ($\log f(x) = -x^2/2$ is concave), and
truncating to $[l, u]$ only shrinks the domain — it cannot break concavity —
so a truncated normal is log-concave too, and `ars` applies to it directly.
This makes it a clean test case for comparing ARS against the most obvious
alternative: sample the *untruncated* $N(0,1)$ and throw away anything
outside $[l, u]$. That naive sampler is itself a (trivial) rejection
sampler, with envelope $g$ equal to the untruncated normal, so it is exactly
correct — but its acceptance rate is just $P(l<Z<u)$, which collapses to
essentially $0$ once $[l, u]$ sits out in the tail. ARS has no such
weakness: its hull is built from the *truncated* target from the start, so
it never proposes a point outside $[l, u]$ in the first place, and its
acceptance rate stays high regardless of where the window sits.

```{r}
#| label: ars-fns
### direct rejection sampling for a truncated normal: draw from N(0,1),
### keep values inside [lb, ub]; vectorized batch-and-refill
sample_tnorm_drs <- function(n, lb = -Inf, ub = Inf) {
    out <- numeric(0)
    while (length(out) < n) {
        x <- rnorm(n - length(out))
        out <- c(out, x[x >= lb & x <= ub])
    }
    out
}

### adaptive rejection sampling via the ars package
sample_tnorm_ars <- function(n, lb, ub) {
    logf <- function(x) dnorm(x, log = TRUE)
    fprima <- function(x) -x
    ars(n, f = logf, fprima = fprima, x = c(lb, (lb + ub) / 2, ub),
        lb = TRUE, ub = TRUE, xlb = lb, xub = ub)
}
```

```{r}
#| label: tbl-ars-timing
#| tbl-cap: "Adaptive rejection sampling (ars) vs naive rejection sampling
#|   from an untruncated N(0,1): naive sampling degrades catastrophically
#|   as the truncation window moves into the tail, while ars does not,
#|   because it never wastes draws outside [lb, ub] in the first place."
set.seed(4)
n <- 1000
windows <- list(c(-5, -4), c(-50, -40), c(100, 110))
timing <- lapply(windows, function(w) {
    t_ars <- system.time(sample_tnorm_ars(n, w[1], w[2]))["elapsed"]
    t_naive <- if (w[1] > -10 && w[2] < 10) system.time(sample_tnorm_drs(n, w[1], w[2]))["elapsed"] else NA
    c(lb = w[1], ub = w[2], ars_seconds = t_ars, naive_seconds = t_naive)
})
knitr::kable(do.call(rbind, timing), digits = 4,
             col.names = c("lower bound", "upper bound", "ars (s)", "naive rejection (s)"))
```

For $[-50,-40]$ or $[100,110]$ the naive sampler's acceptance probability is
astronomically small ($P(-50<Z<-40) \approx 10^{-221}$), so it is left as
`NA` rather than run to timeout; `ars` samples these windows in a fraction of
a second because its envelope is built from the truncated density itself,
never from the untruncated normal.

```{r}
#| label: fig-ars
#| fig-width: 10
#| fig-height: 3.6
#| fig-cap: "Histograms of adaptive-rejection-sampled draws against the true
#|   truncated-normal density (red), in three increasingly extreme
#|   truncation windows: ars matches the target exactly regardless of where
#|   the window sits, only the cost per draw changes (@tbl-ars-timing)."
### log(pnorm(ub) - pnorm(lb)), numerically stable even when the naive
### difference underflows to 0 for windows far out in either tail
log_pnorm_diff <- function(lb, ub) {
    if (ub <= 0) {
        log_pu <- pnorm(ub, log.p = TRUE); log_pl <- pnorm(lb, log.p = TRUE)
        log_pu + log1p(-exp(log_pl - log_pu))
    } else if (lb >= 0) {
        log_su <- pnorm(ub, lower.tail = FALSE, log.p = TRUE)
        log_sl <- pnorm(lb, lower.tail = FALSE, log.p = TRUE)
        log_sl + log1p(-exp(log_su - log_sl))
    } else {
        log(pnorm(ub) - pnorm(lb))
    }
}
tnorm_density <- function(x, lb, ub) exp(dnorm(x, log = TRUE) - log_pnorm_diff(lb, ub))

par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
set.seed(5)
for (w in windows) {
    x <- sample_tnorm_ars(n, w[1], w[2])
    hist(x, breaks = 25, freq = FALSE, col = "steelblue", border = "white",
         main = sprintf("[%g, %g]", w[1], w[2]), xlab = "x")
    curve(tnorm_density(x, w[1], w[2]), add = TRUE, lwd = 3, col = "firebrick", n = 300)
    if (identical(w, windows[[1]])) {
        legend("topright", bty = "n", lwd = 3, col = "firebrick", cex = 0.8,
               legend = "true density")
    }
}
```

### Animation for Illustrating ARS {#sec-ars-app}

The `ars` package above is a black box: it returns samples but not the
sequence of hulls it built along the way. The app below reimplements the
same idea directly — start with a few tangent lines to $\log f$, sample from
the piecewise-exponential envelope they define, and whenever a candidate is
*rejected*, add it as a new tangent point and rebuild the hull — one
candidate at a time, so every step of the mechanism is visible. **Next**
advances a single candidate; **Play** advances automatically. The top panel
shows $\log f$ against the current hull (in log scale the hull is exactly
piecewise-*linear*, since it is built from tangent lines), marking the most
recent candidate as accepted (green dot) or rejected (black x, becoming a
new tangent point). The bottom panel is a histogram of the accepted draws so
far against the true truncated normal density. Controls: the target sample
size, and the truncation limits $[l, u]$ (changing the limits restarts the
run). This app runs live only in the HTML edition of this book; readers of
other formats can follow the link below it.

::: {.content-visible when-format="html"}

```{shinylive-r}
#| label: ars-hull-app
#| standalone: true
#| viewerHeight: 800

library(shiny)

h <- function(x) -x^2 / 2          # log unnormalized N(0,1) density
hprime <- function(x) -x

build_hull <- function(xs, lb, ub) {
    xs <- sort(xs)
    hx <- h(xs); hpx <- hprime(xs)
    k <- length(xs)
    z <- numeric(k + 1); z[1] <- lb; z[k + 1] <- ub
    if (k > 1) for (i in 1:(k - 1)) {
        z[i + 1] <- (hx[i + 1] - hx[i] - xs[i + 1] * hpx[i + 1] + xs[i] * hpx[i]) /
            (hpx[i] - hpx[i + 1])
    }
    list(x = xs, hx = hx, hp = hpx, z = z)
}

### log-area under the exponential of a single tangent-line piece, between its
### breakpoints -- computed entirely in log space so it stays finite even far
### out in the tail, where exp(hull value) itself would underflow to exactly 0
log_piece_area <- function(hull, j) {
    a <- hull$hx[j]; b <- hull$hp[j]; xj <- hull$x[j]
    lo <- hull$z[j]; hi <- hull$z[j + 1]
    val_lo <- a + b * (lo - xj); val_hi <- a + b * (hi - xj)
    if (abs(b) < 1e-10) return(val_lo + log(hi - lo))
    m <- max(val_lo, val_hi); d <- min(val_lo, val_hi)
    m + log1p(-exp(d - m)) - log(abs(b))
}

### draw one point from the piecewise-exponential envelope: pick a piece
### proportional to its area, then invert its (exponential) CDF within it --
### both steps done on the log scale for the same underflow-safety reason
sample_hull <- function(hull) {
    k <- length(hull$x)
    log_areas <- vapply(seq_len(k), function(j) log_piece_area(hull, j), numeric(1))
    log_areas[!is.finite(log_areas)] <- -Inf
    areas <- exp(log_areas - max(log_areas))    # relative areas, safely in (0, 1]
    j <- sample.int(k, 1, prob = areas)
    a <- hull$hx[j]; b <- hull$hp[j]; xj <- hull$x[j]
    lo <- hull$z[j]; hi <- hull$z[j + 1]
    U <- runif(1)
    if (abs(b) < 1e-10) {
        x <- lo + U * (hi - lo)
        logg <- a
    } else {
        val_lo <- a + b * (lo - xj); val_hi <- a + b * (hi - xj)
        m <- max(val_lo, val_hi)
        val_x <- m + log((1 - U) * exp(val_lo - m) + U * exp(val_hi - m))
        x <- xj + (val_x - a) / b
        logg <- val_x
    }
    list(x = x, logg = logg)
}

### piecewise-linear hull value (log scale) at a vector of x's
hull_val <- function(hull, xgrid) {
    j <- findInterval(xgrid, hull$z, rightmost.closed = TRUE)
    j <- pmax(1, pmin(j, length(hull$x)))
    hull$hx[j] + hull$hp[j] * (xgrid - hull$x[j])
}

init_hull <- function(lb, ub) {
    x_init <- sort(unique(pmin(pmax(c(lb + 0.25 * (ub - lb), (lb + ub) / 2,
                                       ub - 0.25 * (ub - lb)), lb + 1e-6), ub - 1e-6)))
    build_hull(x_init, lb, ub)
}

### smart plotting window: view = [mode - 2, mode + 2], clipped to [lb, ub],
### where "mode" is the peak of the density *within* the truncation window --
### 0 if the window straddles it, otherwise whichever edge is nearest 0 (the
### window is then entirely on one side of the peak, so the density is
### monotonic and highest right at that edge). A wide window far out in the
### tail (e.g. [100, 200]) would otherwise dilute the whole plot into a
### mostly-flat, uninformative sliver near that one edge.
smart_xlim <- function(lb, ub) {
    m <- if (lb >= 0) lb else if (ub <= 0) ub else 0
    c(max(lb, m - 2), min(ub, m + 2))
}

### log(pnorm(ub) - pnorm(lb)), numerically stable even when the naive
### difference underflows to 0 for windows far out in either tail
log_pnorm_diff <- function(lb, ub) {
    if (ub <= 0) {
        log_pu <- pnorm(ub, log.p = TRUE); log_pl <- pnorm(lb, log.p = TRUE)
        log_pu + log1p(-exp(log_pl - log_pu))
    } else if (lb >= 0) {
        log_su <- pnorm(ub, lower.tail = FALSE, log.p = TRUE)
        log_sl <- pnorm(lb, lower.tail = FALSE, log.p = TRUE)
        log_sl + log1p(-exp(log_su - log_sl))
    } else {
        log(pnorm(ub) - pnorm(lb))
    }
}
tnorm_density <- function(x, lb, ub) exp(dnorm(x, log = TRUE) - log_pnorm_diff(lb, ub))

ui <- fluidPage(
    titlePanel(NULL),
    sidebarLayout(
        sidebarPanel(
            width = 3,
            numericInput("n", "sample size (target)", value = 200, min = 1, step = 10),
            numericInput("l", "lower limit l", value = -2, step = 0.1),
            numericInput("u", "upper limit u", value = 2, step = 0.1),
            fluidRow(
                column(6, actionButton("play_pause", "▶ Play", width = "100%")),
                column(6, actionButton("next_step", "Next ↷", width = "100%"))
            ),
            br(),
            actionButton("restart", "↺ Restart", width = "100%"),
            helpText("Each step draws one candidate from the current hull. ",
                     "Accepted points (green) join the histogram; rejected ",
                     "points (black x) become new tangent lines, tightening ",
                     "the hull. Changing the limits restarts the run."),
            textOutput("status")
        ),
        mainPanel(
            width = 9,
            plotOutput("panels", height = "620px")
        )
    )
)

server <- function(input, output, session) {

    rv <- reactiveValues(hull = NULL, samples = numeric(0), tries = 0, target_n = 200,
                          lb = -2, ub = 2, done = FALSE, playing = FALSE,
                          last_x = NULL, last_accept = NA)

    reset_run <- function() {
        lb <- if (!is.null(input$l) && !is.na(input$l)) input$l else -2
        ub <- if (!is.null(input$u) && !is.na(input$u)) input$u else 2
        if (ub - lb < 0.05) ub <- lb + 0.05      # guard against a degenerate window
        rv$hull <- init_hull(lb, ub)
        rv$samples <- numeric(0)
        rv$tries <- 0
        rv$lb <- lb; rv$ub <- ub
        rv$target_n <- if (!is.null(input$n) && !is.na(input$n) && input$n >= 1) input$n else 200
        rv$done <- FALSE
        rv$playing <- FALSE
        rv$last_x <- NULL; rv$last_accept <- NA
        updateActionButton(session, "play_pause", label = "▶ Play")
    }
    isolate(reset_run())

    do_step <- function() {
        if (rv$done) return(invisible(NULL))
        draw <- sample_hull(rv$hull)
        x <- draw$x; logg <- draw$logg; logf <- h(x)
        rv$tries <- rv$tries + 1
        accept <- log(runif(1)) < logf - logg
        rv$last_x <- x; rv$last_accept <- accept
        if (accept) {
            rv$samples <- c(rv$samples, x)
        } else {
            rv$hull <- build_hull(sort(c(rv$hull$x, x)), rv$lb, rv$ub)
        }
        if (length(rv$samples) >= rv$target_n) {
            rv$done <- TRUE
            rv$playing <- FALSE
            updateActionButton(session, "play_pause", label = "▶ Play")
        }
    }

    observeEvent(input$l, { reset_run() }, ignoreInit = TRUE)
    observeEvent(input$u, { reset_run() }, ignoreInit = TRUE)
    observeEvent(input$restart, { reset_run() })

    observeEvent(input$n, {
        req(input$n)
        rv$target_n <- input$n
        if (length(rv$samples) < rv$target_n) rv$done <- FALSE
    }, ignoreInit = TRUE)

    observeEvent(input$play_pause, {
        if (rv$done) return()
        rv$playing <- !rv$playing
        updateActionButton(session, "play_pause",
                            label = if (rv$playing) "⏸ Pause" else "▶ Play")
    })

    observeEvent(input$next_step, {
        rv$playing <- FALSE
        updateActionButton(session, "play_pause", label = "▶ Play")
        do_step()
    })

    ### auto-advance one step every 120ms while playing
    observe({
        if (isTRUE(rv$playing) && !rv$done) {
            invalidateLater(120, session)
            isolate(do_step())
        }
    })

    output$status <- renderText({
        sprintf("%d / %d samples accepted  |  %d proposals so far  |  %d tangent points",
                length(rv$samples), rv$target_n, rv$tries, length(rv$hull$x))
    })

    output$panels <- renderPlot({
        hull <- rv$hull; lb <- rv$lb; ub <- rv$ub
        dl <- smart_xlim(lb, ub)
        xg <- seq(dl[1], dl[2], length.out = 400)
        hull_log <- hull_val(hull, xg); target_log <- h(xg)

        par(mfrow = c(2, 1), mar = c(4, 4, 3, 1))

        plot(xg, hull_log, type = "l", lwd = 2, col = "steelblue",
             ylim = range(c(hull_log, target_log)), xlab = "x",
             ylab = "log density (unnormalized)",
             main = sprintf("log target vs. hull (%d tangent points)", length(hull$x)))
        lines(xg, target_log, lwd = 2, col = "firebrick")
        points(hull$x, hull$hx, pch = 19, col = "gray20", cex = 0.9)
        abline(v = hull$z[-c(1, length(hull$z))], col = adjustcolor("gray40", 0.4), lty = 3)
        abline(v = c(lb, ub), col = "darkorange", lty = 2)
        if (!is.null(rv$last_x)) {
            points(rv$last_x, h(rv$last_x), pch = if (rv$last_accept) 19 else 4,
                   col = if (rv$last_accept) "forestgreen" else "black", cex = 1.6, lwd = 2)
        }
        legend("bottomleft", bty = "n", cex = 0.85,
               lwd = c(2, 2, 1, NA, NA), lty = c(1, 1, 2, NA, NA), pch = c(NA, NA, NA, 19, 4),
               col = c("steelblue", "firebrick", "darkorange", "forestgreen", "black"),
               legend = c("log hull", "log target", "truncation limits", "last: accepted", "last: rejected"))

        if (length(rv$samples) == 0) {
            plot(0, 0, type = "n", xlim = dl, ylim = c(0, 1), xlab = "x", ylab = "",
                 main = "no samples accepted yet")
        } else {
            ### scale the bin count so ~30 bins fall within the *visible* view,
            ### not the full [lb, ub] window -- otherwise a wide window (e.g.
            ### [100, 200] zoomed to [100, 102]) puts everything in one giant bin
            n_bins <- min(5000, max(30, ceiling(30 * (ub - lb) / (dl[2] - dl[1]))))
            breaks <- seq(lb, ub, length.out = n_bins + 1)
            hist(rv$samples, breaks = breaks, freq = FALSE, col = "steelblue", border = "white",
                 xlim = dl, xlab = "x",
                 main = sprintf("%d / %d samples%s", length(rv$samples), rv$target_n,
                                 if (rv$done) "  (done)" else ""))
            curve(tnorm_density(x, lb, ub), add = TRUE, lwd = 3, col = "firebrick", n = 300)
            legend("topright", bty = "n", lwd = 3, col = "firebrick", legend = "true truncated-normal density")
        }
    })
}

shinyApp(ui, server)
```

:::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit10_rejection_sampling.html#sec-ars-app>.
:::

## Summary

- Rejection sampling turns draws from an envelope $g \ge f$ into *exact* iid
  draws from $f$, using only the ratio $f/g$; no normalizing constants are
  needed. The geometric view — uniform points under $g$, keep those under
  $f$ — explains why this works and why efficiency is exactly the ratio of
  the two areas.
- A good envelope is tight (high acceptance) and easy to sample; finding one
  by hand, as in the Cauchy envelope for the Gamma distribution above,
  requires matching the target's tails and curvature and can be delicate.
- Adaptive rejection sampling automates envelope construction for any
  log-concave target via tangent lines to $\log f$, tightening the envelope
  on every rejection.
- Limitations: an envelope with a bounded ratio $g/f$ may not exist when $f$
  has heavier tails than any convenient $g$, and in high dimensions the
  acceptance rate of a product envelope typically decays exponentially with
  dimension, for the same curse-of-dimensionality reason quadrature grids
  fail (Unit 8). These limitations motivate **importance sampling**, which
  reweights instead of rejecting, and **Markov chain Monte Carlo**, which
  builds a chain that adapts to the target as it runs — the subjects of the
  next two chapters.
