---
title: Introduction to R Programming
---

# Vector

R's basic data structure is the vector. A single number is just a vector of length one.

```{r}
a <- 1

a

b <- 2:10

b
```

Two vectors can be joined together with `c()` ("combine"):

```{r}
c <- c(a,b)

c
```

Arithmetic operators applied to a vector act element by element:

```{r}
1/c

c^2

c^2 + 1
```

To apply an arbitrary function to every element of a vector, either the function is already *vectorized* (like `log`), or we can use `sapply` to apply it element-wise:

```{r}
log(c)

sapply(c,log)
```

When two vectors are combined with an operator, R again works element by element. If the vectors have different lengths, the shorter one is recycled:

```{r}
d <- (1:10)*10

d

c + d

c * d

d ^ c
```

As a concrete example, here is the sample variance of `c` computed directly from its definition, and then with the built-in `var` function:

```{r}
sum((c - mean(c))^2)/(length(c)-1)

var(c)
```

## Subsetting a vector

Elements of a vector are extracted with `[ ]`, using either positions or a logical condition:

```{r}
c

c[2]

c[c(2,3)]

c[c(3,2)]

c[c > 5]
```

The condition itself is a logical vector, which is often useful to inspect on its own:

```{r}
c > 5

c[c > 5 & c < 10]

c[as.logical((c > 8) + (c < 3))]

log(c)

c[log(c) < 2]
```

A subset selected this way can also be assigned new values, which modifies the original vector in place:

```{r}
c[log(c) < 2] <- 3

c
```

## Generating sequences and repeated values

`seq` generalizes the `:` operator, letting us control the step size or the number of points, and it behaves more predictably in edge cases:

```{r}
seq(0,10,by=1)

seq(0,10,length=20)

1:10
```

For example, `1:n` does *not* produce an empty sequence when `n` is 0 -- it silently counts backwards, which is a common source of bugs in loops:

```{r}
n <- 0

1:n

#seq(1,n,by=1)
#Error in seq.default(1, n, by = 1) : wrong sign in 'by' argument
```

`rep` repeats a vector, either as a whole or element by element:

```{r}
c<- 1:5

c

rep(c,5)

rep(c,each=5)
```

# Strings

Character vectors are created and combined just like numeric ones. `paste` and `paste0` glue strings together, optionally recycling shorter arguments:

```{r}
A <- c("a","b","c")

A

paste("a","b",sep="")

paste0("a","b",sep="")

paste(A,c("d","e"))

paste(A,10)

paste(A,10,sep="")

paste0(A,10)

paste0(A,1:10)
```

`sprintf` gives C-style formatted strings, which is convenient for generating file names in a loop:

```{r}
sprintf("unit%g.pdf", 1:10)

sprintf("unit%g.%s", 1:10, "html")
```

## Selecting strings with pattern matching

`grep` returns the positions of the elements matching a pattern, which can then be used to subset a character vector -- here we list files in the working directory and pick out ones with particular extensions or prefixes:

```{r}
filelist <- list.files(); filelist

filelist[grep( "*.pdf",filelist)]
filelist[grep( "*.Rmd",filelist)]
unit.files <- filelist[grep( "^unit",filelist)]; unit.files
unit.pdf.files <- unit.files[grep("*.pdf", unit.files)]; unit.pdf.files
unit.html.files <- unit.files[grep("*.html", unit.files)]; unit.html.files
```

# Special Values

R has special values for results that are not ordinary numbers. `0/0` is "not a number" (`NaN`), `log(0)` is negative infinity (not finite), and `NA` marks a missing value:

```{r}
a <- 0/0

a

is.nan(a)

b <- log(0)

b

is.finite(b)

c <- c(0:4,NA)

c

is.na(c)
```

# Matrices

A matrix is created either by filling in values directly or by reshaping a vector into rows and columns. By default values are filled by column; `byrow = TRUE` fills by row instead:

```{r}

A <- matrix(0,4,5)

A
A <- matrix(1:20,4,5)

B <- matrix(1:20,4,5,byrow = T)
```

## Subsetting and modifying a matrix

Just like vectors, matrices can be subset with `[rows, columns]`, and the resulting subset can be assigned new values:

```{r}
D <- A[c(1,4),c(2,3)]

A[c(1,4),c(2,3)] <- 1

A[c(1,4),c(2,3)] <- 101:104

A[c(1,4),c(2,3)] <- matrix (1001:1004, 2,2)

a<- A[4,]

b<- A[3:4,]

A[1,1]
```

Note that extracting a single row or column normally drops the matrix dimension and returns a plain vector; `drop = FALSE` keeps it as a matrix:

```{r}
a2<- A[4,, drop = FALSE]
```

## Combining matrices

`array` is another way to build a matrix (or higher-dimensional array) from a vector. `cbind` and `rbind` glue matrices together by columns or by rows:

```{r}
A2 <- array(1:20,dim=c(4,5))

A2

cbind(A,A2)

rbind(A,A2)
```

## Matrix arithmetic

`t()` transposes a matrix. Arithmetic operators again act element by element, and when a matrix is combined with a shorter vector, the vector is recycled down the columns:

```{r}
t(A)

A

A + 1

x <- 1:5

A*x
```

The logic here is that R first coerces the matrix `A` into a vector by joining its columns, and then repeats the shorter vector `x` as many times as needed to match that length. This means the recycling only lines up neatly with the columns when the vector's length divides the number of rows -- as in the next example, where `x` has length 3 and `A` has 4 rows:

```{r}
x <- 1:3

A*x

A^2
```

## Matrix multiplication and linear algebra

Matrix multiplication uses `%*%`, and `solve` computes either a matrix inverse or the solution of a linear system:

```{r}
A <- matrix(sample(1:20),4,5)

A

B <- matrix(sample(1:20),5,4)

B

C <- A %*% B

C

solve(C)
```

To solve $Cx = d$ for $x$, it is both faster and numerically more stable to call `solve(C, d)` directly than to explicitly compute the inverse first:

```{r}
x <- 1:4

d <- C %*% x

solve(C,d)

#altenative way (but not recommended)
solve(C) %*% d
```

The singular value decomposition $C = UDV'$ gives another route to quantities like the determinant, here obtained as the product of the singular values:

```{r}
svd.C <- svd(C)

svd.C

prod(svd.C$d)
```

# Data frame

A data frame stores columns of possibly different types (numbers, factors, strings) side by side, indexed by name. Columns can be pulled out with `[, ]`, by name, or with `$`:

```{r}
name <- c("john","peter","jennifer")

gender <- factor(c("m","m","f"))

hw1 <- c(60,60,80)

hw2 <- c(40,50,30)

grades <- data.frame(name,gender,hw1,hw2)


grades[,"gender"]

grades[,2]
```

## Subsetting a data frame

Rows and columns can be selected simultaneously, and `subset` offers a more readable way to filter rows by a condition on the columns:

```{r}
grades[1,2]

grades[,"name"]

grades$name

grades[grades$gender=="m",]
subset (grades, hw1 >60)

grades[,"hw1"]
```

`tapply` splits a column into groups defined by another column and applies a function within each group -- here, computing the mean of `hw1` for each `gender`:

```{r}
tapply(grades[,"hw1"], grades[,"gender"],mean)
```

# List

Unlike a vector, a list can hold components of different types and sizes -- here a numeric vector, a matrix, and a character vector:

```{r}

a <- 1:10

b <- matrix(1:10,2,5)

c <- c("name1","name2")

alst <- list(aa=a,b=b,c=c)

names (alst)
str(alst)
```

A component of a list is accessed by name with `$`, or by position with double brackets `[[ ]]`:

```{r}
alst$aa


alst[[2]]
```

Two lists can be joined into one with `c()`, just as with vectors:

```{r}
blst <- list(d=2:10*10)

ablst <- c(alst,blst)

ablst
```

# Reading and saving data to harddrive

R can read plain numbers with `scan`, and tabular data with `read.csv`. Objects can then be written back to disk in several formats:

```{r}
a<- scan (text = "3 4 5.3 3")
numbers <- scan (file="numbers.txt")
mtcars <- read.csv ("mtcars.csv")
```

`save`/`load` store and restore one or more R objects together in a single `.RData` file. Note that `load` silently overwrites any existing objects with the same names:

```{r}
save (mtcars, numbers, file = "mtcars.RData")
load ("mtcars.RData")
```

A data frame can also be written out as a plain csv file, and any single R object can be saved to its own `.RDS` file:

```{r}
write.csv(mtcars, file = "mtcars2.csv")

alist <- list (A = rnorm (100), B = letters[1:10])
saveRDS(alist, file = "alist.RDS")
blist <- readRDS("alist.RDS")
## note that a list is not erased by readRDS
identical(alist, blist)
```

# Function

We can package a computation into a reusable function. Here is a hand-written function that finds the maximum value of a numeric vector and the position at which it occurs:

```{r}

find.max <- function(x)
{
    n <- length(x)

    x.m <- x[1]
    ix.m <- 1

    if(n > 1)
    {
        for( i in seq(2,n,by=1) )
        {
            if(x[i] > x.m)
            {
                x.m <- x[i]
                ix.m <- i
            }
        }
    }

    #return the maximum value and the index
    list(max=x.m,index.max=ix.m)
}
```

We can now call `find.max` just like any built-in function, and compare its result with R's own built-in tools for the same task (`max`, `which.max`, `order`, `sort`):

```{r}
a <- rnorm (5); a
find.max(a)

max(a)
which.max(a)
order (a)
sort (a)
sort(a, index.return=TRUE)
```

# Graphics

Base R plotting builds a figure with a sequence of calls: `plot` starts a new plot, and functions like `points` add to the current one. Plotting parameters such as the layout of multiple panels are set with `par`:

```{r}

demofun1 <- function(x)
{
    ( 1 + 2*x^2 + 3*x^3 + exp(x) ) / x^2
}
demofun2 <- function(x)
{
    ( 1 + 2*(10-x)^2 + 3*(10-x)^3 + exp(10-x) ) / (10-x)^2
}

# plot in R windows (for quick look)
#specify plotting parameters
par(mfrow=c(1,2), mar = c(4,4,3,1))
x <- seq(0,10,by=0.1)
#make "Plot 1"
plot(x, demofun1(x), type="p", pch = 1, ylab="y", main="Plot 1")
#add another line to "Plot 1"
points(x, demofun2(x), type="l", lty = 1)
#make "plot 2"
plot(x, demofun1(x), type="b", pch = 3, lty=1, ylab="y", main="Plot 2")
#add another line to "Plot 2"
points(x, demofun2(x), type="b", pch = 4, lty = 2)
```

The same sequence of calls can be redirected into a file instead of the screen by opening a graphics device such as `pdf()` before plotting, and closing it with `dev.off()` afterwards:

```{r}
# save plot in a file (for publication)
pdf ("afig.pdf", height=4.8, width=10)
#specify plotting parameters
par(mfrow=c(1,2), mar = c(4,4,3,1))
x <- seq(0,10,by=0.1)
#make "Plot 1"
plot(x, demofun1(x), type="p", pch = 1, ylab="y", main="Plot 1")
#add another line to "Plot 1"
points(x, demofun2(x), type="l", lty = 1)
#make "plot 2"
plot(x, demofun1(x), type="b", pch = 3, lty=1, ylab="y", main="Plot 2")
#add another line to "Plot 2"
points(x, demofun2(x), type="b", pch = 4, lty = 2)

dev.off()
```

# Modern data workflows: ggplot2 and gt

Base graphics build a plot with a sequence of side-effecting calls, as above. The **ggplot2** package instead builds a plot declaratively, by mapping columns of a data frame to visual properties ("aesthetics") and adding layers with `+`. The **gt** package does the analogous thing for tables: it turns a data frame into a polished, publication-ready table instead of a plain printout.

```{r}
library(ggplot2)
library(gt)

mtcars <- read.csv("mtcars.csv")
```

## A scatter plot with ggplot2

We map `wt` (weight) to the x-axis, `mpg` (fuel efficiency) to the y-axis, and the number of cylinders `cyl` to color, then add a fitted regression line for each group with `geom_smooth`:

```{r}
mtcars$cyl <- factor(mtcars$cyl)

ggplot(mtcars, aes(x = wt, y = mpg, color = cyl)) +
    geom_point(size = 2) +
    geom_smooth(method = "lm", se = FALSE) +
    labs(
        title = "Fuel Efficiency vs. Weight",
        x = "Weight (1000 lbs)",
        y = "Miles per Gallon",
        color = "Cylinders"
    ) +
    theme_minimal()
```

## Facets and boxplots with ggplot2

Layers combine freely: here a boxplot of `mpg` by number of gears is split into one panel per transmission type (`am`) using `facet_wrap`:

```{r}
mtcars$am <- factor(mtcars$am, labels = c("Automatic", "Manual"))

ggplot(mtcars, aes(x = factor(gear), y = mpg, fill = factor(gear))) +
    geom_boxplot(show.legend = FALSE) +
    facet_wrap(~ am) +
    labs(
        title = "MPG by Number of Gears, split by Transmission",
        x = "Number of Gears",
        y = "Miles per Gallon"
    ) +
    theme_minimal()
```

## A summary table with gt

We first use base R (`aggregate`) to compute, for each cylinder count, the average and standard deviation of `mpg` and `hp`. Piping the resulting data frame into `gt()` turns it into a formatted table with a title, readable column labels, and rounded numbers:

```{r}
summary.tbl <- aggregate(
    cbind(mpg, hp) ~ cyl,
    data = mtcars,
    FUN = function(x) c(mean = mean(x), sd = sd(x))
)
summary.tbl <- do.call(data.frame, summary.tbl)
names(summary.tbl) <- c("cyl", "mpg_mean", "mpg_sd", "hp_mean", "hp_sd")

summary.tbl |>
    gt(rowname_col = "cyl") |>
    tab_header(
        title = "Engine Performance by Cylinder Count",
        subtitle = "Mean and standard deviation of mpg and horsepower"
    ) |>
    fmt_number(columns = -cyl, decimals = 1) |>
    cols_label(
        mpg_mean = "Mean MPG",
        mpg_sd = "SD MPG",
        hp_mean = "Mean HP",
        hp_sd = "SD HP"
    ) |>
    tab_spanner(label = "Miles per Gallon", columns = c(mpg_mean, mpg_sd)) |>
    tab_spanner(label = "Horsepower", columns = c(hp_mean, hp_sd)) |>
    tab_stubhead(label = "Cylinders")
```
