---
title: "Importance Sampling"
format:
  html:
    toc: true
    tbl-colwidths: true
filters:
  - shinylive
editor_options:
  chunk_output_type: console
code-fold: true
---

```{r}
#| label: setup
#| message: false
library(ars)
```

Rejection sampling turns draws from an envelope $g \ge f$ into exact draws from
$f$, but every rejected proposal is wasted, and in the tails of a poor
envelope almost every proposal is rejected. **Importance sampling** keeps
every proposal instead of discarding most of them: draw from an easier
function $g$ and attach a *weight* to each draw that corrects for sampling
from the wrong distribution. The price is that the resulting sample is no
longer an exact iid sample from $f$ — it is a *weighted* sample — but nothing
is ever thrown away, and no envelope condition $g \ge f$ is required.

## The Reweighting Idea

### The goal

As in the last two chapters, $f(\theta)$ is an **unnormalized** function of
interest — typically $f(\theta) = L(\theta)\pi(\theta)$ for a likelihood $L$
and prior $\pi$ — and we want two kinds of integral:
$$
A = E_f\big[a(\theta)\big] = \frac{\int a(\theta)\,f(\theta)\,d\theta}{\int f(\theta)\,d\theta},
\qquad\qquad
C_f = \int f(\theta)\,d\theta,
$$
a posterior expectation and a normalizing constant (the marginal likelihood
when $f = L\pi$). Plain Monte Carlo needs draws from $f/C_f$, which we
generally cannot produce.

Suppose instead we can draw from another function $g(\theta)$ (also possibly
unnormalized, with $C_g = \int g(\theta)\,d\theta$). Wherever $f$ puts mass,
$g$ must too:
$$
\{\theta : g(\theta) > 0\} \ \supseteq\ \{\theta : f(\theta) > 0\}.
$$
Unlike rejection sampling, $g$ need **not** dominate $f$ everywhere — there is
no requirement $g \ge f$. This is what makes importance sampling applicable
in situations where no usable envelope exists.

### Two estimators

Define the **importance weight** $w(\theta) = f(\theta)/g(\theta)$. For any
$a$,
$$
E_f\big[a(\theta)\big]
= \int a(\theta)\,\frac{f(\theta)}{C_f}\,d\theta
= \frac{C_g}{C_f}\, E_g\big[a(\theta)\,w(\theta)\big],
$$
and taking $a \equiv 1$ gives $C_f/C_g = E_g[w(\theta)]$. Combining the two,
$$
E_f\big[a(\theta)\big] = \frac{E_g\big[a(\theta)\,w(\theta)\big]}{E_g\big[w(\theta)\big]}.
$$
Draw $\theta_1,\ldots,\theta_S \overset{iid}{\sim} g/C_g$ and compute
$w_s = w(\theta_s)$. Two estimators follow directly:

- **Ratio of normalizing constants**: $\widehat{C_f/C_g} = \frac1S\sum_s w_s$,
  unbiased for $C_f/C_g$; if $g$ is already a normalized density ($C_g=1$)
  this estimates $C_f$ itself.
- **Self-normalized expectation**: $\hat A = \sum_{s=1}^S a(\theta_s)\,\tilde w_s$,
  with normalized weights $\tilde w_s = w_s / \sum_j w_j$ summing to $1$. This
  turns the proposal sample into a *weighted sample from $f$*: any unknown
  constants in $f$ and $g$ cancel in the ratio, so $\hat A$ needs only $f$ and
  $g$ up to a proportionality constant. It is a ratio of two averages and so
  is biased with bias $O(1/S)$, but consistent.

### Effective sample size

$\hat A$ is a weighted average of $a(\theta_s)$: draws where $f \gg g$ count
more, draws where $f \ll g$ count less. How many *equally weighted* draws is
the weighted sample worth?
$$
S_{\text{eff}} = \frac{1}{\displaystyle\sum_{s=1}^S \tilde w_s^2}.
$$
If the weights are all equal, $\tilde w_s = 1/S$ and $S_{\text{eff}} =
1/(S \cdot 1/S^2) = S$: the proposal is shaped just like the target, and every
draw counts fully. If one weight dominates all the others, $\tilde w_s
\approx 1$ for that draw and $S_{\text{eff}} \approx 1$: only one draw is
doing any work. A variance heuristic makes this precise:
$\text{Var}_g(\hat A) \approx \text{Var}_f(a)/S_{\text{eff}}$, the same form as
for an iid sample of size $S_{\text{eff}}$ from $f$ itself.

The catch, central to this chapter, is that $S_{\text{eff}}$ is computed
**only from the weights actually observed**. A proposal that never visits the
part of $\theta$-space where $f$ and $g$ disagree most will report a
reassuring $S_{\text{eff}}$ while being badly wrong — this is exactly the
subject of the "bad stories" below.

```{r}
#| label: ess-helpers
### effective sample size from raw (possibly unnormalized) weights
ess <- function(w) sum(w)^2 / sum(w^2)
### numerically stable log(sum(exp(lx)))
log_sum_exp <- function(lx) { m <- max(lx); m + log(sum(exp(lx - m))) }
### effective sample size computed from LOG weights (avoids under/overflow)
log_ess <- function(lw) exp(2 * log_sum_exp(lw) - log_sum_exp(2 * lw))
```

## Example 1: Tail Probability of a Student-$t$ Distribution {#sec-ars-tnorm}

The target is a standard Student-$t_{\nu}$ distribution, with the degrees of
freedom $\nu$ a **tuning parameter of the target** that controls how heavy
its tail is: $\nu=1$ is the standard Cauchy, and $\nu\to\infty$ recovers the
standard normal. The goal, as before, is $P(X>c)$ for a cutoff $c$ out in the
tail, estimated from draws of a truncated-normal proposal generated by
adaptive rejection sampling (ARS).

As in the ARS section of the previous chapter, $\log\varphi$ is concave, and
restricting to $(c,\infty)$ preserves concavity, so ARS applies directly to
the truncated normal proposal. `rtnorm_ars()` below is the same idea
(standardize to $z=(x-\mu)/\sigma$, build a tangent-line upper hull and a
chord lower hull, adapt on rejection, freeze the hull once it reaches
`max_hull` points), rewritten as a vectorized batch-and-refill sampler so it
is fast enough to redraw on every slider move below.

```{r}
#| label: rtnorm-ars
## ---------------------------------------------------------------------
##  rtnorm_ars(): adaptive rejection sampling from
##      g(x)  proportional to  phi((x - mu) / sigma) * I(lo < x < hi)
##  hi may be Inf. Returns a numeric vector with attributes "hull" (final
##  abscissae, on the original scale) and "acc_rate" (accepted draws per
##  envelope draw).
## ---------------------------------------------------------------------
rtnorm_ars <- function(n, mu = 0, sigma = 1, lo = 0, hi = Inf,
                       max_hull = 25, batch = 4096) {

    stopifnot(n >= 1, sigma > 0, is.finite(lo), hi > lo, max_hull >= 3)

    a  <- (lo - mu) / sigma                # standardized lower bound
    b  <- (hi - mu) / sigma                # standardized upper bound (maybe Inf)
    h  <- function(z) -z^2 / 2             # log density, up to a constant
    hp <- function(z) -z                   # its derivative

    ## ---- upper hull: tangents at the abscissae x ---------------------
    ## Piece j is the tangent at x[j], valid on [z[j], z[j+1]], where the
    ## z[] are the tangent intersections, z[1] = a and z[k+1] = b.
    build <- function(x) {
        x <- sort(unique(x))
        x[x == 0] <- 1e-8                  # keep every slope non-zero
        x <- sort(unique(x))
        k  <- length(x)
        hx <- h(x)
        sx <- hp(x)

        j  <- seq_len(k - 1)
        ## general ARS intersection formula; for the Gaussian it reduces
        ## to the midpoint (x[j] + x[j+1]) / 2
        zz <- (hx[j + 1] - hx[j] - x[j + 1] * sx[j + 1] + x[j] * sx[j]) /
              (sx[j] - sx[j + 1])
        z  <- c(a, zz, b)

        ul <- hx + sx * (z[1:k] - x)       # hull at the left end of piece j
        ur <- hx + sx * (z[2:(k + 1)] - x) # hull at the right end (-Inf on the last)

        ## log mass of each piece, |exp(ul) - exp(ur)| / |slope|, computed
        ## with a per-piece offset so that nothing under- or overflows
        M  <- pmax(ul, ur)
        lp <- M + log(abs(exp(ul - M) - exp(ur - M))) - log(abs(sx))
        lp[!is.finite(lp)] <- -Inf

        list(x = x, hx = hx, sx = sx, z = z, ul = ul, ur = ur, M = M,
             p = exp(lp - max(lp)))
    }

    ## ---- draw m points from the piecewise-exponential envelope -------
    draw <- function(H, m) {
        j  <- sample.int(length(H$x), m, replace = TRUE, prob = H$p)
        v  <- runif(m)
        Mj <- H$M[j]
        el <- exp(H$ul[j] - Mj)
        er <- exp(H$ur[j] - Mj)
        lu <- Mj + log(el + v * (er - el))      # inverse cdf within the piece
        list(z = H$x[j] + (lu - H$hx[j]) / H$sx[j], u = lu)
    }

    ## ---- lower hull (squeeze): chords between adjacent abscissae -----
    squeeze <- function(H, zs) {
        k   <- length(H$x)
        i   <- findInterval(zs, H$x)
        out <- rep(-Inf, length(zs))
        ok  <- i >= 1 & i <= k - 1
        if (any(ok)) {
            ii <- i[ok]; zz <- zs[ok]
            out[ok] <- ((H$x[ii + 1] - zz) * H$hx[ii] +
                        (zz - H$x[ii]) * H$hx[ii + 1]) /
                       (H$x[ii + 1] - H$x[ii])
        }
        out
    }

    ## ---- initial abscissae -------------------------------------------
    ## Five points spanning (a, b) when b is finite; otherwise the same
    ## local scale 1 / max(|a|, 1) as before, reflecting the near-
    ## exponential behaviour of the Gaussian tail just above a large lower
    ## bound with no upper bound to anchor to.
    if (is.finite(b)) {
        x0 <- a + (b - a) * c(1e-3, 0.25, 0.5, 0.75, 1 - 1e-3)
    } else {
        s0 <- 1 / max(abs(a), 1)
        x0 <- c(a + 1e-3 * s0, a + 0.5 * s0, a + 2 * s0, max(a, 0) + 1.5)
    }
    H  <- build(x0)

    out <- numeric(n); got <- 0L; tries <- 0L

    ## ---- phase 1: adaptive, one draw at a time ------------------------
    while (got < n && length(H$x) < max_hull) {
        d  <- draw(H, 1); tries <- tries + 1L
        lw <- log(runif(1))
        if (lw <= squeeze(H, d$z) - d$u) {          # cheap squeeze test
            got <- got + 1L; out[got] <- d$z
        } else {                                    # full test, then adapt
            if (lw <= h(d$z) - d$u) { got <- got + 1L; out[got] <- d$z }
            H <- build(c(H$x, d$z))
        }
    }

    ## ---- phase 2: envelope frozen, vectorized rejection ---------------
    while (got < n) {
        m    <- max(batch, n - got)
        d    <- draw(H, m); tries <- tries + m
        zs   <- d$z[log(runif(m)) <= h(d$z) - d$u]
        take <- min(length(zs), n - got)
        if (take > 0L) out[got + seq_len(take)] <- zs[seq_len(take)]
        got <- got + take
    }

    structure(mu + sigma * out,
              hull     = mu + sigma * H$x,
              acc_rate = n / tries)
}
```

The contrast with the target is the point of the exercise: ARS needs
log-concavity, and $\log f_{t_{\nu}}$ is simply **not concave**, for any
finite $\nu$ — it is convex out in the tails, and only the limit
$\nu\to\infty$ (the normal) is concave everywhere. So ARS cannot sample the
target directly at any finite $\nu$; the truncated-normal proposal above is a
substitute, not a target-side match, which is exactly why the sample needs
importance weights at all.

### The estimator, and when its variance is infinite

A one-sided tail $(c,\infty)$ is a bad target for this proposal, so consider
instead a general interval $(c_{1},c_{2})$, with $c_{2}=\infty$ allowed as a
special case. Let $g$ be the normalized truncated normal on $(c_{1},c_{2})$
and $f(x)=f_{t_{\nu}}(x)I(c_{1}<x<c_{2})$. Draw $X_{1},\dots,X_{S}\sim g$,
weight each by $w(x)=f(x)/g(x)$, and average:
$$
\widehat{C} \;=\; \frac{1}{S}\sum_{i=1}^{S} w(X_{i}).
$$
This is unbiased for $C_{f}=P(c_{1}<X<c_{2})$ for *any* $g$ that is positive
wherever $f$ is — that is the whole mechanism of importance sampling, and it
does not care what $c_{1},c_{2},\mu,\sigma,\nu$ are.

Whether $\widehat C$ is *precise* is a different question, governed by
$$
\operatorname{Var}_{g}\bigl(w(X)\bigr)
 \;=\; \int_{c_{1}}^{c_{2}}\frac{f_{t_{\nu}}(x)^{2}}{g(x)}\,dx \;-\; C_{f}^{2} .
$$
When $c_{2}$ is finite, $f$ and $g$ share the same bounded interval, the
ratio $f/g$ is bounded on it, and the integral is finite — a well-behaved
estimator, whatever $\nu$ is. Send $c_{2}\to\infty$, though, and the tails
stop matching: $f_{t_{\nu}}(x)^{2}$ decays only polynomially
($\propto x^{-2(\nu+1)}$) while $g$'s Gaussian tail decays much faster, so the
integrand grows without bound and the integral diverges — for *every*
$\mu,\sigma$ and *every* finite $\nu$. The app below lets you flip between
the two regimes with a single checkbox, and, as before, $S_{\text{eff}}$ need
not warn you when you have landed in the bad one.

### An interactive view {#sec-t-is-app}

The app below puts $c_{1}$, $c_{2}$, $\mu$, $\sigma$ and the target's degrees
of freedom $\nu$ under slider control (a checkbox sends $c_{2}$ to $\infty$),
and displays, on a common $x$-axis: the two truncated densities on a log
scale with a rug of the draws, and the normalized importance weights
$\tilde w_{i}=w_{i}/\sum_{j}w_{j}$ at the sampled points. The legend reports
the true $P(c_{1}<X<c_{2})$, the estimate, and its nominal 95% interval,
colored green when the interval covers the truth and red when it does not.
Leave $c_{2}$ finite and the estimator behaves; check the box and watch it
misbehave exactly as the variance calculation predicts, while
$S_{\text{eff}}$ in the bottom panel need not warn you either way. This app
runs live only in the HTML edition of this book; readers of other formats can
follow the link below it.

::: {#fig-t-is-app .content-visible when-format="html"}

```{shinylive-r}
#| label: is-t-app
#| standalone: true
#| viewerHeight: 800

library(shiny)

## ---------------------------------------------------------------------
##  rtnorm_ars(): adaptive rejection sampling from
##      g(x)  proportional to  phi((x - mu) / sigma) * I(lo < x < hi)
##  hi may be Inf. Returns a numeric vector with attributes "hull" (final
##  abscissae, on the original scale) and "acc_rate" (accepted draws per
##  envelope draw).
## ---------------------------------------------------------------------
rtnorm_ars <- function(n, mu = 0, sigma = 1, lo = 0, hi = Inf,
                       max_hull = 25, batch = 4096) {

    stopifnot(n >= 1, sigma > 0, is.finite(lo), hi > lo, max_hull >= 3)

    a  <- (lo - mu) / sigma                # standardized lower bound
    b  <- (hi - mu) / sigma                # standardized upper bound (maybe Inf)
    h  <- function(z) -z^2 / 2             # log density, up to a constant
    hp <- function(z) -z                   # its derivative

    ## ---- upper hull: tangents at the abscissae x ---------------------
    ## Piece j is the tangent at x[j], valid on [z[j], z[j+1]], where the
    ## z[] are the tangent intersections, z[1] = a and z[k+1] = b.
    build <- function(x) {
        x <- sort(unique(x))
        x[x == 0] <- 1e-8                  # keep every slope non-zero
        x <- sort(unique(x))
        k  <- length(x)
        hx <- h(x)
        sx <- hp(x)

        j  <- seq_len(k - 1)
        ## general ARS intersection formula; for the Gaussian it reduces
        ## to the midpoint (x[j] + x[j+1]) / 2
        zz <- (hx[j + 1] - hx[j] - x[j + 1] * sx[j + 1] + x[j] * sx[j]) /
              (sx[j] - sx[j + 1])
        z  <- c(a, zz, b)

        ul <- hx + sx * (z[1:k] - x)       # hull at the left end of piece j
        ur <- hx + sx * (z[2:(k + 1)] - x) # hull at the right end (-Inf on the last)

        ## log mass of each piece, |exp(ul) - exp(ur)| / |slope|, computed
        ## with a per-piece offset so that nothing under- or overflows
        M  <- pmax(ul, ur)
        lp <- M + log(abs(exp(ul - M) - exp(ur - M))) - log(abs(sx))
        lp[!is.finite(lp)] <- -Inf

        list(x = x, hx = hx, sx = sx, z = z, ul = ul, ur = ur, M = M,
             p = exp(lp - max(lp)))
    }

    ## ---- draw m points from the piecewise-exponential envelope -------
    draw <- function(H, m) {
        j  <- sample.int(length(H$x), m, replace = TRUE, prob = H$p)
        v  <- runif(m)
        Mj <- H$M[j]
        el <- exp(H$ul[j] - Mj)
        er <- exp(H$ur[j] - Mj)
        lu <- Mj + log(el + v * (er - el))      # inverse cdf within the piece
        list(z = H$x[j] + (lu - H$hx[j]) / H$sx[j], u = lu)
    }

    ## ---- lower hull (squeeze): chords between adjacent abscissae -----
    squeeze <- function(H, zs) {
        k   <- length(H$x)
        i   <- findInterval(zs, H$x)
        out <- rep(-Inf, length(zs))
        ok  <- i >= 1 & i <= k - 1
        if (any(ok)) {
            ii <- i[ok]; zz <- zs[ok]
            out[ok] <- ((H$x[ii + 1] - zz) * H$hx[ii] +
                        (zz - H$x[ii]) * H$hx[ii + 1]) /
                       (H$x[ii + 1] - H$x[ii])
        }
        out
    }

    ## ---- initial abscissae -------------------------------------------
    ## Five points spanning (a, b) when b is finite; otherwise the same
    ## local scale 1 / max(|a|, 1) as before, reflecting the near-
    ## exponential behaviour of the Gaussian tail just above a large lower
    ## bound with no upper bound to anchor to.
    if (is.finite(b)) {
        x0 <- a + (b - a) * c(1e-3, 0.25, 0.5, 0.75, 1 - 1e-3)
    } else {
        s0 <- 1 / max(abs(a), 1)
        x0 <- c(a + 1e-3 * s0, a + 0.5 * s0, a + 2 * s0, max(a, 0) + 1.5)
    }
    H  <- build(x0)

    out <- numeric(n); got <- 0L; tries <- 0L

    ## ---- phase 1: adaptive, one draw at a time ------------------------
    while (got < n && length(H$x) < max_hull) {
        d  <- draw(H, 1); tries <- tries + 1L
        lw <- log(runif(1))
        if (lw <= squeeze(H, d$z) - d$u) {          # cheap squeeze test
            got <- got + 1L; out[got] <- d$z
        } else {                                    # full test, then adapt
            if (lw <= h(d$z) - d$u) { got <- got + 1L; out[got] <- d$z }
            H <- build(c(H$x, d$z))
        }
    }

    ## ---- phase 2: envelope frozen, vectorized rejection ---------------
    while (got < n) {
        m    <- max(batch, n - got)
        d    <- draw(H, m); tries <- tries + m
        zs   <- d$z[log(runif(m)) <= h(d$z) - d$u]
        take <- min(length(zs), n - got)
        if (take > 0L) out[got + seq_len(take)] <- zs[seq_len(take)]
        got <- got + take
    }

    structure(mu + sigma * out,
              hull     = mu + sigma * H$x,
              acc_rate = n / tries)
}

ui <- fluidPage(
    titlePanel("Importance sampling for a Student-t interval with a truncated-normal proposal"),
    sidebarLayout(
        sidebarPanel(
            width = 3,
            sliderInput("df", HTML("target df &nu;"), min = 1, max = 30, value = 3, step = 1),
            sliderInput("c1", HTML("c<sub>1</sub>"), min = -5, max = 10, value = 1, step = 0.25),
            sliderInput("c2", HTML("c<sub>2</sub>"), min = -5, max = 15, value = 3, step = 0.25),
            checkboxInput("c2inf", HTML("c<sub>2</sub> = &infin;  (infinite-variance case)"), FALSE),
            sliderInput("mu", HTML("proposal &mu;"), min = -5, max = 20, value = 0, step = 0.25),
            sliderInput("sg", HTML("proposal &sigma;"), min = 0.25, max = 10, value = 1, step = 0.25),
            selectInput("S", "sample size  S",
                        choices = c(200, 1000, 5000, 20000), selected = 1000),
            actionButton("new", "New sample"),
            checkboxInput("logw", "log scale for weights", FALSE)
        ),
        mainPanel(width = 9, plotOutput("panels", height = "680px"))
    )
)

server <- function(input, output) {

    sim <- reactive({
        S  <- as.integer(input$S)
        c1 <- input$c1
        c2 <- if (input$c2inf) Inf else input$c2
        validate(need(c2 > c1, "c2 must be greater than c1"))
        mu <- input$mu; sg <- input$sg; df <- input$df
        set.seed(input$new + 1)

        X  <- rtnorm_ars(S, mu, sg, c1, c2)

        ## log importance weight  log f_t(x; df) - log g(x),  with g the
        ## normalized truncated normal on (c1, c2)
        lZ   <- log(pnorm(c2, mu, sg) - pnorm(c1, mu, sg))
        logw <- dt(X, df, log = TRUE) - (dnorm(X, mu, sg, log = TRUE) - lZ)

        m    <- max(logw)
        v    <- exp(logw - m)                       # rescaled weights
        Chat <- mean(v) * exp(m)                    # unbiased estimate of P(c1 < X < c2)
        se   <- sd(v) / sqrt(S) * exp(m)            # nominal standard error
        wn   <- v / sum(v)                          # normalized weights

        list(X = X, wn = wn, S = S, c1 = c1, c2 = c2, mu = mu, sg = sg, df = df,
             Chat = Chat, se = se, ess = 1 / sum(wn^2),
             truth = pt(c2, df) - pt(c1, df))
    })

    output$panels <- renderPlot({
        s  <- sim()
        c1 <- s$c1; c2 <- s$c2; X <- s$X
        lo <- c1
        hi <- if (is.finite(c2)) c2 else max(max(X) + 0.5, c1 + 5)
        xs <- seq(lo, hi, length.out = 800)

        ## both densities normalized on (c1, c2)
        ft <- dt(xs, s$df) / s$truth
        gt <- dnorm(xs, s$mu, s$sg) / (pnorm(c2, s$mu, s$sg) - pnorm(c1, s$mu, s$sg))
        yl <- range(c(ft, gt)[c(ft, gt) > 0])
        yl[1] <- max(yl[1], 1e-12)

        lwr <- s$Chat - 1.96 * s$se
        upr <- s$Chat + 1.96 * s$se
        covered <- lwr <= s$truth && s$truth <= upr
        ci_col  <- if (covered) "darkgreen" else "firebrick"

        layout(matrix(1:2, 2), heights = c(3, 2))

        ## ---- top panel: the two densities ----------------------------
        par(mar = c(0.4, 5, 3, 1))
        plot(xs, ft, type = "n", log = "y", ylim = yl, xlim = c(lo, hi),
             xaxt = "n", xlab = "", ylab = "density (log scale)",
             main = "target and proposal, both normalized on (c1, c2)")
        if (!is.finite(c2)) {
            rect(max(X), yl[1], hi, yl[2] * 10,
                 col = adjustcolor("grey60", 0.18), border = NA)
            text(max(X), yl[2], "  no draw ever reached here", adj = c(0, 1.4),
                 col = "grey35", cex = 0.95)
        }
        lines(xs, ft, lwd = 4, col = "firebrick")
        lines(xs, gt, lwd = 3, col = "steelblue")
        rug(X, col = adjustcolor("steelblue", 0.25))
        legend("bottomleft", inset = c(0.02, 0.05), bty = "n", cex = 1.05,
               lwd = c(4, 3, NA, NA, NA),
               col = c("firebrick", "steelblue", NA, NA, NA),
               text.col = c("black", "black", "black", "black", ci_col),
               legend = c(
                   sprintf("target: t(df=%g) on (c1, c2)", s$df),
                   sprintf("proposal: N(%.2f, %.2f^2) on (c1, c2)", s$mu, s$sg),
                   sprintf("true  P = %.5g", s$truth),
                   sprintf("est.  P = %.5g", s$Chat),
                   sprintf("95%% CI = (%.5g, %.5g)", lwr, upr)))

        ## ---- bottom panel: normalized weights ------------------------
        par(mar = c(4.5, 5, 0.4, 1))
        wn <- s$wn
        if (input$logw) {
            wl <- c(max(min(wn[wn > 0]), 1e-12), max(wn))
            plot(X, wn, type = "n", log = "y", ylim = wl, xlim = c(lo, hi),
                 xlab = "x", ylab = "normalized weight (log scale)")
            segments(X, wl[1], X, wn, col = adjustcolor("grey25", 0.5))
        } else {
            plot(X, wn, type = "n", ylim = c(0, max(wn) * 1.08),
                 xlim = c(lo, hi), xlab = "x", ylab = "normalized weight")
            segments(X, 0, X, wn, col = adjustcolor("grey25", 0.5))
        }
        abline(h = 1 / s$S, lty = 2, col = "darkgreen", lwd = 2)
        legend("bottomright", inset = c(0.02, 0.05), bty = "n", cex = 1.05,
               lty = c(2, NA, NA), lwd = c(2, NA, NA),
               col = c("darkgreen", NA, NA),
               legend = c("1 / S  (perfectly balanced weights)",
                          sprintf("ESS = %.1f   (%.2f%% of S = %d)",
                                  s$ess, 100 * s$ess / s$S, s$S),
                          sprintf("largest weight = %.1f%% of the total",
                                  100 * max(wn))))
    })
}

shinyApp(ui, server)
```

**Shinylive App for Importance Sampling of a Student-$t$ Interval.** Sliders
control $c_{1}$, $c_{2}$ (or $c_{2}=\infty$ via the checkbox), the target's
degrees of freedom $\nu$, and the proposal's $\mu,\sigma$; the top panel
compares the (normalized) target and proposal densities on a log scale, with
the CI in the legend colored by whether it covers the truth, and the bottom
panel shows the normalized importance weights at the sampled points.

:::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit12_importance_sampling.html#sec-t-is-app>.
:::

::: {.callout-note}
### Reproducing this figure
The chunk above uses the [shinylive](https://github.com/quarto-ext/shinylive)
extension, which compiles the app to WebAssembly so that it runs in the
reader's browser with no Shiny server (`quarto add quarto-ext/shinylive`, then
`filters: [shinylive]` in the document or project YAML, already set for this
book). A `{shinylive-r}` chunk cannot see objects defined elsewhere in the
document, so the body of `rtnorm_ars()` is pasted inside the app chunk rather
than reused from the `rtnorm-ars` chunk above; only base R and `shiny` are
available inside the app. To develop the app interactively instead, an
ordinary `{r}` chunk with `shinyApp(ui, server)` at the end can be run directly
in RStudio or Positron, which opens it in the viewer pane.
:::

Several settings repay attention.

- **$c_{2}$ finite (the default).** Target and proposal share the same
  bounded support, so the weights have finite variance: raising $S$ improves
  the estimate at the usual $1/\sqrt S$ rate, and the 95% interval covers
  about as often as it claims to.
- **Check "$c_{2}=\infty$".** This reproduces the divergent case derived
  above. Raising $S$ no longer shrinks a variance that does not exist, and
  repeated *New sample* draws swing far more than the reported standard error
  admits. With large $\nu$ the run-away region is rarely reached, so the app
  can still look fine by chance; with small $\nu$ (Cauchy, $\nu=1$) it bites
  often — but $S_{\text{eff}}$ need not warn you either way.
- **Small $\sigma$, or $\mu$ far outside $(c_{1},c_{2})$.** Draws pile up
  wherever the proposal happens to put mass inside the interval, the weights
  are nearly equal, $S_{\text{eff}}$ stays high, and the estimate can still
  miss badly: uniform weights certify only that the proposal is internally
  consistent, not that it matches the target's shape there.
- **Large $\sigma$.** One draw's weight can dominate the sum,
  $S_{\text{eff}}$ collapses toward $1$ — visibly unreliable, which is an
  improvement on being invisibly unreliable.


## Example 2: Marginal Likelihood via Importance Sampling

### Setup

The [quadrature](unit09_integral.qmd) and Laplace-approximation chapters
compute the marginal likelihood $P(y) = \int L(\theta)\pi(\theta)\,d\theta$ of
the same conjugate-free Gaussian model — unknown mean $\mu$ and
$w=\log\sigma^2$, independent normal priors — by a product grid and by a
mode-and-Hessian approximation, respectively. Importance sampling offers a
third route, and one that comes with a built-in accuracy diagnostic
($S_{\text{eff}}$) that neither of the other two methods provides directly.

Take $g(\mu,w)$ to be the **Gaussian approximation to the posterior**
obtained by a Laplace fit: locate the posterior mode $(\hat\mu,\hat w)$ by
optimization, take $g = N_2\big((\hat\mu,\hat w),\, H^{-1}\big)$ for $H$ the
Hessian of $-\log[L\pi]$ at the mode, and importance-sample from $g$ with
$f = L\pi$:
$$
\widehat{P(y)} = \frac{1}{S}\sum_{s=1}^S w_s, \qquad
w_s = \frac{L(\theta_s)\pi(\theta_s)}{g(\theta_s)}, \qquad \theta_s \sim g.
$$
Because $g$ is shaped like the posterior near its mode, this is a "good
story" proposal by construction — in sharp contrast to sampling directly from
the (typically much wider, uninformative) prior, which is exactly the
`log_marlik_mc` function used elsewhere in this book as a naive-Monte-Carlo
check: it is importance sampling with $g = \pi$, and its weights are just the
likelihood values $L(\theta_s)$.

### An efficient implementation

```{r}
#| label: marlik-functions
log_lik   <- function(x, mu, w) sum(dnorm(x, mu, exp(w / 2), log = TRUE))
log_prior <- function(mu, w, mu_0, sigma_mu, w_0, sigma_w) {
    dnorm(mu, mu_0, sigma_mu, log = TRUE) + dnorm(w, w_0, sigma_w, log = TRUE)
}
neg_log_post <- function(theta, x, mu_0, sigma_mu, w_0, sigma_w) {
    -log_lik(x, theta[1], theta[2]) -
        log_prior(theta[1], theta[2], mu_0, sigma_mu, w_0, sigma_w)
}

### log density of N(mu, A^{-1}) at every column of a 2 x S matrix of points at once,
### replacing a per-point apply() call with one vectorized quadratic form
log_dmvnorm_batch <- function(Theta, mu, A) {
    Tc <- Theta - mu
    quad <- colSums((A %*% Tc) * Tc)
    0.5 * (-nrow(Theta) * log(2 * pi) + sum(log(svd(A)$d)) - quad)
}

### log marginal likelihood by importance sampling from a Laplace-shaped Gaussian
log_marlik_imps <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    fit <- nlm(neg_log_post, c(mean(x), log(stats::var(x))), hessian = TRUE,
               x = x, mu_0 = mu_0, sigma_mu = sigma_mu, w_0 = w_0, sigma_w = sigma_w)
    mu_hat <- fit$estimate; H <- fit$hessian
    L <- t(chol(solve(H)))                        # Sigma = H^{-1} = L L'
    Theta <- L %*% matrix(rnorm(2 * S), 2, S) + mu_hat

    log_g <- log_dmvnorm_batch(Theta, mu_hat, H)
    log_f <- -apply(Theta, 2, neg_log_post, x = x, mu_0 = mu_0, sigma_mu = sigma_mu,
                     w_0 = w_0, sigma_w = sigma_w)
    lw <- log_f - log_g
    structure(log_sum_exp(lw) - log(S), ess = log_ess(lw))
}

### log marginal likelihood by naive Monte Carlo from the prior (g = pi): a
### second, independent importance-sampling estimator with an unfavourable g
log_marlik_prior <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    mus <- rnorm(S, mu_0, sigma_mu); ws <- rnorm(S, w_0, sigma_w)
    lw <- vapply(seq_len(S), function(i) log_lik(x, mus[i], ws[i]), numeric(1))
    structure(log_sum_exp(lw) - log(S), ess = log_ess(lw))
}
```

Replacing the two `apply()` calls used in earlier drafts of this code — one
for the proposal log-density, one for the log-posterior — with a single
vectorized quadratic form (`log_dmvnorm_batch`) removes $S$ separate R
function calls for the proposal evaluation; only the data-dependent
log-likelihood still needs one call per draw, since it sums over the $n$
observations for each $\theta_s$ separately.

### Checking accuracy

```{r}
#| label: tbl-marlik-check
#| tbl-cap: "Log marginal likelihood by importance sampling (Laplace-shaped proposal, S = 200) across 500 replications, for two simulated data sets."
set.seed(4)
x_small <- rnorm(10);  x_large <- rnorm(200)

check_tbl <- do.call(rbind, lapply(list(n10 = x_small, n200 = x_large), function(x) {
    reps <- replicate(500, log_marlik_imps(x, 0, 5, 0, 5, 200), simplify = FALSE)
    vals <- vapply(reps, as.numeric, numeric(1))
    esss <- vapply(reps, attr, numeric(1), which = "ess")
    data.frame(mean_log_ml = mean(vals), sd_log_ml = sd(vals), mean_ess = mean(esss))
}))
check_tbl <- cbind(dataset = c("n = 10", "n = 200"), check_tbl)
knitr::kable(check_tbl, digits = 4)
```

The Monte Carlo standard deviation across replications is small relative to
the estimate itself, and the effective sample size stays close to the nominal
$S=200$ in both cases — evidence that the Laplace-shaped proposal tracks the
posterior well even though it is, after all, only a second-order
approximation to it.

### Why the proposal matters: Laplace-shaped versus prior sampling

The prior is a valid proposal for the same target — every posterior draw is
also a prior draw with positive density — so `log_marlik_prior` is a
legitimate importance-sampling estimator too. But the prior is far more
diffuse than the posterior, especially once $n$ is not tiny, so it is a much
poorer match to $f = L\pi$.

```{r}
#| label: tbl-marlik-ess-vs-n
#| tbl-cap: "Effective sample size (of S = 500 draws) for the Laplace-shaped proposal versus prior sampling, as the sample size n of the simulated data grows. Both target the same posterior for (mu, w)."
set.seed(5)
ns <- c(5, 20, 100, 400)
ess_tbl <- do.call(rbind, lapply(ns, function(n) {
    x <- rnorm(n)
    data.frame(n = n,
               ess_laplace_proposal = attr(log_marlik_imps(x, 0, 5, 0, 5, 500), "ess"),
               ess_prior_sampling   = attr(log_marlik_prior(x, 0, 5, 0, 5, 500), "ess"))
}))
knitr::kable(ess_tbl, digits = c(0, 1, 1))
```

As $n$ grows the posterior sharpens around the truth while the prior does
not, so prior sampling's effective sample size deteriorates steadily; the
Laplace-shaped proposal, which re-centers and re-scales itself around the
posterior mode for every data set, keeps pace. This is the marginal-likelihood
instance of the same lesson as the missed-mode example above: a fixed,
data-independent proposal becomes an increasingly bad match to a target that
concentrates as more data arrive, and only a proposal that adapts to the
target avoids the resulting collapse in efficiency.

## Summary

- Importance sampling reweights draws from a proposal $g$ by
  $w(\theta) = f(\theta)/g(\theta)$ instead of accepting or rejecting them,
  using $\widehat{C_f/C_g} = \frac1S\sum_s w_s$ and the self-normalized
  $\hat A = \sum_s a(\theta_s)\tilde w_s$; only the support condition
  $\{g>0\}\supseteq\{f>0\}$ is required, not an envelope $g \ge f$.
- A good proposal shaped like the target on its support gives near-constant
  weights and $S_{\text{eff}} \approx S$ (the truncated-normal and
  Laplace-proposal examples), and a single weighted sample can answer several
  questions about the same target — a probability and an expectation alike.
- A proposal that misses a mode or part of the support fails **silently**: the
  observed weights can look perfectly healthy while both $C_f/C_g$ and
  $E_f[a]$ are badly wrong, and no amount of extra sampling fixes it.
  $S_{\text{eff}}$ diagnoses only the noisy failure (wildly varying observed
  weights), never the silent one.
- For the marginal likelihood, a proposal built from a Laplace approximation
  at the posterior mode remains efficient as the data accumulate and the
  posterior sharpens, while a fixed, data-independent proposal such as the
  prior does not — reinforcing why the mode-finding methods from earlier
  chapters remain useful even once we turn to sampling-based methods.
