---
title: "Rejection Sampling"
format:
  html:
    toc: true
    tbl-colwidths: true
filters:
  - shinylive
editor_options:
  chunk_output_type: console
code-fold: true
---

```{r}
#| label: setup
#| message: false
library(ars)
```

Monte Carlo integration estimates $E_f[a(\theta)] = \int a(\theta) f(\theta)\,d\theta$ from
iid draws $\theta_1,\ldots,\theta_n \sim f$. The catch is the sampling step
itself: R has built-in samplers for standard families, but a posterior
$f(\theta) \propto L(\theta)\pi(\theta)$ is typically known only up to a
constant, and there is no generic way to draw from it directly. **Rejection
sampling** solves this by producing *exact* iid draws from $f$ using draws
from an easier distribution $g$, following the recipe *reject, reweight,
repeat*.

## The Envelope and the Algorithm

### The envelope condition

Suppose we can find a function $g(x)$ such that

- $g(x) \ge f(x)$ for every $x$ — an **envelope** of $f$, and
- we can sample from the normalized density $g(x)/\!\int g(x)\,dx$.

Neither $f$ nor $g$ needs to be normalized: $f$ may be an unnormalized
posterior and $g$ a scaled density $M\,g_0(x)$ for a sampler $g_0$ we already
have. The overall acceptance rate of the algorithm below turns out to be the
ratio of the two areas,
$$
P(\text{accept}) = \frac{\int f(x)\,dx}{\int g(x)\,dx},
$$
so a tight envelope wastes few draws and a loose one wastes many — the entire
design problem is finding a $g$ that is both easy to sample and close to $f$.

### The algorithm

**Input**: target $f$, envelope $g \ge f$ with a sampler for
$g(x)/\!\int g(x)\,dx$. For each draw needed, repeat until acceptance:

1. Draw $x \sim g(x)/\!\int g(x)\,dx$.
2. Compute $r = f(x)/g(x)$.
3. Draw $U \sim \text{Unif}(0,1)$.
4. If $U < r$, accept $x$; otherwise return to step 1.

Only the ratio $f/g$ is ever evaluated, so unknown normalizing constants in
$f$ (and in $g$) cancel automatically as long as the *versions actually used*
satisfy $g \ge f$. The number of proposals needed per accepted draw is
geometric with mean $1/P(\text{accept})$.

### An equivalent geometric picture

An illuminating restatement: draw $X \sim g$, then draw a height
$V \sim \text{Unif}(0, g(X))$ (equivalently $V = U\,g(X)$), and accept $X$ if
$V < f(X)$. This is the same algorithm, since
$V < f(X) \iff U < f(X)/g(X)$, but it exposes *why* rejection sampling works:

- Steps "draw $X\sim g$, then $V\sim\text{Unif}(0,g(X))$" place $(X,V)$
  **uniformly** in the region under $g$, $G = \{(x,v): 0<v<g(x)\}$.
- Keeping only points with $v < f(x)$ restricts to the sub-region under $f$,
  $F = \{(x,v): 0<v<f(x)\} \subset G$.
- A uniform point on $G$, conditioned on landing in $F$, is uniform on $F$.
  The $x$-coordinate of a uniform point on $F$ has density proportional to
  the *height* of $F$ at $x$, which is $f(x)$. That is exactly a draw from
  $f/\!\int f$.

```{r}
#| label: fig-envelope-scatter
#| fig-width: 10
#| fig-height: 4.4
#| fig-cap: "Left: target $f(x)=e^{-x^2/2}$ (unnormalized standard normal) under
#|   its Laplace envelope $g(x) = e^{1/2-|x|}$, with the two areas shaded.
#|   Right: 600 candidate points drawn uniformly under $g$; the accepted ones
#|   (red, under $f$ too) are uniformly distributed under $f$, so their
#|   $x$-coordinates are draws from the normalized $f$."
f_demo <- function(x) exp(-x^2 / 2)
g_demo <- function(x) exp(0.5 - abs(x))
xs <- seq(-6, 6, length.out = 500)

par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(xs, g_demo(xs), type = "n", ylim = c(0, 1.75), xlab = "x", ylab = "")
polygon(c(xs, rev(xs)), c(g_demo(xs), rep(0, 500)), col = adjustcolor("steelblue", 0.3), border = NA)
polygon(c(xs, rev(xs)), c(f_demo(xs), rep(0, 500)), col = adjustcolor("firebrick", 0.4), border = NA)
lines(xs, g_demo(xs), col = "steelblue", lwd = 3); lines(xs, f_demo(xs), col = "firebrick", lwd = 3)
legend("topright", bty = "n", fill = adjustcolor(c("steelblue", "firebrick"), 0.35), border = NA,
       legend = c("area under envelope g", "area under target f"))

set.seed(1)
N <- 600
rlaplace <- function(n) { u <- runif(n); ifelse(u < 0.5, log(2 * u), -log(2 * (1 - u))) }
xp <- rlaplace(N); vp <- runif(N) * g_demo(xp); acc <- vp < f_demo(xp)
plot(xs, g_demo(xs), type = "l", lwd = 3, col = "steelblue", ylim = c(0, 1.75), xlab = "x", ylab = "v")
lines(xs, f_demo(xs), lwd = 3, col = "firebrick")
points(xp[!acc], vp[!acc], pch = 4, col = adjustcolor("gray40", 0.7), cex = 0.7)
points(xp[acc], vp[acc], pch = 19, col = adjustcolor("firebrick", 0.7), cex = 0.7)
legend("topright", bty = "n", pch = c(19, 4), col = c("firebrick", "gray40"),
       legend = c(sprintf("accepted (%d)", sum(acc)), sprintf("rejected (%d)", sum(!acc))))
```

## Example: Sampling a Normal from a Laplace Envelope

The figure above already uses a concrete, closed-form example worth stating
in full, because it is simple enough to check every claim exactly. The target
is the unnormalized standard normal density $f(x) = e^{-x^2/2}$. Take
$g_0(x) = \tfrac12 e^{-|x|}$, the standard Laplace (double-exponential)
density, which is trivial to sample by inversion. Since
$e^{-x^2/2} \le e^{1/2-|x|}$ for all $x$ (equality exactly at $|x| = 1$, where
the two curves in the figure touch), the envelope
$$
g(x) = 2e^{1/2}\,g_0(x) = e^{1/2 - |x|} \ \ge\ f(x)
$$
is valid, and the exact acceptance rate is
$$
P(\text{accept}) = \frac{\int f(x)\,dx}{\int g(x)\,dx}
 = \frac{\sqrt{2\pi}}{2e^{1/2}} \approx 0.760.
$$

```{r}
#| label: rnorm-reject
#| tbl-cap: "Theoretical vs. simulated acceptance rate and sample moments for the Laplace-envelope normal sampler."
rnorm_reject <- function(n) {
    out <- numeric(0); tries <- 0
    while (length(out) < n) {
        m <- n - length(out)
        x <- rlaplace(m); tries <- tries + m
        out <- c(out, x[runif(m) < f_demo(x) / g_demo(x)])
    }
    list(x = out, acceptance = n / tries)
}
set.seed(2)
res <- rnorm_reject(2e5)
knitr::kable(data.frame(
    quantity = c("acceptance rate", "sample mean", "sample sd"),
    theoretical = c(sqrt(2 * pi) / (2 * sqrt(exp(1))), 0, 1),
    simulated = c(res$acceptance, mean(res$x), sd(res$x))
), digits = 4)
```

Roughly 3 in 4 proposals are kept — a tight, efficient envelope, and the
benchmark against which the next example's Cauchy envelope should be judged.

### Trying it interactively: tight vs. loose envelopes {#sec-rejection-app}

The app below scales the envelope by a factor $c \ge 1$,
$g_c(x) = c\, e^{1/2-|x|}$: larger $c$ keeps the envelope valid but wastes
more draws (acceptance rate $= 0.760/c$). Dragging $c$ below $1$ shows what
an *invalid* envelope looks like — the target pokes through it, and rejection
sampling would silently under-represent that region. This app runs live only
in the HTML edition of this book; readers of other formats can follow the
link below it.

::: {.content-visible when-format="html"}

```{shinylive-r}
#| label: envelope-tightness-app
#| standalone: true

library(shiny)

f_demo <- function(x) exp(-x^2 / 2)
g_demo <- function(x, c) c * exp(0.5 - abs(x))
rlaplace <- function(n) { u <- runif(n); ifelse(u < 0.5, log(2 * u), -log(2 * (1 - u))) }

ui <- fluidPage(
    titlePanel(NULL),
    sidebarLayout(
        sidebarPanel(
            width = 4,
            sliderInput("c", "envelope scale c (g = c * e^(1/2 - |x|))",
                        min = 0.6, max = 3, value = 1.2, step = 0.05),
            helpText("c >= 1 keeps g above f everywhere: a valid, if looser, ",
                     "envelope. c < 1 breaks the envelope condition near ",
                     "|x| = 1, where the target actually exceeds it."),
            actionButton("resample", "draw new points"),
            tableOutput("tab")
        ),
        mainPanel(
            width = 8,
            plotOutput("scatter", height = "440px")
        )
    )
)

server <- function(input, output, session) {

    pts <- eventReactive(list(input$c, input$resample), {
        set.seed(input$resample + 1)
        N <- 500
        x <- rlaplace(N)
        v <- runif(N) * g_demo(x, input$c)
        list(x = x, v = v, acc = v < f_demo(x))
    })

    output$scatter <- renderPlot({
        p <- pts()
        xs <- seq(-6, 6, length.out = 500)
        par(mar = c(4, 4, 2, 1))
        plot(xs, g_demo(xs, input$c), type = "l", lwd = 3, col = "steelblue",
             ylim = c(0, max(2.2, 1.05 * input$c * exp(0.5))),
             xlab = "x", ylab = "v", main = sprintf("c = %.2f", input$c))
        lines(xs, f_demo(xs), lwd = 3, col = "firebrick")
        bad <- f_demo(xs) > g_demo(xs, input$c)
        if (any(bad)) lines(xs[bad], f_demo(xs)[bad], col = "black", lwd = 5)
        points(p$x[!p$acc], p$v[!p$acc], pch = 4, col = adjustcolor("gray40", 0.6), cex = 0.7)
        points(p$x[p$acc], p$v[p$acc], pch = 19, col = adjustcolor("firebrick", 0.7), cex = 0.7)
        legend("topright", bty = "n", lwd = c(3, 3, 5), col = c("firebrick", "steelblue", "black"),
               legend = c("target f", "envelope g", "f exceeds g here: envelope invalid"))
    })

    output$tab <- renderTable({
        p <- pts()
        data.frame(quantity = c("theoretical acceptance", "observed acceptance", "envelope valid?"),
                   value = c(sprintf("%.3f", 0.7602 / input$c), sprintf("%.3f", mean(p$acc)),
                             ifelse(input$c >= 1, "yes", "NO (c < 1)")))
    }, colnames = FALSE)
}

shinyApp(ui, server)
```

:::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit10_rejection_sampling.html#sec-rejection-app>.
:::

::: {.callout-note}
### Reproducing this figure
The chunk above uses the [shinylive](https://github.com/quarto-ext/shinylive)
extension, compiled to WebAssembly so it runs in the reader's browser with no
Shiny server (`quarto add quarto-ext/shinylive`, then `filters: [shinylive]`
in the document or project YAML). Only base R and `shiny` are available
inside the app.
:::

## Example: Sampling a Gamma Distribution

The Laplace envelope above was designed to match a normal target. A
different target needs a different envelope. Here is one built for
$\text{Gamma}(\alpha, 1)$, $\alpha > 1$, whose unnormalized log-density is
$(\alpha-1)\log x - x$.

```{r}
#| label: gamma-envelope-fns
### log of a function that is (for alpha > 2) always above the Gamma(alpha, 1) log-density
log_g_gamma <- function(x, alpha) {
    (alpha - 1) * (log(alpha - 1) - 1) - log(1 + (x - (alpha - 1))^2 / (2 * alpha - 1))
}
```

### Where the envelope comes from

`log_g_gamma` is not an arbitrary formula. Up to an additive constant that
depends only on $\alpha$, it is the log-density of a **rescaled, shifted
Cauchy** distribution: writing $Y = (\alpha-1) + \sqrt{2\alpha-1}\,C$ for a
standard Cauchy $C$, a change of variables gives
$\log f_Y(y) = -\log\pi - \tfrac12\log(2\alpha-1) - \log\!\big(1 + (y-(\alpha-1))^2/(2\alpha-1)\big)$,
which differs from `log_g_gamma(y, alpha)` only by a term that does not
depend on $y$. We can check this numerically rather than take it on faith:

```{r}
#| label: verify-cauchy-link
x_check <- seq(0.1, 10, by = 0.37)
alpha_check <- 3.3
log_f_Y <- dcauchy((x_check - (alpha_check - 1)) / sqrt(2 * alpha_check - 1), log = TRUE) -
    log(sqrt(2 * alpha_check - 1))
range(log_g_gamma(x_check, alpha_check) - log_f_Y)   # constant across x confirms the claim
```

So sampling the envelope is just sampling a (rescaled) Cauchy, and the
constant in `log_g_gamma` was chosen so that $g$ touches the Gamma density
exactly at the shared mode $x = \alpha - 1$. Whether $g$ stays *above* the
Gamma density everywhere else is a separate question, and the docstring's
claim "alpha must be > 2" is a claim about exactly that:

```{r}
#| label: tbl-envelope-threshold
#| tbl-cap: "Minimum of log g(x) - log f(x) over x > 0, for several alpha. Negative means the envelope is violated somewhere."
alphas_check <- c(1.1, 1.5, 1.9, 1.99, 2.0, 2.01, 2.1, 2.5, 3, 4.5)
xv <- seq(1e-6, 60, length.out = 2e4)
gaps <- vapply(alphas_check, function(a) min(log_g_gamma(xv, a) - dgamma(xv, a, log = TRUE)), numeric(1))
knitr::kable(data.frame(alpha = alphas_check, min_log_gap = gaps, envelope_valid = gaps >= 0), digits = 5)
```

The gap is exactly $0$ at $\alpha = 2$ and negative below it: the Cauchy-based
envelope is tangent to the Gamma density at $\alpha = 2$ and crosses below it
for smaller $\alpha$, so $\alpha > 2$ is precisely the validity condition, not
a conservative rule of thumb.

### Sampling function

A batch-and-refill implementation is both clearer and much faster than
drawing one candidate at a time in an R loop: generate a vector of
candidates, vectorize the accept/reject test, keep the survivors, and top up
only the shortfall.

```{r}
#| label: gamma-sampler
sample_gamma_rej <- function(n, alpha, max_tries = 2e6) {
    out <- numeric(0); tries <- 0
    while (length(out) < n) {
        if (tries > max_tries) stop("acceptance rate too low for this alpha: envelope is too loose")
        m <- n - length(out)
        x <- rcauchy(m) * sqrt(2 * alpha - 1) + (alpha - 1)
        tries <- tries + m
        keep <- log(runif(m)) < dgamma(x, shape = alpha, log = TRUE) - log_g_gamma(x, alpha)
        out <- c(out, x[keep])
    }
    structure(out[seq_len(n)], accept.rate = n / tries)
}
```

### Test cases and efficiency

```{r}
#| label: tbl-gamma-tests
#| tbl-cap: "Acceptance rate falls sharply as alpha grows: the Cauchy envelope
#|   is tuned to match curvature at alpha = 2 and gets looser as the Gamma
#|   density becomes more Gaussian-shaped further from that point."
set.seed(3)
alphas_test <- c(2.1, 2.5, 4.5)
gamma_tests <- lapply(alphas_test, function(a) sample_gamma_rej(2000, a))
knitr::kable(data.frame(
    alpha = alphas_test,
    accept_rate = vapply(gamma_tests, attr, numeric(1), "accept.rate"),
    ks_stat = mapply(function(x, a) suppressWarnings(ks.test(x, "pgamma", a))$statistic,
                      gamma_tests, alphas_test)
), digits = 4, col.names = c("alpha", "acceptance rate", "KS statistic vs true Gamma"))
```

```{r}
#| label: fig-gamma-tests
#| fig-width: 10
#| fig-height: 3.6
#| fig-cap: "Q-Q plots of rejection-sampled draws against the true Gamma(alpha,1)
#|   quantiles, for three of the test cases: the envelope produces exact draws
#|   regardless of alpha, only the cost per draw changes."
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
for (i in 1:3) {
    a <- alphas_test[i]
    qqplot(gamma_tests[[i]], rgamma(2000, a), xlab = "rejection sample", ylab = "rgamma sample",
           main = sprintf("alpha = %g (accept %.1f%%)", a, 100 * attr(gamma_tests[[i]], "accept.rate")))
    abline(0, 1, col = "firebrick", lwd = 2)
}
```

The Q-Q plots confirm exactness at every $\alpha$: rejection sampling never
trades accuracy for speed, only speed changes. Efficient random-number
generators for the Gamma distribution (as used inside `rgamma` itself) use
envelopes that stay tight across a much wider range of $\alpha$; see Marsaglia
& Tsang (2000, *ACM TOMS*) for the algorithm R actually uses.

## Adaptive Rejection Sampling for Log-Concave Targets

Designing an envelope by hand, as above, is the hard part of rejection
sampling. For a broad and important class of targets it can be automated.

### Log-concavity and the tangent hull

A density $f$ is **log-concave** if $\log f(x)$ is concave. Many posteriors
are: the normal, the gamma with shape $\ge 1$, the beta with both parameters
$\ge 1$, logistic-regression posteriors under normal priors, and any product
of log-concave factors (so most conjugate and many non-conjugate exponential-
family posteriors qualify).

A concave function lies below every one of its tangent lines. So for tangent
lines $\ell_1,\ldots,\ell_k$ to $\log f$ at points $x_1 < \cdots < x_k$, the
piecewise-linear **upper hull**
$$
\log g(x) = \min_j \ell_j(x)
$$
satisfies $\log g \ge \log f$ everywhere, hence $g = e^{\log g} \ge f$: a
valid envelope, built automatically from a handful of tangent points and
their slopes (the log-density's derivative), with no hand-crafted formula.
Because $g$ is piecewise-exponential, both its normalizing integral and its
inverse CDF are closed-form on each piece, so sampling from $g$ is cheap.

### The adaptive algorithm

**Adaptive rejection sampling** (Gilks & Wild, 1992) refines the hull as it
goes:

1. Start with a few tangent points, at least one on each side of the mode.
2. Draw $x \sim g$, $U \sim \text{Unif}(0,1)$; accept if $U < f(x)/g(x)$.
3. Whenever a point is **rejected**, add it to the tangent set and rebuild
   the hull.

Each rejection adds a tangent exactly where the envelope was loose, so the
acceptance rate rises toward $1$ as sampling proceeds — the envelope adapts
itself to the target instead of being fixed in advance. Only $\log f$ and its
derivative are required (a derivative-free variant uses secants in place of
tangents). The `ars` package implements this for any user-supplied
log-concave `f`.

### Example: Truncated Normal

The normal density is log-concave, so `ars` applies directly to a normal
truncated to $[l, u]$ — useful whenever a parameter has a hard bound (a
variance, a probability, a physically constrained rate).

```{r}
#| label: ars-fns
### direct rejection sampling for a truncated normal: draw from N(0,1),
### keep values inside [lb, ub]; vectorized batch-and-refill
sample_tnorm_drs <- function(n, lb = -Inf, ub = Inf) {
    out <- numeric(0)
    while (length(out) < n) {
        x <- rnorm(n - length(out))
        out <- c(out, x[x >= lb & x <= ub])
    }
    out
}

### adaptive rejection sampling via the ars package
sample_tnorm_ars <- function(n, lb, ub) {
    logf <- function(x) dnorm(x, log = TRUE)
    fprima <- function(x) -x
    ars(n, f = logf, fprima = fprima, x = c(lb, (lb + ub) / 2, ub),
        lb = TRUE, ub = TRUE, xlb = lb, xub = ub)
}
```

```{r}
#| label: tbl-ars-timing
#| tbl-cap: "Adaptive rejection sampling (ars) vs naive rejection sampling
#|   from an untruncated N(0,1): naive sampling degrades catastrophically
#|   as the truncation window moves into the tail, while ars does not,
#|   because it never wastes draws outside [lb, ub] in the first place."
set.seed(4)
n <- 1000
windows <- list(c(-5, -4), c(-50, -40), c(100, 110))
timing <- lapply(windows, function(w) {
    t_ars <- system.time(sample_tnorm_ars(n, w[1], w[2]))["elapsed"]
    t_naive <- if (w[1] > -10 && w[2] < 10) system.time(sample_tnorm_drs(n, w[1], w[2]))["elapsed"] else NA
    c(lb = w[1], ub = w[2], ars_seconds = t_ars, naive_seconds = t_naive)
})
knitr::kable(do.call(rbind, timing), digits = 4,
             col.names = c("lower bound", "upper bound", "ars (s)", "naive rejection (s)"))
```

For $[-50,-40]$ or $[100,110]$ the naive sampler's acceptance probability is
astronomically small ($P(-50<Z<-40) \approx 10^{-221}$), so it is left as
`NA` rather than run to timeout; `ars` samples these windows in a fraction of
a second because its envelope is built from the truncated density itself,
never from the untruncated normal.

```{r}
#| label: fig-ars
#| fig-width: 10
#| fig-height: 3.6
#| fig-cap: "Histograms of adaptive-rejection-sampled draws in three
#|   increasingly extreme truncation windows: ars has no trouble with
#|   any of them."
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
set.seed(5)
for (w in windows) {
    x <- sample_tnorm_ars(n, w[1], w[2])
    hist(x, breaks = 25, col = "steelblue", border = "white",
         main = sprintf("[%g, %g]", w[1], w[2]), xlab = "x")
}
```

## Summary

- Rejection sampling turns draws from an envelope $g \ge f$ into *exact* iid
  draws from $f$, using only the ratio $f/g$; no normalizing constants are
  needed. The geometric view — uniform points under $g$, keep those under
  $f$ — explains why this works and why efficiency is exactly the ratio of
  the two areas.
- A good envelope is tight (high acceptance) and easy to sample; finding one
  by hand, as in the Cauchy envelope for the Gamma distribution above,
  requires matching the target's tails and curvature and can be delicate.
- Adaptive rejection sampling automates envelope construction for any
  log-concave target via tangent lines to $\log f$, tightening the envelope
  on every rejection.
- Limitations: an envelope with a bounded ratio $g/f$ may not exist when $f$
  has heavier tails than any convenient $g$, and in high dimensions the
  acceptance rate of a product envelope typically decays exponentially with
  dimension, for the same curse-of-dimensionality reason quadrature grids
  fail (Unit 8). These limitations motivate **importance sampling**, which
  reweights instead of rejecting, and **Markov chain Monte Carlo**, which
  builds a chain that adapts to the target as it runs — the subjects of the
  next two chapters.
