---
title: "Importance Sampling"
format:
  html:
    toc: true
    tbl-colwidths: true
editor_options:
  chunk_output_type: console
code-fold: true
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(ars)
library(latex2exp)
```

Rejection sampling turns draws from an envelope $g \ge f$ into exact draws from
$f$, but every rejected proposal is wasted, and in the tails of a poor
envelope almost every proposal is rejected. **Importance sampling** keeps
every proposal instead of discarding most of them: draw from an easier
function $g$ and attach a *weight* to each draw that corrects for sampling
from the wrong distribution. The price is that the resulting sample is no
longer an exact iid sample from $f$ — it is a *weighted* sample — but nothing
is ever thrown away, and no envelope condition $g \ge f$ is required.

## Importance Reweighting Formulas

### Objective

Let $f(\theta)$ denote an unnormalized function of interest. In Bayesian inference, this typically takes the form $f(\theta) = L(\theta)\pi(\theta)$, representing the product of a likelihood function $L$ and a prior distribution $\pi$. The objective is generally to evaluate two types of integrals:


$$A = E_f\big[a(\theta)\big] = \frac{\int a(\theta)\,f(\theta)\,d\theta}{\int f(\theta)\,d\theta}, \qquad\qquad C_f = \int f(\theta)\,d\theta,$$


which correspond to a posterior expectation and a normalizing constant (the marginal likelihood), respectively. Standard Monte Carlo methods require sampling directly from the normalized target density $f/C_f$, which is computationally intractable in most complex settings.

Suppose we can instead sample from an alternative proposal distribution $g(\theta)$, which may also be unnormalized, with a corresponding normalizing constant $C_g = \int g(\theta)\,d\theta$. The absolute continuity requirement dictates that $g(\theta)$ must possess support wherever $f(\theta)$ does:


$$\{\theta : g(\theta) > 0\} \ \supseteq\ \{\theta : f(\theta) > 0\}.$$


Unlike rejection sampling, the proposal $g$ is not required to strictly dominate $f$ uniformly; there is no prerequisite that $g \ge f$. This characteristic renders importance sampling feasible in situations where constructing a bounding envelope is mathematically or computationally impractical.

### Fundamental Estimators

We define the unnormalized **importance weight** as $w(\theta) = f(\theta)/g(\theta)$. For any arbitrary function $a(\theta)$, the expectation under $f$ can be expressed as:


$$E_f\big[a(\theta)\big] = \int a(\theta)\,\frac{f(\theta)}{C_f}\,d\theta = \frac{C_g}{C_f}\, E_g\big[a(\theta)\,w(\theta)\big].$$


By evaluating the special case where $a \equiv 1$, we obtain the ratio of normalizing constants $C_f/C_g = E_g[w(\theta)]$. Combining these relations yields:


$$E_f\big[a(\theta)\big] = \frac{E_g\big[a(\theta)\,w(\theta)\big]}{E_g\big[w(\theta)\big]}.$$


Given a sequence of independent and identically distributed draws $\theta_1,\ldots,\theta_S \sim g/C_g$, we compute the corresponding weights $w_s = w(\theta_s)$. This operation leads directly to two primary estimators:

* **Ratio of Normalizing Constants:** The estimator $\widehat{C_f/C_g} = \frac{1}{S}\sum_{s=1}^S w_s$ provides an unbiased estimate for the ratio $C_f/C_g$. If $g$ is a properly normalized density (such that $C_g=1$), this term serves as an unbiased estimator for the marginal likelihood $C_f$ itself.
* **Self-Normalized Expectation:** The estimator $\hat{A} = \sum_{s=1}^S a(\theta_s)\,\tilde{w}_s$ utilizes the normalized weights $\tilde{w}_s = w_s / \sum_{j=1}^S w_j$, which inherently sum to $1$. This methodology transforms the proposal draws into a weighted sample from the target density $f$. Because any unknown proportionality constants in $f$ and $g$ cancel mathematically within the ratio, computing $\hat{A}$ only requires $f$ and $g$ to be specified up to a proportionality constant. While this estimator is formally biased due to being a ratio of random variables (with a bias of order $O(1/S)$), it is asymptotically consistent.

### Effective Sample Size (ESS)

The estimator $\hat{A}$ operates as a weighted average over $a(\theta_s)$. Draws located in regions where $f(\theta) \gg g(\theta)$ disproportionately influence the estimate, whereas draws in regions where $f(\theta) \ll g(\theta)$ contribute minimally. To quantify the efficiency of this weighted sample relative to an ideal scenario of independent draws, we compute the Effective Sample Size (ESS):


$$S_{\text{eff}} = \frac{1}{\displaystyle\sum_{s=1}^S \tilde{w}_s^2}.$$


If the normalized weights are perfectly uniform ($\tilde{w}_s = 1/S$), then $S_{\text{eff}} = S$, indicating optimal efficiency. Conversely, if a single weight heavily dominates the sample, $S_{\text{eff}} \approx 1$, implying that the inference relies almost entirely on a single draw.

**Theoretical Derivation.** The self-normalized estimator can be expressed as the ratio $\hat{A} = \bar{N}/\bar{D}$, where $N_s = w_s a(\theta_s)$ and $D_s = w_s$, utilizing $w = f/g$. Applying the delta method to this ratio yields a first-order variance approximation:


$$\operatorname{Var}_g(\hat{A}) \approx \frac{1}{S}E_g\!\bigl[w^{2}(a-A)^{2}\bigr] = \frac{1}{S}E_f\!\bigl[w(a-A)^{2}\bigr],$$


which relies on the identity $E_g[w\,h] = E_f[h]$. Assuming that the weight $w$ varies minimally across the primary support of $(a-A)^2$, we can approximate this integral by factoring it as $E_g[w^2]\operatorname{Var}_f(a)/S$ (substituting $E_f[w] = E_g[w^2]$). This produces the relationship:


$$\operatorname{Var}_g(\hat{A}) \approx \frac{\operatorname{Var}_f(a)}{S_{\text{eff}}^{\star}}, \qquad S_{\text{eff}}^{\star} := \frac{S}{E_g[w^{2}]},$$


where $S_{\text{eff}}^{\star}$ denotes the theoretical (or population) ESS. This quantity is an inherent property of the relationship between the distributions $f$ and $g$, independent of the target function $a$ or finite-sample effects. It represents the exact mathematical metric that reveals severe inefficiencies in pathological scenarios (e.g., missed target modes or heavy-tailed discrepancies), even when the empirically observed $S_{\text{eff}}$ appears deceptively stable.

**Empirical Estimation.** By substituting the sample moments $\frac{1}{S}\sum w_s$ for $E_g[w]$ and $\frac{1}{S}\sum w_s^2$ for $E_g[w^2]$, the theoretical metric $S_{\text{eff}}^{\star}$ transitions into the standard Kish's formula: $(\sum w_s)^2 / \sum w_s^2 = 1 / \sum \tilde{w}_s^2$. This demonstrates that Kish's formula functions as a principled plug-in estimator rather than an ad hoc heuristic.

**Limitations and Caveats.** The Kish approximation is not uniformly precise. For instance, utilizing the parameters $f = N(0,1)$, $g = N(0,2^2)$, $a(\theta) = \theta^2$, and $S = 4000$, empirical simulation yields $\operatorname{Var}(\hat{A}) \approx 3.15 \times 10^{-4}$, which aligns closely with exact delta-method calculations. However, the theoretical approximation $\operatorname{Var}_f(a)/S_{\text{eff}}^{\star}$ yields $7.56 \times 10^{-4}$ — an overestimation by more than a factor of two, caused by the negative correlation between $w$ and $(a-A)^2$ in this specific configuration. More critically, the empirical ESS ($\widehat{S_{\text{eff}}}$) is derived exclusively from the observed weights. If the proposal distribution fails to adequately explore regions where $f$ and $g$ diverge significantly, the estimator may yield a falsely reassuring $\widehat{S_{\text{eff}}}$, thereby masking severe sampling inadequacies.

```{r}
#| label: ess-helpers
### effective sample size from raw (possibly unnormalized) weights
ess <- function(w) sum(w)^2 / sum(w^2)
### numerically stable log(sum(exp(lx)))
log_sum_exp <- function(lx) { m <- max(lx); m + log(sum(exp(lx - m))) }
### effective sample size computed from LOG weights (avoids under/overflow)
log_ess <- function(lw) exp(2 * log_sum_exp(lw) - log_sum_exp(2 * lw))

```
## Example 1: Same Support, but a Mode Is Missed

$$
f(\theta) = \tfrac12\varphi(\theta+d) + \tfrac12\varphi(\theta-d), \qquad
g_{\sigma}(\theta) = \tfrac{1}{\sigma}\varphi\!\left(\tfrac{\theta+d}{\sigma}\right),
\qquad d = 5 .
$$

Both are positive on all of $\mathbb{R}$, so the support requirement holds. But
for $\sigma$ of order one the proposal concentrates around $-d$ and never
produces draws near the second mode at $+d$. For $\theta$ near $-d$ and
$\sigma=1$,
$$
w(\theta) = \frac{f(\theta)}{g_{1}(\theta)}
          = \tfrac12 + \tfrac12\,\frac{\varphi(\theta-d)}{\varphi(\theta+d)}
          = \tfrac12 + \tfrac12 e^{2d\theta} \;\approx\; \tfrac12 ,
$$
so every observed weight is nearly equal, $\tilde w_{i}\approx 1/S$ and
$S_{\text{eff}}\approx S$, while $\widehat{C_{f}/C_{g}}\approx\tfrac12$ against a
truth of $1$ and $\widehat{E}_{f}(\theta)\approx -d$ against a truth of $0$. The
draws that would expose the error carry $w\approx\tfrac12 e^{2d\theta}\gg 1$ near
$\theta=+d$ and have probability $\Phi(-2d)\approx 10^{-23}$ under $g_{1}$: they
never appear.

Both panels of @fig-missed-mode are drawn on the same $\theta$ axis. The lower
panel plots the *normalized weight function* $\theta\mapsto w(\theta)/\sum_{j}w_{j}$,
whose values at the sampled points are exactly the plotted stems, so the sample
can be seen sitting on the flat part of a function that rises without bound to
the right.

```{r}
#| label: fig-missed-mode
#| fig-cap: "Demonstrates a silent failure of importance sampling: when the proposal misses a mode of the target, the weights can look well behaved while the estimate is wrong.
#|   Top: the bimodal target f (firebrick) against the proposal g (steelblue)
#|   on a log density scale, with a rug of S = 2000 draws. Bottom: normalized
#|   importance weights at the sampled points, with the normalized weight function
#|   through them. Every observed weight is essentially 1/2, so S_eff is close to
#|   its nominal value while the estimate of C_f/C_g is half the truth."
#| fig-width: 9
#| fig-height: 7
d <- 5; sg <- 1; S <- 2000
set.seed(11)

## log of the mixture density, computed stably
log_f <- function(th, d) {
    a <- dnorm(th, -d, log = TRUE); b <- dnorm(th, d, log = TRUE)
    m <- pmax(a, b)
    m + log(exp(a - m) + exp(b - m)) - log(2)
}

## exact log E_g(w^2) for this pair; infinite when 2 - 1/sigma^2 <= 0,
## i.e. when sigma <= 1/sqrt(2).  Since E_g(w) = 1 exactly, the ideal
## effective fraction is 1 / E_g(w^2).
log_Ew2 <- function(d, sg) {
    A <- 2 - 1 / sg^2
    if (A <= 0) return(Inf)
    lterm <- function(a, b) {
        m <- (a + b) / 2
        B <- 2 * m + d / sg^2
        C <- 2 * m^2 - d^2 / sg^2
        dnorm(a, b, sqrt(2), log = TRUE) + log(sg) + 0.5 * log(2) +
            0.5 * log(2 * pi / A) - 0.5 * (C - B^2 / A)
    }
    lv <- c(lterm(-d, -d), log(2) + lterm(-d, d), lterm(d, d))
    m  <- max(lv)
    m + log(sum(exp(lv - m))) - log(4)
}

## ---- sample and weight ------------------------------------------------
th   <- rnorm(S, -d, sg)
logw <- log_f(th, d) - dnorm(th, -d, sg, log = TRUE)
mx   <- max(logw); v <- exp(logw - mx)
Chat <- mean(v) * exp(mx)                       # estimate of C_f / C_g, truth 1
se   <- sd(v) / sqrt(S) * exp(mx)
wn   <- v / sum(v)                              # normalized weights
ess_obs <- 1 / sum(wn^2)
Ehat <- sum(wn * th)                            # estimate of E_f(theta), truth 0
ess_x <- 100 * exp(-log_Ew2(d, sg))             # exact effective fraction

## ---- shared axis ------------------------------------------------------
xr <- range(c(th, -d - 4, d + 4))
xs <- seq(xr[1], xr[2], length.out = 1000)
fx <- exp(log_f(xs, d)); gx <- dnorm(xs, -d, sg)
lwc <- log_f(xs, d) - dnorm(xs, -d, sg, log = TRUE) - (mx + log(sum(v)))

layout(matrix(1:2, 2), heights = c(3, 2))

## ---- top panel: the two densities ------------------------------------
par(mar = c(0.4, 5, 3, 1))
yl <- c(1e-10, max(fx, gx) * 2)
plot(xs, fx, type = "n", log = "y", ylim = yl, xlim = xr, xaxt = "n",
     xlab = "", ylab = "density (log scale)",
     main = "bimodal target and a proposal centred on one mode")
lines(xs, fx, lwd = 4, col = "red")
lines(xs, gx, lwd = 3, col = "grey60")
rug(th, col = adjustcolor("grey45", 0.35))
legend("bottomright", bty = "n", cex = 1.0, text.col = "navy",
       lwd = c(4, 3, NA, NA, NA), col = c("grey25", "grey60", NA, NA, NA),
       legend = TeX(c(
           r"(target $f$)",
           sprintf(r"(proposal $g$, sd $= %.2f$)", sg),
           sprintf(r"($C_f/C_g$: true 1, est. %.3f)", Chat),
           sprintf(r"(95%% CI $(%.3f, %.3f)$)", Chat - 1.96 * se, Chat + 1.96 * se),
           sprintf(r"($E_f(\theta)$: true 0, est. %.2f)", Ehat))))

## ---- bottom panel: normalized weights --------------------------------
par(mar = c(4.5, 5, 0.4, 1))
wl <- c(min(wn) / 5, max(wn) * 5)
plot(th, wn, type = "n", log = "y", ylim = wl, xlim = xr,
     xlab = expression(theta), ylab = "normalized weight (log scale)")
lines(xs, exp(pmin(lwc, 700)), lwd = 2, col = "grey25")
segments(th, wl[1], th, wn, col = adjustcolor("grey45", 0.5))
abline(h = 1 / S, lty = 2, col = "grey55", lwd = 2)
legend("bottomright", bty = "n", cex = 1.0, text.col = "navy",
       lty = c(2, 1, NA, NA), lwd = c(2, 2, NA, NA),
       col = c("grey55", "grey25", NA, NA),
       legend = TeX(c(
           r"($1/S$)",
           r"(normalized weight function)",
           sprintf(r"(observed $S_{eff}$ = %.1f%% of S)", 100 * ess_obs / S),
           sprintf(r"(exact $S_{eff}$ = %.3g%% of S)", ess_x))))
```

The last two legend entries are the whole lesson. The observed effective sample
size is a statistic of the weights that actually appeared, and it reports
near-perfect efficiency. The exact effective fraction $1/E_{g}(w^{2})$, available
here in closed form because both densities are Gaussian mixtures, is smaller by
some forty orders of magnitude. At $\sigma=1$ the variance is finite, so the
central limit theorem does apply and the estimator is consistent; it would
simply need $S$ of order $10^{43}$ before its sampling distribution resembled a
normal one. Finiteness of the variance is itself fragile: the exact second
moment diverges once $2-\sigma^{-2}\le 0$, that is for every
$\sigma\le 1/\sqrt{2}$.

### Shinylive App for the Failure of Importance Sampling {#sec-missed-mode-app}

There is a [shinylive app](../shinyliveapps_compstat/app_missed-mode.html) to illustrate how importance sampling can fail without warning when the proposal is too narrow to cover the target.

## Example 2: Tail Probability of a Student-$t$ Distribution {#sec-ars-tnorm}

The target is a standard Student-$t_{\nu}$ distribution, with the degrees of
freedom $\nu$ a **tuning parameter of the target** that controls how heavy
its tail is: $\nu=1$ is the standard Cauchy, and $\nu\to\infty$ recovers the
standard normal. The goal is $P(c_{1}<X<c_{2})$ for an interval $(c_{1},c_{2})$
— possibly one-sided, $c_{2}=\infty$ — estimated from draws of a
truncated-normal proposal generated by adaptive rejection sampling (ARS).

As in the ARS section of the previous chapter, $\log\varphi$ is concave, and
restricting to $(c_{1},c_{2})$ (or one-sided, $(c_{1},\infty)$) preserves
concavity, so the `ars` package applies directly: `rtnorm_ars()` below just
hands it the log-density and its derivative, plus the truncation bounds.

```{r}
#| label: rtnorm-ars
## ---------------------------------------------------------------------
##  rtnorm_ars(): draw from N(mu, sigma^2) truncated to (lo, hi) (hi may
##  be Inf) via the ars package. log-phi is concave, so ars() applies
##  directly; x0 gives it three starting abscissae inside (lo, hi), with
##  the rightmost past mu when hi = Inf so the upper piece has negative
##  slope and stays integrable.
## ---------------------------------------------------------------------
rtnorm_ars <- function(n, mu = 0, sigma = 1, lo = 0, hi = Inf) {
    logf   <- function(x) dnorm(x, mu, sigma, log = TRUE)
    fprima <- function(x) -(x - mu) / sigma^2
    if (is.finite(hi)) {
        x0 <- lo + (hi - lo) * c(0.1, 0.5, 0.9)
    } else {
        right <- max(mu, lo) + 3 * sigma
        x0 <- sort(unique(c(lo + 0.1 * sigma, max(lo + 0.2 * sigma, mu + 1e-6), right)))
    }
    ars(n, f = logf, fprima = fprima, x = x0,
        lb = TRUE, xlb = lo, ub = is.finite(hi), xub = if (is.finite(hi)) hi else 0)
}
```

The `ars` package is itself available inside the browser — webR fetches it
from CRAN's WebAssembly mirror the first time `library(ars)` runs — so the
interactive app below calls this exact function, not a hand-rolled
substitute.

The contrast with the target is the point of the exercise: ARS needs
log-concavity, and $\log f_{t_{\nu}}$ is simply **not concave**, for any
finite $\nu$ — it is convex out in the tails, and only the limit
$\nu\to\infty$ (the normal) is concave everywhere. So ARS cannot sample the
target directly at any finite $\nu$; the truncated-normal proposal above is a
substitute, not a target-side match, which is exactly why the sample needs
importance weights at all.

### The estimator, and when its variance is infinite

A one-sided tail $(c,\infty)$ is a bad target for this proposal, so consider
instead a general interval $(c_{1},c_{2})$, with $c_{2}=\infty$ allowed as a
special case. Let $g$ be the normalized truncated normal on $(c_{1},c_{2})$
and $f(x)=f_{t_{\nu}}(x)I(c_{1}<x<c_{2})$. Draw $X_{1},\dots,X_{S}\sim g$,
weight each by $w(x)=f(x)/g(x)$, and average:
$$
\widehat{C} \;=\; \frac{1}{S}\sum_{i=1}^{S} w(X_{i}).
$$
This is unbiased for $C_{f}=P(c_{1}<X<c_{2})$ for *any* $g$ that is positive
wherever $f$ is — that is the whole mechanism of importance sampling, and it
does not care what $c_{1},c_{2},\mu,\sigma,\nu$ are.

Whether $\widehat C$ is *precise* is a different question, governed by
$$
\operatorname{Var}_{g}\bigl(w(X)\bigr)
 \;=\; \int_{c_{1}}^{c_{2}}\frac{f_{t_{\nu}}(x)^{2}}{g(x)}\,dx \;-\; C_{f}^{2} .
$$
When $c_{2}$ is finite, $f$ and $g$ share the same bounded interval, the
ratio $f/g$ is bounded on it, and the integral is finite — a well-behaved
estimator, whatever $\nu$ is. Send $c_{2}\to\infty$, though, and the tails
stop matching: $f_{t_{\nu}}(x)^{2}$ decays only polynomially
($\propto x^{-2(\nu+1)}$) while $g$'s Gaussian tail decays much faster, so the
integrand grows without bound and the integral diverges — for *every*
$\mu,\sigma$ and *every* finite $\nu$. The app below lets you flip between
the two regimes with a single checkbox, and, as before, $S_{\text{eff}}$ need
not warn you when you have landed in the bad one.

`is_t_interval()` turns the estimator above into code: draw $X$ from
`rtnorm_ars()`, form the log-weight $\log f_{t_\nu}(X) - \log g(X)$, and
report $\widehat C$, its standard error, and $S_{\text{eff}}$.

```{r}
#| label: is-t-interval
## ---------------------------------------------------------------------
##  is_t_interval(): importance-sampling estimate of P(c1 < X < c2) for a
##  standard t_df target, using X ~ N(mu, sigma^2) truncated to (c1, c2)
##  as the proposal.
## ---------------------------------------------------------------------
is_t_interval <- function(S, df, c1, c2, mu, sigma, seed = NULL) {
    if (!is.null(seed)) set.seed(seed)
    X <- rtnorm_ars(S, mu, sigma, c1, c2)

    ## log importance weight  log f_t(x; df) - log g(x),  with g the
    ## normalized truncated normal on (c1, c2)
    lZ   <- log(pnorm(c2, mu, sigma) - pnorm(c1, mu, sigma))
    logw <- dt(X, df, log = TRUE) - (dnorm(X, mu, sigma, log = TRUE) - lZ)

    m    <- max(logw)
    v    <- exp(logw - m)                       # rescaled weights
    Chat <- mean(v) * exp(m)                    # unbiased estimate of P(c1 < X < c2)
    se   <- sd(v) / sqrt(S) * exp(m)            # nominal standard error
    wn   <- v / sum(v)                          # normalized weights

    list(X = X, wn = wn, Chat = Chat, se = se, ess = 1 / sum(wn^2),
         truth = pt(c2, df) - pt(c1, df))
}
```

Four settings make the point concretely: a finite interval is well-behaved
at any $\nu$, while $c_{2}=\infty$ reproduces the divergence — badly at
$\nu=1$, but deceptively well at $\nu=30$, exactly as @tbl-is-t-examples
shows.

```{r}
#| label: tbl-is-t-examples
#| tbl-cap: "Shows how the choice of bounding constants affects importance sampling of a Student-$t$ interval, including a case where the weights have infinite variance.
#|   Four calls to `is_t_interval()` with S = 5000: a finite c2 is
#|   well-behaved at every df, c2 = Inf reproduces the divergent case,
#|   badly at df = 1 and only deceptively well at df = 30."
settings <- data.frame(df = c(1, 10, 1, 30), c1 = c(1, 1, 1, 1), c2 = c(3, 3, Inf, Inf))
res <- lapply(seq_len(nrow(settings)), function(i) with(settings[i, ],
    is_t_interval(S = 5000, df = df, c1 = c1, c2 = c2, mu = 0, sigma = 1, seed = 1)))

knitr::kable(data.frame(
    df       = settings$df,
    c1       = settings$c1,
    c2       = ifelse(is.finite(settings$c2), as.character(settings$c2), "Inf"),
    truth    = sapply(res, `[[`, "truth"),
    estimate = sapply(res, `[[`, "Chat"),
    se       = sapply(res, `[[`, "se"),
    ess_pct  = 100 * sapply(res, `[[`, "ess") / 5000
), digits = 4, col.names = c("df", "c1", "c2", "truth", "estimate", "s.e.", "ESS %"))
```

### Shinylive App for Importance Sampling of a Student-$t$ Interval {#sec-t-is-app}

There is a [shinylive app](../shinyliveapps_compstat/app_t-is.html) to show how the choice of constants and proposal determines whether importance sampling of a Student-$t$ interval is reliable, by displaying the target, the proposal, and the importance weights.

## Example 3: Marginal Likelihood of Normal Data via Importance Sampling


The [quadrature](unit08_integral.qmd) and Laplace-approximation chapters
compute the marginal likelihood $P(y) = \int L(\theta)\pi(\theta)\,d\theta$ of
the same conjugate-free Gaussian model — unknown mean $\mu$ and
$w=\log\sigma^2$, independent normal priors — by a product grid and by a
mode-and-Hessian approximation, respectively. Importance sampling offers a
third route, and one that comes with a built-in accuracy diagnostic
($S_{\text{eff}}$) that neither of the other two methods provides directly.

Take $g(\mu,w)$ to be the **Gaussian approximation to the posterior**
obtained by a Laplace fit: locate the posterior mode $(\hat\mu,\hat w)$ by
optimization, take $g = N_2\big((\hat\mu,\hat w),\, H^{-1}\big)$ for $H$ the
Hessian of $-\log[L\pi]$ at the mode, and importance-sample from $g$ with
$f = L\pi$:
$$
\widehat{P(y)} = \frac{1}{S}\sum_{s=1}^S w_s, \qquad
w_s = \frac{L(\theta_s)\pi(\theta_s)}{g(\theta_s)}, \qquad \theta_s \sim g.
$$
Because $g$ is shaped like the posterior near its mode, this is a "good
story" proposal by construction — in sharp contrast to sampling directly from
the (typically much wider, uninformative) prior, which is exactly the
`log_marlik_mc` function used elsewhere in this book as a naive-Monte-Carlo
check: it is importance sampling with $g = \pi$, and its weights are just the
likelihood values $L(\theta_s)$.

**Implementation**

```{r}
#| label: marlik-functions
log_lik   <- function(x, mu, w) sum(dnorm(x, mu, exp(w / 2), log = TRUE))
log_prior <- function(mu, w, mu_0, sigma_mu, w_0, sigma_w) {
    dnorm(mu, mu_0, sigma_mu, log = TRUE) + dnorm(w, w_0, sigma_w, log = TRUE)
}
neg_log_post <- function(theta, x, mu_0, sigma_mu, w_0, sigma_w) {
    -log_lik(x, theta[1], theta[2]) -
        log_prior(theta[1], theta[2], mu_0, sigma_mu, w_0, sigma_w)
}

### log density of N(mu, A^{-1}) at every column of a 2 x S matrix of points at once,
### replacing a per-point apply() call with one vectorized quadratic form
log_dmvnorm_batch <- function(Theta, mu, A) {
    Tc <- Theta - mu
    quad <- colSums((A %*% Tc) * Tc)
    0.5 * (-nrow(Theta) * log(2 * pi) + sum(log(svd(A)$d)) - quad)
}

### log marginal likelihood by importance sampling from a Laplace-shaped Gaussian.
### Besides S_eff, attaches "se": the standard error of the *log* estimate,
### sd(v) / (sqrt(S) * mean(v)) for v = exp(lw - max(lw)) -- the delta-method
### transfer of the raw estimator's SE onto the log scale, exactly as used for
### the log estimate in the t-interval app above.
log_marlik_imps <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    fit <- nlm(neg_log_post, c(mean(x), log(stats::var(x))), hessian = TRUE,
               x = x, mu_0 = mu_0, sigma_mu = sigma_mu, w_0 = w_0, sigma_w = sigma_w)
    mu_hat <- fit$estimate; H <- fit$hessian
    L <- t(chol(solve(H)))                        # Sigma = H^{-1} = L L'
    Theta <- L %*% matrix(rnorm(2 * S), 2, S) + mu_hat

    log_g <- log_dmvnorm_batch(Theta, mu_hat, H)
    log_f <- -apply(Theta, 2, neg_log_post, x = x, mu_0 = mu_0, sigma_mu = sigma_mu,
                     w_0 = w_0, sigma_w = sigma_w)
    lw <- log_f - log_g
    m  <- max(lw); v <- exp(lw - m)
    structure(log_sum_exp(lw) - log(S),
              ess = log_ess(lw), se = sd(v) / (sqrt(S) * mean(v)))
}

### log marginal likelihood by naive Monte Carlo from the prior (g = pi): a
### second, independent importance-sampling estimator with an unfavourable g
log_marlik_prior <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    mus <- rnorm(S, mu_0, sigma_mu); ws <- rnorm(S, w_0, sigma_w)
    lw <- vapply(seq_len(S), function(i) log_lik(x, mus[i], ws[i]), numeric(1))
    m  <- max(lw); v <- exp(lw - m)
    structure(log_sum_exp(lw) - log(S),
              ess = log_ess(lw), se = sd(v) / (sqrt(S) * mean(v)))
}

### "truth": log marginal likelihood by a midpoint grid over (mu, w), centred
### and scaled by the Laplace fit. The likelihood term is vectorized over the
### whole mu-grid at once via the sufficient statistic sum((x - mu)^2), so --
### like log_dmvnorm_batch above -- no R-level loop over grid points, or over
### data points beyond the one sum(), is needed.
log_marlik_grid <- function(x, mu_0, sigma_mu, w_0, sigma_w, n_grid = 400, width = 8) {
    n   <- length(x)
    fit <- nlm(neg_log_post, c(mean(x), log(stats::var(x))), hessian = TRUE,
               x = x, mu_0 = mu_0, sigma_mu = sigma_mu, w_0 = w_0, sigma_w = sigma_w)
    se  <- sqrt(diag(solve(fit$hessian)))
    mu_grid <- fit$estimate[1] + seq(-width, width, length.out = n_grid) * se[1]
    w_grid  <- fit$estimate[2] + seq(-width, width, length.out = n_grid) * se[2]

    SS <- sum(x^2) - 2 * mu_grid * sum(x) + n * mu_grid^2   # sum((x - mu)^2)
    log_lik_grid   <- outer(SS, w_grid,
                             function(ss, w) -0.5 * n * log(2 * pi) - n * w / 2 - 0.5 * exp(-w) * ss)
    log_prior_grid <- outer(dnorm(mu_grid, mu_0, sigma_mu, log = TRUE),
                             dnorm(w_grid, w_0, sigma_w, log = TRUE), "+")

    h_mu <- diff(mu_grid[1:2]); h_w <- diff(w_grid[1:2])
    log_sum_exp(as.vector(log_lik_grid + log_prior_grid)) + log(h_mu) + log(h_w)
}
```


`log_marlik_grid()` supplies ground truth by quadrature — a midpoint grid
over $(\mu,w)$, centred and scaled by the same Laplace fit used for the
importance proposal, wide enough (8 posterior SDs each way, on a $400\times
400$ grid) that refining it further does not move the answer. Comparing both
proposals against it in a single run, with a 95% CI built from the standard
error of that run, makes the difference between them concrete rather than
just asserted.

**Results across sample sizes**

Both proposals target the same $f=L\pi$, and the prior is a legitimate
proposal for it: every posterior draw has positive prior density, so the
support condition holds. The two differ only in how well their shape matches
the target's, and the following table varies the amount of data to make that
difference visible.

```{r}
#| label: tbl-marlik-merged
#| tbl-cap: "Compares importance sampling proposals for estimating a marginal likelihood, showing that a Laplace-shaped proposal is more reliable than the prior as $n$ grows.
#|   Log marginal likelihood for a freshly simulated data set at each
#|   sample size n: truth by grid quadrature, against single runs of importance
#|   sampling (S = 2000) from a Laplace-shaped proposal and from the prior."
set.seed(4)
ns <- c(5, 20, 100, 400)
S  <- 2000

## one proposal, one data set: estimate, interval and S_eff as a % of S
summarize_is <- function(fn, x, S) {
    est <- fn(x, 0, 5, 0, 5, S)
    se  <- attr(est, "se")
    c(estimate = sprintf("%.2f", as.numeric(est)),
      ci       = sprintf("(%.2f, %.2f)",
                         as.numeric(est) - 1.96 * se,
                         as.numeric(est) + 1.96 * se),
      ess      = sprintf("%.1f", 100 * attr(est, "ess") / S))
}

merged_tbl <- do.call(rbind, lapply(ns, function(n) {
    x   <- rnorm(n)
    lap <- summarize_is(log_marlik_imps,  x, S)
    pri <- summarize_is(log_marlik_prior, x, S)
    data.frame(n     = n,
               truth = sprintf("%.2f", log_marlik_grid(x, 0, 5, 0, 5)),
               lap_estimate = lap[["estimate"]],
               lap_ci       = lap[["ci"]],
               lap_ess      = lap[["ess"]],
               pri_estimate = pri[["estimate"]],
               pri_ci       = pri[["ci"]],
               pri_ess      = pri[["ess"]],
               row.names    = NULL)
}))

## repeated names are intentional: the proposal is carried by the banner
names(merged_tbl) <- c("n", "truth",
                       "estimate", "95% CI", "ESS %",
                       "estimate", "95% CI", "ESS %")

tinytable::tt(merged_tbl) |>
    tinytable::group_tt(j = list("Laplace-shaped proposal" = 3:5,
                                 "prior sampling"          = 6:8))
```

The Laplace-shaped proposal re-centers and re-scales itself around the
posterior mode for every data set. Its interval stays tight and centred on the
truth at every $n$, and $S_{\text{eff}}$ stays close to the nominal $S$: a
second-order approximation to the posterior is evidently a good match to it.
Prior sampling degrades monotonically. The posterior concentrates at rate
$n^{-1/2}$ while the prior does not move, so the fraction of prior mass lying
where the likelihood is appreciable shrinks with $n$, the weights become ever
more concentrated on a handful of draws, and $S_{\text{eff}}$ falls to a small
fraction of one percent. Its interval may still cover the truth, but it is only
as trustworthy as those few draws happen to be from one run to the next.

This is the marginal-likelihood instance of the lesson drawn from the
missed-mode example: a fixed, data-independent proposal becomes an increasingly
poor match to a target that concentrates as data accumulate, and only a
proposal that adapts to the target avoids the resulting collapse in efficiency.
Note also the contrast in how the two failures announce themselves. Prior
sampling fails *noisily* — $S_{\text{eff}}$ reports the collapse directly, which
is exactly what the diagnostic is built to catch. The earlier tail examples
failed *silently*, with $S_{\text{eff}}$ near its nominal value while the
estimate was badly wrong, because there the weights were uniform among the
draws actually observed and the trouble lay entirely outside the region the
proposal ever visited.

### Shinylive App for Sorted Importance Weights {#sec-marlik-app}

There is a [shinylive app](../shinyliveapps_compstat/app_marlik.html) to show how the distribution of the importance weights reveals the quality of a proposal for estimating a marginal likelihood, comparing a Laplace-shaped proposal with the prior.

## Example 4: Area of an Irregular Region

The two fundamental estimators from [Fundamental Estimators] were stated for
a generic unnormalized target $f$, but nothing in their derivation assumes
$f$ is a likelihood or a density. Taking $f(\theta) = \mathbf 1_A(\theta)$,
the indicator of a region $A \subset \mathbb R^2$, collapses $C_f = \int
f(\theta)\,d\theta$ to $\text{Area}(A)$, and the ratio-of-normalizing-constants
estimator $\widehat{C_f/C_g} = \frac1S\sum_s w_s$ becomes an importance-sampling
estimator of that area. When $g$ is uniform on a bounding box, this is exactly
the "throw darts and count hits" trick used earlier to estimate $\pi$ as the
area of a circle — importance sampling only starts to matter once the region
is irregular, or small relative to its bounding box, so that the choice of
$g$ affects how many of the $S$ draws actually land somewhere useful.

Consider a region built by a Boolean combination of two disks — a crescent
$A = D_1 \setminus D_2$, where $D_1$ is the unit disk centred at the origin
and $D_2$ is a disk of radius $r_2 = 0.9$ centred at $(d, 0)$ with $d = 0.3$,
overlapping most of $D_1$. This is a **constructive solid geometry (CSG)**
shape: testing whether a point lies in $A$ is a one-line computation, but the
area of $A$ itself has no elementary formula once more than a couple of
primitives are combined this way — exactly the situation CAD and engineering
software face after a handful of Boolean operations on solids. Two circles
are simple enough that a closed form still exists, from the standard
circle-intersection formula; it is kept here only to check the Monte Carlo
estimates against ground truth.

```{r}
#| label: crescent-functions
r1 <- 1; r2 <- 0.9; d <- 0.3
in_A <- function(x, y) (x^2 + y^2 <= r1^2) & ((x - d)^2 + y^2 > r2^2)

### exact area of A, from the circle-circle intersection formula; used only
### to check the Monte Carlo estimates below
circle_intersection_area <- function(r1, r2, d) {
    r1^2 * acos((d^2 + r1^2 - r2^2) / (2 * d * r1)) +
    r2^2 * acos((d^2 + r2^2 - r1^2) / (2 * d * r2)) -
    0.5 * sqrt((-d + r1 + r2) * (d + r1 - r2) * (d - r1 + r2) * (d + r1 + r2))
}
area_true <- pi * r1^2 - circle_intersection_area(r1, r2, d)

### bounding box for the naive proposal: generous, with no attempt to hug A
xlim <- c(-2, 2); ylim <- c(-2, 2)

### importance-sampling estimate of Area(A) under one of two proposals:
### "box" samples uniformly from the bounding box; "disk" samples uniformly
### from D1, which is known to contain A even though A itself is not known
### in closed form
crescent_is <- function(proposal, S, seed) {
    set.seed(seed)
    if (proposal == "box") {
        x <- runif(S, xlim[1], xlim[2]); y <- runif(S, ylim[1], ylim[2])
        g <- 1 / ((xlim[2] - xlim[1]) * (ylim[2] - ylim[1]))
    } else {
        rr <- r1 * sqrt(runif(S)); th <- runif(S, 0, 2 * pi)
        x <- rr * cos(th); y <- rr * sin(th)
        g <- 1 / (pi * r1^2)
    }
    w <- in_A(x, y) / g
    list(x = x, y = y, hit = in_A(x, y),
         estimate = mean(w), me = 1.96 * sd(w) / sqrt(S), ess = ess(w))
}
```

```{r}
#| label: fig-crescent-darts
#| echo: true
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Compares a naive bounding-box proposal with a proposal matched to
#|   the target's support for estimating the area of an irregular region.
#|   1,500 darts thrown at the crescent $A = D_1 \\setminus D_2$ (outlined in
#|   black) from a generous bounding box (left) and from $D_1$ itself (right);
#|   blue points land in $A$, red points miss."
res_box_vis  <- crescent_is("box",  1500, seed = 1)
res_disk_vis <- crescent_is("disk", 1500, seed = 1)
theta_seq <- seq(0, 2 * pi, length.out = 200)

par(mfrow = c(1, 2))
plot(res_box_vis$x, res_box_vis$y, asp = 1, pch = 20,
     col = ifelse(res_box_vis$hit, "steelblue", "tomato"),
     xlab = "x", ylab = "y",
     main = paste(sum(res_box_vis$hit), "of 1500 hit A (box)"))
rect(xlim[1], ylim[1], xlim[2], ylim[2], border = "black", lwd = 2)
lines(r1 * cos(theta_seq), r1 * sin(theta_seq), lwd = 2)
lines(d + r2 * cos(theta_seq), r2 * sin(theta_seq), lwd = 2)

plot(res_disk_vis$x, res_disk_vis$y, asp = 1, pch = 20,
     col = ifelse(res_disk_vis$hit, "steelblue", "tomato"),
     xlab = "x", ylab = "y", xlim = xlim, ylim = ylim,
     main = paste(sum(res_disk_vis$hit), "of 1500 hit A (disk)"))
lines(r1 * cos(theta_seq), r1 * sin(theta_seq), lwd = 2)
lines(d + r2 * cos(theta_seq), r2 * sin(theta_seq), lwd = 2)
par(mfrow = c(1, 1))
```

```{r}
#| label: tbl-crescent-is
#| tbl-cap: "Shows that both proposals are unbiased for the true area, and
#|   that only the effective sample size distinguishes them: the disk
#|   proposal wastes draws only on $D_2$, while the box proposal also wastes
#|   draws outside $D_1$ entirely.
#|   Importance-sampling estimates of Area(A) with S = 5000, box versus disk
#|   proposal."
S <- 5000
res_box  <- crescent_is("box",  S, seed = 1)
res_disk <- crescent_is("disk", S, seed = 1)

knitr::kable(data.frame(
    proposal = c("Bounding box (naive)", "Disk $D_1$ (importance)"),
    truth    = rep(area_true, 2),
    estimate = c(res_box$estimate, res_disk$estimate),
    ME      = c(res_box$me, res_disk$me),
    ess_pct  = 100 * c(res_box$ess, res_disk$ess) / S
), digits = 4, col.names = c("Proposal $g$", "True area", "Estimate", "M.E.", "ESS %"))
```

Both rows of @tbl-crescent-is are unbiased for the true area — importance
sampling never needed to know $A$'s exact shape, only a proposal whose
support covers it. What changes is efficiency: the box proposal's effective
sample size is under 6% of $S$, because most darts land outside $D_1$
entirely and carry weight zero, while the disk proposal's effective sample
size is close to 28% of $S$ — five times larger, since every draw at least
lands in $D_1$ and only the ones also landing in $D_2$ are wasted. The
standard error falls by roughly the square root of that factor. Unlike
@fig-missed-mode, no bias is at stake here — a bounding box that is too
generous costs precision, not correctness — but the lesson is the same:
match the proposal to where the target actually lives.

## Advanced Applications of Importance Sampling

### The physics: partition functions and free energy

Statistical physics supplies a second reading of the same estimator. A system
with energy $U(x)$ at inverse temperature $\beta$ has partition function and
free energy
$$
Z(\beta) = \int e^{-\beta U(x)}\,dx, \qquad F(\beta) = -\frac{1}{\beta}\log Z(\beta).
$$
$e^{-\beta U(x)}/Z(\beta)$ is the Boltzmann distribution at temperature
$1/\beta$: large $\beta$ (low temperature) concentrates mass where $U$ is
small, small $\beta$ (high temperature) spreads it out.

To move between two temperatures, importance-sample from $g(x)=e^{-\beta_0
U(x)}$ to $f(x) = e^{-\beta_1 U(x)}$. The weight is $w(x) =
f(x)/g(x)=e^{-(\beta_1-\beta_0)U(x)}$, and the ratio-of-normalizing-constants
estimator from the start of the chapter reads
$$
\frac{Z(\beta_1)}{Z(\beta_0)} = E_{\beta_0}\!\left[e^{-(\beta_1-\beta_0)U(x)}\right],
$$
an expectation under the $\beta_0$-tempered distribution.

#### The Bayesian analogue

Set $U(\theta) = -\log L(\theta)$. Then $\beta=0$ gives $e^{-\beta U}\propto
1$ — up to normalization the prior, taking $g=\pi(\theta)$ — and $\beta=1$
gives $e^{-U}=L(\theta)$, so $f=L(\theta)\pi(\theta)$ is the (unnormalized)
posterior. The ratio $Z(1)/Z(0)$ is exactly the marginal likelihood
$C_f/C_g$ of the previous section:
$$
P(y) = \frac{C_f}{C_g} = E_\pi\!\big[L(\theta)\big] \approx \frac1S\sum_{s=1}^S L(\theta_s), \qquad \theta_s\sim\pi,
$$
and $-\log P(y)$ is, in this language, the free energy of the posterior at
$\beta=1$. This is not a new estimator — it is `log_marlik_prior()` from
above, rederived from a different starting point. The physics vocabulary
just names the failure mode precisely: jumping straight from $\beta=0$ to
$\beta=1$ asks one importance-sampling step to bridge the *entire*
temperature range in a single move, and the wider the prior is relative to
the likelihood, the more that one jump resembles the badly matched proposals
seen earlier in this chapter.

#### Stepping stones: many small jumps instead of one big one

The fix follows from the algebra of the ratio itself. Chaining intermediate
temperatures $\beta_0=0<\beta_1<\dots<\beta_K=1$,
$$
\frac{Z(1)}{Z(0)} = \prod_{k=1}^K \frac{Z(\beta_k)}{Z(\beta_{k-1})}, \qquad
\log Z(1)-\log Z(0) = \sum_{k=1}^K \log\frac{Z(\beta_k)}{Z(\beta_{k-1})},
$$
and each factor is its own small importance-sampling problem: sample from
the tempered distribution at $\beta_{k-1}$, weight by
$e^{-(\beta_k-\beta_{k-1})U(\theta)}$. If the temperatures are close enough
together, $g_{\beta_{k-1}}$ and $f_{\beta_k}$ nearly coincide and every step
is a good-story proposal with near-constant weights — the same principle as
building an envelope from many well-fitting pieces rather than one loose
one. This chain of intermediate distributions is a **stepping-stones**
estimator (a discretization of thermodynamic integration, and the same
schedule idea behind simulated annealing and parallel tempering).

### Bayesian Leave-One-Out Cross-Validation

#### The leave-one-out predictive density

Cross-validation asks how well a model predicts observations it was not
fitted to. The Bayesian version of a single leave-one-out fold is the
**leave-one-out predictive density**
$$
p(y_i\mid y_{-i}) = \int p(y_i\mid\theta)\,p(\theta\mid y_{-i})\,d\theta,
$$
the posterior predictive density for $y_i$ under the posterior that never
saw $y_i$. Summed in log over $i=1,\dots,n$, these give the expected log
pointwise predictive density, a standard measure of a model's out-of-sample
fit. Computed directly it requires refitting the model $n$ times — one full
posterior per left-out observation — exactly the cost importance sampling is
suited to avoid.

#### Reusing the full posterior as the proposal

Take $g(\theta)=p(\theta\mid y)$, the **full** posterior — already available
from a single fit — as the proposal, and target the leave-one-out posterior
$f(\theta)=p(\theta\mid y_{-i})$. Since $p(\theta\mid
y)\propto p(y_i\mid\theta)\,p(\theta\mid y_{-i})$,
$$
f(\theta) = p(\theta\mid y_{-i}) \;\propto\; \frac{p(\theta\mid y)}{p(y_i\mid\theta)}, \qquad
w_i(\theta) = \frac{f(\theta)}{g(\theta)} \;\propto\; \frac{1}{p(y_i\mid\theta)}.
$$
Reusing the self-normalized estimator from the start of the chapter with
$a(\theta)=p(y_i\mid\theta)$,
$$
\widehat{p(y_i\mid y_{-i})} =
\frac{\sum_{s} p(y_i\mid\theta_s)\,w_i(\theta_s)}{\sum_s w_i(\theta_s)} =
\frac{\sum_s p(y_i\mid\theta_s)/p(y_i\mid\theta_s)}{\sum_s 1/p(y_i\mid\theta_s)} =
\frac{S}{\sum_{s=1}^S 1/p(y_i\mid\theta_s)},
$$
the **harmonic mean** of the likelihood contributions $p(y_i\mid\theta_s)$
over draws from the *full* posterior (Gelfand, Dey & Chang, 1992) — one
weighted average per left-out observation, all reusing the same $S$
posterior draws from the single original fit.



## Summary

- Importance sampling reweights draws from a proposal $g$ by
  $w(\theta) = f(\theta)/g(\theta)$ instead of accepting or rejecting them,
  using $\widehat{C_f/C_g} = \frac1S\sum_s w_s$ and the self-normalized
  $\hat A = \sum_s a(\theta_s)\tilde w_s$; only the support condition
  $\{g>0\}\supseteq\{f>0\}$ is required, not an envelope $g \ge f$.
- A good proposal shaped like the target on its support gives near-constant
  weights and $S_{\text{eff}} \approx S$ (the truncated-normal and
  Laplace-proposal examples), and a single weighted sample can answer several
  questions about the same target — a probability and an expectation alike.
- A proposal that misses a mode or part of the support fails **silently**: the
  observed weights can look perfectly healthy while both $C_f/C_g$ and
  $E_f[a]$ are badly wrong, and no amount of extra sampling fixes it.
  $S_{\text{eff}}$ diagnoses only the noisy failure (wildly varying observed
  weights), never the silent one.
- For the marginal likelihood, a proposal built from a Laplace approximation
  at the posterior mode remains efficient as the data accumulate and the
  posterior sharpens, while a fixed, data-independent proposal such as the
  prior does not — reinforcing why the mode-finding methods from earlier
  chapters remain useful even once we turn to sampling-based methods.
- Recast as free energy, the marginal likelihood's prior-versus-posterior
