---
title: "Laplace Approximation"
format:
  html:
    toc: true
    tbl-colwidths: true
filters:
  - shinylive
editor_options:
  chunk_output_type: console
code-fold: true
---

The previous chapter approximated an integral by a weighted sum of function
values on a grid. A grid gets expensive fast: an $n$-point-per-axis product
grid costs $n^d$ evaluations, so quadrature is only practical for $d \lesssim
3$. The **Laplace approximation** takes a completely different route: instead
of evaluating the integrand everywhere, it looks only at the single most
important point — the mode — and replaces the whole integrand there by the
Gaussian that matches its height and curvature. No grid is needed at any
dimension, at the price of assuming the integrand is roughly bell-shaped.

## The Goal and the Idea

Many Bayesian calculations are posterior expectations
$$
E_f\big[a(\theta)\big] = \frac{\int a(\theta)\,f(\theta)\,d\theta}{\int f(\theta)\,d\theta},
$$
where $\theta \in \mathbb{R}^p$, $a(\theta)$ is some quantity of interest (a
parameter, a prediction, an indicator), and $f(\theta)$ is an **unnormalized**
posterior, typically $f(\theta) = L(\theta)\pi(\theta)$. Both numerator and
denominator are $p$-dimensional integrals.

Write $h(\theta) = -\log f(\theta)$, so $f(\theta) = e^{-h(\theta)}$. For
independent data, $h$ is (up to an additive constant) minus the
log-likelihood plus minus the log-prior — a sum of $n$ terms — so as $n$ grows
it becomes large and sharply curved at its minimum. This is the same
large-sample regularity that makes the MLE asymptotically normal
(seen in the maximum-likelihood chapters); here it makes the *posterior* asymptotically normal too. The
Laplace approximation makes this precise by expanding $h$ to second order at
its minimum and integrating the resulting Gaussian exactly — no grid, no
sampling, just one optimization and one Hessian.

## Derivation

Let $\theta_0 = \arg\min_\theta h(\theta)$, the posterior mode, found
numerically (`optim(..., hessian = TRUE)` or `nlm(..., hessian = TRUE)` in
R — exactly the optimizers used to find the MLE in earlier chapters). Expand
$h$ around $\theta_0$:
$$
h(\theta) = h(\theta_0) + \underbrace{\nabla h(\theta_0)^\top}_{=\,0\text{ at the minimum}}(\theta - \theta_0)
 + \tfrac12(\theta-\theta_0)^\top H (\theta - \theta_0) + O\big(\|\theta-\theta_0\|^3\big),
\qquad H = \nabla^2 h(\theta_0).
$$
$H$, the Hessian of $-\log f$ at the mode, is positive definite at a genuine
minimum; for $f = L\pi$ it is the observed information plus the curvature of
the log-prior. Dropping the cubic remainder and exponentiating,
$$
f(\theta) \approx \tilde f(\theta) = f(\theta_0)\exp\Big\{-\tfrac12(\theta-\theta_0)^\top H(\theta-\theta_0)\Big\}
 = \underbrace{f(\theta_0)(2\pi)^{p/2}|H|^{-1/2}}_{\text{constant}}
   \cdot \underbrace{(2\pi)^{-p/2}|H|^{1/2}e^{-\frac12(\theta-\theta_0)^\top H(\theta-\theta_0)}}_{N_p(\theta_0,\,H^{-1})\text{ density}}.
$$
Two results follow immediately, since the bracketed factor is a properly
normalized $N_p(\theta_0, H^{-1})$ density:

$$
\boxed{\ \theta \mid \text{data} \ \approx\ N_p\big(\theta_0,\ H^{-1}\big) \ }
\qquad\qquad
\boxed{\ \int f(\theta)\,d\theta \ \approx\ f(\theta_0)\,(2\pi)^{p/2}\,|H|^{-1/2} \ }
$$

Only the mode and the Hessian of $-\log f$ at the mode are needed — both come
for free from any optimizer that returns a Hessian. The relative error of the
second result is $O(1/n)$ for regular models; the error of the first is
governed by how skewed the true posterior actually is.

## Worked Example: A Poisson Rate with a Gamma Prior

To see the approximation against ground truth, use a model with a known exact
answer. For $y_1,\ldots,y_n \mid \lambda \sim \text{Poisson}(\lambda)$ and
$\lambda \sim \text{Gamma}(\alpha,\beta)$, the posterior is exactly
$\text{Gamma}(\alpha + \sum y_i,\ \beta+n)$, so both the exact log marginal
likelihood and the exact posterior are available in closed form and the
Laplace error can be measured directly.

```{r}
#| label: poisson-gamma-functions
alpha <- 2; beta <- 1

### Laplace approximation: mode, Hessian, and the two boxed results above
laplace_poisson <- function(y) {
  h  <- function(l) -(sum(dpois(y, l, log = TRUE)) + dgamma(l, alpha, beta, log = TRUE))
  op <- optim(mean(y) + 0.5, h, method = "L-BFGS-B", lower = 1e-6, hessian = TRUE)
  c(mode = op$par, se = sqrt(1 / op$hessian[1, 1]),
    log_ml = -op$value + 0.5 * log(2 * pi) - 0.5 * log(op$hessian[1, 1]))
}

### exact log marginal likelihood, from the Gamma-Poisson conjugate identity
exact_log_ml <- function(y) {
  n <- length(y); s <- sum(y)
  alpha * log(beta) - lgamma(alpha) + lgamma(alpha + s) -
    (alpha + s) * log(beta + n) - sum(lfactorial(y))
}
```

```{r}
#| label: tbl-poisson-gamma-error
#| tbl-cap: "Laplace log marginal likelihood versus the exact value, and the posterior mean versus the Laplace mode, as n grows. The error shrinks roughly like 1/n: it falls by about a factor of 4-5 each time n increases by that factor."
set.seed(42)
res <- t(sapply(c(5, 20, 100, 500), function(n) {
  y <- rpois(n, 3)
  lap <- laplace_poisson(y)
  c(n = n, laplace_log_ml = lap["log_ml"], exact_log_ml = exact_log_ml(y),
    error = lap["log_ml"] - exact_log_ml(y),
    laplace_mode = lap["mode"], posterior_mean = (alpha + sum(y)) / (beta + n))
}))
knitr::kable(res, digits = 4, row.names = FALSE,
             col.names = c("n", "Laplace log P(y)", "exact log P(y)", "error",
                           "Laplace mode", "posterior mean"))
```

The error is a few thousandths at $n=5$ and shrinks to a few ten-thousandths
by $n=500$; the Laplace mode also converges to the posterior mean as the
Gamma posterior becomes more symmetric. Plotting the exact posterior against
its Laplace approximation makes this concrete:

```{r}
#| label: fig-poisson-gamma-density
#| fig-width: 10
#| fig-height: 4.4
#| fig-cap: "Exact Gamma posterior (red) and the Laplace normal approximation (green) for n=5 and n=100. At n=5 the exact posterior is visibly right-skewed and the Gaussian is a rough match; by n=100 the two are nearly indistinguishable, illustrating the O(1/n) convergence in the table above."
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
set.seed(42)
for (n in c(5, 100)) {
  y <- rpois(n, 3); a1 <- alpha + sum(y); b1 <- beta + n
  lap <- laplace_poisson(y)
  curve(dgamma(x, a1, b1), max(0, lap["mode"] - 4 * lap["se"]), lap["mode"] + 4 * lap["se"],
        lwd = 3, col = "firebrick", xlab = expression(lambda), ylab = "density",
        main = sprintf("n = %d", n))
  curve(dnorm(x, lap["mode"], lap["se"]), add = TRUE, lwd = 3, col = "forestgreen")
  if (n == 5) legend("topright", bty = "n", lwd = 3, col = c("firebrick", "forestgreen"),
                      legend = c("exact posterior", "Laplace approximation"))
}
```

### Linear functionals: posterior mean and credible intervals

If $a(\theta) = A^\top\theta + a_0$ is linear, replacing $f$ by $\tilde f$
gives $E_f[a(\theta)] \approx A^\top\theta_0 + a_0$: the posterior mean is
approximated by the posterior mode, and $\text{Var}(A^\top\theta) \approx
A^\top H^{-1}A$ gives Wald-style credible intervals $\theta_{0j} \pm
z_{\alpha/2}\sqrt{(H^{-1})_{jj}}$ — exactly the Bayesian counterpart of an
MLE plus its Wald standard error from the maximum-likelihood chapters.

```{r}
#| label: tbl-poisson-gamma-ci
#| tbl-cap: "Laplace mode/SE-based 95% interval versus the exact Gamma posterior mean/sd-based interval, n = 20."
set.seed(42)
n <- 20; y <- rpois(n, 3); lap <- laplace_poisson(y)
a1 <- alpha + sum(y); b1 <- beta + n
comp <- data.frame(
  quantity = c("Laplace (mode / se)", "exact posterior (mean / sd)"),
  center = c(lap["mode"], a1 / b1),
  spread = c(lap["se"], sqrt(a1) / b1),
  lower95 = c(lap["mode"] - 1.96 * lap["se"], qgamma(0.025, a1, b1)),
  upper95 = c(lap["mode"] + 1.96 * lap["se"], qgamma(0.975, a1, b1))
)
knitr::kable(comp, digits = 4)
```

The two intervals agree closely even at $n=20$: the mild right-skew of the
Gamma posterior visible in @fig-poisson-gamma-density has not yet noticeably
distorted a Wald-style interval built from a symmetric Gaussian.

### BIC as a large-$n$ simplification

With $f(\theta) = L(\theta;D)\pi(\theta)$, the Laplace approximation to the
log marginal likelihood is
$$
\log P(D) \approx \log\big[L(\theta_0;D)\,\pi(\theta_0)\big] + \frac p2\log(2\pi) - \frac12\log|H|.
$$
For large $n$ the Hessian is dominated by the likelihood, $H \approx
n\,I(\theta_0)$ with $I$ the per-observation Fisher information, so $|H|
\approx n^p|I(\theta_0)|$ and
$$
\log P(D) \approx \log L(\theta_0;D) - \frac p2\log n
 + \underbrace{\log\pi(\theta_0) + \frac p2\log(2\pi) - \frac12\log|I(\theta_0)|}_{O(1),\text{ dropped}}.
$$
Multiplying by $-2$ and replacing the posterior mode by the MLE (the two
coincide to $O(1/n)$) gives the **Bayesian information criterion**,
$\text{BIC} = -2\log L(\hat\theta_{\text{MLE}}) + p\log n$ — a model-comparison
score that needs no prior at all, because the dropped $O(1)$ terms do not grow
with $n$ and so wash out of *differences* between models fit to the same
data. For the $n=20$ Poisson example from the credible-interval check above,
$-2\times(\text{Laplace log }P(y))$ and the MLE-based BIC differ by about
`r round(-2*laplace_poisson(y)["log_ml"] - (-2*sum(dpois(y, mean(y), log=TRUE)) + log(n)), 2)` —
a constant offset consistent with the dropped $O(1)$ terms, not something
that grows with $n$.

## When Laplace Works Well, and When It Does Not

**Works well** when $n$ is large relative to $p$ and the posterior is
unimodal and roughly symmetric, as in the $n=100$ panel above. Parameters
should live on an unbounded scale — transform first ($\log\sigma$,
$\text{logit}\,p$) so the matching Gaussian is not truncated at a boundary,
remembering to include the Jacobian of the transform in $f$ (exactly the
$w=\log\sigma^2$ reparametrization used throughout this book since the
multivariate optimization chapter).

**Struggles** when the posterior is skewed (small $n$, a parameter near a
boundary) — the mode is then a poor stand-in for the mean and the Gaussian
tails are wrong, as the $n=5$ panel above already shows mildly; when the
posterior is multimodal, since the approximation only sees the one mode an
optimizer happened to find; and in hierarchical models where $p$ grows with
$n$, so the $O(1/n)$ error does not vanish (nested Laplace approximation,
INLA, handles this by integrating out the high-dimensional part
analytically).

## Comparing Laplace, Quadrature, and Monte Carlo on the Same Problem

The previous chapter (Numerical Quadrature) computed the marginal likelihood of a normal sample with
unknown mean $\mu$ and log-variance $w=\log\sigma^2$,
$$
P(y) = \int_{-\infty}^{\infty}\!\int_{-\infty}^{\infty} L(\mu,w)\,\pi(\mu,w)\,d\mu\,dw,
\qquad \mu\sim N(\mu_0,\sigma_\mu^2),\ w\sim N(w_0,\sigma_w^2),
$$
by a product midpoint rule on a logistic-transformed $(0,1)^2$ grid, checked
against naive Monte Carlo. The Laplace approximation attacks the same
integral by optimization instead of a grid: find the joint mode of
$(\mu,w)$ and its $2\times2$ Hessian, then apply the boxed formula with
$p=2$.

```{r}
#| label: laplace-gaussian-functions
### log likelihood and log prior for the normal/log-variance model of Unit08
log_lik_gauss  <- function(x, mu, w) sum(dnorm(x, mu, exp(w), log = TRUE))
log_prior_gauss <- function(mu, w, mu0, sigma_mu, w0, sigma_w) {
  dnorm(mu, mu0, sigma_mu, log = TRUE) + dnorm(w, w0, sigma_w, log = TRUE)
}
neg_log_post <- function(theta, x, mu0, sigma_mu, w0, sigma_w) {
  -log_lik_gauss(x, theta[1], theta[2]) - log_prior_gauss(theta[1], theta[2], mu0, sigma_mu, w0, sigma_w)
}

### Laplace approximation for this two-parameter posterior
laplace_gaussian <- function(x, mu0, sigma_mu, w0, sigma_w) {
  op <- nlm(neg_log_post, p = c(mean(x), log(sd(x))), hessian = TRUE,
            x = x, mu0 = mu0, sigma_mu = sigma_mu, w0 = w0, sigma_w = sigma_w)
  H <- op$hessian; p <- length(op$estimate)
  list(mode = op$estimate, se = sqrt(diag(solve(H))),
       log_ml = -op$minimum + p / 2 * log(2 * pi) - 0.5 * log(det(H)))
}

### naive Monte Carlo (samples from the prior; see Unit08 for the log-sum-exp helper)
log_sum_exp <- function(lx) { m <- max(lx); m + log(sum(exp(lx - m))) }
log_mar_mc <- function(x, mu0, sigma_mu, w0, sigma_w, iters) {
  mus <- rnorm(iters, mu0, sigma_mu); ws <- rnorm(iters, w0, sigma_w)
  ll <- vapply(seq_len(iters), function(i) log_lik_gauss(x, mus[i], ws[i]), numeric(1))
  log_sum_exp(ll) - log(iters)
}

### product midpoint quadrature on the logistic-transformed square (Unit08's method)
logit_inv     <- function(u) log(u / (1 - u))
log_der_logit <- function(u) -log(u) - log(1 - u)
log_g <- function(u, v, x, mu0, sigma_mu, w0, sigma_w) {
  mu <- logit_inv(u); w <- logit_inv(v)
  log_lik_gauss(x, mu, w) + log_prior_gauss(mu, w, mu0, sigma_mu, w0, sigma_w) +
    log_der_logit(u) + log_der_logit(v)
}
log_mar_mid <- function(x, mu0, sigma_mu, w0, sigma_w, n) {
  h <- 1 / n; grid <- (seq_len(n) - 0.5) * h
  G <- outer(grid, grid, Vectorize(function(u, v) log_g(u, v, x, mu0, sigma_mu, w0, sigma_w)))
  log_sum_exp(G) + 2 * log(h)
}
```

```{r}
#| label: tbl-three-way
#| tbl-cap: "Log marginal likelihood by three independent methods, for a vague prior N(0, 100) on mu and N(0, 5) on w, n = 50 observations per scenario. Quadrature uses a 100x100 grid (unchanged from Unit08); Monte Carlo uses 1e6 prior draws."
set.seed(7)
scenarios <- list(
  c(desc = "well-centred: mean 5",    mean = 5,   sd = 1),
  c(desc = "well-centred: mean -5",   mean = -5,  sd = 1),
  c(desc = "far from origin: mean -50, sd 4",  mean = -50, sd = 4),
  c(desc = "far from origin: mean -50, sd 10", mean = -50, sd = 10)
)
three_way <- t(sapply(scenarios, function(s) {
  x <- rnorm(50, mean = as.numeric(s["mean"]), sd = as.numeric(s["sd"]))
  lap <- laplace_gaussian(x, 0, 100, 0, 5)
  c(Laplace = lap$log_ml,
    Monte_Carlo = log_mar_mc(x, 0, 100, 0, 5, 1e6),
    Quadrature = log_mar_mid(x, 0, 100, 0, 5, 100))
}))
knitr::kable(cbind(scenario = sapply(scenarios, `[`, "desc"), round(three_way, 3)),
             row.names = FALSE, col.names = c("scenario", "Laplace", "Monte Carlo", "Quadrature (n=100 grid)"))
```

The pattern is striking. For data centred near the origin, all three methods
agree to within a unit or two on the log scale. But once the data are far
from the origin (mean $-50$), **Laplace and Monte Carlo still agree closely
with each other, while the fixed-grid quadrature is off by well over 100
log-units** — the same boundary-squeeze failure diagnosed at the end of
Unit08: the logistic transform used there is centred at $0$, so a posterior
sitting at $\mu\approx-50$ gets squeezed into a sliver of the unit square that
a $100\times100$ grid cannot resolve. Laplace approximation has no such
failure mode here, and the reason is structural, not a matter of tuning:
**it has no fixed grid to be squeezed against**. It optimizes to find
$\theta_0$ wherever the mode actually is, then builds its approximation
*around that point* — so it automatically "recentres" itself for every
dataset, which is exactly the manual fix (recentre the transform on a
location estimate) that Unit08 proposed for quadrature. This is Laplace's
core practical advantage over quadrature: robustness to *where* the posterior
happens to sit, at the cost of assuming it looks Gaussian once you get there.

### Explore it interactively {#sec-laplace-app}

The app below lets you move the true mean of the simulated data and compare
the three log marginal likelihood estimates live, so you can watch
quadrature's error grow while Laplace and Monte Carlo continue to track each
other. This app runs live only in the HTML edition of this book; readers of
other formats can follow the link below it.

::: {.content-visible when-format="html"}

```{shinylive-r}
#| label: laplace-vs-quadrature-app
#| standalone: true
#| viewerHeight: 800

library(shiny)

log_lik_gauss  <- function(x, mu, w) sum(dnorm(x, mu, exp(w), log = TRUE))
log_prior_gauss <- function(mu, w, mu0, sigma_mu, w0, sigma_w) {
  dnorm(mu, mu0, sigma_mu, log = TRUE) + dnorm(w, w0, sigma_w, log = TRUE)
}
neg_log_post <- function(theta, x, mu0, sigma_mu, w0, sigma_w) {
  -log_lik_gauss(x, theta[1], theta[2]) - log_prior_gauss(theta[1], theta[2], mu0, sigma_mu, w0, sigma_w)
}
laplace_gaussian <- function(x, mu0, sigma_mu, w0, sigma_w) {
  op <- nlm(neg_log_post, p = c(mean(x), log(sd(x))), hessian = TRUE,
            x = x, mu0 = mu0, sigma_mu = sigma_mu, w0 = w0, sigma_w = sigma_w)
  H <- op$hessian; p <- length(op$estimate)
  list(mode = op$estimate, log_ml = -op$minimum + p / 2 * log(2 * pi) - 0.5 * log(det(H)))
}
log_sum_exp <- function(lx) { m <- max(lx); m + log(sum(exp(lx - m))) }
log_mar_mc <- function(x, mu0, sigma_mu, w0, sigma_w, iters) {
  mus <- rnorm(iters, mu0, sigma_mu); ws <- rnorm(iters, w0, sigma_w)
  ll <- vapply(seq_len(iters), function(i) log_lik_gauss(x, mus[i], ws[i]), numeric(1))
  log_sum_exp(ll) - log(iters)
}
logit_inv     <- function(u) log(u / (1 - u))
log_der_logit <- function(u) -log(u) - log(1 - u)
log_g <- function(u, v, x, mu0, sigma_mu, w0, sigma_w) {
  mu <- logit_inv(u); w <- logit_inv(v)
  log_lik_gauss(x, mu, w) + log_prior_gauss(mu, w, mu0, sigma_mu, w0, sigma_w) +
    log_der_logit(u) + log_der_logit(v)
}
log_mar_mid <- function(x, mu0, sigma_mu, w0, sigma_w, n) {
  h <- 1 / n; grid <- (seq_len(n) - 0.5) * h
  G <- outer(grid, grid, Vectorize(function(u, v) log_g(u, v, x, mu0, sigma_mu, w0, sigma_w)))
  log_sum_exp(G) + 2 * log(h)
}

ui <- fluidPage(
  titlePanel(NULL),
  sidebarLayout(
    sidebarPanel(
      width = 4,
      sliderInput("true_mean", "true mean of the simulated data", min = -60, max = 10, value = 5, step = 2),
      sliderInput("n_grid", "quadrature grid points per axis", min = 10, max = 100, value = 100, step = 10),
      helpText("Prior: N(0, 100) on mu, N(0, 5) on w = log(sd). As the true mean ",
               "moves away from 0, the fixed logistic-transform grid squeezes the ",
               "posterior into a corner and the quadrature estimate breaks away, ",
               "while Laplace (which re-optimizes for every dataset) tracks Monte Carlo."),
      tableOutput("tab")
    ),
    mainPanel(width = 8, plotOutput("bars", height = "420px"))
  )
)

server <- function(input, output, session) {
  dat <- reactive({ set.seed(1); rnorm(50, input$true_mean, 2) })
  vals <- reactive({
    x <- dat()
    c(Laplace = laplace_gaussian(x, 0, 100, 0, 5)$log_ml,
      Monte_Carlo = { set.seed(2); log_mar_mc(x, 0, 100, 0, 5, 2e5) },
      Quadrature = log_mar_mid(x, 0, 100, 0, 5, input$n_grid))
  })
  output$bars <- renderPlot({
    v <- vals()
    par(mar = c(4, 4, 2, 1))
    bp <- barplot(v, col = c("forestgreen", "steelblue", "firebrick"),
                  ylab = "estimated log P(y)", main = "three methods, same data")
    text(bp, v, sprintf("%.1f", v), pos = ifelse(v >= 0, 3, 1))
  })
  output$tab <- renderTable({
    v <- vals()
    data.frame(method = names(v), `log P(y)` = as.numeric(v), check.names = FALSE)
  }, digits = 2, rownames = FALSE)
}

shinyApp(ui, server)
```

:::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit09_laplace.html#sec-laplace-app>.
:::

::: {.callout-note}
### Reproducing this figure
As in earlier chapters, this app uses the
[shinylive](https://github.com/quarto-ext/shinylive) extension
(`quarto add quarto-ext/shinylive`, with `filters: [shinylive]` in the
document YAML), compiled to WebAssembly so it runs in the browser with no
Shiny server. Only base R and `shiny` are available inside the app.
:::

## Summary

- The Laplace approximation expands $h=-\log f$ to second order at its
  minimum $\theta_0$ and integrates the resulting Gaussian exactly:
  $\theta\mid\text{data} \approx N_p(\theta_0, H^{-1})$ and
  $\int f(\theta)\,d\theta \approx f(\theta_0)(2\pi)^{p/2}|H|^{-1/2}$.
- It needs only an optimizer and a Hessian — the same machinery already used
  to find the MLE — and its cost does not grow with dimension the way a
  quadrature grid's does.
- BIC is the large-$n$ simplification of the Laplace marginal likelihood,
  dropping $O(1)$ terms that do not depend on the sample size.
- On the running Gaussian marginal-likelihood example, Laplace matched Monte
  Carlo everywhere quadrature was tested, and remained accurate exactly where
  quadrature's fixed grid broke down — because Laplace re-optimizes for every
  dataset instead of committing to one transform in advance.
- It is deterministic, fast, and accurate to $O(1/n)$, but blind to
  skewness and multimodality; the Monte Carlo methods of the next chapters
  remove these limitations at higher computational cost.
