---
title: "Optimization for Maximum Likelihood Estimation"
format:
  html:
    toc: true
    tbl-colwidths: true
filters:
  - shinylive
editor_options:
  chunk_output_type: console
code-fold: true
---

This chapter develops maximum likelihood estimation and the numerical methods
used to find it, first for a scalar parameter and then for a parameter vector.
The univariate case, where the log-likelihood can be plotted and a root can be
bracketed, motivates the two ideas that carry over unchanged to higher
dimensions: replacing the log-likelihood locally by a quadratic and solving for
its maximum (Newton–Raphson), and using the curvature of that quadratic to
report a standard error. The multivariate case, where neither plotting nor
bracketing is available, shows what survives of that idea and what has to be
added — safeguards against divergence, and a wider menu of derivative-free and
gradient-only methods for when the Hessian is unavailable or too expensive.

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false
library(latex2exp)   # TeX() renders LaTeX in plot titles, axis labels and legends
set.seed(812)
```

## Maximum Likelihood: Theory

### Likelihood, Score, and Hessian

A model specifies the density or probability of the data $D$ given a parameter $\theta$, written $f(D \mid \theta)$. As a function of $D$ with $\theta$ fixed it is the sampling distribution; as a function of $\theta$ with $D$ observed it is the likelihood. For independent observations $x_1, \ldots, x_n$,
$$L(\theta) = \prod_{i=1}^{n} f(x_i \mid \theta), \qquad \ell(\theta) = \log L(\theta) = \sum_{i=1}^{n} \log f(x_i \mid \theta).$$

We always work with $\ell$: the product underflows to zero for even moderate $n$, and sums are easier to differentiate. The maximum likelihood estimator maximizes $\ell$, and where $\ell$ is smooth with an interior maximum, $\hat\theta$ solves the **score equation**
$$S(\theta) = \ell'(\theta) = 0, \qquad \ell''(\hat\theta) < 0 .$$

Three quantities recur throughout the chapter:

- the **score** $S(\theta) = \ell'(\theta)$, which locates the maximum;
- the **observed information** $J(\theta) = -\ell''(\theta)$, the curvature of the log-likelihood at $\theta$;
- the **expected information** $n I(\theta) = -E_D\left[\ell''(\theta; D)\right]$, its expectation under the model.

The second derivative therefore plays a double role. It determines how a Newton-type algorithm steps toward the maximum, and it determines how precisely the data pin the parameter down. An optimizer that computes the curvature has computed the standard error at the same time.

### Closed Form: the Normal Mean

Let $X_1, \ldots, X_n$ be iid $N(\mu, \sigma_0^{2})$ with $\sigma_0^{2}$ known. Using the identity $\sum_i (x_i - \mu)^2 = \sum_i (x_i - \bar x)^2 + n(\bar x - \mu)^2$,
$$\ell(\mu) = -\frac{n}{2}\log\left(2\pi\sigma_0^{2}\right) - \frac{1}{2\sigma_0^{2}}\sum_{i=1}^{n}(x_i - \mu)^{2} = C - \frac{n}{2\sigma_0^{2}}(\mu - \bar x)^{2} .$$

Differentiating,
$$S(\mu) = \frac{n}{\sigma_0^{2}}\left(\bar x - \mu\right), \qquad \ell''(\mu) = -\frac{n}{\sigma_0^{2}}, \qquad \hat\mu = \bar x .$$

Every feature of the general theory is visible here in exact form. The log-likelihood is a downward parabola, so it has a unique maximum. The score is linear, so the score equation is solved in one line. The curvature $J(\mu) = n/\sigma_0^{2}$ does not depend on $\mu$ or on the data, so observed and expected information coincide, $J = nI$, and both grow linearly with $n$: doubling the sample doubles the sharpness of the peak and halves the variance of $\hat\mu$.

```{r}
#| label: normal-loglik-score-curvature
#| fig-width: 10
#| fig-height: 4
x20 <- rnorm(20, mean = 1); x50 <- rnorm(50, mean = 1)
mu <- seq(0, 2, length.out = 200)

ll_norm <- function(mu, x) vapply(mu, function(m) sum(dnorm(x, m, log = TRUE)), numeric(1))

par(mfrow = c(1, 3), mar = c(4, 4.5, 2.5, 1))

matplot(mu, cbind(ll_norm(mu, x20) - max(ll_norm(mu, x20)),
                  ll_norm(mu, x50) - max(ll_norm(mu, x50))),
        type = "l", lty = 1, lwd = 2, col = c(1, 2),
        xlab = TeX(r"($\mu$)"), ylab = TeX(r"($\ell(\mu) - \ell(\hat{\mu})$)"),
        main = TeX(r"(log-likelihood)"))
legend("bottom", bty = "n", lty = 1, lwd = 2, col = c(1, 2),
       legend = TeX(c(r"($n = 20$)", r"($n = 50$)")))

matplot(mu, cbind(20 * (mean(x20) - mu), 50 * (mean(x50) - mu)),
        type = "l", lty = 1, lwd = 2, col = c(1, 2),
        xlab = TeX(r"($\mu$)"), ylab = TeX(r"($S(\mu)$)"), main = TeX(r"(score)"))
abline(h = 0, lty = 3)

matplot(mu, cbind(rep(20, 200), rep(50, 200)), type = "l", lty = 1, lwd = 2, col = c(1, 2),
        ylim = c(0, 60), xlab = TeX(r"($\mu$)"), ylab = TeX(r"($-\ell''(\mu)$)"),
        main = TeX(r"(curvature)"))
```

The score crosses zero exactly where $\ell$ peaks, and the larger sample gives a more peaked log-likelihood, a steeper score, and a larger curvature.

### Consistency, Asymptotic Normality, and Fisher Information

Under regularity conditions — the support of $f$ not depending on $\theta$, the true value interior to the parameter space, and two derivatives passing under the integral sign — the MLE has three properties as $n \to \infty$, with $\theta_*$ the true parameter.

**Consistency.** $\hat\theta \to \theta_*$ in probability. The reason is that $\ell(\theta)/n$ converges to $E_{\theta_*}[\log f(X \mid \theta)]$, which by the information inequality is maximized at $\theta = \theta_*$: the log-likelihood surface converges to one whose peak is at the truth.

**Asymptotic normality.**
$$\hat\theta \;\dot\sim\; N\left(\theta_*,\ \left[n I(\theta_*)\right]^{-1}\right), \qquad I(\theta) = -E_X\left[\frac{\partial^{2}}{\partial\theta^{2}}\log f(X \mid \theta)\right] = E_X\left[\left(\frac{\partial}{\partial\theta}\log f(X\mid\theta)\right)^{2}\right].$$

A sketch: expand $S(\hat\theta) = 0$ about $\theta_*$ to get $\hat\theta - \theta_* \approx S(\theta_*)/\{-\ell''(\theta_*)\}$. The numerator is a sum of iid mean-zero terms, hence asymptotically normal with variance $nI(\theta_*)$; the denominator converges to $nI(\theta_*)$ by the law of large numbers. Dividing gives variance $[nI(\theta_*)]^{-1}$.

**Asymptotic efficiency.** No consistent estimator has smaller asymptotic variance — the Cramér–Rao bound is attained.

Since $\theta_*$ is unknown, the variance is estimated by substituting $\hat\theta$. Two versions are available: the expected information $nI(\hat\theta)$, which requires the expectation analytically, and the observed information $J(\hat\theta) = -\ell''(\hat\theta)$, which requires only the curvature that a Newton step already computes. By the law of large numbers $J(\hat\theta)/n \approx I(\hat\theta)$, and both give
$$\widehat{\mathrm{SE}}(\hat\theta) = J(\hat\theta)^{-1/2} \quad\text{or}\quad \left[nI(\hat\theta)\right]^{-1/2}, \qquad \hat\theta \pm 1.96\,\widehat{\mathrm{SE}}(\hat\theta).$$

These are asymptotic statements. A flat log-likelihood — small curvature — gives a large standard error, and a log-likelihood that is far from quadratic gives a standard error that should not be trusted at all; the last section of this chapter shows a case of exactly that.

### Demonstration: the Cauchy Location Model

The Cauchy density $f(x \mid \theta) = 1/\{\pi[1 + (x - \theta)^2]\}$ has no closed-form MLE:
$$\ell(\theta) = -n\log\pi - \sum_{i=1}^{n}\log\left(1 + (x_i - \theta)^{2}\right), \qquad
S(\theta) = \sum_{i=1}^{n}\frac{2(x_i - \theta)}{1 + (x_i - \theta)^{2}} .$$

The score equation is a polynomial equation of degree $2n - 1$. Plotting $\ell$ for samples of increasing size shows the peak sharpening around the truth, which is consistency and growing information in one picture.

```{r}
#| label: cauchy-loglik-by-n
#| fig-width: 9
#| fig-height: 5
log_like_cauchy <- function(theta, x) sum(dcauchy(x, theta, log = TRUE))

thetas <- seq(15, 25, length.out = 200)
n_set <- c(50, 100, 200)

ll_cauchy <- sapply(n_set, function(n) {
  d <- rcauchy(n, location = 20)
  sapply(thetas, log_like_cauchy, x = d)
})
ll_cauchy <- sweep(ll_cauchy, 2, apply(ll_cauchy, 2, max))   # align peaks at 0

par(mar = c(4, 4.5, 2.5, 1))
matplot(thetas, ll_cauchy, type = "l", lty = 1, lwd = 2, col = c(1, 2, 4), ylim = c(-20, 0),
        xlab = TeX(r"($\theta$)"), ylab = TeX(r"($\ell(\theta) - \ell(\hat{\theta})$)"),
        main = TeX(r"(Cauchy log-likelihood, $\theta_* = 20$)"))
abline(v = 20, lty = 3)
legend("bottomright", bty = "n", lty = 1, lwd = 2, col = c(1, 2, 4),
       legend = TeX(sprintf(r"($n = %d$)", n_set)))
```

The variability of the MLE across repeated samples is the sampling distribution the theory describes. Repeating the experiment 50 times at two sample sizes shows the curves clustering more tightly, and their maxima concentrating, as $n$ grows.

```{r}
#| label: cauchy-loglik-variability
#| fig-width: 10
#| fig-height: 4.5
#| cache: true
loglik_curves <- function(n, nrep = 50)
  replicate(nrep, {
    d <- rcauchy(n, location = 20)
    lc <- sapply(thetas, log_like_cauchy, x = d)
    lc - max(lc)
  })

curves_small <- loglik_curves(20)
curves_large <- loglik_curves(200)

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
for (obj in list(list(m = curves_small, n = 20), list(m = curves_large, n = 200)))
  matplot(thetas, obj$m, type = "l", lty = 1, col = adjustcolor(1, 0.35), ylim = c(-10, 0),
          xlab = TeX(r"($\theta$)"), ylab = TeX(r"($\ell(\theta) - \ell(\hat{\theta})$)"),
          main = TeX(sprintf(r"(sample size $n = %d$)", obj$n)))

### sampling distribution of the MLE at the two sample sizes
mle_at <- function(m) thetas[apply(m, 2, which.max)]
rbind(`n = 20`  = c(mean = mean(mle_at(curves_small)), sd = sd(mle_at(curves_small))),
      `n = 200` = c(mean = mean(mle_at(curves_large)), sd = sd(mle_at(curves_large))))
```

The standard deviation falls by roughly the factor $\sqrt{10}$ predicted by $[nI(\theta_*)]^{-1/2}$. Two features of these curves also warn us about computation: for small $n$ the log-likelihood has several local maxima, and its shape away from the peak is far from quadratic. Numerical optimization is required, and it must be used with care.

## Univariate Optimization Algorithms

### Three Equivalent Formulations

The same task can be posed three ways, and the algorithms below attack different versions of it:

- maximize $\ell(\theta)$;
- minimize $f(\theta) = -\ell(\theta)$;
- find a root of $S(\theta) = \ell'(\theta)$.

Root-finders are usually faster because they use derivative information, but they locate any stationary point, including minima and saddle points. Direct optimizers need only function values and always move toward a maximum, but converge more slowly. Which to prefer depends on whether derivatives are available and whether the function is smooth.

Each method below is illustrated with a schematic showing its first few iterations. To keep the pictures comparable they all use the same small problem: the per-observation log-likelihood of the zero-truncated Poisson model of the next section, with $\bar x$ fixed at 1.8. The root of its score is marked in every figure.

```{r}
#| label: demo-problem
xbar_d <- 1.8                                             # fixed sample mean

ll_d <- function(l) xbar_d * log(l) - l - log(1 - exp(-l))          # log-likelihood
S_d  <- function(l) xbar_d / l - 1 / (1 - exp(-l))                  # score
dS_d <- function(l) -xbar_d / l^2 + exp(-l) / (1 - exp(-l))^2       # its derivative
I_d  <- function(l) 1 / (l * (1 - exp(-l))) - exp(-l) / (1 - exp(-l))^2   # information
g_d  <- function(l) xbar_d * (1 - exp(-l))                          # fixed-point map

lstar <- uniroot(S_d, c(0.2, 5))$root
c(root = lstar, contraction_factor = xbar_d * exp(-lstar))
```

### Fixed-Point (Simple) Iteration

Rearrange the score equation algebraically into the form $\theta = g(\theta)$ and iterate $\theta_{k+1} = g(\theta_k)$. If $g$ maps an interval into itself and $|g'(\theta^{*})| < 1$ at the fixed point, the mean value theorem gives
$$\left|\theta_{k+1} - \theta^{*}\right| = \left|g(\theta_k) - g(\theta^{*})\right| \le \left|g'(\xi)\right| \left|\theta_k - \theta^{*}\right| ,$$
so the error contracts by roughly the constant factor $|g'(\theta^{*})|$ at each step. This is **linear convergence**: a fixed number of digits is gained per iteration, and the closer $|g'|$ is to 1 the slower the progress.

```{r}
#| label: fixed-point
fixed_point <- function(g, x0, tol = 1e-10, maxit = 500) {
  for (k in 1:maxit) {
    x1 <- g(x0)
    if (abs(x1 - x0) < tol) return(c(root = x1, iter = k))
    x0 <- x1
  }
  stop("not converged")
}
```

The mechanism is easiest to see as a **cobweb**: from $\lambda_k$ on the diagonal, move vertically to the curve $g$ to obtain $\lambda_{k+1}$, then horizontally back to the diagonal to start the next step. The iterates walk toward the intersection of $g$ with the $45^{\circ}$ line, and the shallower $g$ is at that crossing — the smaller $|g'|$ — the larger each stride toward it.

```{r}
#| label: scheme-fixed-point
#| fig-width: 8
#| fig-height: 5
lam <- seq(0.2, 3, length.out = 300)

par(mar = c(4, 4.5, 2.5, 1))
plot(lam, g_d(lam), type = "l", lwd = 2, xlim = c(0.2, 3), ylim = c(0.2, 3),
     xlab = TeX(r"($\lambda_k$)"), ylab = TeX(r"($g(\lambda_k) = \lambda_{k+1}$)"),
     main = TeX(r"(fixed-point iteration: the cobweb)"))
abline(0, 1, lty = 2)

x0 <- 2.8
for (k in 1:5) {
  x1 <- g_d(x0)
  segments(x0, x0, x0, x1, col = 2, lwd = 2)          # up or down to the curve
  segments(x0, x1, x1, x1, col = 2, lwd = 2)          # across to the diagonal
  points(x0, x1, pch = 19, col = 2, cex = 0.7)
  if (k <= 3) text(x0, 0.25, TeX(sprintf(r"($\lambda_%d$)", k - 1)), col = 2, cex = 0.9)
  x0 <- x1
}
points(lstar, lstar, pch = 19, cex = 1.2)
text(lstar, lstar, TeX(r"($\hat{\lambda}$)"), pos = 4)
```

The method needs no derivatives, but the rearrangement must be found by hand, and different rearrangements of the same equation can converge at different speeds or diverge outright. Geometrically, a rearrangement with $|g'| > 1$ at the crossing produces a cobweb that spirals outward instead of inward.

### Bisection

If $S$ is continuous and $S(a)$ and $S(b)$ have opposite signs, a root lies in $[a, b]$. Evaluate $S$ at the midpoint and retain whichever half still changes sign.

```{r}
#| label: bisection
bisection <- function(S, a, b, tol = 1e-10) {
  if (S(a) * S(b) > 0) stop("endpoints do not bracket a root")
  k <- 0
  while (b - a > tol) {
    m <- (a + b) / 2
    if (S(a) * S(m) <= 0) b <- m else a <- m
    k <- k + 1
  }
  c(root = (a + b) / 2, iter = k)
}
```

The schematic below draws the bracket at each iteration as a bar beneath the score curve. The dot is the midpoint, and the dotted line carries it up to the curve, where its sign decides which half survives.

```{r}
#| label: scheme-bisection
#| fig-width: 8
#| fig-height: 5
par(mar = c(4, 4.5, 2.5, 1))
plot(lam, S_d(lam), type = "l", lwd = 2, ylim = c(-1.6, 3.5),
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($S(\lambda)$)"),
     main = TeX(r"(bisection: each step halves the bracket)"))
abline(h = 0, lty = 3)
points(lstar, 0, pch = 19, cex = 1.2)

a <- 0.3; b <- 3
ylev <- seq(-0.5, -1.4, length.out = 4)
for (k in 1:4) {
  m <- (a + b) / 2
  segments(a, ylev[k], b, ylev[k], col = 4, lwd = 3)
  points(c(a, b), rep(ylev[k], 2), pch = "|", col = 4)
  segments(m, ylev[k], m, S_d(m), lty = 3, col = 2)
  points(m, ylev[k], pch = 19, col = 2, cex = 0.8)
  text(0.25, ylev[k], k, cex = 0.8, col = 4)
  if (S_d(a) * S_d(m) <= 0) b <- m else a <- m
}
```

The bracket halves at every step, so after $k$ iterations the error is at most $(b - a)/2^{k}$ and the iteration count is known in advance: about $\log_2\{(b-a)/\text{tol}\}$, regardless of the shape of $S$. Bisection cannot diverge, which makes it the natural fallback when a faster method misbehaves. R's `uniroot` uses Brent's method, which retains the bracketing guarantee while interpolating for speed.

### Newton–Raphson

#### Derivation

Approximate the score by its tangent line at $\theta_k$ and solve for where the tangent crosses zero:
$$S(\theta) \approx S(\theta_k) + S'(\theta_k)(\theta - \theta_k) = 0
\implies \theta_{k+1} = \theta_k - \frac{S(\theta_k)}{S'(\theta_k)} = \theta_k - \frac{\ell'(\theta_k)}{\ell''(\theta_k)} .$$

Equivalently, replace $\ell$ by its second-order Taylor expansion at $\theta_k$ and jump to the maximum of that parabola. For an exactly quadratic $\ell(\theta) = -a(\theta - u)^2 + C$ with $a > 0$ we have $\ell' = -2a(\theta - u)$ and $\ell'' = -2a$, so
$$\theta_{k+1} = \theta_k - \frac{-2a(\theta_k - u)}{-2a} = u$$
in one step from any starting value. The normal-mean example of the previous section is exactly of this form: Newton converges immediately.

The left panel below traces three steps. At each iterate, the tangent to the score is followed down to the horizontal axis, and where it lands is the next iterate. As the curve straightens near the root, the tangent becomes an excellent approximation and the steps land almost exactly on the answer.

The right panel contrasts one Newton step with one step of the method of scoring from the same point. Newton uses the actual slope $S'(\lambda_k)$; scoring uses $-nI(\lambda_k)$, the slope averaged over hypothetical datasets. The two lines differ away from the root and coincide as $\lambda_k$ approaches it.

```{r}
#| label: scheme-newton
#| fig-width: 10
#| fig-height: 4.8
par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))

### three Newton steps
plot(lam, S_d(lam), type = "l", lwd = 2, ylim = c(-1, 3.5),
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($S(\lambda)$)"),
     main = TeX(r"(Newton-Raphson: follow the tangent)"))
abline(h = 0, lty = 3)
points(lstar, 0, pch = 19, cex = 1.2)

x0 <- 0.45
for (k in 1:3) {
  x1 <- x0 - S_d(x0) / dS_d(x0)
  segments(x0, S_d(x0), x1, 0, col = 2, lwd = 2)
  segments(x1, 0, x1, S_d(x1), lty = 3, col = "grey40")
  points(c(x0, x1), c(S_d(x0), 0), pch = 19, col = 2, cex = 0.7)
  text(x0, S_d(x0), TeX(sprintf(r"($\lambda_%d$)", k - 1)), pos = 3, col = 2, cex = 0.9)
  x0 <- x1
}

## Newton step versus scoring step from the same point
plot(lam, S_d(lam), type = "l", lwd = 2, ylim = c(-1, 3.5),
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($S(\lambda)$)"),
     main = TeX(r"(observed versus expected curvature)"))
abline(h = 0, lty = 3)
points(lstar, 0, pch = 19, cex = 1.2)

x0 <- 0.45
x_nr <- x0 - S_d(x0) / dS_d(x0)          # slope  S'(lambda)
x_fs <- x0 + S_d(x0) / I_d(x0)           # slope -nI(lambda)
segments(x0, S_d(x0), x_nr, 0, col = 2, lwd = 2)
segments(x0, S_d(x0), x_fs, 0, col = 4, lwd = 2, lty = 5)
points(c(x0, x_nr, x_fs), c(S_d(x0), 0, 0), pch = 19, cex = 0.7, col = c(1, 2, 4))
legend("topright", bty = "n", lwd = 2, lty = c(1, 5), col = c(2, 4),
       legend = TeX(c(r"(Newton: slope $S'(\lambda_k)$)", r"(scoring: slope $-nI(\lambda_k)$)")))
```

### Advantage: quadratic convergence

Let $e_k = \theta_k - \hat\theta$ and expand $S(\hat\theta) = 0$ about $\theta_k$:
$$0 = S(\theta_k) - S'(\theta_k)e_k + \tfrac12 S''(\xi)e_k^{2}
\implies e_{k+1} = e_k - \frac{S(\theta_k)}{S'(\theta_k)} = \frac{S''(\xi)}{2S'(\theta_k)}\,e_k^{2} .$$

The error is squared at each step, so the number of correct digits roughly doubles. Two conditions appear in the constant: $S'(\hat\theta) \ne 0$, meaning the maximum is a genuine peak and not a flat spot, and $S''$ bounded near the root. Asymptotic normality says that log-likelihoods are approximately quadratic near the MLE, which is why Newton is so effective there.

```{r}
#| label: newton
newton <- function(S, dS, x0, tol = 1e-10, maxit = 100, trace = FALSE) {
  for (k in 1:maxit) {
    x1 <- x0 - S(x0) / dS(x0)
    if (trace) cat(sprintf("iter %2d: %.10f\n", k, x1))
    if (!is.finite(x1)) stop("diverged")
    if (abs(x1 - x0) < tol) return(c(root = x1, iter = k))
    x0 <- x1
  }
  stop("not converged")
}
```

### Problem: divergence

Newton uses only local information, so the quadratic model can be badly wrong away from the maximum. Three failures occur:

- **wrong curvature** — where $\ell'' > 0$ the model is a bowl, and the step heads toward a minimum or off to infinity;
- **overshooting** — where the score is nearly flat, $S'(\theta_k) \approx 0$ produces an enormous step, possibly outside the parameter space;
- **multiple roots** — the iteration converges to whichever stationary point it happens to reach, which need not be the global maximum.

The single-observation Cauchy likelihood is the standard illustration. With one observation at zero, $f(\theta) = -\ell(\theta) = \log(1 + \theta^{2})$ and $s(\theta) = f'(\theta) = 2\theta/(1 + \theta^{2})$, whose only root is $\theta = 0$. The score rises, turns over at $\theta = 1$, and decays to zero in the tails — exactly the flat region that defeats Newton.

```{r}
#| label: cauchy-single-obs
#| fig-width: 10
#| fig-height: 4.5
lik <- function(t) log(1 + t^2)                    # negative log-likelihood
s   <- function(t) 2 * t / (1 + t^2)               # its derivative
ds  <- function(t) 2 * (1 - t^2) / (1 + t^2)^2

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
curve(lik, -5, 5, n = 400, lwd = 2, xlab = TeX(r"($\theta$)"), ylab = TeX(r"($-\ell(\theta)$)"),
      main = TeX(r"(negative log-likelihood)"))
curve(s, -5, 5, n = 400, lwd = 2, xlab = TeX(r"($\theta$)"), ylab = TeX(r"($s(\theta)$)"),
      main = TeX(r"(derivative, with the flat tails)"))
abline(h = 0, lty = 3); abline(v = c(-1, 1), col = 2, lty = 2)
```

Tracing the iterates from three starting values shows the boundary at $|\theta_0| = 1$, where the derivative $s$ turns over:

```{r}
#| label: newton-divergence-trace
newton_trace <- function(s, ds, x0, iters = 8) {
  path <- numeric(iters + 1); path[1] <- x0
  for (k in 1:iters) {
    x0 <- x0 - s(x0) / ds(x0)
    path[k + 1] <- if (is.finite(x0)) x0 else NA
  }
  path
}

starts <- c(0.5, 0.8, 1.2)
traces <- vapply(starts, function(v) newton_trace(s, ds, v), numeric(9))
colnames(traces) <- sprintf("start = %.1f", starts)
rownames(traces) <- paste("iter", 0:8)
signif(traces, 5)
```

From $\theta_0 = 0.5$, inside the region where the quadratic model is adequate, Newton converges quadratically to zero. From $\theta_0 = 0.8$ the first step already overshoots into the tail, and from $\theta_0 = 1.2$ — beyond the inflection point of $s$ — the iterates run away, each step landing further out where the derivative is flatter still.

```{r}
#| label: newton-animation
#| eval: false
## Interactive versions of the same three trajectories (requires the animation
## package and a graphics device that can write animations)
library(animation)
ani.options(interval = 1, nmax = 50)
newton.method(FUN = s, init = 0.5, rg = c(-1.5, 1.5))
newton.method(FUN = s, init = 0.8, rg = c(-8, 1.5))
newton.method(FUN = s, init = 1.2, rg = c(-2, 20))
```

### Safeguards

Three remedies cover most cases. Start from a robust estimate — the sample median for a location model. Apply **step halving**: accept the full Newton step only if it improves the log-likelihood, otherwise halve it until it does. Or replace the observed curvature by the expected information, giving the **method of scoring**,
$$\theta_{k+1} = \theta_k + \frac{S(\theta_k)}{n I(\theta_k)} ,$$
in which the divisor is positive by construction, so every step ascends. Near the MLE $nI(\hat\theta) \approx -\ell''(\hat\theta)$, so scoring inherits Newton's speed while being far more stable away from the optimum. Its cost is that the expectation must be available in closed form.

```{r}
#| label: safe-newton
safe_newton <- function(ll, S, dS, x0, lower = -Inf, upper = Inf,
                        tol = 1e-10, maxit = 100, max_halve = 40) {
  f0 <- ll(x0)
  for (k in 1:maxit) {
    step <- -S(x0) / dS(x0)
    if (dS(x0) > 0) step <- sign(S(x0))          # wrong curvature: just move uphill
    for (h in 0:max_halve) {                     # halve until feasible and improving
      x1 <- x0 + step / 2^h
      if (x1 > lower && x1 < upper && is.finite(ll(x1)) && ll(x1) >= f0) break
    }
    if (abs(x1 - x0) < tol) return(c(root = x1, iter = k, halvings = h))
    x0 <- x1; f0 <- ll(x1)
  }
  stop("not converged")
}
```

### Golden Section Search

Golden section minimizes $f = -\ell$ directly, using **no derivative at all**, which makes it the method of choice when $\ell$ is not differentiable or its derivative is unreliable. Keep a bracket $[x_l, x_u]$ containing the minimum of a unimodal $f$, with two interior points placed at the golden ratio $g = (3 - \sqrt5)/2 \approx 0.382$ of the way in from each end. Comparing the two interior values discards one end, and the ratio is chosen precisely so that the surviving interior point is correctly positioned for the next iteration: one new evaluation per step, with the bracket shrinking by the factor $0.618$.

```{r}
#| label: golden
golden <- function(f, brack.int, eps = 1e-4, ...) {
  g <- (3 - sqrt(5)) / 2
  xl <- min(brack.int); xu <- max(brack.int)
  tmp <- g * (xu - xl)
  xmu <- xu - tmp; xml <- xl + tmp
  fl <- f(xml, ...); fu <- f(xmu, ...)
  while (abs(xu - xl) > (1e-5 + abs(xl)) * eps) {
    if (fl < fu) {
      xu <- xmu; xmu <- xml; fu <- fl
      xml <- xl + g * (xu - xl); fl <- f(xml, ...)
    } else {
      xl <- xml; xml <- xmu; fl <- fu
      xmu <- xu - g * (xu - xl); fu <- f(xmu, ...)
    }
  }
  if (fl < fu) xml else xmu
}
```

The schematic makes the reuse visible. Each bar is the current bracket with its two interior points; the point that survives into the next iteration is drawn in blue and sits at exactly the right position there, so only the red point has to be evaluated afresh.

```{r}
#| label: scheme-golden
#| fig-width: 8
#| fig-height: 5
f_d <- function(l) -ll_d(l)
gr <- (3 - sqrt(5)) / 2

par(mar = c(4, 4.5, 2.5, 1))
plot(lam, f_d(lam), type = "l", lwd = 2, ylim = c(0.2, 1.25),
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($-\ell(\lambda)$)"),
     main = TeX(r"(golden section: one new evaluation per step)"))
points(lstar, f_d(lstar), pch = 19, cex = 1.2)

xl <- 0.3; xu <- 3
ylev <- seq(0.44, 0.24, length.out = 3)
for (k in 1:3) {
  xml <- xl + gr * (xu - xl); xmu <- xu - gr * (xu - xl)
  segments(xl, ylev[k], xu, ylev[k], col = 4, lwd = 3)
  points(c(xl, xu), rep(ylev[k], 2), pch = "|", col = 4)
  keep_lower <- f_d(xml) < f_d(xmu)
  points(xml, ylev[k], pch = 19, cex = 0.9, col = if (keep_lower) 4 else 2)
  points(xmu, ylev[k], pch = 19, cex = 0.9, col = if (keep_lower) 2 else 4)
  text(0.25, ylev[k], k, cex = 0.8, col = 4)
  if (keep_lower) xu <- xmu else xl <- xml
}
```

The first test is the negative log-likelihood of the classical genetic linkage model, with cell counts $(1997, 906, 904, 32)$ and probabilities $\left((2+\theta)/4, (1-\theta)/4, (1-\theta)/4, \theta/4\right)$; it is smooth, so any method would work. The second, $f(x) = |x - 10|$, is not differentiable at its minimum, so no root-finder applies — golden section handles it without modification.

```{r}
#| label: golden-tests
#| fig-width: 10
#| fig-height: 4.5
f1 <- function(theta) -1997 * log(2 + theta) - 1810 * log(1 - theta) - 32 * log(theta)
f2 <- function(x) abs(x - 10)

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
curve(f1, 0.001, 0.999, n = 400, lwd = 2, xlab = TeX(r"($\theta$)"), ylab = TeX(r"($-\ell(\theta)$)"),
      main = TeX(r"(genetic linkage model)"))
abline(v = golden(f1, c(0.001, 0.999)), col = 2, lty = 2, lwd = 2)

curve(f2, 0, 20, n = 400, lwd = 2, xlab = TeX(r"($x$)"), ylab = TeX(r"($|x - 10|$)"),
      main = TeX(r"(not differentiable at the minimum)"))
abline(v = golden(f2, c(0, 20)), col = 2, lty = 2, lwd = 2)

c(linkage = golden(f1, c(0.001, 0.999)), abs_value = golden(f2, c(0, 20)))
```

### Summary of the Methods

| method | needs | convergence | fails when |
|:--|:--|:--|:--|
| fixed point | a rearrangement $\theta = g(\theta)$ | linear, factor $\lvert g'\rvert$ | $\lvert g'\rvert \ge 1$ |
| bisection | $S$ and a sign-changing bracket | linear, factor $1/2$ | no bracket available |
| Newton–Raphson | $S$ and $S'$ | quadratic near the root | flat or wrong curvature |
| method of scoring | $S$ and $I(\theta)$ | linear to quadratic | $E[\cdot]$ unavailable |
| golden section | $\ell$ only, and a bracket | linear, factor $0.618$ | $\ell$ not unimodal |

## Example: Poisson Data With Zeros Unobserved

### The Model

Counts are often recorded only when positive: a study of hospital stays sees admitted patients, a survey of litters sees litters that exist. If $Y \sim \mathrm{Poisson}(\lambda)$ and we observe $X = Y \mid Y > 0$,
$$P(x \mid \lambda) = \frac{e^{-\lambda}\lambda^{x}}{x!}\cdot\frac{1}{1 - e^{-\lambda}}, \qquad x = 1, 2, \ldots$$

Dropping the factorial terms, which do not involve $\lambda$,
$$\ell(\lambda) = \left(\sum_i x_i\right)\log\lambda - n\left[\lambda + \log\left(1 - e^{-\lambda}\right)\right],$$
$$S(\lambda) = \frac{\sum_i x_i}{\lambda} - \frac{n}{1 - e^{-\lambda}}, \qquad
S'(\lambda) = -\frac{\sum_i x_i}{\lambda^{2}} + \frac{n e^{-\lambda}}{\left(1 - e^{-\lambda}\right)^{2}} .$$

Since $E(X) = \lambda/(1 - e^{-\lambda})$, taking $-E[S'(\lambda)]$ gives the expected information
$$n I(\lambda) = \frac{n}{\lambda\left(1 - e^{-\lambda}\right)} - \frac{n e^{-\lambda}}{\left(1 - e^{-\lambda}\right)^{2}} .$$

The score equation has no closed-form root, but the likelihood is unimodal — an ideal problem on which to compare all the methods against a single known answer.

```{r}
#| label: nzp-functions
## log likelihood, omitting factorial terms; vectorized in lambda
nzp.log.likelihood <- function(n, lambda)
  sum(n) * log(lambda) - length(n) * (lambda + log(1 - exp(-lambda)))

nzp.scorefunction <- function(n, lambda)
  sum(n) / lambda - length(n) / (1 - exp(-lambda))

nzp.dscore <- function(n, lambda)
  -sum(n) / lambda^2 + length(n) * exp(-lambda) / (1 - exp(-lambda))^2

nzp.information <- function(n, lambda)
  length(n) / (lambda * (1 - exp(-lambda))) - length(n) * exp(-lambda) / (1 - exp(-lambda))^2

## simulated data: Poisson counts with the zeros discarded
x <- rpois(100, lambda = 1.5)
xp <- x[x > 0]

c(n_observed = length(xp), sample_mean = mean(xp), true_lambda = 1.5)
```

The sample mean overshoots 1.5, because the zeros that would have pulled it down were never recorded.

```{r}
#| label: nzp-loglik-score
#| fig-width: 10
#| fig-height: 4.5
lambdas <- seq(0.1, 8, length.out = 300)

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
plot(lambdas, nzp.log.likelihood(xp, lambdas), type = "l", lwd = 2,
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($\ell(\lambda)$)"), main = TeX(r"(log-likelihood)"))
plot(lambdas, nzp.scorefunction(xp, lambdas), type = "l", lwd = 2,
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($S(\lambda)$)"), main = TeX(r"(score)"))
abline(h = 0, lty = 3)
```

### The Three Iterations

Each method below returns the whole sequence of iterates together with the log-likelihood attained, so the trajectories can be compared directly.

**Simple iteration** rearranges $S(\lambda) = 0$ as
$$\frac{\bar x}{\lambda} = \frac{1}{1 - e^{-\lambda}} \iff \lambda = \bar x\left(1 - e^{-\lambda}\right) = g(\lambda),$$
with contraction factor $g'(\lambda) = \bar x e^{-\lambda}$.

```{r}
#| label: nzp-iterations
nzp.simple.iteration <- function(n, r, lambda0 = mean(n)) {
  mean.n <- mean(n); lambda <- numeric(r + 1); lambda[1] <- lambda0
  for (i in 1:r) lambda[i + 1] <- mean.n * (1 - exp(-lambda[i]))
  data.frame(lambda = lambda, log.lik = nzp.log.likelihood(n, lambda))
}

nzp.newton.raphson <- function(n, r, lambda0 = mean(n)) {
  mean.n <- mean(n); lambda <- numeric(r + 1); lambda[1] <- lambda0
  for (i in 1:r) {
    e <- exp(-lambda[i])
    lambda[i + 1] <- lambda[i] -
      (mean.n / lambda[i] - 1 / (1 - e)) / (e / (1 - e)^2 - mean.n / lambda[i]^2)
  }
  data.frame(lambda = lambda, log.lik = nzp.log.likelihood(n, lambda))
}

nzp.method.of.scoring <- function(n, r, lambda0 = mean(n)) {
  mean.n <- mean(n); lambda <- numeric(r + 1); lambda[1] <- lambda0
  for (i in 1:r) {
    e <- exp(-lambda[i])
    lambda[i + 1] <- lambda[i] -
      (mean.n / lambda[i] - 1 / (1 - e)) / ((e / (1 - e) - 1 / lambda[i]) / (1 - e))
  }
  data.frame(lambda = lambda, log.lik = nzp.log.likelihood(n, lambda))
}
```

### Comparison From Several Starting Values

The three methods are run for 15 iterations from a sensible start, a start far below, and a start far above.

```{r}
#| label: nzp-compare-starts
starts <- c(mean(xp), 0.1, 100)

trace_tab <- function(fun, r = 15)
  vapply(starts, function(s) fun(xp, r, s)$lambda, numeric(r + 1))

paths <- list(simple = trace_tab(nzp.simple.iteration),
              newton = trace_tab(nzp.newton.raphson),
              scoring = trace_tab(nzp.method.of.scoring))

### first six iterates from each starting value
for (m in names(paths)) {
  cat("\n", m, "\n")
  tab <- paths[[m]][1:6, ]
  dimnames(tab) <- list(paste("iter", 0:5), sprintf("start = %.4g", starts))
  print(signif(tab, 6))
}
```

The pattern is the one the theory predicts. Newton–Raphson takes the fewest iterations from a good start, and its digits lock in from the left as the error is squared each step. Simple iteration converges from every start, including the extreme ones, but slowly and at a steady rate. The method of scoring is intermediate: always ascending, and nearly as fast as Newton once close.

```{r}
#| label: nzp-convergence-plot
#| fig-width: 9
#| fig-height: 5
lhat <- nzp.newton.raphson(xp, 30, mean(xp))$lambda[31]

err <- sapply(paths, function(p) abs(p[, 1] - lhat))   # from the mean(xp) start

par(mar = c(4, 4.5, 2.5, 1))
matplot(0:15, pmax(err, 1e-17), log = "y", type = "b", pch = 19, lty = 1, col = c(4, 1, 2),
        xlab = TeX(r"(iteration $k$)"), ylab = TeX(r"($|\lambda_k - \hat{\lambda}|$)"),
        main = TeX(r"(error decay from $\lambda_0 = \bar{x}$)"))
legend("bottomleft", bty = "n", pch = 19, lty = 1, col = c(4, 1, 2),
       legend = c("simple iteration", "Newton-Raphson", "method of scoring"))
```

On a log scale a linear method is a straight line; Newton's curve bends downward until it reaches the floor set by machine precision.

```{r}
#| label: nzp-builtin-check
S_xp  <- function(l) nzp.scorefunction(xp, l)
dS_xp <- function(l) nzp.dscore(xp, l)
ll_xp <- function(l) nzp.log.likelihood(xp, l)

c(newton      = unname(newton(S_xp, dS_xp, mean(xp))["root"]),
  fixed_point = unname(fixed_point(function(l) mean(xp) * (1 - exp(-l)), mean(xp))["root"]),
  bisection   = unname(bisection(S_xp, 0.1, 20)["root"]),
  golden      = golden(function(l) -ll_xp(l), c(0.1, 20), eps = 1e-8),
  uniroot     = uniroot(S_xp, c(0.1, 20), tol = 1e-10)$root,
  optimize    = optimize(function(l) -ll_xp(l), c(0.1, 20), tol = 1e-10)$minimum)
```

All six agree, which is the point of running them together: the methods differ in cost and robustness, not in the answer.

### Standard Error From the Curvature

The optimizer has already produced everything inference needs.

```{r}
#| label: nzp-standard-error
se_obs <- 1 / sqrt(-nzp.dscore(xp, lhat))
se_exp <- 1 / sqrt(nzp.information(xp, lhat))

c(mle = lhat, se_observed = se_obs, se_expected = se_exp)
lhat + c(-1.96, 1.96) * se_obs                     # 95% Wald interval
```

```{r}
#| label: nzp-quadratic-approx
#| fig-width: 9
#| fig-height: 5
lam <- seq(lhat - 4 * se_obs, lhat + 4 * se_obs, length.out = 300)
quad <- -0.5 * (-nzp.dscore(xp, lhat)) * (lam - lhat)^2

par(mar = c(4, 4.5, 2.5, 1))
plot(lam, ll_xp(lam) - ll_xp(lhat), type = "l", lwd = 2,
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($\ell(\lambda) - \ell(\hat{\lambda})$)"),
     main = TeX(r"(log-likelihood and its quadratic approximation at $\hat{\lambda}$)"))
lines(lam, quad, col = 2, lwd = 2, lty = 2)
abline(h = -1.92, col = 4, lty = 3)
abline(v = lhat + c(-1.96, 1.96) * se_obs, col = 4, lty = 3)
legend("bottom", bty = "n", lwd = 2, lty = c(1, 2), col = c(1, 2),
       legend = TeX(c(r"($\ell(\lambda)$)", r"(quadratic from $J(\hat{\lambda})$)")))
```

The parabola implied by the observed information tracks the log-likelihood closely over the width of the interval, so the Wald interval is trustworthy here.

### Evaluating the Naive Estimator

All of this machinery is worth the effort only if the MLE beats the easy alternative. The naive estimator ignores the truncation and reports $\bar x$. Simulating many truncated samples gives the sampling distributions of both estimators.

```{r}
#| label: nzp-mle-vs-naive
#| cache: true
#| fig-width: 10
#| fig-height: 4.5
mle_nzp <- function(n)
  uniroot(function(l) nzp.scorefunction(n, l), c(1e-6, 50), tol = 1e-10)$root

sim_nzp <- function(lambda, n_gen = 100, B = 500)
  vapply(1:B, function(b) {
    y <- rpois(n_gen, lambda); n <- y[y > 0]
    c(naive = mean(n), mle = mle_nzp(n))
  }, c(naive = 0, mle = 0))

for (lambda_true in c(0.5, 1.5)) {
  est <- sim_nzp(lambda_true)
  tab <- t(apply(est, 1, function(e)
    c(bias = mean(e) - lambda_true, sd = sd(e), rmse = sqrt(mean((e - lambda_true)^2)))))
  cat("\nlambda =", lambda_true, "\n"); print(round(tab, 4))
}

est <- sim_nzp(1.5)
par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
for (m in c("naive", "mle")) {
  hist(est[m, ], breaks = 25, col = "grey90",
       xlim = range(est), xlab = TeX(r"($\hat{\lambda}$)"),
       main = TeX(sprintf(r"(%s, bias $= %.3f$)", m, mean(est[m, ]) - 1.5)))
  abline(v = 1.5, col = 2, lwd = 2, lty = 2)
}
```

The naive estimator is biased upward, and severely so when $\lambda$ is small, since that is when most of the probability mass sits on the discarded zero. Its bias does not shrink with sample size: it reflects the wrong model, not too little data. The MLE is centred on the truth with a standard deviation matching the information-based standard error of the previous section. The optimization was the means; the correct estimator was the end.

## Postscript: Interval Data and a Flat Likelihood

Not every likelihood is well approximated by a parabola, and when it is not, the information-based standard error misleads. Suppose each Poisson observation is known only to lie in an interval $[l_i, h_i]$. Then
$$\ell(\lambda) = \sum_{i=1}^{n} \log \Pr\left(l_i \le X_i \le h_i \mid \lambda\right)
= \sum_{i=1}^{n} \log \sum_{x = l_i}^{h_i} \frac{e^{-\lambda}\lambda^{x}}{x!} .$$

The inner sum is computed on the log scale with the log-sum-exp trick to avoid underflow, and the maximization is delegated to `nlm`, which returns a numerical Hessian and hence a standard error.

```{r}
#| label: interval-poisson-functions
log_sum_exp <- function(lx) { mlx <- max(lx); mlx + log(sum(exp(lx - mlx))) }

poisson_log_likelihood <- function(lambda, low, high)
  sum(vapply(seq_along(low),
             function(i) log_sum_exp(dpois(low[i]:high[i], lambda, log = TRUE)),
             numeric(1)))
### equivalently, and faster: sum(log(ppois(high, lambda) - ppois(low - 1, lambda)))

poisson_mle <- function(low, high) {
  if (length(low) != length(high)) stop("low and high have different lengths")
  if (any(floor(low) != low) || any(floor(high) != high)) stop("interval ends are not integers")

  fit <- nlm(function(l) -poisson_log_likelihood(l, low, high),
             p = (mean(low) + mean(high)) / 2, hessian = TRUE)
  est <- fit$estimate; sd <- sqrt(1 / fit$hessian[1, 1])
  list(est = est, sd = sd, CI = est + c(-1.96, 1.96) * sd)
}
```

```{r}
#| label: interval-poisson-fits
#| cache: true
pn <- rpois(1000, lambda = 20)

tight <- poisson_mle(pmax(pn - 1, 0), pn + 4)      # narrow intervals: much information
wide  <- poisson_mle(pmax(pn - 5, 0), pn + 5)     # wide intervals: little information

rbind(tight = c(est = tight$est, sd = tight$sd, tight$CI),
      wide  = c(est = wide$est,  sd = wide$sd,  wide$CI))
```

```{r}
#| label: interval-poisson-loglik
#| fig-width: 10
#| fig-height: 4.5
lam_grid <- seq(1, 50, by = 0.5)

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
for (iv in list(list(lo = pmax(pn - 1, 0), hi = pn + 4, lab = "narrow intervals"),
                list(lo = pmax(pn - 50, 0), hi = pn + 5, lab = "wide intervals"))) {
  lp <- vapply(lam_grid, poisson_log_likelihood, numeric(1), low = iv$lo, high = iv$hi)
  plot(lam_grid, lp - max(lp), type = "l", lwd = 2, ylim = c(-50, 0),
       xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($\ell(\lambda) - \ell(\hat{\lambda})$)"),
       main = TeX(sprintf(r"(%s)", iv$lab)))
  abline(h = -1.92, col = 4, lty = 3)
}
```

With narrow intervals the log-likelihood is sharply peaked and the standard error is small and reliable. With wide intervals, most of which include the lower boundary at zero, the log-likelihood is nearly flat over a broad range: the curvature at the maximum is tiny, the reported standard error is large, and — because the curve is visibly asymmetric and non-quadratic — even that large value understates the true uncertainty. The information-based standard error is an asymptotic approximation resting on local quadratic behaviour, and this is what its failure looks like. When the log-likelihood is this shape, report a likelihood-based interval, read off where $\ell$ drops by 1.92, instead of a Wald interval.

## A Practical Newton-Raphson Algorithm for Finding $\sqrt{a}$

**Derivation.** To find $\sqrt{a}$, take $f(x) = x^2 - a$, whose positive root is $\sqrt a$. Then $f'(x) = 2x$, and the Newton step is

$$x_{k+1} = x_k - \frac{f(x_k)}{f'(x_k)} = x_k - \frac{x_k^{2} - a}{2x_k} = \frac{1}{2}\left(x_k + \frac{a}{x_k}\right).$$

The final form is the Babylonian rule: if $x_k$ is too small then $a/x_k$ is too large, and the root lies between them, so average the pair. Newton's tangent line lands exactly on the midpoint.

The error follows the general quadratic law. Writing $e_k = x_k - \sqrt a$,

$$e_{k+1} = \frac{1}{2}\left(x_k + \frac{a}{x_k}\right) - \sqrt a = \frac{\left(x_k - \sqrt a\right)^{2}}{2x_k} = \frac{e_k^{2}}{2x_k},$$

so the number of correct digits roughly doubles each step.

```{r}
#| label: babylonian-sqrt
babylonian <- function(a, x0 = 1, iters = 6)
  Reduce(function(x, k) (x + a / x) / 2, 1:iters, init = x0, accumulate = TRUE)

traces <- sapply(c(10, 20, 50), babylonian)
dimnames(traces) <- list(paste("iter", 0:6), sprintf("sqrt(%d)", c(10, 20, 50)))

print(traces, digits = 12)
sqrt(c(10, 20, 50))                       # for comparison
signif(abs(traces - rep(sqrt(c(10, 20, 50)), each = 7)), 3)    # errors
```

Starting from $x_0 = 1$ in every case, the first two steps do the coarse work — the larger $a$ is, the further the opening leap — and from the third step on the error squares: roughly $10^{-1}$, then $10^{-3}$, then $10^{-6}$, then machine precision. Six iterations suffice for all three, though $\sqrt{50}$ spends one more step than $\sqrt{10}$ recovering from the poor start.

## The Newton–Raphson Algorithm in Several Dimensions

### Derivation from a quadratic approximation

In the previous section the parameter was a scalar, and the log-likelihood could
be inspected by plotting it. Once $\theta \in \mathbb{R}^p$ neither is true: we
cannot see the surface, and the notion of "bracketing a root" that underlies
bisection and golden section has no direct multivariate analogue. What survives
is the local quadratic approximation, and the Newton–Raphson algorithm is its
direct consequence.

This section develops the multivariate Newton–Raphson method and its
statistical variant, Fisher scoring, applies both to logistic regression,
examines interactively how and why the iteration can diverge, and closes with a
survey of the derivative-free and gradient-only alternatives.

Write the log-likelihood as $\ell(\theta)$ for $\theta \in \mathbb{R}^p$, and
define

$$
S(\theta) = \frac{\partial \ell(\theta)}{\partial \theta}
\quad (p \times 1, \text{ the score}),
\qquad
H(\theta) = \frac{\partial^2 \ell(\theta)}{\partial \theta \, \partial \theta^{\top}}
\quad (p \times p, \text{ the Hessian}).
$$

The negative Hessian $J(\theta) = -H(\theta)$ is the **observed information
matrix**.

Expand $\ell$ to second order about the current iterate $\theta^{(k)}$:

$$
\ell(\theta) \;\approx\; \ell(\theta^{(k)})
 + S(\theta^{(k)})^{\top}(\theta - \theta^{(k)})
 - \tfrac{1}{2}\,(\theta - \theta^{(k)})^{\top} J(\theta^{(k)}) \,(\theta - \theta^{(k)}).
$$

This is a quadratic in $\theta$. If $J(\theta^{(k)})$ is positive definite the
quadratic has a unique maximum, found by setting its derivative to zero:

$$
S(\theta^{(k)}) - J(\theta^{(k)})(\theta - \theta^{(k)}) = 0 .
$$

Solving for $\theta$ gives the **Newton–Raphson update**:

$$
\boxed{\;\theta^{(k+1)} = \theta^{(k)} + J(\theta^{(k)})^{-1} S(\theta^{(k)}) \;}
$$

The algorithm therefore replaces the log-likelihood by the quadratic that
matches its value, slope and curvature at the current point, and jumps to the
maximum of that quadratic. Everything that follows — the fast convergence, and
the modes of failure — comes from how good or bad that replacement is.

Three remarks are worth making explicit.

*The inverse is never formed.* Write the update as
$\theta^{(k+1)} = \theta^{(k)} + \delta^{(k)}$ where $\delta^{(k)}$ solves the
linear system $J(\theta^{(k)})\,\delta = S(\theta^{(k)})$. In R this is
`solve(J, S)`, not `solve(J) %*% S`. The former is faster and numerically more
stable; the latter computes $p^2$ numbers in order to use $p$ of them.

*The step is not the gradient direction.* $\delta^{(k)} = J^{-1}S$ points along
the gradient only when $J$ is a multiple of the identity. The information matrix
rescales and rotates the ascent direction to account for differing curvature and
correlation among the parameters, which is exactly why Newton–Raphson is
insensitive to the scaling of the parameters while gradient descent is not.

*Positive definiteness is an assumption, not a guarantee.* Away from the maximum
$J(\theta^{(k)})$ can be indefinite or singular, in which case the "maximum of
the quadratic" is a saddle point or does not exist, and $\delta^{(k)}$ may point
downhill. @sec-divergence examines this.

#### Fisher scoring

The observed information $J(\theta)$ depends on the data. Its expectation under
the model,

$$
I(\theta) = E_{\theta}\{J(\theta)\} = -E_{\theta}\left\{ \frac{\partial^2 \ell(\theta)}{\partial \theta \, \partial \theta^{\top}} \right\},
$$

is the **expected (Fisher) information matrix**. Replacing $J$ by $I$ in the
update gives the **method of scoring**:

$$
\theta^{(k+1)} = \theta^{(k)} + I(\theta^{(k)})^{-1} S(\theta^{(k)}).
$$

The two algorithms have different iterates but the same fixed point, since both
stop where $S(\theta) = 0$. The practical differences:

| | Newton–Raphson | Fisher scoring |
|:---|:---|:---|
| Curvature matrix | Observed information $J(\theta)$ | Expected information $I(\theta)$ |
| Derivation | Second derivative of $\ell$ | Expectation of that second derivative |
| Positive definite? | Only guaranteed near the maximum | Always, wherever the model is identified |
| Algebra | Can be messy | Often simpler, as data-dependent terms average out |
| Local convergence | Quadratic | Linear, but typically with a very small rate constant |
| Stability from poor starts | Less stable | More stable |

: Newton–Raphson compared with Fisher scoring {#tbl-nr-scoring}

The guaranteed positive definiteness of $I(\theta)$ is the main practical
argument for scoring: the update direction is always an ascent direction, so the
iteration cannot be sent uphill-seeking in a downhill direction by an
ill-behaved Hessian.

For generalised linear models with the **canonical link** the two coincide
exactly, because the second derivative of the log-likelihood does not involve
the response. Logistic regression uses the canonical logit link, so the example
below is simultaneously Newton–Raphson, Fisher scoring, and — as we note at the
end of the section — iteratively reweighted least squares.

### Convergence properties

Near a maximum $\hat\theta$ at which $J(\hat\theta)$ is positive definite,
Newton–Raphson converges **quadratically**: the error satisfies

$$
\|\theta^{(k+1)} - \hat\theta\| \;\le\; C \,\|\theta^{(k)} - \hat\theta\|^{2}
$$

for some constant $C$ depending on the third derivatives. Informally, the number
of correct digits doubles at each iteration, which is why five or six iterations
are usually enough and why a well-started run appears to converge instantly.

The qualification "near" is essential. The theorem is local: it guarantees
nothing about a start far from $\hat\theta$, and nothing at all if $J$ fails to
be positive definite along the way.

### Standard errors as a by-product

At convergence the same matrix that drove the iteration supplies the asymptotic
variance. Under the usual regularity conditions,

$$
\hat\theta \;\dot\sim\; N\!\left(\theta, \; J(\hat\theta)^{-1}\right),
$$

so $\operatorname{SE}(\hat\theta_j) = \sqrt{[J(\hat\theta)^{-1}]_{jj}}$. This is
a genuine practical advantage of Newton-type methods over derivative-free ones:
the curvature information needed for inference has already been computed. Methods
that never form a Hessian require a separate numerical differentiation step, which
is what `hessian = TRUE` requests from `nlm` and `optim`.

## Worked Example: Logistic Regression

### The model and its derivatives

With a single covariate $z_i$ and binary response $r_i$, let
$u_i = \beta_0 + \beta_1 z_i$ and $p_i = e^{u_i}/(1 + e^{u_i})$. The
log-likelihood is

$$
\ell(\beta) = \sum_{i=1}^{n}\left\{ r_i u_i - \log\left(1 + e^{u_i}\right) \right\}.
$$

Writing $X$ for the $n \times 2$ design matrix with rows $(1, z_i)$, the score
and information are

$$
S(\beta) = X^{\top}(r - p), \qquad
J(\beta) = X^{\top} W X, \qquad W = \operatorname{diag}\{p_i(1 - p_i)\}.
$$

Note that $J$ does not involve $r$. This is the canonical-link property
mentioned above, and it means $J(\beta) = I(\beta)$ identically.

Two numerical points, both carried over from the chapter on computer arithmetic.
First, $\log(1 + e^{u})$ overflows for $u \gtrsim 710$; the stable form is
$\log(1+e^u) = \max(u,0) + \log(1 + e^{-|u|})$. Second, $p_i(1-p_i) \to 0$ when
$|u_i|$ is large, so $W$ — and hence $J$ — degenerates toward singularity when
the fitted probabilities are extreme. That observation is the whole explanation
of the divergence we are about to see.

### Implementation

```{r}
#| label: logistic-fns
#| message: false

### numerically stable log(1 + exp(u))
log1pexp <- function(u) pmax(u, 0) + log1p(exp(-abs(u)))

### negative log-likelihood, negative score, and information matrix
lik_score_info <- function(b, r, z) {
    u <- b[1] + b[2] * z
    p <- plogis(u)
    w <- p * (1 - p)
    list(
        neg_loglik = -sum(u * r - log1pexp(u)),
        neg_score  = -c(sum(r - p), sum(z * (r - p))),
        inf        = matrix(c(sum(w), sum(z * w),
                              sum(z * w), sum(z * z * w)), 2, 2)
    )
}

neg_logp_logistic <- function(b, z, r) {
    u <- b[1] + b[2] * z
    -sum(u * r - log1pexp(u))
}
```

Because we minimise the *negative* log-likelihood, the update
$\beta^{(k+1)} = \beta^{(k)} + J^{-1}S$ becomes
$\beta^{(k+1)} = \beta^{(k)} - J^{-1} \times (\text{negative score})$, which is
the form coded below.

```{r}
#| label: logistic-nr

mle_logistic_nr <- function(b0, no_iter, r, z, debug = FALSE) {
    out <- matrix(NA_real_, no_iter + 1, 3,
                  dimnames = list(NULL, c("beta0", "beta1", "neg_loglike")))
    b <- b0
    for (i in seq_len(no_iter + 1)) {
        q <- lik_score_info(b, r, z)
        out[i, ] <- c(b, q$neg_loglik)
        if (debug) print(out[i, ])
        if (i > no_iter) break
        step <- tryCatch(solve(q$inf, q$neg_score),
                         error = function(e) rep(NA_real_, 2))
        if (any(!is.finite(step))) break
        b <- b - step
    }
    out
}
```

```{r}
#| label: logistic-data
#| fig-cap: "Simulated logistic data: true success probability and observed responses."

gen_logistic_data <- function(b, n) {
    z <- sort(runif(n, -2, 2))
    p <- plogis(b[1] + b[2] * z)
    r <- rbinom(n, 1, p)
    plot(z, p, type = "l", ylim = c(0, 1), ylab = "P(r = 1)")
    points(z, r, col = r + 1)
    list(z = z, r = r)
}

set.seed(1)
data <- gen_logistic_data(c(0, 1.5), 200)
```

A start close to the truth converges in a handful of iterations:

```{r}
#| label: logistic-nr-good
round(mle_logistic_nr(c(0, 3), 8, data$r, data$z), 6)
```

The doubling of correct digits predicted by the quadratic-convergence result is
visible in the final rows.

### Comparison with `nlm` and `glm`

```{r}
#| label: logistic-nlm
logit_nlm <- nlm(neg_logp_logistic, c(0, 0),
                 z = data$z, r = data$r, hessian = TRUE)
logit_nlm$estimate
sqrt(diag(solve(logit_nlm$hessian)))   # asymptotic standard errors
```

```{r}
#| label: logistic-glm
summary(glm(r ~ z, family = binomial(), data = data))$coefficients
```

The three routes agree to printing accuracy. `glm` is doing the same arithmetic
as `mle_logistic_nr`: for a GLM the Newton–Raphson step can be rearranged as a
weighted least-squares fit of the working response
$u_i + (r_i - p_i)/\{p_i(1-p_i)\}$ on $X$ with weights $p_i(1-p_i)$, giving
**iteratively reweighted least squares (IRLS)**. This is not a different
algorithm but a different arrangement of the same one, and it explains why
`glm` reports "Fisher Scoring iterations" in its output.

## Divergence {#sec-divergence}

### What goes wrong

The quadratic approximation is trustworthy only in a neighbourhood of the
current iterate. Starting from a large $|\beta_1|$, nearly all fitted
probabilities are pushed to $0$ or $1$, so the weights $p_i(1-p_i)$ are
essentially zero, $J$ is nearly singular, and $J^{-1}$ is enormous. The
resulting step is huge, the new iterate lands far outside the region where the
quadratic was a reasonable fit, and the process can oscillate or run off to
infinity — even though the logistic log-likelihood is globally concave and has a
unique maximum.

The interactive figure below lets you choose a starting value and step through
the iterations. Press the play button on the iteration slider to animate the
path over the contours of the negative log-likelihood. This app runs live only
in the HTML edition of this book; readers of other formats can follow the link
below it.

[]{#sec-nr-divergence-app}

::: {.content-visible when-format="html"}

```{shinylive-r}
#| label: nr-divergence-app
#| standalone: true
#| viewerHeight: 800

library(shiny)

### ---------- fixed data ----------------------------------------------------
set.seed(1)
n <- 200
z <- sort(runif(n, -2, 2))
r <- rbinom(n, 1, plogis(0 + 1.5 * z))

log1pexp <- function(u) pmax(u, 0) + log1p(exp(-abs(u)))
negll <- function(b) {
    u <- b[1] + b[2] * z
    -sum(u * r - log1pexp(u))
}

### ---------- contour grid (computed once) ----------------------------------
B0 <- seq(-3, 3, by = 0.1)
B1 <- seq(-8, 8, by = 0.2)
Z <- outer(B0, B1, Vectorize(function(a, b) negll(c(a, b))))

### ---------- Newton-Raphson path -------------------------------------------
nr_path <- function(b0, maxit = 15, halving = FALSE) {
    P <- matrix(NA_real_, maxit + 1, 3,
                dimnames = list(NULL, c("beta0", "beta1", "negll")))
    b <- b0
    P[1, ] <- c(b, negll(b))
    for (i in seq_len(maxit)) {
        u <- b[1] + b[2] * z
        p <- plogis(u)
        w <- p * (1 - p)
        g <- -c(sum(r - p), sum(z * (r - p)))
        J <- matrix(c(sum(w), sum(z * w), sum(z * w), sum(z * z * w)), 2, 2)
        step <- tryCatch(solve(J, g), error = function(e) rep(NA_real_, 2))
        if (any(!is.finite(step))) break
        if (halving) {
            f0 <- P[i, 3]; s <- 1
            repeat {
                bn <- b - s * step
                fn <- negll(bn)
                if ((is.finite(fn) && fn <= f0) || s < 1e-6) break
                s <- s / 2
            }
            b <- bn
        } else {
            b <- b - step
        }
        if (any(!is.finite(b))) break
        P[i + 1, ] <- c(b, negll(b))
    }
    P
}

### ---------- UI -------------------------------------------------------------
ui <- fluidPage(
    titlePanel(NULL),
    sidebarLayout(
        sidebarPanel(
            width = 4,
            numericInput("b0", HTML("starting &beta;<sub>0</sub>"),
                         value = 0, step = 0.5),
            numericInput("b1", HTML("starting &beta;<sub>1</sub>"),
                         value = 5, step = 0.5),
            checkboxInput("halve", "apply step halving", FALSE),
            sliderInput("iter", "iteration", min = 0, max = 15, value = 0,
                        step = 1, animate = animationOptions(interval = 900)),
            helpText("Try (0, 3) and then (0, 5). The second start makes the ",
                     "information matrix nearly singular, so the first step ",
                     "overshoots badly."),
            tableOutput("tab")
        ),
        mainPanel(
            width = 8,
            plotOutput("contour", height = "380px"),
            plotOutput("trace", height = "200px")
        )
    )
)

### ---------- server ---------------------------------------------------------
server <- function(input, output, session) {

    path <- reactive({
        nr_path(c(input$b0, input$b1), maxit = 15, halving = input$halve)
    })

    shown <- reactive({
        P <- path()
        k <- min(input$iter + 1, nrow(P))
        P[seq_len(k), , drop = FALSE]
    })

    output$contour <- renderPlot({
        P <- shown()
        par(mar = c(4, 4, 2, 1))
        contour(B0, B1, Z, nlevels = 40, drawlabels = FALSE,
                col = "grey70",
                xlab = expression(beta[0]), ylab = expression(beta[1]),
                main = "negative log-likelihood surface and NR path")
        ok <- complete.cases(P)
        if (sum(ok) > 1) {
            lines(P[ok, 1], P[ok, 2], col = "steelblue", lwd = 2)
            points(P[ok, 1], P[ok, 2], pch = 21, bg = "steelblue", cex = 1.1)
        }
        last <- P[max(which(complete.cases(P))), , drop = FALSE]
        points(last[1], last[2], pch = 21, bg = "firebrick", cex = 1.8)
        outside <- last[1] < min(B0) || last[1] > max(B0) ||
                   last[2] < min(B1) || last[2] > max(B1)
        if (isTRUE(outside)) {
            mtext(sprintf("current iterate (%.1f, %.1f) is off the plotted region",
                          last[1], last[2]), side = 3, line = -1.5,
                  col = "firebrick", cex = 0.9)
        }
        points(0, 1.5, pch = 3, col = "darkgreen", lwd = 2)
    })

    output$trace <- renderPlot({
        P <- path()
        k <- min(input$iter + 1, nrow(P))
        par(mar = c(4, 4, 1, 1))
        yl <- range(P[, 3], na.rm = TRUE)
        plot(0:(nrow(P) - 1), P[, 3], type = "n", ylim = yl,
             xlab = "iteration", ylab = "negative log-likelihood")
        lines(0:(k - 1), P[seq_len(k), 3], type = "b", pch = 19,
              col = "steelblue")
    })

    output$tab <- renderTable({
        P <- shown()
        d <- as.data.frame(P)
        d$iteration <- seq_len(nrow(d)) - 1
        d <- d[, c("iteration", "beta0", "beta1", "negll")]
        tail(d, 5)
    }, digits = 4, rownames = FALSE)
}

shinyApp(ui, server)
```

:::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit06_MLE.html#sec-nr-divergence-app>.
:::

::: {.callout-note}
### Reproducing this figure
The chunk above uses the [shinylive](https://github.com/quarto-ext/shinylive)
extension, which compiles the app to WebAssembly so that it runs in the reader's
browser with no Shiny server. Install it once per project with

```bash
quarto add quarto-ext/shinylive
```

and add `filters: [shinylive]` to the document or project YAML. Only packages
available in webR may be used inside the app, so the code above is restricted to
base R and `shiny`.
:::

### Safeguards

The divergence is not a defect of the Newton direction but of the Newton *step
length*. The standard remedies all keep the direction and control the distance.

**Step halving (backtracking).** Take $\delta^{(k)}$ as usual but move
$\theta^{(k+1)} = \theta^{(k)} + s\,\delta^{(k)}$, starting from $s = 1$ and
halving until the log-likelihood actually increases. This costs a few extra
function evaluations, never damages quadratic convergence near the maximum
(where $s = 1$ is accepted on the first try), and rescues most divergent runs.
Toggle it in the app above to see the effect.

**Ridging (Levenberg–Marquardt).** Replace $J$ by $J + \lambda \mathbb{I}$ with
$\lambda > 0$. Large $\lambda$ shrinks the step toward a small gradient step;
$\lambda \to 0$ recovers Newton. This also repairs indefiniteness, and is the
approach `nlm` takes internally.

**Trust regions.** Maximise the quadratic approximation subject to
$\|\theta - \theta^{(k)}\| \le \Delta_k$, adapting $\Delta_k$ according to how
well the quadratic predicted the realised change. This is `nlminb` and the
`PORT` routines.

**Reparameterisation and good starts.** Often the cheapest fix. Centring and
scaling covariates improves conditioning of $J$ substantially, and for GLMs
the null-model fit is a reliable starting point — which is what `glm` uses when
`start` is not supplied.

## Other Multivariate Optimization Methods

Newton–Raphson is the method of choice when the second derivatives are available
and $p$ is modest. Outside that regime, three considerations push toward
alternatives: derivatives may be unavailable or painful to code; the $O(p^3)$
cost of solving the Newton system and the $O(p^2)$ storage of $J$ become
prohibitive for large $p$; and the objective may be non-smooth, noisy or
multimodal.

### Classification by derivative information

Methods are most usefully organised by what they require of the objective.

| Order | Requires | Methods | Typical use |
|---:|:---|:---|:---|
| 0 | Function values only | Nelder–Mead, simulated annealing, genetic algorithms | Non-smooth, discontinuous, or awkward-to-differentiate objectives; small $p$ |
| 1 | Gradient | Steepest descent, conjugate gradient, quasi-Newton (BFGS, L-BFGS), stochastic gradient descent | The workhorse class; large $p$; gradients cheap by automatic differentiation |
| 2 | Gradient and Hessian | Newton–Raphson, Fisher scoring, IRLS, trust-region Newton | Statistical models of modest dimension where the Hessian is also wanted for inference |

: Optimization methods classified by the derivative information required {#tbl-orders}

### The main alternatives

**Nelder–Mead (simplex).** Maintains $p+1$ points forming a simplex and moves it
downhill by reflecting, expanding, contracting and shrinking. It needs no
derivatives and tolerates mild non-smoothness, but converges slowly, degrades
badly as $p$ grows past roughly 10–20, and can stall at non-stationary points.
It is `optim`'s default, which makes it by far the most-used and most-overused
method in R. Reasonable as a first look at an unfamiliar objective; a poor
choice for a likelihood you will fit thousands of times.

**Steepest descent with line search.** Move along $-\nabla f$, choosing the
distance by a one-dimensional search. Conceptually the simplest gradient method
and the one that best illustrates the role of a line search, but the
convergence is only linear, and the rate deteriorates with the condition number
of the Hessian: on an elongated valley the iterates zig-zag across it rather
than travelling along it. Rarely used unmodified.

**Conjugate gradient.** Corrects the zig-zag by choosing each direction to be
conjugate to the previous ones with respect to the local curvature, without ever
forming or storing a Hessian. Storage is $O(p)$, which is its defining
advantage. Fletcher–Reeves and Polak–Ribière differ only in the formula for the
correction coefficient. The method is sensitive to the accuracy of the line
search and usually needs periodic restarting.

**Quasi-Newton (BFGS and L-BFGS).** Builds an approximation to $J^{-1}$ from
successive gradient differences, so it attains superlinear convergence with only
first derivatives. BFGS stores a full $p \times p$ approximation; L-BFGS keeps
only the last few update vectors, reducing storage to $O(mp)$ and making it the
default choice for high-dimensional smooth problems. For most likelihood
problems where the Hessian is awkward to derive, BFGS with an analytic gradient
is the best practical compromise.

**Stochastic gradient descent.** For an objective that is a sum over $n$
observations, estimate the gradient from a small random subset ("minibatch") at
each step. The per-iteration cost becomes independent of $n$, at the price of a
noisy trajectory requiring a decreasing step-size sequence
($\sum_k \alpha_k = \infty$, $\sum_k \alpha_k^2 < \infty$). Modern variants —
momentum, AdaGrad, RMSProp, Adam — adapt the step size per coordinate. SGD is
the foundation of neural-network fitting and is treated in its own right in a
later chapter; for classical likelihood problems with $n$ in the thousands it
offers nothing over BFGS.

### Comparison

| Method | Derivatives | Storage | Cost per iteration | Local rate | Robust start | Hessian for SEs |
|:---|:---|:---|:---|:---|:---|:---|
| Nelder–Mead | none | $O(p^2)$ | $O(p)$ evaluations | sublinear | good | no |
| Steepest descent | gradient | $O(p)$ | 1 gradient + line search | linear (slow) | good | no |
| Conjugate gradient | gradient | $O(p)$ | 1 gradient + line search | linear (fast) | moderate | no |
| BFGS | gradient | $O(p^2)$ | 1 gradient + $O(p^2)$ | superlinear | moderate | approximate |
| L-BFGS | gradient | $O(mp)$ | 1 gradient + $O(mp)$ | superlinear | moderate | no |
| Newton–Raphson | gradient, Hessian | $O(p^2)$ | 1 Hessian + $O(p^3)$ | quadratic | poor | exact |
| Fisher scoring / IRLS | score, expected information | $O(p^2)$ | 1 weighted LS fit | near-quadratic | fair | exact |
| Stochastic gradient | minibatch gradient | $O(p)$ | $O(\text{batch})$ | sublinear | good | no |

: Comparison of multivariate optimization methods. $p$ is the number of parameters and $m$ the L-BFGS memory length (typically 5–20). {#tbl-comparison}

### Implementations in R

| Function | Package | Methods available | Notes |
|:---|:---|:---|:---|
| `optim` | stats | Nelder–Mead, BFGS, CG, L-BFGS-B, SANN | General workhorse; supply `gr` or the gradient is differenced; `hessian = TRUE` for inference; L-BFGS-B allows box constraints |
| `nlm` | stats | Newton-type with ridging | Uses analytic gradient/Hessian if attached as attributes, otherwise numerical; returns `$hessian` |
| `nlminb` | stats | Trust-region (PORT) | Reliable on badly scaled problems; supports box constraints |
| `optimize`, `uniroot` | stats | Golden section, Brent | One dimension only |
| `glm` | stats | IRLS (Fisher scoring) | The specialised case; use `start` when separation is suspected |
| `nls` | stats | Gauss–Newton, Golub–Pereyra | Nonlinear least squares; `port` algorithm adds constraints |
| `optimx` | optimx | Unified front end to a dozen optimizers | Useful for comparing methods on the same objective |
| `nloptr` | nloptr | Large NLopt collection, with constraints | Global and derivative-free options |
| `Rsolnp`, `alabama` | — | Nonlinear equality and inequality constraints | For genuinely constrained problems |

: R interfaces to multivariate optimizers {#tbl-r-functions}

### Choosing a method

| Situation | Suggested approach |
|:---|:---|
| Hessian derivable, $p$ small, standard errors wanted | Newton–Raphson or Fisher scoring |
| GLM or a model expressible as one | `glm` (IRLS) |
| Gradient available, $p$ moderate | BFGS via `optim(method = "BFGS")` |
| Gradient available, $p$ large | L-BFGS |
| Derivatives unavailable, $p < 10$ | Nelder–Mead as a first pass, then refine |
| Objective noisy or non-smooth | Nelder–Mead or simulated annealing |
| Divergence or poor conditioning suspected | `nlm` or `nlminb`; rescale the parameters |
| Objective a sum over very many observations | Stochastic gradient descent and variants |
| Multiple local maxima suspected | Multiple random starts with any of the above |

: Practical guidance on method selection {#tbl-choosing}

The last row deserves emphasis. Every method in this chapter is a *local*
method: each converges to a stationary point near where it started, and none
can certify that the point found is the global maximum. Restarting from a
dispersed set of initial values and comparing the attained log-likelihoods is
the only general safeguard, and it should be routine for any likelihood that is
not known to be concave.
