---
title: "Laplace Approximation"
format:
  html:
    toc: true
    tbl-colwidths: true
filters:
  - shinylive
editor_options:
  chunk_output_type: console
code-fold: true
---

The previous chapter approximated an integral by a weighted sum of function
values on a grid. A grid gets expensive fast: an $n$-point-per-axis product
grid costs $n^d$ evaluations, so quadrature is only practical for $d \lesssim
3$. The **Laplace approximation** takes a completely different route: instead
of evaluating the integrand everywhere, it looks only at the single most
important point — the mode — and replaces the whole integrand there by the
Gaussian that matches its height and curvature. No grid is needed at any
dimension, at the price of assuming the integrand is roughly bell-shaped.

## The Goal and the Idea

Many Bayesian calculations are posterior expectations
$$
E_f\big[a(\theta)\big] = \frac{\int a(\theta)\,f(\theta)\,d\theta}{\int f(\theta)\,d\theta},
$$
where $\theta \in \mathbb{R}^p$, $a(\theta)$ is some quantity of interest (a
parameter, a prediction, an indicator), and $f(\theta)$ is an **unnormalized**
posterior, typically $f(\theta) = L(\theta)\pi(\theta)$. Both numerator and
denominator are $p$-dimensional integrals.

Write $h(\theta) = -\log f(\theta)$, so $f(\theta) = e^{-h(\theta)}$. For
independent data, $h$ is (up to an additive constant) minus the
log-likelihood plus minus the log-prior — a sum of $n$ terms — so as $n$ grows
it becomes large and sharply curved at its minimum. This is the same
large-sample regularity that makes the MLE asymptotically normal
(seen in the maximum-likelihood chapters); here it makes the *posterior* asymptotically normal too. The
Laplace approximation makes this precise by expanding $h$ to second order at
its minimum and integrating the resulting Gaussian exactly — no grid, no
sampling, just one optimization and one Hessian.

## Derivation

Let $\theta_0 = \arg\min_\theta h(\theta)$, the posterior mode, found
numerically (`optim(..., hessian = TRUE)` or `nlm(..., hessian = TRUE)` in
R — exactly the optimizers used to find the MLE in earlier chapters). Expand
$h$ around $\theta_0$:
$$
h(\theta) = h(\theta_0) + \underbrace{\nabla h(\theta_0)^\top}_{=\,0\text{ at the minimum}}(\theta - \theta_0)
 + \tfrac12(\theta-\theta_0)^\top H (\theta - \theta_0) + O\big(\|\theta-\theta_0\|^3\big),
\qquad H = \nabla^2 h(\theta_0).
$$
$H$, the Hessian of $-\log f$ at the mode, is positive definite at a genuine
minimum; for $f = L\pi$ it is the observed information plus the curvature of
the log-prior. Dropping the cubic remainder and exponentiating,
$$
f(\theta) \approx \tilde f(\theta) = f(\theta_0)\exp\Big\{-\tfrac12(\theta-\theta_0)^\top H(\theta-\theta_0)\Big\}
 = \underbrace{f(\theta_0)(2\pi)^{p/2}|H|^{-1/2}}_{\text{constant}}
   \cdot \underbrace{(2\pi)^{-p/2}|H|^{1/2}e^{-\frac12(\theta-\theta_0)^\top H(\theta-\theta_0)}}_{N_p(\theta_0,\,H^{-1})\text{ density}}.
$$
Two results follow immediately, since the bracketed factor is a properly
normalized $N_p(\theta_0, H^{-1})$ density:

$$
\boxed{\ \theta \mid \text{data} \ \approx\ N_p\big(\theta_0,\ H^{-1}\big) \ }
\qquad\qquad
\boxed{\ \int f(\theta)\,d\theta \ \approx\ f(\theta_0)\,(2\pi)^{p/2}\,|H|^{-1/2} \ }
$$

Only the mode and the Hessian of $-\log f$ at the mode are needed — both come
for free from any optimizer that returns a Hessian. The relative error of the
second result is $O(1/n)$ for regular models; the error of the first is
governed by how skewed the true posterior actually is.

## Worked Example: A Poisson Rate with a Gamma Prior

To see the approximation against ground truth, use a model with a known exact
answer: the Gamma–Poisson conjugate pair from the Bayesian Inference chapter.
For $y_1,\ldots,y_n \mid \lambda \sim \text{Poisson}(\lambda)$ and
$\lambda \sim \text{Gamma}(\alpha,\beta)$, the posterior is exactly
$\text{Gamma}(\alpha + \sum y_i,\ \beta+n)$, and the marginal likelihood
derived there in closed form,
$\log P(\mathbf y) = \alpha\log\beta - \log\Gamma(\alpha) + \log\Gamma(\alpha+S)
- (\alpha+S)\log(\beta+n) - \sum_i\log(y_i!)$ with $S=\sum_i y_i$, is exactly
`exact_log_ml` below — so both the exact log marginal likelihood and the
exact posterior are available to measure the Laplace error directly.

```{r}
#| label: poisson-gamma-functions
alpha <- 2; beta <- 1

### Laplace approximation: mode, Hessian, and the two boxed results above
laplace_poisson <- function(y) {
  h  <- function(l) -(sum(dpois(y, l, log = TRUE)) + dgamma(l, alpha, beta, log = TRUE))
  op <- optim(mean(y) + 0.5, h, method = "L-BFGS-B", lower = 1e-6, hessian = TRUE)
  c(mode = op$par, se = sqrt(1 / op$hessian[1, 1]),
    log_ml = -op$value + 0.5 * log(2 * pi) - 0.5 * log(op$hessian[1, 1]))
}

### exact log marginal likelihood, from the Gamma-Poisson conjugate identity
exact_log_ml <- function(y) {
  n <- length(y); s <- sum(y)
  alpha * log(beta) - lgamma(alpha) + lgamma(alpha + s) -
    (alpha + s) * log(beta + n) - sum(lfactorial(y))
}
```

```{r}
#| label: tbl-poisson-gamma-error
#| tbl-cap: "Assesses the accuracy of the Laplace approximation by comparing it with the exact marginal likelihood in a conjugate example where the exact answer is known.
#|   Laplace log marginal likelihood versus the exact value, and the posterior mean versus the Laplace mode, as n grows. The error shrinks roughly like 1/n: it falls by about a factor of 4-5 each time n increases by that factor."
set.seed(42)
res <- t(sapply(c(5, 20, 100, 500), function(n) {
  y <- rpois(n, 3)
  lap <- laplace_poisson(y)
  c(n = n, laplace_log_ml = lap["log_ml"], exact_log_ml = exact_log_ml(y),
    error = lap["log_ml"] - exact_log_ml(y),
    laplace_mode = lap["mode"], posterior_mean = (alpha + sum(y)) / (beta + n))
}))
knitr::kable(res, digits = 4, row.names = FALSE,
             col.names = c("n", "Laplace log P(y)", "exact log P(y)", "error",
                           "Laplace mode", "posterior mean"))
```

The error is a few thousandths at $n=5$ and shrinks to a few ten-thousandths
by $n=500$; the Laplace mode also converges to the posterior mean as the
Gamma posterior becomes more symmetric. Plotting the exact posterior against
its Laplace approximation makes this concrete:

```{r}
#| label: fig-poisson-gamma-density
#| fig-width: 10
#| fig-height: 4.4
#| fig-cap: "Shows how the Laplace approximation replaces a posterior by a normal density centred at the mode, and how well it fits at small and large sample sizes.
#|   Exact Gamma posterior (red) and the Laplace normal approximation (green) for n=5 and n=100. At n=5 the exact posterior is visibly right-skewed and the Gaussian is a rough match; by n=100 the two are nearly indistinguishable, illustrating the O(1/n) convergence in the table above."
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
set.seed(42)
for (n in c(5, 100)) {
  y <- rpois(n, 3); a1 <- alpha + sum(y); b1 <- beta + n
  lap <- laplace_poisson(y)
  curve(dgamma(x, a1, b1), max(0, lap["mode"] - 4 * lap["se"]), lap["mode"] + 4 * lap["se"],
        lwd = 3, col = "firebrick", xlab = expression(lambda), ylab = "density",
        main = sprintf("n = %d", n))
  curve(dnorm(x, lap["mode"], lap["se"]), add = TRUE, lwd = 3, col = "forestgreen")
  if (n == 5) legend("topright", bty = "n", lwd = 3, col = c("firebrick", "forestgreen"),
                      legend = c("exact posterior", "Laplace approximation"))
}
```

### Linear functionals: posterior mean and credible intervals

If $a(\theta) = A^\top\theta + a_0$ is linear, replacing $f$ by $\tilde f$
gives $E_f[a(\theta)] \approx A^\top\theta_0 + a_0$: the posterior mean is
approximated by the posterior mode, and $\text{Var}(A^\top\theta) \approx
A^\top H^{-1}A$ gives Wald-style credible intervals $\theta_{0j} \pm
z_{\alpha/2}\sqrt{(H^{-1})_{jj}}$ — exactly the Bayesian counterpart of an
MLE plus its Wald standard error from the maximum-likelihood chapters.

```{r}
#| label: tbl-poisson-gamma-ci
#| tbl-cap: "Compares interval estimates based on the Laplace approximation with the exact posterior intervals, to see how much accuracy is lost.
#|   Laplace mode/SE-based 95% interval versus the exact Gamma posterior mean/sd-based interval, n = 20."
set.seed(42)
n <- 20; y <- rpois(n, 3); lap <- laplace_poisson(y)
a1 <- alpha + sum(y); b1 <- beta + n
comp <- data.frame(
  quantity = c("Laplace (mode / se)", "exact posterior (mean / sd)"),
  center = c(lap["mode"], a1 / b1),
  spread = c(lap["se"], sqrt(a1) / b1),
  lower95 = c(lap["mode"] - 1.96 * lap["se"], qgamma(0.025, a1, b1)),
  upper95 = c(lap["mode"] + 1.96 * lap["se"], qgamma(0.975, a1, b1))
)
knitr::kable(comp, digits = 4)
```

The two intervals agree closely even at $n=20$: the mild right-skew of the
Gamma posterior visible in @fig-poisson-gamma-density has not yet noticeably
distorted a Wald-style interval built from a symmetric Gaussian.

### BIC as a large-$n$ simplification

With $f(\theta) = L(\theta;D)\pi(\theta)$, the Laplace approximation to the
log marginal likelihood is
$$
\log P(D) \approx \log\big[L(\theta_0;D)\,\pi(\theta_0)\big] + \frac p2\log(2\pi) - \frac12\log|H|.
$$
For large $n$ the Hessian is dominated by the likelihood, $H \approx
n\,I(\theta_0)$ with $I$ the per-observation Fisher information, so $|H|
\approx n^p|I(\theta_0)|$ and
$$
\log P(D) \approx \log L(\theta_0;D) - \frac p2\log n
 + \underbrace{\log\pi(\theta_0) + \frac p2\log(2\pi) - \frac12\log|I(\theta_0)|}_{O(1),\text{ dropped}}.
$$
Multiplying by $-2$ and replacing the posterior mode by the MLE (the two
coincide to $O(1/n)$) gives the **Bayesian information criterion**,
$\text{BIC} = -2\log L(\hat\theta_{\text{MLE}}) + p\log n$ — a model-comparison
score that needs no prior at all, because the dropped $O(1)$ terms do not grow
with $n$ and so wash out of *differences* between models fit to the same
data. For the $n=20$ Poisson example from the credible-interval check above,
$-2\times(\text{Laplace log }P(y))$ and the MLE-based BIC differ by about
`r round(-2*laplace_poisson(y)["log_ml"] - (-2*sum(dpois(y, mean(y), log=TRUE)) + log(n)), 2)` —
a constant offset consistent with the dropped $O(1)$ terms, not something
that grows with $n$.

### When Laplace Works Well, and When It Does Not

**Works well** when $n$ is large relative to $p$ and the posterior is
unimodal and roughly symmetric, as in the $n=100$ panel above. Parameters
should live on an unbounded scale — transform first ($\log\sigma$,
$\text{logit}\,p$) so the matching Gaussian is not truncated at a boundary,
remembering to include the Jacobian of the transform in $f$ (exactly the
$w=\log\sigma^2$ reparametrization used throughout this book since the
multivariate optimization chapter).

**Struggles** when the posterior is skewed (small $n$, a parameter near a
boundary) — the mode is then a poor stand-in for the mean and the Gaussian
tails are wrong, as the $n=5$ panel above already shows mildly; when the
posterior is multimodal, since the approximation only sees the one mode an
optimizer happened to find; and in hierarchical models where $p$ grows with
$n$, so the $O(1/n)$ error does not vanish (nested Laplace approximation,
INLA, handles this by integrating out the high-dimensional part
analytically).

## Example: Comparing Laplace, Quadrature, and Monte Carlo on Computing Marginal Likelihood

To compare the three methods on an apples-to-apples problem, reuse the
single-parameter example from the end of Unit 08: a single unknown mean
$\mu$, with likelihood
$$
L(\mu) = N\big(\hat\mu;\,\mu,\,1/n_{\text{eff}}\big),
$$
treating $\hat\mu$ as the sample mean of $n_{\text{eff}}$ observations from
$N(\mu,1)$, under a prior $\pi(\mu) = N(\mu;0,5^2)$. The marginal likelihood
is
$$
P(y) = \int_{-\infty}^{\infty} z(\mu)\,d\mu, \qquad z(\mu) = \pi(\mu)L(\mu).
$$
Because this is a normal-normal conjugate pair, $P(y)$ has a closed form, so
a genuine **true** value is available for comparison — computed here by
numerical integration, exactly as in Unit08, rather than by writing out the
closed-form expression. **Laplace** attacks the integral by optimization
instead of a grid: find the mode of $z(\mu)$ and the second derivative of
$-\log z$ there, then apply the boxed formula with $p=1$. **Quadrature**
attacks it with the logistic-transform midpoint rule from Unit08. **Monte
Carlo with $\pi$** attacks it by drawing $\mu$ directly from the prior and
averaging the likelihood — the same naive estimator used throughout Unit08.

```{r}
#| label: laplace-mu-functions
n_eff <- 3; sigma_prior <- 5                   # same setup as the Unit08 app

### true value: exact up to numerical-integration error, since z is a
### normal-normal conjugate kernel
log_mar_true_mu <- function(mu_hat) {
  z <- function(mu) dnorm(mu_hat, mu, 1 / sqrt(n_eff)) * dnorm(mu, 0, sigma_prior)
  log(integrate(z, -Inf, Inf)$value)
}

### Laplace: p = 1 case of the boxed formula, mode and Hessian from nlm()
laplace_mu <- function(mu_hat) {
  h  <- function(mu) -(dnorm(mu_hat, mu, 1 / sqrt(n_eff), log = TRUE) +
                          dnorm(mu, 0, sigma_prior, log = TRUE))
  op <- nlm(h, p = mu_hat, hessian = TRUE)
  c(mode = op$estimate, log_ml = -op$minimum + 0.5 * log(2 * pi) - 0.5 * log(op$hessian[1, 1]))
}

### naive Monte Carlo: draw mu from the prior, average the likelihood
log_sum_exp <- function(lx) { m <- max(lx); m + log(sum(exp(lx - m))) }
log_mar_mc_mu <- function(mu_hat, iters) {
  mus <- rnorm(iters, 0, sigma_prior)
  log_sum_exp(dnorm(mu_hat, mus, 1 / sqrt(n_eff), log = TRUE)) - log(iters)
}

### midpoint rule on the logistic-transformed (0, 1) grid, as in Unit08
log_mar_mid_mu <- function(mu_hat, n) {
  h  <- 1 / n
  u  <- (seq_len(n) - 0.5) * h
  mu <- log(u / (1 - u))
  log_z <- dnorm(mu_hat, mu, 1 / sqrt(n_eff), log = TRUE) +
    dnorm(mu, 0, sigma_prior, log = TRUE) - log(u * (1 - u))
  log_sum_exp(log_z) + log(h)
}
```

```{r}
#| label: tbl-three-way
#| tbl-cap: "Compares the Laplace approximation with grid quadrature and naive Monte Carlo against the true value as the posterior moves away from the prior mean, to show where each method succeeds or fails.
#|   Log marginal likelihood by four methods, for data locations mu_hat = 0, -5, -20 (n_eff = 3, sigma = 1, prior N(0, 5^2) on mu). Quadrature uses a 100-point midpoint grid; Monte Carlo uses 1e6 prior draws; both are unchanged from Unit08's app."
set.seed(7)
mu_hats <- c(0, -5, -20)
three_way <- t(sapply(mu_hats, function(m) {
  c(True = log_mar_true_mu(m),
    Laplace = laplace_mu(m)["log_ml"],
    Monte_Carlo_pi = log_mar_mc_mu(m, 1e6),
    Quadrature = log_mar_mid_mu(m, 100))
}))
knitr::kable(cbind(`mu` = mu_hats, round(three_way, 3)), row.names = FALSE,
             col.names = c("mu", "True", "Laplace", "Monte Carlo (pi)", "Quadrature (n=100)"))
```

At $\mu=0$ the posterior sits right where both the logistic grid and the
prior put their effort, so all four numbers should agree closely. As $\mu$
moves to $-5$ and then $-20$, the fixed grid — centred at $0$ by
construction — squeezes the posterior into an ever-thinner sliver of the
unit interval, and quadrature is the first to break away from the true
value, exactly the failure diagnosed at the end of Unit08. Naive Monte Carlo
is built on the same fragile assumption in a different guise: it needs prior
draws *near* $\mu$ to contribute any likelihood at all, so once $\mu$ is
several prior standard deviations from $0$, a fixed budget of draws from
$\pi$ can also become too sparse there to be trusted, even though nothing
about the Monte Carlo estimator's derivation assumed $\mu$ would be small.
Laplace has neither failure mode: it optimizes to find the mode wherever it
actually is and builds its approximation *around that point*, so it
automatically "recentres" itself for every $\mu$ — the same fix Unit08
proposed by hand for quadrature, done here for free by the optimizer.

## Adaptive Gauss–Hermite Quadrature {#sec-laplace-transform-app}

A fixed logistic transform ($u=1/(1+e^{-\mu})$) fails when the posterior mass moves far from $\mu=0$. **Adaptive Gauss-Hermite quadrature (AGHQ)** solves this by relocating and rescaling the quadrature grid using the Laplace approximation's mode ($\hat\mu_L$) and standard deviation ($s_L=1/\sqrt{H}$). 

Think of it as a smart, localized search grid: instead of laying a massive grid over the entire parameter space, AGHQ finds the peak of the posterior (the mode), measures its exact curvature (the Hessian), and places a custom-scaled grid directly over that localized area.

By swapping the logistic function for the CDF and density of $N(\hat\mu_L, s_L^2)$, the transform becomes:
$$
u = \Phi\!\left(\frac{\mu-\hat\mu_L}{s_L}\right), \qquad \mu = \hat\mu_L + s_L\,\Phi^{-1}(u), \qquad \frac{du}{d\mu} = \phi\!\left(\frac{\mu-\hat\mu_L}{s_L}\right)
$$

Applying a midpoint rule on this adapted grid yields:
$$
P(y) \approx \frac1n\sum_{i=1}^n \frac{z(\mu_i)}{\phi\big((\mu_i-\hat\mu_L)/s_L\big)/s_L}, \qquad \mu_i = \hat\mu_L + s_L\,\Phi^{-1}(u_i), \qquad u_i = \big(i-\tfrac12\big)/n
$$

### Shinylive App for Adaptive Gauss–Hermite Quadrature

::::: {.content-visible when-format="html"}
::: {#fig-laplace-transform-app}

```{shinylive-r}
#| label: laplace-transform-app
#| standalone: true
#| viewerHeight: 760

library(shiny)

ui <- fluidPage(
    titlePanel("Shinylive App for Adaptive Gauss–Hermite Quadrature"),
    sidebarLayout(
        sidebarPanel(
            width = 4,
            numericInput("mu_hat", "mu_hat (data location)", value = 8, step = 1),
            sliderInput("n_pts", "number of grid points in u (over 0 to 1)",
                        min = 2, max = 25, value = 9, step = 1),
            helpText("The grid is a midpoint rule, uniform in u = Phi((mu - ",
                     "mode_L)/se_L), where mode_L and se_L come from the ",
                     "Laplace approximation recomputed for the current ",
                     "mu_hat. Try mu_hat = 8 or 20 (values that broke the ",
                     "plain logistic grid in the earlier app) and watch the ",
                     "quadrature estimate stay close to the true value.")
        ),
        mainPanel(
            width = 8,
            plotOutput("squeeze", height = "560px")
        )
    )
)

server <- function(input, output, session) {

    output$squeeze <- renderPlot({
        n_eff <- 3; sigma_prior <- 5
        L     <- function(mu) dnorm(input$mu_hat, mu, 1 / sqrt(n_eff))
        prior <- function(mu) dnorm(mu, 0, sigma_prior)
        z     <- function(mu) L(mu) * prior(mu)

        log_sum_exp <- function(lx) { m <- max(lx); m + log(sum(exp(lx - m))) }

        ### true value, exactly as in the Unit08 / earlier Unit09 apps
        log_marlik <- log(integrate(z, -Inf, Inf)$value)

        ### Laplace approximation for THIS mu_hat: mode and se define the transform
        h_neg <- function(mu) -(dnorm(input$mu_hat, mu, 1 / sqrt(n_eff), log = TRUE) +
                                   dnorm(mu, 0, sigma_prior, log = TRUE))
        op       <- nlm(h_neg, p = input$mu_hat, hessian = TRUE)
        lap_mode <- op$estimate
        lap_se   <- 1 / sqrt(op$hessian[1, 1])

        lo <- min(-8, input$mu_hat - 6)
        hi <- max(14, input$mu_hat + 6)
        mu_grid <- seq(lo, hi, length.out = 500)

        u_pts  <- (seq_len(input$n_pts) - 0.5) / input$n_pts   # midpoint rule, uniform on (0, 1)
        mu_pts <- qnorm(u_pts, lap_mode, lap_se)                # mu = Laplace-CDF^{-1}(u)

        ### quadrature estimate built from this Laplace-shaped grid
        log_z           <- function(mu) dnorm(input$mu_hat, mu, 1 / sqrt(n_eff), log = TRUE) +
            dnorm(mu, 0, sigma_prior, log = TRUE)
        h_step          <- 1 / input$n_pts
        log_jac         <- dnorm(mu_pts, lap_mode, lap_se, log = TRUE)   # log du/dmu
        log_marlik_quad <- log_sum_exp(log_z(mu_pts) - log_jac) + log(h_step)

        par(mar = c(4, 4, 2, 4))
        plot(mu_grid, z(mu_grid), type = "l", lwd = 3, col = "firebrick",
             xlab = expression(mu), ylab = expression(z(mu) == pi(mu) * L(mu)))
        polygon(c(mu_grid, rev(mu_grid)), c(z(mu_grid), rep(0, length(mu_grid))),
                col = adjustcolor("firebrick", 0.15), border = NA)
        segments(mu_pts, 0, mu_pts, z(mu_pts), col = "gray30", lwd = 1.5)
        points(mu_pts, z(mu_pts), pch = 19, col = "gray30")

        par(new = TRUE)
        plot(mu_grid, pnorm(mu_grid, lap_mode, lap_se), type = "l", lwd = 3, col = "steelblue",
             axes = FALSE, xlab = "", ylab = "", ylim = c(0, 1))
        axis(4, col.axis = "steelblue", col.ticks = "steelblue")
        mtext(expression(u == Phi((mu - hat(mu)[L]) / s[L])), side = 4, line = 2.5, col = "steelblue")

        guide_col <- adjustcolor("gray50", 0.6)
        segments(mu_pts, 0, mu_pts, u_pts, col = guide_col, lty = 3)
        segments(mu_pts, u_pts, max(mu_grid), u_pts, col = guide_col, lty = 3)
        points(mu_pts, u_pts, pch = 19, col = "gray30")

        legend("topleft", bty = "n", lwd = c(3, 3, NA, NA), col = c("firebrick", "steelblue", NA, NA),
               legend = c(expression(z(mu) == pi(mu) * L(mu)),
                          expression(u == Phi((mu - hat(mu)[L]) / s[L])),
                          as.expression(bquote("true"~log~P(y) == .(round(log_marlik, 3)))),
                          as.expression(bquote("quadrature, this grid"~log~P(y) == .(round(log_marlik_quad, 3))))))
    })
}

shinyApp(ui, server)
```

**Shinylive App Illustrating Adaptive Gauss–Hermite Quadrature.** Shows how centring and scaling the quadrature grid with the Laplace approximation lets a few grid points capture the posterior wherever it sits. Because $u_i$ is uniform in *Laplace-normalized* space, the grid automatically concentrates where the true posterior mass lives.

:::
:::::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit09_laplace.html#sec-laplace-transform-app>.
:::

## Summary

- The Laplace approximation expands $h=-\log f$ to second order at its
  minimum $\theta_0$ and integrates the resulting Gaussian exactly:
  $\theta\mid\text{data} \approx N_p(\theta_0, H^{-1})$ and
  $\int f(\theta)\,d\theta \approx f(\theta_0)(2\pi)^{p/2}|H|^{-1/2}$.
- It needs only an optimizer and a Hessian — the same machinery already used
  to find the MLE — and its cost does not grow with dimension the way a
  quadrature grid's does.
- BIC is the large-$n$ simplification of the Laplace marginal likelihood,
  dropping $O(1)$ terms that do not depend on the sample size.
- On the running Gaussian marginal-likelihood example, Laplace tracked the
  true value everywhere it was tested, including where quadrature's fixed
  grid broke down and even where naive Monte Carlo's fixed proposal started
  to run short of relevant draws — because Laplace re-optimizes for every
  dataset instead of committing to one grid or one proposal in advance.
- It is deterministic, fast, and accurate to $O(1/n)$, but blind to
  skewness and multimodality; the Monte Carlo methods of the next chapters
  remove these limitations at higher computational cost.
