---
title: "Importance Sampling"
format:
  html:
    toc: true
    tbl-colwidths: true
filters:
  - shinylive
editor_options:
  chunk_output_type: console
code-fold: true
---

```{r}
#| label: setup
#| message: false
library(ars)
```

Rejection sampling turns draws from an envelope $g \ge f$ into exact draws from
$f$, but every rejected proposal is wasted, and in the tails of a poor
envelope almost every proposal is rejected. **Importance sampling** keeps
every proposal instead of discarding most of them: draw from an easier
function $g$ and attach a *weight* to each draw that corrects for sampling
from the wrong distribution. The price is that the resulting sample is no
longer an exact iid sample from $f$ — it is a *weighted* sample — but nothing
is ever thrown away, and no envelope condition $g \ge f$ is required.

## The Reweighting Idea

### The goal

As in the last two chapters, $f(\theta)$ is an **unnormalized** function of
interest — typically $f(\theta) = L(\theta)\pi(\theta)$ for a likelihood $L$
and prior $\pi$ — and we want two kinds of integral:
$$
A = E_f\big[a(\theta)\big] = \frac{\int a(\theta)\,f(\theta)\,d\theta}{\int f(\theta)\,d\theta},
\qquad\qquad
C_f = \int f(\theta)\,d\theta,
$$
a posterior expectation and a normalizing constant (the marginal likelihood
when $f = L\pi$). Plain Monte Carlo needs draws from $f/C_f$, which we
generally cannot produce.

Suppose instead we can draw from another function $g(\theta)$ (also possibly
unnormalized, with $C_g = \int g(\theta)\,d\theta$). Wherever $f$ puts mass,
$g$ must too:
$$
\{\theta : g(\theta) > 0\} \ \supseteq\ \{\theta : f(\theta) > 0\}.
$$
Unlike rejection sampling, $g$ need **not** dominate $f$ everywhere — there is
no requirement $g \ge f$. This is what makes importance sampling applicable
in situations where no usable envelope exists.

### Two estimators

Define the **importance weight** $w(\theta) = f(\theta)/g(\theta)$. For any
$a$,
$$
E_f\big[a(\theta)\big]
= \int a(\theta)\,\frac{f(\theta)}{C_f}\,d\theta
= \frac{C_g}{C_f}\, E_g\big[a(\theta)\,w(\theta)\big],
$$
and taking $a \equiv 1$ gives $C_f/C_g = E_g[w(\theta)]$. Combining the two,
$$
E_f\big[a(\theta)\big] = \frac{E_g\big[a(\theta)\,w(\theta)\big]}{E_g\big[w(\theta)\big]}.
$$
Draw $\theta_1,\ldots,\theta_S \overset{iid}{\sim} g/C_g$ and compute
$w_s = w(\theta_s)$. Two estimators follow directly:

- **Ratio of normalizing constants**: $\widehat{C_f/C_g} = \frac1S\sum_s w_s$,
  unbiased for $C_f/C_g$; if $g$ is already a normalized density ($C_g=1$)
  this estimates $C_f$ itself.
- **Self-normalized expectation**: $\hat A = \sum_{s=1}^S a(\theta_s)\,\tilde w_s$,
  with normalized weights $\tilde w_s = w_s / \sum_j w_j$ summing to $1$. This
  turns the proposal sample into a *weighted sample from $f$*: any unknown
  constants in $f$ and $g$ cancel in the ratio, so $\hat A$ needs only $f$ and
  $g$ up to a proportionality constant. It is a ratio of two averages and so
  is biased with bias $O(1/S)$, but consistent.

### Effective sample size

$\hat A$ is a weighted average of $a(\theta_s)$: draws where $f \gg g$ count
more, draws where $f \ll g$ count less. How many *equally weighted* draws is
the weighted sample worth?
$$
S_{\text{eff}} = \frac{1}{\displaystyle\sum_{s=1}^S \tilde w_s^2}.
$$
If the weights are all equal, $\tilde w_s = 1/S$ and $S_{\text{eff}} =
1/(S \cdot 1/S^2) = S$: the proposal is shaped just like the target, and every
draw counts fully. If one weight dominates all the others, $\tilde w_s
\approx 1$ for that draw and $S_{\text{eff}} \approx 1$: only one draw is
doing any work. A variance heuristic makes this precise:
$\text{Var}_g(\hat A) \approx \text{Var}_f(a)/S_{\text{eff}}$, the same form as
for an iid sample of size $S_{\text{eff}}$ from $f$ itself.

The catch, central to this chapter, is that $S_{\text{eff}}$ is computed
**only from the weights actually observed**. A proposal that never visits the
part of $\theta$-space where $f$ and $g$ disagree most will report a
reassuring $S_{\text{eff}}$ while being badly wrong — this is exactly the
subject of the "bad stories" below.

```{r}
#| label: ess-helpers
### effective sample size from raw (possibly unnormalized) weights
ess <- function(w) sum(w)^2 / sum(w^2)
### numerically stable log(sum(exp(lx)))
log_sum_exp <- function(lx) { m <- max(lx); m + log(sum(exp(lx - m))) }
### effective sample size computed from LOG weights (avoids under/overflow)
log_ess <- function(lw) exp(2 * log_sum_exp(lw) - log_sum_exp(2 * lw))
```

## Example 1: Tail Probability of Cauchy {#sec-ars-tnorm}



Write the proposal on the standardized scale $z=(x-\mu)/\sigma$ with
standardized cutoff $a=(c-\mu)/\sigma$, so that
$$
h(z) \;=\; \log\varphi(z) \;=\; -\tfrac{1}{2}z^{2} + \text{const},
\qquad h'(z) = -z , \qquad h''(z) = -1 < 0 .
$$
Restricting $z$ to $(a,\infty)$ intersects the support with a convex set and
therefore preserves concavity of $h$. A single sampler on the $z$ scale
consequently serves every parameter combination: only $a$ changes, and draws
are returned as $x=\mu+\sigma z$.

ARS exploits concavity twice. Tangent lines to $h$ at abscissae
$x_{1}<\dots<x_{k}$ in $(a,\infty)$ lie *above* $h$, so their pointwise
minimum is an **upper hull** $u(z)\ge h(z)$; exponentiating gives a
piecewise-exponential envelope that is trivial to sample by inversion within
each piece. Chords joining adjacent $(x_{j},h(x_{j}))$ lie *below* $h$, so
they supply a **lower hull** $\ell(z)\le h(z)$, the *squeeze*. Each candidate
$z^{*}$ drawn from the envelope is accepted at once if
$\log U \le \ell(z^{*})-u(z^{*})$, and only otherwise is $h(z^{*})$ evaluated;
when it is evaluated, $z^{*}$ is appended to the abscissae and the hull is
rebuilt, so the envelope tightens as sampling proceeds. Two structural
requirements are worth noting. The piece boundaries are the tangent
intersections
$$
z_{j} \;=\;
\frac{h(x_{j+1}) - h(x_{j}) - x_{j+1}h'(x_{j+1}) + x_{j}h'(x_{j})}
     {h'(x_{j}) - h'(x_{j+1})} ,
$$
which for the Gaussian collapse to the midpoints $(x_{j}+x_{j+1})/2$; and the
envelope is integrable on the unbounded right piece only if
$h'(x_{k})<0$, so the largest abscissa must exceed the mode. The lower end of
the domain is the finite cutoff $a$, where no such condition is needed.

The implementation below adapts until the hull reaches `max_hull` points and
then freezes it, after which the remaining draws are generated by ordinary
vectorized rejection from a fixed envelope. Roughly two dozen tangents give
acceptance rates above $0.95$ across the whole parameter range used here.

```{r}
#| label: rtnorm-ars
## ---------------------------------------------------------------------
##  rtnorm_ars(): adaptive rejection sampling from
##      g(x)  proportional to  phi((x - mu) / sigma) * I(x > c)
##  Returns a numeric vector with attributes "hull" (final abscissae, on
##  the original scale) and "acc_rate" (accepted draws per envelope draw).
## ---------------------------------------------------------------------
rtnorm_ars <- function(n, mu = 0, sigma = 1, c = 0,
                       max_hull = 25, batch = 4096) {

    stopifnot(n >= 1, sigma > 0, is.finite(c), max_hull >= 3)

    a  <- (c - mu) / sigma                 # standardized cutoff
    h  <- function(z) -z^2 / 2             # log density, up to a constant
    hp <- function(z) -z                   # its derivative

    ## ---- upper hull: tangents at the abscissae x ---------------------
    ## Piece j is the tangent at x[j], valid on [z[j], z[j+1]], where the
    ## z[] are the tangent intersections, z[1] = a and z[k+1] = Inf.
    build <- function(x) {
        x <- sort(unique(x))
        x[x == 0] <- 1e-8                  # keep every slope non-zero
        x <- sort(unique(x))
        k  <- length(x)
        hx <- h(x)
        sx <- hp(x)

        j  <- seq_len(k - 1)
        ## general ARS intersection formula; for the Gaussian it reduces
        ## to the midpoint (x[j] + x[j+1]) / 2
        zz <- (hx[j + 1] - hx[j] - x[j + 1] * sx[j + 1] + x[j] * sx[j]) /
              (sx[j] - sx[j + 1])
        z  <- c(a, zz, Inf)

        ul <- hx + sx * (z[1:k] - x)       # hull at the left end of piece j
        ur <- hx + sx * (z[2:(k + 1)] - x) # hull at the right end (-Inf on the last)

        ## log mass of each piece, |exp(ul) - exp(ur)| / |slope|, computed
        ## with a per-piece offset so that nothing under- or overflows
        M  <- pmax(ul, ur)
        lp <- M + log(abs(exp(ul - M) - exp(ur - M))) - log(abs(sx))
        lp[!is.finite(lp)] <- -Inf

        list(x = x, hx = hx, sx = sx, z = z, ul = ul, ur = ur, M = M,
             p = exp(lp - max(lp)))
    }

    ## ---- draw m points from the piecewise-exponential envelope -------
    draw <- function(H, m) {
        j  <- sample.int(length(H$x), m, replace = TRUE, prob = H$p)
        v  <- runif(m)
        Mj <- H$M[j]
        el <- exp(H$ul[j] - Mj)
        er <- exp(H$ur[j] - Mj)
        lu <- Mj + log(el + v * (er - el))      # inverse cdf within the piece
        list(z = H$x[j] + (lu - H$hx[j]) / H$sx[j], u = lu)
    }

    ## ---- lower hull (squeeze): chords between adjacent abscissae -----
    squeeze <- function(H, zs) {
        k   <- length(H$x)
        i   <- findInterval(zs, H$x)
        out <- rep(-Inf, length(zs))
        ok  <- i >= 1 & i <= k - 1
        if (any(ok)) {
            ii <- i[ok]; zz <- zs[ok]
            out[ok] <- ((H$x[ii + 1] - zz) * H$hx[ii] +
                        (zz - H$x[ii]) * H$hx[ii + 1]) /
                       (H$x[ii + 1] - H$x[ii])
        }
        out
    }

    ## ---- initial abscissae -------------------------------------------
    ## All must lie in (a, Inf) and the largest must have a negative slope,
    ## so that the final piece of the envelope is integrable.  The local
    ## scale 1 / max(|a|, 1) reflects the near-exponential behaviour of the
    ## Gaussian tail just above a large cutoff.
    s0 <- 1 / max(abs(a), 1)
    H  <- build(c(a + 1e-3 * s0, a + 0.5 * s0, a + 2 * s0, max(a, 0) + 1.5))

    out <- numeric(n); got <- 0L; tries <- 0L

    ## ---- phase 1: adaptive, one draw at a time ------------------------
    while (got < n && length(H$x) < max_hull) {
        d  <- draw(H, 1); tries <- tries + 1L
        lw <- log(runif(1))
        if (lw <= squeeze(H, d$z) - d$u) {          # cheap squeeze test
            got <- got + 1L; out[got] <- d$z
        } else {                                    # full test, then adapt
            if (lw <= h(d$z) - d$u) { got <- got + 1L; out[got] <- d$z }
            H <- build(c(H$x, d$z))
        }
    }

    ## ---- phase 2: envelope frozen, vectorized rejection ---------------
    while (got < n) {
        m    <- max(batch, n - got)
        d    <- draw(H, m); tries <- tries + m
        zs   <- d$z[log(runif(m)) <= h(d$z) - d$u]
        take <- min(length(zs), n - got)
        if (take > 0L) out[got + seq_len(take)] <- zs[seq_len(take)]
        got <- got + take
    }

    structure(mu + sigma * out,
              hull     = mu + sigma * H$x,
              acc_rate = n / tries)
}
```

The contrast with the target is the point of the exercise. ARS requires
log-concavity, and the standard Cauchy is not log-concave:
$$
\bigl(\log f_{\text{Cauchy}}\bigr)''(x) \;=\; \frac{2(x^{2}-1)}{(1+x^{2})^{2}} ,
$$
which is *positive* for $|x|>1$, so $\log f_{\text{Cauchy}}$ is convex in both
tails and no tangent line lies above it there. The hull construction fails
outright. This is not an incidental obstacle: log-concavity forces tails to
decay at least exponentially, which is precisely the property that disqualifies
the normal as a proposal for a Cauchy target. The densities ARS handles most
comfortably are, tail for tail, the ones least able to serve as importance
proposals for heavy-tailed targets.

### The estimator and its variance

With $g$ the normalized truncated normal on $(c,\infty)$ and
$f(x)=f_{\text{Cauchy}}(x)I(x>c)$, the weights are $w(x)=f(x)/g(x)$ and
$$
\widehat{C} \;=\; \frac{1}{S}\sum_{i=1}^{S} w(X_{i}), \qquad X_{i}\sim g ,
$$
is unbiased for $C_{f}=P(X>c)$ for every $(\mu,\sigma)$. Its variance is
$$
\operatorname{Var}_{g}\bigl(w(X)\bigr)
 \;=\; \int_{c}^{\infty}\frac{f_{\text{Cauchy}}(x)^{2}}{g(x)}\,dx \;-\; C_{f}^{2} ,
$$
and since $f_{\text{Cauchy}}(x)^{2}\asymp \pi^{-2}x^{-4}$ while
$g(x)\propto\exp\{-(x-\mu)^{2}/2\sigma^{2}\}$, the integrand grows like
$x^{-4}\exp\{(x-\mu)^{2}/2\sigma^{2}\}$. The integral diverges for **every**
finite $\mu$ and every $\sigma>0$. No choice of proposal parameters repairs the
estimator; the parameters govern only how quickly the damage becomes visible.
Reported standard errors and confidence intervals are estimating a quantity
that does not exist, and the sample standard deviation of the weights, being
finite by construction, gives no warning of this.

### An interactive view

The app below puts $c$, $\mu$ and $\sigma$ under slider control and displays,
on a common $x$-axis: the two normalized truncated densities on a log scale
with a rug of the draws, and the normalized importance weights
$\tilde w_{i}=w_{i}/\sum_{j}w_{j}$ at the sampled points. Because both
densities are normalized, the vertical gap between the curves on the log scale
is $\log w(x)-\log C_{f}$, so the visual divergence of the curves *is* the
growth of the weight function. The region beyond the largest draw is shaded:
the proposal's shape there is never interrogated by the sample, and it is
exactly there that the target carries the mass the estimator is missing. The
legends report the true $P(X>c)$, the estimate with its nominal 95% interval,
and $S_{\text{eff}}=1/\sum_{i}\tilde w_{i}^{2}$.

```{r}
#| label: is-cauchy-app
#| eval: false
library(shiny)

ui <- fluidPage(
    titlePanel("Importance sampling for a Cauchy tail with a truncated-normal proposal"),
    sidebarLayout(
        sidebarPanel(
            width = 3,
            sliderInput("cc", "tail cutoff  c", min = 0, max = 10,
                        value = 2, step = 0.25),
            tags$hr(style = "margin:6px 0"),
            sliderInput("mu", HTML("proposal mean &mu;"), min = -5, max = 20,
                        value = 0, step = 0.25),
            sliderInput("sg", HTML("proposal sd &sigma;"), min = 0.25, max = 10,
                        value = 1, step = 0.25),
            tags$hr(style = "margin:6px 0"),
            selectInput("S", "sample size  S",
                        choices = c(200, 1000, 5000, 20000), selected = 1000),
            numericInput("seed", "seed", value = 1, min = 1, step = 1),
            actionButton("new", "New sample"),
            checkboxInput("logw", "log scale for the weights", FALSE),
            tags$hr(style = "margin:6px 0"),
            htmlOutput("readout")
        ),
        mainPanel(width = 9, plotOutput("panels", height = "680px"))
    )
)

server <- function(input, output) {

    sim <- reactive({
        S  <- as.integer(input$S)
        cc <- input$cc; mu <- input$mu; sg <- input$sg
        set.seed(input$seed + 1000 * input$new)

        X  <- rtnorm_ars(S, mu, sg, cc)

        ## log importance weight  log f_Cauchy(x) - log g(x),  with g the
        ## normalized truncated normal on (c, Inf)
        lZ   <- pnorm(cc, mu, sg, lower.tail = FALSE, log.p = TRUE)
        logw <- dcauchy(X, log = TRUE) - (dnorm(X, mu, sg, log = TRUE) - lZ)

        m    <- max(logw)
        v    <- exp(logw - m)                       # rescaled weights
        Chat <- mean(v) * exp(m)                    # unbiased estimate of P(X > c)
        se   <- sd(v) / sqrt(S) * exp(m)            # nominal standard error
        wn   <- v / sum(v)                          # normalized weights

        list(X = X, wn = wn, S = S, cc = cc, mu = mu, sg = sg,
             Chat = Chat, se = se, ess = 1 / sum(wn^2),
             truth = pcauchy(cc, lower.tail = FALSE),
             acc = attr(X, "acc_rate"))
    })

    output$panels <- renderPlot({
        s  <- sim()
        cc <- s$cc; X <- s$X
        lo <- cc
        hi <- max(max(X) + 0.5, cc + 5)
        xs <- seq(lo, hi, length.out = 800)

        ## both densities normalized on (c, Inf)
        ft <- dcauchy(xs) / s$truth
        gt <- dnorm(xs, s$mu, s$sg) / pnorm(cc, s$mu, s$sg, lower.tail = FALSE)
        yl <- range(c(ft, gt)[c(ft, gt) > 0])
        yl[1] <- max(yl[1], 1e-12)

        lwr <- s$Chat - 1.96 * s$se
        upr <- s$Chat + 1.96 * s$se
        covered <- lwr <= s$truth && s$truth <= upr

        layout(matrix(1:2, 2), heights = c(3, 2))

        ## ---- top panel: the two densities ----------------------------
        par(mar = c(0.4, 5, 3, 1))
        plot(xs, ft, type = "n", log = "y", ylim = yl, xlim = c(lo, hi),
             xaxt = "n", xlab = "", ylab = "density (log scale)",
             main = "target and proposal, both normalized on (c, Inf)")
        rect(max(X), yl[1], hi, yl[2] * 10,
             col = adjustcolor("grey60", 0.18), border = NA)
        lines(xs, ft, lwd = 4, col = "firebrick")
        lines(xs, gt, lwd = 3, col = "steelblue")
        rug(X, col = adjustcolor("steelblue", 0.25))
        legend("topright", bty = "n", cex = 1.05,
               lwd = c(4, 3, NA, NA, NA),
               col = c("firebrick", "steelblue", NA, NA, NA),
               legend = c(
                   "target: Cauchy, truncated at c",
                   sprintf("proposal: N(%.2f, %.2f^2), truncated at c", s$mu, s$sg),
                   sprintf("true  P(X > c) = %.5g", s$truth),
                   sprintf("est.  P(X > c) = %.5g", s$Chat),
                   sprintf("95%% CI = (%.5g, %.5g)%s", lwr, upr,
                           if (covered) "" else "   << misses the truth")))
        text(max(X), yl[2], "  no draw ever reached here", adj = c(0, 1.4),
             col = "grey35", cex = 0.95)

        ## ---- bottom panel: normalized weights ------------------------
        par(mar = c(4.5, 5, 0.4, 1))
        wn <- s$wn
        if (input$logw) {
            wl <- c(max(min(wn[wn > 0]), 1e-12), max(wn))
            plot(X, wn, type = "n", log = "y", ylim = wl, xlim = c(lo, hi),
                 xlab = "x", ylab = "normalized weight (log scale)")
            segments(X, wl[1], X, wn, col = adjustcolor("grey25", 0.5))
        } else {
            plot(X, wn, type = "n", ylim = c(0, max(wn) * 1.08),
                 xlim = c(lo, hi), xlab = "x", ylab = "normalized weight")
            segments(X, 0, X, wn, col = adjustcolor("grey25", 0.5))
        }
        abline(h = 1 / s$S, lty = 2, col = "darkgreen", lwd = 2)
        legend("topleft", bty = "n", cex = 1.05,
               lty = c(2, NA, NA), lwd = c(2, NA, NA),
               col = c("darkgreen", NA, NA),
               legend = c("1 / S  (perfectly balanced weights)",
                          sprintf("ESS = %.1f   (%.2f%% of S = %d)",
                                  s$ess, 100 * s$ess / s$S, s$S),
                          sprintf("largest weight = %.1f%% of the total",
                                  100 * max(wn))))
    })

    output$readout <- renderUI({
        s   <- sim()
        lwr <- s$Chat - 1.96 * s$se; upr <- s$Chat + 1.96 * s$se
        ok  <- lwr <= s$truth && s$truth <= upr
        HTML(sprintf(
            "<b>true</b> %.6g<br><b>estimate</b> %.6g<br>
             <b>95%% CI</b> (%.6g, %.6g)<br>
             <b>covers?</b> <span style='color:%s'>%s</span><br>
             <b>ESS</b> %.1f (%.2f%%)<br>
             <b>ARS acceptance</b> %.3f",
            s$truth, s$Chat, lwr, upr,
            if (ok) "green" else "red", if (ok) "yes" else "no",
            s$ess, 100 * s$ess / s$S, s$acc))
    })
}

shinyApp(ui, server)
```

The chunk is marked `eval: false` so that the document renders statically. To
run the app, execute the `rtnorm-ars` chunk and then the `is-cauchy-app` chunk
interactively in RStudio or Positron; `shinyApp()` opens it in the viewer pane.

Two alternatives make the app live in the rendered document. Adding
`server: shiny` to the YAML header turns the document into a Shiny document:
the interface elements are then written as ordinary chunks, the reactive code
moves into a chunk marked `#| context: server`, the function definition moves
into one marked `#| context: setup`, and the document is served with
`quarto serve`. Alternatively the `quarto-shinylive` extension compiles the app
to WebAssembly and embeds it in a self-contained HTML file with no server, which
suits distribution to students; in that case the chunk is marked
`{shinylive-r}` with `#| standalone: true`, and the body of `rtnorm_ars()` must
be pasted inside the app chunk, since a shinylive chunk cannot see objects
defined elsewhere in the document.

Several settings repay attention.

- **Small $\sigma$, or $\mu$ far below $c$.** The draws pile up just above the
  cutoff, the weights are nearly equal among them, and $S_{\text{eff}}$ is
  close to $S$ — while the estimate sits far below the truth and the interval
  excludes it. This is the diagnostic failing in its most misleading form:
  uniform weights certify that the proposal is internally consistent, not that
  it has found the target's mass.
- **Large $\sigma$.** Draws occasionally reach past the crossing of the two
  curves, one weight then dominates the sum, $S_{\text{eff}}$ collapses toward
  $1$, and the estimate jumps. The estimator is now visibly unreliable, which
  is an improvement on being invisibly unreliable.
- **Repeated resampling at fixed $S$.** Pressing *New sample* varies the
  estimate far more than the reported standard error admits, and the nominal
  95% interval misses the truth far more often than 5% of the time.
- **Raising $S$ a hundredfold.** The nominal standard error falls by a factor
  of ten, the actual spread across resamples does not, and $S_{\text{eff}}$ as a
  *percentage* of $S$ deteriorates rather than improving.
- **Raising $c$.** The mismatch worsens monotonically, since the polynomial and
  Gaussian tails separate faster the further out one starts.


## Example 2: Marginal Likelihood via Importance Sampling

### Setup

The [quadrature](unit09_integral.qmd) and Laplace-approximation chapters
compute the marginal likelihood $P(y) = \int L(\theta)\pi(\theta)\,d\theta$ of
the same conjugate-free Gaussian model — unknown mean $\mu$ and
$w=\log\sigma^2$, independent normal priors — by a product grid and by a
mode-and-Hessian approximation, respectively. Importance sampling offers a
third route, and one that comes with a built-in accuracy diagnostic
($S_{\text{eff}}$) that neither of the other two methods provides directly.

Take $g(\mu,w)$ to be the **Gaussian approximation to the posterior**
obtained by a Laplace fit: locate the posterior mode $(\hat\mu,\hat w)$ by
optimization, take $g = N_2\big((\hat\mu,\hat w),\, H^{-1}\big)$ for $H$ the
Hessian of $-\log[L\pi]$ at the mode, and importance-sample from $g$ with
$f = L\pi$:
$$
\widehat{P(y)} = \frac{1}{S}\sum_{s=1}^S w_s, \qquad
w_s = \frac{L(\theta_s)\pi(\theta_s)}{g(\theta_s)}, \qquad \theta_s \sim g.
$$
Because $g$ is shaped like the posterior near its mode, this is a "good
story" proposal by construction — in sharp contrast to sampling directly from
the (typically much wider, uninformative) prior, which is exactly the
`log_marlik_mc` function used elsewhere in this book as a naive-Monte-Carlo
check: it is importance sampling with $g = \pi$, and its weights are just the
likelihood values $L(\theta_s)$.

### An efficient implementation

```{r}
#| label: marlik-functions
log_lik   <- function(x, mu, w) sum(dnorm(x, mu, exp(w / 2), log = TRUE))
log_prior <- function(mu, w, mu_0, sigma_mu, w_0, sigma_w) {
    dnorm(mu, mu_0, sigma_mu, log = TRUE) + dnorm(w, w_0, sigma_w, log = TRUE)
}
neg_log_post <- function(theta, x, mu_0, sigma_mu, w_0, sigma_w) {
    -log_lik(x, theta[1], theta[2]) -
        log_prior(theta[1], theta[2], mu_0, sigma_mu, w_0, sigma_w)
}

### log density of N(mu, A^{-1}) at every column of a 2 x S matrix of points at once,
### replacing a per-point apply() call with one vectorized quadratic form
log_dmvnorm_batch <- function(Theta, mu, A) {
    Tc <- Theta - mu
    quad <- colSums((A %*% Tc) * Tc)
    0.5 * (-nrow(Theta) * log(2 * pi) + sum(log(svd(A)$d)) - quad)
}

### log marginal likelihood by importance sampling from a Laplace-shaped Gaussian
log_marlik_imps <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    fit <- nlm(neg_log_post, c(mean(x), log(stats::var(x))), hessian = TRUE,
               x = x, mu_0 = mu_0, sigma_mu = sigma_mu, w_0 = w_0, sigma_w = sigma_w)
    mu_hat <- fit$estimate; H <- fit$hessian
    L <- t(chol(solve(H)))                        # Sigma = H^{-1} = L L'
    Theta <- L %*% matrix(rnorm(2 * S), 2, S) + mu_hat

    log_g <- log_dmvnorm_batch(Theta, mu_hat, H)
    log_f <- -apply(Theta, 2, neg_log_post, x = x, mu_0 = mu_0, sigma_mu = sigma_mu,
                     w_0 = w_0, sigma_w = sigma_w)
    lw <- log_f - log_g
    structure(log_sum_exp(lw) - log(S), ess = log_ess(lw))
}

### log marginal likelihood by naive Monte Carlo from the prior (g = pi): a
### second, independent importance-sampling estimator with an unfavourable g
log_marlik_prior <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    mus <- rnorm(S, mu_0, sigma_mu); ws <- rnorm(S, w_0, sigma_w)
    lw <- vapply(seq_len(S), function(i) log_lik(x, mus[i], ws[i]), numeric(1))
    structure(log_sum_exp(lw) - log(S), ess = log_ess(lw))
}
```

Replacing the two `apply()` calls used in earlier drafts of this code — one
for the proposal log-density, one for the log-posterior — with a single
vectorized quadratic form (`log_dmvnorm_batch`) removes $S$ separate R
function calls for the proposal evaluation; only the data-dependent
log-likelihood still needs one call per draw, since it sums over the $n$
observations for each $\theta_s$ separately.

### Checking accuracy

```{r}
#| label: tbl-marlik-check
#| tbl-cap: "Log marginal likelihood by importance sampling (Laplace-shaped proposal, S = 200) across 500 replications, for two simulated data sets."
set.seed(4)
x_small <- rnorm(10);  x_large <- rnorm(200)

check_tbl <- do.call(rbind, lapply(list(n10 = x_small, n200 = x_large), function(x) {
    reps <- replicate(500, log_marlik_imps(x, 0, 5, 0, 5, 200), simplify = FALSE)
    vals <- vapply(reps, as.numeric, numeric(1))
    esss <- vapply(reps, attr, numeric(1), which = "ess")
    data.frame(mean_log_ml = mean(vals), sd_log_ml = sd(vals), mean_ess = mean(esss))
}))
check_tbl <- cbind(dataset = c("n = 10", "n = 200"), check_tbl)
knitr::kable(check_tbl, digits = 4)
```

The Monte Carlo standard deviation across replications is small relative to
the estimate itself, and the effective sample size stays close to the nominal
$S=200$ in both cases — evidence that the Laplace-shaped proposal tracks the
posterior well even though it is, after all, only a second-order
approximation to it.

### Why the proposal matters: Laplace-shaped versus prior sampling

The prior is a valid proposal for the same target — every posterior draw is
also a prior draw with positive density — so `log_marlik_prior` is a
legitimate importance-sampling estimator too. But the prior is far more
diffuse than the posterior, especially once $n$ is not tiny, so it is a much
poorer match to $f = L\pi$.

```{r}
#| label: tbl-marlik-ess-vs-n
#| tbl-cap: "Effective sample size (of S = 500 draws) for the Laplace-shaped proposal versus prior sampling, as the sample size n of the simulated data grows. Both target the same posterior for (mu, w)."
set.seed(5)
ns <- c(5, 20, 100, 400)
ess_tbl <- do.call(rbind, lapply(ns, function(n) {
    x <- rnorm(n)
    data.frame(n = n,
               ess_laplace_proposal = attr(log_marlik_imps(x, 0, 5, 0, 5, 500), "ess"),
               ess_prior_sampling   = attr(log_marlik_prior(x, 0, 5, 0, 5, 500), "ess"))
}))
knitr::kable(ess_tbl, digits = c(0, 1, 1))
```

As $n$ grows the posterior sharpens around the truth while the prior does
not, so prior sampling's effective sample size deteriorates steadily; the
Laplace-shaped proposal, which re-centers and re-scales itself around the
posterior mode for every data set, keeps pace. This is the marginal-likelihood
instance of the same lesson as the missed-mode example above: a fixed,
data-independent proposal becomes an increasingly bad match to a target that
concentrates as more data arrive, and only a proposal that adapts to the
target avoids the resulting collapse in efficiency.

## Summary

- Importance sampling reweights draws from a proposal $g$ by
  $w(\theta) = f(\theta)/g(\theta)$ instead of accepting or rejecting them,
  using $\widehat{C_f/C_g} = \frac1S\sum_s w_s$ and the self-normalized
  $\hat A = \sum_s a(\theta_s)\tilde w_s$; only the support condition
  $\{g>0\}\supseteq\{f>0\}$ is required, not an envelope $g \ge f$.
- A good proposal shaped like the target on its support gives near-constant
  weights and $S_{\text{eff}} \approx S$ (the truncated-normal and
  Laplace-proposal examples), and a single weighted sample can answer several
  questions about the same target — a probability and an expectation alike.
- A proposal that misses a mode or part of the support fails **silently**: the
  observed weights can look perfectly healthy while both $C_f/C_g$ and
  $E_f[a]$ are badly wrong, and no amount of extra sampling fixes it.
  $S_{\text{eff}}$ diagnoses only the noisy failure (wildly varying observed
  weights), never the silent one.
- For the marginal likelihood, a proposal built from a Laplace approximation
  at the posterior mode remains efficient as the data accumulate and the
  posterior sharpens, while a fixed, data-independent proposal such as the
  prior does not — reinforcing why the mode-finding methods from earlier
  chapters remain useful even once we turn to sampling-based methods.
