---
title: "Multivariate Optimization Methods for Finding MLE"
format:
  html:
    code-fold: show
    toc: true
    tbl-colwidths: true
filters:
  - shinylive
---

In the previous chapter the parameter was a scalar, and the log-likelihood could
be inspected by plotting it. Once $\theta \in \mathbb{R}^p$ neither is true: we
cannot see the surface, and the notion of "bracketing a root" that underlies
bisection and golden section has no direct multivariate analogue. What survives
is the local quadratic approximation, and the Newton–Raphson algorithm is its
direct consequence.

This chapter develops the multivariate Newton–Raphson method and its
statistical variant, Fisher scoring, applies both to logistic regression,
examines interactively how and why the iteration can diverge, and closes with a
survey of the derivative-free and gradient-only alternatives.

## The Newton–Raphson Algorithm in Several Dimensions

### Derivation from a quadratic approximation

Write the log-likelihood as $\ell(\theta)$ for $\theta \in \mathbb{R}^p$, and
define

$$
S(\theta) = \frac{\partial \ell(\theta)}{\partial \theta}
\quad (p \times 1, \text{ the score}),
\qquad
H(\theta) = \frac{\partial^2 \ell(\theta)}{\partial \theta \, \partial \theta^{\top}}
\quad (p \times p, \text{ the Hessian}).
$$

The negative Hessian $J(\theta) = -H(\theta)$ is the **observed information
matrix**.

Expand $\ell$ to second order about the current iterate $\theta^{(k)}$:

$$
\ell(\theta) \;\approx\; \ell(\theta^{(k)})
 + S(\theta^{(k)})^{\top}(\theta - \theta^{(k)})
 - \tfrac{1}{2}\,(\theta - \theta^{(k)})^{\top} J(\theta^{(k)}) \,(\theta - \theta^{(k)}).
$$

This is a quadratic in $\theta$. If $J(\theta^{(k)})$ is positive definite the
quadratic has a unique maximum, found by setting its derivative to zero:

$$
S(\theta^{(k)}) - J(\theta^{(k)})(\theta - \theta^{(k)}) = 0 .
$$

Solving for $\theta$ gives the **Newton–Raphson update**:

$$
\boxed{\;\theta^{(k+1)} = \theta^{(k)} + J(\theta^{(k)})^{-1} S(\theta^{(k)}) \;}
$$

The algorithm therefore replaces the log-likelihood by the quadratic that
matches its value, slope and curvature at the current point, and jumps to the
maximum of that quadratic. Everything that follows — the fast convergence, and
the modes of failure — comes from how good or bad that replacement is.

Three remarks are worth making explicit.

*The inverse is never formed.* Write the update as
$\theta^{(k+1)} = \theta^{(k)} + \delta^{(k)}$ where $\delta^{(k)}$ solves the
linear system $J(\theta^{(k)})\,\delta = S(\theta^{(k)})$. In R this is
`solve(J, S)`, not `solve(J) %*% S`. The former is faster and numerically more
stable; the latter computes $p^2$ numbers in order to use $p$ of them.

*The step is not the gradient direction.* $\delta^{(k)} = J^{-1}S$ points along
the gradient only when $J$ is a multiple of the identity. The information matrix
rescales and rotates the ascent direction to account for differing curvature and
correlation among the parameters, which is exactly why Newton–Raphson is
insensitive to the scaling of the parameters while gradient descent is not.

*Positive definiteness is an assumption, not a guarantee.* Away from the maximum
$J(\theta^{(k)})$ can be indefinite or singular, in which case the "maximum of
the quadratic" is a saddle point or does not exist, and $\delta^{(k)}$ may point
downhill. @sec-divergence examines this.

#### Fisher scoring

The observed information $J(\theta)$ depends on the data. Its expectation under
the model,

$$
I(\theta) = E_{\theta}\{J(\theta)\} = -E_{\theta}\left\{ \frac{\partial^2 \ell(\theta)}{\partial \theta \, \partial \theta^{\top}} \right\},
$$

is the **expected (Fisher) information matrix**. Replacing $J$ by $I$ in the
update gives the **method of scoring**:

$$
\theta^{(k+1)} = \theta^{(k)} + I(\theta^{(k)})^{-1} S(\theta^{(k)}).
$$

The two algorithms have different iterates but the same fixed point, since both
stop where $S(\theta) = 0$. The practical differences:

| | Newton–Raphson | Fisher scoring |
|---|---|---|
| Curvature matrix | Observed information $J(\theta)$ | Expected information $I(\theta)$ |
| Derivation | Second derivative of $\ell$ | Expectation of that second derivative |
| Positive definite? | Only guaranteed near the maximum | Always, wherever the model is identified |
| Algebra | Can be messy | Often simpler, as data-dependent terms average out |
| Local convergence | Quadratic | Linear, but typically with a very small rate constant |
| Stability from poor starts | Less stable | More stable |

: Newton–Raphson compared with Fisher scoring {#tbl-nr-scoring}

The guaranteed positive definiteness of $I(\theta)$ is the main practical
argument for scoring: the update direction is always an ascent direction, so the
iteration cannot be sent uphill-seeking in a downhill direction by an
ill-behaved Hessian.

For generalised linear models with the **canonical link** the two coincide
exactly, because the second derivative of the log-likelihood does not involve
the response. Logistic regression uses the canonical logit link, so the example
below is simultaneously Newton–Raphson, Fisher scoring, and — as we note at the
end of the section — iteratively reweighted least squares.

### Convergence properties

Near a maximum $\hat\theta$ at which $J(\hat\theta)$ is positive definite,
Newton–Raphson converges **quadratically**: the error satisfies

$$
\|\theta^{(k+1)} - \hat\theta\| \;\le\; C \,\|\theta^{(k)} - \hat\theta\|^{2}
$$

for some constant $C$ depending on the third derivatives. Informally, the number
of correct digits doubles at each iteration, which is why five or six iterations
are usually enough and why a well-started run appears to converge instantly.

The qualification "near" is essential. The theorem is local: it guarantees
nothing about a start far from $\hat\theta$, and nothing at all if $J$ fails to
be positive definite along the way.

### Standard errors as a by-product

At convergence the same matrix that drove the iteration supplies the asymptotic
variance. Under the usual regularity conditions,

$$
\hat\theta \;\dot\sim\; N\!\left(\theta, \; J(\hat\theta)^{-1}\right),
$$

so $\operatorname{SE}(\hat\theta_j) = \sqrt{[J(\hat\theta)^{-1}]_{jj}}$. This is
a genuine practical advantage of Newton-type methods over derivative-free ones:
the curvature information needed for inference has already been computed. Methods
that never form a Hessian require a separate numerical differentiation step, which
is what `hessian = TRUE` requests from `nlm` and `optim`.

## Worked Example: Logistic Regression

### The model and its derivatives

With a single covariate $z_i$ and binary response $r_i$, let
$u_i = \beta_0 + \beta_1 z_i$ and $p_i = e^{u_i}/(1 + e^{u_i})$. The
log-likelihood is

$$
\ell(\beta) = \sum_{i=1}^{n}\left\{ r_i u_i - \log\left(1 + e^{u_i}\right) \right\}.
$$

Writing $X$ for the $n \times 2$ design matrix with rows $(1, z_i)$, the score
and information are

$$
S(\beta) = X^{\top}(r - p), \qquad
J(\beta) = X^{\top} W X, \qquad W = \operatorname{diag}\{p_i(1 - p_i)\}.
$$

Note that $J$ does not involve $r$. This is the canonical-link property
mentioned above, and it means $J(\beta) = I(\beta)$ identically.

Two numerical points, both carried over from the chapter on computer arithmetic.
First, $\log(1 + e^{u})$ overflows for $u \gtrsim 710$; the stable form is
$\log(1+e^u) = \max(u,0) + \log(1 + e^{-|u|})$. Second, $p_i(1-p_i) \to 0$ when
$|u_i|$ is large, so $W$ — and hence $J$ — degenerates toward singularity when
the fitted probabilities are extreme. That observation is the whole explanation
of the divergence we are about to see.

### Implementation

```{r}
#| label: logistic-fns
#| message: false

### numerically stable log(1 + exp(u))
log1pexp <- function(u) pmax(u, 0) + log1p(exp(-abs(u)))

### negative log-likelihood, negative score, and information matrix
lik_score_info <- function(b, r, z) {
    u <- b[1] + b[2] * z
    p <- plogis(u)
    w <- p * (1 - p)
    list(
        neg_loglik = -sum(u * r - log1pexp(u)),
        neg_score  = -c(sum(r - p), sum(z * (r - p))),
        inf        = matrix(c(sum(w), sum(z * w),
                              sum(z * w), sum(z * z * w)), 2, 2)
    )
}

neg_logp_logistic <- function(b, z, r) {
    u <- b[1] + b[2] * z
    -sum(u * r - log1pexp(u))
}
```

Because we minimise the *negative* log-likelihood, the update
$\beta^{(k+1)} = \beta^{(k)} + J^{-1}S$ becomes
$\beta^{(k+1)} = \beta^{(k)} - J^{-1} \times (\text{negative score})$, which is
the form coded below.

```{r}
#| label: logistic-nr

mle_logistic_nr <- function(b0, no_iter, r, z, debug = FALSE) {
    out <- matrix(NA_real_, no_iter + 1, 3,
                  dimnames = list(NULL, c("beta0", "beta1", "neg_loglike")))
    b <- b0
    for (i in seq_len(no_iter + 1)) {
        q <- lik_score_info(b, r, z)
        out[i, ] <- c(b, q$neg_loglik)
        if (debug) print(out[i, ])
        if (i > no_iter) break
        step <- tryCatch(solve(q$inf, q$neg_score),
                         error = function(e) rep(NA_real_, 2))
        if (any(!is.finite(step))) break
        b <- b - step
    }
    out
}
```

```{r}
#| label: logistic-data
#| fig-cap: "Simulated logistic data: true success probability and observed responses."

gen_logistic_data <- function(b, n) {
    z <- sort(runif(n, -2, 2))
    p <- plogis(b[1] + b[2] * z)
    r <- rbinom(n, 1, p)
    plot(z, p, type = "l", ylim = c(0, 1), ylab = "P(r = 1)")
    points(z, r, col = r + 1)
    list(z = z, r = r)
}

set.seed(1)
data <- gen_logistic_data(c(0, 1.5), 200)
```

A start close to the truth converges in a handful of iterations:

```{r}
#| label: logistic-nr-good
round(mle_logistic_nr(c(0, 3), 8, data$r, data$z), 6)
```

The doubling of correct digits predicted by the quadratic-convergence result is
visible in the final rows.

### Comparison with `nlm` and `glm`

```{r}
#| label: logistic-nlm
logit_nlm <- nlm(neg_logp_logistic, c(0, 0),
                 z = data$z, r = data$r, hessian = TRUE)
logit_nlm$estimate
sqrt(diag(solve(logit_nlm$hessian)))   # asymptotic standard errors
```

```{r}
#| label: logistic-glm
summary(glm(r ~ z, family = binomial(), data = data))$coefficients
```

The three routes agree to printing accuracy. `glm` is doing the same arithmetic
as `mle_logistic_nr`: for a GLM the Newton–Raphson step can be rearranged as a
weighted least-squares fit of the working response
$u_i + (r_i - p_i)/\{p_i(1-p_i)\}$ on $X$ with weights $p_i(1-p_i)$, giving
**iteratively reweighted least squares (IRLS)**. This is not a different
algorithm but a different arrangement of the same one, and it explains why
`glm` reports "Fisher Scoring iterations" in its output.

## Divergence {#sec-divergence}

### What goes wrong

The quadratic approximation is trustworthy only in a neighbourhood of the
current iterate. Starting from a large $|\beta_1|$, nearly all fitted
probabilities are pushed to $0$ or $1$, so the weights $p_i(1-p_i)$ are
essentially zero, $J$ is nearly singular, and $J^{-1}$ is enormous. The
resulting step is huge, the new iterate lands far outside the region where the
quadratic was a reasonable fit, and the process can oscillate or run off to
infinity — even though the logistic log-likelihood is globally concave and has a
unique maximum.

The interactive figure below lets you choose a starting value and step through
the iterations. Press the play button on the iteration slider to animate the
path over the contours of the negative log-likelihood.

```{shinylive-r}
#| label: nr-divergence-app
#| standalone: true
#| viewerHeight: 640

library(shiny)

### ---------- fixed data ----------------------------------------------------
set.seed(1)
n <- 200
z <- sort(runif(n, -2, 2))
r <- rbinom(n, 1, plogis(0 + 1.5 * z))

log1pexp <- function(u) pmax(u, 0) + log1p(exp(-abs(u)))
negll <- function(b) {
    u <- b[1] + b[2] * z
    -sum(u * r - log1pexp(u))
}

### ---------- contour grid (computed once) ----------------------------------
B0 <- seq(-3, 3, by = 0.1)
B1 <- seq(-8, 8, by = 0.2)
Z <- outer(B0, B1, Vectorize(function(a, b) negll(c(a, b))))

### ---------- Newton-Raphson path -------------------------------------------
nr_path <- function(b0, maxit = 15, halving = FALSE) {
    P <- matrix(NA_real_, maxit + 1, 3,
                dimnames = list(NULL, c("beta0", "beta1", "negll")))
    b <- b0
    P[1, ] <- c(b, negll(b))
    for (i in seq_len(maxit)) {
        u <- b[1] + b[2] * z
        p <- plogis(u)
        w <- p * (1 - p)
        g <- -c(sum(r - p), sum(z * (r - p)))
        J <- matrix(c(sum(w), sum(z * w), sum(z * w), sum(z * z * w)), 2, 2)
        step <- tryCatch(solve(J, g), error = function(e) rep(NA_real_, 2))
        if (any(!is.finite(step))) break
        if (halving) {
            f0 <- P[i, 3]; s <- 1
            repeat {
                bn <- b - s * step
                fn <- negll(bn)
                if ((is.finite(fn) && fn <= f0) || s < 1e-6) break
                s <- s / 2
            }
            b <- bn
        } else {
            b <- b - step
        }
        if (any(!is.finite(b))) break
        P[i + 1, ] <- c(b, negll(b))
    }
    P
}

### ---------- UI -------------------------------------------------------------
ui <- fluidPage(
    titlePanel(NULL),
    sidebarLayout(
        sidebarPanel(
            width = 4,
            numericInput("b0", HTML("starting &beta;<sub>0</sub>"),
                         value = 0, step = 0.5),
            numericInput("b1", HTML("starting &beta;<sub>1</sub>"),
                         value = 5, step = 0.5),
            checkboxInput("halve", "apply step halving", FALSE),
            sliderInput("iter", "iteration", min = 0, max = 15, value = 0,
                        step = 1, animate = animationOptions(interval = 900)),
            helpText("Try (0, 3) and then (0, 5). The second start makes the ",
                     "information matrix nearly singular, so the first step ",
                     "overshoots badly."),
            tableOutput("tab")
        ),
        mainPanel(
            width = 8,
            plotOutput("contour", height = "380px"),
            plotOutput("trace", height = "200px")
        )
    )
)

### ---------- server ---------------------------------------------------------
server <- function(input, output, session) {

    path <- reactive({
        nr_path(c(input$b0, input$b1), maxit = 15, halving = input$halve)
    })

    shown <- reactive({
        P <- path()
        k <- min(input$iter + 1, nrow(P))
        P[seq_len(k), , drop = FALSE]
    })

    output$contour <- renderPlot({
        P <- shown()
        par(mar = c(4, 4, 2, 1))
        contour(B0, B1, Z, nlevels = 40, drawlabels = FALSE,
                col = "grey70",
                xlab = expression(beta[0]), ylab = expression(beta[1]),
                main = "negative log-likelihood surface and NR path")
        ok <- complete.cases(P)
        if (sum(ok) > 1) {
            lines(P[ok, 1], P[ok, 2], col = "steelblue", lwd = 2)
            points(P[ok, 1], P[ok, 2], pch = 21, bg = "steelblue", cex = 1.1)
        }
        last <- P[max(which(complete.cases(P))), , drop = FALSE]
        points(last[1], last[2], pch = 21, bg = "firebrick", cex = 1.8)
        outside <- last[1] < min(B0) || last[1] > max(B0) ||
                   last[2] < min(B1) || last[2] > max(B1)
        if (isTRUE(outside)) {
            mtext(sprintf("current iterate (%.1f, %.1f) is off the plotted region",
                          last[1], last[2]), side = 3, line = -1.5,
                  col = "firebrick", cex = 0.9)
        }
        points(0, 1.5, pch = 3, col = "darkgreen", lwd = 2)
    })

    output$trace <- renderPlot({
        P <- path()
        k <- min(input$iter + 1, nrow(P))
        par(mar = c(4, 4, 1, 1))
        yl <- range(P[, 3], na.rm = TRUE)
        plot(0:(nrow(P) - 1), P[, 3], type = "n", ylim = yl,
             xlab = "iteration", ylab = "negative log-likelihood")
        lines(0:(k - 1), P[seq_len(k), 3], type = "b", pch = 19,
              col = "steelblue")
    })

    output$tab <- renderTable({
        P <- shown()
        d <- as.data.frame(P)
        d$iteration <- seq_len(nrow(d)) - 1
        d <- d[, c("iteration", "beta0", "beta1", "negll")]
        tail(d, 5)
    }, digits = 4, rownames = FALSE)
}

shinyApp(ui, server)
```

::: {.callout-note}
### Reproducing this figure
The chunk above uses the [shinylive](https://github.com/quarto-ext/shinylive)
extension, which compiles the app to WebAssembly so that it runs in the reader's
browser with no Shiny server. Install it once per project with

```bash
quarto add quarto-ext/shinylive
```

and add `filters: [shinylive]` to the document or project YAML. Only packages
available in webR may be used inside the app, so the code above is restricted to
base R and `shiny`.
:::

### Safeguards

The divergence is not a defect of the Newton direction but of the Newton *step
length*. The standard remedies all keep the direction and control the distance.

**Step halving (backtracking).** Take $\delta^{(k)}$ as usual but move
$\theta^{(k+1)} = \theta^{(k)} + s\,\delta^{(k)}$, starting from $s = 1$ and
halving until the log-likelihood actually increases. This costs a few extra
function evaluations, never damages quadratic convergence near the maximum
(where $s = 1$ is accepted on the first try), and rescues most divergent runs.
Toggle it in the app above to see the effect.

**Ridging (Levenberg–Marquardt).** Replace $J$ by $J + \lambda \mathbb{I}$ with
$\lambda > 0$. Large $\lambda$ shrinks the step toward a small gradient step;
$\lambda \to 0$ recovers Newton. This also repairs indefiniteness, and is the
approach `nlm` takes internally.

**Trust regions.** Maximise the quadratic approximation subject to
$\|\theta - \theta^{(k)}\| \le \Delta_k$, adapting $\Delta_k$ according to how
well the quadratic predicted the realised change. This is `nlminb` and the
`PORT` routines.

**Reparameterisation and good starts.** Often the cheapest fix. Centring and
scaling covariates improves conditioning of $J$ substantially, and for GLMs
the null-model fit is a reliable starting point — which is what `glm` uses when
`start` is not supplied.

## Other Multivariate Optimization Methods

Newton–Raphson is the method of choice when the second derivatives are available
and $p$ is modest. Outside that regime, three considerations push toward
alternatives: derivatives may be unavailable or painful to code; the $O(p^3)$
cost of solving the Newton system and the $O(p^2)$ storage of $J$ become
prohibitive for large $p$; and the objective may be non-smooth, noisy or
multimodal.

### Classification by derivative information

Methods are most usefully organised by what they require of the objective.

| Order | Requires | Methods | Typical use |
|---|---|---|---|
| 0 | Function values only | Nelder–Mead, simulated annealing, genetic algorithms | Non-smooth, discontinuous, or awkward-to-differentiate objectives; small $p$ |
| 1 | Gradient | Steepest descent, conjugate gradient, quasi-Newton (BFGS, L-BFGS), stochastic gradient descent | The workhorse class; large $p$; gradients cheap by automatic differentiation |
| 2 | Gradient and Hessian | Newton–Raphson, Fisher scoring, IRLS, trust-region Newton | Statistical models of modest dimension where the Hessian is also wanted for inference |

: Optimization methods classified by the derivative information required {#tbl-orders}

### The main alternatives

**Nelder–Mead (simplex).** Maintains $p+1$ points forming a simplex and moves it
downhill by reflecting, expanding, contracting and shrinking. It needs no
derivatives and tolerates mild non-smoothness, but converges slowly, degrades
badly as $p$ grows past roughly 10–20, and can stall at non-stationary points.
It is `optim`'s default, which makes it by far the most-used and most-overused
method in R. Reasonable as a first look at an unfamiliar objective; a poor
choice for a likelihood you will fit thousands of times.

**Steepest descent with line search.** Move along $-\nabla f$, choosing the
distance by a one-dimensional search. Conceptually the simplest gradient method
and the one that best illustrates the role of a line search, but the
convergence is only linear, and the rate deteriorates with the condition number
of the Hessian: on an elongated valley the iterates zig-zag across it rather
than travelling along it. Rarely used unmodified.

**Conjugate gradient.** Corrects the zig-zag by choosing each direction to be
conjugate to the previous ones with respect to the local curvature, without ever
forming or storing a Hessian. Storage is $O(p)$, which is its defining
advantage. Fletcher–Reeves and Polak–Ribière differ only in the formula for the
correction coefficient. The method is sensitive to the accuracy of the line
search and usually needs periodic restarting.

**Quasi-Newton (BFGS and L-BFGS).** Builds an approximation to $J^{-1}$ from
successive gradient differences, so it attains superlinear convergence with only
first derivatives. BFGS stores a full $p \times p$ approximation; L-BFGS keeps
only the last few update vectors, reducing storage to $O(mp)$ and making it the
default choice for high-dimensional smooth problems. For most likelihood
problems where the Hessian is awkward to derive, BFGS with an analytic gradient
is the best practical compromise.

**Stochastic gradient descent.** For an objective that is a sum over $n$
observations, estimate the gradient from a small random subset ("minibatch") at
each step. The per-iteration cost becomes independent of $n$, at the price of a
noisy trajectory requiring a decreasing step-size sequence
($\sum_k \alpha_k = \infty$, $\sum_k \alpha_k^2 < \infty$). Modern variants —
momentum, AdaGrad, RMSProp, Adam — adapt the step size per coordinate. SGD is
the foundation of neural-network fitting and is treated in its own right in a
later chapter; for classical likelihood problems with $n$ in the thousands it
offers nothing over BFGS.

### Comparison

| Method | Derivatives | Storage | Cost per iteration | Local rate | Robust start | Hessian for SEs |
|---|---|---|---|---|---|---|
| Nelder–Mead | none | $O(p^2)$ | $O(p)$ evaluations | sublinear | good | no |
| Steepest descent | gradient | $O(p)$ | 1 gradient + line search | linear (slow) | good | no |
| Conjugate gradient | gradient | $O(p)$ | 1 gradient + line search | linear (fast) | moderate | no |
| BFGS | gradient | $O(p^2)$ | 1 gradient + $O(p^2)$ | superlinear | moderate | approximate |
| L-BFGS | gradient | $O(mp)$ | 1 gradient + $O(mp)$ | superlinear | moderate | no |
| Newton–Raphson | gradient, Hessian | $O(p^2)$ | 1 Hessian + $O(p^3)$ | quadratic | poor | exact |
| Fisher scoring / IRLS | score, expected information | $O(p^2)$ | 1 weighted LS fit | near-quadratic | fair | exact |
| Stochastic gradient | minibatch gradient | $O(p)$ | $O(\text{batch})$ | sublinear | good | no |

: Comparison of multivariate optimization methods. $p$ is the number of parameters and $m$ the L-BFGS memory length (typically 5–20). {#tbl-comparison}

### Implementations in R

| Function | Package | Methods available | Notes |
|---|---|---|---|
| `optim` | stats | Nelder–Mead, BFGS, CG, L-BFGS-B, SANN | General workhorse; supply `gr` or the gradient is differenced; `hessian = TRUE` for inference; L-BFGS-B allows box constraints |
| `nlm` | stats | Newton-type with ridging | Uses analytic gradient/Hessian if attached as attributes, otherwise numerical; returns `$hessian` |
| `nlminb` | stats | Trust-region (PORT) | Reliable on badly scaled problems; supports box constraints |
| `optimize`, `uniroot` | stats | Golden section, Brent | One dimension only |
| `glm` | stats | IRLS (Fisher scoring) | The specialised case; use `start` when separation is suspected |
| `nls` | stats | Gauss–Newton, Golub–Pereyra | Nonlinear least squares; `port` algorithm adds constraints |
| `optimx` | optimx | Unified front end to a dozen optimizers | Useful for comparing methods on the same objective |
| `nloptr` | nloptr | Large NLopt collection, with constraints | Global and derivative-free options |
| `Rsolnp`, `alabama` | — | Nonlinear equality and inequality constraints | For genuinely constrained problems |

: R interfaces to multivariate optimizers {#tbl-r-functions}

### Choosing a method

| Situation | Suggested approach |
|---|---|
| Hessian derivable, $p$ small, standard errors wanted | Newton–Raphson or Fisher scoring |
| GLM or a model expressible as one | `glm` (IRLS) |
| Gradient available, $p$ moderate | BFGS via `optim(method = "BFGS")` |
| Gradient available, $p$ large | L-BFGS |
| Derivatives unavailable, $p < 10$ | Nelder–Mead as a first pass, then refine |
| Objective noisy or non-smooth | Nelder–Mead or simulated annealing |
| Divergence or poor conditioning suspected | `nlm` or `nlminb`; rescale the parameters |
| Objective a sum over very many observations | Stochastic gradient descent and variants |
| Multiple local maxima suspected | Multiple random starts with any of the above |

: Practical guidance on method selection {#tbl-choosing}

The last row deserves emphasis. Every method in this chapter is a *local*
method: each converges to a stationary point near where it started, and none
can certify that the point found is the global maximum. Restarting from a
dispersed set of initial values and comparing the attained log-likelihoods is
the only general safeguard, and it should be routine for any likelihood that is
not known to be concave.

## Exercises

1. Verify algebraically that for logistic regression the observed and expected
   information matrices coincide, and explain why this fails for probit
   regression.

2. Add step halving to `mle_logistic_nr` and confirm that the start
   $\beta^{(0)} = (0, 5)$ now converges. How many halvings are needed on the
   first iteration?

3. Derive the IRLS form of the Newton–Raphson update for logistic regression,
   and verify numerically that one iteration of your derivation matches one
   iteration of `mle_logistic_nr`.

4. Generate a separable data set (choose $r_i = 1$ exactly when $z_i > 0$).
   Describe what happens to $\hat\beta_1$, to the information matrix, and to
   the standard errors reported by `glm`.

5. Compare `optim` with `method = "Nelder-Mead"`, `"BFGS"` and `"CG"` on the
   logistic negative log-likelihood, recording the number of function and
   gradient evaluations reported in `$counts`. Which is cheapest in total work?
