---
title: "Introduction to Bayesian Inference"
format:
  html:
    toc: true
    tbl-colwidths: true
filters:
  - shinylive
editor_options:
  chunk_output_type: console
code-fold: true
---

The last three chapters found a single best-fitting parameter value — the
MLE — by optimization: gradient methods and Newton's method for a smooth
log-likelihood, EM when part of the data is missing. Bayesian inference asks
a different question: not just *where* is the best $\theta$, but how
uncertain are we about it, given whatever we already believed before seeing
the data? The answer is a full distribution over $\theta$, not a point, and
extracting anything useful from that distribution — a mean, a probability, a
prediction — is almost always a matter of computing an integral or an
optimization. That is why Bayesian inference sets the agenda for the rest of
this book: numerical quadrature, Laplace's method, rejection sampling,
importance sampling, and Markov chain Monte Carlo are all, at bottom,
different strategies for computing with a posterior distribution that has no
closed form.

## The Bayesian Framework

Bayesian inference treats the unknown parameter $\theta$ itself as a random
variable and proceeds in four steps:

1. **Specify a prior** $\pi(\theta)$ — what is known or believed about
   $\theta$ before seeing any data.
2. **Specify a probability model for the data**, $P(y_1,\ldots,y_n\mid\theta)$,
   exactly the sampling model used to build a likelihood in the MLE chapters.
3. **Find the posterior** by Bayes' rule,
$$
P(\theta\mid y_{1:n}) = \frac{P(y_{1:n}\mid\theta)\,\pi(\theta)}
{\displaystyle\int P(y_{1:n}\mid\theta)\,\pi(\theta)\,d\theta}.
$$
4. **Interpret the posterior**: summarize $\theta$, make predictions.

Writing $L(\theta) = P(y_{1:n}\mid\theta)$ for the likelihood and
$f(\theta) = L(\theta)\pi(\theta)$ for the **unnormalized posterior** — the
notation used throughout the rest of this book — step 3 becomes
$$
P(\theta\mid y_{1:n}) = \frac{f(\theta)}{P(y_{1:n})}, \qquad
P(y_{1:n}) = \int f(\theta)\,d\theta.
$$
$P(y_{1:n})$, the **marginal likelihood**, is the likelihood averaged over
the prior rather than maximized over $\theta$; it does not depend on
$\theta$, so it plays no role in *shaping* the posterior, but it is usually
the hardest single quantity in the whole framework to compute — it is a
$p$-dimensional integral with no generic closed form, and it is the running
example of the next three chapters (numerical quadrature, Laplace
approximation, and the Monte Carlo methods after that).

```{r}
#| label: fig-prior-likelihood-posterior
#| fig-width: 10
#| fig-height: 4.6
#| fig-cap: "Left: a flat-ish prior $\\pi(\\theta)$ (dashed) and a likelihood
#|   $L(\\theta)$ (solid) that is much more concentrated — the data are far
#|   more informative than the prior here. Right: the posterior
#|   $\\propto L(\\theta)\\pi(\\theta)$, essentially the (renormalized)
#|   likelihood since the prior is so flat over the region where $L$ has
#|   mass. Shaded tails cut off $\\alpha/2$ of the posterior probability each,
#|   giving a $(1-\\alpha)$ credible interval $(\\theta_{\\alpha/2},\\theta_{1-\\alpha/2})$."
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
L <- function(t) dnorm(t, 3, 1); pri <- function(t) dnorm(t, 2, 4) * 2.5
curve(L, -2, 8, lwd = 3, col = "firebrick", xlab = expression(theta), ylab = "", yaxt = "n",
      main = "prior and likelihood", ylim = c(0, 0.45))
curve(pri, add = TRUE, lwd = 3, col = "gray40", lty = 2)
legend("topleft", bty = "n", lwd = 3, lty = c(2, 1), col = c("gray40", "firebrick"),
       legend = c(expression(pi(theta)), expression(L(theta))))
post <- function(t) L(t) * pri(t); Z <- integrate(post, -10, 15)$value
pd <- function(t) post(t) / Z
curve(pd, -2, 8, lwd = 3, col = "steelblue", xlab = expression(theta), ylab = "", yaxt = "n",
      main = expression("posterior" %prop% L(theta) * pi(theta)), ylim = c(0, 0.45))
tt <- seq(-2, 8, length.out = 1000); cdf <- cumsum(pd(tt)) * diff(tt)[1]
q <- c(tt[which.min(abs(cdf - 0.025))], tt[which.min(abs(cdf - 0.975))])
for (side in 1:2) { xs <- if (side == 1) seq(-2, q[1], length.out = 100) else seq(q[2], 8, length.out = 100)
  polygon(c(xs, rev(xs)), c(pd(xs), rep(0, 100)), col = adjustcolor("steelblue", 0.35), border = NA) }
abline(v = q, lty = 3); text(q, 0.42, c(expression(theta[alpha/2]), expression(theta[1-alpha/2])))
text(q[1] - 0.5, 0.05, expression(alpha/2)); text(q[2] + 0.5, 0.05, expression(alpha/2))
```

## Summarizing the Posterior

### Point estimates from loss functions

A single number for $\theta$ is a decision, and a principled way to make a
decision is to minimize expected loss. Given a loss function
$L(\theta,\hat\theta)$ measuring the cost of reporting $\hat\theta$ when the
truth is $\theta$, the **Bayes estimate** minimizes the *posterior expected*
loss,
$$
\hat\theta = \arg\min_{\hat\theta}\; E\big[L(\theta,\hat\theta)\mid y_{1:n}\big]
= \arg\min_{\hat\theta} \int L(\theta,\hat\theta)\,P(\theta\mid y_{1:n})\,d\theta.
$$
Three standard losses give three familiar summaries:

|Loss $L(\theta,\hat\theta)$|Bayes estimate $\hat\theta$|
|:---|:---|
|$(\theta-\hat\theta)^2$ (squared error)|posterior mean $E(\theta\mid y_{1:n})$|
|$\lvert\theta-\hat\theta\rvert$ (absolute error)|posterior median|
|$I(\theta\ne\hat\theta)$ (0–1 loss)|posterior mode (MAP)|

**Interval estimates** go one step further: a $(1-\alpha)$ **credible
interval** $(\theta_{\alpha/2},\theta_{1-\alpha/2})$ contains $\theta$ with
posterior probability $1-\alpha$ — a direct probability statement about
$\theta$ itself, unlike a frequentist confidence interval, whose
$1-\alpha$ coverage is a statement about the *procedure* across repeated
samples, not about $\theta$ given the data actually observed.

Every entry in that table is either an integral (a mean, a median, a
probability) or an optimization (the mode) over the posterior. This is the
sense in which Bayesian inference is inherently computational: even the most
basic summaries require operations that, outside of conjugate models like
the three worked below, have no closed form.

### Posterior prediction

Often the real target is not $\theta$ itself but a future observation. The
**posterior predictive distribution** of $y_{n+1}$ is
$$
P(y_{n+1}\mid y_{1:n}) = \frac{\int P(y_{n+1},y_{1:n}\mid\theta)\,\pi(\theta)\,d\theta}
{\int P(y_{1:n}\mid\theta)\,\pi(\theta)\,d\theta}.
$$
If $y_1,\ldots,y_n,y_{n+1}$ are conditionally independent given $\theta$ —
the usual iid-sampling assumption — the joint density factors and this
simplifies to
$$
P(y_{n+1}\mid y_{1:n}) = \int P(y_{n+1}\mid\theta)\,P(\theta\mid y_{1:n})\,d\theta,
$$
an *average* of the sampling density over the posterior, rather than the
sampling density evaluated at one plug-in estimate $\hat\theta$. Given
posterior draws $\theta_1,\ldots,\theta_J\sim P(\theta\mid y_{1:n})$ (the
subject of the Monte Carlo chapters later in the book),
$$
P(y_{n+1}\mid y_{1:n}) \approx \frac1J\sum_{j=1}^J P(y_{n+1}\mid\theta_j).
$$
Because it mixes over the whole posterior instead of conditioning on a
single $\hat\theta$, the predictive distribution is necessarily wider than
any individual $P(y_{n+1}\mid\theta_j)$: it carries both the sampling
variability *and* the remaining uncertainty about $\theta$. The Poisson–Gamma
example below makes this concrete with a closed-form predictive distribution
that is visibly overdispersed relative to a plug-in Poisson.

```{r}
#| label: fig-predictive
#| fig-width: 9
#| fig-height: 4.4
#| fig-cap: "Red: the sampling density $P(y_{n+1}\\mid\\theta_j)$ for ten
#|   posterior draws $\\theta_j\\sim N(3, 0.8^2)$. Black: their average, the
#|   posterior predictive density $P(y_{n+1}\\mid y_{1:n})$. The predictive
#|   is visibly wider than any single red curve — it inherits both the
#|   spread of $P(y_{n+1}\\mid\\theta)$ and the spread of the $\\theta_j$."
par(mar = c(4, 4, 1, 1)); J <- 10; th <- rnorm(J, 3, 0.8)
curve(dnorm(x, th[1]), -2, 8, col = adjustcolor("firebrick", 0.6), lwd = 1.5, ylim = c(0, 0.42),
      xlab = expression(y[n+1]), ylab = "density")
for (j in 2:J) curve(dnorm(x, th[j]), add = TRUE, col = adjustcolor("firebrick", 0.6), lwd = 1.5)
curve(sapply(x, function(v) mean(dnorm(v, th))), add = TRUE, lwd = 4)
legend("topright", bty = "n", lwd = c(1.5, 4), col = c("firebrick", "black"),
       legend = c(expression(P(y[n+1]*"|"*theta[j])), expression(P(y[n+1]*"|"*y[1:n]))))
```

## Example: Bernoulli Data with a Beta Prior

### Prior

Let $y_1,\ldots,y_n\mid\theta \overset{iid}{\sim}\text{Bern}(\theta)$, with
$\theta\in(0,1)$ a success probability, and give $\theta$ a
$\text{Beta}(\alpha_1,\alpha_0)$ prior, $\alpha_1,\alpha_0>0$ given:
$$
\pi(\theta) = \frac{\Gamma(\alpha_0+\alpha_1)}{\Gamma(\alpha_0)\Gamma(\alpha_1)}\,
\theta^{\alpha_1-1}(1-\theta)^{\alpha_0-1}, \qquad 0<\theta<1,
$$
$$
E(\theta) = \frac{\alpha_1}{\alpha_0+\alpha_1}, \qquad
V(\theta) = \frac{\alpha_0\alpha_1}{(\alpha_0+\alpha_1)^2(\alpha_0+\alpha_1+1)}.
$$
The parameters have a direct reading as *pseudo-data*: $\alpha_1$ behaves
like a prior count of successes, $\alpha_0$ a prior count of failures, and
$\alpha_0+\alpha_1$ a prior "sample size" — the more of it, the more
concentrated $\pi(\theta)$ is around $\alpha_1/(\alpha_0+\alpha_1)$, exactly
as a real sample of that size would concentrate a Bernoulli proportion.
$\text{Beta}(1,1)$ is $\text{Unif}(0,1)$, a genuinely flat prior; small
$\alpha$'s (e.g. $\text{Beta}(0.5,0.5)$) give a U-shaped prior that expects
$\theta$ near $0$ or $1$; large, unequal $\alpha$'s concentrate the prior
away from $\tfrac12$.

```{r}
#| label: fig-beta-priors
#| fig-width: 9
#| fig-height: 4.8
#| fig-cap: "Six Beta priors spanning flat, U-shaped, and concentrated shapes.
#|   Beta(1,1) is uniform; Beta(0.5,0.5) is U-shaped, favouring $\\theta$ near
#|   the boundaries; Beta(2,5) and Beta(30,10) are increasingly concentrated,
#|   at $\\theta\\approx2/7$ and $\\theta\\approx0.75$ respectively."
par(mar = c(4, 4, 1, 1))
pars <- list(c(1, 1), c(2, 1), c(0.5, 0.5), c(2, 5), c(10, 10), c(30, 10))
cols <- c("black", "firebrick", "darkorange", "forestgreen", "steelblue", "purple")
plot(NA, xlim = c(0, 1), ylim = c(0, 6), xlab = expression(theta), ylab = expression(pi(theta)))
for (k in seq_along(pars)) curve(dbeta(x, pars[[k]][1], pars[[k]][2]), add = TRUE, lwd = 3, col = cols[k], n = 501)
legend("top", bty = "n", lwd = 3, col = cols, ncol = 3,
       legend = sapply(pars, function(p) sprintf("Beta(%g, %g)", p[1], p[2])))
```

### Likelihood and posterior

With $P(y_i\mid\theta) = \theta^{y_i}(1-\theta)^{1-y_i}$ and
$\mathbf y = (y_1,\ldots,y_n)$, the likelihood depends on the data only
through the total number of successes:
$$
L(\theta) = \prod_{i=1}^n \theta^{y_i}(1-\theta)^{1-y_i} = \theta^{n_1}(1-\theta)^{n_0},
\qquad n_1 = \sum_i y_i,\ \ n_0 = n - n_1.
$$
Multiplying by the prior and collecting powers of $\theta$ and $1-\theta$,
$$
f(\theta) = \pi(\theta)L(\theta) \;\propto\; \theta^{\alpha_1-1}(1-\theta)^{\alpha_0-1}\cdot\theta^{n_1}(1-\theta)^{n_0}
= \theta^{(\alpha_1+n_1)-1}(1-\theta)^{(\alpha_0+n_0)-1}.
$$
This unnormalized posterior is, up to the constant $P(\mathbf y)$, exactly
the Beta kernel with updated parameters $\alpha_1' = \alpha_1+n_1$,
$\alpha_0' = \alpha_0+n_0$ — and since a Beta density is the *only*
normalized distribution with that kernel,
$$
\theta\mid\mathbf y \sim \text{Beta}(\alpha_1+n_1,\ \alpha_0+n_0)
$$
follows immediately, with no integration needed. This is what it means for
the Beta prior to be **conjugate** to the Bernoulli likelihood: updating the
prior to the posterior is nothing but arithmetic on the two parameters,
$\alpha_1 \to \alpha_1+n_1$, $\alpha_0\to\alpha_0+n_0$, and the marginal
likelihood $P(\mathbf y)$ never has to be computed explicitly, because
matching the two Beta kernels pins down the normalizing constant for free.
Few models have this property; it is a stroke of algebraic luck rather than
a general feature of Bayesian inference, which is exactly why the rest of
this book is needed.

### Posterior summaries

The pseudo-data reading of $\alpha_1,\alpha_0$ carries through to the
posterior mean, which is an exact weighted average of the prior mean and the
MLE $\hat\theta_{\text{MLE}} = n_1/n$:
$$
E(\theta\mid\mathbf y) = \frac{\alpha_1+n_1}{\alpha_0+\alpha_1+n}
= \underbrace{\frac{\alpha_0+\alpha_1}{\alpha_0+\alpha_1+n}}_{w}\cdot\frac{\alpha_1}{\alpha_0+\alpha_1}
+ (1-w)\cdot\hat\theta_{\text{MLE}}.
$$
The weight $w$ on the prior mean is the prior's pseudo-sample-size relative
to the total (prior plus real) sample size; as $n\to\infty$, $w\to0$ and the
posterior mean converges to the MLE regardless of the prior. The posterior
variance,
$$
V(\theta\mid\mathbf y) = \frac{(\alpha_1+n_1)(\alpha_0+n_0)}{(n+\alpha_0+\alpha_1)^2(n+\alpha_0+\alpha_1+1)} = O\!\left(\frac1n\right),
$$
shrinks at the same $1/n$ rate as the sampling variance of the MLE,
$\hat\theta_{\text{MLE}}\,\dot\sim\,N\big(\theta_0,\,1/(nI(\theta_0))\big)$
from the MLE chapters — for large $n$ the two approaches to uncertainty
agree, with the prior's influence vanishing.

### The marginal likelihood in closed form

Because conjugacy gives the *whole* posterior in closed form, it also gives
the marginal likelihood $P(\mathbf y)=\int_0^1 f(\theta)\,d\theta$ in closed
form, by integrating the unnormalized Beta kernel directly against its own
normalizing constant:
$$
P(\mathbf y) = \frac{\Gamma(\alpha_0+\alpha_1)}{\Gamma(\alpha_0)\Gamma(\alpha_1)}
\int_0^1 \theta^{(\alpha_1+n_1)-1}(1-\theta)^{(\alpha_0+n_0)-1}\,d\theta
= \frac{\Gamma(\alpha_0+\alpha_1)}{\Gamma(\alpha_0)\Gamma(\alpha_1)}\cdot
\frac{\Gamma(\alpha_1+n_1)\,\Gamma(\alpha_0+n_0)}{\Gamma(\alpha_0+\alpha_1+n)},
$$
using $\int_0^1\theta^{a-1}(1-\theta)^{b-1}\,d\theta = B(a,b) =
\Gamma(a)\Gamma(b)/\Gamma(a+b)$. On the log scale, with
$\log B(a,b)=\log\Gamma(a)+\log\Gamma(b)-\log\Gamma(a+b)$,
$$
\log P(\mathbf y) = \log B(\alpha_1+n_1,\ \alpha_0+n_0) - \log B(\alpha_1,\ \alpha_0).
$$
This exact Beta–Bernoulli conjugate identity is the same trick used for
Gamma–Poisson and Normal–Normal below: matching a recognizable kernel
against its own normalizing constant hands over $P(\mathbf y)$ for free,
with no numerical integration anywhere.

### Posterior predictive, in closed form

For a new $y_{n+1}\mid\theta\sim\text{Bern}(\theta)$, the general predictive
formula from earlier in this chapter simplifies beautifully in the
conjugate case, because $E(\theta\mid\mathbf y)$ *is* the predictive success
probability:
$$
P(y_{n+1}=1\mid\mathbf y) = \int_0^1 \theta\, P(\theta\mid\mathbf y)\,d\theta = E(\theta\mid\mathbf y),
\qquad y_{n+1}\mid\mathbf y \sim \text{Bern}\big(E(\theta\mid\mathbf y)\big).
$$
This exact equality ($E(\theta\mid\mathbf y)$ doing double duty as both a
point estimate of $\theta$ and the predictive probability of success) is
special to Bernoulli data; the Poisson–Gamma and Normal–Normal examples
below give predictive distributions that are genuinely wider than their
plug-in sampling models.

### Worked numbers: the complete inference output

Take a near-flat $\text{Beta}(1,1)$ prior and $n=10$ trials with $y=7$
successes. The table below collects everything derived above — marginal
likelihood, posterior, its mean/sd and credible interval, and the
predictive distribution — in one place, and the same five quantities are
computed the same way for the Gamma–Poisson and Normal–Normal examples that
follow:

```{r}
#| label: bb-example
a1 <- 1; a0 <- 1                 # Beta(1, 1): uniform prior
n_bb <- 10; y_bb <- 7            # 7 successes out of 10 trials
post_a1 <- a1 + y_bb; post_a0 <- a0 + (n_bb - y_bb)
bb_logmarg <- lbeta(post_a1, post_a0) - lbeta(a1, a0)   # log P(y), exact
bb_mean <- post_a1 / (post_a1 + post_a0)
bb_sd   <- sqrt(post_a1 * post_a0 / ((post_a1 + post_a0)^2 * (post_a1 + post_a0 + 1)))
bb_ci   <- qbeta(c(0.025, 0.975), post_a1, post_a0)
bb_pred_mean <- bb_mean                        # predictive: Bernoulli(bb_mean)
bb_pred_sd   <- sqrt(bb_pred_mean * (1 - bb_pred_mean))
```

```{r}
#| label: tbl-bb-summary
#| tbl-cap: "Complete inference output, Beta(1,1) prior with n = 10, y = 7."
knitr::kable(data.frame(
  quantity = c("prior mean", "MLE ($y/n$)", "log marginal likelihood",
               "posterior distribution", "posterior mean", "posterior sd",
               "95% credible interval", "predictive distribution",
               "predictive mean", "predictive sd"),
  value = c(sprintf("%.3f", a1 / (a1 + a0)), sprintf("%.3f", y_bb / n_bb),
            sprintf("%.3f", bb_logmarg),
            sprintf("Beta(%d, %d)", post_a1, post_a0),
            sprintf("%.3f", bb_mean), sprintf("%.3f", bb_sd),
            sprintf("(%.3f, %.3f)", bb_ci[1], bb_ci[2]),
            sprintf("Bernoulli(%.3f)", bb_pred_mean),
            sprintf("%.3f", bb_pred_mean), sprintf("%.3f", bb_pred_sd))
), col.names = c("quantity", "value"))
```

With a uniform prior the posterior mean $0.667$ sits between the prior mean
$0.5$ and the MLE $0.7$, pulled almost all the way to the MLE because the
prior carries only $\alpha_0+\alpha_1=2$ pseudo-observations against $10$
real ones. The 95% credible interval, $(0.390, 0.891)$, is read directly off
the Beta CDF (`qbeta`) — no simulation, no quadrature, because conjugacy
already handed over the exact posterior family. The predictive distribution
$\text{Bern}(0.667)$ has the same mean as the posterior but a nonzero sd
($0.471$), because even a Bernoulli outcome is uncertain even once $\theta$
is known exactly.

```{r}
#| label: fig-beta-posterior-narrowing
#| fig-width: 9
#| fig-height: 4.6
#| fig-cap: "Prior Beta(2,2) (dashed) and posteriors after $n=10,50,200$
#|   observations with $\\bar y=0.7$ held fixed. The posterior narrows at
#|   rate $1/\\sqrt n$ and moves onto the MLE as $n$ grows, exactly the
#|   $w\\to0$ argument above made visible."
par(mar = c(4, 4, 1, 1)); a1p <- 2; a0p <- 2; ybar <- 0.7
ns <- c(10, 50, 200); cols <- c("firebrick", "darkorange", "steelblue")
plot(NA, xlim = c(0, 1), ylim = c(0, 13), xlab = expression(theta), ylab = "density")
curve(dbeta(x, a1p, a0p), add = TRUE, lwd = 3, lty = 2, col = "gray40")
for (k in seq_along(ns)) curve(dbeta(x, a1p + ns[k] * ybar, a0p + ns[k] * (1 - ybar)), add = TRUE, lwd = 3, col = cols[k])
abline(v = ybar, lty = 3)
legend("topleft", bty = "n", lwd = 3, lty = c(2, 1, 1, 1), col = c("gray40", cols),
       legend = c("prior Beta(2, 2)", paste0("posterior, n = ", ns)))
text(ybar, 12.5, expression(hat(theta)[MLE]), pos = 4)
```

## Example: Poisson Data with a Gamma Prior

### Prior

Now let $y_1,\ldots,y_n\mid\lambda \overset{iid}{\sim}\text{Poisson}(\lambda)$,
$\lambda>0$ a rate, with a $\text{Gamma}(\alpha,\beta)$ prior (shape
$\alpha$, rate $\beta$),
$$
\pi(\lambda) = \frac{\beta^\alpha}{\Gamma(\alpha)}\lambda^{\alpha-1}e^{-\beta\lambda}, \qquad \lambda>0,
\qquad E(\lambda) = \frac{\alpha}{\beta}, \qquad V(\lambda) = \frac{\alpha}{\beta^2}.
$$
The pseudo-data reading again applies, with the two parameters playing
different roles than in the Beta case: $\alpha$ behaves like a prior total
count and $\beta$ like a prior total *exposure* (a number of pseudo
time-units or pseudo-observations already seen with rate $\lambda$), so
$\alpha/\beta$ is a prior average rate — the same "counts per exposure"
structure as $\hat\lambda_{\text{MLE}} = \sum y_i / n$, with $\beta$ playing
the role $n$ plays in the likelihood.

### Likelihood and posterior

With $P(y_i\mid\lambda) = \lambda^{y_i}e^{-\lambda}/y_i!$, the likelihood
depends on the data only through the total count $S=\sum_i y_i$:
$$
L(\lambda) = \prod_{i=1}^n \frac{\lambda^{y_i}e^{-\lambda}}{y_i!}
= \frac{\lambda^{S}e^{-n\lambda}}{\prod_i y_i!}, \qquad S = \sum_{i=1}^n y_i.
$$
Multiplying by the prior and collecting powers of $\lambda$ and $e^{-\lambda}$,
$$
f(\lambda) = \pi(\lambda)L(\lambda) \;\propto\; \lambda^{\alpha-1}e^{-\beta\lambda}\cdot\lambda^{S}e^{-n\lambda}
= \lambda^{(\alpha+S)-1}e^{-(\beta+n)\lambda},
$$
the kernel of a $\text{Gamma}(\alpha+S,\ \beta+n)$ density. The Gamma prior
is therefore **conjugate** to the Poisson likelihood, exactly as Beta was to
Bernoulli, and by the same argument (matching a recognizable kernel pins
down the normalizing constant automatically):
$$
\lambda\mid\mathbf y \sim \text{Gamma}(\alpha+S,\ \beta+n).
$$
The posterior mean has precisely the weighted-average structure seen in the
Bernoulli case, with $\beta$ now playing the role $\alpha_0+\alpha_1$ played
there:
$$
E(\lambda\mid\mathbf y) = \frac{\alpha+S}{\beta+n}
= \underbrace{\frac{\beta}{\beta+n}}_{w}\cdot\frac{\alpha}{\beta}
+ (1-w)\cdot\underbrace{\frac{S}{n}}_{\hat\lambda_{\text{MLE}}},
$$
a weighted average of the prior mean and the MLE, with prior weight
$w=\beta/(\beta+n)\to0$ as $n\to\infty$ — the same shrinkage story as
Beta–Bernoulli, now driven by $\beta$ (prior pseudo-exposure) instead of
$\alpha_0+\alpha_1$ (prior pseudo-sample-size).

### The marginal likelihood in closed form

Because conjugacy gives the *whole* posterior in closed form, it also gives
the marginal likelihood $P(\mathbf y)=\int f(\lambda)\,d\lambda$ in closed
form, by integrating the unnormalized Gamma kernel directly against its own
normalizing constant:
$$
P(\mathbf y) = \left(\prod_{i=1}^n \frac1{y_i!}\right)\frac{\beta^\alpha}{\Gamma(\alpha)}
\int_0^\infty \lambda^{(\alpha+S)-1}e^{-(\beta+n)\lambda}\,d\lambda
= \left(\prod_{i=1}^n \frac1{y_i!}\right)\frac{\beta^\alpha\,\Gamma(\alpha+S)}{\Gamma(\alpha)\,(\beta+n)^{\alpha+S}},
$$
using $\int_0^\infty \lambda^{k-1}e^{-c\lambda}\,d\lambda = \Gamma(k)/c^k$.
On the log scale,
$$
\log P(\mathbf y) = \alpha\log\beta - \log\Gamma(\alpha) + \log\Gamma(\alpha+S)
- (\alpha+S)\log(\beta+n) - \sum_{i=1}^n \log(y_i!).
$$
This exact Gamma–Poisson conjugate identity is what makes the Poisson-rate
model a useful *test case* for approximate methods: the Laplace
approximation chapter later in this book applies its mode-and-curvature
approximation to this same model and checks the result against precisely
this formula, since here — unlike in most applications — the truth is known.

### Posterior predictive, in closed form

For a new count $y_{n+1}\mid\lambda\sim\text{Poisson}(\lambda)$, integrating
the Poisson sampling density against the $\text{Gamma}(\alpha+S,\beta+n)$
posterior gives a **Negative Binomial** predictive distribution,
$$
P(y_{n+1}\mid\mathbf y) = \int_0^\infty \frac{\lambda^{y_{n+1}}e^{-\lambda}}{y_{n+1}!}\,
\frac{(\beta+n)^{\alpha+S}}{\Gamma(\alpha+S)}\lambda^{(\alpha+S)-1}e^{-(\beta+n)\lambda}\,d\lambda
= \text{NegBin}\!\left(\text{size}=\alpha+S,\ \ p=\frac{\beta+n}{\beta+n+1}\right),
$$
by the same Gamma-integral identity used for the marginal likelihood above.
Its mean matches the posterior mean of $\lambda$ exactly (as it must, by the
tower property), but its variance is *larger* than that of a plug-in
Poisson at the posterior mean rate:
$$
E(y_{n+1}\mid\mathbf y) = \frac{\alpha+S}{\beta+n}, \qquad
V(y_{n+1}\mid\mathbf y) = \frac{\alpha+S}{\beta+n}\left(1+\frac1{\beta+n}\right)
= E(y_{n+1}\mid\mathbf y)\cdot\Big(1+\tfrac{1}{\beta+n}\Big).
$$
A plug-in Poisson at $\lambda=E(\lambda\mid\mathbf y)$ would have variance
*equal* to its mean — the extra factor $1+1/(\beta+n)$ is entirely the
posterior uncertainty about $\lambda$ leaking into the prediction, exactly
the "wider than any single $P(y_{n+1}\mid\theta_j)$" phenomenon from the
general framework above, now with an explicit formula for *how much* wider.

### Worked numbers: the complete inference output

Take $\text{Gamma}(2,1)$ as the prior (prior mean $2$, i.e. a mild belief
that the rate is around $2$, with pseudo-exposure $\beta=1$) and $n=8$
simulated counts from a true rate of $3$. The table collects the same five
quantities as the Beta–Bernoulli table above — marginal likelihood,
posterior, its mean/sd and credible interval, and the predictive
distribution:

```{r}
#| label: pg-example
alpha <- 2; beta <- 1                       # Gamma(2, 1) prior
set.seed(42)
n_pg <- 8; y_pg <- rpois(n_pg, 3)           # 8 counts from a true rate of 3
S_pg <- sum(y_pg)
post_alpha <- alpha + S_pg; post_beta <- beta + n_pg
pg_logmarg <- alpha * log(beta) - lgamma(alpha) + lgamma(post_alpha) -
  post_alpha * log(post_beta) - sum(lgamma(y_pg + 1))     # log P(y), exact
pg_mean <- post_alpha / post_beta
pg_sd   <- sqrt(post_alpha) / post_beta
pg_ci   <- qgamma(c(0.025, 0.975), post_alpha, post_beta)
pg_nb_prob   <- post_beta / (post_beta + 1)    # predictive: NegBin(post_alpha, pg_nb_prob)
pg_pred_mean <- pg_mean
pg_pred_sd   <- sqrt(pg_mean * (1 + 1 / post_beta))
```

```{r}
#| label: tbl-pg-summary
#| tbl-cap: "Complete inference output, Gamma(2,1) prior, n = 8 Poisson counts."
knitr::kable(data.frame(
  quantity = c("data $y_{1:8}$", "sum $S$", "prior mean", "MLE ($S/n$)",
               "log marginal likelihood", "posterior distribution",
               "posterior mean", "posterior sd", "95% credible interval",
               "predictive distribution", "predictive mean", "predictive sd"),
  value = c(paste(y_pg, collapse = ", "), as.character(S_pg),
            sprintf("%.3f", alpha / beta), sprintf("%.3f", S_pg / n_pg),
            sprintf("%.3f", pg_logmarg),
            sprintf("Gamma(%d, %d)", post_alpha, post_beta),
            sprintf("%.3f", pg_mean), sprintf("%.3f", pg_sd),
            sprintf("(%.3f, %.3f)", pg_ci[1], pg_ci[2]),
            sprintf("NegBin(%d, %.3f)", post_alpha, pg_nb_prob),
            sprintf("%.3f", pg_pred_mean), sprintf("%.3f", pg_pred_sd))
), col.names = c("quantity", "value"))
```

The MLE, $29/8=3.625$, and the prior mean, $2$, are pulled toward each other
by the posterior mean $3.444$ — mostly toward the MLE, since the prior's
pseudo-exposure $\beta=1$ is small next to the real $n=8$. Compare this to
the true generating rate of $3$: both the MLE and the posterior mean
overshoot it here (this particular simulated sample happened to run a bit
high), and the 95% credible interval $(2.340, 4.759)$ still comfortably
covers it. With $\beta+n=9$, the predictive sd $1.957$ exceeds the plug-in
Poisson sd $\sqrt{3.444}=1.856$ by the factor $\sqrt{1+1/9}$ — the same
overdispersion story made numeric.

### Overdispersion, visualized

The gap is small here because $n$ is already large relative to the prior,
but it widens as $n$ shrinks or the prior gets vaguer (smaller $\beta$). The
figure below makes the same point on the density scale: the predictive has
visibly heavier tails than a plug-in Poisson with the same mean.

```{r}
#| label: fig-pg-predictive-dispersion
#| fig-width: 9
#| fig-height: 4.4
#| fig-cap: "Posterior predictive $P(y_{n+1}\\mid\\mathbf y)$ (Negative
#|   Binomial, red) against a plug-in Poisson at the posterior mean rate
#|   (gray). Same mean, but the predictive has heavier tails: the two bars
#|   noticeably diverge above $y_{n+1}=7$, and $V(y_{n+1}\\mid\\mathbf y)$
#|   exceeds the plug-in variance by the factor $1+1/(\\beta+n)=1.111$
#|   derived above."
ks <- 0:12
nb_prob <- post_beta / (post_beta + 1)
pred_nb     <- dnbinom(ks, size = post_alpha, prob = nb_prob)
pred_plugin <- dpois(ks, pg_mean)
par(mar = c(4, 4, 1, 1))
barplot(rbind(pred_nb, pred_plugin), beside = TRUE, col = c("firebrick", "gray60"),
        names.arg = ks, xlab = expression(y[n+1]), ylab = expression(P(y[n+1])))
legend("topright", bty = "n", fill = c("firebrick", "gray60"),
       legend = c("posterior predictive (Negative Binomial)", "plug-in Poisson at posterior mean"))
```

## Example: Normal Data with a Normal Prior (Known Variance)

### Prior

Now let $y_1,\ldots,y_n\mid\mu \overset{iid}{\sim} N(\mu,\sigma^2)$, with
$\sigma^2$ **known/fixed** and only the mean $\mu\in\mathbb R$ unknown. Give
$\mu$ a Normal prior, parametrized through a prior pseudo-sample-size
$\kappa_0>0$ so that it lines up with the pseudo-data reading used in the two
examples above:
$$
\pi(\mu) = N\!\Big(\mu_0,\ \frac{\sigma^2}{\kappa_0}\Big), \qquad
E(\mu) = \mu_0, \qquad V(\mu) = \frac{\sigma^2}{\kappa_0}.
$$
$\mu_0$ is the prior guess for $\mu$, and $\kappa_0$ counts how many
pseudo-observations (each with the known variance $\sigma^2$) that guess is
worth: large $\kappa_0$ means a confident, concentrated prior; $\kappa_0\to0$
recovers a flat, uninformative prior on $\mu$.

### Likelihood and posterior

With $P(y_i\mid\mu) = (2\pi\sigma^2)^{-1/2}\exp\{-(y_i-\mu)^2/(2\sigma^2)\}$,
the likelihood depends on the data only through the sample mean
$\bar y = \frac1n\sum_i y_i$:
$$
L(\mu) \;\propto\; \exp\left\{-\frac1{2\sigma^2}\sum_{i=1}^n(y_i-\mu)^2\right\}
\;\propto\; \exp\left\{-\frac{n(\mu-\bar y)^2}{2\sigma^2}\right\}.
$$
Multiplying by the prior and completing the square in $\mu$,
$$
f(\mu) = \pi(\mu)L(\mu) \;\propto\;
\exp\left\{-\frac{(\mu-\mu_0)^2}{2\sigma^2/\kappa_0} - \frac{n(\mu-\bar y)^2}{2\sigma^2}\right\}
\;\propto\; \exp\left\{-\frac{(\mu-\mu_n)^2}{2\sigma^2/\kappa_n}\right\},
$$
the kernel of a Normal density with
$$
\kappa_n = \kappa_0+n, \qquad \mu_n = \frac{\kappa_0\mu_0+n\bar y}{\kappa_0+n}.
$$
The Normal prior is therefore **conjugate** to the Normal likelihood (known
variance), exactly as Beta was to Bernoulli and Gamma to Poisson:
$$
\mu\mid\mathbf y \sim N\!\Big(\mu_n,\ \frac{\sigma^2}{\kappa_n}\Big).
$$
As in the other two models, the posterior mean is a pseudo-sample-size
weighted average of the prior mean and the MLE $\hat\mu_{\text{MLE}}=\bar y$,
$$
\mu_n = \underbrace{\frac{\kappa_0}{\kappa_0+n}}_{w}\cdot\mu_0 + (1-w)\cdot\bar y,
$$
with prior weight $w=\kappa_0/(\kappa_0+n)\to0$ as $n\to\infty$, and the
posterior "pseudo-sample-sizes" simply **add**: $\kappa_n=\kappa_0+n$ —
posterior precision (inverse variance, in units of $1/\sigma^2$) is prior
precision plus data precision.

### The marginal likelihood in closed form

Because conjugacy gives the *whole* posterior in closed form, it also gives
the marginal likelihood $P(\mathbf y)=\int f(\mu)\,d\mu$ in closed form.
Writing the likelihood with its full normalizing constant,
$L(\mu)=(2\pi\sigma^2)^{-n/2}\exp\{-SS/(2\sigma^2)\}\exp\{-n(\mu-\bar y)^2/(2\sigma^2)\}$
with $SS=\sum_{i=1}^n(y_i-\bar y)^2$, and integrating the same
Gaussian-kernel product used to find the posterior above (a standard
Gaussian-product identity: $\int\exp\{-a(\mu-m_1)^2/2-b(\mu-m_2)^2/2\}\,d\mu
=\sqrt{2\pi/(a+b)}\exp\{-ab(m_1-m_2)^2/(2(a+b))\}$) gives
$$
P(\mathbf y) = (2\pi\sigma^2)^{-n/2}\sqrt{\frac{\kappa_0}{\kappa_n}}\,
\exp\left\{-\frac{SS}{2\sigma^2} - \frac{\kappa_0 n(\mu_0-\bar y)^2}{2\sigma^2\kappa_n}\right\}.
$$
On the log scale,
$$
\log P(\mathbf y) = -\frac n2\log(2\pi\sigma^2) + \frac12\log\frac{\kappa_0}{\kappa_n}
- \frac{SS}{2\sigma^2} - \frac{\kappa_0 n(\mu_0-\bar y)^2}{2\sigma^2\kappa_n}.
$$
As in the other two models, matching the Normal kernel of $f(\mu)$ against
its own normalizing constant hands over $P(\mathbf y)$ with no numerical
integration.

### Posterior predictive, in closed form

For a new $y_{n+1}\mid\mu\sim N(\mu,\sigma^2)$, integrating the Normal
sampling density against the $N(\mu_n,\sigma^2/\kappa_n)$ posterior — a sum
of independent Normals — gives another Normal predictive distribution,
$$
y_{n+1}\mid\mathbf y \sim N\!\Big(\mu_n,\ \sigma^2\big(1+\tfrac1{\kappa_n}\big)\Big),
$$
with mean equal to the posterior mean of $\mu$ (as it must, by the tower
property) and variance **larger** than the known sampling variance
$\sigma^2$ by exactly the factor $1+1/\kappa_n$ — the same form as the
Gamma–Poisson predictive-variance inflation above, now for a continuous
outcome: the extra $1/\kappa_n$ is the leftover posterior uncertainty about
$\mu$ leaking into the prediction.

### Worked numbers: the complete inference output

Take a weakly informative prior, $\mu_0=0$ with $\kappa_0=1$ (worth one
pseudo-observation), known $\sigma=2$, and $n=15$ observations simulated
from a true mean of $3$. The table collects the same five quantities as the
Beta–Bernoulli and Gamma–Poisson tables above:

```{r}
#| label: nn-example
sigma_nn <- 2                                  # known/fixed sigma
mu0 <- 0; kappa0 <- 1                          # N(0, sigma^2/1) prior
set.seed(123)
n_nn <- 15; y_nn <- rnorm(n_nn, 3, sigma_nn)   # 15 draws from a true mean of 3
ybar_nn <- mean(y_nn)
SS_nn <- sum((y_nn - ybar_nn)^2)
post_kappa <- kappa0 + n_nn
post_mu  <- (kappa0 * mu0 + n_nn * ybar_nn) / post_kappa
post_var <- sigma_nn^2 / post_kappa
post_sd  <- sqrt(post_var)
nn_ci    <- qnorm(c(0.025, 0.975), post_mu, post_sd)
nn_logmarg <- -n_nn / 2 * log(2 * pi * sigma_nn^2) + 0.5 * log(kappa0 / post_kappa) -
  SS_nn / (2 * sigma_nn^2) -
  kappa0 * n_nn * (mu0 - ybar_nn)^2 / (2 * sigma_nn^2 * post_kappa)   # log P(y), exact
pred_var <- post_var + sigma_nn^2              # predictive: N(post_mu, pred_var)
pred_sd  <- sqrt(pred_var)
```

```{r}
#| label: tbl-nn-summary
#| tbl-cap: "Complete inference output, $N(0, \\sigma^2/1)$ prior with known
#|   $\\sigma=2$, n = 15."
knitr::kable(data.frame(
  quantity = c("prior mean", "MLE ($\\bar y$)", "log marginal likelihood",
               "posterior distribution", "posterior mean", "posterior sd",
               "95% credible interval", "predictive distribution",
               "predictive mean", "predictive sd"),
  value = c(sprintf("%.3f", mu0), sprintf("%.3f", ybar_nn),
            sprintf("%.3f", nn_logmarg),
            sprintf("N(%.3f, %.3f^2)", post_mu, post_sd),
            sprintf("%.3f", post_mu), sprintf("%.3f", post_sd),
            sprintf("(%.3f, %.3f)", nn_ci[1], nn_ci[2]),
            sprintf("N(%.3f, %.3f^2)", post_mu, pred_sd),
            sprintf("%.3f", post_mu), sprintf("%.3f", pred_sd))
), col.names = c("quantity", "value"))
```

With only $\kappa_0=1$ pseudo-observation against $n=15$ real ones, the
posterior mean $`r sprintf("%.3f", post_mu)`$ sits almost on top of the MLE
$\bar y=`r sprintf("%.3f", ybar_nn)`$, barely pulled toward the prior mean
$0$ — the same shrinkage story as the two examples above, now driven by
$\kappa_0$ instead of $\alpha_0+\alpha_1$ or $\beta$. The predictive sd
$`r sprintf("%.3f", pred_sd)`$ exceeds the known $\sigma=2$ by the factor
$\sqrt{1+1/\kappa_n}=\sqrt{`r sprintf("%.3f", 1 + 1 / post_kappa)`}$ derived
above — the continuous analogue of the Gamma–Poisson overdispersion factor.

### Overdispersion, visualized

```{r}
#| label: fig-nn-predictive-dispersion
#| fig-width: 9
#| fig-height: 4.4
#| fig-cap: "Posterior predictive $N(\\mu_n, \\sigma^2(1+1/\\kappa_n))$ (red,
#|   solid) against a plug-in $N(\\mu_n,\\sigma^2)$ at the posterior mean
#|   (gray, dashed). Same mean, but the predictive is visibly wider — the
#|   extra spread is entirely the remaining posterior uncertainty about
#|   $\\mu$."
par(mar = c(4, 4, 1, 1)); pred_sd <- sqrt(post_var + sigma_nn^2)
curve(dnorm(x, post_mu, pred_sd), post_mu - 4 * pred_sd, post_mu + 4 * pred_sd,
      lwd = 3, col = "firebrick", xlab = expression(y[n+1]), ylab = "density")
curve(dnorm(x, post_mu, sigma_nn), add = TRUE, lwd = 3, lty = 2, col = "gray40")
legend("topright", bty = "n", lwd = 3, lty = c(1, 2), col = c("firebrick", "gray40"),
       legend = c("posterior predictive", "plug-in Normal at posterior mean"))
```

## Explore it interactively {#sec-bayes-explorer}

This app
runs live only in the HTML edition of this book; readers of other formats
can follow the link below it.

::: {#fig-bayes-explorer-app .content-visible when-format="html"}

```{shinylive-r}
#| label: bayes-explorer-app
#| standalone: true
#| viewerHeight: 800

library(shiny)

ui <- fluidPage(
  titlePanel(NULL),
  sidebarLayout(
    sidebarPanel(
      width = 4,
      selectInput("model", "Model",
                  choices = c("Bernoulli - Beta" = "bb", "Poisson - Gamma" = "pg",
                              "Normal (known sigma) - Normal" = "nn")),
      conditionalPanel(
        "input.model == 'bb'",
        numericInput("a1", "prior alpha1 (successes)", value = 1, min = 0.1, step = 0.5),
        numericInput("a0", "prior alpha0 (failures)", value = 1, min = 0.1, step = 0.5),
        numericInput("n_bb", "n (trials)", value = 10, min = 1, step = 1),
        numericInput("y_bb", "y (successes observed)", value = 7, min = 0, step = 1)
      ),
      conditionalPanel(
        "input.model == 'pg'",
        numericInput("alpha", "prior alpha (shape)", value = 2, min = 0.1, step = 0.5),
        numericInput("beta", "prior beta (rate)", value = 1, min = 0.1, step = 0.5),
        numericInput("n_pg", "n (observations)", value = 8, min = 1, step = 1),
        numericInput("S_pg", "sum of counts", value = 29, min = 0, step = 1)
      ),
      conditionalPanel(
        "input.model == 'nn'",
        numericInput("mu0", "prior mu0 (mean)", value = 0, step = 0.5),
        numericInput("kappa0", "prior kappa0 (pseudo-sample-size)", value = 1, min = 0.1, step = 0.5),
        numericInput("sigma_nn", "sigma (known)", value = 2, min = 0.1, step = 0.5),
        numericInput("n_nn", "n (observations)", value = 15, min = 1, step = 1),
        numericInput("ybar_nn", "sample mean y-bar", value = 3, step = 0.5)
      ),
      helpText("Dashed = prior, solid = posterior. The dotted vertical line ",
               "is the MLE (y/n for Bernoulli, S/n for Poisson, y-bar for Normal). ",
               "*Log marginal likelihood is exact for Bernoulli; for Poisson ",
               "and Normal it is shown up to an additive constant that ",
               "depends only on the individual data points (not part of the ",
               "summary inputs here) and cancels when comparing priors for ",
               "the same data.")
    ),
    mainPanel(
      width = 8,
      plotOutput("densplot", height = "380px"),
      tableOutput("summtab")
    )
  )
)

server <- function(input, output, session) {

  output$densplot <- renderPlot({
    par(mar = c(4, 4, 2, 1))
    if (input$model == "bb") {
      a1 <- input$a1; a0 <- input$a0; n <- input$n_bb; y <- input$y_bb
      post_a1 <- a1 + y; post_a0 <- a0 + (n - y)
      xs <- seq(0, 1, length.out = 500)
      ylim <- c(0, max(dbeta(xs, a1, a0), dbeta(xs, post_a1, post_a0)) * 1.1)
      curve(dbeta(x, a1, a0), 0, 1, lwd = 2, lty = 2, col = "gray40",
            xlab = expression(theta), ylab = "density", ylim = ylim)
      curve(dbeta(x, post_a1, post_a0), add = TRUE, lwd = 3, col = "steelblue")
      abline(v = y / n, lty = 3)
      legend("topright", bty = "n", lwd = c(2, 3), lty = c(2, 1), col = c("gray40", "steelblue"),
             legend = c("prior", "posterior"))
    } else if (input$model == "pg") {
      alpha <- input$alpha; beta <- input$beta; n <- input$n_pg; S <- input$S_pg
      post_alpha <- alpha + S; post_beta <- beta + n
      hi <- max(qgamma(0.999, alpha, beta), qgamma(0.999, post_alpha, post_beta))
      xs <- seq(0, hi, length.out = 500)
      ylim <- c(0, max(dgamma(xs, alpha, beta), dgamma(xs, post_alpha, post_beta)) * 1.1)
      curve(dgamma(x, alpha, beta), 0, hi, lwd = 2, lty = 2, col = "gray40",
            xlab = expression(lambda), ylab = "density", ylim = ylim)
      curve(dgamma(x, post_alpha, post_beta), add = TRUE, lwd = 3, col = "firebrick")
      abline(v = S / n, lty = 3)
      legend("topright", bty = "n", lwd = c(2, 3), lty = c(2, 1), col = c("gray40", "firebrick"),
             legend = c("prior", "posterior"))
    } else {
      mu0 <- input$mu0; kappa0 <- input$kappa0; sigma <- input$sigma_nn
      n <- input$n_nn; ybar <- input$ybar_nn
      post_kappa <- kappa0 + n
      post_mu <- (kappa0 * mu0 + n * ybar) / post_kappa
      post_sd <- sigma / sqrt(post_kappa)
      prior_sd <- sigma / sqrt(kappa0)
      lo <- min(mu0 - 3 * prior_sd, post_mu - 3 * post_sd)
      hi <- max(mu0 + 3 * prior_sd, post_mu + 3 * post_sd)
      xs <- seq(lo, hi, length.out = 500)
      ylim <- c(0, max(dnorm(xs, mu0, prior_sd), dnorm(xs, post_mu, post_sd)) * 1.1)
      curve(dnorm(x, mu0, prior_sd), lo, hi, lwd = 2, lty = 2, col = "gray40",
            xlab = expression(mu), ylab = "density", ylim = ylim)
      curve(dnorm(x, post_mu, post_sd), add = TRUE, lwd = 3, col = "darkorchid")
      abline(v = ybar, lty = 3)
      legend("topright", bty = "n", lwd = c(2, 3), lty = c(2, 1), col = c("gray40", "darkorchid"),
             legend = c("prior", "posterior"))
    }
  })

  output$summtab <- renderTable({
    if (input$model == "bb") {
      a1 <- input$a1; a0 <- input$a0; n <- input$n_bb; y <- input$y_bb
      post_a1 <- a1 + y; post_a0 <- a0 + (n - y)
      logmarg    <- lbeta(post_a1, post_a0) - lbeta(a1, a0)         # exact
      pred_mean  <- post_a1 / (post_a1 + post_a0)
      pred_sd    <- sqrt(pred_mean * (1 - pred_mean))
      data.frame(
        quantity = c("prior mean", "MLE (y/n)", "log marginal likelihood*",
                     "posterior distribution", "posterior mean", "posterior sd",
                     "95% credible interval", "predictive distribution",
                     "predictive mean", "predictive sd"),
        value = c(sprintf("%.3f", a1 / (a1 + a0)),
                  sprintf("%.3f", y / n),
                  sprintf("%.3f", logmarg),
                  sprintf("Beta(%.2f, %.2f)", post_a1, post_a0),
                  sprintf("%.3f", pred_mean),
                  sprintf("%.3f", sqrt(post_a1 * post_a0 / ((post_a1 + post_a0)^2 * (post_a1 + post_a0 + 1)))),
                  sprintf("(%.3f, %.3f)", qbeta(0.025, post_a1, post_a0), qbeta(0.975, post_a1, post_a0)),
                  sprintf("Bernoulli(%.3f)", pred_mean),
                  sprintf("%.3f", pred_mean), sprintf("%.3f", pred_sd)))
    } else if (input$model == "pg") {
      alpha <- input$alpha; beta <- input$beta; n <- input$n_pg; S <- input$S_pg
      post_alpha <- alpha + S; post_beta <- beta + n
      logmarg   <- alpha * log(beta) - lgamma(alpha) + lgamma(post_alpha) -
        post_alpha * log(post_beta)                                 # up to -sum(log(y_i!))
      nb_prob   <- post_beta / (post_beta + 1)
      pred_mean <- post_alpha / post_beta
      pred_sd   <- sqrt(pred_mean * (1 + 1 / post_beta))
      data.frame(
        quantity = c("prior mean", "MLE (S/n)", "log marginal likelihood*",
                     "posterior distribution", "posterior mean", "posterior sd",
                     "95% credible interval", "predictive distribution",
                     "predictive mean", "predictive sd"),
        value = c(sprintf("%.3f", alpha / beta),
                  sprintf("%.3f", S / n),
                  sprintf("%.3f", logmarg),
                  sprintf("Gamma(%.2f, %.2f)", post_alpha, post_beta),
                  sprintf("%.3f", pred_mean),
                  sprintf("%.3f", sqrt(post_alpha) / post_beta),
                  sprintf("(%.3f, %.3f)", qgamma(0.025, post_alpha, post_beta), qgamma(0.975, post_alpha, post_beta)),
                  sprintf("NegBin(%.2f, %.3f)", post_alpha, nb_prob),
                  sprintf("%.3f", pred_mean), sprintf("%.3f", pred_sd)))
    } else {
      mu0 <- input$mu0; kappa0 <- input$kappa0; sigma <- input$sigma_nn
      n <- input$n_nn; ybar <- input$ybar_nn
      post_kappa <- kappa0 + n
      post_mu <- (kappa0 * mu0 + n * ybar) / post_kappa
      post_sd <- sigma / sqrt(post_kappa)
      logmarg <- -n / 2 * log(2 * pi * sigma^2) + 0.5 * log(kappa0 / post_kappa) -
        kappa0 * n * (mu0 - ybar)^2 / (2 * sigma^2 * post_kappa)     # up to -SS/(2 sigma^2)
      pred_sd <- sqrt(post_sd^2 + sigma^2)
      data.frame(
        quantity = c("prior mean", "MLE (y-bar)", "log marginal likelihood*",
                     "posterior distribution", "posterior mean", "posterior sd",
                     "95% credible interval", "predictive distribution",
                     "predictive mean", "predictive sd"),
        value = c(sprintf("%.3f", mu0),
                  sprintf("%.3f", ybar),
                  sprintf("%.3f", logmarg),
                  sprintf("N(%.3f, %.3f^2)", post_mu, post_sd),
                  sprintf("%.3f", post_mu),
                  sprintf("%.3f", post_sd),
                  sprintf("(%.3f, %.3f)", qnorm(0.025, post_mu, post_sd), qnorm(0.975, post_mu, post_sd)),
                  sprintf("N(%.3f, %.3f^2)", post_mu, pred_sd),
                  sprintf("%.3f", post_mu), sprintf("%.3f", pred_sd)))
    }
  }, colnames = FALSE, striped = TRUE)
}

shinyApp(ui, server)
```

**Shinylive App Illustrating Conjugate Bayesian Updating.** The app below lets you adjust the prior, sample size, and observed data for
any of the three conjugate models and watch prior and posterior move
together, along with the posterior summary table computed above.

:::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit07_bayesian_inference.html#sec-bayes-explorer>.
:::

## Where Computation Enters

For the Beta–Bernoulli, Gamma–Poisson, and Normal–Normal models, everything
above is closed-form: the posterior, its mean and variance, credible
intervals, and even the marginal likelihood and predictive distribution.
Conjugacy is a special convenience, restricted to a handful of
prior/likelihood pairings; for a general model, the same quantities require
genuine numerical work:

|Quantity|Requires|
|:---|:---|
|Normalizing constant $P(\mathbf y) = \int L(\theta)\pi(\theta)\,d\theta$|integration over $\theta$|
|Posterior mean, variance, probabilities|integration over $\theta$|
|Posterior mode|optimization|
|Posterior predictive $\int P(y_{n+1}\mid\theta)P(\theta\mid\mathbf y)\,d\theta$|integration over $\theta$|
|Credible intervals|posterior quantiles|

The strategy depends on the dimension of $\theta$. For **one-dimensional**
$\theta$, a deterministic grid of evaluation points — **numerical
quadrature**, the next chapter — is cheap and accurate, and **Laplace's
method** (the chapter after that) offers a grid-free alternative built from
just the posterior mode and its curvature, at the cost of assuming the
posterior is roughly bell-shaped. For **multi-dimensional** $\theta$, a grid
becomes exponentially expensive, and the standard approach is Monte Carlo:
approximate every integral by an average over posterior draws
$\theta_1,\ldots,\theta_J\sim P(\theta\mid\mathbf y)$, as prediction already
did above. Producing those draws — by **rejection sampling**, **importance
sampling**, and finally **Markov chain Monte Carlo** — is the subject of the
remaining chapters of this book.

## Summary

- Bayesian inference treats $\theta$ as random, combines a prior $\pi(\theta)$
  with a likelihood $L(\theta)$ via Bayes' rule, and reports a *distribution*
  $P(\theta\mid\mathbf y)\propto f(\theta)=L(\theta)\pi(\theta)$ rather than a
  single point estimate.
- Every standard summary — posterior mean/median/mode, credible intervals,
  predictive distributions — is an integral or an optimization over that
  posterior, which is why Bayesian inference is a computational subject.
- The Beta–Bernoulli, Gamma–Poisson, and Normal–Normal (known variance)
  pairs are **conjugate**: the posterior is available in closed form (Beta,
  Gamma, and Normal respectively), the posterior mean is an exact weighted
  average of the prior mean and the MLE with the prior's pseudo-sample-size
  as weight, and the marginal likelihood and predictive distribution are
  closed-form too (Negative Binomial for Gamma–Poisson, Normal with inflated
  variance for Normal–Normal) — no integration needed anywhere.
- The Gamma–Poisson predictive distribution is visibly overdispersed
  relative to a plug-in Poisson at the posterior mean rate, by an exact
  factor $1+1/(\beta+n)$: real posterior uncertainty about $\theta$ widens
  predictions, a general feature the closed form makes precise here.
- Conjugacy is the exception, not the rule. The rest of this book develops
  general-purpose tools for the same four computations — quadrature,
  Laplace approximation, rejection sampling, importance sampling, and MCMC —
  for models where no such algebraic shortcut exists.
