---
title: "Numerical Quadrature"
format:
  html:
    toc: true
    tbl-colwidths: true
filters:
  - shinylive
editor_options:
  chunk_output_type: console
code-fold: true
---

```{r}
#| label: setup
#| message: false
library(metRology)
```

The previous two chapters found the mode of a log-likelihood or log-posterior
by optimization. Many Bayesian calculations instead need an *area*: a tail
probability, a posterior expectation, or the normalizing constant of a
posterior density. This chapter develops **numerical quadrature** — approximating
an integral by a weighted sum of function values at a finite set of points —
and applies it to the central example of the rest of the book, the marginal
likelihood of a Bayesian model. The next chapter (Laplace's method) revisits
the same example with a method based on the mode instead of a grid.

## The Problem and Transformations

### The general setup

The goal is to approximate
$$
A = \int_a^b f(x)\,dx
$$
for finite $a,b$ by evaluating $f$ at $n$ points $x_1,\ldots,x_n$ and forming a
weighted sum
$$
\tilde A_n = \sum_{i=1}^n w_i\, f(x_i).
$$
The quadrature rules below differ only in how the points and weights are
chosen. Typical statistical uses are probabilities $P(a < X < b)$,
expectations $E[h(X)]$, and normalizing constants such as the marginal
likelihood $P(y) = \int L(\theta)\pi(\theta)\,d\theta$ in Bayesian inference.

### Infinite ranges: the logistic transformation

A grid needs a finite interval, so when the natural range of integration is
$(-\infty,\infty)$ we first change variables. The logistic function maps
$x \in (-\infty,\infty)$ to $y \in (0,1)$,
$$
y = \frac{1}{1+e^{-x}}, \qquad x = t(y) = \log\frac{y}{1-y}, \qquad
t'(y) = \frac{1}{y(1-y)},
$$
and the standard change-of-variables identity gives
$$
\int_{-\infty}^{\infty} f(x)\,dx
 = \int_0^1 f\big(t(y)\big)\,t'(y)\,dy
 = \int_0^1 g(y)\,dy, \qquad
g(y) = \frac{f\big(\log\frac{y}{1-y}\big)}{y(1-y)}.
$$
Any quadrature rule for a finite interval now applies to $g$ on $(0,1)$. Two
consequences matter later in the chapter:

- Points equally spaced in $y$ correspond to points in $x$ that are dense
  near $x=0$ and increasingly sparse in the tails — the grid automatically
  concentrates effort where the logistic curve is steep.
- The factor $1/\{y(1-y)\}$ blows up at $y=0$ and $y=1$, so $g$ is typically
  undefined at the endpoints even when $f$ is perfectly well behaved. This is
  the reason the midpoint rule, which never evaluates the integrand at the
  endpoints of a panel, is the natural choice here.

For a half-line $(0,\infty)$ the same idea applies with $x = -\log(1-y)$ or
$x = y/(1-y)$ in place of the logistic map.

```{r}
#| label: fig-logistic-transform
#| fig-width: 10
#| fig-height: 4.2
#| fig-cap: "Equally spaced points in $y$ (red ticks, right panel) map to unequally
#|   spaced points in $x$ (red ticks, middle panel). The two shaded areas are equal:
#|   the transform changes where we look, not the value of the integral."
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))

curve(1 / (1 + exp(-x)), -6, 6, lwd = 3, xlab = "x", ylab = "y",
      main = expression(y == 1 / (1 + e^-x)))
ys <- seq(0.05, 0.95, by = 0.1)
xs <- log(ys / (1 - ys))
segments(xs, 0, xs, ys, lty = 3, col = "gray50")
segments(-6, ys, xs, ys, lty = 3, col = "gray50")
points(xs, ys, pch = 19, col = "firebrick")

f <- function(x) 0.6 * dnorm(x, -1, 1) + 0.4 * dnorm(x, 2, 1.5)
curve(f, -6, 6, lwd = 3, xlab = "x", ylab = "f(x)", main = "original integrand")
xs2 <- seq(-6, 6, length.out = 400)
polygon(c(xs2, rev(xs2)), c(f(xs2), rep(0, 400)), col = adjustcolor("firebrick", 0.3), border = NA)
rug(xs, col = "firebrick", lwd = 2)

g <- function(y) f(log(y / (1 - y))) / (y * (1 - y))
curve(g, 0.001, 0.999, lwd = 3, n = 501, xlab = "y", ylab = "g(y)",
      main = "transformed integrand on (0, 1)")
ys2 <- seq(0.001, 0.999, length.out = 400)
polygon(c(ys2, rev(ys2)), c(g(ys2), rep(0, 400)), col = adjustcolor("steelblue", 0.3), border = NA)
rug(ys, col = "firebrick", lwd = 2)
```

### Probabilities as a special case

For $X$ with density $\varphi$, e.g. $\varphi(x) = \frac{1}{\sqrt{2\pi}}e^{-x^2/2}$,
$$
P(a < X < b) = \int_a^b \varphi(x)\,dx = \int_{-\infty}^{\infty} \varphi^{(c)}(x)\,dx,
\qquad \varphi^{(c)}(x) = \varphi(x)\,I(a < x < b).
$$
Either integrate $\varphi$ directly over the finite interval $[a,b]$, or
integrate the *truncated* function $\varphi^{(c)}$ over the whole line, letting
the indicator do the cutting. The second view is the more useful one: the same
machinery — transform to $(0,1)$, apply a quadrature rule — computes
$P(X \in S)$ for an arbitrary set $S$ and, more generally, $E[h(X)] = \int
h(x)\varphi(x)\,dx$ for an arbitrary $h$.

## Quadrature Rules on a Finite Interval

All three rules below partition $[a,b]$ into $n$ panels of width $h_n =
(b-a)/n$ and differ only in how $f$ is approximated *within* each panel.

### Midpoint rule

Approximate $f$ on each panel by its value at the panel's midpoint (a constant,
i.e. a rectangle):
$$
\tilde A_n = h_n \sum_{i=1}^n f\!\Big(a + \big(i - \tfrac12\big)h_n\Big).
$$
For smooth $f$ the error is $A - \tilde A_n = \dfrac{(b-a)h_n^2}{24}f''(\xi) =
O(h_n^2)$ for some $\xi \in (a,b)$. The rule never evaluates $f$ at the panel
edges, so in particular it never touches $a$ or $b$ — convenient when $f$ is
singular or undefined at the endpoints, exactly the situation created by the
logistic transform above.

### Trapezoidal rule

Approximate $f$ on $[x_{i-1},x_i]$ by the straight line through the two
endpoint values, so each panel contributes a trapezoid,
$\tfrac12[f(x_{i-1})+f(x_i)](x_i-x_{i-1})$. Summing over panels, interior
points are shared by two neighbouring panels and each contributes once
(not twice) to the total:
$$
\tilde A_n = \tfrac12 h_n\big[f(a) + f(b)\big] + h_n \sum_{i=1}^{n-1} f(a+ih_n).
$$
The error is $A - \tilde A_n = -\dfrac{(b-a)h_n^2}{12}f''(\xi) = O(h_n^2)$ —
the same order as the midpoint rule, but with twice the constant and the
opposite sign. It is exact for linear $f$.

### Simpson's rule

Take $n = 2m$ even. On each *pair* of panels $[x_0,x_1]$ with $x_1-x_0=2h_n$,
replace $f$ by the quadratic through the two endpoints and the midpoint;
integrating that quadratic exactly gives
$$
\int_{x_0}^{x_1} f(x)\,dx \approx \frac{x_1-x_0}{6}
\Big[f(x_0) + 4f\big(\tfrac{x_0+x_1}{2}\big) + f(x_1)\Big].
$$
Summing over the $m$ panel-pairs, with each even interior point shared by two
pairs,
$$
\tilde A_{2m} = \frac{h_n}{3}\left[f(a) + f(b)
 + 4\sum_{i=1}^{m} f\big(a+(2i-1)h_n\big)
 + 2\sum_{i=1}^{m-1} f\big(a+2ih_n\big)\right].
$$
The error is $A - \tilde A_{2m} = -\dfrac{(b-a)h_n^4}{180}f^{(4)}(\xi) =
O(h_n^4)$, two orders faster than the other two rules, and Simpson's rule is
exact for cubics. It can also be written as a weighted average of the other
two rules on the same panel pair: $\text{Simpson} = \tfrac23\,\text{midpoint} +
\tfrac13\,\text{trapezoid}$.

@tbl-rules-compared summarizes the three rules.

| Rule | Approximates $f$ on a panel by | Uses endpoints? | Error | Exact for |
|:---|:---|:---|:---|:---|
| Midpoint | constant at panel midpoint | no | $O(h_n^2)$ | linear |
| Trapezoidal | line through panel endpoints | yes | $O(h_n^2)$ | linear |
| Simpson | quadratic through 3 points (2 panels) | yes | $O(h_n^4)$ | cubic |

: The three quadrature rules compared {#tbl-rules-compared}

### Implementation and accuracy

All three rules are one line of vectorized R each — no explicit loop over
panels is needed:

```{r}
#| label: rules-implementation
midpoint  <- function(f, a, b, n) {
    h <- (b - a) / n
    h * sum(f(a + (seq_len(n) - 0.5) * h))
}
trapezoid <- function(f, a, b, n) {
    h <- (b - a) / n
    x <- a + (0:n) * h
    h * (sum(f(x)) - 0.5 * (f(a) + f(b)))
}
simpson   <- function(f, a, b, n) {
    stopifnot(n %% 2 == 0)
    h <- (b - a) / n
    x <- a + (0:n) * h
    h / 3 * sum(f(x) * c(1, rep(c(4, 2), n / 2 - 1), 4, 1))
}
```

We check all three against a case with a known answer, $P(0.5 < Z < 2)$ for
$Z \sim N(0,1)$:

```{r}
#| label: tbl-rules-accuracy
#| tbl-cap: "Error of each rule with only $n=10$ panels, relative to `pnorm`."
exact <- pnorm(2) - pnorm(0.5)
errs <- sapply(list(midpoint = midpoint, trapezoid = trapezoid, simpson = simpson),
               function(rule) rule(dnorm, 0.5, 2, 10) - exact)
knitr::kable(t(errs), digits = 10, row.names = FALSE)
```

With only 10 panels, Simpson's rule is already accurate to about $10^{-7}$,
several orders of magnitude better than the other two.

```{r}
#| label: fig-convergence
#| fig-width: 7.5
#| fig-height: 4.2
#| fig-cap: "Absolute error versus number of panels, log-log scale. The slopes
#|   are the exponents of $1/n$: $-2$ for midpoint and trapezoid, $-4$ for
#|   Simpson, until floating-point rounding error dominates near $10^{-15}$."
ns  <- 2^(1:9)
err <- sapply(list(midpoint, trapezoid, simpson),
              function(rule) sapply(ns, function(n) abs(rule(dnorm, 0.5, 2, n) - exact)))
matplot(ns, err, type = "b", log = "xy", pch = 19, lty = 1, lwd = 2,
        col = c("steelblue", "firebrick", "forestgreen"),
        xlab = "number of panels n", ylab = "absolute error")
legend("bottomleft", bty = "n", lwd = 2, pch = 19,
       col = c("steelblue", "firebrick", "forestgreen"),
       legend = c("midpoint, slope -2", "trapezoid, slope -2", "Simpson, slope -4"))
```

For a smooth, unimodal integrand like a normal density, Simpson's rule
dominates. The rest of this chapter nonetheless uses the **midpoint rule**,
because the application of interest — Bayesian marginal likelihood after a
logistic transform — produces an integrand that is undefined exactly at the
endpoints $0$ and $1$, which only the midpoint rule avoids evaluating.

## Application: The Marginal Likelihood

### Setup

With likelihood $L(\theta) = \prod_{i=1}^n P(y_i \mid \theta)$ and prior
$\pi(\theta)$, Bayes' rule is
$$
P(\theta \mid y) = \frac{L(\theta)\pi(\theta)}{P(y)}, \qquad
P(y) = \int L(\theta)\pi(\theta)\,d\theta = E_\pi[L(\theta)].
$$
$P(y)$, the **marginal likelihood** (the likelihood averaged over the prior),
is the normalizing constant of the posterior and the basis of Bayes factors
for comparing models. It has no closed form outside conjugate settings, so we
approximate it by quadrature. Because it depends on the *whole* prior — not
just where the likelihood happens to be large — its value is far more
sensitive to prior choice than a posterior mean is, as the next figure shows.

```{r}
#| label: fig-prior-sensitivity
#| fig-width: 10
#| fig-height: 4
#| fig-cap: "The same likelihood (red) combined with three priors (blue).
#|   $P(y)$ is the area under $L(\\theta)\\pi(\\theta)$ (shaded), computed
#|   here by `integrate` for a single parameter as a sanity check."
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
L <- function(t) dnorm(t, 2, 1)
priors <- list(c(2, 1.5, "prior near likelihood"),
               c(-5, 1, "prior far away"),
               c(0, 10, "vague prior N(0, 100)"))
for (pr in priors) {
    m <- as.numeric(pr[1]); s <- as.numeric(pr[2])
    pi_fn <- function(t) dnorm(t, m, s)
    ml <- integrate(function(t) L(t) * pi_fn(t), -Inf, Inf)$value
    curve(L, -9, 8, lwd = 3, col = "firebrick", ylim = c(0, 0.42),
          xlab = expression(theta), ylab = "",
          main = sprintf("%s\nP(y) = %.4f", pr[3], ml))
    curve(pi_fn, add = TRUE, lwd = 3, col = "steelblue")
    ts <- seq(-9, 8, length.out = 300)
    polygon(c(ts, rev(ts)), c(L(ts) * pi_fn(ts), rep(0, 300)),
            col = adjustcolor("gray30", 0.4), border = NA)
}
```

A prior centred where the likelihood is large gives a much bigger marginal
likelihood than one that is far away or merely diffuse — which is exactly why
$P(y)$ is useful for comparing models, and why we must be able to compute it
even when the parameter is multivariate and there is no `integrate`.

### A two-parameter example: mean and variance unknown

Let $y_1,\ldots,y_n \mid \mu, \sigma \overset{iid}{\sim} t_\nu(\mu,\sigma)$, a
location-scale $t$ distribution with $\nu$ degrees of freedom (the special
case $\nu=\infty$ is the normal). Reparametrize the scale as $w = \log \sigma$
so that both parameters range over $(-\infty,\infty)$, and put independent
normal priors on them:
$$
\mu \sim N(\mu_0, \sigma_\mu^2), \qquad w \sim N(w_0, \sigma_w^2).
$$
The marginal likelihood is the two-dimensional integral
$$
P(y) = \int_{-\infty}^{\infty}\!\int_{-\infty}^{\infty}
 L(\mu, w)\,\pi(\mu, w)\, d\mu\, dw,
\qquad L(\mu,w) = \prod_{i=1}^n t_\nu\!\big(y_i;\ \mu,\ e^w\big),
$$
which has no closed form because the normal prior on $w$ is not conjugate to
the likelihood. We handle it with a **product midpoint rule**: apply the
logistic transform to each of $\mu$ and $w$ separately,
$$
\mu = \log\frac{u}{1-u}, \qquad w = \log\frac{v}{1-v}, \qquad
d\mu\,dw = \frac{du\,dv}{u(1-u)\,v(1-v)},
$$
giving
$$
P(y) = \int_0^1\!\int_0^1 g(u,v)\,du\,dv, \qquad
g(u,v) = \frac{L(\mu(u),w(v))\,\pi(\mu(u),w(v))}{u(1-u)\,v(1-v)},
$$
and evaluate $g$ on an $n \times n$ grid $u_j = (j-\tfrac12)h$, $v_k =
(k-\tfrac12)h$ with $h=1/n$:
$$
P(y) \approx h^2 \sum_{j=1}^n \sum_{k=1}^n g(u_j, v_k).
$$
The midpoint rule again earns its place here: $1/\{u(1-u)v(1-v)\} \to \infty$
at the boundary of the square, and the midpoint grid never touches it.

### Computing on the log scale

$L(\theta)$ is a product of $n$ densities and underflows to exactly $0$ in
floating point for even moderate $n$, so every calculation below is carried
out on the log scale, and sums of exponentials are combined with
**log-sum-exp**:
$$
\log \sum_i e^{\ell_i} = M + \log \sum_i e^{\ell_i - M}, \qquad M = \max_i \ell_i.
$$
Subtracting the maximum before exponentiating makes the largest term exactly
$e^0=1$, so nothing overflows; terms far below $M$ underflow harmlessly to
zero instead of corrupting the sum.

```{r}
#| label: log-sum-exp
log_sum_exp <- function(lx) {
    m <- max(lx)
    m + log(sum(exp(lx - m)))
}
```

### An efficient implementation

The functions below compute $\log g(u,v)$ on the whole grid at once with
`outer()`, rather than looping in R or nesting two calls to a generic 1-D
integrator (the nested approach used in earlier drafts of this material is
correct but re-evaluates the logistic transform and re-dispatches a function
call $n^2$ times through two layers of `sapply`; building the grid directly
is both clearer and several times faster for the grid sizes used here).

```{r}
#| label: marlik-functions
### logistic link and its log-derivative, both vectorized
logit_inv     <- function(u) log(u / (1 - u))
log_der_logit <- function(u) -log(u) - log(1 - u)

### log likelihood of a location-scale t sample (df = Inf is the normal)
log_lik <- function(x, mu, w, df = Inf) sum(dt.scaled(x, df, mu, exp(w), log = TRUE))

### log joint prior on (mu, w)
log_prior <- function(mu, w, mu_0, sigma_mu, w_0, sigma_w) {
    dnorm(mu, mu_0, sigma_mu, log = TRUE) + dnorm(w, w_0, sigma_w, log = TRUE)
}

### log of the transformed integrand g(u, v) at one grid point
log_g <- function(u, v, x, mu_0, sigma_mu, w_0, sigma_w, df = Inf) {
    mu <- logit_inv(u); w <- logit_inv(v)
    log_lik(x, mu, w, df) + log_prior(mu, w, mu_0, sigma_mu, w_0, sigma_w) +
        log_der_logit(u) + log_der_logit(v)
}

### log marginal likelihood by the product midpoint rule on an n x n grid
log_marlik_mid <- function(x, mu_0, sigma_mu, w_0, sigma_w, n, df = Inf) {
    h <- 1 / n
    grid <- (seq_len(n) - 0.5) * h
    G <- outer(grid, grid, Vectorize(function(u, v)
        log_g(u, v, x, mu_0, sigma_mu, w_0, sigma_w, df)))
    log_sum_exp(G) + 2 * log(h)
}

### log marginal likelihood by naive Monte Carlo (for validating the above)
log_marlik_mc <- function(x, mu_0, sigma_mu, w_0, sigma_w, iters_mc, df = Inf) {
    mus <- rnorm(iters_mc, mu_0, sigma_mu)
    ws  <- rnorm(iters_mc, w_0, sigma_w)
    v_log_lik <- vapply(seq_len(iters_mc), function(i) log_lik(x, mus[i], ws[i], df), numeric(1))
    log_sum_exp(v_log_lik) - log(iters_mc)
}
```

`log_marlik_mc` draws $\theta$ from the prior and averages the likelihood —
a direct Monte Carlo estimate of $E_\pi[L(\theta)]$ — and serves purely as an
independent check on the quadrature answer, since it makes no use of a grid
or a transformation and so cannot share their failure modes.

### Visualizing the grid and the integrand

```{r}
#| label: fig-grid
#| fig-width: 10
#| fig-height: 4.4
#| fig-cap: "Left: the posterior $L(\\mu,w)\\pi(\\mu,w)$ on its natural scale.
#|   Right: the transformed integrand $g(u,v)$ on the unit square, with the
#|   midpoint grid at $n=20$ overlaid. Equal probability mass under the two
#|   surfaces corresponds to equal volume, even though the shapes differ."
set.seed(1)
y <- rnorm(20, 1, 2)
n <- 20; h <- 1 / n; grid <- (seq_len(n) - 0.5) * h
G  <- outer(grid, grid, Vectorize(function(u, v) log_g(u, v, y, 0, 5, 0, 2)))

mg <- seq(-6, 6, length.out = 80); wg <- seq(-3, 4, length.out = 80)
Gm <- outer(mg, wg, Vectorize(function(m, w) log_lik(y, m, w) + log_prior(m, w, 0, 5, 0, 2)))

par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
contour(mg, wg, exp(Gm - max(Gm)), nlevels = 8,
        xlab = expression(mu), ylab = expression(w == log~sigma^2),
        main = expression(L(mu, w)~pi(mu, w)))
image(grid, grid, exp(G - max(G)), col = hcl.colors(30, "YlOrRd", rev = TRUE),
      xlab = "u", ylab = "v", main = "g(u, v) with midpoint grid")
abline(v = seq(0, 1, by = h), h = seq(0, 1, by = h), col = adjustcolor("gray30", 0.4))
points(expand.grid(grid, grid), pch = ".", cex = 1.5)
```

### Checking the quadrature against Monte Carlo

```{r}
#| label: tbl-quad-vs-mc
#| tbl-cap: "Estimated log marginal likelihood: product midpoint rule versus naive Monte Carlo, for two simulated data sets."
set.seed(2)
x1 <- rt.scaled(50, mean = 2, sd = 2, df = 2)
x2 <- rt.scaled(100, mean = 2, sd = 2, df = Inf)

comparison <- data.frame(
    dataset = c("n=50, t (df=2) data", "n=100, normal data"),
    midpoint_n100 = c(log_marlik_mid(x1, 0, 10, 0, 10, 100),
                       log_marlik_mid(x2, 0, 10, 0, 10, 100)),
    monte_carlo   = c(log_marlik_mc(x1, 0, 10, 0, 10, 1e5),
                       log_marlik_mc(x2, 0, 10, 0, 10, 1e5))
)
knitr::kable(comparison, digits = 4)
```

The two independent methods agree to well within a tenth of a log-unit, which
is reassuring: they share no code path, so agreement is evidence that neither
the grid nor the Monte Carlo sample is too coarse for this posterior.

### Convergence as the grid is refined

```{r}
#| label: tbl-quad-convergence
#| tbl-cap: "Estimated log marginal likelihood as the number of grid points per axis increases, for the $n=100$ normal-data example. By $n=40$ per axis the estimate has stabilized to the precision shown."
ns_grid <- seq(10, 90, by = 10)
est <- vapply(ns_grid, function(k) log_marlik_mid(x2, 0, 10, 0, 10, k), numeric(1))
knitr::kable(data.frame(`grid points per axis` = ns_grid, `log P(y)` = est,
                         check.names = FALSE), digits = 4)
```

### Comparing priors

Holding the data fixed, the marginal likelihood falls as the prior on $\mu$
either narrows around the wrong value or spreads out indefinitely:

```{r}
#| label: tbl-compare-priors
#| tbl-cap: "Log marginal likelihood for data centred at 0: rows are the prior mean $\\mu_0$, columns the prior sd $\\sigma_\\mu$ (w-prior fixed at N(0,1))."
set.seed(3)
x <- rnorm(100)
mu0_vals   <- c(0, -5)
sigma_vals <- c(0.01, 0.1, 1, 10, 100, 1000)
log_P_y <- outer(mu0_vals, sigma_vals,
                  Vectorize(function(m, s) log_marlik_mid(x, m, s, 0, 1, 100)))
rownames(log_P_y) <- mu0_vals
colnames(log_P_y) <- sigma_vals
knitr::kable(log_P_y, digits = 2, align = rep("r", length(sigma_vals)))
```

With $\mu_0 = 0$ (the right neighbourhood for data centred at $0$), the marginal likelihood is highest for a moderately informative prior and degrades slowly as the prior widens. With $\mu_0 = -5$ (the wrong neighbourhood), a *narrow* prior is actively harmful — it insists on a region the data reject — while a wide prior recovers because it eventually puts enough mass back near the truth. This is the general shape of Bayesian model comparison: strong, correct priors win; strong, wrong priors lose; vague priors are safe but not optimal.

### Comparing likelihoods

The marginal likelihood also compares different likelihoods for the same data — here, whether $\nu$ (the degrees of freedom of the $t$ model) matches how the data were actually generated:

```{r}
#| label: tbl-compare-models
#| tbl-cap: "Log marginal likelihood under four values of the t-model degrees of freedom nu, fit to data simulated from a Gaussian and from a heavy-tailed t(df=2)."
dfs <- c(Inf, 2, 1, 0.5)
set.seed(4)
x_normal <- rt.scaled(100, mean = 2, sd = 2, df = Inf)
x_heavy  <- rt.scaled(100, mean = 2, sd = 2, df = 2)
model_compare <- data.frame(
    df_fitted        = dfs,
    normal_data       = vapply(dfs, function(d) log_marlik_mid(x_normal, 0, 10, 0, 10, 100, df = d), numeric(1)),
    heavy_tailed_data = vapply(dfs, function(d) log_marlik_mid(x_heavy,  0, 10, 0, 10, 100, df = d), numeric(1))
)
knitr::kable(model_compare, digits = 3,
             col.names = c("df of fitted model", "data simulated as normal", "data simulated as t(df=2)"))
```

Each column favours the $\nu$ that actually generated it: normal data are
best explained by $\nu=\infty$, and $t(\text{df}=2)$ data are best explained by
$\nu=2$, with the wrong tail weight costing several log-likelihood units
either way. As the earlier prior comparison showed, the prior also shifts
these numbers — but the *ranking* of the correct model over the incorrect
ones survives even under a badly misspecified prior on $(\mu,w)$, which is
why marginal likelihoods remain useful for model comparison despite their
prior-sensitivity.

## Difficulties of Numerical Quadrature


### When the grid misses the posterior's mode {#sec-missed-mode}

The app models a single unknown mean $\mu$. The likelihood
$L(\mu) = N\big(\hat\mu;\,\mu,\,1/n_{\text{eff}}\big)$ treats $\hat\mu$ — the
box you can type into — as the sample mean of $n_{\text{eff}}$ observations
from $N(\mu,1)$; the prior is $\pi(\mu) = N(\mu; 0, 5^2)$. The marginal
likelihood is
$$
P(y) = \int_{-\infty}^{\infty} z(\mu)\,d\mu, \qquad z(\mu) = \pi(\mu)L(\mu).
$$
Because this is a normal-normal conjugate pair, $P(y)$ has a closed form —
but the app instead gets it by numerical integration, shown as the "true"
value in the legend. To approximate $P(y)$ with a finite quadrature rule,
apply the same logistic transform used throughout this chapter,
$u = 1/(1+e^{-\mu})$, so that $d\mu = du/\{u(1-u)\}$ and
$$
P(y) = \int_0^1 \frac{z\big(\mathrm{logit}(u)\big)}{u(1-u)}\,du.
$$
The midpoint rule with $n$ points $u_i = (i-\tfrac12)/n$ turns this into a sum
of value-times-Jacobian terms,
$$
P(y) \approx \frac1n \sum_{i=1}^n \frac{z(\mu_i)}{u_i(1-u_i)},
\qquad \mu_i = \mathrm{logit}(u_i),
$$
and that sum is exactly what the picture shows: drag the slider to change
$n$, and each grey point is one term of it — placed at $\mu_i$ by the guide
lines through the logistic curve, then dropped onto $z(\mu)$ to reveal the
$z(\mu_i)$ factor it contributes, before being scaled by the Jacobian
$1/\{u_i(1-u_i)\}$. Move $\mu$ away from $0$ and every term lands where
$z(\mu)\approx 0$, however many points you add — which is why the quadrature
estimate in the legend drifts far from the true value. This app runs live
only in the HTML edition of this book; readers of other formats can follow
the link below it.

::: {.content-visible when-format="html"}

```{shinylive-r}
#| label: boundary-squeeze-app
#| standalone: true
#| viewerHeight: 600

library(shiny)

ui <- fluidPage(
    titlePanel(NULL),
    sidebarLayout(
        sidebarPanel(
            width = 4,
            numericInput("mu_hat", "mu (data location)", value = 8, step = 1),
            sliderInput("n_pts", "number of grid points in u (over 0 to 1)",
                        min = 2, max = 25, value = 9, step = 1),
            helpText("The grid is a midpoint rule, uniform in u over (0, 1); ",
                     "each point's mu is recovered as logit(u). Try mu = 0 ",
                     "first (the grid sits right under the mass, and the two ",
                     "legend numbers agree closely), then move mu up to 8 or ",
                     "20: the quadrature estimate drifts far from the true ",
                     "value however many grid points you add.")
        ),
        mainPanel(
            width = 8,
            plotOutput("squeeze", height = "560px")
        )
    )
)

server <- function(input, output, session) {

    output$squeeze <- renderPlot({
        n_eff <- 3                                  # likelihood as if 3 obs, sigma = 1
        L     <- function(mu) dnorm(input$mu_hat, mu, 1 / sqrt(n_eff))
        prior <- function(mu) dnorm(mu, 0, 5)
        z     <- function(mu) L(mu) * prior(mu)

        log_sum_exp <- function(lx) { m <- max(lx); m + log(sum(exp(lx - m))) }

        ### this z(mu) is a normal-normal conjugate kernel, so the true log
        ### marginal likelihood is available directly by numerical integration
        log_marlik <- log(integrate(z, -Inf, Inf)$value)

        lo <- min(-8, input$mu_hat - 6)
        hi <- max(14, input$mu_hat + 6)
        mu_grid <- seq(lo, hi, length.out = 500)

        u_pts  <- (seq_len(input$n_pts) - 0.5) / input$n_pts   # midpoint rule, uniform on (0, 1)
        mu_pts <- log(u_pts / (1 - u_pts))          # mu = logit(u)

        ### the midpoint-rule quadrature estimate actually built from this grid:
        ### integral of z(mu) dmu = integral of z(logit(u)) / (u(1-u)) du
        log_z    <- function(mu) dnorm(input$mu_hat, mu, 1 / sqrt(n_eff), log = TRUE) +
            dnorm(mu, 0, 5, log = TRUE)
        h        <- 1 / input$n_pts
        log_marlik_quad <- log_sum_exp(log_z(mu_pts) - log(u_pts * (1 - u_pts))) + log(h)

        par(mar = c(4, 4, 2, 4))
        plot(mu_grid, z(mu_grid), type = "l", lwd = 3, col = "firebrick",
             xlab = expression(mu), ylab = expression(z(mu) == pi(mu) * L(mu)))
        polygon(c(mu_grid, rev(mu_grid)), c(z(mu_grid), rep(0, length(mu_grid))),
                col = adjustcolor("firebrick", 0.15), border = NA)
        segments(mu_pts, 0, mu_pts, z(mu_pts), col = "gray30", lwd = 1.5)
        points(mu_pts, z(mu_pts), pch = 19, col = "gray30")

        par(new = TRUE)
        plot(mu_grid, plogis(mu_grid), type = "l", lwd = 3, col = "steelblue",
             axes = FALSE, xlab = "", ylab = "", ylim = c(0, 1))
        axis(4, col.axis = "steelblue", col.ticks = "steelblue")
        mtext(expression(u == 1 / (1 + e^-mu)), side = 4, line = 2.5, col = "steelblue")

        guide_col <- adjustcolor("gray50", 0.6)
        segments(mu_pts, 0, mu_pts, u_pts, col = guide_col, lty = 3)
        segments(mu_pts, u_pts, max(mu_grid), u_pts, col = guide_col, lty = 3)
        points(mu_pts, u_pts, pch = 19, col = "gray30")

        legend("topleft", bty = "n", lwd = c(3, 3, NA, NA), col = c("firebrick", "steelblue", NA, NA),
               legend = c(expression(z(mu) == pi(mu) * L(mu)),
                          expression(u == 1 / (1 + e^-mu)),
                          as.expression(bquote("true"~log~P(y) == .(round(log_marlik, 3)))),
                          as.expression(bquote("quadrature, this grid"~log~P(y) == .(round(log_marlik_quad, 3))))))
    })
}

shinyApp(ui, server)
```

:::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit08_integral.html#sec-missed-mode>.
:::

### The same failure in the full two-parameter model

The app fixes $\sigma$ to isolate the mechanism, but the product midpoint
rule over $(\mu, w)$ introduced earlier in this chapter — `log_marlik_mid` —
fails the same way once both the mean and the variance are unknown, and
`log_marlik_mc` again serves as the independent check. @tbl-mode-far-away
compares the two for data centred at $\mu=-5$ and at $\mu=-20$ (both with
$\sigma=2$, $n=100$), against a prior on $\mu$ centred at $0$ with
$\sigma_\mu=10$ — reasonably vague, but not recentred on either dataset.

```{r}
#| label: tbl-mode-far-away
#| tbl-cap: "Log marginal likelihood for data centred at mu = -5 and mu = -20 (sigma = 2, n = 100), prior mu_0 = 0, sigma_mu = 10, w_0 = 0, sigma_w = 10: the product midpoint rule (n = 100 points per axis) against a naive Monte Carlo run of 100,000 draws from the prior."
set.seed(6)
x_m5  <- rnorm(100, -5, 2)
x_m20 <- rnorm(100, -20, 2)

mode_far_tbl <- data.frame(
    data        = c("mu = -5", "mu = -20"),
    midpoint    = c(log_marlik_mid(x_m5,  0, 10, 0, 10, 100),
                     log_marlik_mid(x_m20, 0, 10, 0, 10, 100)),
    monte_carlo = c(log_marlik_mc(x_m5,  0, 10, 0, 10, 1e5),
                     log_marlik_mc(x_m20, 0, 10, 0, 10, 1e5))
)
knitr::kable(mode_far_tbl, digits = 3,
             col.names = c("data", "midpoint (n = 100/axis)", "Monte Carlo (N = 1e5)"))
```

At $\mu=-5$ the two methods should still be close, since the logistic grid
still has some points in the region the posterior occupies; at $\mu=-20$ the
posterior for $\mu$ has been pushed close enough to the boundary of the
transformed square that the midpoint estimate can disagree with the Monte
Carlo answer by a large margin on the log scale — the two-dimensional analogue
of exactly what the app above shows in one dimension.

### The curse of dimensionality

A product grid with $n$ points per axis needs $n^d$ evaluations in $d$
dimensions:

| $d$ | $n=20$ | $n=50$ |
|---:|---:|---:|
| 1 | 20 | 50 |
| 2 | 400 | 2,500 |
| 5 | $3.2\times10^6$ | $3.1\times10^8$ |
| 10 | $1.0\times10^{13}$ | $9.8\times10^{16}$ |

: Grid size grows exponentially with dimension {#tbl-curse}

Meanwhile the posterior occupies only a tiny, often tilted, region inside that
grid — most evaluations are wasted on points where $L(\theta)\pi(\theta)
\approx 0$ — and the error of a product rule, expressed in terms of the total
number of evaluations $N=n^d$, degrades to $O(N^{-2/d})$.

```{r}
#| label: fig-curse
#| fig-width: 10
#| fig-height: 4.2
#| fig-cap: "Left: a correlated 2-D posterior inside a 12x12 grid; only the
#|   highlighted points carry non-negligible mass. Right: evaluations needed
#|   for a 20-point-per-axis grid as dimension increases."
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
S <- matrix(c(1, 0.85, 0.85, 1), 2) * 0.15
Si <- solve(S)
grd <- seq(-3, 3, length.out = 150)
Z <- outer(grd, grd, Vectorize(function(a, b) exp(-0.5 * c(a, b) %*% Si %*% c(a, b))))
contour(grd, grd, Z, nlevels = 6, xlab = expression(theta[1]), ylab = expression(theta[2]),
        main = "posterior inside a 12 x 12 grid", drawlabels = FALSE, col = "firebrick", lwd = 2)
gr <- seq(-3, 3, length.out = 13)
abline(v = gr, h = gr, col = adjustcolor("gray30", 0.5))
pts <- expand.grid(gr, gr)
points(pts, pch = 19, cex = 0.5)
inside <- apply(pts, 1, function(p) exp(-0.5 * p %*% Si %*% p)) > 0.01
points(pts[inside, ], pch = 19, col = "firebrick", cex = 0.9)
mtext(sprintf("%d of %d grid points carry non-negligible mass", sum(inside), nrow(pts)),
      side = 1, line = 2.6, cex = 0.85)

plot(1:20, sapply(1:20, function(d) 20^d), log = "y", type = "b", pch = 19,
     xlab = "dimension d", ylab = "evaluations with 20 points per axis")
abline(h = 1e12, lty = 2)
text(3, 3e12, "one trillion", adj = 0, cex = 0.9)
```

**Conclusion.** Quadrature is a reliable, cheap tool for $d \lesssim 3$, and it
is what the rest of this book uses whenever a low-dimensional integral needs
checking against another method. Beyond that, Monte Carlo methods — whose
error is $O(N^{-1/2})$ *regardless of dimension* — take over, at the cost of
losing the very fast convergence rates of Simpson's rule; this trade-off
motivates the sampling methods developed from the next few chapters onward.

## Summary

- Quadrature approximates $\int_a^b f\,dx$ by a weighted sum of $f$ at grid
  points; infinite ranges are handled by a change of variable such as the
  logistic transform, always carrying along the Jacobian $t'(y)$.
- The midpoint and trapezoidal rules have $O(h_n^2)$ error; Simpson's rule has
  $O(h_n^4)$. The midpoint rule's extra virtue — never evaluating the
  integrand at a panel edge — matters whenever the transformed integrand is
  singular at the boundary, as it is here.
- The marginal likelihood $P(y) = \int L(\theta)\pi(\theta)\,d\theta$ is the
  key statistical use case in this book: always compute it on the log scale
  with log-sum-exp, and validate a quadrature answer against an independent
  Monte Carlo estimate.
- Quadrature fails when the integrand is peaked relative to the grid, or when
  a transform squeezes the mass of interest into a region the grid cannot
  resolve; both problems are cured by locating the mass first (optimization)
  and centring the grid or transform on it — the idea taken up systematically
  by the Laplace approximation in the next chapter.
- Product grids cost $n^d$ evaluations, so quadrature is practical only for
  small $d$; higher-dimensional problems require Monte Carlo methods.
