---
title: Univariate Optimization for MLE
code-fold: true
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false
library(latex2exp)   # TeX() renders LaTeX in plot titles, axis labels and legends
set.seed(812)
```

## Maximum Likelihood: Theory

### Likelihood, Score, and Hessian

A model specifies the density or probability of the data $D$ given a parameter $\theta$, written $f(D \mid \theta)$. As a function of $D$ with $\theta$ fixed it is the sampling distribution; as a function of $\theta$ with $D$ observed it is the likelihood. For independent observations $x_1, \ldots, x_n$,
$$L(\theta) = \prod_{i=1}^{n} f(x_i \mid \theta), \qquad \ell(\theta) = \log L(\theta) = \sum_{i=1}^{n} \log f(x_i \mid \theta).$$

We always work with $\ell$: the product underflows to zero for even moderate $n$, and sums are easier to differentiate. The maximum likelihood estimator maximizes $\ell$, and where $\ell$ is smooth with an interior maximum, $\hat\theta$ solves the **score equation**
$$S(\theta) = \ell'(\theta) = 0, \qquad \ell''(\hat\theta) < 0 .$$

Three quantities recur throughout the chapter:

- the **score** $S(\theta) = \ell'(\theta)$, which locates the maximum;
- the **observed information** $J(\theta) = -\ell''(\theta)$, the curvature of the log-likelihood at $\theta$;
- the **expected information** $n I(\theta) = -E_D\left[\ell''(\theta; D)\right]$, its expectation under the model.

The second derivative therefore plays a double role. It determines how a Newton-type algorithm steps toward the maximum, and it determines how precisely the data pin the parameter down. An optimizer that computes the curvature has computed the standard error at the same time.

### Closed Form: the Normal Mean

Let $X_1, \ldots, X_n$ be iid $N(\mu, \sigma_0^{2})$ with $\sigma_0^{2}$ known. Using the identity $\sum_i (x_i - \mu)^2 = \sum_i (x_i - \bar x)^2 + n(\bar x - \mu)^2$,
$$\ell(\mu) = -\frac{n}{2}\log\left(2\pi\sigma_0^{2}\right) - \frac{1}{2\sigma_0^{2}}\sum_{i=1}^{n}(x_i - \mu)^{2} = C - \frac{n}{2\sigma_0^{2}}(\mu - \bar x)^{2} .$$

Differentiating,
$$S(\mu) = \frac{n}{\sigma_0^{2}}\left(\bar x - \mu\right), \qquad \ell''(\mu) = -\frac{n}{\sigma_0^{2}}, \qquad \hat\mu = \bar x .$$

Every feature of the general theory is visible here in exact form. The log-likelihood is a downward parabola, so it has a unique maximum. The score is linear, so the score equation is solved in one line. The curvature $J(\mu) = n/\sigma_0^{2}$ does not depend on $\mu$ or on the data, so observed and expected information coincide, $J = nI$, and both grow linearly with $n$: doubling the sample doubles the sharpness of the peak and halves the variance of $\hat\mu$.

```{r}
#| label: normal-loglik-score-curvature
#| fig-width: 10
#| fig-height: 4
x20 <- rnorm(20, mean = 1); x50 <- rnorm(50, mean = 1)
mu <- seq(0, 2, length.out = 200)

ll_norm <- function(mu, x) vapply(mu, function(m) sum(dnorm(x, m, log = TRUE)), numeric(1))

par(mfrow = c(1, 3), mar = c(4, 4.5, 2.5, 1))

matplot(mu, cbind(ll_norm(mu, x20) - max(ll_norm(mu, x20)),
                  ll_norm(mu, x50) - max(ll_norm(mu, x50))),
        type = "l", lty = 1, lwd = 2, col = c(1, 2),
        xlab = TeX(r"($\mu$)"), ylab = TeX(r"($\ell(\mu) - \ell(\hat{\mu})$)"),
        main = TeX(r"(log-likelihood)"))
legend("bottom", bty = "n", lty = 1, lwd = 2, col = c(1, 2),
       legend = TeX(c(r"($n = 20$)", r"($n = 50$)")))

matplot(mu, cbind(20 * (mean(x20) - mu), 50 * (mean(x50) - mu)),
        type = "l", lty = 1, lwd = 2, col = c(1, 2),
        xlab = TeX(r"($\mu$)"), ylab = TeX(r"($S(\mu)$)"), main = TeX(r"(score)"))
abline(h = 0, lty = 3)

matplot(mu, cbind(rep(20, 200), rep(50, 200)), type = "l", lty = 1, lwd = 2, col = c(1, 2),
        ylim = c(0, 60), xlab = TeX(r"($\mu$)"), ylab = TeX(r"($-\ell''(\mu)$)"),
        main = TeX(r"(curvature)"))
```

The score crosses zero exactly where $\ell$ peaks, and the larger sample gives a more peaked log-likelihood, a steeper score, and a larger curvature.

### Consistency, Asymptotic Normality, and Fisher Information

Under regularity conditions — the support of $f$ not depending on $\theta$, the true value interior to the parameter space, and two derivatives passing under the integral sign — the MLE has three properties as $n \to \infty$, with $\theta_*$ the true parameter.

**Consistency.** $\hat\theta \to \theta_*$ in probability. The reason is that $\ell(\theta)/n$ converges to $E_{\theta_*}[\log f(X \mid \theta)]$, which by the information inequality is maximized at $\theta = \theta_*$: the log-likelihood surface converges to one whose peak is at the truth.

**Asymptotic normality.**
$$\hat\theta \;\dot\sim\; N\left(\theta_*,\ \left[n I(\theta_*)\right]^{-1}\right), \qquad I(\theta) = -E_X\left[\frac{\partial^{2}}{\partial\theta^{2}}\log f(X \mid \theta)\right] = E_X\left[\left(\frac{\partial}{\partial\theta}\log f(X\mid\theta)\right)^{2}\right].$$

A sketch: expand $S(\hat\theta) = 0$ about $\theta_*$ to get $\hat\theta - \theta_* \approx S(\theta_*)/\{-\ell''(\theta_*)\}$. The numerator is a sum of iid mean-zero terms, hence asymptotically normal with variance $nI(\theta_*)$; the denominator converges to $nI(\theta_*)$ by the law of large numbers. Dividing gives variance $[nI(\theta_*)]^{-1}$.

**Asymptotic efficiency.** No consistent estimator has smaller asymptotic variance — the Cramér–Rao bound is attained.

Since $\theta_*$ is unknown, the variance is estimated by substituting $\hat\theta$. Two versions are available: the expected information $nI(\hat\theta)$, which requires the expectation analytically, and the observed information $J(\hat\theta) = -\ell''(\hat\theta)$, which requires only the curvature that a Newton step already computes. By the law of large numbers $J(\hat\theta)/n \approx I(\hat\theta)$, and both give
$$\widehat{\mathrm{SE}}(\hat\theta) = J(\hat\theta)^{-1/2} \quad\text{or}\quad \left[nI(\hat\theta)\right]^{-1/2}, \qquad \hat\theta \pm 1.96\,\widehat{\mathrm{SE}}(\hat\theta).$$

These are asymptotic statements. A flat log-likelihood — small curvature — gives a large standard error, and a log-likelihood that is far from quadratic gives a standard error that should not be trusted at all; the last section of this chapter shows a case of exactly that.

### Demonstration: the Cauchy Location Model

The Cauchy density $f(x \mid \theta) = 1/\{\pi[1 + (x - \theta)^2]\}$ has no closed-form MLE:
$$\ell(\theta) = -n\log\pi - \sum_{i=1}^{n}\log\left(1 + (x_i - \theta)^{2}\right), \qquad
S(\theta) = \sum_{i=1}^{n}\frac{2(x_i - \theta)}{1 + (x_i - \theta)^{2}} .$$

The score equation is a polynomial equation of degree $2n - 1$. Plotting $\ell$ for samples of increasing size shows the peak sharpening around the truth, which is consistency and growing information in one picture.

```{r}
#| label: cauchy-loglik-by-n
#| fig-width: 9
#| fig-height: 5
log_like_cauchy <- function(theta, x) sum(dcauchy(x, theta, log = TRUE))

thetas <- seq(15, 25, length.out = 200)
n_set <- c(50, 100, 200)

ll_cauchy <- sapply(n_set, function(n) {
  d <- rcauchy(n, location = 20)
  sapply(thetas, log_like_cauchy, x = d)
})
ll_cauchy <- sweep(ll_cauchy, 2, apply(ll_cauchy, 2, max))   # align peaks at 0

par(mar = c(4, 4.5, 2.5, 1))
matplot(thetas, ll_cauchy, type = "l", lty = 1, lwd = 2, col = c(1, 2, 4), ylim = c(-20, 0),
        xlab = TeX(r"($\theta$)"), ylab = TeX(r"($\ell(\theta) - \ell(\hat{\theta})$)"),
        main = TeX(r"(Cauchy log-likelihood, $\theta_* = 20$)"))
abline(v = 20, lty = 3)
legend("bottomright", bty = "n", lty = 1, lwd = 2, col = c(1, 2, 4),
       legend = TeX(sprintf(r"($n = %d$)", n_set)))
```

The variability of the MLE across repeated samples is the sampling distribution the theory describes. Repeating the experiment 50 times at two sample sizes shows the curves clustering more tightly, and their maxima concentrating, as $n$ grows.

```{r}
#| label: cauchy-loglik-variability
#| fig-width: 10
#| fig-height: 4.5
#| cache: true
loglik_curves <- function(n, nrep = 50)
  replicate(nrep, {
    d <- rcauchy(n, location = 20)
    lc <- sapply(thetas, log_like_cauchy, x = d)
    lc - max(lc)
  })

curves_small <- loglik_curves(20)
curves_large <- loglik_curves(200)

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
for (obj in list(list(m = curves_small, n = 20), list(m = curves_large, n = 200)))
  matplot(thetas, obj$m, type = "l", lty = 1, col = adjustcolor(1, 0.35), ylim = c(-10, 0),
          xlab = TeX(r"($\theta$)"), ylab = TeX(r"($\ell(\theta) - \ell(\hat{\theta})$)"),
          main = TeX(sprintf(r"(sample size $n = %d$)", obj$n)))

### sampling distribution of the MLE at the two sample sizes
mle_at <- function(m) thetas[apply(m, 2, which.max)]
rbind(`n = 20`  = c(mean = mean(mle_at(curves_small)), sd = sd(mle_at(curves_small))),
      `n = 200` = c(mean = mean(mle_at(curves_large)), sd = sd(mle_at(curves_large))))
```

The standard deviation falls by roughly the factor $\sqrt{10}$ predicted by $[nI(\theta_*)]^{-1/2}$. Two features of these curves also warn us about computation: for small $n$ the log-likelihood has several local maxima, and its shape away from the peak is far from quadratic. Numerical optimization is required, and it must be used with care.

## Univariate Optimization Algorithms

### Three Equivalent Formulations

The same task can be posed three ways, and the algorithms below attack different versions of it:

- maximize $\ell(\theta)$;
- minimize $f(\theta) = -\ell(\theta)$;
- find a root of $S(\theta) = \ell'(\theta)$.

Root-finders are usually faster because they use derivative information, but they locate any stationary point, including minima and saddle points. Direct optimizers need only function values and always move toward a maximum, but converge more slowly. Which to prefer depends on whether derivatives are available and whether the function is smooth.

Each method below is illustrated with a schematic showing its first few iterations. To keep the pictures comparable they all use the same small problem: the per-observation log-likelihood of the zero-truncated Poisson model of the next section, with $\bar x$ fixed at 1.8. The root of its score is marked in every figure.

```{r}
#| label: demo-problem
xbar_d <- 1.8                                             # fixed sample mean

ll_d <- function(l) xbar_d * log(l) - l - log(1 - exp(-l))          # log-likelihood
S_d  <- function(l) xbar_d / l - 1 / (1 - exp(-l))                  # score
dS_d <- function(l) -xbar_d / l^2 + exp(-l) / (1 - exp(-l))^2       # its derivative
I_d  <- function(l) 1 / (l * (1 - exp(-l))) - exp(-l) / (1 - exp(-l))^2   # information
g_d  <- function(l) xbar_d * (1 - exp(-l))                          # fixed-point map

lstar <- uniroot(S_d, c(0.2, 5))$root
c(root = lstar, contraction_factor = xbar_d * exp(-lstar))
```

### Fixed-Point (Simple) Iteration

Rearrange the score equation algebraically into the form $\theta = g(\theta)$ and iterate $\theta_{k+1} = g(\theta_k)$. If $g$ maps an interval into itself and $|g'(\theta^{*})| < 1$ at the fixed point, the mean value theorem gives
$$\left|\theta_{k+1} - \theta^{*}\right| = \left|g(\theta_k) - g(\theta^{*})\right| \le \left|g'(\xi)\right| \left|\theta_k - \theta^{*}\right| ,$$
so the error contracts by roughly the constant factor $|g'(\theta^{*})|$ at each step. This is **linear convergence**: a fixed number of digits is gained per iteration, and the closer $|g'|$ is to 1 the slower the progress.

```{r}
#| label: fixed-point
fixed_point <- function(g, x0, tol = 1e-10, maxit = 500) {
  for (k in 1:maxit) {
    x1 <- g(x0)
    if (abs(x1 - x0) < tol) return(c(root = x1, iter = k))
    x0 <- x1
  }
  stop("not converged")
}
```

The mechanism is easiest to see as a **cobweb**: from $\lambda_k$ on the diagonal, move vertically to the curve $g$ to obtain $\lambda_{k+1}$, then horizontally back to the diagonal to start the next step. The iterates walk toward the intersection of $g$ with the $45^{\circ}$ line, and the shallower $g$ is at that crossing — the smaller $|g'|$ — the larger each stride toward it.

```{r}
#| label: scheme-fixed-point
#| fig-width: 8
#| fig-height: 5
lam <- seq(0.2, 3, length.out = 300)

par(mar = c(4, 4.5, 2.5, 1))
plot(lam, g_d(lam), type = "l", lwd = 2, xlim = c(0.2, 3), ylim = c(0.2, 3),
     xlab = TeX(r"($\lambda_k$)"), ylab = TeX(r"($g(\lambda_k) = \lambda_{k+1}$)"),
     main = TeX(r"(fixed-point iteration: the cobweb)"))
abline(0, 1, lty = 2)

x0 <- 2.8
for (k in 1:5) {
  x1 <- g_d(x0)
  segments(x0, x0, x0, x1, col = 2, lwd = 2)          # up or down to the curve
  segments(x0, x1, x1, x1, col = 2, lwd = 2)          # across to the diagonal
  points(x0, x1, pch = 19, col = 2, cex = 0.7)
  if (k <= 3) text(x0, 0.25, TeX(sprintf(r"($\lambda_%d$)", k - 1)), col = 2, cex = 0.9)
  x0 <- x1
}
points(lstar, lstar, pch = 19, cex = 1.2)
text(lstar, lstar, TeX(r"($\hat{\lambda}$)"), pos = 4)
```

The method needs no derivatives, but the rearrangement must be found by hand, and different rearrangements of the same equation can converge at different speeds or diverge outright. Geometrically, a rearrangement with $|g'| > 1$ at the crossing produces a cobweb that spirals outward instead of inward.

### Bisection

If $S$ is continuous and $S(a)$ and $S(b)$ have opposite signs, a root lies in $[a, b]$. Evaluate $S$ at the midpoint and retain whichever half still changes sign.

```{r}
#| label: bisection
bisection <- function(S, a, b, tol = 1e-10) {
  if (S(a) * S(b) > 0) stop("endpoints do not bracket a root")
  k <- 0
  while (b - a > tol) {
    m <- (a + b) / 2
    if (S(a) * S(m) <= 0) b <- m else a <- m
    k <- k + 1
  }
  c(root = (a + b) / 2, iter = k)
}
```

The schematic below draws the bracket at each iteration as a bar beneath the score curve. The dot is the midpoint, and the dotted line carries it up to the curve, where its sign decides which half survives.

```{r}
#| label: scheme-bisection
#| fig-width: 8
#| fig-height: 5
par(mar = c(4, 4.5, 2.5, 1))
plot(lam, S_d(lam), type = "l", lwd = 2, ylim = c(-1.6, 3.5),
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($S(\lambda)$)"),
     main = TeX(r"(bisection: each step halves the bracket)"))
abline(h = 0, lty = 3)
points(lstar, 0, pch = 19, cex = 1.2)

a <- 0.3; b <- 3
ylev <- seq(-0.5, -1.4, length.out = 4)
for (k in 1:4) {
  m <- (a + b) / 2
  segments(a, ylev[k], b, ylev[k], col = 4, lwd = 3)
  points(c(a, b), rep(ylev[k], 2), pch = "|", col = 4)
  segments(m, ylev[k], m, S_d(m), lty = 3, col = 2)
  points(m, ylev[k], pch = 19, col = 2, cex = 0.8)
  text(0.25, ylev[k], k, cex = 0.8, col = 4)
  if (S_d(a) * S_d(m) <= 0) b <- m else a <- m
}
```

The bracket halves at every step, so after $k$ iterations the error is at most $(b - a)/2^{k}$ and the iteration count is known in advance: about $\log_2\{(b-a)/\text{tol}\}$, regardless of the shape of $S$. Bisection cannot diverge, which makes it the natural fallback when a faster method misbehaves. R's `uniroot` uses Brent's method, which retains the bracketing guarantee while interpolating for speed.

### Newton–Raphson

#### Derivation

Approximate the score by its tangent line at $\theta_k$ and solve for where the tangent crosses zero:
$$S(\theta) \approx S(\theta_k) + S'(\theta_k)(\theta - \theta_k) = 0
\implies \theta_{k+1} = \theta_k - \frac{S(\theta_k)}{S'(\theta_k)} = \theta_k - \frac{\ell'(\theta_k)}{\ell''(\theta_k)} .$$

Equivalently, replace $\ell$ by its second-order Taylor expansion at $\theta_k$ and jump to the maximum of that parabola. For an exactly quadratic $\ell(\theta) = -a(\theta - u)^2 + C$ with $a > 0$ we have $\ell' = -2a(\theta - u)$ and $\ell'' = -2a$, so
$$\theta_{k+1} = \theta_k - \frac{-2a(\theta_k - u)}{-2a} = u$$
in one step from any starting value. The normal-mean example of the previous section is exactly of this form: Newton converges immediately.

The left panel below traces three steps. At each iterate, the tangent to the score is followed down to the horizontal axis, and where it lands is the next iterate. As the curve straightens near the root, the tangent becomes an excellent approximation and the steps land almost exactly on the answer.

The right panel contrasts one Newton step with one step of the method of scoring from the same point. Newton uses the actual slope $S'(\lambda_k)$; scoring uses $-nI(\lambda_k)$, the slope averaged over hypothetical datasets. The two lines differ away from the root and coincide as $\lambda_k$ approaches it.

```{r}
#| label: scheme-newton
#| fig-width: 10
#| fig-height: 4.8
par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))

### three Newton steps
plot(lam, S_d(lam), type = "l", lwd = 2, ylim = c(-1, 3.5),
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($S(\lambda)$)"),
     main = TeX(r"(Newton-Raphson: follow the tangent)"))
abline(h = 0, lty = 3)
points(lstar, 0, pch = 19, cex = 1.2)

x0 <- 0.45
for (k in 1:3) {
  x1 <- x0 - S_d(x0) / dS_d(x0)
  segments(x0, S_d(x0), x1, 0, col = 2, lwd = 2)
  segments(x1, 0, x1, S_d(x1), lty = 3, col = "grey40")
  points(c(x0, x1), c(S_d(x0), 0), pch = 19, col = 2, cex = 0.7)
  text(x0, S_d(x0), TeX(sprintf(r"($\lambda_%d$)", k - 1)), pos = 3, col = 2, cex = 0.9)
  x0 <- x1
}

## Newton step versus scoring step from the same point
plot(lam, S_d(lam), type = "l", lwd = 2, ylim = c(-1, 3.5),
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($S(\lambda)$)"),
     main = TeX(r"(observed versus expected curvature)"))
abline(h = 0, lty = 3)
points(lstar, 0, pch = 19, cex = 1.2)

x0 <- 0.45
x_nr <- x0 - S_d(x0) / dS_d(x0)          # slope  S'(lambda)
x_fs <- x0 + S_d(x0) / I_d(x0)           # slope -nI(lambda)
segments(x0, S_d(x0), x_nr, 0, col = 2, lwd = 2)
segments(x0, S_d(x0), x_fs, 0, col = 4, lwd = 2, lty = 5)
points(c(x0, x_nr, x_fs), c(S_d(x0), 0, 0), pch = 19, cex = 0.7, col = c(1, 2, 4))
legend("topright", bty = "n", lwd = 2, lty = c(1, 5), col = c(2, 4),
       legend = TeX(c(r"(Newton: slope $S'(\lambda_k)$)", r"(scoring: slope $-nI(\lambda_k)$)")))
```

### Advantage: quadratic convergence

Let $e_k = \theta_k - \hat\theta$ and expand $S(\hat\theta) = 0$ about $\theta_k$:
$$0 = S(\theta_k) - S'(\theta_k)e_k + \tfrac12 S''(\xi)e_k^{2}
\implies e_{k+1} = e_k - \frac{S(\theta_k)}{S'(\theta_k)} = \frac{S''(\xi)}{2S'(\theta_k)}\,e_k^{2} .$$

The error is squared at each step, so the number of correct digits roughly doubles. Two conditions appear in the constant: $S'(\hat\theta) \ne 0$, meaning the maximum is a genuine peak and not a flat spot, and $S''$ bounded near the root. Asymptotic normality says that log-likelihoods are approximately quadratic near the MLE, which is why Newton is so effective there.

```{r}
#| label: newton
newton <- function(S, dS, x0, tol = 1e-10, maxit = 100, trace = FALSE) {
  for (k in 1:maxit) {
    x1 <- x0 - S(x0) / dS(x0)
    if (trace) cat(sprintf("iter %2d: %.10f\n", k, x1))
    if (!is.finite(x1)) stop("diverged")
    if (abs(x1 - x0) < tol) return(c(root = x1, iter = k))
    x0 <- x1
  }
  stop("not converged")
}
```

### Problem: divergence

Newton uses only local information, so the quadratic model can be badly wrong away from the maximum. Three failures occur:

- **wrong curvature** — where $\ell'' > 0$ the model is a bowl, and the step heads toward a minimum or off to infinity;
- **overshooting** — where the score is nearly flat, $S'(\theta_k) \approx 0$ produces an enormous step, possibly outside the parameter space;
- **multiple roots** — the iteration converges to whichever stationary point it happens to reach, which need not be the global maximum.

The single-observation Cauchy likelihood is the standard illustration. With one observation at zero, $f(\theta) = -\ell(\theta) = \log(1 + \theta^{2})$ and $s(\theta) = f'(\theta) = 2\theta/(1 + \theta^{2})$, whose only root is $\theta = 0$. The score rises, turns over at $\theta = 1$, and decays to zero in the tails — exactly the flat region that defeats Newton.

```{r}
#| label: cauchy-single-obs
#| fig-width: 10
#| fig-height: 4.5
lik <- function(t) log(1 + t^2)                    # negative log-likelihood
s   <- function(t) 2 * t / (1 + t^2)               # its derivative
ds  <- function(t) 2 * (1 - t^2) / (1 + t^2)^2

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
curve(lik, -5, 5, n = 400, lwd = 2, xlab = TeX(r"($\theta$)"), ylab = TeX(r"($-\ell(\theta)$)"),
      main = TeX(r"(negative log-likelihood)"))
curve(s, -5, 5, n = 400, lwd = 2, xlab = TeX(r"($\theta$)"), ylab = TeX(r"($s(\theta)$)"),
      main = TeX(r"(derivative, with the flat tails)"))
abline(h = 0, lty = 3); abline(v = c(-1, 1), col = 2, lty = 2)
```

Tracing the iterates from three starting values shows the boundary at $|\theta_0| = 1$, where the derivative $s$ turns over:

```{r}
#| label: newton-divergence-trace
newton_trace <- function(s, ds, x0, iters = 8) {
  path <- numeric(iters + 1); path[1] <- x0
  for (k in 1:iters) {
    x0 <- x0 - s(x0) / ds(x0)
    path[k + 1] <- if (is.finite(x0)) x0 else NA
  }
  path
}

starts <- c(0.5, 0.8, 1.2)
traces <- vapply(starts, function(v) newton_trace(s, ds, v), numeric(9))
colnames(traces) <- sprintf("start = %.1f", starts)
rownames(traces) <- paste("iter", 0:8)
signif(traces, 5)
```

From $\theta_0 = 0.5$, inside the region where the quadratic model is adequate, Newton converges quadratically to zero. From $\theta_0 = 0.8$ the first step already overshoots into the tail, and from $\theta_0 = 1.2$ — beyond the inflection point of $s$ — the iterates run away, each step landing further out where the derivative is flatter still.

```{r}
#| label: newton-animation
#| eval: false
## Interactive versions of the same three trajectories (requires the animation
## package and a graphics device that can write animations)
library(animation)
ani.options(interval = 1, nmax = 50)
newton.method(FUN = s, init = 0.5, rg = c(-1.5, 1.5))
newton.method(FUN = s, init = 0.8, rg = c(-8, 1.5))
newton.method(FUN = s, init = 1.2, rg = c(-2, 20))
```

### Safeguards

Three remedies cover most cases. Start from a robust estimate — the sample median for a location model. Apply **step halving**: accept the full Newton step only if it improves the log-likelihood, otherwise halve it until it does. Or replace the observed curvature by the expected information, giving the **method of scoring**,
$$\theta_{k+1} = \theta_k + \frac{S(\theta_k)}{n I(\theta_k)} ,$$
in which the divisor is positive by construction, so every step ascends. Near the MLE $nI(\hat\theta) \approx -\ell''(\hat\theta)$, so scoring inherits Newton's speed while being far more stable away from the optimum. Its cost is that the expectation must be available in closed form.

```{r}
#| label: safe-newton
safe_newton <- function(ll, S, dS, x0, lower = -Inf, upper = Inf,
                        tol = 1e-10, maxit = 100, max_halve = 40) {
  f0 <- ll(x0)
  for (k in 1:maxit) {
    step <- -S(x0) / dS(x0)
    if (dS(x0) > 0) step <- sign(S(x0))          # wrong curvature: just move uphill
    for (h in 0:max_halve) {                     # halve until feasible and improving
      x1 <- x0 + step / 2^h
      if (x1 > lower && x1 < upper && is.finite(ll(x1)) && ll(x1) >= f0) break
    }
    if (abs(x1 - x0) < tol) return(c(root = x1, iter = k, halvings = h))
    x0 <- x1; f0 <- ll(x1)
  }
  stop("not converged")
}
```

### Golden Section Search

Golden section minimizes $f = -\ell$ directly, using **no derivative at all**, which makes it the method of choice when $\ell$ is not differentiable or its derivative is unreliable. Keep a bracket $[x_l, x_u]$ containing the minimum of a unimodal $f$, with two interior points placed at the golden ratio $g = (3 - \sqrt5)/2 \approx 0.382$ of the way in from each end. Comparing the two interior values discards one end, and the ratio is chosen precisely so that the surviving interior point is correctly positioned for the next iteration: one new evaluation per step, with the bracket shrinking by the factor $0.618$.

```{r}
#| label: golden
golden <- function(f, brack.int, eps = 1e-4, ...) {
  g <- (3 - sqrt(5)) / 2
  xl <- min(brack.int); xu <- max(brack.int)
  tmp <- g * (xu - xl)
  xmu <- xu - tmp; xml <- xl + tmp
  fl <- f(xml, ...); fu <- f(xmu, ...)
  while (abs(xu - xl) > (1e-5 + abs(xl)) * eps) {
    if (fl < fu) {
      xu <- xmu; xmu <- xml; fu <- fl
      xml <- xl + g * (xu - xl); fl <- f(xml, ...)
    } else {
      xl <- xml; xml <- xmu; fl <- fu
      xmu <- xu - g * (xu - xl); fu <- f(xmu, ...)
    }
  }
  if (fl < fu) xml else xmu
}
```

The schematic makes the reuse visible. Each bar is the current bracket with its two interior points; the point that survives into the next iteration is drawn in blue and sits at exactly the right position there, so only the red point has to be evaluated afresh.

```{r}
#| label: scheme-golden
#| fig-width: 8
#| fig-height: 5
f_d <- function(l) -ll_d(l)
gr <- (3 - sqrt(5)) / 2

par(mar = c(4, 4.5, 2.5, 1))
plot(lam, f_d(lam), type = "l", lwd = 2, ylim = c(0.2, 1.25),
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($-\ell(\lambda)$)"),
     main = TeX(r"(golden section: one new evaluation per step)"))
points(lstar, f_d(lstar), pch = 19, cex = 1.2)

xl <- 0.3; xu <- 3
ylev <- seq(0.44, 0.24, length.out = 3)
for (k in 1:3) {
  xml <- xl + gr * (xu - xl); xmu <- xu - gr * (xu - xl)
  segments(xl, ylev[k], xu, ylev[k], col = 4, lwd = 3)
  points(c(xl, xu), rep(ylev[k], 2), pch = "|", col = 4)
  keep_lower <- f_d(xml) < f_d(xmu)
  points(xml, ylev[k], pch = 19, cex = 0.9, col = if (keep_lower) 4 else 2)
  points(xmu, ylev[k], pch = 19, cex = 0.9, col = if (keep_lower) 2 else 4)
  text(0.25, ylev[k], k, cex = 0.8, col = 4)
  if (keep_lower) xu <- xmu else xl <- xml
}
```

The first test is the negative log-likelihood of the classical genetic linkage model, with cell counts $(1997, 906, 904, 32)$ and probabilities $\left((2+\theta)/4, (1-\theta)/4, (1-\theta)/4, \theta/4\right)$; it is smooth, so any method would work. The second, $f(x) = |x - 10|$, is not differentiable at its minimum, so no root-finder applies — golden section handles it without modification.

```{r}
#| label: golden-tests
#| fig-width: 10
#| fig-height: 4.5
f1 <- function(theta) -1997 * log(2 + theta) - 1810 * log(1 - theta) - 32 * log(theta)
f2 <- function(x) abs(x - 10)

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
curve(f1, 0.001, 0.999, n = 400, lwd = 2, xlab = TeX(r"($\theta$)"), ylab = TeX(r"($-\ell(\theta)$)"),
      main = TeX(r"(genetic linkage model)"))
abline(v = golden(f1, c(0.001, 0.999)), col = 2, lty = 2, lwd = 2)

curve(f2, 0, 20, n = 400, lwd = 2, xlab = TeX(r"($x$)"), ylab = TeX(r"($|x - 10|$)"),
      main = TeX(r"(not differentiable at the minimum)"))
abline(v = golden(f2, c(0, 20)), col = 2, lty = 2, lwd = 2)

c(linkage = golden(f1, c(0.001, 0.999)), abs_value = golden(f2, c(0, 20)))
```

### Summary of the Methods

| method | needs | convergence | fails when |
|:--|:--|:--|:--|
| fixed point | a rearrangement $\theta = g(\theta)$ | linear, factor $\lvert g'\rvert$ | $\lvert g'\rvert \ge 1$ |
| bisection | $S$ and a sign-changing bracket | linear, factor $1/2$ | no bracket available |
| Newton–Raphson | $S$ and $S'$ | quadratic near the root | flat or wrong curvature |
| method of scoring | $S$ and $I(\theta)$ | linear to quadratic | $E[\cdot]$ unavailable |
| golden section | $\ell$ only, and a bracket | linear, factor $0.618$ | $\ell$ not unimodal |

## Example: Poisson Data With Zeros Unobserved

### The Model

Counts are often recorded only when positive: a study of hospital stays sees admitted patients, a survey of litters sees litters that exist. If $Y \sim \mathrm{Poisson}(\lambda)$ and we observe $X = Y \mid Y > 0$,
$$P(x \mid \lambda) = \frac{e^{-\lambda}\lambda^{x}}{x!}\cdot\frac{1}{1 - e^{-\lambda}}, \qquad x = 1, 2, \ldots$$

Dropping the factorial terms, which do not involve $\lambda$,
$$\ell(\lambda) = \left(\sum_i x_i\right)\log\lambda - n\left[\lambda + \log\left(1 - e^{-\lambda}\right)\right],$$
$$S(\lambda) = \frac{\sum_i x_i}{\lambda} - \frac{n}{1 - e^{-\lambda}}, \qquad
S'(\lambda) = -\frac{\sum_i x_i}{\lambda^{2}} + \frac{n e^{-\lambda}}{\left(1 - e^{-\lambda}\right)^{2}} .$$

Since $E(X) = \lambda/(1 - e^{-\lambda})$, taking $-E[S'(\lambda)]$ gives the expected information
$$n I(\lambda) = \frac{n}{\lambda\left(1 - e^{-\lambda}\right)} - \frac{n e^{-\lambda}}{\left(1 - e^{-\lambda}\right)^{2}} .$$

The score equation has no closed-form root, but the likelihood is unimodal — an ideal problem on which to compare all the methods against a single known answer.

```{r}
#| label: nzp-functions
## log likelihood, omitting factorial terms; vectorized in lambda
nzp.log.likelihood <- function(n, lambda)
  sum(n) * log(lambda) - length(n) * (lambda + log(1 - exp(-lambda)))

nzp.scorefunction <- function(n, lambda)
  sum(n) / lambda - length(n) / (1 - exp(-lambda))

nzp.dscore <- function(n, lambda)
  -sum(n) / lambda^2 + length(n) * exp(-lambda) / (1 - exp(-lambda))^2

nzp.information <- function(n, lambda)
  length(n) / (lambda * (1 - exp(-lambda))) - length(n) * exp(-lambda) / (1 - exp(-lambda))^2

## simulated data: Poisson counts with the zeros discarded
x <- rpois(100, lambda = 1.5)
xp <- x[x > 0]

c(n_observed = length(xp), sample_mean = mean(xp), true_lambda = 1.5)
```

The sample mean overshoots 1.5, because the zeros that would have pulled it down were never recorded.

```{r}
#| label: nzp-loglik-score
#| fig-width: 10
#| fig-height: 4.5
lambdas <- seq(0.1, 8, length.out = 300)

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
plot(lambdas, nzp.log.likelihood(xp, lambdas), type = "l", lwd = 2,
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($\ell(\lambda)$)"), main = TeX(r"(log-likelihood)"))
plot(lambdas, nzp.scorefunction(xp, lambdas), type = "l", lwd = 2,
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($S(\lambda)$)"), main = TeX(r"(score)"))
abline(h = 0, lty = 3)
```

### The Three Iterations

Each method below returns the whole sequence of iterates together with the log-likelihood attained, so the trajectories can be compared directly.

**Simple iteration** rearranges $S(\lambda) = 0$ as
$$\frac{\bar x}{\lambda} = \frac{1}{1 - e^{-\lambda}} \iff \lambda = \bar x\left(1 - e^{-\lambda}\right) = g(\lambda),$$
with contraction factor $g'(\lambda) = \bar x e^{-\lambda}$.

```{r}
#| label: nzp-iterations
nzp.simple.iteration <- function(n, r, lambda0 = mean(n)) {
  mean.n <- mean(n); lambda <- numeric(r + 1); lambda[1] <- lambda0
  for (i in 1:r) lambda[i + 1] <- mean.n * (1 - exp(-lambda[i]))
  data.frame(lambda = lambda, log.lik = nzp.log.likelihood(n, lambda))
}

nzp.newton.raphson <- function(n, r, lambda0 = mean(n)) {
  mean.n <- mean(n); lambda <- numeric(r + 1); lambda[1] <- lambda0
  for (i in 1:r) {
    e <- exp(-lambda[i])
    lambda[i + 1] <- lambda[i] -
      (mean.n / lambda[i] - 1 / (1 - e)) / (e / (1 - e)^2 - mean.n / lambda[i]^2)
  }
  data.frame(lambda = lambda, log.lik = nzp.log.likelihood(n, lambda))
}

nzp.method.of.scoring <- function(n, r, lambda0 = mean(n)) {
  mean.n <- mean(n); lambda <- numeric(r + 1); lambda[1] <- lambda0
  for (i in 1:r) {
    e <- exp(-lambda[i])
    lambda[i + 1] <- lambda[i] -
      (mean.n / lambda[i] - 1 / (1 - e)) / ((e / (1 - e) - 1 / lambda[i]) / (1 - e))
  }
  data.frame(lambda = lambda, log.lik = nzp.log.likelihood(n, lambda))
}
```

### Comparison From Several Starting Values

The three methods are run for 15 iterations from a sensible start, a start far below, and a start far above.

```{r}
#| label: nzp-compare-starts
starts <- c(mean(xp), 0.1, 100)

trace_tab <- function(fun, r = 15)
  vapply(starts, function(s) fun(xp, r, s)$lambda, numeric(r + 1))

paths <- list(simple = trace_tab(nzp.simple.iteration),
              newton = trace_tab(nzp.newton.raphson),
              scoring = trace_tab(nzp.method.of.scoring))

### first six iterates from each starting value
for (m in names(paths)) {
  cat("\n", m, "\n")
  tab <- paths[[m]][1:6, ]
  dimnames(tab) <- list(paste("iter", 0:5), sprintf("start = %.4g", starts))
  print(signif(tab, 6))
}
```

The pattern is the one the theory predicts. Newton–Raphson takes the fewest iterations from a good start, and its digits lock in from the left as the error is squared each step. Simple iteration converges from every start, including the extreme ones, but slowly and at a steady rate. The method of scoring is intermediate: always ascending, and nearly as fast as Newton once close.

```{r}
#| label: nzp-convergence-plot
#| fig-width: 9
#| fig-height: 5
lhat <- nzp.newton.raphson(xp, 30, mean(xp))$lambda[31]

err <- sapply(paths, function(p) abs(p[, 1] - lhat))   # from the mean(xp) start

par(mar = c(4, 4.5, 2.5, 1))
matplot(0:15, pmax(err, 1e-17), log = "y", type = "b", pch = 19, lty = 1, col = c(4, 1, 2),
        xlab = TeX(r"(iteration $k$)"), ylab = TeX(r"($|\lambda_k - \hat{\lambda}|$)"),
        main = TeX(r"(error decay from $\lambda_0 = \bar{x}$)"))
legend("bottomleft", bty = "n", pch = 19, lty = 1, col = c(4, 1, 2),
       legend = c("simple iteration", "Newton-Raphson", "method of scoring"))
```

On a log scale a linear method is a straight line; Newton's curve bends downward until it reaches the floor set by machine precision.

```{r}
#| label: nzp-builtin-check
S_xp  <- function(l) nzp.scorefunction(xp, l)
dS_xp <- function(l) nzp.dscore(xp, l)
ll_xp <- function(l) nzp.log.likelihood(xp, l)

c(newton      = unname(newton(S_xp, dS_xp, mean(xp))["root"]),
  fixed_point = unname(fixed_point(function(l) mean(xp) * (1 - exp(-l)), mean(xp))["root"]),
  bisection   = unname(bisection(S_xp, 0.1, 20)["root"]),
  golden      = golden(function(l) -ll_xp(l), c(0.1, 20), eps = 1e-8),
  uniroot     = uniroot(S_xp, c(0.1, 20), tol = 1e-10)$root,
  optimize    = optimize(function(l) -ll_xp(l), c(0.1, 20), tol = 1e-10)$minimum)
```

All six agree, which is the point of running them together: the methods differ in cost and robustness, not in the answer.

### Standard Error From the Curvature

The optimizer has already produced everything inference needs.

```{r}
#| label: nzp-standard-error
se_obs <- 1 / sqrt(-nzp.dscore(xp, lhat))
se_exp <- 1 / sqrt(nzp.information(xp, lhat))

c(mle = lhat, se_observed = se_obs, se_expected = se_exp)
lhat + c(-1.96, 1.96) * se_obs                     # 95% Wald interval
```

```{r}
#| label: nzp-quadratic-approx
#| fig-width: 9
#| fig-height: 5
lam <- seq(lhat - 4 * se_obs, lhat + 4 * se_obs, length.out = 300)
quad <- -0.5 * (-nzp.dscore(xp, lhat)) * (lam - lhat)^2

par(mar = c(4, 4.5, 2.5, 1))
plot(lam, ll_xp(lam) - ll_xp(lhat), type = "l", lwd = 2,
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($\ell(\lambda) - \ell(\hat{\lambda})$)"),
     main = TeX(r"(log-likelihood and its quadratic approximation at $\hat{\lambda}$)"))
lines(lam, quad, col = 2, lwd = 2, lty = 2)
abline(h = -1.92, col = 4, lty = 3)
abline(v = lhat + c(-1.96, 1.96) * se_obs, col = 4, lty = 3)
legend("bottom", bty = "n", lwd = 2, lty = c(1, 2), col = c(1, 2),
       legend = TeX(c(r"($\ell(\lambda)$)", r"(quadratic from $J(\hat{\lambda})$)")))
```

The parabola implied by the observed information tracks the log-likelihood closely over the width of the interval, so the Wald interval is trustworthy here.

### Evaluating the Naive Estimator

All of this machinery is worth the effort only if the MLE beats the easy alternative. The naive estimator ignores the truncation and reports $\bar x$. Simulating many truncated samples gives the sampling distributions of both estimators.

```{r}
#| label: nzp-mle-vs-naive
#| cache: true
#| fig-width: 10
#| fig-height: 4.5
mle_nzp <- function(n)
  uniroot(function(l) nzp.scorefunction(n, l), c(1e-6, 50), tol = 1e-10)$root

sim_nzp <- function(lambda, n_gen = 100, B = 500)
  vapply(1:B, function(b) {
    y <- rpois(n_gen, lambda); n <- y[y > 0]
    c(naive = mean(n), mle = mle_nzp(n))
  }, c(naive = 0, mle = 0))

for (lambda_true in c(0.5, 1.5)) {
  est <- sim_nzp(lambda_true)
  tab <- t(apply(est, 1, function(e)
    c(bias = mean(e) - lambda_true, sd = sd(e), rmse = sqrt(mean((e - lambda_true)^2)))))
  cat("\nlambda =", lambda_true, "\n"); print(round(tab, 4))
}

est <- sim_nzp(1.5)
par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
for (m in c("naive", "mle")) {
  hist(est[m, ], breaks = 25, col = "grey90",
       xlim = range(est), xlab = TeX(r"($\hat{\lambda}$)"),
       main = TeX(sprintf(r"(%s, bias $= %.3f$)", m, mean(est[m, ]) - 1.5)))
  abline(v = 1.5, col = 2, lwd = 2, lty = 2)
}
```

The naive estimator is biased upward, and severely so when $\lambda$ is small, since that is when most of the probability mass sits on the discarded zero. Its bias does not shrink with sample size: it reflects the wrong model, not too little data. The MLE is centred on the truth with a standard deviation matching the information-based standard error of the previous section. The optimization was the means; the correct estimator was the end.

## Postscript: Interval Data and a Flat Likelihood

Not every likelihood is well approximated by a parabola, and when it is not, the information-based standard error misleads. Suppose each Poisson observation is known only to lie in an interval $[l_i, h_i]$. Then
$$\ell(\lambda) = \sum_{i=1}^{n} \log \Pr\left(l_i \le X_i \le h_i \mid \lambda\right)
= \sum_{i=1}^{n} \log \sum_{x = l_i}^{h_i} \frac{e^{-\lambda}\lambda^{x}}{x!} .$$

The inner sum is computed on the log scale with the log-sum-exp trick to avoid underflow, and the maximization is delegated to `nlm`, which returns a numerical Hessian and hence a standard error.

```{r}
#| label: interval-poisson-functions
log_sum_exp <- function(lx) { mlx <- max(lx); mlx + log(sum(exp(lx - mlx))) }

poisson_log_likelihood <- function(lambda, low, high)
  sum(vapply(seq_along(low),
             function(i) log_sum_exp(dpois(low[i]:high[i], lambda, log = TRUE)),
             numeric(1)))
### equivalently, and faster: sum(log(ppois(high, lambda) - ppois(low - 1, lambda)))

poisson_mle <- function(low, high) {
  if (length(low) != length(high)) stop("low and high have different lengths")
  if (any(floor(low) != low) || any(floor(high) != high)) stop("interval ends are not integers")

  fit <- nlm(function(l) -poisson_log_likelihood(l, low, high),
             p = (mean(low) + mean(high)) / 2, hessian = TRUE)
  est <- fit$estimate; sd <- sqrt(1 / fit$hessian[1, 1])
  list(est = est, sd = sd, CI = est + c(-1.96, 1.96) * sd)
}
```

```{r}
#| label: interval-poisson-fits
#| cache: true
pn <- rpois(1000, lambda = 20)

tight <- poisson_mle(pmax(pn - 1, 0), pn + 4)      # narrow intervals: much information
wide  <- poisson_mle(pmax(pn - 5, 0), pn + 5)     # wide intervals: little information

rbind(tight = c(est = tight$est, sd = tight$sd, tight$CI),
      wide  = c(est = wide$est,  sd = wide$sd,  wide$CI))
```

```{r}
#| label: interval-poisson-loglik
#| fig-width: 10
#| fig-height: 4.5
lam_grid <- seq(1, 50, by = 0.5)

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
for (iv in list(list(lo = pmax(pn - 1, 0), hi = pn + 4, lab = "narrow intervals"),
                list(lo = pmax(pn - 50, 0), hi = pn + 5, lab = "wide intervals"))) {
  lp <- vapply(lam_grid, poisson_log_likelihood, numeric(1), low = iv$lo, high = iv$hi)
  plot(lam_grid, lp - max(lp), type = "l", lwd = 2, ylim = c(-50, 0),
       xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($\ell(\lambda) - \ell(\hat{\lambda})$)"),
       main = TeX(sprintf(r"(%s)", iv$lab)))
  abline(h = -1.92, col = 4, lty = 3)
}
```

With narrow intervals the log-likelihood is sharply peaked and the standard error is small and reliable. With wide intervals, most of which include the lower boundary at zero, the log-likelihood is nearly flat over a broad range: the curvature at the maximum is tiny, the reported standard error is large, and — because the curve is visibly asymmetric and non-quadratic — even that large value understates the true uncertainty. The information-based standard error is an asymptotic approximation resting on local quadratic behaviour, and this is what its failure looks like. When the log-likelihood is this shape, report a likelihood-based interval, read off where $\ell$ drops by 1.92, instead of a Wald interval.

## A Practical Newton-Raphson Algorithm for Finding $\sqrt{a}$

**Derivation.** To find $\sqrt{a}$, take $f(x) = x^2 - a$, whose positive root is $\sqrt a$. Then $f'(x) = 2x$, and the Newton step is

$$x_{k+1} = x_k - \frac{f(x_k)}{f'(x_k)} = x_k - \frac{x_k^{2} - a}{2x_k} = \frac{1}{2}\left(x_k + \frac{a}{x_k}\right).$$

The final form is the Babylonian rule: if $x_k$ is too small then $a/x_k$ is too large, and the root lies between them, so average the pair. Newton's tangent line lands exactly on the midpoint.

The error follows the general quadratic law. Writing $e_k = x_k - \sqrt a$,

$$e_{k+1} = \frac{1}{2}\left(x_k + \frac{a}{x_k}\right) - \sqrt a = \frac{\left(x_k - \sqrt a\right)^{2}}{2x_k} = \frac{e_k^{2}}{2x_k},$$

so the number of correct digits roughly doubles each step.

```{r}
#| label: babylonian-sqrt
babylonian <- function(a, x0 = 1, iters = 6)
  Reduce(function(x, k) (x + a / x) / 2, 1:iters, init = x0, accumulate = TRUE)

traces <- sapply(c(10, 20, 50), babylonian)
dimnames(traces) <- list(paste("iter", 0:6), sprintf("sqrt(%d)", c(10, 20, 50)))

print(traces, digits = 12)
sqrt(c(10, 20, 50))                       # for comparison
signif(abs(traces - rep(sqrt(c(10, 20, 50)), each = 7)), 3)    # errors
```

Starting from $x_0 = 1$ in every case, the first two steps do the coarse work — the larger $a$ is, the further the opening leap — and from the third step on the error squares: roughly $10^{-1}$, then $10^{-3}$, then $10^{-6}$, then machine precision. Six iterations suffice for all three, though $\sqrt{50}$ spends one more step than $\sqrt{10}$ recovering from the poor start.