---
title: "Importance Sampling"
format:
  html:
    toc: true
    tbl-colwidths: true
filters:
  - shinylive
editor_options:
  chunk_output_type: console
code-fold: true
---

```{r}
#| label: setup
#| message: false
library(ars)
```

Rejection sampling turns draws from an envelope $g \ge f$ into exact draws from
$f$, but every rejected proposal is wasted, and in the tails of a poor
envelope almost every proposal is rejected. **Importance sampling** keeps
every proposal instead of discarding most of them: draw from an easier
function $g$ and attach a *weight* to each draw that corrects for sampling
from the wrong distribution. The price is that the resulting sample is no
longer an exact iid sample from $f$ — it is a *weighted* sample — but nothing
is ever thrown away, and no envelope condition $g \ge f$ is required.

## The Reweighting Idea

### The goal

As in the last two chapters, $f(\theta)$ is an **unnormalized** function of
interest — typically $f(\theta) = L(\theta)\pi(\theta)$ for a likelihood $L$
and prior $\pi$ — and we want two kinds of integral:
$$
A = E_f\big[a(\theta)\big] = \frac{\int a(\theta)\,f(\theta)\,d\theta}{\int f(\theta)\,d\theta},
\qquad\qquad
C_f = \int f(\theta)\,d\theta,
$$
a posterior expectation and a normalizing constant (the marginal likelihood
when $f = L\pi$). Plain Monte Carlo needs draws from $f/C_f$, which we
generally cannot produce.

Suppose instead we can draw from another function $g(\theta)$ (also possibly
unnormalized, with $C_g = \int g(\theta)\,d\theta$). Wherever $f$ puts mass,
$g$ must too:
$$
\{\theta : g(\theta) > 0\} \ \supseteq\ \{\theta : f(\theta) > 0\}.
$$
Unlike rejection sampling, $g$ need **not** dominate $f$ everywhere — there is
no requirement $g \ge f$. This is what makes importance sampling applicable
in situations where no usable envelope exists.

### Two estimators

Define the **importance weight** $w(\theta) = f(\theta)/g(\theta)$. For any
$a$,
$$
E_f\big[a(\theta)\big]
= \int a(\theta)\,\frac{f(\theta)}{C_f}\,d\theta
= \frac{C_g}{C_f}\, E_g\big[a(\theta)\,w(\theta)\big],
$$
and taking $a \equiv 1$ gives $C_f/C_g = E_g[w(\theta)]$. Combining the two,
$$
E_f\big[a(\theta)\big] = \frac{E_g\big[a(\theta)\,w(\theta)\big]}{E_g\big[w(\theta)\big]}.
$$
Draw $\theta_1,\ldots,\theta_S \overset{iid}{\sim} g/C_g$ and compute
$w_s = w(\theta_s)$. Two estimators follow directly:

- **Ratio of normalizing constants**: $\widehat{C_f/C_g} = \frac1S\sum_s w_s$,
  unbiased for $C_f/C_g$; if $g$ is already a normalized density ($C_g=1$)
  this estimates $C_f$ itself.
- **Self-normalized expectation**: $\hat A = \sum_{s=1}^S a(\theta_s)\,\tilde w_s$,
  with normalized weights $\tilde w_s = w_s / \sum_j w_j$ summing to $1$. This
  turns the proposal sample into a *weighted sample from $f$*: any unknown
  constants in $f$ and $g$ cancel in the ratio, so $\hat A$ needs only $f$ and
  $g$ up to a proportionality constant. It is a ratio of two averages and so
  is biased with bias $O(1/S)$, but consistent.

### Effective sample size

$\hat A$ is a weighted average of $a(\theta_s)$: draws where $f \gg g$ count
more, draws where $f \ll g$ count less. How many *equally weighted* draws is
the weighted sample worth?
$$
S_{\text{eff}} = \frac{1}{\displaystyle\sum_{s=1}^S \tilde w_s^2}.
$$
If the weights are all equal, $\tilde w_s = 1/S$ and $S_{\text{eff}} =
1/(S \cdot 1/S^2) = S$: the proposal is shaped just like the target, and every
draw counts fully. If one weight dominates all the others, $\tilde w_s
\approx 1$ for that draw and $S_{\text{eff}} \approx 1$: only one draw is
doing any work. A variance heuristic makes this precise:
$\text{Var}_g(\hat A) \approx \text{Var}_f(a)/S_{\text{eff}}$, the same form as
for an iid sample of size $S_{\text{eff}}$ from $f$ itself.

The catch, central to this chapter, is that $S_{\text{eff}}$ is computed
**only from the weights actually observed**. A proposal that never visits the
part of $\theta$-space where $f$ and $g$ disagree most will report a
reassuring $S_{\text{eff}}$ while being badly wrong — this is exactly the
subject of the "bad stories" below.

```{r}
#| label: ess-helpers
### effective sample size from raw (possibly unnormalized) weights
ess <- function(w) sum(w)^2 / sum(w^2)
### numerically stable log(sum(exp(lx)))
log_sum_exp <- function(lx) { m <- max(lx); m + log(sum(exp(lx - m))) }
### effective sample size computed from LOG weights (avoids under/overflow)
log_ess <- function(lw) exp(2 * log_sum_exp(lw) - log_sum_exp(2 * lw))
```

## Example 1: Same Support, but a Mode Is Missed

$$
f(\theta) = \tfrac12\varphi(\theta+d) + \tfrac12\varphi(\theta-d), \qquad
g_{\sigma}(\theta) = \tfrac{1}{\sigma}\varphi\!\left(\tfrac{\theta+d}{\sigma}\right),
\qquad d = 5 .
$$

Both are positive on all of $\mathbb{R}$, so the support requirement holds. But
for $\sigma$ of order one the proposal concentrates around $-d$ and never
produces draws near the second mode at $+d$. For $\theta$ near $-d$ and
$\sigma=1$,
$$
w(\theta) = \frac{f(\theta)}{g_{1}(\theta)}
          = \tfrac12 + \tfrac12\,\frac{\varphi(\theta-d)}{\varphi(\theta+d)}
          = \tfrac12 + \tfrac12 e^{2d\theta} \;\approx\; \tfrac12 ,
$$
so every observed weight is nearly equal, $\tilde w_{i}\approx 1/S$ and
$S_{\text{eff}}\approx S$, while $\widehat{C_{f}/C_{g}}\approx\tfrac12$ against a
truth of $1$ and $\widehat{E}_{f}(\theta)\approx -d$ against a truth of $0$. The
draws that would expose the error carry $w\approx\tfrac12 e^{2d\theta}\gg 1$ near
$\theta=+d$ and have probability $\Phi(-2d)\approx 10^{-23}$ under $g_{1}$: they
never appear.

Both panels of @fig-missed-mode are drawn on the same $\theta$ axis. The lower
panel plots the *normalized weight function* $\theta\mapsto w(\theta)/\sum_{j}w_{j}$,
whose values at the sampled points are exactly the plotted stems, so the sample
can be seen sitting on the flat part of a function that rises without bound to
the right.

```{r}
#| label: fig-missed-mode
#| fig-cap: "Top: the bimodal target f (firebrick) against the proposal g (steelblue)
#|   on a log density scale, with a rug of S = 2000 draws. Bottom: normalized
#|   importance weights at the sampled points, with the normalized weight function
#|   through them. Every observed weight is essentially 1/2, so S_eff is close to
#|   its nominal value while the estimate of C<sub>f</sub>/C<sub>g</sub> is half the truth."
#| fig-width: 9
#| fig-height: 7
d <- 5; sg <- 1; S <- 2000
set.seed(11)

## log of the mixture density, computed stably
log_f <- function(th, d) {
    a <- dnorm(th, -d, log = TRUE); b <- dnorm(th, d, log = TRUE)
    m <- pmax(a, b)
    m + log(exp(a - m) + exp(b - m)) - log(2)
}

## exact log E_g(w^2) for this pair; infinite when 2 - 1/sigma^2 <= 0,
## i.e. when sigma <= 1/sqrt(2).  Since E_g(w) = 1 exactly, the ideal
## effective fraction is 1 / E_g(w^2).
log_Ew2 <- function(d, sg) {
    A <- 2 - 1 / sg^2
    if (A <= 0) return(Inf)
    lterm <- function(a, b) {
        m <- (a + b) / 2
        B <- 2 * m + d / sg^2
        C <- 2 * m^2 - d^2 / sg^2
        dnorm(a, b, sqrt(2), log = TRUE) + log(sg) + 0.5 * log(2) +
            0.5 * log(2 * pi / A) - 0.5 * (C - B^2 / A)
    }
    lv <- c(lterm(-d, -d), log(2) + lterm(-d, d), lterm(d, d))
    m  <- max(lv)
    m + log(sum(exp(lv - m))) - log(4)
}

## ---- sample and weight ------------------------------------------------
th   <- rnorm(S, -d, sg)
logw <- log_f(th, d) - dnorm(th, -d, sg, log = TRUE)
mx   <- max(logw); v <- exp(logw - mx)
Chat <- mean(v) * exp(mx)                       # estimate of C_f / C_g, truth 1
se   <- sd(v) / sqrt(S) * exp(mx)
wn   <- v / sum(v)                              # normalized weights
ess  <- 1 / sum(wn^2)
Ehat <- sum(wn * th)                            # estimate of E_f(theta), truth 0
ess_x <- 100 * exp(-log_Ew2(d, sg))             # exact effective fraction

## ---- shared axis ------------------------------------------------------
xr <- range(c(th, -d - 4, d + 4))
xs <- seq(xr[1], xr[2], length.out = 1000)
fx <- exp(log_f(xs, d)); gx <- dnorm(xs, -d, sg)
lwc <- log_f(xs, d) - dnorm(xs, -d, sg, log = TRUE) - (mx + log(sum(v)))

layout(matrix(1:2, 2), heights = c(3, 2))

## ---- top panel: the two densities ------------------------------------
par(mar = c(0.4, 5, 3, 1))
yl <- c(1e-10, max(fx, gx) * 2)
plot(xs, fx, type = "n", log = "y", ylim = yl, xlim = xr, xaxt = "n",
     xlab = "", ylab = "density (log scale)",
     main = "bimodal target and a proposal centred on one mode")
lines(xs, fx, lwd = 4, col = "grey25")
lines(xs, gx, lwd = 3, col = "grey60")
rug(th, col = adjustcolor("grey45", 0.35))
legend("bottomright", bty = "n", cex = 1.0, text.col = "navy",
       lwd = c(4, 3, NA, NA, NA), col = c("grey25", "grey60", NA, NA, NA),
       legend = c("target  f", sprintf("proposal  g,  sd = %.2f", sg),
                  sprintf("C_f/C_g:   true 1,   est. %.3f", Chat),
                  sprintf("95%% CI  (%.3f, %.3f)", Chat - 1.96 * se, Chat + 1.96 * se),
                  sprintf("E_f(theta):  true 0,   est. %.2f", Ehat)))

## ---- bottom panel: normalized weights --------------------------------
par(mar = c(4.5, 5, 0.4, 1))
wl <- c(min(wn) / 5, max(wn) * 5)
plot(th, wn, type = "n", log = "y", ylim = wl, xlim = xr,
     xlab = expression(theta), ylab = "normalized weight (log scale)")
lines(xs, exp(pmin(lwc, 700)), lwd = 2, col = "grey25")
segments(th, wl[1], th, wn, col = adjustcolor("grey45", 0.5))
abline(h = 1 / S, lty = 2, col = "grey55", lwd = 2)
legend("bottomright", bty = "n", cex = 1.0, text.col = "navy",
       lty = c(2, 1, NA, NA), lwd = c(2, 2, NA, NA),
       col = c("grey55", "grey25", NA, NA),
       legend = c("1 / S", "normalized weight function",
                  sprintf("observed ESS = %.1f%% of S", 100 * ess / S),
                  sprintf("exact ESS = %.3g%% of S", ess_x)))
```

The last two legend entries are the whole lesson. The observed effective sample
size is a statistic of the weights that actually appeared, and it reports
near-perfect efficiency. The exact effective fraction $1/E_{g}(w^{2})$, available
here in closed form because both densities are Gaussian mixtures, is smaller by
some forty orders of magnitude. At $\sigma=1$ the variance is finite, so the
central limit theorem does apply and the estimator is consistent; it would
simply need $S$ of order $10^{43}$ before its sampling distribution resembled a
normal one. Finiteness of the variance is itself fragile: the exact second
moment diverges once $2-\sigma^{-2}\le 0$, that is for every
$\sigma\le 1/\sqrt{2}$.

### An interactive view

@fig-missed-mode-app varies $\sigma$ with everything else held fixed. Three regimes appear.
Below $\sigma=1/\sqrt{2}$ the variance is infinite and the exact effective
fraction is reported as zero. Between there and roughly $\sigma=3$ the failure is
silent in the sense above: observed $S_{\text{eff}}$ near $S$, estimate near
$\tfrac12$. Past $\sigma\approx4$ the proposal is over-dispersed enough to place
draws near $+d$; the observed $S_{\text{eff}}$ then *drops*, which looks like
deterioration and is in fact the diagnostic finally working, and the estimate
moves onto the truth. The cure for a missed mode is a proposal that is too
diffuse rather than too well matched.

::: {#fig-missed-mode-app}

```{shinylive-r}
#| label: app-missed-mode
#| standalone: true
#| viewerHeight: 720
library(shiny)

d <- 5

log_f <- function(th, d) {
    a <- dnorm(th, -d, log = TRUE); b <- dnorm(th, d, log = TRUE)
    m <- pmax(a, b)
    m + log(exp(a - m) + exp(b - m)) - log(2)
}

log_Ew2 <- function(d, sg) {
    A <- 2 - 1 / sg^2
    if (A <= 0) return(Inf)
    lterm <- function(a, b) {
        m <- (a + b) / 2
        B <- 2 * m + d / sg^2
        C <- 2 * m^2 - d^2 / sg^2
        dnorm(a, b, sqrt(2), log = TRUE) + log(sg) + 0.5 * log(2) +
            0.5 * log(2 * pi / A) - 0.5 * (C - B^2 / A)
    }
    lv <- c(lterm(-d, -d), log(2) + lterm(-d, d), lterm(d, d))
    m  <- max(lv)
    m + log(sum(exp(lv - m))) - log(4)
}

ui <- fluidPage(
    titlePanel("A missed mode: the proposal's standard deviation"),
    sidebarLayout(
        sidebarPanel(
            width = 3,
            sliderInput("sg", HTML("proposal sd &sigma;"), min = 0.25, max = 8,
                        value = 1, step = 0.05),
            selectInput("S", "sample size  S",
                        choices = c(500, 2000, 10000), selected = 2000),
            numericInput("seed", "seed", value = 11, min = 1, step = 1),
            actionButton("new", "New sample"),
            tags$hr(style = "margin:8px 0"),
            htmlOutput("readout")
        ),
        mainPanel(width = 9, plotOutput("panels", height = "660px"))
    )
)

server <- function(input, output) {

    sim <- reactive({
        S  <- as.integer(input$S); sg <- input$sg
        set.seed(input$seed + 1000 * input$new)

        th   <- rnorm(S, -d, sg)
        logw <- log_f(th, d) - dnorm(th, -d, sg, log = TRUE)
        mx   <- max(logw); v <- exp(logw - mx)
        wn   <- v / sum(v)

        list(th = th, wn = wn, S = S, sg = sg,
             Chat = mean(v) * exp(mx),
             se   = sd(v) / sqrt(S) * exp(mx),
             ess  = 1 / sum(wn^2),
             Ehat = sum(wn * th),
             lsum = mx + log(sum(v)),
             ess_x = 100 * exp(-log_Ew2(d, sg)))
    })

    output$panels <- renderPlot({
        s  <- sim(); th <- s$th; sg <- s$sg
        xr <- range(c(th, -d - 4, d + 4))
        xs <- seq(xr[1], xr[2], length.out = 1000)
        fx <- exp(log_f(xs, d)); gx <- dnorm(xs, -d, sg)
        lwc <- log_f(xs, d) - dnorm(xs, -d, sg, log = TRUE) - s$lsum

        layout(matrix(1:2, 2), heights = c(3, 2))

        par(mar = c(0.4, 5, 3, 1))
        yl <- c(1e-10, max(fx, gx) * 2)
        plot(xs, fx, type = "n", log = "y", ylim = yl, xlim = xr, xaxt = "n",
             xlab = "", ylab = "density (log scale)",
             main = "bimodal target and a proposal centred on one mode")
        lines(xs, fx, lwd = 4, col = "grey25")
        lines(xs, gx, lwd = 3, col = "grey60")
        rug(th, col = adjustcolor("grey45", 0.35))
        legend("bottomright", bty = "n", cex = 1.05, text.col = "navy",
               lwd = c(4, 3, NA, NA, NA),
               col = c("grey25", "grey60", NA, NA, NA),
               legend = c("target  f", sprintf("proposal  g,  sd = %.2f", sg),
                          sprintf("C_f/C_g:   true 1,   est. %.3f", s$Chat),
                          sprintf("95%% CI  (%.3f, %.3f)",
                                  s$Chat - 1.96 * s$se, s$Chat + 1.96 * s$se),
                          sprintf("E_f(theta):  true 0,   est. %.2f", s$Ehat)))

        par(mar = c(4.5, 5, 0.4, 1))
        wn <- s$wn
        wl <- c(min(wn) / 5, max(wn) * 5)
        plot(th, wn, type = "n", log = "y", ylim = wl, xlim = xr,
             xlab = expression(theta), ylab = "normalized weight (log scale)")
        lines(xs, exp(pmin(lwc, 700)), lwd = 2, col = "grey25")
        segments(th, wl[1], th, wn, col = adjustcolor("grey45", 0.5))
        abline(h = 1 / s$S, lty = 2, col = "grey55", lwd = 2)
        legend("bottomright", bty = "n", cex = 1.05, text.col = "navy",
               lty = c(2, 1, NA, NA), lwd = c(2, 2, NA, NA),
               col = c("grey55", "grey25", NA, NA),
               legend = c("1 / S", "normalized weight function",
                          sprintf("observed ESS = %.1f%% of S", 100 * s$ess / s$S),
                          sprintf("exact ESS = %.3g%% of S", s$ess_x)))
    })

    output$readout <- renderUI({
        s <- sim()
        HTML(sprintf(
            "<b>C<sub>f</sub>/C<sub>g</sub></b><br>true 1, est. %.4f<br>
             <b>95%% CI</b> (%.4f, %.4f)<br>
             <b>E<sub>f</sub>(theta)</b><br>true 0, est. %.3f<br>
             <b>observed ESS</b> %.1f%%<br>
             <b>exact ESS</b> %.3g%%<br>
             <b>Var<sub>f</sub>(w)</b> %s",
            s$Chat, s$Chat - 1.96 * s$se, s$Chat + 1.96 * s$se, s$Ehat,
            100 * s$ess / s$S, s$ess_x,
            if (s$sg <= 1 / sqrt(2)) "infinite" else "finite"))
    })
}

shinyApp(ui, server)
```

Interactive version of @fig-missed-mode. The proposal's standard deviation
$\sigma$ is under slider control and both panels are redrawn for each setting.
For $\sigma \le 1/\sqrt{2}$ the weight variance is infinite; near $\sigma = 1$
the failure is silent, with the observed effective sample size close to $S$
while the estimate of $C_f/C_g$ sits at one half; past $\sigma \approx 4$ the
proposal reaches the second mode, the observed effective sample size falls, and
the estimate moves onto the truth.

:::


## Example 2: Tail Probability of a Student-$t$ Distribution {#sec-ars-tnorm}

The target is a standard Student-$t_{\nu}$ distribution, with the degrees of
freedom $\nu$ a **tuning parameter of the target** that controls how heavy
its tail is: $\nu=1$ is the standard Cauchy, and $\nu\to\infty$ recovers the
standard normal. The goal is $P(c_{1}<X<c_{2})$ for an interval $(c_{1},c_{2})$
— possibly one-sided, $c_{2}=\infty$ — estimated from draws of a
truncated-normal proposal generated by adaptive rejection sampling (ARS).

As in the ARS section of the previous chapter, $\log\varphi$ is concave, and
restricting to $(c_{1},c_{2})$ (or one-sided, $(c_{1},\infty)$) preserves
concavity, so the `ars` package applies directly: `rtnorm_ars()` below just
hands it the log-density and its derivative, plus the truncation bounds.

```{r}
#| label: rtnorm-ars
## ---------------------------------------------------------------------
##  rtnorm_ars(): draw from N(mu, sigma^2) truncated to (lo, hi) (hi may
##  be Inf) via the ars package. log-phi is concave, so ars() applies
##  directly; x0 gives it three starting abscissae inside (lo, hi), with
##  the rightmost past mu when hi = Inf so the upper piece has negative
##  slope and stays integrable.
## ---------------------------------------------------------------------
rtnorm_ars <- function(n, mu = 0, sigma = 1, lo = 0, hi = Inf) {
    logf   <- function(x) dnorm(x, mu, sigma, log = TRUE)
    fprima <- function(x) -(x - mu) / sigma^2
    if (is.finite(hi)) {
        x0 <- lo + (hi - lo) * c(0.1, 0.5, 0.9)
    } else {
        right <- max(mu, lo) + 3 * sigma
        x0 <- sort(unique(c(lo + 0.1 * sigma, max(lo + 0.2 * sigma, mu + 1e-6), right)))
    }
    ars(n, f = logf, fprima = fprima, x = x0,
        lb = TRUE, xlb = lo, ub = is.finite(hi), xub = if (is.finite(hi)) hi else 0)
}
```

The `ars` package is itself available inside the browser — webR fetches it
from CRAN's WebAssembly mirror the first time `library(ars)` runs — so the
interactive app below calls this exact function, not a hand-rolled
substitute.

The contrast with the target is the point of the exercise: ARS needs
log-concavity, and $\log f_{t_{\nu}}$ is simply **not concave**, for any
finite $\nu$ — it is convex out in the tails, and only the limit
$\nu\to\infty$ (the normal) is concave everywhere. So ARS cannot sample the
target directly at any finite $\nu$; the truncated-normal proposal above is a
substitute, not a target-side match, which is exactly why the sample needs
importance weights at all.

### The estimator, and when its variance is infinite

A one-sided tail $(c,\infty)$ is a bad target for this proposal, so consider
instead a general interval $(c_{1},c_{2})$, with $c_{2}=\infty$ allowed as a
special case. Let $g$ be the normalized truncated normal on $(c_{1},c_{2})$
and $f(x)=f_{t_{\nu}}(x)I(c_{1}<x<c_{2})$. Draw $X_{1},\dots,X_{S}\sim g$,
weight each by $w(x)=f(x)/g(x)$, and average:
$$
\widehat{C} \;=\; \frac{1}{S}\sum_{i=1}^{S} w(X_{i}).
$$
This is unbiased for $C_{f}=P(c_{1}<X<c_{2})$ for *any* $g$ that is positive
wherever $f$ is — that is the whole mechanism of importance sampling, and it
does not care what $c_{1},c_{2},\mu,\sigma,\nu$ are.

Whether $\widehat C$ is *precise* is a different question, governed by
$$
\operatorname{Var}_{g}\bigl(w(X)\bigr)
 \;=\; \int_{c_{1}}^{c_{2}}\frac{f_{t_{\nu}}(x)^{2}}{g(x)}\,dx \;-\; C_{f}^{2} .
$$
When $c_{2}$ is finite, $f$ and $g$ share the same bounded interval, the
ratio $f/g$ is bounded on it, and the integral is finite — a well-behaved
estimator, whatever $\nu$ is. Send $c_{2}\to\infty$, though, and the tails
stop matching: $f_{t_{\nu}}(x)^{2}$ decays only polynomially
($\propto x^{-2(\nu+1)}$) while $g$'s Gaussian tail decays much faster, so the
integrand grows without bound and the integral diverges — for *every*
$\mu,\sigma$ and *every* finite $\nu$. The app below lets you flip between
the two regimes with a single checkbox, and, as before, $S_{\text{eff}}$ need
not warn you when you have landed in the bad one.

`is_t_interval()` turns the estimator above into code: draw $X$ from
`rtnorm_ars()`, form the log-weight $\log f_{t_\nu}(X) - \log g(X)$, and
report $\widehat C$, its standard error, and $S_{\text{eff}}$.

```{r}
#| label: is-t-interval
## ---------------------------------------------------------------------
##  is_t_interval(): importance-sampling estimate of P(c1 < X < c2) for a
##  standard t_df target, using X ~ N(mu, sigma^2) truncated to (c1, c2)
##  as the proposal.
## ---------------------------------------------------------------------
is_t_interval <- function(S, df, c1, c2, mu, sigma, seed = NULL) {
    if (!is.null(seed)) set.seed(seed)
    X <- rtnorm_ars(S, mu, sigma, c1, c2)

    ## log importance weight  log f_t(x; df) - log g(x),  with g the
    ## normalized truncated normal on (c1, c2)
    lZ   <- log(pnorm(c2, mu, sigma) - pnorm(c1, mu, sigma))
    logw <- dt(X, df, log = TRUE) - (dnorm(X, mu, sigma, log = TRUE) - lZ)

    m    <- max(logw)
    v    <- exp(logw - m)                       # rescaled weights
    Chat <- mean(v) * exp(m)                    # unbiased estimate of P(c1 < X < c2)
    se   <- sd(v) / sqrt(S) * exp(m)            # nominal standard error
    wn   <- v / sum(v)                          # normalized weights

    list(X = X, wn = wn, Chat = Chat, se = se, ess = 1 / sum(wn^2),
         truth = pt(c2, df) - pt(c1, df))
}
```

Four settings make the point concretely: a finite interval is well-behaved
at any $\nu$, while $c_{2}=\infty$ reproduces the divergence — badly at
$\nu=1$, but deceptively well at $\nu=30$, exactly as @tbl-is-t-examples
shows.

```{r}
#| label: tbl-is-t-examples
#| tbl-cap: "Four calls to `is_t_interval()` with S = 5000: a finite c2 is
#|   well-behaved at every df, c2 = Inf reproduces the divergent case,
#|   badly at df = 1 and only deceptively well at df = 30."
settings <- data.frame(df = c(1, 10, 1, 30), c1 = c(1, 1, 1, 1), c2 = c(3, 3, Inf, Inf))
res <- lapply(seq_len(nrow(settings)), function(i) with(settings[i, ],
    is_t_interval(S = 5000, df = df, c1 = c1, c2 = c2, mu = 0, sigma = 1, seed = 1)))

knitr::kable(data.frame(
    df       = settings$df,
    c1       = settings$c1,
    c2       = ifelse(is.finite(settings$c2), as.character(settings$c2), "Inf"),
    truth    = sapply(res, `[[`, "truth"),
    estimate = sapply(res, `[[`, "Chat"),
    se       = sapply(res, `[[`, "se"),
    ess_pct  = 100 * sapply(res, `[[`, "ess") / 5000
), digits = 4, col.names = c("df", "c1", "c2", "truth", "estimate", "s.e.", "ESS %"))
```

### An interactive view {#sec-t-is-app}

The app below puts $c_{1}$, $c_{2}$, $\mu$, $\sigma$ and the target's degrees
of freedom $\nu$ under slider control (a checkbox sends $c_{2}$ to $\infty$),
and displays, on a common $x$-axis: the two truncated densities on a log
scale with a rug of the draws, and the normalized importance weights
$\tilde w_{i}=w_{i}/\sum_{j}w_{j}$ at the sampled points. The legend reports
the true $P(c_{1}<X<c_{2})$, the estimate, and its nominal 95% interval,
colored green when the interval covers the truth and red when it does not.
Leave $c_{2}$ finite and the estimator behaves; check the box and watch it
misbehave exactly as the variance calculation predicts, while
$S_{\text{eff}}$ in the bottom panel need not warn you either way. This app
runs live only in the HTML edition of this book; readers of other formats can
follow the link below it.

::: {#fig-t-is-app .content-visible when-format="html"}

```{shinylive-r}
#| label: is-t-app
#| standalone: true
#| viewerHeight: 850

library(shiny)
library(ars)

## rtnorm_ars() and is_t_interval(): identical to the versions defined
## earlier in the chapter (a {shinylive-r} chunk cannot see objects
## defined elsewhere in the document, so they are pasted in verbatim).
rtnorm_ars <- function(n, mu = 0, sigma = 1, lo = 0, hi = Inf) {
    logf   <- function(x) dnorm(x, mu, sigma, log = TRUE)
    fprima <- function(x) -(x - mu) / sigma^2
    if (is.finite(hi)) {
        x0 <- lo + (hi - lo) * c(0.1, 0.5, 0.9)
    } else {
        right <- max(mu, lo) + 3 * sigma
        x0 <- sort(unique(c(lo + 0.1 * sigma, max(lo + 0.2 * sigma, mu + 1e-6), right)))
    }
    ars(n, f = logf, fprima = fprima, x = x0,
        lb = TRUE, xlb = lo, ub = is.finite(hi), xub = if (is.finite(hi)) hi else 0)
}

is_t_interval <- function(S, df, c1, c2, mu, sigma, seed = NULL) {
    if (!is.null(seed)) set.seed(seed)
    X <- rtnorm_ars(S, mu, sigma, c1, c2)

    lZ   <- log(pnorm(c2, mu, sigma) - pnorm(c1, mu, sigma))
    logw <- dt(X, df, log = TRUE) - (dnorm(X, mu, sigma, log = TRUE) - lZ)

    m    <- max(logw)
    v    <- exp(logw - m)
    Chat <- mean(v) * exp(m)
    se   <- sd(v) / sqrt(S) * exp(m)
    wn   <- v / sum(v)

    list(X = X, wn = wn, Chat = Chat, se = se, ess = 1 / sum(wn^2),
         truth = pt(c2, df) - pt(c1, df))
}

ui <- fluidPage(
    titlePanel("Importance sampling for Estimating a Student-t Probabilities with a truncated-normal proposal"),
    sidebarLayout(
        sidebarPanel(
            width = 3,
            sliderInput("df", HTML("target df &nu;"), min = 1, max = 30, value = 3, step = 1),
            sliderInput("c1", HTML("c<sub>1</sub>"), min = -5, max = 10, value = 1, step = 0.25),
            sliderInput("c2", HTML("c<sub>2</sub>"), min = -5, max = 15, value = 3, step = 0.25),
            checkboxInput("c2inf", HTML("c<sub>2</sub> = &infin;  (infinite-variance case)"), FALSE),
            sliderInput("mu", HTML("proposal &mu;"), min = -5, max = 20, value = 0, step = 0.25),
            sliderInput("sg", HTML("proposal &sigma;"), min = 0.25, max = 10, value = 1, step = 0.25),
            selectInput("S", "sample size  S",
                        choices = c(200, 1000, 5000, 20000), selected = 1000),
            actionButton("new", "New sample"),
            checkboxInput("logw", "log scale for weights", FALSE)
        ),
        mainPanel(width = 9, plotOutput("panels", height = "680px"))
    )
)

server <- function(input, output) {

    sim <- reactive({
        S  <- as.integer(input$S)
        c1 <- input$c1
        c2 <- if (input$c2inf) Inf else input$c2
        validate(need(c2 > c1, "c2 must be greater than c1"))
        mu <- input$mu; sg <- input$sg; df <- input$df

        r <- is_t_interval(S, df, c1, c2, mu, sg, seed = input$new + 1)
        c(r, list(S = S, c1 = c1, c2 = c2, mu = mu, sg = sg, df = df))
    })

    output$panels <- renderPlot({
        s  <- sim()
        c1 <- s$c1; c2 <- s$c2; X <- s$X
        lo <- c1
        hi <- if (is.finite(c2)) c2 else max(max(X) + 0.5, c1 + 5)
        xs <- seq(lo, hi, length.out = 800)

        ## both densities normalized on (c1, c2)
        ft <- dt(xs, s$df) / s$truth
        gt <- dnorm(xs, s$mu, s$sg) / (pnorm(c2, s$mu, s$sg) - pnorm(c1, s$mu, s$sg))
        yl <- range(c(ft, gt)[c(ft, gt) > 0])
        yl[1] <- max(yl[1], 1e-12)

        lwr <- s$Chat - 1.96 * s$se
        upr <- s$Chat + 1.96 * s$se
        covered <- lwr <= s$truth && s$truth <= upr
        ci_col  <- if (covered) "darkgreen" else "firebrick"

        layout(matrix(1:2, 2), heights = c(3, 2))

        ## ---- top panel: the two densities ----------------------------
        par(mar = c(0.4, 5, 3, 1))
        plot(xs, ft, type = "n", log = "y", ylim = yl, xlim = c(lo, hi),
             xaxt = "n", xlab = "", ylab = "density (log scale)",
             main = "target and proposal, both normalized on (c1, c2)")
        if (!is.finite(c2)) {
            rect(max(X), yl[1], hi, yl[2] * 10,
                 col = adjustcolor("grey60", 0.18), border = NA)
            text(max(X), yl[2], "  no draw ever reached here", adj = c(0, 1.4),
                 col = "grey35", cex = 0.95)
        }
        lines(xs, ft, lwd = 4, col = "firebrick")
        lines(xs, gt, lwd = 3, col = "steelblue")
        rug(X, col = adjustcolor("steelblue", 0.25))
        legend("bottomleft", inset = c(0.02, 0.05), bty = "n", cex = 1.05,
               lwd = c(4, 3, NA, NA, NA),
               col = c("firebrick", "steelblue", NA, NA, NA),
               text.col = c("black", "black", "black", "black", ci_col),
               legend = c(
                   sprintf("target: t(df=%g) on (c1, c2)", s$df),
                   sprintf("proposal: N(%.2f, %.2f^2) on (c1, c2)", s$mu, s$sg),
                   sprintf("true  P = %.5g", s$truth),
                   sprintf("est.   P = %.5g", s$Chat),
                   sprintf("95%% CI = (%.5g, %.5g)", lwr, upr)))

        ## ---- bottom panel: normalized weights ------------------------
        par(mar = c(4.5, 5, 0.4, 1))
        wn <- s$wn
        if (input$logw) {
            wl <- c(max(min(wn[wn > 0]), 1e-12), max(wn))
            plot(X, wn, type = "n", log = "y", ylim = wl, xlim = c(lo, hi),
                 xlab = "x", ylab = "normalized weight (log scale)")
            segments(X, wl[1], X, wn, col = adjustcolor("grey50", 0.6))
        } else {
            plot(X, wn, type = "n", ylim = c(0, max(wn) * 1.08),
                 xlim = c(lo, hi), xlab = "x", ylab = "normalized weight")
            segments(X, 0, X, wn, col = adjustcolor("grey50", 0.6))
        }
        abline(h = 1 / s$S, lty = 2, col = "darkgreen", lwd = 2)
        legend("topright", inset = c(0.02, 0.05), bty = "n", cex = 1.05,
               lty = c(2, NA, NA), lwd = c(2, NA, NA),
               col = c("darkgreen", NA, NA),
               text.col = "blue",
               legend = c("1 / S  (perfectly balanced weights)",
                          sprintf("ESS = %.1f   (%.2f%% of S = %d)",
                                  s$ess, 100 * s$ess / s$S, s$S),
                          sprintf("largest weight = %.1f%% of the total",
                                  100 * max(wn))))
    })
}

shinyApp(ui, server)
```

**Shinylive App for Importance Sampling of a Student-$t$ Interval.** Sliders
control $c_{1}$, $c_{2}$ (or $c_{2}=\infty$ via the checkbox), the target's
degrees of freedom $\nu$, and the proposal's $\mu,\sigma$; the top panel
compares the (normalized) target and proposal densities on a log scale, with
the CI in the legend colored by whether it covers the truth, and the bottom
panel shows the normalized importance weights at the sampled points.

:::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit12_importance_sampling.html#sec-t-is-app>.
:::


Several settings repay attention.

- **$c_{2}$ finite (the default).** Target and proposal share the same
  bounded support, so the weights have finite variance: raising $S$ improves
  the estimate at the usual $1/\sqrt S$ rate, and the 95% interval covers
  about as often as it claims to.
- **Check "$c_{2}=\infty$".** This reproduces the divergent case derived
  above. Raising $S$ no longer shrinks a variance that does not exist, and
  repeated *New sample* draws swing far more than the reported standard error
  admits. With large $\nu$ the run-away region is rarely reached, so the app
  can still look fine by chance; with small $\nu$ (Cauchy, $\nu=1$) it bites
  often — but $S_{\text{eff}}$ need not warn you either way.
- **Small $\sigma$, or $\mu$ far outside $(c_{1},c_{2})$.** Draws pile up
  wherever the proposal happens to put mass inside the interval, the weights
  are nearly equal, $S_{\text{eff}}$ stays high, and the estimate can still
  miss badly: uniform weights certify only that the proposal is internally
  consistent, not that it matches the target's shape there.
- **Large $\sigma$.** One draw's weight can dominate the sum,
  $S_{\text{eff}}$ collapses toward $1$ — visibly unreliable, which is an
  improvement on being invisibly unreliable.


## Example 3: Marginal Likelihood of Normal Data via Importance Sampling


The [quadrature](unit09_integral.qmd) and Laplace-approximation chapters
compute the marginal likelihood $P(y) = \int L(\theta)\pi(\theta)\,d\theta$ of
the same conjugate-free Gaussian model — unknown mean $\mu$ and
$w=\log\sigma^2$, independent normal priors — by a product grid and by a
mode-and-Hessian approximation, respectively. Importance sampling offers a
third route, and one that comes with a built-in accuracy diagnostic
($S_{\text{eff}}$) that neither of the other two methods provides directly.

Take $g(\mu,w)$ to be the **Gaussian approximation to the posterior**
obtained by a Laplace fit: locate the posterior mode $(\hat\mu,\hat w)$ by
optimization, take $g = N_2\big((\hat\mu,\hat w),\, H^{-1}\big)$ for $H$ the
Hessian of $-\log[L\pi]$ at the mode, and importance-sample from $g$ with
$f = L\pi$:
$$
\widehat{P(y)} = \frac{1}{S}\sum_{s=1}^S w_s, \qquad
w_s = \frac{L(\theta_s)\pi(\theta_s)}{g(\theta_s)}, \qquad \theta_s \sim g.
$$
Because $g$ is shaped like the posterior near its mode, this is a "good
story" proposal by construction — in sharp contrast to sampling directly from
the (typically much wider, uninformative) prior, which is exactly the
`log_marlik_mc` function used elsewhere in this book as a naive-Monte-Carlo
check: it is importance sampling with $g = \pi$, and its weights are just the
likelihood values $L(\theta_s)$.

**Implementation**

```{r}
#| label: marlik-functions
log_lik   <- function(x, mu, w) sum(dnorm(x, mu, exp(w / 2), log = TRUE))
log_prior <- function(mu, w, mu_0, sigma_mu, w_0, sigma_w) {
    dnorm(mu, mu_0, sigma_mu, log = TRUE) + dnorm(w, w_0, sigma_w, log = TRUE)
}
neg_log_post <- function(theta, x, mu_0, sigma_mu, w_0, sigma_w) {
    -log_lik(x, theta[1], theta[2]) -
        log_prior(theta[1], theta[2], mu_0, sigma_mu, w_0, sigma_w)
}

### log density of N(mu, A^{-1}) at every column of a 2 x S matrix of points at once,
### replacing a per-point apply() call with one vectorized quadratic form
log_dmvnorm_batch <- function(Theta, mu, A) {
    Tc <- Theta - mu
    quad <- colSums((A %*% Tc) * Tc)
    0.5 * (-nrow(Theta) * log(2 * pi) + sum(log(svd(A)$d)) - quad)
}

### log marginal likelihood by importance sampling from a Laplace-shaped Gaussian.
### Besides S_eff, attaches "se": the standard error of the *log* estimate,
### sd(v) / (sqrt(S) * mean(v)) for v = exp(lw - max(lw)) -- the delta-method
### transfer of the raw estimator's SE onto the log scale, exactly as used for
### the log estimate in the t-interval app above.
log_marlik_imps <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    fit <- nlm(neg_log_post, c(mean(x), log(stats::var(x))), hessian = TRUE,
               x = x, mu_0 = mu_0, sigma_mu = sigma_mu, w_0 = w_0, sigma_w = sigma_w)
    mu_hat <- fit$estimate; H <- fit$hessian
    L <- t(chol(solve(H)))                        # Sigma = H^{-1} = L L'
    Theta <- L %*% matrix(rnorm(2 * S), 2, S) + mu_hat

    log_g <- log_dmvnorm_batch(Theta, mu_hat, H)
    log_f <- -apply(Theta, 2, neg_log_post, x = x, mu_0 = mu_0, sigma_mu = sigma_mu,
                     w_0 = w_0, sigma_w = sigma_w)
    lw <- log_f - log_g
    m  <- max(lw); v <- exp(lw - m)
    structure(log_sum_exp(lw) - log(S),
              ess = log_ess(lw), se = sd(v) / (sqrt(S) * mean(v)))
}

### log marginal likelihood by naive Monte Carlo from the prior (g = pi): a
### second, independent importance-sampling estimator with an unfavourable g
log_marlik_prior <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    mus <- rnorm(S, mu_0, sigma_mu); ws <- rnorm(S, w_0, sigma_w)
    lw <- vapply(seq_len(S), function(i) log_lik(x, mus[i], ws[i]), numeric(1))
    m  <- max(lw); v <- exp(lw - m)
    structure(log_sum_exp(lw) - log(S),
              ess = log_ess(lw), se = sd(v) / (sqrt(S) * mean(v)))
}

### "truth": log marginal likelihood by a midpoint grid over (mu, w), centred
### and scaled by the Laplace fit. The likelihood term is vectorized over the
### whole mu-grid at once via the sufficient statistic sum((x - mu)^2), so --
### like log_dmvnorm_batch above -- no R-level loop over grid points, or over
### data points beyond the one sum(), is needed.
log_marlik_grid <- function(x, mu_0, sigma_mu, w_0, sigma_w, n_grid = 400, width = 8) {
    n   <- length(x)
    fit <- nlm(neg_log_post, c(mean(x), log(stats::var(x))), hessian = TRUE,
               x = x, mu_0 = mu_0, sigma_mu = sigma_mu, w_0 = w_0, sigma_w = sigma_w)
    se  <- sqrt(diag(solve(fit$hessian)))
    mu_grid <- fit$estimate[1] + seq(-width, width, length.out = n_grid) * se[1]
    w_grid  <- fit$estimate[2] + seq(-width, width, length.out = n_grid) * se[2]

    SS <- sum(x^2) - 2 * mu_grid * sum(x) + n * mu_grid^2   # sum((x - mu)^2)
    log_lik_grid   <- outer(SS, w_grid,
                             function(ss, w) -0.5 * n * log(2 * pi) - n * w / 2 - 0.5 * exp(-w) * ss)
    log_prior_grid <- outer(dnorm(mu_grid, mu_0, sigma_mu, log = TRUE),
                             dnorm(w_grid, w_0, sigma_w, log = TRUE), "+")

    h_mu <- diff(mu_grid[1:2]); h_w <- diff(w_grid[1:2])
    log_sum_exp(as.vector(log_lik_grid + log_prior_grid)) + log(h_mu) + log(h_w)
}
```


`log_marlik_grid()` supplies ground truth by quadrature — a midpoint grid
over $(\mu,w)$, centred and scaled by the same Laplace fit used for the
importance proposal, wide enough (8 posterior SDs each way, on a $400\times
400$ grid) that refining it further does not move the answer. Comparing both
proposals against it in a single run, with a 95% CI built from the standard
error of that run, makes the difference between them concrete rather than
just asserted.

**Results across sample sizes**

Both proposals target the same $f=L\pi$, and the prior is a legitimate
proposal for it: every posterior draw has positive prior density, so the
support condition holds. The two differ only in how well their shape matches
the target's, and the following table varies the amount of data to make that
difference visible.

```{r}
#| label: tbl-marlik-merged
#| tbl-cap: "Log marginal likelihood for a freshly simulated data set at each
#|   sample size n: truth by grid quadrature, against single runs of importance
#|   sampling (S = 2000) from a Laplace-shaped proposal and from the prior."
set.seed(4)
ns <- c(5, 20, 100, 400)
S  <- 2000

## one proposal, one data set: estimate, interval and S_eff as a % of S
summarize_is <- function(fn, x, S) {
    est <- fn(x, 0, 5, 0, 5, S)
    se  <- attr(est, "se")
    c(estimate = sprintf("%.2f", as.numeric(est)),
      ci       = sprintf("(%.2f, %.2f)",
                         as.numeric(est) - 1.96 * se,
                         as.numeric(est) + 1.96 * se),
      ess      = sprintf("%.1f", 100 * attr(est, "ess") / S))
}

merged_tbl <- do.call(rbind, lapply(ns, function(n) {
    x   <- rnorm(n)
    lap <- summarize_is(log_marlik_imps,  x, S)
    pri <- summarize_is(log_marlik_prior, x, S)
    data.frame(n     = n,
               truth = sprintf("%.2f", log_marlik_grid(x, 0, 5, 0, 5)),
               lap_estimate = lap[["estimate"]],
               lap_ci       = lap[["ci"]],
               lap_ess      = lap[["ess"]],
               pri_estimate = pri[["estimate"]],
               pri_ci       = pri[["ci"]],
               pri_ess      = pri[["ess"]],
               row.names    = NULL)
}))

## repeated names are intentional: the proposal is carried by the banner
names(merged_tbl) <- c("n", "truth",
                       "estimate", "95% CI", "ESS %",
                       "estimate", "95% CI", "ESS %")

tinytable::tt(merged_tbl) |>
    tinytable::group_tt(j = list("Laplace-shaped proposal" = 3:5,
                                 "prior sampling"          = 6:8))
```

The Laplace-shaped proposal re-centers and re-scales itself around the
posterior mode for every data set. Its interval stays tight and centred on the
truth at every $n$, and $S_{\text{eff}}$ stays close to the nominal $S$: a
second-order approximation to the posterior is evidently a good match to it.
Prior sampling degrades monotonically. The posterior concentrates at rate
$n^{-1/2}$ while the prior does not move, so the fraction of prior mass lying
where the likelihood is appreciable shrinks with $n$, the weights become ever
more concentrated on a handful of draws, and $S_{\text{eff}}$ falls to a small
fraction of one percent. Its interval may still cover the truth, but it is only
as trustworthy as those few draws happen to be from one run to the next.

This is the marginal-likelihood instance of the lesson drawn from the
missed-mode example: a fixed, data-independent proposal becomes an increasingly
poor match to a target that concentrates as data accumulate, and only a
proposal that adapts to the target avoids the resulting collapse in efficiency.
Note also the contrast in how the two failures announce themselves. Prior
sampling fails *noisily* — $S_{\text{eff}}$ reports the collapse directly, which
is exactly what the diagnostic is built to catch. The earlier tail examples
failed *silently*, with $S_{\text{eff}}$ near its nominal value while the
estimate was badly wrong, because there the weights were uniform among the
draws actually observed and the trouble lay entirely outside the region the
proposal ever visited.

### Seeing it: sorted weights {#sec-marlik-app}

A single number like $S_{\text{eff}}$ summarizes the weight distribution;
looking at the weights themselves is more direct. The app below draws $S$
importance samples from each proposal and plots their *normalized weights,
sorted from largest to smallest* against rank$/S$, in two separate panels,
because the two proposals' weights live on such different scales that a
shared axis would flatten the well-behaved one to a barely visible line. The
weight axis defaults to a log scale — a proposal shaped like the target gives
a nearly flat curve close to $1/S$ (every draw contributes about equally),
while a poorly matched proposal gives a curve that falls off a cliff, with a
handful of draws near rank $0$ carrying nearly all the weight; a checkbox
switches both panels to a linear scale instead, where that same cliff
collapses to a single spike against an otherwise invisible floor — a useful
reminder that the log scale is doing real work, not just spreading the
picture out. Each panel's legend reports the grid-quadrature truth and that
proposal's estimate and 95% CI, colored green when the interval covers the
truth and red when it does not. This app runs live only in the HTML edition
of this book; readers of other formats can follow the link below it.

::: {#fig-marlik-app .content-visible when-format="html"}

```{shinylive-r}
#| label: marlik-weights-app
#| standalone: true
#| viewerHeight: 820

library(shiny)

log_sum_exp <- function(lx) { m <- max(lx); m + log(sum(exp(lx - m))) }

log_lik   <- function(x, mu, w) sum(dnorm(x, mu, exp(w / 2), log = TRUE))
log_prior <- function(mu, w, mu_0, sigma_mu, w_0, sigma_w) {
    dnorm(mu, mu_0, sigma_mu, log = TRUE) + dnorm(w, w_0, sigma_w, log = TRUE)
}
neg_log_post <- function(theta, x, mu_0, sigma_mu, w_0, sigma_w) {
    -log_lik(x, theta[1], theta[2]) -
        log_prior(theta[1], theta[2], mu_0, sigma_mu, w_0, sigma_w)
}
log_dmvnorm_batch <- function(Theta, mu, A) {
    Tc <- Theta - mu
    quad <- colSums((A %*% Tc) * Tc)
    0.5 * (-nrow(Theta) * log(2 * pi) + sum(log(svd(A)$d)) - quad)
}

## same grid-quadrature "truth" as log_marlik_grid() above
marlik_truth <- function(x, mu_0, sigma_mu, w_0, sigma_w, n_grid = 400, width = 8) {
    n   <- length(x)
    fit <- nlm(neg_log_post, c(mean(x), log(stats::var(x))), hessian = TRUE,
               x = x, mu_0 = mu_0, sigma_mu = sigma_mu, w_0 = w_0, sigma_w = sigma_w)
    se  <- sqrt(diag(solve(fit$hessian)))
    mu_grid <- fit$estimate[1] + seq(-width, width, length.out = n_grid) * se[1]
    w_grid  <- fit$estimate[2] + seq(-width, width, length.out = n_grid) * se[2]
    SS <- sum(x^2) - 2 * mu_grid * sum(x) + n * mu_grid^2
    log_lik_grid   <- outer(SS, w_grid,
                             function(ss, w) -0.5 * n * log(2 * pi) - n * w / 2 - 0.5 * exp(-w) * ss)
    log_prior_grid <- outer(dnorm(mu_grid, mu_0, sigma_mu, log = TRUE),
                             dnorm(w_grid, w_0, sigma_w, log = TRUE), "+")
    h_mu <- diff(mu_grid[1:2]); h_w <- diff(w_grid[1:2])
    log_sum_exp(as.vector(log_lik_grid + log_prior_grid)) + log(h_mu) + log(h_w)
}

## Laplace-shaped proposal: draw, weight, and return the log estimate, its
## standard error (delta method), and the sorted normalized weights
sim_laplace <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    fit <- nlm(neg_log_post, c(mean(x), log(stats::var(x))), hessian = TRUE,
               x = x, mu_0 = mu_0, sigma_mu = sigma_mu, w_0 = w_0, sigma_w = sigma_w)
    mu_hat <- fit$estimate; H <- fit$hessian
    L <- t(chol(solve(H)))
    Theta <- L %*% matrix(rnorm(2 * S), 2, S) + mu_hat
    log_g <- log_dmvnorm_batch(Theta, mu_hat, H)
    log_f <- -apply(Theta, 2, neg_log_post, x = x, mu_0 = mu_0, sigma_mu = sigma_mu,
                     w_0 = w_0, sigma_w = sigma_w)
    lw <- log_f - log_g
    m  <- max(lw); v <- exp(lw - m); wn <- v / sum(v)
    list(est = log_sum_exp(lw) - log(S), se = sd(v) / (sqrt(S) * mean(v)),
         ess = 1 / sum(wn^2), wn = sort(wn, decreasing = TRUE))
}

## prior proposal (g = pi): same bookkeeping, unfavourable g
sim_prior <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    mus <- rnorm(S, mu_0, sigma_mu); ws <- rnorm(S, w_0, sigma_w)
    lw  <- vapply(seq_len(S), function(i) log_lik(x, mus[i], ws[i]), numeric(1))
    m   <- max(lw); v <- exp(lw - m); wn <- v / sum(v)
    list(est = log_sum_exp(lw) - log(S), se = sd(v) / (sqrt(S) * mean(v)),
         ess = 1 / sum(wn^2), wn = sort(wn, decreasing = TRUE))
}

ui <- fluidPage(
    titlePanel("Sorted importance weights: Laplace-shaped proposal vs. prior"),
    sidebarLayout(
        sidebarPanel(
            width = 3,
            sliderInput("n", "data size  n", min = 5, max = 200, value = 20, step = 5),
            sliderInput("sig_mu", HTML("prior sd on &mu;"), min = 1, max = 10, value = 5, step = 0.5),
            sliderInput("sig_w", HTML("prior sd on w"), min = 1, max = 10, value = 5, step = 0.5),
            selectInput("S", "importance sample size  S",
                        choices = c(500, 2000, 5000, 20000), selected = 2000),
            actionButton("new", "New sample"),
            checkboxInput("logw", "log scale for weights", TRUE)
        ),
        mainPanel(width = 9, plotOutput("panel", height = "720px"))
    )
)

server <- function(input, output) {

    sim <- reactive({
        n <- input$n; S <- as.integer(input$S)
        sig_mu <- input$sig_mu; sig_w <- input$sig_w
        set.seed(input$new + 1)
        x <- rnorm(n)

        list(S = S,
             truth = marlik_truth(x, 0, sig_mu, 0, sig_w),
             lap   = sim_laplace(x, 0, sig_mu, 0, sig_w, S),
             pri   = sim_prior(x, 0, sig_mu, 0, sig_w, S))
    })

    output$panel <- renderPlot({
        s <- sim(); S <- s$S
        rank_frac <- seq_len(S) / S
        uselog <- input$logw

        lwr_lap <- s$lap$est - 1.96 * s$lap$se; upr_lap <- s$lap$est + 1.96 * s$lap$se
        lwr_pri <- s$pri$est - 1.96 * s$pri$se; upr_pri <- s$pri$est + 1.96 * s$pri$se
        cov_lap <- if (lwr_lap <= s$truth && s$truth <= upr_lap) "darkgreen" else "firebrick"
        cov_pri <- if (lwr_pri <= s$truth && s$truth <= upr_pri) "darkgreen" else "firebrick"

        draw_panel <- function(wn, col, title, xlab, est, lwr, upr, ess_pct, cov_col) {
            if (uselog) {
                floor <- max(min(wn[wn > 0]), 1e-300)
                wn    <- pmax(wn, floor)
                ylim  <- range(wn)
                ylab  <- "normalized weight (log scale)"
            } else {
                ylim <- c(0, max(wn))
                ylab <- "normalized weight"
            }
            plot(rank_frac, wn, type = "n", log = if (uselog) "y" else "",
                 ylim = ylim, xlab = xlab, ylab = ylab, main = title)
            lines(rank_frac, wn, lwd = 3, col = col)
            abline(h = 1 / S, lty = 2, col = "darkgreen", lwd = 2)
            legend("topright", inset = c(0.02, 0.05), bty = "n", cex = 1.05,
                   text.col = c("black", cov_col),
                   legend = c(
                       sprintf("truth: log P(y) = %.3f", s$truth),
                       sprintf("est=%.3f  CI=(%.3f,%.3f)  ESS=%.2f%%", est, lwr, upr, ess_pct)))
        }

        layout(matrix(1:2, 2))

        par(mar = c(2.5, 5, 3, 1))
        draw_panel(s$lap$wn, "steelblue", "Laplace-shaped proposal", "",
                   s$lap$est, lwr_lap, upr_lap, 100 * s$lap$ess / S, cov_lap)

        par(mar = c(4.5, 5, 2.5, 1))
        draw_panel(s$pri$wn, "firebrick", "prior proposal",
                   "rank / S  (draws sorted by weight, largest first)",
                   s$pri$est, lwr_pri, upr_pri, 100 * s$pri$ess / S, cov_pri)
    })
}

shinyApp(ui, server)
```

**Shinylive App for Sorted Importance Weights.** Sliders control the
simulated data size $n$ and the prior spreads on $\mu$ and $w$; the top panel
plots the sorted, normalized importance weights from the Laplace-shaped
proposal and the bottom panel the same for the prior — each on its own axis,
since one proposal's weights routinely span far more orders of magnitude
than the other's — with a checkbox toggling both between log and linear
scale, and the dashed line at $1/S$ marking perfectly balanced weights in
both.

:::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit12_importance_sampling.html#sec-marlik-app>.
:::

Raise $n$ with the sliders at their defaults and watch the prior's curve
rotate from a gentle slope into a cliff at the very first few ranks, while the
Laplace curve barely moves — the same collapse as
@tbl-marlik-ess-vs-n, now visible in the shape of the curve rather than
only in a single summary number.

## Summary

- Importance sampling reweights draws from a proposal $g$ by
  $w(\theta) = f(\theta)/g(\theta)$ instead of accepting or rejecting them,
  using $\widehat{C_f/C_g} = \frac1S\sum_s w_s$ and the self-normalized
  $\hat A = \sum_s a(\theta_s)\tilde w_s$; only the support condition
  $\{g>0\}\supseteq\{f>0\}$ is required, not an envelope $g \ge f$.
- A good proposal shaped like the target on its support gives near-constant
  weights and $S_{\text{eff}} \approx S$ (the truncated-normal and
  Laplace-proposal examples), and a single weighted sample can answer several
  questions about the same target — a probability and an expectation alike.
- A proposal that misses a mode or part of the support fails **silently**: the
  observed weights can look perfectly healthy while both $C_f/C_g$ and
  $E_f[a]$ are badly wrong, and no amount of extra sampling fixes it.
  $S_{\text{eff}}$ diagnoses only the noisy failure (wildly varying observed
  weights), never the silent one.
- For the marginal likelihood, a proposal built from a Laplace approximation
  at the posterior mode remains efficient as the data accumulate and the
  posterior sharpens, while a fixed, data-independent proposal such as the
  prior does not — reinforcing why the mode-finding methods from earlier
  chapters remain useful even once we turn to sampling-based methods.
