---
title: "The EM Algorithm"
format:
  html:
    toc: true
    tbl-colwidths: true
filters:
  - shinylive
editor_options:
  chunk_output_type: console
code-fold: true
---

```{r}
#| label: setup
#| include: false
set.seed(812)

### censored Poisson example used for the illustrations in Section 1 and
### developed fully in Section 3: m exact observations, ncen observations
### known only to satisfy y < 2
yobs_d <- c(3, 2, 4, 1, 3, 2, 5, 3)
m_d    <- length(yobs_d)
ncen_d <- 12
n_d    <- m_d + ncen_d

lobs  <- function(l) sum(dpois(yobs_d, l, log = TRUE)) + ncen_d * log(ppois(1, l))
lcomp <- function(l, z) -n_d * l + log(l) * (sum(yobs_d) + sum(z)) - sum(lfactorial(yobs_d))
Qfun  <- function(l, lt) lcomp(l, rep(lt / (1 + lt), ncen_d))
lam   <- seq(0.4, 3.2, length.out = 300)
```

## Overview of EM

### Setup and motivation

Let $Y$ denote the observed data and $\theta$ the parameter, with likelihood
$L(\theta) = P(Y \mid \theta)$. In many models this quantity is itself an
integral (or sum) over latent or missing data $Z$:

$$
P(Y \mid \theta) = \int P(Y, Z \mid \theta) \, dZ .
$$

The integral is what makes direct maximization awkward: it destroys the product
structure that would otherwise make the log-likelihood a sum of simple terms.
The EM algorithm avoids it by working with the complete-data likelihood, which
retains that structure, and correcting for the fact that $Z$ is unobserved.

| | Likelihood | Log-likelihood |
|---|---|---|
| Observed | $L_{\text{obs}}(\theta) = P(Y \mid \theta)$ | $\ell_{\text{obs}}(\theta) = \log P(Y \mid \theta)$ |
| Complete | $L_{\text{comp}}(\theta) = P(Y, Z \mid \theta)$ | $\ell_{\text{comp}}(\theta; Y, Z) = \log P(Y, Z \mid \theta)$ |

: Observed and complete likelihoods {#tbl-terminology}

### The algorithm

Start from $\hat\theta^{(0)}$. For $t = 0, 1, 2, \ldots$:

**E-step.** Replace $\ell_{\text{obs}}(\theta)$ by the expected complete
log-likelihood,

$$
Q\!\left(\theta \mid \hat\theta^{(t)}\right)
= E_{Z \mid Y, \hat\theta^{(t)}}\!\left[\ell_{\text{comp}}(\theta; Y, Z)\right].
$$

**M-step.** Maximize it,

$$
\hat\theta^{(t+1)} = \arg\max_{\theta} \; Q\!\left(\theta \mid \hat\theta^{(t)}\right).
$$

Iterate until $\hat\theta^{(t)}$ converges.

::: {.callout-warning}
### A common misreading
$Q$ is an *average of* $\ell_{\text{comp}}(\theta; Y, Z)$ over
$P(Z \mid Y, \hat\theta^{(t)})$. It is **not**
$\ell_{\text{comp}}\!\left(\theta; Y, E(Z \mid Y, \hat\theta^{(t)})\right)$ —
that is, EM is not "fill in the missing values with their expectations and
maximize". The two coincide only when $\ell_{\text{comp}}$ happens to be linear
in $Z$, which is true in both examples of this chapter and false in general.
:::

### $Q$ as an average of complete log-likelihoods

Fix $\hat\theta^{(t)}$. Each draw $Z^{(s)} \sim P(Z \mid Y, \hat\theta^{(t)})$
gives one curve $\ell_{\text{comp}}(\theta; Y, Z^{(s)})$, and
$Q(\theta \mid \hat\theta^{(t)})$ is their expectation. The figure uses the
censored Poisson model of @sec-poisson with $\lambda^{(t)} = 1.5$.

```{r}
#| label: fig-q-average
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Ten complete-data log-likelihood curves, their average, and the exact $Q$ function."
lt <- 1.5; S <- 10
Z <- replicate(S, rbinom(ncen_d, 1, lt / (1 + lt)))
curves <- sapply(1:S, function(s) sapply(lam, lcomp, z = Z[, s]))
Qexact <- sapply(lam, Qfun, lt = lt)
par(mar = c(4, 4, 1, 1))
matplot(lam, curves, type = "l", lty = 1, lwd = 1, col = "gray60",
        xlab = expression(theta), ylab = "log-likelihood",
        ylim = range(curves) + c(0, 2))
lines(lam, rowMeans(curves), lwd = 3, col = "firebrick")
lines(lam, Qexact, lwd = 3, col = "black", lty = 2)
abline(v = lt, lty = 3)
legend("bottomright", bty = "n", lwd = c(1, 3, 3), lty = c(1, 1, 2),
       col = c("gray60", "firebrick", "black"),
       legend = c(expression(l[comp](theta*";"*Y*","*Z^(s)) ~ "for 10 draws"),
                  "average of the 10 curves",
                  expression(Q(theta*"|"*hat(theta)^(t)) ~ "(exact expectation)")))
```

### $Q$ functions as tangent lower bounds

The geometric picture that explains why EM works: each
$Q(\cdot \mid \hat\theta^{(t)})$, shifted by a constant, lies **below**
$\ell_{\text{obs}}$ and **touches** it at $\hat\theta^{(t)}$. Maximizing the
bound therefore moves uphill on $\ell_{\text{obs}}$ itself.

```{r}
#| label: fig-tangent-bound
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Three EM iterations viewed as successive tangent minorants of the observed log-likelihood."
lo <- sapply(lam, lobs); top <- max(lo)
par(mar = c(4, 4, 1, 1))
plot(lam, lo, type = "l", lwd = 3, xlab = expression(theta),
     ylab = "log-likelihood", ylim = c(top - 16, top + 2.5))
lt <- 0.5; cols <- c("firebrick", "darkorange", "forestgreen")
for (k in 1:3) {
  shift <- lobs(lt) - Qfun(lt, lt)            # make Q touch l_obs at theta^(t)
  lines(lam, sapply(lam, Qfun, lt = lt) + shift, col = cols[k], lwd = 2, lty = 2)
  points(lt, lobs(lt), pch = 19, col = cols[k])
  lt_new <- (sum(yobs_d) + ncen_d * lt / (1 + lt)) / n_d     # M-step
  segments(lt_new, top - 16, lt_new, lobs(lt_new), col = cols[k], lty = 3)
  text(lt, lobs(lt), bquote(hat(theta)^(.(k - 1))), pos = 3, col = cols[k], xpd = NA)
  lt <- lt_new
}
legend("bottomright", bty = "n", lwd = c(3, 2), lty = c(1, 2),
       legend = c(expression(l[obs](theta)),
                  expression(Q(theta*"|"*hat(theta)^(t)) + "const")))
```

### The ascending property

Write $h(Y, Z \mid \theta) = P(Y, Z \mid \theta)$,
$g(Y \mid \theta) = \int h(Y, Z \mid \theta) \, dZ$, and
$K(Z \mid Y, \theta) = h / g$.

::: {.callout-note}
### Theorem
$\ell_{\text{obs}}\!\left(\hat\theta^{(t+1)}\right) \ \ge\ \ell_{\text{obs}}\!\left(\hat\theta^{(t)}\right)$
at every iteration.
:::

**Key identity.** Since
$\log g(Y \mid \theta) = \log h(Y, Z \mid \theta) - \log K(Z \mid Y, \theta)$
holds for every $Z$, taking $E_{Z \mid Y, \theta_0}$ of both sides gives, for any
$\theta_0$,

$$
\ell_{\text{obs}}(\theta)
= \underbrace{E_{Z \mid Y, \theta_0}\!\left[\log h(Y, Z \mid \theta)\right]}_{Q(\theta \mid \theta_0)}
- \underbrace{E_{Z \mid Y, \theta_0}\!\left[\log K(Z \mid Y, \theta)\right]}_{H(\theta \mid \theta_0)} .
$$

The left side does not depend on $Z$, so the expectation leaves it unchanged;
this is the whole content of the step.

**Proof.** Set $\theta_0 = \hat\theta^{(t)}$ and subtract the identity evaluated
at $\theta = \hat\theta^{(t)}$ from the one evaluated at
$\theta = \hat\theta^{(t+1)}$:

$$
\begin{aligned}
\ell_{\text{obs}}\!\left(\hat\theta^{(t+1)}\right) - \ell_{\text{obs}}\!\left(\hat\theta^{(t)}\right)
&= \underbrace{\left[Q\!\left(\hat\theta^{(t+1)} \mid \hat\theta^{(t)}\right) - Q\!\left(\hat\theta^{(t)} \mid \hat\theta^{(t)}\right)\right]}_{\ge\, 0 \text{ by the M-step}} \\
&\quad - \underbrace{\left[H\!\left(\hat\theta^{(t+1)} \mid \hat\theta^{(t)}\right) - H\!\left(\hat\theta^{(t)} \mid \hat\theta^{(t)}\right)\right]}_{\le\, 0 \text{ by Jensen, below}} .
\end{aligned}
$$

**The $H$ term.** We need
$H(\theta \mid \hat\theta^{(t)}) \le H(\hat\theta^{(t)} \mid \hat\theta^{(t)})$
for every $\theta$, that is

$$
E_{Z \mid Y, \hat\theta^{(t)}}\!\left[\log \frac{K(Z \mid Y, \theta)}{K(Z \mid Y, \hat\theta^{(t)})}\right] \le 0 .
$$

Because $\log$ is concave, Jensen's inequality gives $E[\log W] \le \log E[W]$,
so

$$
\begin{aligned}
E_{Z \mid Y, \hat\theta^{(t)}}\!\left[\log \frac{K(Z \mid Y, \theta)}{K(Z \mid Y, \hat\theta^{(t)})}\right]
&\le \log \int \frac{K(Z \mid Y, \theta)}{K(Z \mid Y, \hat\theta^{(t)})} \, K(Z \mid Y, \hat\theta^{(t)}) \, dZ \\
&= \log \int K(Z \mid Y, \theta) \, dZ = \log 1 = 0 . \qquad \blacksquare
\end{aligned}
$$

The quantity on the left is
$-\mathrm{KL}\!\left(K(\cdot \mid Y, \hat\theta^{(t)}) \,\|\, K(\cdot \mid Y, \theta)\right)$.
Note carefully what the theorem does and does not say: the observed likelihood
never decreases, but that guarantees convergence only to a stationary point,
which may be a local maximum or even a saddle point.

### EM as alternating optimization

There is a second formulation that makes the algorithm's structure clearer and
generalizes better. Let $\tilde P(\cdot)$ be any distribution on the hidden data
$Z$, and define

$$
F\!\left(\tilde P, \theta\right)
= \int \tilde P(z) \log P(Y, z \mid \theta) \, dz
- \int \tilde P(z) \log \tilde P(z) \, dz .
$$

Using $P(Y, z \mid \theta) = P(Y \mid \theta) \, P(z \mid Y, \theta)$,

$$
F\!\left(\tilde P, \theta\right)
= \ell_{\text{obs}}(\theta) - \mathrm{KL}\!\left(\tilde P \,\|\, P(\cdot \mid Y, \theta)\right),
\qquad
\mathrm{KL}(P \| q) = \int P(z) \log \frac{P(z)}{q(z)} \, dz \; \ge\; 0 .
$$

Both steps of EM are now coordinate ascent on the single function $F$:

- **E-step** maximizes $F$ over $\tilde P$ with $\theta = \hat\theta^{(t)}$
  fixed. Since $\mathrm{KL} \ge 0$ with equality if and only if
  $\tilde P = P(\cdot \mid Y, \hat\theta^{(t)})$, the E-step sets $\tilde P$ to
  the conditional distribution of $Z$.
- **M-step** maximizes $F$ over $\theta$ with $\tilde P$ fixed. The entropy term
  is free of $\theta$, so this is $\arg\max_\theta Q(\theta \mid \hat\theta^{(t)})$.

The ascending property is then immediate: alternating maximization of a single
objective cannot decrease it. This view also explains the variational methods
used when the E-step is intractable — restrict $\tilde P$ to a manageable family
and accept a bound rather than an equality.

For completeness, $\mathrm{KL} \ge 0$ follows from $\log x \le x - 1$:

$$
\mathrm{KL}(P \| q) = -\int P(z) \log \frac{q(z)}{P(z)} \, dz
\;\ge\; -\int P(z) \left(\frac{q(z)}{P(z)} - 1\right) dz
= -\int \left[q(z) - P(z)\right] dz = 0 .
$$

## Example 1: Two-Component Normal Mixture {#sec-mixture}

### Model and likelihoods

$$
Z_i \sim \text{Bern}(p), \qquad
Y_i \mid Z_i = 1 \sim N(\mu_1, 1), \qquad
Y_i \mid Z_i = 0 \sim N(\mu_0, 1) .
$$

Observed: $y_1, \ldots, y_n$. Hidden: $z_1, \ldots, z_n$. Parameter
$\theta = (\mu_0, \mu_1, p)$.

**Complete likelihood**

$$
P(y, z \mid \theta) = \prod_{i=1}^{n} P(y_i \mid z_i, \theta) \, P(z_i \mid \theta)
= \prod_{i=1}^{n} \varphi(y_i \mid \mu_{z_i}, 1) \, p^{z_i} (1-p)^{1-z_i} .
$$

**Observed likelihood**, summing out each $z_i$,

$$
P(y \mid \theta) = \prod_{i=1}^{n}\left[\varphi(y_i \mid \mu_1, 1)\, p + \varphi(y_i \mid \mu_0, 1)(1 - p)\right] .
$$

The sum inside the product is what makes direct maximization awkward.

```{r}
#| label: fig-mixture-components
#| fig-width: 9
#| fig-height: 4.6
#| fig-cap: "The observed density is a sum of two weighted normal components."
par(mar = c(4, 4, 1, 1)); mu0 <- 0; mu1 <- 4; p <- 0.3
curve((1 - p) * dnorm(x, mu0), -3.5, 7.5, lwd = 2, col = "steelblue",
      ylab = "density", xlab = "y", ylim = c(0, 0.3))
curve(p * dnorm(x, mu1), add = TRUE, lwd = 2, col = "firebrick")
curve((1 - p) * dnorm(x, mu0) + p * dnorm(x, mu1), add = TRUE, lwd = 3)
legend("topright", bty = "n", lwd = c(2, 2, 3),
       col = c("steelblue", "firebrick", "black"),
       legend = c(expression((1-p)*phi(y*"|"*mu[0])),
                  expression(p*phi(y*"|"*mu[1])),
                  expression(f(y*"|"*theta) == "sum")))
```

The marginal density may be bimodal or unimodal depending on $p$ and
$|\mu_1 - \mu_0|$ — a point worth keeping in mind, since the unimodal case is
where EM struggles most.

```{r}
#| label: fig-mixture-shapes
#| fig-width: 10
#| fig-height: 4.4
#| fig-cap: "Separation and mixing proportion determine whether the mixture is visibly bimodal."
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
mix <- function(x, mu0, mu1, p) (1 - p) * dnorm(x, mu0) + p * dnorm(x, mu1)
for (s in list(c(0, 4, 0.5), c(0, 4, 0.15), c(0, 1.5, 0.5))) {
  curve(mix(x, s[1], s[2], s[3]), -4, 7, lwd = 3, xlab = "y", ylab = "density",
        main = bquote(mu[0] == .(s[1]) ~ "," ~ mu[1] == .(s[2]) ~ "," ~ p == .(s[3])))
  y <- ifelse(rbinom(200, 1, s[3]) == 1, rnorm(200, s[2]), rnorm(200, s[1]))
  rug(y, col = "gray40")
}
```

### Complete log-likelihood

With $\log \varphi(y \mid \mu, 1) = -\tfrac{1}{2}(y - \mu)^2 - \log\sqrt{2\pi}$,

$$
\begin{aligned}
\ell_{\text{comp}}(\theta; y, z)
&= \sum_{i=1}^{n}\left[z_i \log p + (1 - z_i)\log(1-p) - \tfrac{1}{2}(y_i - \mu_{z_i})^2\right] + C \\
&= \left(\sum_{i=1}^{n} z_i\right) \log \frac{p}{1-p} + n \log(1-p) \\
&\quad - \frac{1}{2}\sum_{i=1}^{n}\left[z_i (y_i - \mu_1)^2 + (1 - z_i)(y_i - \mu_0)^2\right] + C .
\end{aligned}
$$

Two features drive everything that follows. The expression is **linear in each
$z_i$**, so the E-step needs only $E(Z_i \mid y_i, \hat\theta^{(t)})$ rather than
the full conditional distribution. And given $z$, the parameters **separate**:
$p$ is a Bernoulli proportion while $\mu_1$ and $\mu_0$ are group means.

### E-step: soft classification

Given $\hat\theta^{(t)} = (\hat\mu_0^{(t)}, \hat\mu_1^{(t)}, \hat p^{(t)})$,
Bayes' rule gives

$$
\hat p_i^{(t)} \equiv P\!\left(Z_i = 1 \mid y_i, \hat\theta^{(t)}\right)
= \frac{\varphi\!\left(y_i \mid \hat\mu_1^{(t)}, 1\right) \hat p^{(t)}}
       {\varphi\!\left(y_i \mid \hat\mu_0^{(t)}, 1\right)\!\left(1 - \hat p^{(t)}\right)
        + \varphi\!\left(y_i \mid \hat\mu_1^{(t)}, 1\right) \hat p^{(t)}} ,
$$

and therefore

$$
\begin{aligned}
Q\!\left(\theta \mid \hat\theta^{(t)}\right)
&= \left(\sum_i \hat p_i^{(t)}\right) \log \frac{p}{1-p} + n \log(1-p) \\
&\quad - \frac{1}{2}\sum_i \left[\hat p_i^{(t)} (y_i - \mu_1)^2 + \left(1 - \hat p_i^{(t)}\right)(y_i - \mu_0)^2\right] .
\end{aligned}
$$

Computationally, $\hat p_i^{(t)}$ must be formed on the log scale. Writing
$a_i = \log \varphi(y_i \mid \hat\mu_1^{(t)}, 1) + \log \hat p^{(t)}$ and
$b_i = \log \varphi(y_i \mid \hat\mu_0^{(t)}, 1) + \log(1 - \hat p^{(t)})$, we
have $\hat p_i^{(t)} = \exp\{a_i - \operatorname{lse}(a_i, b_i)\}$ where
$\operatorname{lse}$ is the log-sum-exp function of the chapter on computer
arithmetic. Forming the ratio directly underflows as soon as an observation sits
far from both means.

```{r}
#| label: fig-estep
#| fig-width: 10
#| fig-height: 4.6
#| fig-cap: "Left: the responsibility at each $y_i$ is the red share of the total height. Right: responsibility as a function of $y$."
par(mfrow = c(1, 2), mar = c(4, 4, 1, 1)); mu0 <- 0; mu1 <- 4; p <- 0.3
f0 <- function(x) (1 - p) * dnorm(x, mu0)
f1 <- function(x) p * dnorm(x, mu1)
curve(f0, -3.5, 7.5, lwd = 2, col = "steelblue", ylab = "weighted density",
      xlab = "y", ylim = c(0, 0.3))
curve(f1, add = TRUE, lwd = 2, col = "firebrick")
yi <- c(-0.5, 1.8, 3.2)
for (y in yi) {
  segments(y, 0, y, f0(y), col = "steelblue", lwd = 4)
  segments(y, f0(y), y, f0(y) + f1(y), col = "firebrick", lwd = 4)
  text(y, f0(y) + f1(y), sprintf("%.2f", f1(y) / (f0(y) + f1(y))), pos = 3, cex = 0.9)
}
legend("topright", bty = "n", lwd = 2, col = c("steelblue", "firebrick"),
       legend = c(expression((1-hat(p))*phi(y*"|"*hat(mu)[0])),
                  expression(hat(p)*phi(y*"|"*hat(mu)[1]))))
curve(f1(x) / (f0(x) + f1(x)), -3.5, 7.5, lwd = 3, xlab = "y",
      ylab = expression(P(Z[i] == 1*"|"*y[i]*","*hat(theta)^(t))))
abline(h = c(0, 1), lty = 3); points(yi, f1(yi) / (f0(yi) + f1(yi)), pch = 19)
```

### M-step: weighted proportions and means

$Q$ separates into one term in $p$ and two in $\mu_1$, $\mu_0$, each maximized
in closed form.

| Update | Stationary equation $\Longrightarrow$ solution |
|---|---|
| $p$ | $\dfrac{\partial Q}{\partial p} = \dfrac{\sum_i \hat p_i^{(t)}}{p(1-p)} - \dfrac{n}{1-p} = 0 \;\Longrightarrow\; \hat p^{(t+1)} = \dfrac{1}{n}\sum_{i=1}^{n} \hat p_i^{(t)}$ |
| $\mu_1$ | $\dfrac{\partial Q}{\partial \mu_1} = \sum_i \hat p_i^{(t)}(y_i - \mu_1) = 0 \;\Longrightarrow\; \hat\mu_1^{(t+1)} = \dfrac{\sum_i \hat p_i^{(t)} y_i}{\sum_i \hat p_i^{(t)}}$ |
| $\mu_0$ | weights $1 - \hat p_i^{(t)}$: $\;\hat\mu_0^{(t+1)} = \dfrac{\sum_i \left(1 - \hat p_i^{(t)}\right) y_i}{\sum_i \left(1 - \hat p_i^{(t)}\right)}$ |

: M-step updates for the two-component mixture {#tbl-mstep-mixture}

In words: the M-step is the complete-data MLE with each observation split
between the two groups in proportion to its soft classification. Return to the
E-step with $\hat\theta^{(t+1)}$ and recompute the responsibilities.

### Implementation

```{r}
#| label: mix-functions

log_sum_exp <- function(log_x) {
    max_log_x <- max(log_x)
    max_log_x + log(sum(exp(log_x - max_log_x)))
}

### observed log-likelihood, theta = (p, mu1, mu0)
log_like_obs <- function(theta, Y) {
    log_joint <- cbind(log(theta[1])     + dnorm(Y, theta[2], 1, log = TRUE),
                       log(1 - theta[1]) + dnorm(Y, theta[3], 1, log = TRUE))
    sum(apply(log_joint, 1, log_sum_exp))
}

### E-step: responsibilities P(Z_i = 1 | y_i, theta)
e_step <- function(theta, Y) {
    l1 <- dnorm(Y, theta[2], 1, log = TRUE) + log(theta[1])
    l0 <- dnorm(Y, theta[3], 1, log = TRUE) + log(1 - theta[1])
    exp(l1 - apply(cbind(l1, l0), 1, log_sum_exp))
}

### M-step: closed-form weighted updates
m_step <- function(w, Y) {
    c(mean(w),
      sum(Y * w) / sum(w),
      sum(Y * (1 - w)) / sum(1 - w))
}

em_mixnorm <- function(theta0, Y, no_iters) {
    out <- matrix(NA_real_, no_iters + 1, 4,
                  dimnames = list(NULL, c("p", "mu1", "mu0", "log_lik")))
    theta <- theta0
    out[1, ] <- c(theta, log_like_obs(theta, Y))
    for (i in seq_len(no_iters)) {
        theta <- m_step(e_step(theta, Y), Y)
        out[i + 1, ] <- c(theta, log_like_obs(theta, Y))
    }
    out
}
```

```{r}
#| label: mix-data
#| fig-width: 6
#| fig-height: 4
#| fig-cap: "Simulated mixture data, coloured by the (normally unobserved) component label."

gen_mixnorm <- function(theta, n) {
    Z <- rbinom(n, 1, theta[1])
    Y <- rnorm(n, mean = ifelse(Z == 1, theta[2], theta[3]), sd = 1)
    plot(Y, Z, col = cm.colors(2)[Z + 1], pch = 19,
         ylab = "component label Z")
    Y
}

set.seed(812)
data <- gen_mixnorm(c(0.3, 0, 3), 200)
```

A well-chosen start converges quickly:

```{r}
#| label: mix-run-good
round(em_mixnorm(c(0.5, 0, 3), data, 10), 4)
```

The monotone increase of the last column is the ascending property in action;
it is the single most useful diagnostic when debugging an EM implementation,
since any decrease indicates an error in the E- or M-step.

### Interactive demonstration {#sec-mixture-app}

The figure below replaces the frame-by-frame animation of earlier versions of
this chapter. Choose a data set and a starting value — either from the presets
in the left panel or by typing your own — then step through the fit with
**Prev** and **Next**, or press **Play** to run it automatically. **Regenerate
data** draws a fresh sample from the same model, which is the quickest way to
see how much of the algorithm's behaviour is a property of the model and how
much is an accident of one data set. Changing the data, the sample size or any
starting value resets the display to iteration 0.

Upper panel: the two weighted component densities, with each observation
coloured by its current responsibility $\hat p_i^{(t)}$ and arrows showing where
the M-step moves the two means. Lower panel: the observed log-likelihood, which
must never decrease.

```{shinylive-r}
#| label: em-mixture-app
#| standalone: true
#| viewerHeight: 760

library(shiny)

### ---------- model functions -------------------------------------------------
log_sum_exp <- function(lx) { mx <- max(lx); mx + log(sum(exp(lx - mx))) }

log_like_obs <- function(theta, Y) {
    A <- cbind(log(theta[1])     + dnorm(Y, theta[2], 1, log = TRUE),
               log(1 - theta[1]) + dnorm(Y, theta[3], 1, log = TRUE))
    sum(apply(A, 1, log_sum_exp))
}

e_step <- function(theta, Y) {
    l1 <- dnorm(Y, theta[2], 1, log = TRUE) + log(theta[1])
    l0 <- dnorm(Y, theta[3], 1, log = TRUE) + log(1 - theta[1])
    exp(l1 - apply(cbind(l1, l0), 1, log_sum_exp))
}

m_step <- function(w, Y) {
    c(mean(w), sum(Y * w) / sum(w), sum(Y * (1 - w)) / sum(1 - w))
}

em_path <- function(theta0, Y, no_iters) {
    out <- matrix(NA_real_, no_iters + 1, 4,
                  dimnames = list(NULL, c("p", "mu1", "mu0", "log_lik")))
    th <- theta0
    out[1, ] <- c(th, log_like_obs(th, Y))
    for (i in seq_len(no_iters)) {
        th <- m_step(e_step(th, Y), Y)
        if (any(!is.finite(th))) break
        out[i + 1, ] <- c(th, log_like_obs(th, Y))
    }
    out
}

### ---------- preset starting values -----------------------------------------
presets <- list(
    "near the truth: (0.5, 0, 3)"                 = c(0.5,  0.0,  3.0),
    "far outside the data: (0.5, -10, 5)"         = c(0.5, -10.0, 5.0),
    "both means too high: (0.5, 5, 10)"           = c(0.5,  5.0, 10.0),
    "means nearly equal: (0.5, 0.1, 0.2)"         = c(0.5,  0.1,  0.2),
    "label-switched: (0.5, 3, 0)"                 = c(0.5,  3.0,  0.0),
    "random labels"                               = NA,
    "manual entry"                                = NULL
)

datasets <- list(
    "well separated: p = 0.3, mu = (3, 0)"  = c(0.3, 3, 0),
    "widely separated: p = 0.3, mu = (5, 0)" = c(0.3, 5, 0),
    "overlapping: p = 0.3, mu = (1.5, 0)"    = c(0.3, 1.5, 0),
    "balanced and overlapping: p = 0.5, mu = (1.5, 0)" = c(0.5, 1.5, 0)
)

### ---------- UI ---------------------------------------------------------------
ui <- fluidPage(
    sidebarLayout(
        sidebarPanel(
            width = 4,
            h5("Data"),
            selectInput("dataset", NULL, choices = names(datasets)),
            sliderInput("n", "sample size", 50, 400, 200, step = 50),
            actionButton("newdata", "Regenerate data", width = "100%"),
            tags$hr(),
            h5("Starting value"),
            selectInput("preset", NULL, choices = names(presets),
                        selected = "far outside the data: (0.5, -10, 5)"),
            fluidRow(
                column(4, numericInput("p0",   "p",       0.5, min = 0.01,
                                       max = 0.99, step = 0.05)),
                column(4, numericInput("mu10", "mu1",   -10,   step = 0.5)),
                column(4, numericInput("mu00", "mu0",     5,   step = 0.5))
            ),
            helpText("Selecting a preset fills the three boxes. The boxes are ",
                     "always live: edit them to use your own values."),
            tags$hr(),
            sliderInput("maxit", "iterations to run", 5, 60, 30, step = 5),
            h5(textOutput("stepLabel", inline = TRUE)),
            fluidRow(
                column(4, actionButton("prv",  "\u25c0 Prev", width = "100%")),
                column(4, actionButton("nxt",  "Next \u25b6", width = "100%")),
                column(4, actionButton("play", "Play",         width = "100%"))
            ),
            tags$br(),
            tableOutput("tab")
        ),
        mainPanel(
            width = 8,
            plotOutput("fit",   height = "400px"),
            plotOutput("trace", height = "260px")
        )
    )
)

### ---------- server -----------------------------------------------------------
server <- function(input, output, session) {

    seed <- reactiveVal(812)
    observeEvent(input$newdata, seed(seed() + 1))

    Y <- reactive({
        s <- datasets[[input$dataset]]
        set.seed(seed())
        Z <- rbinom(input$n, 1, s[1])
        rnorm(input$n, mean = ifelse(Z == 1, s[2], s[3]), sd = 1)
    })

    ## preset -> fill the three boxes
    observeEvent(list(input$preset, input$dataset, input$n, seed()), {
        pr <- presets[[input$preset]]
        if (is.null(pr)) return(invisible(NULL))          # manual entry
        if (length(pr) == 1 && is.na(pr)) {               # random labels
            y <- Y()
            lab <- rbinom(length(y), 1, 0.5)
            pr <- c(0.5, mean(y[lab == 1]), mean(y[lab == 0]))
        }
        updateNumericInput(session, "p0",   value = round(pr[1], 3))
        updateNumericInput(session, "mu10", value = round(pr[2], 3))
        updateNumericInput(session, "mu00", value = round(pr[3], 3))
    }, ignoreInit = FALSE)

    ## ---- iteration state driven by the three buttons ----------------------
    t_cur   <- reactiveVal(0)      # iteration currently displayed
    playing <- reactiveVal(FALSE)

    stop_play <- function() {
        playing(FALSE)
        updateActionButton(session, "play", label = "Play")
    }

    ## any change to the data or the starting value restarts the fit
    observeEvent(list(input$dataset, input$n, seed(), input$maxit,
                      input$p0, input$mu10, input$mu00), {
        t_cur(0)
        stop_play()
    })

    observeEvent(input$nxt, {
        stop_play()
        t_cur(min(t_cur() + 1, input$maxit))
    })

    observeEvent(input$prv, {
        stop_play()
        t_cur(max(t_cur() - 1, 0))
    })

    observeEvent(input$play, {
        if (playing()) {
            stop_play()
        } else {
            if (t_cur() >= input$maxit) t_cur(0)   # replay from the start
            playing(TRUE)
            updateActionButton(session, "play", label = "Pause")
        }
    })

    ## the timer: advances one iteration per tick while playing
    observe({
        if (!playing()) return(NULL)
        invalidateLater(700, session)
        isolate({
            if (t_cur() >= input$maxit) stop_play() else t_cur(t_cur() + 1)
        })
    })

    output$stepLabel <- renderText(
        sprintf("Iteration %d of %d", t_cur(), input$maxit)
    )

    theta0 <- reactive({
        p <- input$p0
        validate(need(is.finite(p) && p > 0 && p < 1, "p must lie strictly between 0 and 1"),
                 need(is.finite(input$mu10) && is.finite(input$mu00),
                      "both means must be numeric"))
        c(p, input$mu10, input$mu00)
    })

    path <- reactive(em_path(theta0(), Y(), input$maxit))

    output$fit <- renderPlot({
        y <- Y(); P <- path()
        k  <- min(t_cur() + 1, nrow(P))
        th <- P[k, 1:3]
        validate(need(all(is.finite(th)), "the iteration broke down numerically"))
        w  <- e_step(th, y)
        cw <- cm.colors(100)

        xr <- range(y, th[2:3]) + c(-2, 2)
        xg <- seq(xr[1], xr[2], length.out = 400)
        d1 <- dnorm(xg, th[2]) * th[1]
        d0 <- dnorm(xg, th[3]) * (1 - th[1])

        par(mar = c(4, 4, 3, 1))
        plot(xg, d1, type = "l", lwd = 2, col = cw[100],
             ylim = c(0, max(0.42, d1, d0)), xlab = "y", ylab = "weighted density",
             main = sprintf("step %d:  p = %.3f,  mu1 = %.3f,  mu0 = %.3f,  loglik = %.2f",
                            k - 1, th[1], th[2], th[3], P[k, 4]))
        lines(xg, d0, lwd = 2, col = cw[1])
        points(y, rep(0, length(y)), col = cw[pmax(1, ceiling(w * 100))], pch = 19,
               cex = 0.7)
        if (k < nrow(P) && all(is.finite(P[k + 1, 1:3]))) {
            arrows(th[2], 0.02, P[k + 1, 2], 0.02, length = 0.08, lwd = 2, col = cw[100])
            arrows(th[3], 0.02, P[k + 1, 3], 0.02, length = 0.08, lwd = 2, col = cw[1])
        }
        legend("topright", bty = "n", lwd = 2, col = c(cw[100], cw[1]),
               legend = c(expression(hat(p) * phi(y * "|" * hat(mu)[1])),
                          expression((1 - hat(p)) * phi(y * "|" * hat(mu)[0]))))
    })

    output$trace <- renderPlot({
        P <- path(); k <- min(t_cur() + 1, nrow(P))
        par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
        it <- 0:(nrow(P) - 1)
        plot(it, P[, 4], type = "n", xlab = "iteration",
             ylab = "observed log-likelihood", main = "must never decrease")
        lines(it[1:k], P[1:k, 4], type = "b", pch = 19, col = "firebrick")
        yl <- range(P[, 2:3], na.rm = TRUE)
        plot(it, P[, 2], type = "n", ylim = yl, xlab = "iteration",
             ylab = "component means", main = "parameter paths")
        lines(it[1:k], P[1:k, 2], type = "b", pch = 19, col = "firebrick")
        lines(it[1:k], P[1:k, 3], type = "b", pch = 19, col = "steelblue")
        legend("right", bty = "n", lwd = 2, col = c("firebrick", "steelblue"),
               legend = c(expression(hat(mu)[1]), expression(hat(mu)[0])))
    })

    output$tab <- renderTable({
        P <- path(); k <- min(t_cur() + 1, nrow(P))
        d <- as.data.frame(P[seq_len(k), , drop = FALSE])
        d$iter <- seq_len(k) - 1
        tail(d[, c("iter", "p", "mu1", "mu0", "log_lik")], 5)
    }, digits = 4, rownames = FALSE)
}

shinyApp(ui, server)
```


Four behaviours are worth producing deliberately.

**Slow drift from a distant start.** With $(0.5, -10, 5)$ both means begin far
outside the data. The early iterations move them very little, because every
observation is assigned almost the same responsibility; progress accelerates
only once one mean reaches the data.

**Label switching.** The starts $(0.5, 0, 3)$ and $(0.5, 3, 0)$ converge to the
same fitted mixture with the roles of the two components exchanged. The
likelihood is invariant to this relabelling, so the fitted density is identical
while the parameter estimates differ — a genuine identifiability problem for the
model, not a defect of the algorithm.

**Stalling at a symmetric point.** With $\mu_1^{(0)} = 0.1$ and
$\mu_0^{(0)} = 0.2$ the two components are nearly indistinguishable, every
responsibility is close to $\hat p^{(0)}$, and the algorithm creeps away from the
one-component solution very slowly. That configuration is close to a saddle
point of $\ell_{\text{obs}}$.

**Separation and speed.** Switch the data set from $\mu = (1.5, 0)$ to
$\mu = (5, 0)$. Well-separated components make the responsibilities nearly $0$ or
$1$, the E-step becomes almost a hard classification, and convergence takes a
handful of iterations rather than dozens. This is the practical face of the
general result that EM's linear convergence rate is governed by the fraction of
missing information.

### Comparison with general-purpose optimizers

EM is not the only way to maximize the observed likelihood. Applying a
general-purpose optimizer directly requires attention to the parameter space:
$p$ must lie in $(0,1)$, so we reparameterize with the logit.

```{r}
#| label: mix-compare
#| warning: false

neg_loglike_obs <- function(theta, Y) -log_like_obs(theta, Y)

### transformed parameterization: ttheta[1] = logit(p)
neg_loglike_obs_transf <- function(ttheta, Y) {
    theta <- ttheta
    theta[1] <- 1 / (1 + exp(-ttheta[1]))
    neg_loglike_obs(theta, Y)
}

lp0 <- log(0.3 / 0.7)

### Newton-type, untransformed: p can leave (0, 1)
try(nlm(neg_loglike_obs, p = c(0.3, 0, 3), Y = data)$estimate)
try(nlm(neg_loglike_obs, p = c(0.5, -10, 5), Y = data)$estimate)

### Newton-type, transformed
try(nlm(neg_loglike_obs_transf, p = c(lp0, 0, 3), Y = data)$estimate)
try(nlm(neg_loglike_obs_transf, p = c(0, -10, 5), Y = data)$estimate)

### Nelder-Mead
try(optim(c(lp0, 0, 3), neg_loglike_obs_transf, Y = data, method = "Nelder-Mead")$par)
try(optim(c(0, -10, 5), neg_loglike_obs_transf, Y = data, method = "Nelder-Mead")$par)

### Conjugate gradient
try(optim(c(0, 0, 3), neg_loglike_obs_transf, Y = data, method = "CG")$par)
try(optim(c(0, -10, 5), neg_loglike_obs_transf, Y = data, method = "CG")$par)
```

Applied to the observed log-likelihood, these algorithms are more fragile than
EM when the initial values are poorly chosen: the untransformed runs can step
outside $p \in (0,1)$ and produce `NaN`, and the gradient-based methods can stall
in the flat region that a distant start creates. Nelder–Mead is the most stable
of the three here, at the cost of many more function evaluations.

| | EM | General optimizer on $\ell_{\text{obs}}$ |
|---|---|---|
| Constraints | Respected automatically by the M-step | Require reparameterization |
| Monotonicity | Guaranteed; a decrease signals a coding error | None; a bad step can be accepted |
| Derivatives | None needed | Gradient helpful, Hessian better |
| Convergence rate | Linear, rate set by the missing information | Superlinear or quadratic near the optimum |
| Standard errors | Not produced; need Louis' method or a bootstrap | Hessian available directly |
| Behaviour from poor starts | Slow but stable | Can fail outright |

: EM compared with direct numerical maximization {#tbl-em-vs-optim}

The last row is the practical trade. EM rarely fails and rarely needs tuning,
which is why it dominates in mixture and latent-variable modelling; it is also
slow near the optimum and gives no standard errors, which is why a hybrid — EM
to get close, then a Newton step to polish and to obtain the Hessian — is common
in production code.

## Example 2: Censored Poisson Data {#sec-poisson}

### Truncated distributions in brief

Let $X \sim f(x)$ and $Y = I(X > c)$. Then $Y \sim \text{Bern}\{P(X > c)\}$, and
the conditional distribution of $X$ given $Y$ is $f$ restricted to one side of
$c$ and renormalized:

$$
f(x \mid Y = 1) = \frac{f(x)\, I(x > c)}{\int_c^{\infty} f(t)\,dt},
\qquad
f(x \mid Y = 0) = \frac{f(x)\, I(x \le c)}{\int_{-\infty}^{c} f(t)\,dt} .
$$

For discrete $X$ the integrals are sums. This is exactly the distribution the
E-step needs when an observation is known only to lie on one side of $c$.

```{r}
#| label: fig-truncation
#| fig-width: 10
#| fig-height: 4.8
#| fig-cap: "A density split at a cut point, and the two renormalized conditional densities."
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1)); cc <- 1
curve(dnorm, -3.5, 3.5, lwd = 3, xlab = "x", ylab = "density",
      main = "f(x) with cut point c")
xs <- seq(cc, 3.5, length.out = 100)
polygon(c(xs, rev(xs)), c(dnorm(xs), rep(0, 100)),
        col = adjustcolor("firebrick", 0.3), border = NA)
xs <- seq(-3.5, cc, length.out = 100)
polygon(c(xs, rev(xs)), c(dnorm(xs), rep(0, 100)),
        col = adjustcolor("steelblue", 0.3), border = NA)
abline(v = cc, lty = 2); text(cc, 0.38, "c", pos = 4)
text(1.8, 0.05, "P(X > c)", col = "firebrick")
text(-1, 0.1, expression(P(X <= c)), col = "steelblue")
curve(dnorm(x) * (x > cc) / pnorm(cc, lower.tail = FALSE), -3.5, 3.5, lwd = 3,
      col = "firebrick", n = 1001, xlab = "x", ylab = "density",
      main = "conditional densities", ylim = c(0, 1.6))
curve(dnorm(x) * (x <= cc) / pnorm(cc), add = TRUE, lwd = 3, col = "steelblue", n = 1001)
abline(v = cc, lty = 2)
legend("topleft", bty = "n", lwd = 3, col = c("steelblue", "firebrick"),
       legend = c("f(x | Y = 0)", "f(x | Y = 1)"))
```

### Data and likelihoods

Let $y_i \overset{iid}{\sim} \text{Pois}(\lambda)$ for $i = 1, \ldots, n$. The
first $m$ values are observed exactly; for $i > m$ we know only that $y_i < 2$,
recorded as $x_i = I(y_i < 2) = 1$.

**Observed data**: $y_1, \ldots, y_m$ together with
$x_{m+1} = \cdots = x_n = 1$. **Missing**: $y_{m+1}, \ldots, y_n$.

**Complete likelihood**

$$
\prod_{i=1}^{n} f^{\text{Pois}}(y_i \mid \lambda) \cdot \prod_{i=m+1}^{n} I(y_i < 2) .
$$

**Observed likelihood**

$$
\prod_{i=1}^{m} f^{\text{Pois}}(y_i \mid \lambda) \cdot \prod_{i=m+1}^{n} P(y_i < 2 \mid \lambda),
\qquad P(y_i < 2 \mid \lambda) = e^{-\lambda}(1 + \lambda) .
$$

Taking logarithms and dropping constants,

$$
\ell_{\text{obs}}(\lambda) = m \bar y_{1:m} \log \lambda - n \lambda + (n - m)\log(1 + \lambda) .
$$

The second factor makes the MLE nonlinear in $\lambda$; EM sidesteps it.

### E-step

$$
\ell_{\text{comp}}(\lambda) = -n\lambda + (\log \lambda) \sum_{i=1}^{m} y_i
+ (\log \lambda) \underbrace{\sum_{i=m+1}^{n} y_i}_{\text{missing}}
- \sum_{i=1}^{n} \log(y_i!) .
$$

For $i > m$ we have $y_i \in \{0, 1\}$, and the Poisson pmf truncated to that set
given $\lambda^{(t)}$ is

$$
P\!\left(y_i = 0 \mid x_i = 1\right) = \frac{1}{1 + \lambda^{(t)}},
\qquad
P\!\left(y_i = 1 \mid x_i = 1\right) = \frac{\lambda^{(t)}}{1 + \lambda^{(t)}} ,
$$

so that

$$
\hat y^{(t)} \equiv E\!\left(y_i \mid x_i = 1, \lambda^{(t)}\right) = \frac{\lambda^{(t)}}{1 + \lambda^{(t)}} .
$$

Since $y_i! = 1$ for $y_i \in \{0, 1\}$, the $\log(y_i!)$ term is a known
constant and no further expectation is needed.

```{r}
#| label: fig-poisson-estep
#| fig-width: 10
#| fig-height: 4.4
#| fig-cap: "The conditional distribution of a censored count is the Poisson pmf restricted to $\\{0,1\\}$ and renormalized."
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1)); lt <- 1.5; k <- 0:6
barplot(dpois(k, lt), names.arg = k,
        col = c("firebrick", "firebrick", rep("gray80", 5)),
        xlab = "y", ylab = "probability",
        main = bquote("Pois(" * lambda^(t) == .(lt) * ")"))
barplot(c(1, lt) / (1 + lt), names.arg = 0:1, col = "firebrick",
        xlab = "y", ylab = "probability",
        main = expression("truncated to {0, 1}: " * P(y[i]*"|"*x[i] == 1)))
```

### M-step and the EM update

$$
Q\!\left(\lambda \mid \lambda^{(t)}\right)
= -n\lambda + (\log \lambda)\left[\sum_{i=1}^{m} y_i + (n - m)\,\hat y^{(t)}\right] + C^{(t)} ,
$$

$$
\frac{\partial Q}{\partial \lambda}
= -n + \frac{1}{\lambda}\left[\sum_{i=1}^{m} y_i + (n - m)\,\hat y^{(t)}\right] = 0 ,
$$

$$
\Longrightarrow \quad
\lambda^{(t+1)} = \frac{m\, \bar y_{1:m} + (n - m)\dfrac{\lambda^{(t)}}{1 + \lambda^{(t)}}}{n} .
$$

The update is the ordinary Poisson MLE — the sample mean — with each censored
value replaced by its conditional expectation. As warned in the overview, this
works *only* because $\ell_{\text{comp}}$ is linear in the missing $y_i$.

### Implementation

```{r}
#| label: poisson-em

### observed log-likelihood, up to an additive constant
##   m    : number of exactly observed counts
##   ybar : their mean
##   ncen : number of counts known only to be < 2
loglik_obs_pois <- function(m, ybar, ncen, lambda) {
    m * ybar * log(lambda) - (m + ncen) * lambda + ncen * log(1 + lambda)
}

em_censored_poisson <- function(m, ybar, ncen,
                                lambda0 = (m * ybar + ncen) / (m + ncen),
                                iterations = 20) {
    out <- matrix(NA_real_, iterations + 1, 2,
                  dimnames = list(NULL, c("lambda", "log_lik")))
    lambda <- lambda0
    out[1, ] <- c(lambda, loglik_obs_pois(m, ybar, ncen, lambda))
    for (i in seq_len(iterations)) {
        y_mis  <- lambda / (1 + lambda)                       # E-step
        lambda <- (m * ybar + ncen * y_mis) / (m + ncen)       # M-step
        out[i + 1, ] <- c(lambda, loglik_obs_pois(m, ybar, ncen, lambda))
    }
    out
}
```

```{r}
#| label: poisson-data
set.seed(812)
y    <- rpois(200, 3)
ncen <- sum(y < 2)
m    <- length(y) - ncen
ybar <- mean(y[y >= 2])
c(m = m, ncen = ncen, ybar_observed = ybar)
```

Discarding the censored observations and averaging the rest gives an estimate
clearly above the true mean of $3$: conditioning on $y \ge 2$ removes the small
counts and biases the mean upward. EM recovers them.

```{r}
#| label: poisson-run
round(em_censored_poisson(m, ybar, ncen, lambda0 = 100, iterations = 10), 5)
round(em_censored_poisson(m, ybar, ncen, lambda0 = ybar, iterations = 10), 5)
```

Both starting values reach the same limit, and the log-likelihood increases
monotonically from each. Note how slowly the run from $\lambda^{(0)} = 100$
descends at first — a single-parameter illustration of the same slow-drift
behaviour seen in the mixture app.

### Comparison with general-purpose optimizers

```{r}
#| label: poisson-compare
#| warning: false
neg_loglik_pois <- function(lambda, m, ybar, ncen) {
    -loglik_obs_pois(m, ybar, ncen, lambda)
}

try(nlm(neg_loglik_pois, p = ybar, m = m, ybar = ybar, ncen = ncen)$estimate)
try(nlm(neg_loglik_pois, p = 100,  m = m, ybar = ybar, ncen = ncen)$estimate)
try(optim(ybar, neg_loglik_pois, method = "CG",
          m = m, ybar = ybar, ncen = ncen)$par)
try(optim(100,  neg_loglik_pois, method = "CG",
          m = m, ybar = ybar, ncen = ncen)$par)
```

With one parameter and a smooth concave objective this is an easy problem for
any method; the contrast with the mixture example is instructive, since there
the multimodality and the constraint on $p$ were what made the general
optimizers fragile.

## More Complex Applications

### Model-based clustering

$$
Z_i \sim \text{Categorical}(\pi_1, \ldots, \pi_K),
\qquad Y_i \mid Z_i = k \sim N_d(\mu_k, \Sigma_k) .
$$

This is the multivariate, $K$-component generalization of @sec-mixture. The
E-step computes soft cluster memberships
$P(Z_i = k \mid y_i, \hat\theta^{(t)})$; the M-step returns weighted means,
covariances and proportions. The `mclust` package implements this with a family
of constrained covariance structures.

```{r}
#| label: fig-clustering
#| fig-width: 6
#| fig-height: 4.6
#| fig-cap: "Three-component multivariate normal mixture with unequal covariances."
par(mar = c(4, 4, 1, 1))
ctr <- rbind(c(0, 0), c(4, 3), c(1, 5))
Sig <- list(diag(2) * 0.8, matrix(c(1, 0.7, 0.7, 1), 2), diag(c(1.2, 0.4)))
cols <- c("steelblue", "firebrick", "forestgreen")
ang <- seq(0, 2 * pi, length.out = 100)
plot(NA, xlim = c(-3, 7), ylim = c(-3, 8),
     xlab = expression(y[1]), ylab = expression(y[2]))
for (k in 1:3) {
  R <- chol(Sig[[k]])
  pts <- matrix(rnorm(120), 60) %*% R + rep(ctr[k, ], each = 60)
  points(pts, col = adjustcolor(cols[k], 0.6), pch = 19, cex = 0.8)
  ell <- (cbind(cos(ang), sin(ang)) * 2.1) %*% R + rep(ctr[k, ], each = 100)
  lines(ell, col = cols[k], lwd = 2)
  points(ctr[k, 1], ctr[k, 2], pch = 4, lwd = 3, col = cols[k])
}
```

### Hidden Markov models

Hidden states $h_1 \to h_2 \to \cdots \to h_n$ form a Markov chain, and each
observation depends only on its own state:

$$
P(y_{1:n} \mid h_{1:n}, \theta) = \prod_{i=1}^{n} P(y_i \mid h_i, \theta) .
$$

The E-step is the forward–backward algorithm, which computes
$P(h_i \mid y_{1:n}, \hat\theta^{(t)})$ and
$P(h_i, h_{i+1} \mid y_{1:n}, \hat\theta^{(t)})$ without enumerating all $K^n$
state sequences; the M-step produces weighted transition and emission estimates.
The combination is known as the Baum–Welch algorithm, and it is the clearest
example of an E-step that is itself a nontrivial computation.

```{r}
#| label: fig-hmm
#| fig-width: 10
#| fig-height: 3.4
#| fig-cap: "A two-state hidden Markov chain with normal emissions; background shading shows the hidden state."
par(mar = c(4, 4, 1, 1)); N <- 120
Pm <- matrix(c(0.95, 0.05, 0.1, 0.9), 2, byrow = TRUE)
h <- integer(N); h[1] <- 1
for (i in 2:N) h[i] <- sample(1:2, 1, prob = Pm[h[i - 1], ])
yy <- rnorm(N, mean = c(0, 3)[h])
plot(yy, type = "n", xlab = "i", ylab = expression(y[i]))
r <- rle(h); e <- cumsum(r$lengths); s <- c(1, head(e, -1) + 1)
rect(s - 0.5, par("usr")[3], e + 0.5, par("usr")[4],
     col = adjustcolor(c("steelblue", "firebrick")[r$values], 0.15), border = NA)
lines(yy, col = "gray30")
points(yy, pch = 19, col = c("steelblue", "firebrick")[h], cex = 0.7)
legend("topright", bty = "n", pch = 19, col = c("steelblue", "firebrick"),
       legend = c(expression(h[i] == 1), expression(h[i] == 2)), horiz = TRUE)
```

### Factor analysis and the common pattern

$$
y_i = \Gamma x_i + \varepsilon_i,
\qquad x_i \sim N_q(0, I) \text{ hidden},
\qquad \varepsilon_i \sim N_d(0, \Psi) .
$$

The E-step uses the fact that $x_i \mid y_i, \hat\theta^{(t)}$ is normal with
closed-form mean and covariance; the M-step is a regression of $y$ on the
expected factors.

Across all these applications the recipe is the same:

1. Introduce a hidden variable that makes the complete-data model tractable.
2. **E-step**: compute the conditional distribution of the hidden data, or just
   the expected sufficient statistics if $\ell_{\text{comp}}$ is linear in them.
3. **M-step**: compute the complete-data MLE with those expected sufficient
   statistics in place of the unobserved ones.

The art is entirely in step 1. The same observed-data model often admits several
choices of $Z$, and they lead to EM algorithms with quite different convergence
rates — faster when the augmentation adds little information, slower when the
missing information is a large fraction of the total.

## Summary

- EM maximizes $\ell_{\text{obs}}$ by alternating between forming
  $Q(\theta \mid \hat\theta^{(t)}) = E_{Z \mid Y, \hat\theta^{(t)}}[\ell_{\text{comp}}]$
  and maximizing it.
- Every iteration weakly increases $\ell_{\text{obs}}$. Equivalently, EM is
  coordinate ascent on
  $F(\tilde P, \theta) = \ell_{\text{obs}}(\theta) - \mathrm{KL}(\tilde P \,\|\, P(Z \mid Y, \theta))$.
- When $\ell_{\text{comp}}$ is linear in sufficient statistics of $Z$, the E-step
  reduces to computing their conditional expectations. Otherwise the full
  conditional distribution is required, and "plug in the expectation" is wrong.
- Convergence is linear, often slow, and only to a stationary point. Run from
  several starting values and compare the attained log-likelihoods.
- Standard errors are not a by-product. Use Louis' identity, the SEM algorithm,
  a bootstrap, or a final Newton step.

