---
title: "Optimization for Maximum Likelihood Estimation"
format:
  html:
    toc: true
    tbl-colwidths: true
filters:
  - shinylive
editor_options:
  chunk_output_type: console
code-fold: true
---

This chapter develops maximum likelihood estimation and the numerical methods
used to find it, first for a scalar parameter and then for a parameter vector.
The univariate case, where the log-likelihood can be plotted and a root can be
bracketed, motivates the two ideas that carry over unchanged to higher
dimensions: replacing the log-likelihood locally by a quadratic and solving for
its maximum (Newton–Raphson), and using the curvature of that quadratic to
report a standard error. The multivariate case, where neither plotting nor
bracketing is available, shows what survives of that idea and what has to be
added — safeguards against divergence, and a wider menu of derivative-free and
gradient-only methods (Nelder–Mead, gradient descent, conjugate gradient,
BFGS, and stochastic gradient descent) for when the Hessian is unavailable or
too expensive. Several of these methods are illustrated below with live,
in-browser animations that let you change the objective function, the
starting point, and the algorithm before watching the iterates respond.


```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false
library(latex2exp)   # TeX() renders LaTeX in plot titles, axis labels and legends
set.seed(812)
```

## Maximum Likelihood: Theory

### Likelihood, Score, and Hessian

A model specifies the density or probability of the data $D$ given a parameter $\theta$, written $f(D \mid \theta)$. As a function of $D$ with $\theta$ fixed it is the sampling distribution; as a function of $\theta$ with $D$ observed it is the likelihood. For independent observations $x_1, \ldots, x_n$,
$$
L(\theta) = \prod_{i=1}^{n} f(x_i \mid \theta), \qquad \ell(\theta) = \log L(\theta) = \sum_{i=1}^{n} \log f(x_i \mid \theta).
$$

We always work with $\ell$: the product underflows to zero for even moderate $n$, and sums are easier to differentiate. The maximum likelihood estimator maximizes $\ell$, and where $\ell$ is smooth with an interior maximum, $\hat\theta$ solves the **score equation**.
$$
S(\theta) = \ell'(\theta) = 0, \qquad \ell''(\hat\theta) < 0 .
$$

Three quantities recur throughout the chapter:

- the **score** $S(\theta) = \ell'(\theta)$, which locates the maximum;

- the **observed information** $J(\theta) = -\ell''(\theta)$, the curvature of the log-likelihood at $\theta$;
- the **expected information** $n I(\theta) = -E_D\left[\ell''(\theta; D)\right]$, its expectation under the model.

The second derivative therefore plays a double role. It determines how a Newton-type algorithm steps toward the maximum, and it determines how precisely the data pin the parameter down. An optimizer that computes the curvature has computed the standard error at the same time.

### Example: iid Bernoulli

Before turning to a model with a continuous parameter and an exactly quadratic
log-likelihood, it is worth solving the score equation by hand in the simplest
possible case. Let $Y_1, \ldots, Y_n \overset{iid}{\sim} \mathrm{Bernoulli}(\theta)$,
so that $P(y_i \mid \theta) = \theta^{y_i}(1 - \theta)^{1 - y_i}$. Writing
$n_1 = \sum_i y_i$ for the number of successes and $n_0 = n - n_1$ for the
number of failures,
$$
L(\theta) = \theta^{n_1}(1 - \theta)^{n_0}, \qquad \ell(\theta) = n_1 \log\theta + n_0 \log(1 - \theta).
$$

The score is.
$$
S(\theta) = \frac{n_1}{\theta} - \frac{n_0}{1 - \theta},
$$
and setting $S(\theta) = 0$ gives $n_1(1-\theta) = n_0\theta$, i.e.
$n_1 = (n_1 + n_0)\theta = n\theta$, so
$$
\hat\theta = \frac{n_1}{n} = \overline y.
$$

The second derivative $\ell''(\theta) = -n_1/\theta^{2} - n_0/(1-\theta)^{2}$ is
negative wherever it is defined, confirming that $\overline y$ is a maximum and not
a saddle point. With only three observations, $y = (1, 1, 0)$, the likelihood
can even be tabulated and plotted directly, which makes the maximum at
$\hat\theta = 2/3 = \overline y$ visible without any calculus at all.

| $\theta$ | $P(1, 1, 0 \mid \theta) = \theta^{2}(1-\theta)$ |
|:---:|:---|
| $0$ | $0$ |
| $1/2$ | $1/8 = 0.1250$ |
| $2/3$ | $4/27 = 0.1481$ **(max)** |
| $3/4$ | $9/64 = 0.1406$ |
| $1$ | $0$ |

```{r}
#| label: fig-bernoulli-toy-likelihood
#| fig-width: 5
#| fig-height: 4
#| fig-cap: "Likelihood of $\\theta$ for three Bernoulli observations $y = (1,1,0)$: the maximum sits at $\\hat\\theta = \\bar y = 2/3$."
curve(x^2 * (1 - x), 0, 1, lwd = 2, xlab = TeX(r"($\theta$)"), ylab = "likelihood")
abline(v = 2 / 3, col = 2, lty = 2)
```

### Closed Form: the Normal Mean

Let $X_1, \ldots, X_n$ be iid $N(\mu, \sigma_0^{2})$ with $\sigma_0^{2}$ known. Using the identity $\sum_i (x_i - \mu)^2 = \sum_i (x_i - \overline x)^2 + n(\overline x - \mu)^2$,
$$
\ell(\mu) = -\frac{n}{2}\log\left(2\pi\sigma_0^{2}\right) - \frac{1}{2\sigma_0^{2}}\sum_{i=1}^{n}(x_i - \mu)^{2} = C - \frac{n}{2\sigma_0^{2}}(\mu - \overline x)^{2} .
$$

Differentiating,
$$
S(\mu) = \frac{n}{\sigma_0^{2}}\left(\overline x - \mu\right), \qquad \ell''(\mu) = -\frac{n}{\sigma_0^{2}}, \qquad \hat\mu = \overline x .
$$

Every feature of the general theory is visible here in exact form. The log-likelihood is a downward parabola, so it has a unique maximum. The score is linear, so the score equation is solved in one line. The curvature $J(\mu) = n/\sigma_0^{2}$ does not depend on $\mu$ or on the data, so observed and expected information coincide, $J = nI$, and both grow linearly with $n$: doubling the sample doubles the sharpness of the peak and halves the variance of $\hat\mu$.

```{r}
#| label: fig-normal-loglik-score-curvature
#| fig-width: 10
#| fig-height: 4
#| fig-cap: "Log-likelihood, score, and curvature for a normal mean at $n = 20$ and $n = 50$: the larger sample gives a sharper peak, a steeper score, and larger curvature."
x20 <- rnorm(20, mean = 1); x50 <- rnorm(50, mean = 1)
mu <- seq(0, 2, length.out = 200)

ll_norm <- function(mu, x) vapply(mu, function(m) sum(dnorm(x, m, log = TRUE)), numeric(1))

par(mfrow = c(1, 3), mar = c(4, 4.5, 2.5, 1))

matplot(mu, cbind(ll_norm(mu, x20) - max(ll_norm(mu, x20)),
                  ll_norm(mu, x50) - max(ll_norm(mu, x50))),
        type = "l", lty = 1, lwd = 2, col = c(1, 2),
        xlab = TeX(r"($\mu$)"), ylab = TeX(r"($\ell(\mu) - \ell(\hat{\mu})$)"),
        main = TeX(r"(log-likelihood)"))
legend("bottom", bty = "n", lty = 1, lwd = 2, col = c(1, 2),
       legend = TeX(c(r"($n = 20$)", r"($n = 50$)")))

matplot(mu, cbind(20 * (mean(x20) - mu), 50 * (mean(x50) - mu)),
        type = "l", lty = 1, lwd = 2, col = c(1, 2),
        xlab = TeX(r"($\mu$)"), ylab = TeX(r"($S(\mu)$)"), main = TeX(r"(score)"))
abline(h = 0, lty = 3)

matplot(mu, cbind(rep(20, 200), rep(50, 200)), type = "l", lty = 1, lwd = 2, col = c(1, 2),
        ylim = c(0, 60), xlab = TeX(r"($\mu$)"), ylab = TeX(r"($-\ell''(\mu)$)"),
        main = TeX(r"(curvature)"))
```

The score crosses zero exactly where $\ell$ peaks, and the larger sample gives a more peaked log-likelihood, a steeper score, and a larger curvature.

### Consistency, Asymptotic Normality, and Fisher Information

Under regularity conditions — the support of $f$ not depending on $\theta$, the true value interior to the parameter space, and two derivatives passing under the integral sign — the MLE has three properties as $n \to \infty$, with $\theta_*$ the true parameter.

**Consistency.** $\hat\theta \to \theta_*$ in probability. The reason is that $\ell(\theta)/n$ converges to $E_{\theta_*}[\log f(X \mid \theta)]$, which by the information inequality is maximized at $\theta = \theta_*$: the log-likelihood surface converges to one whose peak is at the truth.
#### Asymptotic normality. {-}
$$
\hat\theta \;\dot\sim\; N\left(\theta_*,\ \left[n I(\theta_*)\right]^{-1}\right), \qquad I(\theta) = -E_X\left[\frac{\partial^{2}}{\partial\theta^{2}}\log f(X \mid \theta)\right] = E_X\left[\left(\frac{\partial}{\partial\theta}\log f(X\mid\theta)\right)^{2}\right].
$$

A sketch: expand $S(\hat\theta) = 0$ about $\theta_*$ to get $\hat\theta - \theta_* \approx S(\theta_*)/\{-\ell''(\theta_*)\}$. The numerator is a sum of iid mean-zero terms, hence asymptotically normal with variance $nI(\theta_*)$; the denominator converges to $nI(\theta_*)$ by the law of large numbers. Dividing gives variance $[nI(\theta_*)]^{-1}$.

**Asymptotic efficiency.** No consistent estimator has smaller asymptotic variance — the Cramér–Rao bound is attained.

Since $\theta_*$ is unknown, the variance is estimated by substituting $\hat\theta$. Two versions are available: the expected information $nI(\hat\theta)$, which requires the expectation analytically, and the observed information $J(\hat\theta) = -\ell''(\hat\theta)$, which requires only the curvature that a Newton step already computes. By the law of large numbers $J(\hat\theta)/n \approx I(\hat\theta)$, and both give
$$
\widehat{\mathrm{SE}}(\hat\theta) = J(\hat\theta)^{-1/2} \quad\text{or}\quad \left[nI(\hat\theta)\right]^{-1/2}, \qquad \hat\theta \pm 1.96\,\widehat{\mathrm{SE}}(\hat\theta).
$$

These are asymptotic statements. A flat log-likelihood — small curvature — gives a large standard error, and a log-likelihood that is far from quadratic gives a standard error that should not be trusted at all; the last section of this chapter shows a case of exactly that.

### Demonstration: the Cauchy Location Model

The Cauchy density $f(x \mid \theta) = 1/\{\pi[1 + (x - \theta)^2]\}$ has no closed-form MLE:
$$
\ell(\theta) = -n\log\pi - \sum_{i=1}^{n}\log\left(1 + (x_i - \theta)^{2}\right), \qquad
S(\theta) = \sum_{i=1}^{n}\frac{2(x_i - \theta)}{1 + (x_i - \theta)^{2}} .
$$

The score equation is a polynomial equation of degree $2n - 1$. Plotting $\ell$ for samples of increasing size shows the peak sharpening around the truth, which is consistency and growing information in one picture.

```{r}
#| label: fig-cauchy-loglik-by-n
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Cauchy log-likelihood for samples of increasing size ($n = 50, 100, 200$) from a true location $\\theta_* = 20$: the peak sharpens around the truth as $n$ grows."
log_like_cauchy <- function(theta, x) sum(dcauchy(x, theta, log = TRUE))

thetas <- seq(15, 25, length.out = 200)
n_set <- c(50, 100, 200)

ll_cauchy <- sapply(n_set, function(n) {
  d <- rcauchy(n, location = 20)
  sapply(thetas, log_like_cauchy, x = d)
})
ll_cauchy <- sweep(ll_cauchy, 2, apply(ll_cauchy, 2, max))   # align peaks at 0

par(mar = c(4, 4.5, 2.5, 1))
matplot(thetas, ll_cauchy, type = "l", lty = 1, lwd = 2, col = c(1, 2, 4), ylim = c(-20, 0),
        xlab = TeX(r"($\theta$)"), ylab = TeX(r"($\ell(\theta) - \ell(\hat{\theta})$)"),
        main = TeX(r"(Cauchy log-likelihood, $\theta_* = 20$)"))
abline(v = 20, lty = 3)
legend("bottomright", bty = "n", lty = 1, lwd = 2, col = c(1, 2, 4),
       legend = TeX(sprintf(r"($n = %d$)", n_set)))
```

The variability of the MLE across repeated samples is the sampling distribution the theory describes. Repeating the experiment 50 times at two sample sizes shows the curves clustering more tightly, and their maxima concentrating, as $n$ grows.

```{r}
#| label: fig-cauchy-loglik-variability
#| fig-width: 10
#| fig-height: 4.5
#| cache: true
#| fig-cap: "50 repeated Cauchy log-likelihood curves at $n = 20$ versus $n = 200$: the curves cluster more tightly, and their maxima concentrate, as $n$ grows."
loglik_curves <- function(n, nrep = 50)
  replicate(nrep, {
    d <- rcauchy(n, location = 20)
    lc <- sapply(thetas, log_like_cauchy, x = d)
    lc - max(lc)
  })

curves_small <- loglik_curves(20)
curves_large <- loglik_curves(200)

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
for (obj in list(list(m = curves_small, n = 20), list(m = curves_large, n = 200)))
  matplot(thetas, obj$m, type = "l", lty = 1, col = adjustcolor(1, 0.35), ylim = c(-10, 0),
          xlab = TeX(r"($\theta$)"), ylab = TeX(r"($\ell(\theta) - \ell(\hat{\theta})$)"),
          main = TeX(sprintf(r"(sample size $n = %d$)", obj$n)))

### sampling distribution of the MLE at the two sample sizes
mle_at <- function(m) thetas[apply(m, 2, which.max)]
rbind(`n = 20`  = c(mean = mean(mle_at(curves_small)), sd = sd(mle_at(curves_small))),
      `n = 200` = c(mean = mean(mle_at(curves_large)), sd = sd(mle_at(curves_large))))
```

The standard deviation falls by roughly the factor $\sqrt{10}$ predicted by $[nI(\theta_*)]^{-1/2}$. Two features of these curves also warn us about computation: for small $n$ the log-likelihood has several local maxima, and its shape away from the peak is far from quadratic. Numerical optimization is required, and it must be used with care.

## Univariate Optimization Algorithms

### Three Equivalent Formulations

The same task can be posed three ways, and the algorithms below attack different versions of it:

- maximize $\ell(\theta)$;

- minimize $f(\theta) = -\ell(\theta)$;
- find a root of $S(\theta) = \ell'(\theta)$.

Root-finders are usually faster because they use derivative information, but they locate any stationary point, including minima and saddle points. Direct optimizers need only function values and always move toward a maximum, but converge more slowly. Which to prefer depends on whether derivatives are available and whether the function is smooth.

Each method below is illustrated with a schematic showing its first few iterations. To keep the pictures comparable they all use the same small problem: the per-observation log-likelihood of the zero-truncated Poisson model of the next section, with $\overline x$ fixed at 1.8. The root of its score is marked in every figure.

```{r}
#| label: demo-problem
xbar_d <- 1.8                                             # fixed sample mean

ll_d <- function(l) xbar_d * log(l) - l - log(1 - exp(-l))          # log-likelihood
S_d  <- function(l) xbar_d / l - 1 / (1 - exp(-l))                  # score
dS_d <- function(l) -xbar_d / l^2 + exp(-l) / (1 - exp(-l))^2       # its derivative
I_d  <- function(l) 1 / (l * (1 - exp(-l))) - exp(-l) / (1 - exp(-l))^2   # information
g_d  <- function(l) xbar_d * (1 - exp(-l))                          # fixed-point map

lstar <- uniroot(S_d, c(0.2, 5))$root
c(root = lstar, contraction_factor = xbar_d * exp(-lstar))
```

### Fixed-Point (Simple) Iteration

Rearrange the score equation algebraically into the form $\theta = g(\theta)$ and iterate $\theta_{k+1} = g(\theta_k)$. If $g$ maps an interval into itself and $|g'(\theta^{*})| < 1$ at the fixed point, the mean value theorem gives
$$\left|\theta_{k+1} - \theta^{*}\right| = \left|g(\theta_k) - g(\theta^{*})\right| \le \left|g'(\xi)\right| \left|\theta_k - \theta^{*}\right| ,$$
so the error contracts by roughly the constant factor $|g'(\theta^{*})|$ at each step. This is **linear convergence**: a fixed number of digits is gained per iteration, and the closer $|g'|$ is to 1 the slower the progress.

```{r}
#| label: fixed-point
fixed_point <- function(g, x0, tol = 1e-10, maxit = 500) {
  for (k in 1:maxit) {
    x1 <- g(x0)
    if (abs(x1 - x0) < tol) return(c(root = x1, iter = k))
    x0 <- x1
  }
  stop("not converged")
}
```

The mechanism is easiest to see as a **cobweb**: from $\lambda_k$ on the diagonal, move vertically to the curve $g$ to obtain $\lambda_{k+1}$, then horizontally back to the diagonal to start the next step. The iterates walk toward the intersection of $g$ with the $45^{\circ}$ line, and the shallower $g$ is at that crossing — the smaller $|g'|$ — the larger each stride toward it.

```{r}
#| label: fig-scheme-fixed-point
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Fixed-point iteration as a cobweb diagram: iterates walk toward the intersection of $g$ with the diagonal."
lam <- seq(0.2, 3, length.out = 300)

par(mar = c(4, 4.5, 2.5, 1))
plot(lam, g_d(lam), type = "l", lwd = 2, xlim = c(0.2, 3), ylim = c(0.2, 3),
     xlab = TeX(r"($\lambda_k$)"), ylab = TeX(r"($g(\lambda_k) = \lambda_{k+1}$)"),
     main = TeX(r"(fixed-point iteration: the cobweb)"))
abline(0, 1, lty = 2)

x0 <- 2.8
for (k in 1:5) {
  x1 <- g_d(x0)
  segments(x0, x0, x0, x1, col = 2, lwd = 2)          # up or down to the curve
  segments(x0, x1, x1, x1, col = 2, lwd = 2)          # across to the diagonal
  points(x0, x1, pch = 19, col = 2, cex = 0.7)
  if (k <= 3) text(x0, 0.25, TeX(sprintf(r"($\lambda_%d$)", k - 1)), col = 2, cex = 0.9)
  x0 <- x1
}
points(lstar, lstar, pch = 19, cex = 1.2)
text(lstar, lstar, TeX(r"($\hat{\lambda}$)"), pos = 4)
```

The method needs no derivatives, but the rearrangement must be found by hand, and different rearrangements of the same equation can converge at different speeds or diverge outright. Geometrically, a rearrangement with $|g'| > 1$ at the crossing produces a cobweb that spirals outward instead of inward.

### Bisection

If $S$ is continuous and $S(a)$ and $S(b)$ have opposite signs, a root lies in $[a, b]$. Evaluate $S$ at the midpoint and retain whichever half still changes sign.

```{r}
#| label: bisection
bisection <- function(S, a, b, tol = 1e-10) {
  if (S(a) * S(b) > 0) stop("endpoints do not bracket a root")
  k <- 0
  while (b - a > tol) {
    m <- (a + b) / 2
    if (S(a) * S(m) <= 0) b <- m else a <- m
    k <- k + 1
  }
  c(root = (a + b) / 2, iter = k)
}
```

The schematic below draws the bracket at each iteration as a bar beneath the score curve. The dot is the midpoint, and the dotted line carries it up to the curve, where its sign decides which half survives.

```{r}
#| label: fig-scheme-bisection
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Bisection on the score function: each iteration halves the bracket that contains the root."
par(mar = c(4, 4.5, 2.5, 1))
plot(lam, S_d(lam), type = "l", lwd = 2, ylim = c(-1.6, 3.5),
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($S(\lambda)$)"),
     main = TeX(r"(bisection: each step halves the bracket)"))
abline(h = 0, lty = 3)
points(lstar, 0, pch = 19, cex = 1.2)

a <- 0.3; b <- 3
ylev <- seq(-0.5, -1.4, length.out = 4)
for (k in 1:4) {
  m <- (a + b) / 2
  segments(a, ylev[k], b, ylev[k], col = 4, lwd = 3)
  points(c(a, b), rep(ylev[k], 2), pch = "|", col = 4)
  segments(m, ylev[k], m, S_d(m), lty = 3, col = 2)
  points(m, ylev[k], pch = 19, col = 2, cex = 0.8)
  text(0.25, ylev[k], k, cex = 0.8, col = 4)
  if (S_d(a) * S_d(m) <= 0) b <- m else a <- m
}
```

The bracket halves at every step, so after $k$ iterations the error is at most $(b - a)/2^{k}$ and the iteration count is known in advance: about $\log_2\{(b-a)/\text{tol}\}$, regardless of the shape of $S$. Bisection cannot diverge, which makes it the natural fallback when a faster method misbehaves. R's `uniroot` uses Brent's method, which retains the bracketing guarantee while interpolating for speed.

### Newton–Raphson

#### Derivation

Approximate the score by its tangent line at $\theta_k$ and solve for where the tangent crosses zero:
$$
S(\theta) \approx S(\theta_k) + S'(\theta_k)(\theta - \theta_k) = 0
\implies \theta_{k+1} = \theta_k - \frac{S(\theta_k)}{S'(\theta_k)} = \theta_k - \frac{\ell'(\theta_k)}{\ell''(\theta_k)} .
$$

Equivalently, replace $\ell$ by its second-order Taylor expansion at $\theta_k$ and jump to the maximum of that parabola. For an exactly quadratic $\ell(\theta) = -a(\theta - u)^2 + C$ with $a > 0$ we have $\ell' = -2a(\theta - u)$ and $\ell'' = -2a$, so
$$
\theta_{k+1} = \theta_k - \frac{-2a(\theta_k - u)}{-2a} = u,
$$
in one step from any starting value. The normal-mean example of the previous section is exactly of this form: Newton converges immediately.

The left panel below traces three steps. At each iterate, the tangent to the score is followed down to the horizontal axis, and where it lands is the next iterate. As the curve straightens near the root, the tangent becomes an excellent approximation and the steps land almost exactly on the answer.

The right panel contrasts one Newton step with one step of the method of scoring from the same point. Newton uses the actual slope $S'(\lambda_k)$; scoring uses $-nI(\lambda_k)$, the slope averaged over hypothetical datasets. The two lines differ away from the root and coincide as $\lambda_k$ approaches it.

```{r}
#| label: fig-scheme-newton
#| fig-width: 10
#| fig-height: 4.8
#| fig-cap: "Newton-Raphson following the tangent to the score (left), and one Newton step compared with one Fisher-scoring step from the same point (right)."
par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))

### three Newton steps
plot(lam, S_d(lam), type = "l", lwd = 2, ylim = c(-1, 3.5),
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($S(\lambda)$)"),
     main = TeX(r"(Newton-Raphson: follow the tangent)"))
abline(h = 0, lty = 3)
points(lstar, 0, pch = 19, cex = 1.2)

x0 <- 0.45
for (k in 1:3) {
  x1 <- x0 - S_d(x0) / dS_d(x0)
  segments(x0, S_d(x0), x1, 0, col = 2, lwd = 2)
  segments(x1, 0, x1, S_d(x1), lty = 3, col = "grey40")
  points(c(x0, x1), c(S_d(x0), 0), pch = 19, col = 2, cex = 0.7)
  text(x0, S_d(x0), TeX(sprintf(r"($\lambda_%d$)", k - 1)), pos = 3, col = 2, cex = 0.9)
  x0 <- x1
}

## Newton step versus scoring step from the same point
plot(lam, S_d(lam), type = "l", lwd = 2, ylim = c(-1, 3.5),
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($S(\lambda)$)"),
     main = TeX(r"(observed versus expected curvature)"))
abline(h = 0, lty = 3)
points(lstar, 0, pch = 19, cex = 1.2)

x0 <- 0.45
x_nr <- x0 - S_d(x0) / dS_d(x0)          # slope  S'(lambda)
x_fs <- x0 + S_d(x0) / I_d(x0)           # slope -nI(lambda)
segments(x0, S_d(x0), x_nr, 0, col = 2, lwd = 2)
segments(x0, S_d(x0), x_fs, 0, col = 4, lwd = 2, lty = 5)
points(c(x0, x_nr, x_fs), c(S_d(x0), 0, 0), pch = 19, cex = 0.7, col = c(1, 2, 4))
legend("topright", bty = "n", lwd = 2, lty = c(1, 5), col = c(2, 4),
       legend = TeX(c(r"(Newton: slope $S'(\lambda_k)$)", r"(scoring: slope $-nI(\lambda_k)$)")))
```

### Advantage: quadratic convergence

Let $e_k = \theta_k - \hat\theta$ and expand $S(\hat\theta) = 0$ about $\theta_k$:
$$
0 = S(\theta_k) - S'(\theta_k)e_k + \tfrac12 S''(\xi)e_k^{2}
\implies e_{k+1} = e_k - \frac{S(\theta_k)}{S'(\theta_k)} = \frac{S''(\xi)}{2S'(\theta_k)}\,e_k^{2} .
$$

The error is squared at each step, so the number of correct digits roughly doubles. Two conditions appear in the constant: $S'(\hat\theta) \ne 0$, meaning the maximum is a genuine peak and not a flat spot, and $S''$ bounded near the root. Asymptotic normality says that log-likelihoods are approximately quadratic near the MLE, which is why Newton is so effective there.

```{r}
#| label: newton
newton <- function(S, dS, x0, tol = 1e-10, maxit = 100, trace = FALSE) {
  for (k in 1:maxit) {
    x1 <- x0 - S(x0) / dS(x0)
    if (trace) cat(sprintf("iter %2d: %.10f\n", k, x1))
    if (!is.finite(x1)) stop("diverged")
    if (abs(x1 - x0) < tol) return(c(root = x1, iter = k))
    x0 <- x1
  }
  stop("not converged")
}
```

### A Classic Illustration: Newton's Method for $\sqrt{a}$

**Derivation.** To find $\sqrt{a}$, take $f(x) = x^2 - a$, whose positive root is $\sqrt a$. Then $f'(x) = 2x$, and the Newton step is,
$$
x_{k+1} = x_k - \frac{f(x_k)}{f'(x_k)} = x_k - \frac{x_k^{2} - a}{2x_k} = \frac{1}{2}\left(x_k + \frac{a}{x_k}\right).
$$

The final form is the Babylonian rule: if $x_k$ is too small then $a/x_k$ is too large, and the root lies between them, so average the pair. Newton's tangent line lands exactly on the midpoint.

The error follows the general quadratic law. Writing $e_k = x_k - \sqrt a$,
$$
e_{k+1} = \frac{1}{2}\left(x_k + \frac{a}{x_k}\right) - \sqrt a = \frac{\left(x_k - \sqrt a\right)^{2}}{2x_k} = \frac{e_k^{2}}{2x_k},
$$

so the number of correct digits roughly doubles each step.

```{r}
#| label: babylonian-sqrt
babylonian <- function(a, x0 = 1, iters = 6)
  Reduce(function(x, k) (x + a / x) / 2, 1:iters, init = x0, accumulate = TRUE)

traces <- sapply(c(10, 20, 50), babylonian)
dimnames(traces) <- list(paste("iter", 0:6), sprintf("sqrt(%d)", c(10, 20, 50)))

print(traces, digits = 12)
sqrt(c(10, 20, 50))                       # for comparison
signif(abs(traces - rep(sqrt(c(10, 20, 50)), each = 7)), 3)    # errors
```

Starting from $x_0 = 1$ in every case, the first two steps do the coarse work — the larger $a$ is, the further the opening leap — and from the third step on the error squares: roughly $10^{-1}$, then $10^{-3}$, then $10^{-6}$, then machine precision. Six iterations suffice for all three, though $\sqrt{50}$ spends one more step than $\sqrt{10}$ recovering from the poor start.

### Problem: divergence

Newton uses only local information, so the quadratic model can be badly wrong away from the maximum. Three failures occur:

- **wrong curvature** — where $\ell'' > 0$ the model is a bowl, and the step heads toward a minimum or off to infinity;

- **overshooting** — where the score is nearly flat, $S'(\theta_k) \approx 0$ produces an enormous step, possibly outside the parameter space;
- **multiple roots** — the iteration converges to whichever stationary point it happens to reach, which need not be the global maximum.

The single-observation Cauchy likelihood is the standard illustration. With one observation at zero, $f(\theta) = -\ell(\theta) = \log(1 + \theta^{2})$ and $s(\theta) = f'(\theta) = 2\theta/(1 + \theta^{2})$, whose only root is $\theta = 0$. The score rises, turns over at $\theta = 1$, and decays to zero in the tails — exactly the flat region that defeats Newton.

```{r}
#| label: fig-cauchy-single-obs
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: "Single-observation Cauchy negative log-likelihood and its derivative: the derivative flattens in the tails, the region that defeats Newton-Raphson."
lik <- function(t) log(1 + t^2)                    # negative log-likelihood
s   <- function(t) 2 * t / (1 + t^2)               # its derivative
ds  <- function(t) 2 * (1 - t^2) / (1 + t^2)^2

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
curve(lik, -5, 5, n = 400, lwd = 2, xlab = TeX(r"($\theta$)"), ylab = TeX(r"($-\ell(\theta)$)"),
      main = TeX(r"(negative log-likelihood)"))
curve(s, -5, 5, n = 400, lwd = 2, xlab = TeX(r"($\theta$)"), ylab = TeX(r"($s(\theta)$)"),
      main = TeX(r"(derivative, with the flat tails)"))
abline(h = 0, lty = 3); abline(v = c(-1, 1), col = 2, lty = 2)
```

Tracing the iterates from three starting values shows the boundary at $|\theta_0| = 1$, where the derivative $s$ turns over:

```{r}
#| label: newton-divergence-trace
newton_trace <- function(s, ds, x0, iters = 8) {
  path <- numeric(iters + 1); path[1] <- x0
  for (k in 1:iters) {
    x0 <- x0 - s(x0) / ds(x0)
    path[k + 1] <- if (is.finite(x0)) x0 else NA
  }
  path
}

starts <- c(0.5, 0.8, 1.2)
traces <- vapply(starts, function(v) newton_trace(s, ds, v), numeric(9))
colnames(traces) <- sprintf("start = %.1f", starts)
rownames(traces) <- paste("iter", 0:8)
signif(traces, 5)
```

From $\theta_0 = 0.5$, inside the region where the quadratic model is adequate, Newton converges quadratically to zero. From $\theta_0 = 0.8$ the first step already overshoots into the tail, and from $\theta_0 = 1.2$ — beyond the inflection point of $s$ — the iterates run away, each step landing further out where the derivative is flatter still.

```{r}
#| label: newton-animation
#| eval: false
## Interactive versions of the same three trajectories (requires the animation
## package and a graphics device that can write animations)
library(animation)
ani.options(interval = 1, nmax = 50)
newton.method(FUN = s, init = 0.5, rg = c(-1.5, 1.5))
newton.method(FUN = s, init = 0.8, rg = c(-8, 1.5))
newton.method(FUN = s, init = 1.2, rg = c(-2, 20))
```

### Univariate Fisher Scoring

Newton's step divides by the *observed* curvature $S'(\theta_k) = \ell''(\theta_k)$, computed from the data at hand, which can be positive — or nearly zero — far from the maximum. Replacing it by the *expected* curvature $-nI(\theta_k)$ gives the **method of scoring**:
$$
\theta_{k+1} = \theta_k + \frac{S(\theta_k)}{nI(\theta_k)}.
$$

Because $nI(\theta) > 0$ for every $\theta$ by construction, the denominator never changes sign: every step of scoring is an ascent step, even at a point where Newton's step would point downhill or blow up. Near the maximum $nI(\hat\theta) \approx -\ell''(\hat\theta)$, so scoring and Newton coincide and both converge quickly; the two methods differ only in how they behave away from the root.

```{r}
#| label: fisher-scoring-univariate
fisher_scoring <- function(S, nI, x0, tol = 1e-10, maxit = 100) {
  for (k in 1:maxit) {
    x1 <- x0 + S(x0) / nI(x0)
    if (abs(x1 - x0) < tol) return(c(root = x1, iter = k))
    x0 <- x1
  }
  stop("not converged")
}

## a poor start: at lambda = 3.5 the truncated-Poisson score is nearly flat
c(newton_step_from_3.5 = 3.5 - S_d(3.5) / dS_d(3.5))    # overshoots to an infeasible negative lambda
rbind(newton  = tryCatch(newton(S_d, dS_d, 3.5), error = function(e) c(root = NA, iter = NA)),
      scoring = fisher_scoring(S_d, I_d, 3.5))
```

From this poor start, Newton's step lands on a negative — and hence infeasible — value of $\lambda$ and never recovers, while scoring's positive-definite denominator lands its very first step near $\lambda \approx 1.53$, already close to the root, and converges normally from there. The price of that safety is that $nI(\theta)$ must be available in closed form; where it is not, Newton with the safeguards below is the fallback.

### Safeguards

Three remedies cover most cases. Start from a robust estimate — the sample median for a location model. Apply **step halving**: accept the full Newton step only if it improves the log-likelihood, otherwise halve it until it does. Or replace the observed curvature by the expected information, giving the **method of scoring**,
$$
\theta_{k+1} = \theta_k + \frac{S(\theta_k)}{n I(\theta_k)} ,
$$
in which the divisor is positive by construction, so every step ascends. Near the MLE $nI(\hat\theta) \approx -\ell''(\hat\theta)$, so scoring inherits Newton's speed while being far more stable away from the optimum. Its cost is that the expectation must be available in closed form.

```{r}
#| label: safe-newton
safe_newton <- function(ll, S, dS, x0, lower = -Inf, upper = Inf,
                        tol = 1e-10, maxit = 100, max_halve = 40) {
  f0 <- ll(x0)
  for (k in 1:maxit) {
    step <- -S(x0) / dS(x0)
    if (dS(x0) > 0) step <- sign(S(x0))          # wrong curvature: just move uphill
    for (h in 0:max_halve) {                     # halve until feasible and improving
      x1 <- x0 + step / 2^h
      if (x1 > lower && x1 < upper && is.finite(ll(x1)) && ll(x1) >= f0) break
    }
    if (abs(x1 - x0) < tol) return(c(root = x1, iter = k, halvings = h))
    x0 <- x1; f0 <- ll(x1)
  }
  stop("not converged")
}
```

### Golden Section Search

Golden section minimizes $f = -\ell$ directly, using **no derivative at all**, which makes it the method of choice when $\ell$ is not differentiable or its derivative is unreliable. Keep a bracket $[x_l, x_u]$ containing the minimum of a unimodal $f$, with two interior points placed at the golden ratio $g = (3 - \sqrt5)/2 \approx 0.382$ of the way in from each end. Comparing the two interior values discards one end, and the ratio is chosen precisely so that the surviving interior point is correctly positioned for the next iteration: one new evaluation per step, with the bracket shrinking by the factor $0.618$.

```{r}
#| label: golden
golden <- function(f, brack.int, eps = 1e-4, ...) {
  g <- (3 - sqrt(5)) / 2
  xl <- min(brack.int); xu <- max(brack.int)
  tmp <- g * (xu - xl)
  xmu <- xu - tmp; xml <- xl + tmp
  fl <- f(xml, ...); fu <- f(xmu, ...)
  while (abs(xu - xl) > (1e-5 + abs(xl)) * eps) {
    if (fl < fu) {
      xu <- xmu; xmu <- xml; fu <- fl
      xml <- xl + g * (xu - xl); fl <- f(xml, ...)
    } else {
      xl <- xml; xml <- xmu; fl <- fu
      xmu <- xu - g * (xu - xl); fu <- f(xmu, ...)
    }
  }
  if (fl < fu) xml else xmu
}
```

The schematic makes the reuse visible. Each bar is the current bracket with its two interior points; the point that survives into the next iteration is drawn in blue and sits at exactly the right position there, so only the red point has to be evaluated afresh.

```{r}
#| label: fig-scheme-golden
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Golden section search: the interior point that survives (blue) is already correctly positioned for the next iteration, so only one new point (red) is evaluated per step."
f_d <- function(l) -ll_d(l)
gr <- (3 - sqrt(5)) / 2

par(mar = c(4, 4.5, 2.5, 1))
plot(lam, f_d(lam), type = "l", lwd = 2, ylim = c(0.2, 1.25),
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($-\ell(\lambda)$)"),
     main = TeX(r"(golden section: one new evaluation per step)"))
points(lstar, f_d(lstar), pch = 19, cex = 1.2)

xl <- 0.3; xu <- 3
ylev <- seq(0.44, 0.24, length.out = 3)
for (k in 1:3) {
  xml <- xl + gr * (xu - xl); xmu <- xu - gr * (xu - xl)
  segments(xl, ylev[k], xu, ylev[k], col = 4, lwd = 3)
  points(c(xl, xu), rep(ylev[k], 2), pch = "|", col = 4)
  keep_lower <- f_d(xml) < f_d(xmu)
  points(xml, ylev[k], pch = 19, cex = 0.9, col = if (keep_lower) 4 else 2)
  points(xmu, ylev[k], pch = 19, cex = 0.9, col = if (keep_lower) 2 else 4)
  text(0.25, ylev[k], k, cex = 0.8, col = 4)
  if (keep_lower) xu <- xmu else xl <- xml
}
```

The first test is the negative log-likelihood of the classical genetic linkage model, with cell counts $(1997, 906, 904, 32)$ and probabilities $\left((2+\theta)/4, (1-\theta)/4, (1-\theta)/4, \theta/4\right)$; it is smooth, so any method would work. The second, $f(x) = |x - 10|$, is not differentiable at its minimum, so no root-finder applies — golden section handles it without modification.

```{r}
#| label: fig-golden-tests
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: "Golden section search applied to a smooth genetic-linkage log-likelihood (left) and to a non-differentiable function $|x-10|$ (right), with the located minimizer marked in red."
f1 <- function(theta) -1997 * log(2 + theta) - 1810 * log(1 - theta) - 32 * log(theta)
f2 <- function(x) abs(x - 10)

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
curve(f1, 0.001, 0.999, n = 400, lwd = 2, xlab = TeX(r"($\theta$)"), ylab = TeX(r"($-\ell(\theta)$)"),
      main = TeX(r"(genetic linkage model)"))
abline(v = golden(f1, c(0.001, 0.999)), col = 2, lty = 2, lwd = 2)

curve(f2, 0, 20, n = 400, lwd = 2, xlab = TeX(r"($x$)"), ylab = TeX(r"($|x - 10|$)"),
      main = TeX(r"(not differentiable at the minimum)"))
abline(v = golden(f2, c(0, 20)), col = 2, lty = 2, lwd = 2)

c(linkage = golden(f1, c(0.001, 0.999)), abs_value = golden(f2, c(0, 20)))
```

### Numerical Derivatives

When $S$ or $S'$ is tedious or error-prone to derive by hand, both can be approximated by finite differences:
$$
\ell'(\theta) \approx \frac{\ell(\theta + h) - \ell(\theta - h)}{2h}, \qquad
\ell''(\theta) \approx \frac{\ell(\theta + h) - 2\ell(\theta) + \ell(\theta - h)}{h^{2}}.
$$

The step $h$ trades off two sources of error: too large and the approximation suffers **truncation error** from the curvature of $\ell$ itself; too small and it suffers **rounding error**, because the numerator subtracts two nearly equal floating-point numbers. A step of $h \approx \sqrt{\varepsilon}\,(1 + |\theta|) \approx 10^{-8}$ balances the two for the first derivative, and $h \approx \varepsilon^{1/3}$ for the second, where $\varepsilon \approx 2 \times 10^{-16}$ is double-precision machine epsilon.

```{r}
#| label: numerical-derivative-check
h <- 1e-5
c(analytic = dS_d(lstar), numeric = (ll_d(lstar + h) - 2 * ll_d(lstar) + ll_d(lstar - h)) / h^2)
```

The two agree to several digits, which is reassuring but also the point: `nlm()` and `optim(method = "BFGS")` fall back on exactly this kind of finite differencing whenever an analytic gradient is not supplied, and `numDeriv::grad()` / `numDeriv::hessian()` implement more accurate variants (Richardson extrapolation) for when a single finite difference is not precise enough.

### Summary of the Methods

| method | needs | convergence | fails when |
|:--|:--|:--|:--|
| fixed point | a rearrangement $\theta = g(\theta)$ | linear, factor $\lvert g'\rvert$ | $\lvert g'\rvert \ge 1$ |
| bisection | $S$ and a sign-changing bracket | linear, factor $1/2$ | no bracket available |
| Newton–Raphson | $S$ and $S'$ | quadratic near the root | flat or wrong curvature |
| method of scoring | $S$ and $I(\theta)$ | linear to quadratic | $E[\cdot]$ unavailable |
| golden section | $\ell$ only, and a bracket | linear, factor $0.618$ | $\ell$ not unimodal |

### Example: Univariate Optimzation for Zero-truncated Poisson Data

#### The Model

Counts are often recorded only when positive: a study of hospital stays sees admitted patients, a survey of litters sees litters that exist. If $Y \sim \mathrm{Poisson}(\lambda)$ and we observe $X = Y \mid Y > 0$,
$$
P(x \mid \lambda) = \frac{e^{-\lambda}\lambda^{x}}{x!}\cdot\frac{1}{1 - e^{-\lambda}}, \qquad x = 1, 2, \ldots.
$$

Dropping the factorial terms, which do not involve $\lambda$,
$$
\ell(\lambda) = \left(\sum_i x_i\right)\log\lambda - n\left[\lambda + \log\left(1 - e^{-\lambda}\right)\right],
$$
$$
S(\lambda) = \frac{\sum_i x_i}{\lambda} - \frac{n}{1 - e^{-\lambda}}, \qquad
S'(\lambda) = -\frac{\sum_i x_i}{\lambda^{2}} + \frac{n e^{-\lambda}}{\left(1 - e^{-\lambda}\right)^{2}} .
$$

Since $E(X) = \lambda/(1 - e^{-\lambda})$, taking $-E[S'(\lambda)]$ gives the expected information,
$$
n I(\lambda) = \frac{n}{\lambda\left(1 - e^{-\lambda}\right)} - \frac{n e^{-\lambda}}{\left(1 - e^{-\lambda}\right)^{2}} .
$$

The score equation has no closed-form root, but the likelihood is unimodal — an ideal problem on which to compare all the methods against a single known answer.

```{r}
#| label: nzp-functions
## log likelihood, omitting factorial terms; vectorized in lambda
nzp.log.likelihood <- function(n, lambda)
  sum(n) * log(lambda) - length(n) * (lambda + log(1 - exp(-lambda)))

nzp.scorefunction <- function(n, lambda)
  sum(n) / lambda - length(n) / (1 - exp(-lambda))

nzp.dscore <- function(n, lambda)
  -sum(n) / lambda^2 + length(n) * exp(-lambda) / (1 - exp(-lambda))^2

nzp.information <- function(n, lambda)
  length(n) / (lambda * (1 - exp(-lambda))) - length(n) * exp(-lambda) / (1 - exp(-lambda))^2

## simulated data: Poisson counts with the zeros discarded
x <- rpois(100, lambda = 1.5)
xp <- x[x > 0]

c(n_observed = length(xp), sample_mean = mean(xp), true_lambda = 1.5)
```

The sample mean overshoots 1.5, because the zeros that would have pulled it down were never recorded.

```{r}
#| label: fig-nzp-loglik-score
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: "Log-likelihood and score for the zero-truncated Poisson model, from simulated data with $\\bar x$ overshooting the true $\\lambda = 1.5$."
lambdas <- seq(0.1, 8, length.out = 300)

par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
plot(lambdas, nzp.log.likelihood(xp, lambdas), type = "l", lwd = 2,
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($\ell(\lambda)$)"), main = TeX(r"(log-likelihood)"))
plot(lambdas, nzp.scorefunction(xp, lambdas), type = "l", lwd = 2,
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($S(\lambda)$)"), main = TeX(r"(score)"))
abline(h = 0, lty = 3)
```

#### The Three Iterations

Each method below returns the whole sequence of iterates together with the log-likelihood attained, so the trajectories can be compared directly.

**Simple iteration** rearranges $S(\lambda) = 0$ as
$$
\frac{\overline x}{\lambda} = \frac{1}{1 - e^{-\lambda}} \iff \lambda = \overline x\left(1 - e^{-\lambda}\right) = g(\lambda),
$$
with contraction factor $g'(\lambda) = \overline x e^{-\lambda}$.

```{r}
#| label: nzp-iterations
nzp.simple.iteration <- function(n, r, lambda0 = mean(n)) {
  mean.n <- mean(n); lambda <- numeric(r + 1); lambda[1] <- lambda0
  for (i in 1:r) lambda[i + 1] <- mean.n * (1 - exp(-lambda[i]))
  data.frame(lambda = lambda, log.lik = nzp.log.likelihood(n, lambda))
}

nzp.newton.raphson <- function(n, r, lambda0 = mean(n)) {
  mean.n <- mean(n); lambda <- numeric(r + 1); lambda[1] <- lambda0
  for (i in 1:r) {
    e <- exp(-lambda[i])
    lambda[i + 1] <- lambda[i] -
      (mean.n / lambda[i] - 1 / (1 - e)) / (e / (1 - e)^2 - mean.n / lambda[i]^2)
  }
  data.frame(lambda = lambda, log.lik = nzp.log.likelihood(n, lambda))
}

nzp.method.of.scoring <- function(n, r, lambda0 = mean(n)) {
  mean.n <- mean(n); lambda <- numeric(r + 1); lambda[1] <- lambda0
  for (i in 1:r) {
    e <- exp(-lambda[i])
    lambda[i + 1] <- lambda[i] -
      (mean.n / lambda[i] - 1 / (1 - e)) / ((e / (1 - e) - 1 / lambda[i]) / (1 - e))
  }
  data.frame(lambda = lambda, log.lik = nzp.log.likelihood(n, lambda))
}
```

#### Comparison From Several Starting Values

The three methods are run for 15 iterations from a sensible start, a start far below, and a start far above.

```{r}
#| label: nzp-compare-starts
starts <- c(mean(xp), 0.1, 100)

trace_tab <- function(fun, r = 15)
  vapply(starts, function(s) fun(xp, r, s)$lambda, numeric(r + 1))

paths <- list(simple = trace_tab(nzp.simple.iteration),
              newton = trace_tab(nzp.newton.raphson),
              scoring = trace_tab(nzp.method.of.scoring))

### first six iterates from each starting value
for (m in names(paths)) {
  cat("\n", m, "\n")
  tab <- paths[[m]][1:6, ]
  dimnames(tab) <- list(paste("iter", 0:5), sprintf("start = %.4g", starts))
  print(signif(tab, 6))
}
```

The pattern is the one the theory predicts. Newton–Raphson takes the fewest iterations from a good start, and its digits lock in from the left as the error is squared each step. Simple iteration converges from every start, including the extreme ones, but slowly and at a steady rate. The method of scoring is intermediate: always ascending, and nearly as fast as Newton once close.

```{r}
#| label: fig-nzp-convergence-plot
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Error decay (log scale) of simple iteration, Newton-Raphson, and the method of scoring for the zero-truncated Poisson MLE: Newton's error curve bends downward, reflecting quadratic convergence."
lhat <- nzp.newton.raphson(xp, 30, mean(xp))$lambda[31]

err <- sapply(paths, function(p) abs(p[, 1] - lhat))   # from the mean(xp) start

par(mar = c(4, 4.5, 2.5, 1))
matplot(0:15, pmax(err, 1e-17), log = "y", type = "b", pch = 19, lty = 1, col = c(4, 1, 2),
        xlab = TeX(r"(iteration $k$)"), ylab = TeX(r"($|\lambda_k - \hat{\lambda}|$)"),
        main = TeX(r"(error decay from $\lambda_0 = \bar{x}$)"))
legend("bottomleft", bty = "n", pch = 19, lty = 1, col = c(4, 1, 2),
       legend = c("simple iteration", "Newton-Raphson", "method of scoring"))
```

On a log scale a linear method is a straight line; Newton's curve bends downward until it reaches the floor set by machine precision.

```{r}
#| label: nzp-builtin-check
S_xp  <- function(l) nzp.scorefunction(xp, l)
dS_xp <- function(l) nzp.dscore(xp, l)
ll_xp <- function(l) nzp.log.likelihood(xp, l)

c(newton      = unname(newton(S_xp, dS_xp, mean(xp))["root"]),
  fixed_point = unname(fixed_point(function(l) mean(xp) * (1 - exp(-l)), mean(xp))["root"]),
  bisection   = unname(bisection(S_xp, 0.1, 20)["root"]),
  golden      = golden(function(l) -ll_xp(l), c(0.1, 20), eps = 1e-8),
  uniroot     = uniroot(S_xp, c(0.1, 20), tol = 1e-10)$root,
  optimize    = optimize(function(l) -ll_xp(l), c(0.1, 20), tol = 1e-10)$minimum)
```

All six agree, which is the point of running them together: the methods differ in cost and robustness, not in the answer.

#### Standard Error From the Curvature

The optimizer has already produced everything inference needs.

```{r}
#| label: nzp-standard-error
se_obs <- 1 / sqrt(-nzp.dscore(xp, lhat))
se_exp <- 1 / sqrt(nzp.information(xp, lhat))

c(mle = lhat, se_observed = se_obs, se_expected = se_exp)
lhat + c(-1.96, 1.96) * se_obs                     # 95% Wald interval
```

```{r}
#| label: fig-nzp-quadratic-approx
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Zero-truncated Poisson log-likelihood versus its quadratic approximation at $\\hat\\lambda$, with the resulting Wald 95% interval marked."
lam <- seq(lhat - 4 * se_obs, lhat + 4 * se_obs, length.out = 300)
quad <- -0.5 * (-nzp.dscore(xp, lhat)) * (lam - lhat)^2

par(mar = c(4, 4.5, 2.5, 1))
plot(lam, ll_xp(lam) - ll_xp(lhat), type = "l", lwd = 2,
     xlab = TeX(r"($\lambda$)"), ylab = TeX(r"($\ell(\lambda) - \ell(\hat{\lambda})$)"),
     main = TeX(r"(log-likelihood and its quadratic approximation at $\hat{\lambda}$)"))
lines(lam, quad, col = 2, lwd = 2, lty = 2)
abline(h = -1.92, col = 4, lty = 3)
abline(v = lhat + c(-1.96, 1.96) * se_obs, col = 4, lty = 3)
legend("bottom", bty = "n", lwd = 2, lty = c(1, 2), col = c(1, 2),
       legend = TeX(c(r"($\ell(\lambda)$)", r"(quadratic from $J(\hat{\lambda})$)")))
```

The parabola implied by the observed information tracks the log-likelihood closely over the width of the interval, so the Wald interval is trustworthy here.

#### Evaluating the Naive Estimator

All of this machinery is worth the effort only if the MLE beats the easy alternative. The naive estimator ignores the truncation and reports $\overline x$. Simulating many truncated samples gives the sampling distributions of both estimators.

```{r}
#| label: fig-nzp-mle-vs-naive
#| cache: true
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: "Sampling distributions of the naive estimator $\\bar x$ versus the MLE for a zero-truncated Poisson mean, $\\lambda_{true} = 1.5$: the naive estimator is badly biased upward, the MLE is centred on the truth."
mle_nzp <- function(n)
  uniroot(function(l) nzp.scorefunction(n, l), c(1e-6, 50), tol = 1e-10)$root

sim_nzp <- function(lambda, n_gen = 100, B = 500)
  vapply(1:B, function(b) {
    y <- rpois(n_gen, lambda); n <- y[y > 0]
    c(naive = mean(n), mle = mle_nzp(n))
  }, c(naive = 0, mle = 0))

for (lambda_true in c(0.5, 1.5)) {
  est <- sim_nzp(lambda_true)
  tab <- t(apply(est, 1, function(e)
    c(bias = mean(e) - lambda_true, sd = sd(e), rmse = sqrt(mean((e - lambda_true)^2)))))
  cat("\nlambda =", lambda_true, "\n"); print(round(tab, 4))
}

est <- sim_nzp(1.5)
par(mfrow = c(1, 2), mar = c(4, 4.5, 2.5, 1))
for (m in c("naive", "mle")) {
  hist(est[m, ], breaks = 25, col = "grey90",
       xlim = range(est), xlab = TeX(r"($\hat{\lambda}$)"),
       main = TeX(sprintf(r"(%s, bias $= %.3f$)", m, mean(est[m, ]) - 1.5)))
  abline(v = 1.5, col = 2, lwd = 2, lty = 2)
}
```

The naive estimator is biased upward, and severely so when $\lambda$ is small, since that is when most of the probability mass sits on the discarded zero. Its bias does not shrink with sample size: it reflects the wrong model, not too little data. The MLE is centred on the truth with a standard deviation matching the information-based standard error of the previous section. The optimization was the means; the correct estimator was the end.

## Multivariate Newton-Raphson Optimzation Methods


### Gradient and Hessian: A Geometric Illustration

The two objects entering the multivariate Newton step have a direct geometric reading. The score $S(\theta)$ is the direction of steepest ascent and is perpendicular to the contour of $\ell$ through $\theta$; the Hessian $H(\theta)$ describes how that direction curves, and at a strict local maximum $J(\theta) = -H(\theta)$ is positive definite. Near $\hat\theta$, $\ell$ is well approximated by the quadratic
$$
\ell(\theta) \approx \ell(\hat\theta) - \tfrac12(\theta - \hat\theta)^\top J(\hat\theta)(\theta - \hat\theta),
$$
whose contours are ellipses centred at $\hat\theta$. Writing the eigendecomposition $J(\hat\theta) = Q\Lambda Q^\top$, the axes of these ellipses point along the eigenvectors in $Q$, and the half-length along axis $j$ is proportional to $1/\sqrt{\lambda_j}$: a large eigenvalue means sharp curvature and a short, well-determined direction, while $\lambda_{\max}/\lambda_{\min}$ measures how elongated — and, once the axes are tilted, how correlated the parameters are.

```{r}
#| label: fig-hessian-ellipse
#| echo: false
#| fig-width: 7
#| fig-height: 7
#| fig-cap: "Contours of a quadratic log-likelihood: gradient arrows (green) point uphill perpendicular to the contours but not toward the maximum unless the ellipses are circular; the dashed axes are the eigenvectors of $J(\\hat\\theta)$, each with half-length $1/\\sqrt{\\lambda_j}$."
A <- matrix(c(2, 1.2, 1.2, 1), 2)
s <- seq(-2, 2, length.out = 80); g <- expand.grid(x = s, y = s)
z <- matrix(apply(g, 1, function(v) -0.5 * drop(t(v) %*% A %*% v)), 80)
par(pty = "s", mar = c(4, 4, 1, 1))
contour(s, s, z, nlevels = 12, xlab = expression(theta[1]), ylab = expression(theta[2]))
## gradient arrows
pts <- rbind(c(1, 1), c(-1.5, 0.5), c(0.5, -1.5))
for (i in 1:3) {
  gr <- -A %*% pts[i, ]
  arrows(pts[i, 1], pts[i, 2], pts[i, 1] + 0.3 * gr[1], pts[i, 2] + 0.3 * gr[2],
         col = 3, lwd = 2, length = 0.1)
}
## principal axes with eigenvalue labels: length 1/sqrt(lambda) on each side
e <- eigen(A); k <- 1.6
for (j in 1:2) {
  v <- e$vectors[, j] * k / sqrt(e$values[j])
  segments(-v[1], -v[2], v[1], v[2], col = "firebrick", lwd = 2, lty = 2)
  text(v[1], v[2], bquote(lambda[.(j)] == .(round(e$values[j], 2))),
       col = "firebrick", pos = if (j == 1) 4 else 2, cex = 0.9)
}
points(0, 0, pch = 19)
```

Every gradient arrow in the figure is perpendicular to the contour it starts on and points generally uphill toward the maximum at the origin, but only an arrow started exactly on a principal axis points straight at it. This is why Newton's step, which multiplies the score by $J(\hat\theta)^{-1}$ rather than simply following the gradient, converges directly to $\hat\theta$: it rescales and rotates the ascent direction to undo exactly this distortion. The plain gradient-ascent methods of a later section have no such correction, and zig-zag across elongated contours like these.

### Naive Newton-Raphson Iterations

In the previous section the parameter was a scalar, and the log-likelihood could
be inspected by plotting it. Once $\theta \in \mathbb{R}^p$ neither is true: we
cannot see the surface, and the notion of "bracketing a root" that underlies
bisection and golden section has no direct multivariate analogue. What survives
is the local quadratic approximation, and the Newton–Raphson algorithm is its
direct consequence.

This section develops the multivariate Newton–Raphson method and its
statistical variant, Fisher scoring, applies both to logistic regression,
examines interactively how and why the iteration can diverge, and closes with a
survey of the derivative-free and gradient-only alternatives.

Write the log-likelihood as $\ell(\theta)$ for $\theta \in \mathbb{R}^p$, and
define.
$$
S(\theta) = \frac{\partial \ell(\theta)}{\partial \theta}
\quad (p \times 1, \text{ the score}),
\qquad
H(\theta) = \frac{\partial^2 \ell(\theta)}{\partial \theta \, \partial \theta^{\top}}
\quad (p \times p, \text{ the Hessian}).
$$

The negative Hessian $J(\theta) = -H(\theta)$ is the **observed information
matrix**.

Expand $\ell$ to second order about the current iterate $\theta^{(k)}$:
$$
\ell(\theta) \;\approx\; \ell(\theta^{(k)})

 + S(\theta^{(k)})^{\top}(\theta - \theta^{(k)})
 - \tfrac{1}{2}\,(\theta - \theta^{(k)})^{\top} J(\theta^{(k)}) \,(\theta - \theta^{(k)}).
$$

This is a quadratic in $\theta$. If $J(\theta^{(k)})$ is positive definite the
quadratic has a unique maximum, found by setting its derivative to zero:
$$
S(\theta^{(k)}) - J(\theta^{(k)})(\theta - \theta^{(k)}) = 0 .
$$

Solving for $\theta$ gives the **Newton–Raphson update**:
$$
\boxed{\;\theta^{(k+1)} = \theta^{(k)} + J(\theta^{(k)})^{-1} S(\theta^{(k)}) \;}.
$$

The algorithm therefore replaces the log-likelihood by the quadratic that
matches its value, slope and curvature at the current point, and jumps to the
maximum of that quadratic. Everything that follows — the fast convergence, and
the modes of failure — comes from how good or bad that replacement is.

Three remarks are worth making explicit.

*The inverse is never formed.* Write the update as
$\theta^{(k+1)} = \theta^{(k)} + \delta^{(k)}$ where $\delta^{(k)}$ solves the
linear system $J(\theta^{(k)})\,\delta = S(\theta^{(k)})$. In R this is
`solve(J, S)`, not `solve(J) %*% S`. The former is faster and numerically more
stable; the latter computes $p^2$ numbers in order to use $p$ of them.

*The step is not the gradient direction.* $\delta^{(k)} = J^{-1}S$ points along
the gradient only when $J$ is a multiple of the identity. The information matrix
rescales and rotates the ascent direction to account for differing curvature and
correlation among the parameters, which is exactly why Newton–Raphson is
insensitive to the scaling of the parameters while gradient descent is not.

*Positive definiteness is an assumption, not a guarantee.* Away from the maximum
$J(\theta^{(k)})$ can be indefinite or singular, in which case the "maximum of
the quadratic" is a saddle point or does not exist, and $\delta^{(k)}$ may point
downhill. @sec-divergence examines this.

#### Fisher scoring

The observed information $J(\theta)$ depends on the data. Its expectation under
the model,
$$
I(\theta) = E_{\theta}\{J(\theta)\} = -E_{\theta}\left\{ \frac{\partial^2 \ell(\theta)}{\partial \theta \, \partial \theta^{\top}} \right\},
$$

is the **expected (Fisher) information matrix**. Replacing $J$ by $I$ in the
update gives the **method of scoring**:
$$
\theta^{(k+1)} = \theta^{(k)} + I(\theta^{(k)})^{-1} S(\theta^{(k)}).
$$

The two algorithms have different iterates but the same fixed point, since both
stop where $S(\theta) = 0$. The practical differences:

| | Newton–Raphson | Fisher scoring |
|:---|:---|:---|
| Curvature matrix | Observed information $J(\theta)$ | Expected information $I(\theta)$ |
| Derivation | Second derivative of $\ell$ | Expectation of that second derivative |
| Positive definite? | Only guaranteed near the maximum | Always, wherever the model is identified |
| Algebra | Can be messy | Often simpler, as data-dependent terms average out |
| Local convergence | Quadratic | Linear, but typically with a very small rate constant |
| Stability from poor starts | Less stable | More stable |

: Newton–Raphson compared with Fisher scoring {#tbl-nr-scoring}

The guaranteed positive definiteness of $I(\theta)$ is the main practical
argument for scoring: the update direction is always an ascent direction, so the
iteration cannot be sent uphill-seeking in a downhill direction by an
ill-behaved Hessian.

For generalised linear models with the **canonical link** the two coincide
exactly, because the second derivative of the log-likelihood does not involve
the response. Logistic regression uses the canonical logit link, so the example
below is simultaneously Newton–Raphson, Fisher scoring, and — as we note at the
end of the section — iteratively reweighted least squares.


### Convergence properties

Near a maximum $\hat\theta$ at which $J(\hat\theta)$ is positive definite,
Newton–Raphson converges **quadratically**: the error satisfies
$$\|\theta^{(k+1)} - \hat\theta\| \;\le\; C \,\|\theta^{(k)} - \hat\theta\|^{2},$$

for some constant $C$ depending on the third derivatives. Informally, the number
of correct digits doubles at each iteration, which is why five or six iterations
are usually enough and why a well-started run appears to converge instantly.

The qualification "near" is essential. The theorem is local: it guarantees
nothing about a start far from $\hat\theta$, and nothing at all if $J$ fails to
be positive definite along the way.

### Standard errors as a by-product

At convergence the same matrix that drove the iteration supplies the asymptotic
variance. Under the usual regularity conditions,
$$
\hat\theta \;\dot\sim\; N\!\left(\theta, \; J(\hat\theta)^{-1}\right),
$$

so $\operatorname{\mathrm{SE}}(\hat\theta_j) = \sqrt{[J(\hat\theta)^{-1}]_{jj}}$. This is
a genuine practical advantage of Newton-type methods over derivative-free ones:
the curvature information needed for inference has already been computed. Methods
that never form a Hessian require a separate numerical differentiation step, which
is what `hessian = TRUE` requests from `nlm` and `optim`.

### Example: Multivariate Newton-Raphson for Logistic Regression

#### The model and its derivatives

With a single covariate $z_i$ and binary response $r_i$, let
$u_i = \beta_0 + \beta_1 z_i$ and $p_i = e^{u_i}/(1 + e^{u_i})$. The
log-likelihood is
$$
\ell(\beta) = \sum_{i=1}^{n}\left\{ r_i u_i - \log\left(1 + e^{u_i}\right) \right\}.
$$

Writing $X$ for the $n \times 2$ design matrix with rows $(1, z_i)$, the score
and information are.
$$
S(\beta) = X^{\top}(r - p), \qquad
J(\beta) = X^{\top} W X, \qquad W = \operatorname{diag}\{p_i(1 - p_i)\}.
$$

Note that $J$ does not involve $r$. This is the canonical-link property
mentioned above, and it means $J(\beta) = I(\beta)$ identically.

Two numerical points, both carried over from the chapter on computer arithmetic.
First, $\log(1 + e^{u})$ overflows for $u \gtrsim 710$; the stable form is
$\log(1+e^u) = \max(u,0) + \log(1 + e^{-|u|})$. Second, $p_i(1-p_i) \to 0$ when
$|u_i|$ is large, so $W$ — and hence $J$ — degenerates toward singularity when
the fitted probabilities are extreme. That observation is the whole explanation
of the divergence we are about to see.

#### Implementation

```{r}
#| label: logistic-fns
#| message: false

### numerically stable log(1 + exp(u))
log1pexp <- function(u) pmax(u, 0) + log1p(exp(-abs(u)))

### negative log-likelihood, negative score, and information matrix
lik_score_info <- function(b, r, z) {
    u <- b[1] + b[2] * z
    p <- plogis(u)
    w <- p * (1 - p)
    list(
        neg_loglik = -sum(u * r - log1pexp(u)),
        neg_score  = -c(sum(r - p), sum(z * (r - p))),
        inf        = matrix(c(sum(w), sum(z * w),
                              sum(z * w), sum(z * z * w)), 2, 2)
    )
}

neg_logp_logistic <- function(b, z, r) {
    u <- b[1] + b[2] * z
    -sum(u * r - log1pexp(u))
}
```

Because we minimise the *negative* log-likelihood, the update
$\beta^{(k+1)} = \beta^{(k)} + J^{-1}S$ becomes
$\beta^{(k+1)} = \beta^{(k)} - J^{-1} \times (\text{negative score})$, which is
the form coded below.

```{r}
#| label: logistic-nr

mle_logistic_nr <- function(b0, no_iter, r, z, debug = FALSE) {
    out <- matrix(NA_real_, no_iter + 1, 3,
                  dimnames = list(NULL, c("beta0", "beta1", "neg_loglike")))
    b <- b0
    for (i in seq_len(no_iter + 1)) {
        q <- lik_score_info(b, r, z)
        out[i, ] <- c(b, q$neg_loglik)
        if (debug) print(out[i, ])
        if (i > no_iter) break
        step <- tryCatch(solve(q$inf, q$neg_score),
                         error = function(e) rep(NA_real_, 2))
        if (any(!is.finite(step))) break
        b <- b - step
    }
    out
}
```

```{r}
#| label: fig-logistic-data
#| fig-cap: "Simulated logistic data: true success probability and observed responses."

gen_logistic_data <- function(b, n) {
    z <- sort(runif(n, -2, 2))
    p <- plogis(b[1] + b[2] * z)
    r <- rbinom(n, 1, p)
    plot(z, p, type = "l", ylim = c(0, 1), ylab = "P(r = 1)")
    points(z, r, col = r + 1)
    list(z = z, r = r)
}

set.seed(1)
data <- gen_logistic_data(c(0, 1.5), 200)
```

A start close to the truth converges in a handful of iterations:

```{r}
#| label: logistic-nr-good
round(mle_logistic_nr(c(0, 3), 8, data$r, data$z), 6)
```

The doubling of correct digits predicted by the quadratic-convergence result is
visible in the final rows.

#### Comparison with `nlm` and `glm`

```{r}
#| label: logistic-nlm
logit_nlm <- nlm(neg_logp_logistic, c(0, 0),
                 z = data$z, r = data$r, hessian = TRUE)
logit_nlm$estimate
sqrt(diag(solve(logit_nlm$hessian)))   # asymptotic standard errors
```

```{r}
#| label: logistic-glm
summary(glm(r ~ z, family = binomial(), data = data))$coefficients
```

The three routes agree to printing accuracy. `glm` is doing the same arithmetic
as `mle_logistic_nr`: for a GLM the Newton–Raphson step can be rearranged as a
weighted least-squares fit of the working response
$u_i + (r_i - p_i)/\{p_i(1-p_i)\}$ on $X$ with weights $p_i(1-p_i)$, giving
**iteratively reweighted least squares (IRLS)**. This is not a different
algorithm but a different arrangement of the same one, and it explains why
`glm` reports "Fisher Scoring iterations" in its output.

### Divergence {#sec-divergence}

#### What goes wrong {#sec-nr-divergence-app}

The quadratic approximation is trustworthy only in a neighbourhood of the
current iterate. Starting from a large $|\beta_1|$, nearly all fitted
probabilities are pushed to $0$ or $1$, so the weights $p_i(1-p_i)$ are
essentially zero, $J$ is nearly singular, and $J^{-1}$ is enormous. The
resulting step is huge, the new iterate lands far outside the region where the
quadratic was a reasonable fit, and the process can oscillate or run off to
infinity — even though the logistic log-likelihood is globally concave and has a
unique maximum.

Press the play button on the iteration slider to animate the
path over the contours of the negative log-likelihood. This app runs live only
in the HTML edition of this book; readers of other formats can follow the link
below it.

::::: {.content-visible when-format="html"}
::: {#fig-nr-divergence-app}

```{shinylive-r}
#| label: nr-divergence-app
#| standalone: true
#| viewerHeight: 800

library(shiny)

### ---------- fixed data ----------------------------------------------------
set.seed(1)
n <- 200
z <- sort(runif(n, -2, 2))
r <- rbinom(n, 1, plogis(0 + 1.5 * z))

log1pexp <- function(u) pmax(u, 0) + log1p(exp(-abs(u)))
negll <- function(b) {
    u <- b[1] + b[2] * z
    -sum(u * r - log1pexp(u))
}

### ---------- contour grid (computed once) ----------------------------------
B0 <- seq(-3, 3, by = 0.1)
B1 <- seq(-8, 8, by = 0.2)
Z <- outer(B0, B1, Vectorize(function(a, b) negll(c(a, b))))

### ---------- Newton-Raphson path -------------------------------------------
nr_path <- function(b0, maxit = 15, halving = FALSE) {
    P <- matrix(NA_real_, maxit + 1, 3,
                dimnames = list(NULL, c("beta0", "beta1", "negll")))
    b <- b0
    P[1, ] <- c(b, negll(b))
    for (i in seq_len(maxit)) {
        u <- b[1] + b[2] * z
        p <- plogis(u)
        w <- p * (1 - p)
        g <- -c(sum(r - p), sum(z * (r - p)))
        J <- matrix(c(sum(w), sum(z * w), sum(z * w), sum(z * z * w)), 2, 2)
        step <- tryCatch(solve(J, g), error = function(e) rep(NA_real_, 2))
        if (any(!is.finite(step))) break
        if (halving) {
            f0 <- P[i, 3]; s <- 1
            repeat {
                bn <- b - s * step
                fn <- negll(bn)
                if ((is.finite(fn) && fn <= f0) || s < 1e-6) break
                s <- s / 2
            }
            b <- bn
        } else {
            b <- b - step
        }
        if (any(!is.finite(b))) break
        P[i + 1, ] <- c(b, negll(b))
    }
    P
}

### ---------- UI -------------------------------------------------------------
ui <- fluidPage(
    titlePanel(NULL),
    sidebarLayout(
        sidebarPanel(
            width = 4,
            numericInput("b0", HTML("starting &beta;<sub>0</sub>"),
                         value = 0, step = 0.5),
            numericInput("b1", HTML("starting &beta;<sub>1</sub>"),
                         value = 5, step = 0.5),
            checkboxInput("halve", "apply step halving", FALSE),
            sliderInput("iter", "iteration", min = 0, max = 15, value = 0,
                        step = 1, animate = animationOptions(interval = 900)),
            helpText("Try (0, 3) and then (0, 5). The second start makes the ",
                     "information matrix nearly singular, so the first step ",
                     "overshoots badly."),
            tableOutput("tab")
        ),
        mainPanel(
            width = 8,
            plotOutput("contour", height = "380px"),
            plotOutput("trace", height = "200px")
        )
    )
)

### ---------- server ---------------------------------------------------------
server <- function(input, output, session) {

    path <- reactive({
        nr_path(c(input$b0, input$b1), maxit = 15, halving = input$halve)
    })

    shown <- reactive({
        P <- path()
        k <- min(input$iter + 1, nrow(P))
        P[seq_len(k), , drop = FALSE]
    })

    output$contour <- renderPlot({
        P <- shown()
        par(mar = c(4, 4, 2, 1))
        contour(B0, B1, Z, nlevels = 40, drawlabels = FALSE,
                col = "grey70",
                xlab = expression(beta[0]), ylab = expression(beta[1]),
                main = "negative log-likelihood surface and NR path")
        ok <- complete.cases(P)
        if (sum(ok) > 1) {
            lines(P[ok, 1], P[ok, 2], col = "steelblue", lwd = 2)
            points(P[ok, 1], P[ok, 2], pch = 21, bg = "steelblue", cex = 1.1)
        }
        last <- P[max(which(complete.cases(P))), , drop = FALSE]
        points(last[1], last[2], pch = 21, bg = "firebrick", cex = 1.8)
        outside <- last[1] < min(B0) || last[1] > max(B0) ||
                   last[2] < min(B1) || last[2] > max(B1)
        if (isTRUE(outside)) {
            mtext(sprintf("current iterate (%.1f, %.1f) is off the plotted region",
                          last[1], last[2]), side = 3, line = -1.5,
                  col = "firebrick", cex = 0.9)
        }
        points(0, 1.5, pch = 3, col = "darkgreen", lwd = 2)
    })

    output$trace <- renderPlot({
        P <- path()
        k <- min(input$iter + 1, nrow(P))
        par(mar = c(4, 4, 1, 1))
        yl <- range(P[, 3], na.rm = TRUE)
        plot(0:(nrow(P) - 1), P[, 3], type = "n", ylim = yl,
             xlab = "iteration", ylab = "negative log-likelihood")
        lines(0:(k - 1), P[seq_len(k), 3], type = "b", pch = 19,
              col = "steelblue")
    })

    output$tab <- renderTable({
        P <- shown()
        d <- as.data.frame(P)
        d$iteration <- seq_len(nrow(d)) - 1
        d <- d[, c("iteration", "beta0", "beta1", "negll")]
        tail(d, 5)
    }, digits = 4, rownames = FALSE)
}

shinyApp(ui, server)
```

**Shinylive App Illustrating Newton–Raphson Divergence.** The interactive figure below lets you choose a starting value and step through
the iterations.

:::
:::::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit05_MLE.html#sec-nr-divergence-app>.

:::

::: {.callout-note}

#### Reproducing this figure
The chunk above uses the [shinylive](https://github.com/quarto-ext/shinylive)
extension, which compiles the app to WebAssembly so that it runs in the reader's
browser with no Shiny server. Install it once per project with

```bash
quarto add quarto-ext/shinylive
```

and add `filters: [shinylive]` to the document or project YAML. Only packages
available in webR may be used inside the app, so the code above is restricted to
base R and `shiny`.

:::

#### Enhancements to Multivariate Newton-Raphson Methods

The divergence is not a defect of the Newton direction but of the Newton *step
length*. The standard remedies all keep the direction and control the distance.

**Step halving (backtracking).** Take $\delta^{(k)}$ as usual but move
$\theta^{(k+1)} = \theta^{(k)} + s\,\delta^{(k)}$, starting from $s = 1$ and
halving until the log-likelihood actually increases. This costs a few extra
function evaluations, never damages quadratic convergence near the maximum
(where $s = 1$ is accepted on the first try), and rescues most divergent runs.
Toggle it in the app above to see the effect.

**Ridging (Levenberg–Marquardt).** Replace $J$ by $J + \lambda \mathbb{I}$ with
$\lambda > 0$. Large $\lambda$ shrinks the step toward a small gradient step;
$\lambda \to 0$ recovers Newton. This also repairs indefiniteness, and is the
approach `nlm` takes internally.

**Trust regions.** Maximise the quadratic approximation subject to
$\|\theta - \theta^{(k)}\| \le \Delta_k$, adapting $\Delta_k$ according to how
well the quadratic predicted the realised change. This is `nlminb` and the
`PORT` routines.

**Reparameterisation and good starts.** Often the cheapest fix. Centring and
scaling covariates improves conditioning of $J$ substantially, and for GLMs
the null-model fit is a reliable starting point — which is what `glm` uses when
`start` is not supplied.

Two more precise versions of these ideas are worth stating explicitly, since together they make up what is sometimes called **safeguarded Newton** — not "Newton–Raphson" in the textbook sense, and not what `nlm()` runs internally, which globalises with a trust region instead.

**A principled choice of ridge.** Write $\delta_k = J_k^{-1}S_k$ for the raw Newton step at $\theta_k$ (with $J_k = J(\theta_k)$, $S_k = S(\theta_k)$). Two things can go wrong once $\theta_k$ is far from $\hat\theta$: the direction may not ascend at all, since $S_k^\top \delta_k = S_k^\top J_k^{-1}S_k > 0$ requires $J_k \succ 0$, which need not hold where $\ell$ is not concave; and even when $J_k \succ 0$, the full step can overshoot, since the quadratic model is trustworthy only near $\theta_k$. Ridging replaces $J_k$ by
$$
\tilde J_k = J_k + \tau I, \qquad \tau = \max\!\big(0,\ \delta - \lambda_{\min}(J_k)\big), \qquad \delta = 10^{-3}\,\lambda_{\max}(J_k),
$$
which shifts every eigenvalue of $J_k$ by $\tau$ while leaving its eigenvectors unchanged. As $\tau \to \infty$ the resulting direction $\tilde J_k^{-1}S_k$ rotates toward $S_k$ itself — plain gradient ascent — so ridging interpolates continuously between Newton and steepest ascent. An alternative, **spectral modification**, decomposes $J_k = V\Lambda V^\top$ and replaces each eigenvalue $\lambda_i$ by $\max(|\lambda_i|, \delta)$: this reverses negative-curvature directions rather than merely damping them, pushing the step away from a saddle instead of just slowing it down. Wherever $J_k \succ 0$ already, both reduce to $\tau = 0$ and $\tilde J_k = J_k$, so the safeguard is inactive exactly where it is not needed.

**A precise stopping rule for step halving: the Armijo condition.** Accepting the halved step $\theta_k + s\,\delta_k$ the first time the log-likelihood merely fails to decrease is workable but crude. The standard criterion instead requires a **sufficient** increase,
$$
\ell(\theta_k + s\,\delta_k) \ \ge\ \ell(\theta_k) + c_1\, s\, S_k^\top \delta_k, \qquad c_1 = 10^{-4},
$$
tested at $s = 1, \rho, \rho^2, \ldots$ for a fixed $\rho \in (0,1)$ (e.g. $\rho = \tfrac12$) until it holds. Because $c_1$ is tiny, $s = 1$ passes immediately whenever the ordinary Newton step already improves $\ell$, so quadratic convergence near $\hat\theta$ is untouched. A companion **curvature (Wolfe) condition**,
$$
S(\theta_k + s\,\delta_k)^\top \delta_k \ \le\ c_2\, S_k^\top \delta_k, \qquad c_2 = 0.9,
$$
additionally rules out steps that are cut *too* short — something plain backtracking, which only ever shrinks $s$, cannot enforce on its own. This second condition matters less for Newton itself than for the gradient-only methods of the next section, where it is exactly what keeps the BFGS update well defined.

::: {.callout-note appearance="minimal"}
**The safeguarded step, in full.** At $\theta_k$: compute $S_k$ and $J_k$; form $\tilde J_k \succ 0$ by ridging or spectral modification; solve $\tilde J_k\,\delta_k = S_k$; choose $s_k$ from $s_0 = 1$ by backtracking against the Armijo condition; set $\theta_{k+1} = \theta_k + s_k\,\delta_k$.

:::

Both safeguards switch themselves off automatically near a well-behaved maximum, and stay active only where they are needed:

| | near the maximum | far away / indefinite |
|:--|:--|:--|
| ridge $\tau$ | $0$ | $> 0$, direction rotates toward $S_k$ |
| step $s_k$ | $1$ | $< 1$ |

so quadratic convergence is retained locally while sufficient increase at every iteration guarantees progress from remote starting points — the combination that step halving, ridging, trust regions, and good starting values each approximate in the plainer forms described above.

## Gradient-Based Methods

Newton–Raphson is the method of choice when the second derivatives are available
and $p$ is modest. Outside that regime, three considerations push toward
alternatives: derivatives may be unavailable or painful to code; the $O(p^3)$
cost of solving the Newton system and the $O(p^2)$ storage of $J$ become
prohibitive for large $p$; and the objective may be non-smooth, noisy or
multimodal.

Throughout this section it is more convenient to **minimise** $f(\theta) = -\ell(\theta)$ rather than maximise $\ell$ directly — the minimum of $f$ is exactly the maximum of $\ell$, and this is also the convention that R's `optim()`, `nlm()`, and `nlminb()` use internally, so the code below matches what those functions do under the hood.

### Classification by derivative information

Methods are most usefully organised by what they require of the objective.

| Order | Requires | Methods | Typical use |
|---:|:---|:---|:---|
| 0 | Function values only | Nelder–Mead, simulated annealing, genetic algorithms | Non-smooth, discontinuous, or awkward-to-differentiate objectives; small $p$ |
| 1 | Gradient | Steepest descent, conjugate gradient, quasi-Newton (BFGS, L-BFGS), stochastic gradient descent | The workhorse class; large $p$; gradients cheap by automatic differentiation |
| 2 | Gradient and Hessian | Newton–Raphson, Fisher scoring, IRLS, trust-region Newton | Statistical models of modest dimension where the Hessian is also wanted for inference |

: Optimization methods classified by the derivative information required {#tbl-orders}

### Steepest Descent with Line Search

Move along $-\nabla f$, the direction of steepest descent, choosing the distance by a one-dimensional search:
$$
\theta_{k+1} = \theta_k + \alpha_k d_k, \qquad d_k = -\nabla f(\theta_k), \qquad \alpha_k = \arg\min_{\alpha > 0} f(\theta_k + \alpha d_k) .
$$

The step length $\alpha_k$ can come from an **exact** line search — a univariate minimization by golden section or `optimize()` along the ray — or an **inexact** one: start at $\alpha = 1$ and halve until the Armijo sufficient-decrease condition $f(\theta_k + \alpha d_k) \le f(\theta_k) + c\,\alpha\,\nabla f(\theta_k)'d_k$ holds. This is conceptually the simplest gradient method, and the one that best illustrates the role of a line search, but its convergence is only linear, and the rate deteriorates with the condition number of the Hessian: on an elongated valley the iterates zig-zag across it rather than travelling along it, as the next figure shows. It is rarely used unmodified.

```{r}
#| label: grad-descent-fn
grad_descent <- function(f, gr, x0, maxit = 1000, tol = 1e-8) {
  path <- x0
  for (k in 1:maxit) {
    d <- -gr(x0); a <- optimize(function(a) f(x0 + a * d), c(0, 10))$minimum   # exact line search
    x1 <- x0 + a * d; path <- rbind(path, x1)
    if (sqrt(sum((x1 - x0)^2)) < tol) break
    x0 <- x1 }
  list(par = x1, iter = k, path = path)
}
```

### Why Gradient Descent Zig-Zags

With an exact line search, consecutive directions are **orthogonal** — $\nabla f(\theta_{k+1})'d_k = 0$, since the search stops exactly where the directional derivative along $d_k$ vanishes. On the elongated contours of the quadratic from @fig-hessian-ellipse, that orthogonality forces the path to zig-zag back and forth across the valley instead of running down its length:

```{r}
#| label: fig-gd-zigzag
#| echo: false
#| fig-cap: "Gradient descent with an exact line search zig-zags across an elongated valley: each step is orthogonal to the next."
fq <- function(v) 0.5 * drop(t(v) %*% A %*% v); gq <- function(v) drop(A %*% v)
gd <- grad_descent(fq, gq, c(2, -1.8), maxit = 30)
par(pty = "s", mar = c(4, 4, 2, 1))
contour(s, s, -z, nlevels = 15, main = sprintf("gradient descent, %d iterations", gd$iter))
lines(gd$path, type = "b", col = 2, pch = 19, cex = 0.6)
```

The number of iterations needed grows with the **condition number** $\lambda_{\max}/\lambda_{\min}$ of the Hessian — precisely the elongation of the ellipses in the earlier figure.

### Conjugate Gradient

The zig-zag is fixed by choosing each new direction to be **conjugate** to the previous ones with respect to the Hessian $Q$: $d_i'Qd_j = 0$ for $i \ne j$. For a quadratic $f(\theta) = \tfrac12\theta'Q\theta - b'\theta$, minimizing along $p$ such directions in succession reaches the exact minimum in **at most $p$ steps**, without ever forming or storing $Q$ — storage is $O(p)$, its defining advantage over Newton.

The nonlinear version builds directions from gradients only:
$$
d_0 = -g_0, \qquad d_{k+1} = -g_{k+1} + \beta_k d_k,
$$
$$
\beta_k^{\mathrm{FR}} = \frac{g_{k+1}'g_{k+1}}{g_k'g_k}\ \text{(Fletcher–Reeves)}, \qquad
\beta_k^{\mathrm{PR}} = \frac{g_{k+1}'(g_{k+1} - g_k)}{g_k'g_k}\ \text{(Polak–Ribière)},
$$
with $\theta_{k+1} = \theta_k + \alpha_k d_k$ from a line search and $g_k = \nabla f(\theta_k)$. Fletcher–Reeves and Polak–Ribière differ only in this correction coefficient. The method is sensitive to the accuracy of the line search, and conjugacy degrades on non-quadratic $f$, so it is usually restarted with $d = -g$ every $p$ iterations.

```{r}
#| label: fig-conjugate-gradient
#| fig-cap: "Conjugate gradient (blue) reaches the minimum of the same elongated quadratic in far fewer iterations than gradient descent (red)."
conj_grad <- function(f, gr, x0, maxit = 1000, tol = 1e-8, restart = length(x0)) {
  g0 <- gr(x0); d <- -g0; path <- x0
  for (k in 1:maxit) {
    a <- optimize(function(a) f(x0 + a * d), c(0, 10))$minimum
    x1 <- x0 + a * d; g1 <- gr(x1); path <- rbind(path, x1)
    if (sqrt(sum(g1^2)) < tol) break
    beta <- if (k %% restart == 0) 0 else max(0, sum(g1 * (g1 - g0)) / sum(g0^2))   # Polak-Ribiere+
    d <- -g1 + beta * d; x0 <- x1; g0 <- g1 }
  list(par = x1, iter = k, path = path)
}
cg <- conj_grad(fq, gq, c(2, -1.8))
par(pty = "s", mar = c(4, 4, 2, 1))
contour(s, s, -z, nlevels = 15, main = sprintf("conjugate gradient, %d iterations", cg$iter))
lines(gd$path, type = "b", col = 2, pch = 19, cex = 0.6); lines(cg$path, type = "b", col = 4, pch = 19, lwd = 2)
legend("topleft", c("gradient descent", "conjugate gradient"), col = c(2, 4), lwd = 2, bty = "n")
```

### Quasi-Newton: BFGS

BFGS sits between the gradient methods above and Newton's method: it builds an approximation $B_k \approx J(\theta_k)$ from successive gradient differences $s_k = \theta_{k+1} - \theta_k$, $y_k = g_{k+1} - g_k$, via a rank-two update that keeps $B_k$ positive definite by construction (which is exactly why the Wolfe curvature condition of an earlier section matters: it is what guarantees $s_k'y_k > 0$, and hence a well-defined update). Steps are $d_k = -B_k^{-1}g_k$ with a line search, giving **superlinear** convergence from only first derivatives. BFGS itself stores a full $p \times p$ approximation; **L-BFGS** keeps only the last few update vectors, reducing storage to $O(mp)$ and making it the default choice for high-dimensional smooth problems. For most likelihood problems where the Hessian is awkward to derive, BFGS with an analytic gradient is the best practical compromise.

Returning to the logistic-regression example, `optim()` can fit it with no Hessian at all — Nelder–Mead needs only the log-likelihood, CG and BFGS also take the analytic gradient already coded for Newton–Raphson:

```{r}
#| label: bfgs-vs-nm-cg
negll_fn  <- function(b) neg_logp_logistic(b, data$z, data$r)
neggr_fn  <- function(b) lik_score_info(b, data$r, data$z)$neg_score

fits <- list(NM   = optim(c(0, 0), negll_fn, method = "Nelder-Mead"),
             CG   = optim(c(0, 0), negll_fn, neggr_fn, method = "CG"),
             BFGS = optim(c(0, 0), negll_fn, neggr_fn, method = "BFGS", hessian = TRUE))
t(sapply(fits, function(f) c(f$par, f_evals = f$counts[1], g_evals = f$counts[2])))
sqrt(diag(solve(fits$BFGS$hessian)))     # SEs from the numerically computed Hessian at the optimum
```

All three land on the same estimate as `mle_logistic_nr` and `glm` above, and `optim(..., hessian = TRUE)` returns a finite-difference Hessian at the solution, so standard errors are available even for a method that never formed one during the iterations.

### Animation: Comparing the Optimizers {#sec-gradient-methods-app}

Press play to watch safeguarded Newton, gradient descent, conjugate gradient, and BFGS attack the same objective from the same start, on six test functions ranging from a well-conditioned quadratic to Rastrigin's many local minima. Click anywhere in the contour panel to try a different starting point.

::::: {.content-visible when-format="html"}
::: {#fig-gradient-methods-app}

```{shinylive-r}
#| standalone: true
#| viewerHeight: 820

library(shiny)

## ---------------------------------------------------------------- objectives
mk <- function(label, fv, xlim, ylim, start, xstar) {
  force(fv)
  list(label = label, fv = fv, f = function(p) fv(p[1], p[2]),
       xlim = xlim, ylim = ylim, start = start,
       xstar = matrix(xstar, ncol = 2))
}

FNS <- list(
  quad = mk(
    "(a) Well-conditioned quadratic",
    function(x, y) 0.5 * (2 * (x - 1)^2 + 1.6 * (x - 1) * (y - 1) +
                          2 * (y - 1)^2),
    c(-3, 4), c(-3, 4), c(-2, 3), c(1, 1)
  ),
  banana = mk(
    "(b) Rosenbrock banana (narrow curved ridge)",
    function(x, y) (1 - x)^2 + 100 * (y - x^2)^2,
    c(-2, 2), c(-1, 3), c(-1.2, 1), c(1, 1)
  ),
  mix = mk(
    "(c) Highly correlated, two modes",
    function(x, y) {
      rho <- 0.9; den <- 1 - rho^2
      q <- function(u, v) (u^2 - 2 * rho * u * v + v^2) / den
      -log(exp(-0.5 * q(x - 1.2, y - 1.2)) +
           0.6 * exp(-0.5 * q(x + 1.5, y + 1.5) / 1.5) + 1e-12)
    },
    c(-4, 4), c(-4, 4), c(-3, -0.5), c(1.2, 1.2)
  ),
  himmel = mk(
    "(d) Himmelblau (four global minima)",
    function(x, y) (x^2 + y - 11)^2 + (x + y^2 - 7)^2,
    c(-5, 5), c(-5, 5), c(-4, 4),
    rbind(c(3, 2), c(-2.805118, 3.131312),
          c(-3.779310, -3.283186), c(3.584428, -1.848126))
  ),
  beale = mk(
    "(e) Beale (flat plateau, sharp valley)",
    function(x, y) (1.5 - x + x * y)^2 + (2.25 - x + x * y^2)^2 +
                   (2.625 - x + x * y^3)^2,
    c(-4.5, 4.5), c(-4.5, 4.5), c(-1, 1), c(3, 0.5)
  ),
  rastrigin = mk(
    "(f) Rastrigin (many local minima)",
    function(x, y) 20 + x^2 - 10 * cos(2 * pi * x) +
                        y^2 - 10 * cos(2 * pi * y),
    c(-5.12, 5.12), c(-5.12, 5.12), c(-3.1, 4.2), c(0, 0)
  )
)

## ------------------------------------------------- numerical derivatives
num_grad <- function(f, x) {
  h <- 1e-5 * (1 + abs(x))
  g <- numeric(2)
  for (i in 1:2) {
    e <- numeric(2); e[i] <- h[i]
    g[i] <- (f(x + e) - f(x - e)) / (2 * h[i])
  }
  g
}

num_hess <- function(f, x) {
  h <- 1e-4 * (1 + abs(x))
  H <- matrix(0, 2, 2)
  for (i in 1:2) for (j in i:2) {
    ei <- numeric(2); ei[i] <- h[i]
    ej <- numeric(2); ej[j] <- h[j]
    H[i, j] <- (f(x + ei + ej) - f(x + ei - ej) -
                f(x - ei + ej) + f(x - ei - ej)) / (4 * h[i] * h[j])
    H[j, i] <- H[i, j]
  }
  H
}

## --------------------------------------------------- Wolfe line search
line_search <- function(f, g, x, d, a0 = 1, c1 = 1e-4, c2 = 0.9) {
  f0 <- f(x); gd0 <- sum(g * d)
  if (!is.finite(gd0) || gd0 >= 0) return(list(alpha = 0, ok = FALSE))
  lo <- 0; hi <- Inf; a <- a0
  for (it in 1:60) {
    fa <- f(x + a * d)
    if (!is.finite(fa) || fa > f0 + c1 * a * gd0) {
      hi <- a; a <- 0.5 * (lo + hi)
    } else {
      ga <- num_grad(f, x + a * d)
      if (sum(ga * d) < c2 * gd0) {
        lo <- a
        a <- if (is.finite(hi)) 0.5 * (lo + hi) else min(2 * a, 1e4)
      } else {
        return(list(alpha = a, ok = TRUE))
      }
    }
    if (a < 1e-14) return(list(alpha = 0, ok = FALSE))
  }
  list(alpha = a, ok = (a > 0))
}

## ------------------------------------------------------------- algorithms
run_alg <- function(alg, FN, x0, maxit = 60, tol = 1e-6) {
  f <- FN$f
  X <- matrix(NA_real_, maxit + 1, 2); G <- matrix(NA_real_, maxit + 1, 2)
  D <- matrix(NA_real_, maxit + 1, 2); Fv <- rep(NA_real_, maxit + 1)
  AL <- rep(NA_real_, maxit + 1)
  x <- x0; B <- diag(2)
  g_prev <- NULL; d_prev <- NULL; s_prev <- NULL
  n <- 0; note <- "maximum number of iterations reached"

  for (k in seq_len(maxit)) {
    g <- num_grad(f, x); gn <- sqrt(sum(g^2))
    X[k, ] <- x; G[k, ] <- g; Fv[k] <- f(x); n <- k
    if (!is.finite(gn)) { note <- "diverged"; break }
    if (gn < tol) { note <- "converged (||grad|| < 1e-6)"; break }

    if (alg == "nr") {
      H <- num_hess(f, x); H <- (H + t(H)) / 2
      ev <- eigen(H, symmetric = TRUE, only.values = TRUE)$values
      delta <- 1e-3 * max(1, max(abs(ev)))
      tau <- max(0, delta - min(ev))
      d <- tryCatch(solve(H + tau * diag(2), -g), error = function(e) -g)
    } else if (alg == "gd") {
      d <- -g
    } else if (alg == "cg") {
      if (is.null(d_prev)) {
        d <- -g
      } else {
        y <- g - g_prev
        beta <- max(0, sum(g * y) / max(sum(g_prev^2), 1e-300))   # PR+
        d <- -g + beta * d_prev
      }
      if (sum(g * d) > -1e-12) d <- -g                            # restart
    } else {                                                      # BFGS
      if (!is.null(s_prev)) {
        y <- g - g_prev; sy <- sum(s_prev * y)
        if (sy > 1e-10) {
          rho <- 1 / sy
          V <- diag(2) - rho * outer(s_prev, y)
          B <- V %*% B %*% t(V) + rho * outer(s_prev, s_prev)
        }
      }
      d <- -as.vector(B %*% g)
      if (sum(g * d) > -1e-12) { B <- diag(2); d <- -g }
    }

    ls <- line_search(f, g, x, d)
    D[k, ] <- d; AL[k] <- ls$alpha
    if (ls$alpha <= 0) { note <- "line search failed (flat or non-descent)"; break }

    s_prev <- ls$alpha * d; g_prev <- g; d_prev <- d
    x <- x + s_prev
  }

  list(X = X[1:n, , drop = FALSE], G = G[1:n, , drop = FALSE],
       D = D[1:n, , drop = FALSE], F = Fv[1:n], A = AL[1:n],
       n = n, note = note)
}

ALGS <- c("Safeguarded Newton (modified Hessian + line search)" = "nr",
          "Gradient descent + line search"    = "gd",
          "Conjugate gradient (PR+)"          = "cg",
          "BFGS (quasi-Newton)"               = "bfgs")

## ---------------------------- widen the data box to the device aspect ratio
expand_box <- function(xlim, ylim, pin) {
  if (length(pin) != 2 || any(!is.finite(pin)) || any(pin <= 0))
    return(list(xlim = xlim, ylim = ylim))
  r <- pin[1] / pin[2]
  wd <- diff(xlim); ht <- diff(ylim)
  if (wd / ht < r) {
    w <- ht * r; m <- mean(xlim); xlim <- m + c(-0.5, 0.5) * w
  } else {
    h <- wd / r; m <- mean(ylim); ylim <- m + c(-0.5, 0.5) * h
  }
  list(xlim = xlim, ylim = ylim)
}

## --------------------------------------------------------------------- UI
ui <- fluidPage(
  fluidRow(
    ## ---- column 1: controls -------------------------------------------
    column(
      3,
      selectInput("alg", "Algorithm", ALGS),
      selectInput("fn", "Objective function",
                  setNames(names(FNS), sapply(FNS, `[[`, "label"))),
      sliderInput("k", "Iteration k", min = 0, max = 1, value = 0, step = 1),
      checkboxInput("showpath", "Show the whole path", TRUE),
      htmlOutput("info")
    ),
    ## ---- column 2: playback -------------------------------------------
    column(
      2,
      div(
        style = "margin-top:26px;",
        actionButton("play", "Play", class = "btn-primary btn-lg",
                     style = "width:100%; margin-bottom:10px;"),
        actionButton("step", "Next", class = "btn-lg",
                     style = "width:100%; margin-bottom:10px;"),
        actionButton("back", "Prev", class = "btn-lg",
                     style = "width:100%;"),
        hr(),
        checkboxInput("autoplay", "Autoplay on Changing Initial", TRUE),
        actionButton("reset", "Reset initial",
                     style = "width:100%; margin-bottom:8px;"),
        helpText("Click inside the contour plot to choose a new initial value.")
      )
    ),
    ## ---- column 3: plots ----------------------------------------------
    column(
      7,
      plotOutput("contour", click = "click", height = "540px"),
      plotOutput("conv", height = "190px")
    )
  )
)

## ----------------------------------------------------------------- server
server <- function(input, output, session) {

  FN <- reactive(FNS[[input$fn]])
  start <- reactiveVal(FNS[[1]]$start)
  playing <- reactiveVal(FALSE)

  observeEvent(input$fn, start(FNS[[input$fn]]$start))
  observeEvent(input$reset, start(FN()$start))
  observeEvent(input$click, {
    p <- c(input$click$x, input$click$y)
    if (all(is.finite(p))) start(p)
  })

  path <- reactive({
    req(start())
    run_alg(input$alg, FN(), start())
  })

  observeEvent(path(), {
    updateSliderInput(session, "k", max = max(1, path()$n - 1), value = 0)
    playing(isTRUE(input$autoplay) && path()$n > 1)
  })

  observeEvent(input$play, {
    if (isTRUE(playing())) {
      playing(FALSE)
    } else {
      if (as.integer(input$k) >= path()$n - 1)
        updateSliderInput(session, "k", value = 0)
      playing(TRUE)
    }
  })

  observeEvent(playing(), {
    updateActionButton(session, "play",
                       label = if (isTRUE(playing())) "Pause" else "Play")
  })

  observe({
    if (!isTRUE(playing())) return()
    kmax <- isolate(path()$n) - 1
    k <- isolate(as.integer(input$k))
    if (k >= kmax) { playing(FALSE); return() }
    invalidateLater(900, session)
    updateSliderInput(session, "k", value = k + 1)
  })

  observeEvent(input$step, {
    playing(FALSE)
    updateSliderInput(session, "k",
                      value = min(as.integer(input$k) + 1, max(0, path()$n - 1)))
  })

  observeEvent(input$back, {
    playing(FALSE)
    updateSliderInput(session, "k", value = max(as.integer(input$k) - 1, 0))
  })

  output$contour <- renderPlot({
    FNc <- FN(); P <- path()
    k <- min(as.integer(input$k), P$n - 1) + 1L

    par(mar = c(4, 4, 2.5, 1))
    plot.new()
    bx <- expand_box(FNc$xlim, FNc$ylim, par("pin"))
    plot.window(xlim = bx$xlim, ylim = bx$ylim, xaxs = "i", yaxs = "i")

    xs <- seq(bx$xlim[1], bx$xlim[2], length.out = 180)
    ys <- seq(bx$ylim[1], bx$ylim[2], length.out = 180)
    z  <- outer(xs, ys, FNc$fv)
    zf <- z[is.finite(z)]
    lv <- unique(quantile(zf, probs = seq(0, 1, length.out = 30)^1.7))
    zr <- matrix(rank(z, na.last = "keep", ties.method = "average"),
                 nrow = length(xs))

    image(xs, ys, zr, add = TRUE,
          col = colorRampPalette(c("#f7fbff", "#9ec4e3"))(128))
    contour(xs, ys, z, levels = lv, add = TRUE,
            col = "grey45", drawlabels = FALSE)
    axis(1); axis(2); box()
    title(main = paste0(names(ALGS)[ALGS == input$alg], "  |  ", FNc$label),
          xlab = expression(x[1]), ylab = expression(x[2]), cex.main = 1.0)

    points(FNc$xstar[, 1], FNc$xstar[, 2], pch = 8, col = "red",
           cex = 1.3, lwd = 2)

    if (isTRUE(input$showpath) && k > 1)
      lines(P$X[1:k, 1], P$X[1:k, 2], col = "grey25", lty = 2)
    points(P$X[1:k, 1], P$X[1:k, 2], pch = 20, col = "grey35", cex = 0.9)

    xk <- P$X[k, ]; gk <- P$G[k, ]; gn <- sqrt(sum(gk^2))
    L <- 0.09 * diff(bx$xlim)                      # fixed display length
    if (gn > 0) {
      u <- -gk / gn
      arrows(xk[1], xk[2], xk[1] + L * u[1], xk[2] + L * u[2],
             col = "grey45", lwd = 3, length = 0.10)
      arrows(xk[1], xk[2], xk[1] - gk[1], xk[2] - gk[2],
             col = "#E8820C", lwd = 3, length = 0.13)
    }
    if (k < P$n) {
      arrows(xk[1], xk[2], P$X[k + 1, 1], P$X[k + 1, 2],
             col = "#2E8B57", lwd = 3, length = 0.13)
      points(P$X[k + 1, 1], P$X[k + 1, 2], pch = 21, bg = "white",
             col = "#2E8B57", cex = 1.2)
    }
    points(xk[1], xk[2], pch = 19, col = "black", cex = 1.3)

    legend("topleft", lwd = 3, cex = 0.9, bg = "#ffffffcc", box.col = NA,
           col = c("grey45", "#E8820C", "#2E8B57"),
           legend = c(expression(-nabla * f(x[k])~"(direction only)"),
                      expression(-nabla * f(x[k])~"(actual length)"),
                      expression(alpha[k] * d[k]~"(step taken)")))
  })

  output$conv <- renderPlot({
    P <- path(); k <- min(as.integer(input$k), P$n - 1) + 1L
    gn <- sqrt(rowSums(P$G^2))
    gn[!is.finite(gn) | gn <= 0] <- NA
    par(mar = c(4, 4.5, 1.5, 1))
    plot(seq_len(P$n) - 1, gn, type = "b", pch = 20, log = "y",
         xlab = "iteration k", ylab = expression("||" * nabla * f * "||"),
         col = "grey40")
    points(k - 1, gn[k], pch = 19, col = "#E8820C", cex = 1.6)
  })

  output$info <- renderUI({
    P <- path(); k <- min(as.integer(input$k), P$n - 1) + 1L
    xk <- P$X[k, ]; gk <- P$G[k, ]; gn <- sqrt(sum(gk^2))
    ang <- NA
    if (k < P$n) {
      sk <- P$X[k + 1, ] - xk
      cs <- sum(-gk * sk) / (gn * sqrt(sum(sk^2)))
      ang <- acos(max(-1, min(1, cs))) * 180 / pi
    }
    fmt <- function(v, d = 4) formatC(v, format = "g", digits = d)
    HTML(paste0(
      "<hr><b>iteration</b> ", k - 1, " of ", P$n - 1, "<br>",
      "<b>x<sub>k</sub></b> = (", fmt(xk[1]), ", ", fmt(xk[2]), ")<br>",
      "<b>f(x<sub>k</sub>)</b> = ", fmt(P$F[k], 6), "<br>",
      "<b>||grad||</b> = ", fmt(gn), "<br>",
      "<b>step size &alpha;<sub>k</sub></b> = ",
      if (k < P$n) fmt(P$A[k]) else "&ndash;", "<br>",
      "<b>angle(grey, green)</b> = ",
      if (is.na(ang)) "&ndash;" else paste0(fmt(ang, 3), "&deg;"),
      "<br><br><i>", P$note, "</i>"
    ))
  })
}

shinyApp(ui, server)
```

**Shinylive App Comparing Gradient Descent and Newton-Raphson Optimization Algorithms.** Choose an algorithm and objective, then step or play through the iterations.

:::
:::::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit05_MLE.html#sec-gradient-methods-app>.

:::

The app is built the same way as the Newton–Raphson divergence app earlier in this chapter (@sec-nr-divergence-app); see the callout there for how to reproduce a shinylive app of your own.

## Nelder–Mead (Simplex Method)

Nelder–Mead maintains a **simplex** of $p+1$ points (a triangle in 2D) and moves it downhill using only function values, with no derivatives at all:

1. Order the vertices $f(\theta_{(1)}) \le \cdots \le f(\theta_{(p+1)})$; let $\overline\theta$ be the centroid of the best $p$.

2. **Reflect** the worst vertex through $\overline\theta$; if the reflected point is the new best, try **expanding** further.
3. If reflection is poor, **contract** toward $\overline\theta$; if that also fails, **shrink** all vertices toward the best.

```{r}
#| label: nelder-mead-rosenbrock
rosen <- function(v) (1 - v[1])^2 + 100 * (v[2] - v[1]^2)^2       # Rosenbrock "banana", min at (1, 1)
nm <- optim(c(-1.2, 1), rosen, method = "Nelder-Mead"); c(nm$par, evals = nm$counts["function"])
```

The method needs no derivatives and tolerates mild non-smoothness, but it converges slowly, degrades badly as $p$ grows past roughly 10–20, and can stall at non-stationary points. It is `optim`'s default, which makes it by far the most-used — and most-overused — method in R: a reasonable first look at an unfamiliar objective, but a poor choice for a likelihood you will fit thousands of times.

### Animation for the Nelder–Mead Algorithm {#sec-nelder-mead-app}

The interactive figure below animates every reflection, expansion, contraction, and shrink on six test objectives; click inside the panel to place a new starting simplex.

::::: {.content-visible when-format="html"}
::: {#fig-nelder-mead-app}

```{shinylive-r}
#| standalone: true
#| viewerHeight: 820

library(shiny)

## ---------------------------------------------------------------- objectives
mk <- function(label, fv, xlim, ylim, start, xstar) {
  force(fv)
  list(label = label, fv = fv, f = function(p) fv(p[1], p[2]),
       xlim = xlim, ylim = ylim, start = start,
       xstar = matrix(xstar, ncol = 2))
}

FNS <- list(
  quad = mk(
    "(a) Well-conditioned quadratic",
    function(x, y) 0.5 * (2 * (x - 1)^2 + 1.6 * (x - 1) * (y - 1) +
                          2 * (y - 1)^2),
    c(-3, 4), c(-3, 4), c(-2, 3), c(1, 1)
  ),
  banana = mk(
    "(b) Rosenbrock banana (narrow curved ridge)",
    function(x, y) (1 - x)^2 + 100 * (y - x^2)^2,
    c(-2, 2), c(-1, 3), c(-1.2, 1), c(1, 1)
  ),
  mix = mk(
    "(c) Highly correlated, two modes",
    function(x, y) {
      rho <- 0.9; den <- 1 - rho^2
      q <- function(u, v) (u^2 - 2 * rho * u * v + v^2) / den
      -log(exp(-0.5 * q(x - 1.2, y - 1.2)) +
           0.6 * exp(-0.5 * q(x + 1.5, y + 1.5) / 1.5) + 1e-12)
    },
    c(-4, 4), c(-4, 4), c(-3, -0.5), c(1.2, 1.2)
  ),
  himmel = mk(
    "(d) Himmelblau (four global minima)",
    function(x, y) (x^2 + y - 11)^2 + (x + y^2 - 7)^2,
    c(-5, 5), c(-5, 5), c(-4, 4),
    rbind(c(3, 2), c(-2.805118, 3.131312),
          c(-3.779310, -3.283186), c(3.584428, -1.848126))
  ),
  beale = mk(
    "(e) Beale (flat plateau, sharp valley)",
    function(x, y) (1.5 - x + x * y)^2 + (2.25 - x + x * y^2)^2 +
                   (2.625 - x + x * y^3)^2,
    c(-4.5, 4.5), c(-4.5, 4.5), c(-1, 1), c(3, 0.5)
  ),
  rastrigin = mk(
    "(f) Rastrigin (many local minima)",
    function(x, y) 20 + x^2 - 10 * cos(2 * pi * x) +
                        y^2 - 10 * cos(2 * pi * y),
    c(-5.12, 5.12), c(-5.12, 5.12), c(-3.1, 4.2), c(0, 0)
  )
)

## ------------------------------------------------------------- Nelder-Mead
## coefficients: reflection a, expansion g, contraction r, shrink s
run_nm <- function(FN, x0, h, a = 1, g = 2, r = 0.5, s = 0.5,
                   maxit = 80, tol = 1e-7) {
  f <- FN$f
  S <- rbind(x0, x0 + c(h, 0), x0 + c(0, h))
  fS <- apply(S, 1, f)
  steps <- list()
  note <- "maximum number of iterations reached"

  for (k in seq_len(maxit)) {
    o <- order(fS); S <- S[o, , drop = FALSE]; fS <- fS[o]
    dia <- max(c(sqrt(sum((S[1, ] - S[2, ])^2)),
                 sqrt(sum((S[1, ] - S[3, ])^2)),
                 sqrt(sum((S[2, ] - S[3, ])^2))))

    if (dia < tol || diff(range(fS)) < 1e-12) {
      steps[[k]] <- list(S = S, fS = fS, dia = dia, op = "converged",
                         cen = colMeans(S[1:2, , drop = FALSE]),
                         xr = NULL, xe = NULL, xc = NULL, Sn = S)
      note <- "converged (simplex collapsed)"
      break
    }

    cen <- colMeans(S[1:2, , drop = FALSE])          # centroid of the best two
    xr <- cen + a * (cen - S[3, ]); fr <- f(xr)
    xe <- NULL; xc <- NULL
    Sn <- S; fn <- fS

    if (fr < fS[1]) {                                 # better than the best
      xe <- cen + g * (xr - cen); fe <- f(xe)
      if (fe < fr) { Sn[3, ] <- xe; fn[3] <- fe; op <- "expansion" }
      else         { Sn[3, ] <- xr; fn[3] <- fr; op <- "reflection" }
    } else if (fr < fS[2]) {                          # middling: accept
      Sn[3, ] <- xr; fn[3] <- fr; op <- "reflection"
    } else {
      if (fr < fS[3]) {                               # outside contraction
        xc <- cen + r * (xr - cen); fc <- f(xc)
        if (fc <= fr) { Sn[3, ] <- xc; fn[3] <- fc; op <- "outside contraction" }
        else op <- "shrink"
      } else {                                        # inside contraction
        xc <- cen + r * (S[3, ] - cen); fc <- f(xc)
        if (fc < fS[3]) { Sn[3, ] <- xc; fn[3] <- fc; op <- "inside contraction" }
        else op <- "shrink"
      }
      if (op == "shrink") {
        Sn[2, ] <- S[1, ] + s * (S[2, ] - S[1, ])
        Sn[3, ] <- S[1, ] + s * (S[3, ] - S[1, ])
        fn[2] <- f(Sn[2, ]); fn[3] <- f(Sn[3, ])
      }
    }

    steps[[k]] <- list(S = S, fS = fS, dia = dia, op = op, cen = cen,
                       xr = xr, xe = xe, xc = xc, Sn = Sn)
    S <- Sn; fS <- fn
  }

  list(steps = steps, n = length(steps), note = note)
}

## ---------------------------- widen the data box to the device aspect ratio
expand_box <- function(xlim, ylim, pin) {
  if (length(pin) != 2 || any(!is.finite(pin)) || any(pin <= 0))
    return(list(xlim = xlim, ylim = ylim))
  rr <- pin[1] / pin[2]
  wd <- diff(xlim); ht <- diff(ylim)
  if (wd / ht < rr) {
    w <- ht * rr; m <- mean(xlim); xlim <- m + c(-0.5, 0.5) * w
  } else {
    hh <- wd / rr; m <- mean(ylim); ylim <- m + c(-0.5, 0.5) * hh
  }
  list(xlim = xlim, ylim = ylim)
}

COL <- c(refl = "#E8820C", exp = "#7B3FBF", con = "#1F6FB4",
         new = "#2E8B57", best = "#2E8B57", worst = "#C81E1E")

## --------------------------------------------------------------------- UI
ui <- fluidPage(
  fluidRow(
    column(
      3,
      selectInput("fn", "Objective function",
                  setNames(names(FNS), sapply(FNS, `[[`, "label"))),
      sliderInput("k", "Iteration k", min = 0, max = 1, value = 0, step = 1),
      sliderInput("h", "Initial simplex size (% of range)",
                  min = 2, max = 40, value = 12, step = 1),
      checkboxInput("showpath", "Keep past simplices", TRUE),
      htmlOutput("info")
    ),
    column(
      2,
      div(
        style = "margin-top:26px;",
        actionButton("play", "Play", class = "btn-primary btn-lg",
                     style = "width:100%; margin-bottom:10px;"),
        actionButton("step", "Next", class = "btn-lg",
                     style = "width:100%; margin-bottom:10px;"),
        actionButton("back", "Prev", class = "btn-lg",
                     style = "width:100%;"),
        hr(),
        checkboxInput("autoplay", "Autoplay on change", FALSE),
        actionButton("reset", "Reset initial",
                     style = "width:100%; margin-bottom:8px;"),
        helpText("Click inside the contour plot to choose a new initial value.")
      )
    ),
    column(
      7,
      plotOutput("contour", click = "click", height = "540px"),
      plotOutput("conv", height = "190px")
    )
  )
)

## ----------------------------------------------------------------- server
server <- function(input, output, session) {

  FN <- reactive(FNS[[input$fn]])
  start <- reactiveVal(FNS[[1]]$start)
  playing <- reactiveVal(FALSE)

  observeEvent(input$fn, start(FNS[[input$fn]]$start))
  observeEvent(input$reset, start(FN()$start))
  observeEvent(input$click, {
    p <- c(input$click$x, input$click$y)
    if (all(is.finite(p))) start(p)
  })

  path <- reactive({
    req(start())
    FNc <- FN()
    run_nm(FNc, start(), h = input$h / 100 * diff(FNc$xlim))
  })

  observeEvent(path(), {
    updateSliderInput(session, "k", max = max(1, path()$n - 1), value = 0)
    playing(isTRUE(input$autoplay) && path()$n > 1)
  })

  observeEvent(input$play, {
    if (isTRUE(playing())) {
      playing(FALSE)
    } else {
      if (as.integer(input$k) >= path()$n - 1)
        updateSliderInput(session, "k", value = 0)
      playing(TRUE)
    }
  })

  observeEvent(playing(), {
    updateActionButton(session, "play",
                       label = if (isTRUE(playing())) "Pause" else "Play")
  })

  observe({
    if (!isTRUE(playing())) return()
    kmax <- isolate(path()$n) - 1
    k <- isolate(as.integer(input$k))
    if (k >= kmax) { playing(FALSE); return() }
    invalidateLater(900, session)
    updateSliderInput(session, "k", value = k + 1)
  })

  observeEvent(input$step, {
    playing(FALSE)
    updateSliderInput(session, "k",
                      value = min(as.integer(input$k) + 1, max(0, path()$n - 1)))
  })

  observeEvent(input$back, {
    playing(FALSE)
    updateSliderInput(session, "k", value = max(as.integer(input$k) - 1, 0))
  })

  output$contour <- renderPlot({
    FNc <- FN(); P <- path()
    k <- min(as.integer(input$k), P$n - 1) + 1L
    st <- P$steps[[k]]

    par(mar = c(4, 4, 3, 1))
    plot.new()
    bx <- expand_box(FNc$xlim, FNc$ylim, par("pin"))
    plot.window(xlim = bx$xlim, ylim = bx$ylim, xaxs = "i", yaxs = "i")

    xs <- seq(bx$xlim[1], bx$xlim[2], length.out = 180)
    ys <- seq(bx$ylim[1], bx$ylim[2], length.out = 180)
    z  <- outer(xs, ys, FNc$fv)
    zf <- z[is.finite(z)]
    lv <- unique(quantile(zf, probs = seq(0, 1, length.out = 30)^1.7))
    zr <- matrix(rank(z, na.last = "keep", ties.method = "average"),
                 nrow = length(xs))

    image(xs, ys, zr, add = TRUE,
          col = colorRampPalette(c("#f7fbff", "#9ec4e3"))(128))
    contour(xs, ys, z, levels = lv, add = TRUE,
            col = "grey45", drawlabels = FALSE)
    axis(1); axis(2); box()
    title(main = paste0("Nelder-Mead  |  ", FNc$label),
          xlab = expression(x[1]), ylab = expression(x[2]), cex.main = 1.0)
    mtext(paste0("iteration ", k - 1, ":  ", st$op), side = 3, line = 0.1,
          cex = 1.1, font = 2)

    points(FNc$xstar[, 1], FNc$xstar[, 2], pch = 8, col = "red",
           cex = 1.3, lwd = 2)

    ## past simplices
    if (isTRUE(input$showpath) && k > 1)
      for (j in seq_len(k - 1))
        polygon(P$steps[[j]]$S[, 1], P$steps[[j]]$S[, 2],
                border = "grey55", lty = 3)

    ## current simplex
    S <- st$S
    polygon(S[, 1], S[, 2], border = "black", lwd = 2,
            col = adjustcolor("white", alpha.f = 0.35))
    points(st$cen[1], st$cen[2], pch = 3, col = "black", lwd = 2, cex = 1.1)
    points(S[, 1], S[, 2], pch = 21, cex = 1.6, lwd = 2, bg = "white",
           col = c(COL["best"], "grey30", COL["worst"]))
    text(S[, 1], S[, 2], labels = c("B", "G", "W"), pos = 3, offset = 0.6,
         font = 2, col = c(COL["best"], "grey30", COL["worst"]))

    ## candidate points generated this iteration
    if (!is.null(st$xr)) {
      segments(S[3, 1], S[3, 2], st$xr[1], st$xr[2],
               col = COL["refl"], lwd = 2, lty = 2)
      points(st$xr[1], st$xr[2], pch = 19, col = COL["refl"], cex = 1.5)
    }
    if (!is.null(st$xe)) {
      segments(st$xr[1], st$xr[2], st$xe[1], st$xe[2],
               col = COL["exp"], lwd = 2, lty = 2)
      points(st$xe[1], st$xe[2], pch = 19, col = COL["exp"], cex = 1.5)
    }
    if (!is.null(st$xc))
      points(st$xc[1], st$xc[2], pch = 19, col = COL["con"], cex = 1.5)

    ## the simplex that results
    if (st$op != "converged") {
      polygon(st$Sn[, 1], st$Sn[, 2], border = COL["new"], lwd = 3, lty = 2)
      if (st$op == "shrink")
        arrows(S[2:3, 1], S[2:3, 2], st$Sn[2:3, 1], st$Sn[2:3, 2],
               col = COL["con"], lwd = 2, length = 0.10)
    }

    legend("topleft", cex = 0.85, bg = "#ffffffcc", box.col = NA,
           pch = c(21, 21, 3, 19, 19, 19, NA),
           lty = c(NA, NA, NA, NA, NA, NA, 2),
           lwd = c(2, 2, 2, NA, NA, NA, 3),
           col = c(COL["best"], COL["worst"], "black", COL["refl"],
                   COL["exp"], COL["con"], COL["new"]),
           legend = c("best vertex B", "worst vertex W",
                      "centroid of B and G", "reflection", "expansion",
                      "contraction / shrink", "next simplex"))
  })

  output$conv <- renderPlot({
    P <- path(); k <- min(as.integer(input$k), P$n - 1) + 1L
    fb <- sapply(P$steps, function(s) s$fS[1])
    fw <- sapply(P$steps, function(s) s$fS[3])
    par(mar = c(4, 4.5, 1.5, 1))
    matplot(seq_len(P$n) - 1, cbind(fb, fw), type = "b", pch = 20, lty = 1,
            col = c(COL["best"], COL["worst"]),
            xlab = "iteration k", ylab = "f at simplex vertices")
    abline(v = k - 1, col = "grey60", lty = 2)
    legend("topright", bty = "n", cex = 0.9, lty = 1, pch = 20,
           col = c(COL["best"], COL["worst"]), legend = c("best", "worst"))
  })

  output$info <- renderUI({
    P <- path(); k <- min(as.integer(input$k), P$n - 1) + 1L
    st <- P$steps[[k]]
    fmt <- function(v, d = 4) formatC(v, format = "g", digits = d)
    HTML(paste0(
      "<hr><b>iteration</b> ", k - 1, " of ", P$n - 1, "<br>",
      "<b>operation</b>: ", st$op, "<br><br>",
      "<b>B</b> = (", fmt(st$S[1, 1]), ", ", fmt(st$S[1, 2]), "),  f = ",
      fmt(st$fS[1], 6), "<br>",
      "<b>G</b> = (", fmt(st$S[2, 1]), ", ", fmt(st$S[2, 2]), "),  f = ",
      fmt(st$fS[2], 6), "<br>",
      "<b>W</b> = (", fmt(st$S[3, 1]), ", ", fmt(st$S[3, 2]), "),  f = ",
      fmt(st$fS[3], 6), "<br><br>",
      "<b>simplex diameter</b> = ", fmt(st$dia), "<br>",
      "<b>f range</b> = ", fmt(diff(range(st$fS))),
      "<br><br><i>", P$note, "</i>"
    ))
  })
}

shinyApp(ui, server)
```

**Shinylive App Illustrating Nelder–Mead.** Reflection, expansion, contraction and shrink steps are colour-coded on the contour plot as the simplex moves.

:::
:::::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit05_MLE.html#sec-nelder-mead-app>.

:::

## Stochastic Gradient Descent

### Motivation: Large $n$

For iid data the log-likelihood and its gradient are **sums over observations**:
$$
\ell(\theta) = \sum_{i=1}^n \ell_i(\theta), \qquad \nabla\ell(\theta) = \sum_{i=1}^n \nabla\ell_i(\theta).
$$

Every method in the two previous sections needs the **full** gradient — an $O(np)$ pass through all the data — at every single iteration, and Newton needs $O(np^2)$ for the Hessian on top of that. With $n$ in the millions and $p$ in the millions, as in a large neural network, even one full gradient evaluation is expensive.

The idea behind stochastic gradient descent is that $\nabla\ell_i(\theta)$ for a randomly chosen $i$ is an **unbiased estimate** of the average gradient,
$$
E_i\big[\nabla\ell_i(\theta)\big] = \frac{1}{n}\nabla\ell(\theta),
$$
so a noisy but cheap gradient estimate is often enough to make progress — the same Monte Carlo idea used elsewhere in this book, now applied to optimization rather than integration.

### The SGD Algorithm

For $t = 1, 2, \ldots$:

1. Sample an index $i_t$ uniformly from $\{1, \ldots, n\}$ — or shuffle the data once and cycle through it, so that one full pass is called an **epoch**.

2. Update in the ascent direction of that single observation's log-likelihood:
$$
\theta_{t+1} = \theta_t + \gamma_t\,\nabla\ell_{i_t}(\theta_t).
$$

**Mini-batch SGD** averages the gradient over a batch $B_t$ of $m$ observations (typically $m = 32$–$512$ in machine-learning practice, though far smaller batches suffice for the simple models in this book):
$$
\theta_{t+1} = \theta_t + \gamma_t\,\frac{1}{m}\sum_{i\in B_t}\nabla\ell_i(\theta_t),
$$
which reduces the variance of the gradient estimate by a factor of $1/m$ at $m$ times the per-step cost, and is what lets the computation vectorize efficiently or run on a GPU. The **learning rate** $\gamma_t$ plays the role that a line search played for the deterministic methods above — a line search itself would need a full pass over the data to evaluate $\ell$, defeating the purpose.

### Learning Rate and Convergence

Because each update uses a noisy gradient, a constant learning rate $\gamma$ makes $\theta_t$ **bounce around** the optimum indefinitely, with a variance proportional to $\gamma$. The classical result of Robbins and Monro (1951) is that $\theta_t$ converges to the optimum provided
$$
\sum_t \gamma_t = \infty \quad\text{and}\quad \sum_t \gamma_t^2 < \infty, \qquad\text{e.g. } \gamma_t = \frac{\gamma_0}{1 + t/t_0},
$$
the first condition allowing the iterate to travel arbitrarily far from its start, and the second forcing the noise to die out fast enough for it to settle.

A learning rate that is too large causes divergence or persistent noise; one that is too small is painfully slow. Convergence is only **sublinear** — far slower than Newton's quadratic rate *per iteration* — but each SGD iteration costs $O(p)$ or $O(mp)$ rather than $O(np)$ for a full gradient or $O(np^2)$ for a Hessian, and early progress is often fast even when final convergence is not. Common refinements include **momentum** (averaging past gradients to damp the noise), **Adam** (per-coordinate adaptive learning rates), and **Polyak–Ruppert averaging** of the iterates, $\overline\theta_T = \tfrac1T\sum_{t\le T}\theta_t$, which restores the $\sqrt n$-efficiency of the MLE that the noisy path alone does not have.

### Animation for Stochastic Gradient Descent {#sec-sgd-app}

The app below trains a small one-hidden-layer neural network ($1 \to H \to 1$, so $3H+1$ parameters) by gradient ascent on the log-likelihood (equivalently, gradient descent on the mean squared error) under two data-generating processes — one inside the network's own model class, one outside it — and lets you compare a mini-batch run against a full-batch run sharing the same data, initial weights, and number of epochs.

::::: {.content-visible when-format="html"}
::: {#fig-sgd-app}

```{shinylive-r}
#| standalone: true
#| viewerHeight: 830

library(shiny)

## ------------------------------------------------------------------- data
## Two data-generating processes. The teacher network lies inside the
## student's model class; the bumps do not.
TEACH <- list(w = c(4, 3, 5), cc = c(4, 0, -5), v = c(1.0, -1.2, 0.8), b = 0)

g_teacher <- function(x) {
  Hm <- tanh(outer(x, TEACH$w) + rep(TEACH$cc, each = length(x)))
  as.vector(Hm %*% TEACH$v + TEACH$b)
}

g_bumps <- function(x)
  1.2 * exp(-9 * (x + 0.7)^2) - 1.0 * exp(-9 * (x - 0.8)^2)

DGPS <- c("Teacher network (1 - 3 tanh - 1)" = "teacher",
          "Two Gaussian bumps (outside the model class)" = "bumps")

gtrue <- function(x, dgp)
  if (identical(dgp, "bumps")) g_bumps(x) else g_teacher(x)

XG <- seq(-2, 2, length.out = 300)

make_data <- function(n, sigma, seed, dgp) {
  set.seed(500 + seed)
  x <- sort(runif(n, -2, 2))
  list(x = x, y = gtrue(x, dgp) + rnorm(n, 0, sigma), n = n, dgp = dgp)
}

## ------------------------------------------------- network: 1 - H tanh - 1
unpack <- function(p, H)
  list(w = p[1:H], cc = p[H + (1:H)], v = p[2 * H + (1:H)], b = p[3 * H + 1])

fnet <- function(p, x, H) {
  P <- unpack(p, H)
  Hm <- tanh(outer(x, P$w) + rep(P$cc, each = length(x)))
  as.vector(Hm %*% P$v + P$b)
}

gnet <- function(p, x, y, H) {
  n <- length(x); P <- unpack(p, H)
  Hm <- tanh(outer(x, P$w) + rep(P$cc, each = n))
  r <- as.vector(Hm %*% P$v + P$b) - y
  S <- (1 - Hm^2) * outer(r, P$v)
  c(as.vector(colSums(S * x)) / n,        # dw
    as.vector(colSums(S)) / n,            # dc
    as.vector(crossprod(Hm, r)) / n,      # dv
    mean(r))                              # db
}

mse <- function(p, H, x, y) 0.5 * mean((fnet(p, x, H) - y)^2)

init_p <- function(H, seed) {
  set.seed(1000 + seed)
  c(rnorm(H, 0, 2), rnorm(H, 0, 1.5), rnorm(H, 0, 0.7), 0)
}

## ---------------------------------------------------------------- training
## m = batch size; m >= n is full-batch gradient descent.
train <- function(p0, H, m, a, nepoch, seed, D) {
  n <- D$n
  m <- max(1, min(m, n))
  B <- ceiling(n / m)                              # updates per epoch
  nup <- B * nepoch
  P <- matrix(NA_real_, nup + 1, length(p0))
  L <- rep(NA_real_, nup + 1); EP <- rep(NA_real_, nup + 1)
  IDX <- vector("list", nup + 1)
  set.seed(seed)
  p <- p0; k <- 0; bad <- FALSE
  for (e in seq_len(nepoch)) {
    perm <- if (B == 1) seq_len(n) else sample.int(n)
    grp <- split(perm, ceiling(seq_len(n) / m))    # last batch may be smaller
    for (b in seq_len(B)) {
      k <- k + 1
      idx <- grp[[b]]
      P[k, ] <- p; IDX[[k]] <- idx
      L[k] <- mse(p, H, D$x, D$y); EP[k] <- (k - 1) / B
      p <- p - a * gnet(p, D$x[idx], D$y[idx], H)
      if (any(!is.finite(p))) { bad <- TRUE; break }
    }
    if (bad) break
  }
  if (!bad) {
    k <- k + 1
    P[k, ] <- p; IDX[[k]] <- seq_len(n)
    L[k] <- mse(p, H, D$x, D$y); EP[k] <- (k - 1) / B
  }
  n_rec <- max(1, k)
  list(P = P[1:n_rec, , drop = FALSE], L = L[1:n_rec], EP = EP[1:n_rec],
       IDX = IDX[1:n_rec], n = n_rec, B = B, m = m, diverged = bad)
}

## ------------------------------------------------------------ weight plot
## Diverging ramp: blue = negative, red = positive, magnitude by saturation.
DIV <- colorRampPalette(c("#08306B", "#2171B5", "#6BAED6", "#DEEBF7",
                          "#F7F7F7",
                          "#FEE0D2", "#FC9272", "#CB181D", "#67000D"))(201)

## compander: maps v in [-wmax, wmax] to [-1, 1], boosting small values.
tone <- function(v, wmax, k) {
  u <- pmax(-1, pmin(1, v / wmax))
  sign(u) * log1p(k * abs(u)) / log1p(k)
}

wcol <- function(v, wmax, k) DIV[round((tone(v, wmax, k) + 1) * 100) + 1]
wlwd <- function(v, wmax, k) 0.8 + 6 * abs(tone(v, wmax, k))

draw_net <- function(p, H, wmax, k, main) {
  P <- unpack(p, H)
  yh <- if (H == 1) 0.5 else seq(0.92, 0.10, length.out = H)
  ix <- 0.05; hx <- 0.55; ox <- 1.05
  nc <- max(1.4, min(4.4, 36 / H))
  plot.new(); plot.window(c(-0.05, 1.2), c(-0.10, 1.02))
  title(main = main, cex.main = 1.05)
  for (j in 1:H)
    segments(ix, 0.5, hx, yh[j], col = wcol(P$w[j], wmax, k),
             lwd = wlwd(P$w[j], wmax, k), lend = 1)
  for (j in 1:H)
    segments(hx, yh[j], ox, 0.5, col = wcol(P$v[j], wmax, k),
             lwd = wlwd(P$v[j], wmax, k), lend = 1)
  points(ix, 0.5, pch = 21, cex = 3.2, bg = "white", lwd = 1.6)
  text(ix, 0.5, "x", font = 2)
  points(rep(hx, H), yh, pch = 21, cex = nc, lwd = 1.4,
         bg = wcol(P$cc, wmax, k))
  if (H <= 10) text(rep(hx, H), yh, 1:H, cex = 0.65)
  points(ox, 0.5, pch = 21, cex = 3.4, lwd = 1.6, bg = wcol(P$b, wmax, k))
  text(ox, 0.5, "y", font = 2)

  ## signed colour bar
  xs <- seq(0.12, 0.98, length.out = 201)
  rect(xs, -0.09, xs + diff(xs)[1], -0.035, col = DIV, border = NA)
  lab <- formatC(wmax, format = "g", digits = 2)
  text(c(0.12, 0.55, 0.98), -0.005,
       c(paste0("-", lab), "0", paste0("+", lab)), cex = 0.72)
}

## --------------------------------------------------------------------- UI
ui <- fluidPage(
  fluidRow(
    ## ---- column 1: training controls -----------------------------------
    column(
      3,
      selectInput("dgp", "True curve", choices = DGPS, selected = "teacher"),
      sliderInput("H", "Hidden neurons", min = 2, max = 16, value = 8,
                  step = 1),
      sliderInput("m", "Batch size m", min = 1, max = 400, value = 20,
                  step = 1),
      sliderInput("a", "Learning rate", min = 0.01, max = 0.60, value = 0.20,
                  step = 0.01),
      sliderInput("nep", "Epochs", min = 10, max = 300, value = 100, step = 10),
      sliderInput("ctr", "Weight colour contrast", min = 1, max = 40,
                  value = 12, step = 1),
      checkboxInput("showgd", "Overlay the full-batch run", TRUE)
    ),
    ## ---- column 2: data generation, then playback ----------------------
    column(
      2,
      div(
        style = "margin-top:10px;",
        tags$h5(tags$b("Data")),
        sliderInput("n", "Sample size n", min = 40, max = 400, value = 200,
                    step = 20),
        sliderInput("sigma", "Noise SD", min = 0.01, max = 0.40, value = 0.08,
                    step = 0.01),
        actionButton("newdata", "Regenerate a dataset",
                     class = "btn-warning",
                     style = "width:100%; margin-bottom:6px;"),

        hr(style = "border-top:2px solid #999; margin:14px 0;"),

        tags$h5(tags$b("Training")),
        actionButton("play", "Play", class = "btn-primary btn-lg",
                     style = "width:100%; margin-bottom:10px;"),
        actionButton("step", "Next", class = "btn-lg",
                     style = "width:100%; margin-bottom:10px;"),
        actionButton("back", "Prev", class = "btn-lg",
                     style = "width:100%;"),
        checkboxInput("byepoch", "Advance a whole epoch", TRUE),
        checkboxInput("autoplay", "Autoplay on change", FALSE),
        actionButton("newinit", "New initial weights",
                     style = "width:100%; margin-bottom:8px;"),
        actionButton("newseed", "New batch sequence",
                     style = "width:100%; margin-bottom:8px;"),
        helpText("Both runs share the data, the initial weights and the",
                 "number of epochs; only m differs.")
      )
    ),
    ## ---- column 3: plots -----------------------------------------------
    column(
      7,
      plotOutput("nets", height = "300px"),
      plotOutput("fit", height = "340px"),
      sliderInput("k", "Update k", min = 0, max = 1, value = 0, step = 1,
                  width = "900px")
    )
  )
)

## ----------------------------------------------------------------- server
server <- function(input, output, session) {

  MAXUP <- 4000
  dseed <- reactiveVal(1); wseed <- reactiveVal(1); bseed <- reactiveVal(1)
  playing <- reactiveVal(FALSE)

  observeEvent(input$newdata, dseed(dseed() + 1))
  observeEvent(input$newinit, wseed(wseed() + 1))
  observeEvent(input$newseed, bseed(bseed() + 1))

  observeEvent(input$n, {
    updateSliderInput(session, "m", max = input$n,
                      value = min(input$m, input$n))
  })

  DAT <- reactive(make_data(input$n, input$sigma, dseed(), input$dgp))

  runs <- reactive({
    D <- DAT(); H <- input$H
    m <- max(1, min(input$m, D$n))
    B <- ceiling(D$n / m)
    ne <- max(1, min(input$nep, floor(MAXUP / B)))
    p0 <- init_p(H, wseed())
    S <- train(p0, H, m, input$a, ne, bseed(), D)
    G <- train(p0, H, D$n, input$a, ne, bseed(), D)
    list(S = S, G = G, H = H, B = B, m = m, D = D, ne = ne,
         capped = ne < input$nep,
         wmax = max(0.2, quantile(abs(c(S$P, G$P)), 0.97, na.rm = TRUE)))
  })

  stride <- reactive(if (isTRUE(input$byepoch)) runs()$B else 1L)

  observeEvent(runs(), {
    updateSliderInput(session, "k", max = max(1, runs()$S$n - 1), value = 0)
    playing(isTRUE(input$autoplay) && runs()$S$n > 1)
  })

  observeEvent(input$play, {
    if (isTRUE(playing())) playing(FALSE)
    else {
      if (as.integer(input$k) >= runs()$S$n - 1)
        updateSliderInput(session, "k", value = 0)
      playing(TRUE)
    }
  })
  observeEvent(playing(), {
    updateActionButton(session, "play",
                       label = if (isTRUE(playing())) "Pause" else "Play")
  })
  observe({
    if (!isTRUE(playing())) return()
    kmax <- isolate(runs()$S$n) - 1
    k <- isolate(as.integer(input$k)); st <- isolate(stride())
    if (k >= kmax) { playing(FALSE); return() }
    invalidateLater(250, session)
    updateSliderInput(session, "k", value = min(k + st, kmax))
  })
  observeEvent(input$step, {
    playing(FALSE)
    updateSliderInput(session, "k",
                      value = min(as.integer(input$k) + stride(),
                                  max(0, runs()$S$n - 1)))
  })
  observeEvent(input$back, {
    playing(FALSE)
    updateSliderInput(session, "k",
                      value = max(as.integer(input$k) - stride(), 0))
  })

  cur <- reactive({
    R <- runs()
    k <- min(as.integer(input$k), R$S$n - 1) + 1L
    ep <- R$S$EP[k]
    kg <- max(1, min(R$G$n, floor(ep) + 1L))
    list(R = R, k = k, kg = kg, ep = ep)
  })

  output$nets <- renderPlot({
    C <- cur(); R <- C$R
    msz <- length(R$S$IDX[[C$k]])
    par(mfrow = c(1, if (isTRUE(input$showgd)) 2 else 1), mar = c(1, 1, 3, 1))
    draw_net(R$S$P[C$k, ], R$H, R$wmax, input$ctr,
             sprintf("m = %d (B = %d),  epoch %.2f", msz, R$B, C$ep))
    if (isTRUE(input$showgd))
      draw_net(R$G$P[C$kg, ], R$H, R$wmax, input$ctr,
               sprintf("m = %d (full batch),  epoch %d", R$D$n, C$kg - 1L))
  })

  output$fit <- renderPlot({
    C <- cur(); R <- C$R; D <- R$D
    par(mfrow = c(1, 2), mar = c(4, 4.2, 2.5, 1))

    idx <- R$S$IDX[[C$k]]
    plot(D$x, D$y, pch = 20, col = "grey75", cex = 0.9,
         xlab = "x", ylab = "y", main = "fitted curve and current batch")
    points(D$x[idx], D$y[idx], pch = 19, col = "#E8820C", cex = 1.2)
    lines(XG, gtrue(XG, D$dgp), col = "red", lty = 2, lwd = 2)
    lines(XG, fnet(R$S$P[C$k, ], XG, R$H), col = "#2E8B57", lwd = 2.5)
    if (isTRUE(input$showgd))
      lines(XG, fnet(R$G$P[C$kg, ], XG, R$H), col = "#7B3FBF", lwd = 2.5)
    legend("topright", bty = "n", cex = 0.85,
           lwd = c(2, 2.5, 2.5), lty = c(2, 1, 1),
           col = c("red", "#2E8B57", "#7B3FBF"),
           legend = c(if (identical(D$dgp, "bumps")) "true curve (bumps)"
                      else "teacher network",
                      paste0("m = ", R$m, " fit"), "full-batch fit"))

    yl <- range(c(R$S$L, if (isTRUE(input$showgd)) R$G$L), na.rm = TRUE)
    plot(R$S$EP, R$S$L, type = "l", log = "y", col = "#2E8B57", lwd = 2,
         ylim = pmax(yl, 1e-7), xlab = "epoch (one pass over all n)",
         ylab = "training loss (all n)", main = "loss per epoch")
    if (isTRUE(input$showgd))
      lines(R$G$EP, R$G$L, col = "#7B3FBF", lwd = 2)
    abline(h = 0.5 * input$sigma^2, col = "grey55", lty = 3)
    points(C$ep, R$S$L[C$k], pch = 19, col = "#E8820C", cex = 1.5)
  })

  output$info <- renderUI({
    C <- cur(); R <- C$R; H <- R$H
    P <- unpack(R$S$P[C$k, ], H)
    fmt <- function(v, d = 3) formatC(v, format = "f", digits = d)
    rows <- paste0(
      "<tr><td>h", 1:H, "</td><td align='right'>", fmt(P$w),
      "</td><td align='right'>", fmt(P$cc),
      "</td><td align='right'>", fmt(P$v), "</td></tr>", collapse = "")
    HTML(paste0(
      "<hr><b>update</b> ", C$k - 1, " of ", R$S$n - 1,
      "  (epoch ", fmt(C$ep, 2), ")<br>",
      "<b>batch size m</b> = ", length(R$S$IDX[[C$k]]), " of ", R$D$n, "<br>",
      "<b>updates per epoch</b> = ", R$B, "<br>",
      "<b>parameters</b> = ", 3 * H + 1, "<br><br>",
      "<b>loss, m = ", R$m, "</b> = ", fmt(R$S$L[C$k], 5), "<br>",
      "<b>loss, full batch</b> = ", fmt(R$G$L[C$kg], 5), "<br>",
      "<b>noise floor</b> &asymp; ", fmt(0.5 * input$sigma^2, 5), "<br>",
      if (isTRUE(R$S$diverged))
        "<b style='color:#C81E1E'>run diverged</b><br>" else "",
      if (isTRUE(R$capped))
        paste0("<i style='color:#C81E1E'>epochs capped at ", R$ne,
               "</i><br>") else "",
      "<br><table style='font-size:90%'>",
      "<tr><th>unit</th><th>w</th><th>bias</th><th>v</th></tr>",
      rows, "</table>",
      "<br><i>output bias = ", fmt(P$b), "</i>"
    ))
  })
}

shinyApp(ui, server)
```

**Shinylive App for Stochastic Gradient Descent.** Compare a mini-batch run against a full-batch run sharing the same data and starting weights.

:::
:::::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit05_MLE.html#sec-sgd-app>.

:::

### Example: SGD for Logistic Regression in R

Returning one last time to the running logistic-regression example, mini-batch SGD needs only the per-observation score already coded for Newton–Raphson, applied to a small random subset at each update:

```{r}
#| label: sgd-logistic
sgd_logistic <- function(z, r, b0, gamma0 = 0.5, t0 = 100, epochs = 30, batch = 10) {
  n <- length(z); b <- b0; t <- 0; path <- b0
  for (e in 1:epochs) {
    idx <- sample(n)
    for (s in seq(1, n, by = batch)) {
      i <- idx[s:min(s + batch - 1, n)]
      t <- t + 1
      u <- b[1] + b[2] * z[i]
      p <- plogis(u)
      grad <- c(sum(r[i] - p), sum(z[i] * (r[i] - p))) / length(i)
      b <- b + gamma0 / (1 + t / t0) * grad
      path <- rbind(path, b)
    }
  }
  list(est = b, path = path)
}

set.seed(2)
beta_true <- c(0, 1.5)                    # the parameters used to simulate data$z, data$r above
sg <- sgd_logistic(data$z, data$r, c(0, 0))
rbind(sgd = sg$est, nlm = logit_nlm$estimate, truth = beta_true)
```

### SGD Trajectory vs. Newton–Raphson

```{r}
#| label: fig-sgd-vs-newton
#| fig-cap: "SGD (fine, noisy trajectory) versus Newton-Raphson (large steps, one per full pass through the data) converging to the same logistic-regression MLE."
nr_path <- mle_logistic_nr(c(0, 0), 8, data$r, data$z)     # beta0, beta1, neg_loglike per Newton iteration
m <- 10; per_epoch <- length(data$z) / m
xs <- 0:(nrow(sg$path) - 1)
xn <- (0:(nrow(nr_path) - 1)) * per_epoch

matplot(xs, sg$path, type = "l", lty = 1, col = c(2, 4),
        ylim = range(sg$path, nr_path[, 1:2]),
        xlab = "mini-batch updates (one epoch = one full pass through the data)",
        ylab = "coefficient",
        main = expression(paste("SGD vs. Newton-Raphson: ", beta[0], " (red), ", beta[1], " (blue)")))
matlines(xn, nr_path[, 1:2], type = "b", lty = 2, pch = 19, col = c(2, 4), lwd = 2)
abline(h = logit_nlm$estimate, col = c(2, 4), lty = 3)
abline(v = seq(0, max(xs), per_epoch), col = "grey85")
legend("right", c("SGD (mini-batch)", "Newton (per full pass)", "MLE"),
       lty = c(1, 2, 3), pch = c(NA, 19, NA), bty = "n", cex = 0.85)
```

Newton needs only a handful of full-data passes to converge on this small problem ($n = 200$, $p = 2$); SGD makes many cheap updates and lands near the MLE with residual noise from its stochastic gradient. Per data pass, Newton is far more efficient here — its advantage narrows, and eventually reverses, only once a full gradient, or the $O(p^3)$ Hessian solve, becomes too expensive to compute at every iteration, which is exactly the regime where SGD earns its keep.

## Comparing All the Methods

The chapter has covered a lot of ground, from bisection on a scalar score to stochastic gradient descent on millions of parameters. The three tables below collect it into a single reference: which method needs what, which R function to reach for, and how to choose among them for a given problem.

### Comparison

| Method | Derivatives | Storage | Cost per iteration | Local rate | Robust start | Hessian for SEs |
|:---|:---|:---|:---|:---|:---|:---|
| Nelder–Mead | none | $O(p^2)$ | $O(p)$ evaluations | sublinear | good | no |
| Steepest descent | gradient | $O(p)$ | 1 gradient + line search | linear (slow) | good | no |
| Conjugate gradient | gradient | $O(p)$ | 1 gradient + line search | linear (fast) | moderate | no |
| BFGS | gradient | $O(p^2)$ | 1 gradient + $O(p^2)$ | superlinear | moderate | approximate |
| L-BFGS | gradient | $O(mp)$ | 1 gradient + $O(mp)$ | superlinear | moderate | no |
| Newton–Raphson | gradient, Hessian | $O(p^2)$ | 1 Hessian + $O(p^3)$ | quadratic | poor | exact |
| Fisher scoring / IRLS | score, expected information | $O(p^2)$ | 1 weighted LS fit | near-quadratic | fair | exact |
| Stochastic gradient | minibatch gradient | $O(p)$ | $O(\text{batch})$ | sublinear | good | no |

: Comparison of multivariate optimization methods. $p$ is the number of parameters and $m$ the L-BFGS memory length (typically 5–20). {#tbl-comparison}

### Implementations in R

| Function | Package | Methods available | Notes |
|:---|:---|:---|:---|
| `optim` | stats | Nelder–Mead, BFGS, CG, L-BFGS-B, SANN | General workhorse; supply `gr` or the gradient is differenced; `hessian = TRUE` for inference; L-BFGS-B allows box constraints |
| `nlm` | stats | Newton-type with ridging | Uses analytic gradient/Hessian if attached as attributes, otherwise numerical; returns `$hessian` |
| `nlminb` | stats | Trust-region (PORT) | Reliable on badly scaled problems; supports box constraints |
| `optimize`, `uniroot` | stats | Golden section, Brent | One dimension only |
| `glm` | stats | IRLS (Fisher scoring) | The specialised case; use `start` when separation is suspected |
| `nls` | stats | Gauss–Newton, Golub–Pereyra | Nonlinear least squares; `port` algorithm adds constraints |
| `optimx` | optimx | Unified front end to a dozen optimizers | Useful for comparing methods on the same objective |
| `nloptr` | nloptr | Large NLopt collection, with constraints | Global and derivative-free options |
| `Rsolnp`, `alabama` | — | Nonlinear equality and inequality constraints | For genuinely constrained problems |

: R interfaces to multivariate optimizers {#tbl-r-functions}

### Choosing a method

| Situation | Suggested approach |
|:---|:---|
| Hessian derivable, $p$ small, standard errors wanted | Newton–Raphson or Fisher scoring |
| GLM or a model expressible as one | `glm` (IRLS) |
| Gradient available, $p$ moderate | BFGS via `optim(method = "BFGS")` |
| Gradient available, $p$ large | L-BFGS |
| Derivatives unavailable, $p < 10$ | Nelder–Mead as a first pass, then refine |
| Objective noisy or non-smooth | Nelder–Mead or simulated annealing |
| Divergence or poor conditioning suspected | `nlm` or `nlminb`; rescale the parameters |
| Objective a sum over very many observations | Stochastic gradient descent and variants |
| Multiple local maxima suspected | Multiple random starts with any of the above |

: Practical guidance on method selection {#tbl-choosing}

The last row deserves emphasis. Every method in this chapter is a *local*
method: each converges to a stationary point near where it started, and none
can certify that the point found is the global maximum. Restarting from a
dispersed set of initial values and comparing the attained log-likelihoods is
the only general safeguard, and it should be routine for any likelihood that is
not known to be concave.
