---
title: Computer Arithmetic
editor_options:
  chunk_output_type: console
code-fold: false
format: html
---

```{r include=FALSE}
options(digits = 22)

```
```{=html}
<style>
.pv table { border-collapse: collapse; margin: 0.6em auto; font-family: monospace; }
.pv td, .pv th { border: 2px solid #555; min-width: 2em; height: 1.5em;
                 text-align: center !important; padding: 0.3em 0.3em; font-size: 0.95em; }
.pv th { border: none; font-size: 0.7em; color: #666; font-weight: normal; }
.pv td.s { background: #fbe4ec; }   /* sign */
.pv td.m { background: #e3eefb; }   /* mantissa / digits */
.pv td.e { background: #e7f6e0; }   /* exponent */
.pv td.x { background: #ffd6d6; color: #b00000; font-weight: bold; } /* lost / invalid */
.pv td.p { border: none; padding: 0; min-width: 0; font-weight: bold; } /* punctuation */
.pv .cap { font-size: 0.85em; color: #444; text-align: center !important; margin-top: -0.3em; }
</style>
```

Statistical computation is arithmetic on a machine that stores every number in a fixed number of binary digits. This chapter describes how integers and real numbers are represented, what goes wrong when a computation exceeds the capacity of that representation (overflow, underflow, and roundoff error), and the standard techniques for writing statistical code that avoids these failures: reordering sums, rewriting formulas, and working on the logarithmic scale.

## Representations of Numbers in Computer



### Integers


Every number is a string of digits in boxes, and each box has a **place value**. In base 10,

::: {.pv}
<table>
<tr><th>10³</th><th>10²</th><th>10¹</th><th>10⁰</th></tr>
<tr><td class="m">1</td><td class="m">3</td><td class="m">4</td><td class="m">8</td></tr>
</table>
<p class="cap">$1348 = 1\times10^3 + 3\times10^2 + 4\times10^1 + 8\times10^0$</p>

:::

A computer uses base 2, so each box, called a **bit**, holds 0 or 1:

::: {.pv}
<table>
<tr><th>2³</th><th>2²</th><th>2¹</th><th>2⁰</th></tr>
<tr><td class="m">1</td><td class="m">1</td><td class="m">0</td><td class="m">1</td></tr>
</table>
<p class="cap">$1101_2 = 1\times2^3 + 1\times2^2 + 0\times2^1 + 1\times2^0 = 13$</p>

:::

Everything that follows is a consequence of one fact: the number of boxes is fixed in advance.

An integer is stored in a fixed number of bits, typically 32 or 64. With $p$ bits the largest unsigned value is obtained by filling every box with 1,
$$
\sum_{i=0}^{p-1} 2^i = 2^p - 1 .
$$
A 4-bit example: the largest storable integer is $1111_2 = 15$, and $15 + 1 = 16 = 10000_2$ needs a fifth box that does not exist.

::: {.pv}
<table>
<tr><th>2⁴</th><th>2³</th><th>2²</th><th>2¹</th><th>2⁰</th></tr>
<tr><td class="x">1</td><td class="m">0</td><td class="m">0</td><td class="m">0</td><td class="m">0</td></tr>
</table>
<p class="cap">The leading 1 has nowhere to go: **integer overflow**</p>
:::

To represent negative values, one of the boxes is used as a **sign bit**, leaving $p-1$ boxes for the magnitude. With 32 bits, a signed integer $u$ is
$$
u = (-1)^{s}\sum_{i=0}^{30} d_i\,2^i, \qquad s\in\{0,1\}.
$$
The largest integer has the sign bit $0$ and the other 31 boxes all equal to $1$:

::: {.pv}
<table>
<tr><th>sign</th><th>2³⁰</th><th>2²⁹</th><th>…</th><th>2¹</th><th>2⁰</th></tr>
<tr><td class="s">+</td><td class="m">1</td><td class="m">1</td><td class="m">…</td><td class="m">1</td><td class="m">1</td></tr>
</table>
<p class="cap">$u^{\max} = \sum_{i=0}^{30} 2^i = 2^{31}-1 = 2147483647$</p>
:::

and the smallest is its negative, $u^{\min} = -(2^{31}-1)$. When an integer computation leaves this range, R does not wrap around silently, as C does; it returns `NA` with a warning.

```{r}
#| label: integer-limits
#| warning: false
.Machine$integer.max

## Check upper bounds
as.integer(2^31 - 1)                    # maximum valid integer
as.integer(2^31)                        # exceeds limit, returns NA
as.integer(2^31 - 1) + as.integer(1)    # overflows to NA

## Check lower bounds
as.integer(-(2^31 - 1))                 # minimum valid integer
as.integer(-2^31)                       # exceeds lower limit, returns NA
```

A practical consequence: products of moderately large integers overflow easily. `50000L * 50000L` is `NA` in R, while `50000 * 50000` (double precision) is fine. R converts to double automatically for most arithmetic, so integer overflow is rarely a problem in R code; it is a common one in C, Java, and database systems.

```{r}
#| label: integer-product
#| warning: false
50000L * 50000L
50000 * 50000
```

### Floating point numbers

A real number is stored in the form of scientific notation: a **sign**, a **mantissa** holding the significant digits, and an **exponent**:
$$
x = (-1)^{s} \times 1.d_1 d_2 \cdots d_p \times 2^{\,e}.
$$

::: {.pv}
<table>
<tr><th>sign</th><th colspan="4">mantissa (fixed number of digits)</th><th></th><th colspan="2">exponent</th></tr>
<tr><td class="s">+</td><td class="m">1</td><td class="p">.</td><td class="m">d₁</td><td class="m">d₂ … d<sub>p</sub></td><td class="p">×2^</td><td class="s">±</td><td class="e">e</td></tr>
</table>

:::

More precisely, a floating point number $x$ is represented by a vector of bits $(s,d_0,\ldots,d_{t-1},e_1,\ldots, e_{k-1})$ as,
$$
x = (-1)^s \sum_{i=0}^{t-1}d_i2^{-i}\times 2^{\sum_{i=0}^{k-1}e_i2^i-2^{k-1}}.
$$

### Zoom Models for Representable Floating Numbers

```{=html}
<style>
.pv td.z  { color: #bbb; border-style: dashed; }          /* alignment zero, not stored */
.pv td.x  { background: #fde0e0; color: #b22222; text-decoration: line-through; } /* digit rounded off */
.pv td.o  { background: #fde0e0; color: #b22222; font-weight: bold; }            /* no room: overflow/underflow */
.pv td.op { border: none; background: none; font-weight: bold; }
.pv td.nb { border: none; background: none; }
.pv tr.sum td:not(.op):not(.nb) { border-top: 3px solid #333; }
</style>
```

A floating point number is a **ruler** and a **zoom setting**. The mantissa is a ruler with a fixed number of equally spaced marks; the exponent stretches it by $10^{e}$ over $[10^{e}, 10^{e+1})$. Stretching adds no marks, so the marks spread apart as the zoom grows.

#### A toy decimal computer

Each number is stored in six boxes: a sign, a two-digit mantissa $d_1.d_2$ ($d_1\in\{1,\ldots,9\}$, $d_2\in\{0,\ldots,9\}$), and an exponent made of a sign and one decimal digit, so $e\in\{-9,\ldots,0,\ldots,+9\}$.

::: {.pv}
<table>
<tr><th>sign</th><th>1</th><th></th><th>10⁻¹</th><th></th><th>±</th><th>exp</th></tr>
<tr><td class="s">+</td><td class="m">3</td><td class="p">.</td><td class="m">1</td><td class="p">×10^</td><td class="s">+</td><td class="e">0</td></tr>
</table>
<p class="cap">$+3.1\times10^{+0}$, the closest storable value to $\pi$</p>
:::

Each of the $19$ zoom levels is a ruler of $90$ marks. At the lowest zoom $e=-9$ the leading digit may also be $0$ (**subnormal** numbers), which fills the gap between $0$ and $1.0\times10^{-9}$.

| zoom $e$ | interval | storable numbers | gap |
|:--:|:--|:--|:--:|
| $-9$ (subnormal) | $(0, 10^{-9})$ | $0.1\times10^{-9}, \ldots, 0.9\times10^{-9}$ | $10^{-10}$ |
| $-9$ | $[10^{-9}, 10^{-8})$ | $1.0\times10^{-9}, 1.1\times10^{-9}, \ldots, 9.9\times10^{-9}$ | $10^{-10}$ |
| $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ |
| $-1$ | $[0.1, 1)$ | $0.10, 0.11, \ldots, 0.99$ | $0.01$ |
| $0$ | $[1, 10)$ | $1.0, 1.1, \ldots, 9.9$ | $0.1$ |
| $+1$ | $[10, 100)$ | $10, 11, \ldots, 99$ | $1$ |
| $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ |
| $+9$ | $[10^{9}, 10^{10})$ | $1.0\times10^{9}, 1.1\times10^{9}, \ldots, 9.9\times10^{9}$ | $10^{8}$ |

With both signs, the $9$ subnormals per sign, and a reserved code for $0$, the machine stores exactly $2\times(19\times90+9)+1=3439$ numbers. The $90$ normal mantissas are listed below; multiplying them by $10^{e}$, $e=-9,\ldots,9$, gives every positive normal number.

```{r}
#| label: toy-all-numbers
m <- outer(1:9, 0:9, function(a, b) sprintf("%.1f",a + b / 10))
dimnames(m) <- list(d1 = 1:9, d2 = 0:9)
m
```

```{r}
#| label: fig-toy-lines
#| echo: false
#| fig-width: 9
#| fig-height: 3.6
#| fig-cap: 'Positive numbers at zoom $e=0$ (top, gap $0.1$) and $e=+1$ (bottom, gap $1$). Grey bands hold the crowded marks of all lower zooms ($e<0$, and $e\le 0$ in the bottom panel); overflow occurs only far to the right, beyond $9.9\times10^{9}$.'
par(mfrow = c(2, 1), mar = c(2.5, 1, 1.5, 1))
m0 <- (10:99) / 10; m1 <- 10:99
plot(NA, xlim = c(0, 10.5), ylim = c(0, 1), axes = FALSE, xlab = "", ylab = "",
     main = "e = 0   (gap 0.1)", cex.main = 0.9)
rect(0, 0.2, 1, 0.8, col = "gray85", border = NA)
segments(m0, 0.3, m0, 0.7, col = "steelblue")
points(0, 0.5, pch = 19)
axis(1, at = 0:10)
plot(NA, xlim = c(0, 105), ylim = c(0, 1), axes = FALSE, xlab = "", ylab = "",
     main = "e = +1   (gap 1)", cex.main = 0.9)
rect(0, 0.2, 10, 0.8, col = "gray85", border = NA)
segments(m1, 0.3, m1, 0.7, col = "darkorange3")
points(0, 0.5, pch = 19)
axis(1, at = seq(0, 100, 10))
```

#### Rounding error and machine epsilon

A real number between two marks is stored as the nearer mark. The three numbers below share the same mantissa boxes and differ only in the exponent, so the absolute error scales with the zoom while the relative error does not.

::: {.pv}
<table>
<tr><th>$x$</th><th>sign</th><th>1</th><th></th><th>10⁻¹</th><th></th><th>±</th><th>exp</th><th>stored</th><th>error</th></tr>
<tr><td class="nb">$0.234$</td><td class="s">+</td><td class="m">2</td><td class="p">.</td><td class="m">3</td><td class="p">×10^</td><td class="s">−</td><td class="e">1</td><td class="nb">$0.23$</td><td class="nb">$0.004$ (1.7%)</td></tr>
<tr><td class="nb">$2.34$</td><td class="s">+</td><td class="m">2</td><td class="p">.</td><td class="m">3</td><td class="p">×10^</td><td class="s">+</td><td class="e">0</td><td class="nb">$2.3$</td><td class="nb">$0.04$ (1.7%)</td></tr>
<tr><td class="nb">$23.4$</td><td class="s">+</td><td class="m">2</td><td class="p">.</td><td class="m">3</td><td class="p">×10^</td><td class="s">+</td><td class="e">1</td><td class="nb">$23$</td><td class="nb">$0.4$ (1.7%)</td></tr>
</table>
:::

The same rounding occurs in arithmetic. To add, the two numbers are aligned by place value; solid boxes are stored digits, dashed boxes are alignment zeros, and the result keeps only the two boxes starting at its leading digit, so digits to the right (red) are rounded off.

:::: {.columns}

::: {.column width="33%"}
::: {.pv}
<table>
<tr><th></th><th>10⁰</th><th></th><th>10⁻¹</th><th>10⁻²</th></tr>
<tr><td class="op"></td><td class="m">1</td><td class="p">.</td><td class="m">0</td><td class="nb"></td></tr>
<tr><td class="op">+</td><td class="z">0</td><td class="p">.</td><td class="m">1</td><td class="m">0</td></tr>
<tr class="sum"><td class="op"></td><td class="m">1</td><td class="p">.</td><td class="m">1</td><td class="x">0</td></tr>
</table>
<p class="cap">$1.0 + 0.10 = 1.1 \ne 1.0$</p>
:::
:::

::: {.column width="33%"}
::: {.pv}
<table>
<tr><th></th><th>10⁰</th><th></th><th>10⁻¹</th><th>10⁻²</th></tr>
<tr><td class="op"></td><td class="m">1</td><td class="p">.</td><td class="m">0</td><td class="nb"></td></tr>
<tr><td class="op">+</td><td class="z">0</td><td class="p">.</td><td class="m">1</td><td class="m">4</td></tr>
<tr class="sum"><td class="op"></td><td class="m">1</td><td class="p">.</td><td class="m">1</td><td class="x">4</td></tr>
</table>
<p class="cap">$1.0 + 0.14 \to 1.1$</p>
:::
:::

::: {.column width="33%"}
::: {.pv}
<table>
<tr><th></th><th>10¹</th><th>10⁰</th><th></th><th>10⁻¹</th><th>10⁻²</th></tr>
<tr><td class="op"></td><td class="m">1</td><td class="m">0</td><td class="p">.</td><td class="nb"></td><td class="nb"></td></tr>
<tr><td class="op">+</td><td class="z">0</td><td class="z">0</td><td class="p">.</td><td class="m">4</td><td class="m">0</td></tr>
<tr class="sum"><td class="op"></td><td class="m">1</td><td class="m">0</td><td class="p">.</td><td class="x">4</td><td class="x">0</td></tr>
</table>
<p class="cap">$10 + 0.40 \to 10$</p>
:::
:::

::::

The **machine epsilon** $\epsilon$ is the gap next to $1$, the smallest increment that the second mantissa box can register when added to $1$; here $\epsilon = 0.1$ (left). Smaller increments fall into the red box: the digit $4$ is lost (middle), and at zoom $e=+1$ the whole addend $0.40$ is lost (right). Relative to the magnitude, the resolution is about $\epsilon$ at every zoom.

#### Largest and smallest numbers

::: {.pv}
<table>
<tr><th></th><th>sign</th><th>1</th><th></th><th>10⁻¹</th><th></th><th>±</th><th>exp</th><th>value</th></tr>
<tr><td class="nb">largest</td><td class="s">+</td><td class="m">9</td><td class="p">.</td><td class="m">9</td><td class="p">×10^</td><td class="s">+</td><td class="e">9</td><td class="nb">$9.9\times10^{9}$</td></tr>
<tr><td class="nb">smallest positive</td><td class="s">+</td><td class="m">0</td><td class="p">.</td><td class="m">1</td><td class="p">×10^</td><td class="s">−</td><td class="e">9</td><td class="nb">$0.1\times10^{-9}$</td></tr>
<tr><td class="nb">largest negative</td><td class="s">−</td><td class="m">0</td><td class="p">.</td><td class="m">1</td><td class="p">×10^</td><td class="s">−</td><td class="e">9</td><td class="nb">$-0.1\times10^{-9}$</td></tr>
<tr><td class="nb">smallest</td><td class="s">−</td><td class="m">9</td><td class="p">.</td><td class="m">9</td><td class="p">×10^</td><td class="s">+</td><td class="e">9</td><td class="nb">$-9.9\times10^{9}$</td></tr>
</table>
<p class="cap">Largest mantissa with the largest exponent; smallest subnormal mantissa with the smallest exponent</p>
:::

**Overflow.** A result with magnitude at least $9.95\times10^{9}$ needs the exponent $+10$, which the single exponent digit cannot hold, so it is returned as $\pm\infty$:

::: {.pv}
<table>
<tr><th></th><th>10¹⁰</th><th>10⁹</th><th>10⁸</th></tr>
<tr><td class="op"></td><td class="nb"></td><td class="m">6</td><td class="m">0</td></tr>
<tr><td class="op">+</td><td class="nb"></td><td class="m">5</td><td class="m">0</td></tr>
<tr class="sum"><td class="op"></td><td class="o">1</td><td class="m">1</td><td class="x">0</td></tr>
</table>
<p class="cap">$6.0\times10^{9} + 5.0\times10^{9} = 1.1\times10^{+10} \to \infty$</p>
:::

::: {.pv}
<table>
<tr><th>sign</th><th>1</th><th></th><th>10⁻¹</th><th></th><th>±</th><th>exp</th></tr>
<tr><td class="s">+</td><td class="m">1</td><td class="p">.</td><td class="m">1</td><td class="p">×10^</td><td class="s">+</td><td class="o">10</td></tr>
</table>
<p class="cap">The exponent $10$ needs two digits but has one box: overflow</p>
:::

**Underflow.** A result with magnitude below $0.05\times10^{-9}$ has no box pattern and is returned as $0$:

::: {.pv}
<table>
<tr><th>sign</th><th>1</th><th></th><th>10⁻¹</th><th>10⁻²</th><th></th><th>±</th><th>exp</th></tr>
<tr><td class="s">+</td><td class="m">0</td><td class="p">.</td><td class="x">0</td><td class="o">1</td><td class="p">×10^</td><td class="s">−</td><td class="e">9</td></tr>
</table>
<p class="cap">$(0.1\times10^{-9})\times0.1 = 0.01\times10^{-9} \to 0$: the digit $1$ falls outside the mantissa</p>
:::

Every later operation treats the underflowed result as exactly zero.

#### Doubles

A double uses the same boxes in base $2$: $1$ sign bit, $11$ exponent bits, and $52$ mantissa bits after an implicit leading $1$. The exponent is stored with a bias of $1023$, so the codes $1,\ldots,2046$ give $e\in[-1022, 1023]$; the code $0$ is reserved for $0$ and the subnormals, and the code $2047$ for `Inf` and `NaN`.

$$
\newcommand{\bx}[1]{\boxed{\mathtt{#1}}\!}
\begin{array}{rccccccl}
 & \text{sign} & \text{mantissa} & & & \text{exponent} & & \\[4pt]
1 & \bx{+} & {\color{gray}1.}\overbrace{\bx{0}\bx{0}\bx{0}\cdots\bx{0}\bx{0}\bx{0}}^{52\text{ bits}} & & \times2^\wedge & \overbrace{\bx{+}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}}^{11\text{ bits}} & & \\
+\;\epsilon & \bx{+} & {\color{gray}0.}\bx{0}\bx{0}\bx{0}\cdots\bx{0}\bx{0}\bx{1} & & \times2^\wedge & \bx{+}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0} & & = 2^{-52}\\ \hline
1+\epsilon & \bx{+} & {\color{gray}1.}\bx{0}\bx{0}\bx{0}\cdots\bx{0}\bx{0}\bx{1} & & \times2^\wedge & \bx{+}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0} & & \ne 1\\[20pt]
1 & \bx{+} & {\color{gray}1.}\bx{0}\bx{0}\bx{0}\cdots\bx{0}\bx{0}\bx{0} & & \times2^\wedge & \bx{+}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0} & & \\
+\;\epsilon/2 & \bx{+} & {\color{gray}0.}\bx{0}\bx{0}\bx{0}\cdots\bx{0}\bx{0}\bx{0} & {\color{red}\boxed{\mathtt{1}}} & \times2^\wedge & \bx{+}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0} & & = 2^{-53}\\ \hline
1+\epsilon/2 & \bx{+} & {\color{gray}1.}\bx{0}\bx{0}\bx{0}\cdots\bx{0}\bx{0}\bx{0} & {\color{red}\boxed{\mathtt{1}}} & \times2^\wedge & \bx{+}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0}\bx{0} & & \to 1
\end{array}
$$

The **machine epsilon** is the smallest number that changes $1$ when added to it: the place value of the last mantissa box at $e=0$, $\epsilon = 0.0\cdots01\times2^{0} = 2^{-52}\approx 2.220446\times10^{-16}$. Adding $\epsilon$ sets that box (top). Adding $\epsilon/2$ needs a $53$rd box (red), which does not exist, so the sum rounds back to $1$ (bottom; the tie is rounded to the even mantissa).

$$
\begin{array}{rcccl}
 & \text{mantissa} & & \text{exponent} & \text{value}\\[6pt]
\begin{array}{r}\text{largest}\end{array}
  & {\color{gray}1.}\underbrace{\bx{1}\bx{1}\bx{1}\cdots\bx{1}\bx{1}\bx{1}}_{52\text{ bits}}
  & \times 2^\wedge
  & \underbrace{\bx{+}\bx{1}\bx{1}\bx{1}\bx{1}\bx{1}\bx{1}\bx{1}\bx{1}\bx{1}\bx{1}}_{11\text{ bits}}
  & \begin{array}{l}(2-2^{-52})\times2^{1023}\\\approx 1.797693\times10^{308}\end{array}\\[16pt]
\begin{array}{r}\text{smallest}\\\text{normal}\end{array}
  & {\color{gray}1.}\underbrace{\bx{0}\bx{0}\bx{0}\cdots\bx{0}\bx{0}\bx{0}}_{52\text{ bits}}
  & \times 2^\wedge
  & \underbrace{\bx{-}\bx{1}\bx{1}\bx{1}\bx{1}\bx{1}\bx{1}\bx{1}\bx{1}\bx{1}\bx{0}}_{11\text{ bits}}
  & \begin{array}{l}2^{-1022}\\\approx 2.225074\times10^{-308}\end{array}\\[16pt]
\begin{array}{r}\text{smallest}\\\text{positive}\end{array}
  & {\color{gray}0.}\underbrace{\bx{0}\bx{0}\bx{0}\cdots\bx{0}\bx{0}\bx{1}}_{52\text{ bits}}
  & \times 2^\wedge
  & \underbrace{\bx{-}\bx{1}\bx{1}\bx{1}\bx{1}\bx{1}\bx{1}\bx{1}\bx{1}\bx{1}\bx{0}}_{11\text{ bits}}
  & \begin{array}{l}2^{-1074}\\\approx 4.940656\times10^{-324}\end{array}
\end{array}
$$

For readability, the picture shows the exponent as a sign box and $10$ magnitude bits; the hardware stores the same $11$ bits as the unsigned code $e+1023$. The grey leading digit is not stored: it is $1$ for normal doubles and $0$ for subnormals, which occur only at $e=-1022$.

- The largest double has a full mantissa and the largest exponent, $+1023$.
- The smallest normal double has mantissa $1.0$ and the smallest exponent, $-1022$.
- The smallest positive double is the subnormal with only the last mantissa box set at $e=-1022$, i.e. $2^{-52}\times2^{-1022}=2^{-1074}$.

- The **machine epsilon** $\epsilon=2^{-52}$ is the gap next to $1$, set by the last mantissa bit. Since the gap in $[2^e, 2^{e+1})$ is $2^{e-52}$, $\epsilon$ is also the relative resolution at every magnitude: a double carries about $15$–$16$ significant decimal digits.
- The **largest double** has all mantissa bits and all usable exponent bits set to $1$. Results beyond it overflow to `Inf`.
- The **smallest normal double** has the smallest usable exponent and mantissa $1.0$. Below it, subnormals fill the gap down to $2^{-1074}$ with evenly spaced but less precise values; results below half of $2^{-1074}$ underflow to $0$.

| quantity | value | R | relevance in computing |
|:--|:--|:--|:--|
| machine epsilon | $2^{-52}\approx2.2\times10^{-16}$ | `.Machine$double.eps` | Doubles cannot be compared exactly (`0.1 + 0.2 != 0.3`); convergence and equality tests use tolerances such as $\sqrt{\epsilon}\approx1.5\times10^{-8}$ (`all.equal`), which is also the usual step size for numerical derivatives |
| largest double | $\approx1.80\times10^{308}$ | `.Machine$double.xmax` | `exp(x)` overflows for $x>709.8$; products of large terms such as likelihood ratios or $n!$ must be computed on the log scale |
| smallest normal double | $2^{-1022}\approx2.2\times10^{-308}$ | `.Machine$double.xmin` | `exp(x)` leaves the normal range for $x<-708.4$; a product of many densities falls here, which is why likelihoods are computed as log-likelihoods and sums of exponentials via log-sum-exp |
| smallest subnormal double | $2^{-1074}\approx4.9\times10^{-324}$ | `2^-1074` | Last value before underflow to $0$; subnormals keep few significant bits, so ratios of such values are unreliable |
| largest consecutive integer | $2^{53}\approx9.0\times10^{15}$ | `2^53` | Beyond it the gap exceeds $1$ and integers are skipped; R's `integer` type stops even earlier at $2^{31}-1$, so larger identifiers must be stored as character strings or `bit64::integer64` |

#### Verify in R

```{r}
#| label: double-limits
c(eps = .Machine$double.eps, xmax = .Machine$double.xmax,
  xmin = .Machine$double.xmin, subnormal = 2^-1074, maxint = 2^53)
```

These are the decimal values in the boxed displays above.

- `eps` $=2^{-52}$ is the place value of the last mantissa box at $e=0$, the $\epsilon$ added to $1$ in the vertical addition.
- `xmax` is the "largest" row: a full mantissa with exponent $+1023$.
- `xmin` is the "smallest normal" row: mantissa $1.0$ with exponent $-1022$.
- `subnormal` is the "smallest positive" row: only the last mantissa box set, with leading digit $0$ and exponent $-1022$.
- `maxint` $=2^{53}$ is where the last mantissa box becomes worth $2$ (see `spacing-demo` below).

```{r}
#| label: machine-eps
1 + 2^-52 == 1
1 + 2^-53 == 1                       # half an epsilon is lost
0.1 + 0.2 == 0.3                     # rounding error
isTRUE(all.equal(0.1 + 0.2, 0.3))    # test with tolerance sqrt(eps)
```

- `1 + 2^-52 == 1` is `FALSE`: the top vertical addition, where $\epsilon$ sets the last mantissa box and $1+\epsilon$ is a different double.
- `1 + 2^-53 == 1` is `TRUE`: the bottom vertical addition, where $\epsilon/2$ needs the nonexistent $53$rd box (red) and the sum rounds back to $1$.
- `0.1 + 0.2 == 0.3` is `FALSE`: $0.1$, $0.2$ and $0.3$ have infinitely repeating binary mantissas, so each is cut at the $52$nd box, and the rounded sum lands on a different mark from the rounded $0.3$.
- `all.equal(...)` is `TRUE`: it tests equality up to a relative tolerance $\sqrt{\epsilon}\approx1.5\times10^{-8}$, the safe way to compare doubles.

```{r}
#| label: over-underflow
2 * .Machine$double.xmax             # overflow: Inf
exp(710)                             # overflow: Inf
2^-1075                              # underflow: 0
prod(rep(0.01, 200))                 # 1e-400 underflows to 0
sum(log(rep(0.01, 200)))             # the log scale is safe
```

- `2 * .Machine$double.xmax` is `Inf`: doubling the "largest" row needs exponent $+1024$, whose code is reserved for `Inf`, as in the toy $6.0\times10^{9}+5.0\times10^{9}$.
- `exp(710)` is `Inf`: $e^{710}$ exceeds `xmax` $\approx e^{709.78}$.
- `2^-1075` is `0`: it is half the "smallest positive" row and would need a box beyond the last one; the tie rounds to the even pattern, which is $0$.
- `prod(rep(0.01, 200))` is `0`: the true value $10^{-400}$ lies below $2^{-1074}\approx4.9\times10^{-324}$. The same happens to a likelihood built as a product of many densities.
- `sum(log(rep(0.01, 200)))` is $-921.03$: on the log scale the product becomes a sum of moderate numbers, far from both limits.

```{r}
#| label: spacing-demo
2^52 + 1 == 2^52                     # gap 1: still distinguishable
2^53 + 1 == 2^53                     # gap 2: the +1 is lost
format(c(2^53, 2^53 + 1, 2^53 + 2), digits = 17)
.Machine$integer.max                 # 2^31 - 1
```

- `2^52 + 1 == 2^52` is `FALSE`: at $e=52$ the last mantissa box is worth $2^{52-52}=1$, so adding $1$ sets it, exactly as $\epsilon$ does at $e=0$.
- `2^53 + 1 == 2^53` is `TRUE`: at $e=53$ the last box is worth $2$, so adding $1$ needs a $53$rd box. This is the $1+\epsilon/2$ addition scaled by $2^{53}$.
- `format(...)` prints `9007199254740992` twice and then `9007199254740994`: above $2^{53}$ only even integers are marks on the ruler.
- `.Machine$integer.max` is $2^{31}-1 = 2147483647$: R's `integer` type is $32$ bits, so integers beyond it, and beyond $2^{53}$ for doubles, must be stored as character strings or `bit64::integer64`.

## Methods for Mitigating Roundoff Error

### Roundoff error

The stored result of $a+b$ (or $a\times b$, $a/b$, …) is the **nearest representable number** to the exact result; digits beyond the mantissa are rounded away, as in the vertical additions above. Each single operation therefore has relative error at most about $\epsilon$, which is harmless on its own. The concern is when these errors **accumulate** over many operations or are **amplified** by the structure of a formula. Two patterns do the amplifying:

- adding a **large + small** number, where the small number is partly or wholly rounded away;

- subtracting **large − large** nearly equal numbers, where the leading digits cancel and only the roundoff errors survive (**catastrophic cancellation**).

### Computing sums: avoiding large + small

Because the mantissa has finite capacity, a term smaller than half the last box of the large number is lost entirely, as in the addition $1+\epsilon/2 \to 1$ shown earlier.

Consider $1.0 + 0.04 + 0.04$ in the toy system, whose exact value is $1.08$. Evaluated left to right, $(1.0 + 0.04) + 0.04$, each addition rounds back to $1.0$, so both small terms are lost. Evaluated as $1.0 + (0.04 + 0.04) = 1.0 + 0.08$, the intermediate sum is exact, and the final addition rounds once, to $1.1$. Adding the small terms among themselves first commits only **one** roundoff error, whereas adding them one at a time to the large number commits one per term, and these accumulate.

The following R demonstration makes the same point at machine precision. `f2` attempts to add $2^{-53}$ (half an epsilon) $2^{13}$ times, but because $1 + 2^{-53} = 1$ at every single step, the final sum is simply $1$ instead of $1 + 2^{-40}$. Grouping the small numbers and adding them together before adding them to the large number (`f6s`) recovers the correct answer.


The following R demonstration makes the same point at machine precision. `f2` attempts to add $2^{-53}$ (half an epsilon) $2^{13}$ times, but because $1 + 2^{-53} = 1$ at every single step, the final sum is simply $1$ instead of $1 + 2^{-40}$. Grouping the small numbers and adding them together before adding them to the large number (`f6s`) recovers the correct answer.

```{r}
#| label: sum-large-small
## the target value
f0 <- 1 + 2^(-40)
f0

## adding 2^-52 (one epsilon) 2^12 times works: each addition changes the sum
f1 <- 1; i <- 1
while (i <= 2^12) { f1 <- f1 + 2^(-52); i <- i + 1 }
f1

## adding 2^-53 (half an epsilon) 2^13 times fails: 1 + 2^-53 rounds to 1 every time
f2 <- 1; i <- 1
while (i <= 2^13) { f2 <- f2 + 2^(-53); i <- i + 1 }
f2

## this difference is not due to number display in R
f0 == f1
f0 == f2

## R's built-in sum() uses pairwise summation and extended precision
f3 <- sum(c(1, rep(2^(-53), 2^13))); f3
f1 == f3
f4 <- sum(c(1, rep(2^(-54), 2^14))); f4
f0 == f4

## the fix: accumulate the small numbers first, then add to 1
f6s <- 0; i <- 1
while (i <= 2^14) { f6s <- f6s + 2^(-54); i <- i + 1 }
f6s
f6 <- f6s + 1; f6
f0 == f6
```
#### Rules for accurate summation {-}
- Add numbers of similar magnitude together; when summing terms of very different sizes, add the smallest first (sort ascending).

- **Pairwise summation**: sum the two halves of the vector recursively, so that partial sums stay of similar size. R's `sum()` uses this together with extended-precision accumulation, which is why `sum()` beat the naive loop above.
- **Kahan compensated summation**: keep track of the part rounded away at each step and add it back at the next step.

```{r}
#| label: summation-methods
x <- c(1, rep(1e-16, 1e4))                 # exact sum is 1 + 1e-12

naive_sum <- function(x) { s <- 0; for (v in x) s <- s + v; s }

kahan_sum <- function(x) {
  s <- 0; c <- 0                            # c holds the running roundoff error
  for (v in x) { y <- v - c; t <- s + y; c <- (t - s) - y; s <- t }
  s }

c(naive   = naive_sum(x) - 1,
  sorted  = naive_sum(sort(x)) - 1,
  kahan   = kahan_sum(x) - 1,
  builtin = sum(x) - 1,
  exact   = 1e-12)
```

The naive loop loses every one of the $10^4$ small terms; sorting recovers most of the answer, and Kahan summation and R's `sum()` recover it to full precision.

We can use `log()` to translate the 52-bit limit into other bases, which tells us at roughly what decimal place precision is lost (about 15 to 16 significant decimal digits).

```{r}
#| label: eps-other-bases
log(2^(-52))               # base e: about -36
1 + exp(-35)
1 + exp(-37)               # evaluates to exactly 1

log(2^(-52), base = 10)    # base 10: about -15.7
1 + 1e-15 == 1
1 + 1e-16 == 1
```

### Computing $e^x$: avoiding large − large

The Taylor series,
$$
e^x \approx \sum_{k=0}^{K}\frac{x^k}{k!} = 1 + x + \frac{x^2}{2!} + \frac{x^3}{3!} + \cdots,
$$
works well for $x > 0$ but fails for large negative $x$. For $x = -20$ the terms alternate in sign and are huge relative to the answer $e^{-20}\approx 2\times10^{-9}$:

| $k$ | 0 | 1 | 2 | … | 20 | … |
|:--|:--:|:--:|:--:|:--:|:--:|:--:|
| $x^k/k!$ | $1$ | $-20$ | $200$ | … | $4.3\times10^{7}$ | … |

Each term carries a roundoff error of relative size $\epsilon \approx 10^{-16}$, i.e. absolute size $\approx 4\times10^{-9}$ for the largest terms, which is **as large as the answer itself**. Subtracting nearly equal large numbers leaves only their errors: catastrophic cancellation.

The function `fexp` below implements the series with a stopping rule based on the relative size of the current term. With `debug = TRUE` it returns the internal steps rather than the final answer.

```{r}
#| label: fexp-naive
#| message: false
library(knitr)

fexp <- function(x, debug = FALSE) {
  i <- 0; expx <- 1; u <- 1
  if (debug) debug_list <- list()
  while (abs(u) > 1e-20 * abs(expx)) {
    i <- i + 1
    u <- u * x / i
    expx <- expx + u
    if (debug) debug_list[[i]] <- c(Step = i, Term_u = u, exp_x = expx)
  }
  if (debug) return(do.call(rbind, debug_list))
  expx
}

test_vals <- c(10, 20, 60, -1, -10, -20, -50)
res1 <- data.frame(x = test_vals, Built_in_Exp = exp(test_vals), Taylor_fexp = sapply(test_vals, fexp))
kable(res1, caption = "Built-in exp() versus the naive Taylor series fexp()")
```

For positive $x$ the series is accurate to all printed digits. For $x = -10$ the answer is already off in the fourth digit, for $x = -20$ it is wrong in the first digit, and `fexp(-50)` yields a completely wrong, negative number. Running it in debug mode shows the massive terms cancelling each other out.

```{r}
#| label: fexp-debug
debug_matrix <- fexp(-50, debug = TRUE)
debug_summary <- rbind(
  head(debug_matrix, 10),
  matrix(NA, nrow = 1, ncol = 3, dimnames = list("...", colnames(debug_matrix))),
  tail(debug_matrix, 10))
kable(debug_summary, caption = "Debug trace for fexp(-50) showing catastrophic cancellation")
```

The largest terms reach about $10^{20}$; their roundoff errors are of order $10^{4}$, and the true answer $e^{-50}\approx 2\times10^{-22}$ is invisible beneath them.

**Fix by reformulating.** For $x>0$ all terms are positive, so no cancellation occurs. Therefore compute $e^{|x|}$ and invert the result if $x$ was negative:
$$
e^{-x} = \frac{1}{e^{x}}, \qquad x > 0 .
$$

```{r}
#| label: fexp-robust
fexp2 <- function(x) {
  xa <- abs(x); i <- 0; expx <- 1; u <- 1
  while (u > 1e-20 * expx) { i <- i + 1; u <- u * xa / i; expx <- expx + u }
  if (x >= 0) expx else 1 / expx
}

test_vals_neg <- c(-10, -20, -50)
res2 <- data.frame(x = test_vals_neg, Built_in_Exp = exp(test_vals_neg), Robust_fexp2 = sapply(test_vals_neg, fexp2))
kable(res2, caption = "Robust fexp2() for negative arguments")
```

The general principle: rearrange a formula so that quantities of similar magnitude and **the same sign** are combined, and avoid computing a small result as the difference of two large numbers.

**A second example of the same principle.** The function $f(x) = e^x - 1$ is needed in interest-rate and survival calculations for small $x$. Computing `exp(x) - 1` directly subtracts two numbers near 1, and for $x = 10^{-10}$ loses six of the sixteen digits; `expm1(x)` uses a series for small $x$ and is exact to machine precision. The same holds for $\log(1+x)$ and `log1p(x)`.

```{r}
#| label: expm1-demo
x <- 1e-10
c(naive = exp(x) - 1, expm1 = expm1(x), series = x + x^2 / 2)
x <- 1e-20
c(naive = log(1 + x), log1p = log1p(x))
```

### Computing variance: avoiding large − large

Two algebraically identical formulas for the sum of squared deviations:
$$
\underbrace{\sum_{i=1}^{n}(x_i - \overline{x})^2}_{\text{two-pass}}
\;=\;
\underbrace{\sum_{i=1}^{n} x_i^2 - \frac{\left(\sum_i x_i\right)^2}{n}}_{\text{one-pass}}.
$$

The one-pass formula needs only $\sum x_i$ and $\sum x_i^2$, computed in a single sweep of the data, which made it attractive for hand calculators and for streaming data. But when the $x_i$ are large and close together, $\sum x_i^2$ and $(\sum x_i)^2/n$ are both huge and nearly equal, and the subtraction is large − large.

Take $x = (1, 2, 3)$: the deviations are $(-1, 0, 1)$, so $\sum(x_i-\overline x)^2 = 2$. Now shift by $M = 10^{8}$, giving $x = (M+1, M+2, M+3)$. The deviations and the answer are unchanged, but the one-pass formula computes
$$
\sum x_i^2 \approx 3\times10^{16}, \qquad \frac{(\sum x_i)^2}{n} \approx 3\times10^{16}, \qquad \text{difference} = 2.
$$
With 16 significant digits, $3\times10^{16}$ is stored with a gap of about $4$ between neighbouring representable numbers, so the true answer $2$ lives entirely in the lost digits.

::: {.pv}
<table>
<tr><th colspan="16">16 mantissa digits</th><th>lost</th></tr>
<tr><td class="m">3</td><td class="p">.</td><td class="m">0</td><td class="m">0</td><td class="m">0</td><td class="m">0</td><td class="m">0</td><td class="m">0</td><td class="m">1</td><td class="m">2</td><td class="m">0</td><td class="m">0</td><td class="m">0</td><td class="m">0</td><td class="m">0</td><td class="m">1</td><td class="x">4</td><td class="p">×10^16</td></tr>
</table>

:::

```{r}
#| label: variance-formulas
## one-pass (computational) formula
var1 <- function(x) { n <- length(x); (sum(x^2) - sum(x)^2 / n) / (n - 1) }

## two-pass formula (centering on the mean)
var2 <- function(x) { n <- length(x); sum((x - mean(x))^2) / (n - 1) }

## small data works fine for both
x <- c(1, 2, 3)
c(var1(x), var2(x), var(x))

## a large shift exposes catastrophic cancellation in var1
x <- c(1, 2, 3) + 1e8
c(var1(x), var2(x), var(x))
x <- c(1, 2, 3) + 1e10
c(var1(x), var2(x), var(x))
```

The failure sets in gradually. The error of the one-pass formula is of order $\epsilon\,M^2 n$, so it becomes visible once $M^2\epsilon$ is comparable to the true variance, i.e. around $M\approx 10^{8}$ for a variance of 1. The following scan shows the one-pass result drifting away from 1 and eventually becoming negative, which is impossible for a variance.

```{r}
#| label: variance-scan
M <- 10^(0:10)
res <- t(sapply(M, function(m) { x <- c(1, 2, 3) + m; c(M = m, one_pass = var1(x), two_pass = var2(x)) }))
kable(res, caption = "One-pass versus two-pass variance as the data are shifted by M", digits = 6)
```
#### Fixes {-}
- **Two-pass**: compute $\overline x$ first, then $\sum (x_i - \overline x)^2$. The deviations are small and of similar size, so no cancellation occurs. This is what `var()` does.

- **Shifted one-pass**: subtract any convenient constant $c \approx \overline x$, for instance $c = x_1$, before applying the one-pass formula. The identity
$$
\sum (x_i-\overline x)^2 = \sum (x_i - c)^2 - \frac{\left(\sum (x_i - c)\right)^2}{n},
$$
holds for every $c$, and with $c$ near the mean the two terms on the right are no longer huge.

- **Welford's online algorithm** updates the running mean $\overline x_k$ and running sum of squares $S_k = \sum_{i\le k}(x_i - \overline x_k)^2$ one observation at a time,
$$
\overline x_k = \overline x_{k-1} + \frac{x_k - \overline x_{k-1}}{k}, \qquad S_k = S_{k-1} + (x_k - \overline x_{k-1})(x_k - \overline x_k),
$$
and is the method of choice for streaming data, where a two-pass computation is impossible.

```{r}
#| label: variance-fixes
var_shifted <- function(x) { n <- length(x); d <- x - x[1]; (sum(d^2) - sum(d)^2 / n) / (n - 1) }

var_welford <- function(x) {
  m <- 0; S <- 0
  for (k in seq_along(x)) { m_old <- m; m <- m + (x[k] - m) / k; S <- S + (x[k] - m_old) * (x[k] - m) }
  S / (length(x) - 1) }

x <- c(1, 2, 3) + 1e10
c(one_pass = var1(x), shifted = var_shifted(x), welford = var_welford(x), two_pass = var2(x))
```

## Methods for Mitigating Overflow and Underflow Errors

### Where overflow and underflow arise in statistics

Likelihoods are **products of many probabilities**, and normalizing constants involve **factorials** or **exponentials**:
$$
\prod_{i=1}^{2000} p(x_i) = \left(\tfrac12\right)^{2000} \approx 10^{-602} \quad\to\quad \text{underflow to } 0,
$$
$$
1000! \approx 4\times10^{2567} \quad\to\quad \text{overflow to } \infty,
$$
$$
\text{logistic: } \frac{e^{x}}{1+e^{x}} \text{ at } x = 1000 \quad\to\quad \frac{\infty}{\infty} = \texttt{NaN}.
$$

Two remedies are available: change the algebraic expression so that the intermediate quantities stay in range, or work on the logarithmic scale.

### Changing the algebraic expression

Two equivalent forms of the logistic (sigmoid) function:
$$
\sigma(x) = \frac{e^{x}}{1+e^{x}} = \frac{1}{1+e^{-x}}.
$$

| $x$ | $e^{x}/(1+e^{x})$ | $1/(1+e^{-x})$ |
|:--:|:--|:--|
| $+1000$ | $e^{x}=\infty$, $\infty/\infty = $ `NaN` ✗ | $e^{-x}=0$, $1/1 = 1$ ✓ |
| $-1000$ | $e^{x}=0$, $0/1 = 0$ ✓ | $e^{-x}=\infty$, $1/\infty = 0$ ✓ |

Evaluating the first form at a large positive value yields $\infty/\infty$; the second form lets $e^{-x}$ underflow harmlessly to $0$ and computes $1/(1+0) = 1$. In this case the second form is safe at both ends, because a division by $\infty$ gives $0$ correctly. In general one picks the form in which the exponential has a **non-positive** argument:
$$
\sigma(x) = \begin{cases} 1/(1+e^{-x}) & x \ge 0\\ e^{x}/(1+e^{x}) & x < 0.\end{cases}
$$

```{r}
#| label: sigmoid-demo
p1 <- function(theta) exp(theta) / (1 + exp(theta))
p2 <- function(theta) 1 / (1 + exp(-theta))

## large positive theta breaks p1, works in p2
theta <- 2000
c(p1 = p1(theta), p2 = p2(theta))

## large negative theta works for both
theta <- -2000
c(p1 = p1(theta), p2 = p2(theta))

## the piecewise form is safe everywhere and keeps full precision in both tails
sigmoid <- function(x) ifelse(x >= 0, 1 / (1 + exp(-x)), exp(x) / (1 + exp(x)))
sigmoid(c(-2000, -20, 0, 20, 2000))
```

**Built-in functions.** Software provides functions written specifically to avoid cancellation and overflow, and they should be used in preference to the naive composition:

| instead of | use | reason |
|:--|:--|:--|
| `log(1 + x)` | `log1p(x)` | for tiny $x$, `1 + x` rounds to 1 |
| `exp(x) - 1` | `expm1(x)` | for tiny $x$, `exp(x)` rounds to 1 |
| `log(factorial(n))` | `lfactorial(n)`, `lgamma(n + 1)` | `factorial(n)` overflows |
| `log(choose(n, k))` | `lchoose(n, k)` | overflow |
| `log(dnorm(x))` | `dnorm(x, log = TRUE)` | `dnorm(x)` underflows |
| `log(pnorm(x))` | `pnorm(x, log.p = TRUE)` | underflow |
| `log(1 - pnorm(x))` | `pnorm(x, lower.tail = FALSE, log.p = TRUE)` | cancellation and underflow |

```{r}
#| label: builtin-demo
c(naive = log(factorial(1000)), lfactorial = lfactorial(1000))
c(naive = log(dnorm(40)), dnorm_log = dnorm(40, log = TRUE))
c(naive = log(1 - pnorm(10)), stable = pnorm(10, lower.tail = FALSE, log.p = TRUE))
```

The last line shows both failures at once: `1 - pnorm(10)` is `1 - 1 = 0` in double precision, so the naive answer is `-Inf`, while the true value is about $-53.2$.

### Representing numbers in logarithmic form

Represent a positive number $x$ by $\ell_x = \log x$:
$$
x = e^{2000} \iff \ell_x = 2000, \qquad x = e^{-2000} \iff \ell_x = -2000 .
$$
Numbers that overflow or underflow in ordinary form are ordinary numbers on the log scale. The cost is that each operation $f$ must be **translated** to the log scale:
$$
x_1,\dots,x_n \xrightarrow{\;f\;} y \qquad\Longleftrightarrow\qquad \ell_1,\dots,\ell_n \xrightarrow{\;\tilde f\;} \ell_y = \log f(e^{\ell_1},\dots,e^{\ell_n}).
$$

**Products and quotients** are the easy case:
$$
\log(x_1 x_2) = \ell_1 + \ell_2, \qquad \log\left(\prod_{i=1}^{n} x_i\right) = \sum_{i=1}^{n} \ell_i, \qquad \log\frac{x_1}{x_2} = \ell_1 - \ell_2.
$$
In particular $\log \prod_i p(x_i) = \sum_i \log p(x_i)$: always compute the log-likelihood, never the likelihood.

```{r}
#| label: log-quotient-demo
log_x <- 800; x <- exp(log_x); x          # overflow
log_y <- 805; y <- exp(log_y); y          # overflow
x / y                                     # Inf/Inf = NaN

log_xovery <- log_x - log_y               # subtracting logs is safe
log_xovery
exp(log_xovery)

## likelihood versus log-likelihood for 2000 fair-coin tosses
p <- rep(0.5, 2000)
c(likelihood = prod(p), log_likelihood = sum(log(p)))
```

### The log-sum-exp trick

Sums are the hard case: $\log(x_1 + x_2) \neq \ell_1 + \ell_2$. We need
$$
\log\left(\sum_{i=1}^{n} x_i\right) = \log\left(\sum_{i=1}^{n} e^{\ell_i}\right),
$$
but the $e^{\ell_i}$ may overflow or underflow. Let $m = \max_i \ell_i$ and factor it out:
$$
\log\left(\sum_{i=1}^{n} e^{\ell_i}\right)
= \log\left(e^{m}\sum_{i=1}^{n} e^{\ell_i - m}\right)
= m + \log\left(\sum_{i=1}^{n} e^{\ell_i - m}\right).
$$

Every $\ell_i - m \le 0$, so $e^{\ell_i - m} \in (0, 1]$ and nothing overflows; the largest term is $e^{0} = 1$, so the sum is at least 1 and cannot underflow. Terms far below the maximum underflow to $0$ individually, but they would have contributed nothing at 16-digit precision anyway.

Illustration with $x_i = e^{2000}, e^{2001}, e^{2002}$, each of which overflows to `Inf`:

| | $i=1$ | $i=2$ | $i=3$ |
|:--|:--:|:--:|:--:|
| $\ell_i$ | 2000 | 2001 | 2002 |
| $\ell_i - m$ ($m = 2002$) | $-2$ | $-1$ | $0$ |
| $e^{\ell_i - m}$ | $0.135$ | $0.368$ | $1$ |
$$
\log\sum_i x_i = 2002 + \log(0.135 + 0.368 + 1) = 2002.408
$$

```{r}
#| label: log-sum-exp
log_sum_exp <- function(log_x) {
  max_log_x <- max(log_x)
  max_log_x + log(sum(exp(log_x - max_log_x)))
}

## values that would overflow directly
log_x <- c(2000, 2010, 2030)
exp(log_x)                                   # Inf
c(direct = log(sum(exp(log_x))), lse = log_sum_exp(log_x))

## values that would underflow directly
log_x <- -c(2000, 2010, 2030)
exp(log_x)                                   # 0
c(direct = log(sum(exp(log_x))), lse = log_sum_exp(log_x))
```

**Example: log-likelihood of a mixture.** The density of a two-component normal mixture is $f(y) = \pi_1\varphi(y;\mu_1,1) + \pi_2\varphi(y;\mu_2,1)$, a sum inside a product. For an observation far from both means, both component densities underflow and the naive log-likelihood is `-Inf`; with log-sum-exp applied observation by observation it is finite and correct.

```{r}
#| label: mixture-loglik
y <- c(-1, 0.5, 2, 45)                      # the last value is far from both means
mu <- c(0, 3); w <- c(0.6, 0.4)
naive <- sum(log(w[1] * dnorm(y, mu[1]) + w[2] * dnorm(y, mu[2])))
stable <- sum(sapply(y, function(yi) log_sum_exp(log(w) + dnorm(yi, mu, log = TRUE))))
c(naive = naive, stable = stable)
```

**Example: log-softmax.** The softmax probabilities $p_k = e^{x_k}/\sum_j e^{x_j}$ appear in multinomial logistic regression and neural networks. Their logarithm is $\log p_k = x_k - \log\sum_j e^{x_j}$, and the second term is a log-sum-exp.

```{r}
#| label: log-softmax
x <- c(1000, 1001, 1005)
naive <- log(exp(x) / sum(exp(x)))
stable <- x - log_sum_exp(x)
rbind(naive, stable, probabilities = exp(stable))
```

### Differences on the log scale

For $x > y > 0$ with $\ell_x = \log x$, $\ell_y = \log y$:
$$
\log(x - y) = \log\left(e^{\ell_x}\left(1 - e^{\ell_y - \ell_x}\right)\right) = \ell_x + \log\left(1 - e^{\ell_y - \ell_x}\right).
$$
Since $\ell_y - \ell_x < 0$, $e^{\ell_y-\ell_x}\in(0,1)$ and there is no overflow. When $x \approx y$, the factor $1 - e^{\ell_y-\ell_x}$ is itself a large − large subtraction; `log1p(-exp(ly - lx))` avoids it.

```{r}
#| label: log-minus-exp
log_minus_exp <- function(log_x, log_y) {
  if (log_x < log_y) stop("the first argument must be the larger")
  log_x + log1p(-exp(log_y - log_x))
}

log_minus_exp(2020, 2000)
exp(2000) - exp(1999)                       # direct subtraction: Inf - Inf = NaN

## nearly equal arguments: log1p keeps the precision that log(1 - .) would lose
lx <- 10; ly <- 10 - 1e-12
c(with_log1p = log_minus_exp(lx, ly), with_log = lx + log(1 - exp(ly - lx)), exact = log(exp(lx) - exp(ly)))
```

A statistical use of the difference formula is the survival function of a discrete distribution, $\log P(X > k) = \log\big(1 - P(X\le k)\big)$, when $P(X\le k)$ is available only as a log-probability close to $0$.

## Summary

- Integers occupy a fixed number of bits; exceeding $2^{31}-1$ gives `NA` in R.

- Doubles store a sign, a 52-bit mantissa (about 16 decimal digits), and an 11-bit exponent (magnitudes $10^{-308}$ to $10^{308}$). Neighbouring doubles are spaced at a relative distance of about $\epsilon = 2^{-52}$; integers above $2^{53}$ are not all representable.
- Overflow produces `Inf`, underflow produces `0`; the latter is silent and therefore more dangerous.

- Roundoff error is amplified by adding large + small (fix: add small terms first, pairwise or compensated summation) and by subtracting large − large (fix: reformulate so that same-sign quantities of similar size are combined, as in $e^{-x} = 1/e^{x}$ and the two-pass variance).
- Overflow and underflow are avoided by rewriting expressions so that exponentials have non-positive arguments, by using purpose-built functions (`log1p`, `expm1`, `lgamma`, `dnorm(log = TRUE)`), and by working on the log scale, where products become sums and sums are handled by the log-sum-exp trick.

## Exercises

**Question 1.** How do you compute
$$
\log\left(\frac{1}{1+e^{-x}}\right),
$$
so as to avoid overflow and underflow of $e^{-x}$, for both large positive and large negative $x$? (Hint: write it as $-\log(1+e^{-x})$ and consider `log1p` and the two forms of the sigmoid.)

**Question 2 (log-softmax).** How do you compute
$$
\log\left(\frac{e^{x_k}}{\sum_{j=1}^{K} e^{x_j}}\right),
$$
so as to avoid underflow and overflow of $e^{x_j}$? Verify your formula against the `log-softmax` example above for $x = (1000, 1001, 1005)$ and for $x = (-1000, -1001, -1005)$.

**Question 3.** The one-pass variance formula fails for $x = (M+1, M+2, M+3)$ with $M = 10^{8}$. Predict the smallest $M$ (a power of 10) at which the one-pass result in R first differs from $2$, then verify using the scan in the variance section.

**Question 4.** Write a function `log_dbinom(k, n, p)` that returns $\log\binom{n}{k}p^k(1-p)^{n-k}$ for $n = 10^5$ without overflow or underflow, using `lchoose` and `log1p`, and compare it with `dbinom(k, n, p, log = TRUE)`.

**Question 5.** Implement Welford's algorithm for the sample covariance of two variables and check it against `cov()` on data shifted by $10^{10}$.
