---
title: "Importance Sampling"
format:
  html:
    toc: true
    tbl-colwidths: true
filters:
  - shinylive
editor_options:
  chunk_output_type: console
code-fold: true
---

```{r}
#| label: setup
#| message: false
library(ars)
```

Rejection sampling turns draws from an envelope $g \ge f$ into exact draws from
$f$, but every rejected proposal is wasted, and in the tails of a poor
envelope almost every proposal is rejected. **Importance sampling** keeps
every proposal instead of discarding most of them: draw from an easier
function $g$ and attach a *weight* to each draw that corrects for sampling
from the wrong distribution. The price is that the resulting sample is no
longer an exact iid sample from $f$ — it is a *weighted* sample — but nothing
is ever thrown away, and no envelope condition $g \ge f$ is required.

## The Reweighting Idea

### The goal

As in the last two chapters, $f(\theta)$ is an **unnormalized** function of
interest — typically $f(\theta) = L(\theta)\pi(\theta)$ for a likelihood $L$
and prior $\pi$ — and we want two kinds of integral:
$$
A = E_f\big[a(\theta)\big] = \frac{\int a(\theta)\,f(\theta)\,d\theta}{\int f(\theta)\,d\theta},
\qquad\qquad
C_f = \int f(\theta)\,d\theta,
$$
a posterior expectation and a normalizing constant (the marginal likelihood
when $f = L\pi$). Plain Monte Carlo needs draws from $f/C_f$, which we
generally cannot produce.

Suppose instead we can draw from another function $g(\theta)$ (also possibly
unnormalized, with $C_g = \int g(\theta)\,d\theta$). Wherever $f$ puts mass,
$g$ must too:
$$
\{\theta : g(\theta) > 0\} \ \supseteq\ \{\theta : f(\theta) > 0\}.
$$
Unlike rejection sampling, $g$ need **not** dominate $f$ everywhere — there is
no requirement $g \ge f$. This is what makes importance sampling applicable
in situations where no usable envelope exists.

### Two estimators

Define the **importance weight** $w(\theta) = f(\theta)/g(\theta)$. For any
$a$,
$$
E_f\big[a(\theta)\big]
= \int a(\theta)\,\frac{f(\theta)}{C_f}\,d\theta
= \frac{C_g}{C_f}\, E_g\big[a(\theta)\,w(\theta)\big],
$$
and taking $a \equiv 1$ gives $C_f/C_g = E_g[w(\theta)]$. Combining the two,
$$
E_f\big[a(\theta)\big] = \frac{E_g\big[a(\theta)\,w(\theta)\big]}{E_g\big[w(\theta)\big]}.
$$
Draw $\theta_1,\ldots,\theta_S \overset{iid}{\sim} g/C_g$ and compute
$w_s = w(\theta_s)$. Two estimators follow directly:

- **Ratio of normalizing constants**: $\widehat{C_f/C_g} = \frac1S\sum_s w_s$,
  unbiased for $C_f/C_g$; if $g$ is already a normalized density ($C_g=1$)
  this estimates $C_f$ itself.
- **Self-normalized expectation**: $\hat A = \sum_{s=1}^S a(\theta_s)\,\tilde w_s$,
  with normalized weights $\tilde w_s = w_s / \sum_j w_j$ summing to $1$. This
  turns the proposal sample into a *weighted sample from $f$*: any unknown
  constants in $f$ and $g$ cancel in the ratio, so $\hat A$ needs only $f$ and
  $g$ up to a proportionality constant. It is a ratio of two averages and so
  is biased with bias $O(1/S)$, but consistent.

### Effective sample size

$\hat A$ is a weighted average of $a(\theta_s)$: draws where $f \gg g$ count
more, draws where $f \ll g$ count less. How many *equally weighted* draws is
the weighted sample worth?
$$
S_{\text{eff}} = \frac{1}{\displaystyle\sum_{s=1}^S \tilde w_s^2}.
$$
If the weights are all equal, $\tilde w_s = 1/S$ and $S_{\text{eff}} =
1/(S \cdot 1/S^2) = S$: the proposal is shaped just like the target, and every
draw counts fully. If one weight dominates all the others, $\tilde w_s
\approx 1$ for that draw and $S_{\text{eff}} \approx 1$: only one draw is
doing any work. A variance heuristic makes this precise:
$\text{Var}_g(\hat A) \approx \text{Var}_f(a)/S_{\text{eff}}$, the same form as
for an iid sample of size $S_{\text{eff}}$ from $f$ itself.

The catch, central to this chapter, is that $S_{\text{eff}}$ is computed
**only from the weights actually observed**. A proposal that never visits the
part of $\theta$-space where $f$ and $g$ disagree most will report a
reassuring $S_{\text{eff}}$ while being badly wrong — this is exactly the
subject of the "bad stories" below.

```{r}
#| label: ess-helpers
### effective sample size from raw (possibly unnormalized) weights
ess <- function(w) sum(w)^2 / sum(w^2)
### numerically stable log(sum(exp(lx)))
log_sum_exp <- function(lx) { m <- max(lx); m + log(sum(exp(lx - m))) }
### effective sample size computed from LOG weights (avoids under/overflow)
log_ess <- function(lw) exp(2 * log_sum_exp(lw) - log_sum_exp(2 * lw))
```

## Estimating a Truncated Normal: Probability and Expectation

### Setup: two quantities, one target

For $X \sim N(0,1)$, take $f(x) = \varphi(x)\,I(l<x<u)$, the standard normal
density truncated to $(l,u)$, so that $C_f = P(l<X<u)$. The same $f$ also
defines a conditional expectation $A = E\big[a(X) \mid l<X<u\big]$ for any
function $a$ — for instance $a(x)=x^2$. Both quantities are instances of the
two estimators from the start of the chapter, $\widehat{C_f/C_g} =
\frac1S\sum_s w_s$ and the self-normalized $\hat A = \sum_s a(\theta_s)\tilde
w_s$, applied to the *same* weighted sample — so one simulation answers both
questions at once, and it is natural to compare their errors side by side.

### Two proposals

- **The untruncated normal itself**, $g(x) = \varphi(x)$: since $f \le g$
  everywhere (the indicator only ever removes mass from $\varphi$), the
  importance weight is $w(x) = I(l<x<u)$ — exactly $0$ or $1$. Averaging $w$
  over draws from $g$ is plain Monte Carlo for $C_f$; averaging $a(x)$ over
  only the draws that land inside $(l,u)$ is *exact rejection sampling* for
  $A$, with no extra machinery, because a weight in $\{0,1\}$ makes
  acceptance deterministic. Both readings of the same draws waste every
  proposal that lands outside $(l,u)$ — fine when $(l,u)$ holds most of the
  mass, ruinous when it is rare.
- **Uniform on $(l,u)$**, $g(x) = I(l\le x\le u)/(u-l)$: every draw lands
  inside the interval, so $w(x) = (u-l)\varphi(x)$ is never $0$. This is the
  *Monte Carlo* version of the midpoint rule from the quadrature chapter — a
  random grid instead of an equally spaced one, evaluating the same weighted
  sum — and it feeds both estimators from a single sample.

```{r}
#| label: tnorm-functions
### naive proposal g = phi: w in {0, 1}; doubles as plain MC (Chat) and exact
### rejection sampling (Ahat, over the accepted subset) from the same draws
est_tnorm_naive <- function(a_fn, l, u, S) {
    X <- rnorm(S)
    inside <- X > l & X < u
    c(Chat = mean(inside), Ahat = if (any(inside)) mean(a_fn(X[inside])) else NA_real_)
}
### uniform proposal on (l, u): every draw counts for both quantities
est_tnorm_unif <- function(a_fn, l, u, S) {
    X <- runif(S, l, u)
    w <- dnorm(X) * (u - l)
    c(Chat = mean(w), Ahat = sum(a_fn(X) * w) / sum(w))
}
```

```{r}
#| label: fig-tnorm-proposals
#| fig-width: 10
#| fig-height: 4.2
#| fig-cap: "Target f (bold) and the two proposals for each interval: the naive proposal (dashed) is the untruncated normal itself — inside the interval it is literally identical to f — while the uniform proposal (flat) has a totally different shape but guarantees every draw lands inside. Right panel is on a log scale because f is minuscule near (5, 6)."
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))

l1 <- -2; u1 <- 2
curve(dnorm, -4, 4, lwd = 2, lty = 2, col = "steelblue", xlab = "x", ylab = "density",
      main = "moderate interval (-2, 2)", ylim = c(0, 0.55))
curve(dnorm, from = l1, to = u1, add = TRUE, lwd = 4, col = "firebrick")
segments(l1, 1 / (u1 - l1), u1, 1 / (u1 - l1), col = "forestgreen", lwd = 3)
legend("topright", bty = "n", lwd = c(4, 3, 2), lty = c(1, 1, 2),
       col = c("firebrick", "forestgreen", "steelblue"),
       legend = c("f = target", "g: uniform", "g: naive (= phi everywhere)"))

l2 <- 5; u2 <- 6
curve(dnorm, 4, 7, log = "y", lwd = 2, lty = 2, col = "steelblue", xlab = "x",
      ylab = "density (log scale)", main = "rare interval (5, 6)", ylim = c(1e-11, 2))
curve(dnorm, from = l2, to = u2, add = TRUE, lwd = 4, col = "firebrick")
segments(l2, 1 / (u2 - l2), u2, 1 / (u2 - l2), col = "forestgreen", lwd = 3)
legend("bottomleft", bty = "n", lwd = c(4, 3, 2), lty = c(1, 1, 2),
       col = c("firebrick", "forestgreen", "steelblue"),
       legend = c("f = target", "g: uniform", "g: naive"))
```

Inside the interval, the naive proposal's shape is not the problem — it is
*exactly* $f$'s shape, since $g=\varphi$ everywhere. Its problem is how rarely
it visits the interval at all: for $(5,6)$ the flat uniform proposal (green)
sits many orders of magnitude above the target near the right edge of its own
support, which is precisely the size of the importance weight needed to
correct for its shape mismatch — a price worth paying, since every one of its
draws lands inside.

### Comparing MSE for both quantities

```{r}
#| label: tbl-tnorm-mse
#| tbl-cap: "Mean squared error of the probability and expectation estimates over 1000 replications, for a moderate and a rare event, S = 100 draws per replication. a(x) = x^2; true E[a] from a 100,000-point midpoint rule."
a_fn <- function(x) x^2
midpoint_ref <- function(l, u, n = 100000) {
    h <- (u - l) / n; xm <- l + (seq_len(n) - 0.5) * h
    sum(a_fn(xm) * dnorm(xm)) * h / (pnorm(u) - pnorm(l))
}

set.seed(1)
cases <- list(c(-2, 2), c(5, 6))
tnorm_mse <- do.call(rbind, lapply(cases, function(A) {
    tp    <- pnorm(A[2]) - pnorm(A[1])
    ref   <- midpoint_ref(A[1], A[2])
    naive <- replicate(1000, est_tnorm_naive(a_fn, A[1], A[2], 100))
    unif  <- replicate(1000, est_tnorm_unif(a_fn, A[1], A[2], 100))
    ok <- !is.na(naive["Ahat", ])
    data.frame(
        interval       = sprintf("(%g, %g)", A[1], A[2]),
        true_prob      = formatC(tp, format = "e", digits = 3),
        mse_prob_naive = formatC(mean((naive["Chat", ] - tp)^2), format = "e", digits = 3),
        mse_prob_unif  = formatC(mean((unif["Chat", ] - tp)^2), format = "e", digits = 3),
        true_Ea        = formatC(ref, format = "e", digits = 3),
        mse_Ea_naive   = if (any(ok)) formatC(mean((naive["Ahat", ok] - ref)^2), format = "e", digits = 3)
                          else sprintf("never accepted (0/%d)", ncol(naive)),
        mse_Ea_unif    = formatC(mean((unif["Ahat", ] - ref)^2), format = "e", digits = 3)
    )
}))
knitr::kable(tnorm_mse, align = c("l", rep("r", 6)),
             col.names = c("interval", "true P", "MSE P (naive)", "MSE P (uniform)",
                           "true E[a]", "MSE E[a] (naive)", "MSE E[a] (uniform)"))
```

For the moderate interval $(-2,2)$ both proposals estimate both quantities
reasonably well. For the rare event $(5,6)$, the naive proposal never once
places a draw inside the interval across any of the 1000 replications — it
cannot even form an estimate of $E[a(X)\mid 5<X<6]$ — and its probability
estimate looks deceptively accurate in MSE only because the true probability
is itself tiny (both estimate and truth round to $0$). The uniform proposal,
which *guarantees* every draw lands in $(5,6)$, stays usable for both
quantities and is an order of magnitude more accurate for the probability
besides.

### An underflowing tail probability

Push the interval further out, to $A = (50,100)$. Now even $\varphi(50)$
itself underflows to exactly $0$ in double precision (its log is about
$-1251$, far below the smallest representable log), so every calculation must
be carried out on the log scale.

```{r}
#| label: exact-far-tail
a <- 50; b <- 100
log_minus_exp <- function(la, lb) la + log1p(-exp(lb - la))
log_exact <- log_minus_exp(pnorm(-a, log = TRUE), pnorm(-b, log = TRUE))
log_exact
```

### Matching the proposal to the tail: a better envelope

A uniform proposal on $(50,100)$ still works in principle, but $\varphi$ decays
so fast across the interval that the weights vary enormously: nearly all of
the probability mass under $\varphi$ restricted to $(50,100)$ sits within a
sliver near $x=50$, so a uniform draw almost never lands where the weight is
large. A **shifted exponential** proposal with rate $\lambda \approx a$
matches that decay and keeps the weights nearly constant:
$$
g(x) = \frac{\lambda e^{-\lambda(x-a)}}{1-e^{-\lambda(b-a)}}\,I(a\le x\le b).
$$

```{r}
#| label: fig-tail-log-weight
#| fig-width: 10
#| fig-height: 4.2
#| fig-cap: "Left: log-density of f and the shifted-exponential proposal (plain density underflows to 0 here, so only the log scale is meaningful). Right: log importance weight log f(x) - log g(x) for each proposal, x - a on a log scale, with a rug of 2000 draws actually generated from each g. The exponential's own draws (orange) stay in the region where its weight curve is flat; the uniform's draws (green) spread into the region where its weight has already collapsed."
lam <- a
log_g_exp <- function(x) log(lam) - lam * (x - a) - log1p(-exp(-lam * (b - a)))
log_g_u   <- function(x) rep(-log(b - a), length(x))
log_f     <- function(x) dnorm(x, log = TRUE)

set.seed(7)
Sfig <- 2000
xfig_u <- runif(Sfig, a, b)
ufig_e <- runif(Sfig); xfig_e <- a - log(1 - ufig_e * (1 - exp(-lam * (b - a)))) / lam

par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))

curve(log_f, a, b, lwd = 4, col = "firebrick", xlab = "x", ylab = "log density",
      main = "log-density of f and the exponential g", n = 500)
curve(log_g_exp, add = TRUE, lwd = 3, col = "darkorange", n = 500)
legend("bottomleft", bty = "n", lwd = 3, col = c("firebrick", "darkorange"),
       legend = c("f = target", "g: shifted exponential"))

d  <- 10^seq(-4, log10(b - a), length.out = 500)
xg <- a + d
plot(d, log_f(xg) - log_g_u(xg), type = "l", lwd = 3, col = "forestgreen", log = "x",
     xlab = "x - a  (log scale)", ylab = "log weight  log f(x) - log g(x)",
     main = "log weight, and where each g's draws land",
     ylim = range(c(log_f(xg) - log_g_u(xg), log_f(xg) - log_g_exp(xg))))
lines(d, log_f(xg) - log_g_exp(xg), lwd = 3, col = "darkorange")
rug(pmax(xfig_u - a, 1e-4), col = adjustcolor("forestgreen", 0.4))
rug(pmax(xfig_e - a, 1e-4), col = adjustcolor("darkorange", 0.4))
legend("bottomleft", bty = "n", lwd = 3, cex = 0.85, col = c("forestgreen", "darkorange"),
       legend = c("uniform: draws reach the collapse", "exponential: draws stay flat"))
```

```{r}
#| label: tbl-tail-proposals
#| tbl-cap: "Two proposals for the same rare event, S = 10,000 draws: log-scale weights avoid the underflow that would occur if densities were multiplied directly."
S <- 1e4; lam <- a
x_u <- runif(S, a, b);  lw_u <- log(b - a) + dnorm(x_u, log = TRUE)
u_e <- runif(S); x_e <- a - log(1 - u_e * (1 - exp(-lam * (b - a)))) / lam        # inverse cdf
lg_e <- log(lam) - lam * (x_e - a) - log1p(-exp(-lam * (b - a)))
lw_e <- dnorm(x_e, log = TRUE) - lg_e

tail_tbl <- data.frame(
    proposal   = c("uniform on (a, b)", "shifted exponential"),
    log_C_f    = c(log_sum_exp(lw_u) - log(S), log_sum_exp(lw_e) - log(S)),
    S_eff      = c(log_ess(lw_u), log_ess(lw_e))
)
knitr::kable(tail_tbl, digits = c(0, 4, 1))
```

Both proposals give a log-probability close to `r round(log_exact, 3)`, but
their effective sample sizes are worlds apart: the exponential proposal keeps
essentially all `r S` draws informative, while the uniform proposal's
effective sample size collapses because almost all of its weight concentrates
on the handful of draws nearest $a$. Shaping $g$ like $f$ is not just
elegant — it is what makes the computation efficient.

### Exact draws instead of weights: adaptive rejection sampling

The naive filter from the setup above is exact rejection sampling too, but
only when $(l,u)$ is not rare — most of its proposals are wasted otherwise, as
just seen for $(5,6)$. **Adaptive rejection sampling** (`ars` package,
applicable because $\log\varphi$ is concave) builds its envelope directly on
$(l,u)$ itself instead of on the whole real line, so it stays exact and fully
efficient no matter how rare the interval is — at the cost of an accept/reject
loop that a single weighted average avoids.

```{r}
#| label: tnorm-ars-function
### exact draws via adaptive rejection sampling (log phi is concave)
sample_tnorm_ars <- function(n, l, u) {
    logf   <- function(x) dnorm(x, log = TRUE)
    fprima <- function(x) -x
    ars(n, f = logf, fprima = fprima, x = c(l, (l + u) / 2, u),
        lb = TRUE, ub = TRUE, xlb = l, xub = u)
}
```

The original version of `sample_tnorm_ars` always drew exactly $10{,}000$
points regardless of the requested `n`; the version above respects `n`, which
matters for the timing comparison next.

```{r}
#| label: tbl-tnorm-speed
#| tbl-cap: "Importance sampling versus adaptive rejection sampling for E[X^2 | l < X < u], averaged over 100 replications at S = 1000 draws each, for the same moderate and rare intervals as above. Timings are milliseconds per replication."
tnorm_speed <- do.call(rbind, lapply(cases, function(A) {
    l <- A[1]; u <- A[2]; ref <- midpoint_ref(l, u)
    t_imps <- system.time(EA_imps <- replicate(100, est_tnorm_unif(a_fn, l, u, 1000)["Ahat"]))
    t_rej  <- system.time(EA_rej  <- replicate(100, mean(a_fn(sample_tnorm_ars(1000, l, u)))))
    data.frame(interval = sprintf("(%g, %g)", l, u), reference = ref,
               mean_imps = mean(EA_imps), mse_imps = mean((EA_imps - ref)^2),
               ms_imps = 1000 * t_imps[["elapsed"]] / 100,
               mean_rej = mean(EA_rej), mse_rej = mean((EA_rej - ref)^2),
               ms_rej = 1000 * t_rej[["elapsed"]] / 100)
}))
knitr::kable(tnorm_speed, digits = c(0, 4, 4, 6, 2, 4, 6, 2))
```

Both estimators agree with the midpoint-rule reference to the precision
shown, including for the rare interval $(5,6)$ where the naive filter above
never once produced an estimate; importance sampling is markedly faster here
because it needs no accept/reject loop or per-interval envelope construction,
at the cost of returning a *weighted* rather than an exact sample.

## Two Bad Stories: When Reweighting Fails Silently

$S_{\text{eff}}$ diagnoses only *one* kind of failure: weights that vary
wildly among the draws actually observed. It says nothing about mass that $g$
never visits at all, and a badly wrong estimate can carry a perfectly healthy
$S_{\text{eff}}$.

### A proposal that misses a mode

Let $f(\theta) = \tfrac12\varphi(\theta+5) + \tfrac12\varphi(\theta-5)$ (a
50/50 mixture, so $C_f=1$ and $E_f(\theta)=0$) and $g(\theta) = \varphi(\theta+5)$.
Both are positive everywhere, so the support requirement holds. But $g$ places
essentially all of its mass near $\theta=-5$ and, for $\theta$ near $-5$,
$$
w(\theta) = \tfrac12 + \tfrac12\,\frac{\varphi(\theta-5)}{\varphi(\theta+5)}
= \tfrac12 + \tfrac12\, e^{10\theta} \ \approx\ \tfrac12,
$$
so every weight the sampler actually sees looks almost identical.

```{r}
#| label: fig-missed-mode
#| fig-width: 10
#| fig-height: 4.2
#| fig-cap: "Left: target f (mixture of two normals) and proposal g (only the left component). Right: normalized weights look almost flat — S_eff is close to S — even though the estimate is badly wrong."
set.seed(2)
S <- 5000
f_mix <- function(t) 0.5 * dnorm(t + 5) + 0.5 * dnorm(t - 5)
g_mix <- function(t) dnorm(t + 5)
th <- rnorm(S, -5)
w  <- f_mix(th) / g_mix(th)
wt <- w / sum(w)

par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
curve(f_mix, -9, 9, lwd = 3, col = "firebrick", xlab = expression(theta), ylab = "",
      main = "target f and proposal g", ylim = c(0, 0.42))
curve(g_mix, add = TRUE, lwd = 3, col = "steelblue")
rug(th[1:300], col = adjustcolor("steelblue", 0.5))
legend("topright", bty = "n", lwd = 3, col = c("firebrick", "steelblue"), legend = c("f", "g"))
hist(wt * S, breaks = 40, col = "gray85", border = "white", xlab = expression(S * tilde(w)[s]),
     main = sprintf("weights: S_eff = %.0f of %d", ess(w), S))
```

```{r}
#| label: tbl-missed-mode
#| tbl-cap: "Missed-mode failure: the effective sample size looks fine, but both quantities of interest are far from the truth."
missed_mode_tbl <- data.frame(
    quantity = c("C_f / C_g (true 1)", "E_f[theta] (true 0)", "S_eff (of S = 5000)"),
    estimate = c(mean(w), sum(th * wt), ess(w))
)
knitr::kable(missed_mode_tbl, digits = 3)
```

The estimate of $C_f/C_g$ comes out near `r round(mean(w), 2)` instead of $1$,
and $E_f[\theta]$ near `r round(sum(th * wt), 2)` instead of $0$ — both wrong
by a wide margin — while $S_{\text{eff}}$ reports essentially the full sample
size. The draws that would reveal the problem, near $\theta=+5$, have
probability under $g$ of order $\Phi(-10) \approx 10^{-23}$: they simply never
appear, at any practical $S$.

### A narrower-tailed proposal: approximating a Cauchy survival probability

The same failure in a subtler form: instead of a finite interval, let $X$
follow a standard Cauchy distribution and take $f(x) = f_{\text{Cauchy}}(x)\,
I(x>t)$ for $t=2$, so $C_f = P(X>t)$ is the Cauchy **survival probability**
beyond $t$. A tempting proposal is a **truncated standard normal** on
$(t,\infty)$ — easy to sample, and the support condition $\{g>0\}\supseteq
\{f>0\}$ holds exactly, so nothing looks obviously wrong at first glance.

```{r}
#| label: fig-cauchy-normal-tails
#| fig-cap: "Target (Cauchy tail beyond t = 2, firebrick) and proposal (truncated normal beyond t = 2, steelblue), log density scale, with a rug of 2000 draws from the proposal. The densities cross near x = 3.8; beyond that the proposal falls away far faster than the target, and its own draws almost never reach past x = 5."
t <- 2
Zt <- 1 - pnorm(t)
f_tail <- function(x) dcauchy(x)
g_tail <- function(x) dnorm(x) / Zt

set.seed(1)
X_g <- qnorm(pnorm(t) + runif(2000) * Zt)

curve(f_tail, t, 8, log = "y", lwd = 4, col = "firebrick", xlab = "x", ylab = "density (log scale)",
      main = "target: Cauchy tail vs. proposal: truncated normal", ylim = c(1e-10, 3), n = 500)
curve(g_tail, add = TRUE, lwd = 3, col = "steelblue", n = 500)
rug(X_g, col = adjustcolor("steelblue", 0.35))
legend("bottomleft", bty = "n", lwd = c(4, 3), col = c("firebrick", "steelblue"),
       legend = c("f: Cauchy(0,1), x > 2", "g: N(0,1), x > 2 (truncated)"))
```

The densities cross near $x\approx3.8$; to the right, $\varphi$'s Gaussian
decay $e^{-x^2/2}$ falls away far faster than the Cauchy's polynomial tail
$1/(\pi x^2)$, so the importance weight $w(x)=f(x)/g(x)$ grows without bound —
by $x=6$ it is already of order $10^4$, by $x=8$ of order $10^{10}$. This
generalizes the too-narrow-support failure to tails: a truncated normal is,
in effect, narrower than a Cauchy no matter how far out its own truncation
point sits, because it assigns vanishing *relative* density to the region
that actually carries the heavy tail's mass. Formally, $\text{Var}_g(w) =
\int f(x)^2/g(x)\,dx = \infty$ here — the ratio estimator is still unbiased,
but its Monte Carlo variance is infinite. The reverse mismatch — a
*heavier*-tailed proposal than the target, such as a Cauchy or low-df
Student-$t$ proposal for a normal target — is harmless: it only ever produces
occasional small weights, never unbounded ones.

```{r}
#| label: tbl-cauchy-normal-mse
#| tbl-cap: "Estimated P(X > 2) for the Cauchy using the truncated-normal proposal, over 500 replications at S = 1000 and 200 replications at S = 100,000. The median stays badly low at both sample sizes, and effective sample size as a percentage of S gets worse, not better, as S grows."
true_surv <- 1 - pcauchy(t)
est_cauchy_normal <- function(S) {
    X <- qnorm(pnorm(t) + runif(S) * Zt)
    w <- dcauchy(X) / (dnorm(X) / Zt)
    c(Chat = mean(w), ess = ess(w))
}

set.seed(2)
cauchy_tbl <- do.call(rbind, lapply(list(c(500, 1000), c(200, 1e5)), function(spec) {
    reps <- t(replicate(spec[1], est_cauchy_normal(spec[2])))
    data.frame(S = spec[2], reps = spec[1],
               mean_est = mean(reps[, "Chat"]), median_est = median(reps[, "Chat"]),
               sd_est = sd(reps[, "Chat"]), mean_ess_pct = 100 * mean(reps[, "ess"]) / spec[2])
}))
knitr::kable(cbind(true = round(true_surv, 4), cauchy_tbl), digits = c(4, 0, 0, 4, 4, 4, 2))
```

The true survival probability is `r round(true_surv, 3)`, but the *typical*
(median) run badly undershoots it at both sample sizes, and the standard
deviation across replications shrinks far slower than the $1/\sqrt{S}$ rate a
finite-variance estimator would give — a hundredfold increase in $S$ buys
nowhere near a tenfold reduction in spread, and $S_{\text{eff}}$ as a
*percentage* of $S$ gets worse, not better, as $S$ grows. Occasionally a draw
lands far enough into the tail to carry a huge weight and swing that one
replication's estimate close to, or past, the truth; but such draws are rare
enough under the truncated normal that most practical runs never see one, and
there is no way to tell from a single run's $S_{\text{eff}}$ whether it got
lucky. Together with the missed-mode example,
the lesson generalizes beyond finite supports: $S_{\text{eff}}$ diagnoses
variability among the weights actually observed, but says nothing about how a
proposal's shape compares to the target's far outside the region the proposal
is likely to visit — whether that region is a missed mode or, as here, a
heavier tail.

### Explore it interactively {#sec-importance-app}

This app runs live
only in the HTML edition of this book; readers of other formats can follow
the link below it.

::: {#fig-importance-app .content-visible when-format="html"}

```{shinylive-r}
#| label: missed-mode-app
#| standalone: true
#| viewerHeight: 800

library(shiny)

f_mix <- function(t) 0.5 * dnorm(t + 5) + 0.5 * dnorm(t - 5)
ess   <- function(w) sum(w)^2 / sum(w^2)

ui <- fluidPage(
    titlePanel(NULL),
    sidebarLayout(
        sidebarPanel(
            width = 4,
            sliderInput("mean_g", "proposal mean", min = -9, max = 9, value = -5, step = 0.5),
            sliderInput("sd_g", "proposal sd", min = 0.5, max = 8, value = 1, step = 0.5),
            helpText("Target f is a 50/50 mixture of N(-5, 1) and N(5, 1), so the ",
                     "truth is C_f/C_g = 1 and E_f(theta) = 0 when g integrates to 1. ",
                     "A narrow proposal centred on one mode gives a deceptively healthy ",
                     "S_eff while both estimates are wrong; only a proposal wide enough ",
                     "or centred between the modes recovers the truth."),
            tableOutput("tab")
        ),
        mainPanel(
            width = 8,
            plotOutput("curves", height = "260px"),
            plotOutput("weights", height = "260px")
        )
    )
)

server <- function(input, output, session) {

    draws <- reactive({
        set.seed(1)
        th <- rnorm(5000, input$mean_g, input$sd_g)
        w  <- f_mix(th) / dnorm(th, input$mean_g, input$sd_g)
        list(th = th, w = w)
    })

    output$curves <- renderPlot({
        d <- draws()
        par(mar = c(4, 4, 2, 1))
        curve(f_mix, -12, 12, lwd = 3, col = "firebrick", xlab = expression(theta), ylab = "",
              main = "target f and proposal g", ylim = c(0, 0.45))
        curve(dnorm(x, input$mean_g, input$sd_g), add = TRUE, lwd = 3, col = "steelblue")
        rug(d$th[1:300], col = adjustcolor("steelblue", 0.5))
    })

    output$weights <- renderPlot({
        d <- draws(); wt <- d$w / sum(d$w)
        par(mar = c(4, 4, 2, 1))
        hist(wt * length(d$w), breaks = 40, col = "gray85", border = "white",
             xlab = expression(S * tilde(w)[s]),
             main = sprintf("S_eff = %.0f of %d", ess(d$w), length(d$w)))
    })

    output$tab <- renderTable({
        d <- draws(); wt <- d$w / sum(d$w)
        data.frame(quantity = c("C_f / C_g (true 1)", "E_f[theta] (true 0)", "S_eff (of 5000)"),
                   estimate = c(mean(d$w), sum(d$th * wt), ess(d$w)))
    }, digits = 3, rownames = FALSE)
}

shinyApp(ui, server)
```

**Shinylive App Illustrating Importance Sampling with a Missed Mode.** The app below reproduces the missed-mode example with an adjustable proposal:
move its mean and spread and watch the effective sample size stay high while
the estimated mean drifts away from the true value of $0$ whenever the
proposal fails to reach both components of the target.

:::

::: {.content-visible unless-format="html"}
This section contains a live, in-browser Shiny app that only runs in the HTML
edition of this book. View it at
<https://longhaisk.github.io/compstat/chapters/unit11_importance_sampling.html#sec-importance-app>.
:::

::: {.callout-note}
### Reproducing this figure
See the callout in [Numerical Quadrature](unit09_integral.qmd) for how to
install the `shinylive` extension used above.
:::

## Application: Marginal Likelihood via Importance Sampling

### Setup

The [quadrature](unit09_integral.qmd) and Laplace-approximation chapters
compute the marginal likelihood $P(y) = \int L(\theta)\pi(\theta)\,d\theta$ of
the same conjugate-free Gaussian model — unknown mean $\mu$ and
$w=\log\sigma^2$, independent normal priors — by a product grid and by a
mode-and-Hessian approximation, respectively. Importance sampling offers a
third route, and one that comes with a built-in accuracy diagnostic
($S_{\text{eff}}$) that neither of the other two methods provides directly.

Take $g(\mu,w)$ to be the **Gaussian approximation to the posterior**
obtained by a Laplace fit: locate the posterior mode $(\hat\mu,\hat w)$ by
optimization, take $g = N_2\big((\hat\mu,\hat w),\, H^{-1}\big)$ for $H$ the
Hessian of $-\log[L\pi]$ at the mode, and importance-sample from $g$ with
$f = L\pi$:
$$
\widehat{P(y)} = \frac{1}{S}\sum_{s=1}^S w_s, \qquad
w_s = \frac{L(\theta_s)\pi(\theta_s)}{g(\theta_s)}, \qquad \theta_s \sim g.
$$
Because $g$ is shaped like the posterior near its mode, this is a "good
story" proposal by construction — in sharp contrast to sampling directly from
the (typically much wider, uninformative) prior, which is exactly the
`log_marlik_mc` function used elsewhere in this book as a naive-Monte-Carlo
check: it is importance sampling with $g = \pi$, and its weights are just the
likelihood values $L(\theta_s)$.

### An efficient implementation

```{r}
#| label: marlik-functions
log_lik   <- function(x, mu, w) sum(dnorm(x, mu, exp(w / 2), log = TRUE))
log_prior <- function(mu, w, mu_0, sigma_mu, w_0, sigma_w) {
    dnorm(mu, mu_0, sigma_mu, log = TRUE) + dnorm(w, w_0, sigma_w, log = TRUE)
}
neg_log_post <- function(theta, x, mu_0, sigma_mu, w_0, sigma_w) {
    -log_lik(x, theta[1], theta[2]) -
        log_prior(theta[1], theta[2], mu_0, sigma_mu, w_0, sigma_w)
}

### log density of N(mu, A^{-1}) at every column of a 2 x S matrix of points at once,
### replacing a per-point apply() call with one vectorized quadratic form
log_dmvnorm_batch <- function(Theta, mu, A) {
    Tc <- Theta - mu
    quad <- colSums((A %*% Tc) * Tc)
    0.5 * (-nrow(Theta) * log(2 * pi) + sum(log(svd(A)$d)) - quad)
}

### log marginal likelihood by importance sampling from a Laplace-shaped Gaussian
log_marlik_imps <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    fit <- nlm(neg_log_post, c(mean(x), log(stats::var(x))), hessian = TRUE,
               x = x, mu_0 = mu_0, sigma_mu = sigma_mu, w_0 = w_0, sigma_w = sigma_w)
    mu_hat <- fit$estimate; H <- fit$hessian
    L <- t(chol(solve(H)))                        # Sigma = H^{-1} = L L'
    Theta <- L %*% matrix(rnorm(2 * S), 2, S) + mu_hat

    log_g <- log_dmvnorm_batch(Theta, mu_hat, H)
    log_f <- -apply(Theta, 2, neg_log_post, x = x, mu_0 = mu_0, sigma_mu = sigma_mu,
                     w_0 = w_0, sigma_w = sigma_w)
    lw <- log_f - log_g
    structure(log_sum_exp(lw) - log(S), ess = log_ess(lw))
}

### log marginal likelihood by naive Monte Carlo from the prior (g = pi): a
### second, independent importance-sampling estimator with an unfavourable g
log_marlik_prior <- function(x, mu_0, sigma_mu, w_0, sigma_w, S) {
    mus <- rnorm(S, mu_0, sigma_mu); ws <- rnorm(S, w_0, sigma_w)
    lw <- vapply(seq_len(S), function(i) log_lik(x, mus[i], ws[i]), numeric(1))
    structure(log_sum_exp(lw) - log(S), ess = log_ess(lw))
}
```

Replacing the two `apply()` calls used in earlier drafts of this code — one
for the proposal log-density, one for the log-posterior — with a single
vectorized quadratic form (`log_dmvnorm_batch`) removes $S$ separate R
function calls for the proposal evaluation; only the data-dependent
log-likelihood still needs one call per draw, since it sums over the $n$
observations for each $\theta_s$ separately.

### Checking accuracy

```{r}
#| label: tbl-marlik-check
#| tbl-cap: "Log marginal likelihood by importance sampling (Laplace-shaped proposal, S = 200) across 500 replications, for two simulated data sets."
set.seed(4)
x_small <- rnorm(10);  x_large <- rnorm(200)

check_tbl <- do.call(rbind, lapply(list(n10 = x_small, n200 = x_large), function(x) {
    reps <- replicate(500, log_marlik_imps(x, 0, 5, 0, 5, 200), simplify = FALSE)
    vals <- vapply(reps, as.numeric, numeric(1))
    esss <- vapply(reps, attr, numeric(1), which = "ess")
    data.frame(mean_log_ml = mean(vals), sd_log_ml = sd(vals), mean_ess = mean(esss))
}))
check_tbl <- cbind(dataset = c("n = 10", "n = 200"), check_tbl)
knitr::kable(check_tbl, digits = 4)
```

The Monte Carlo standard deviation across replications is small relative to
the estimate itself, and the effective sample size stays close to the nominal
$S=200$ in both cases — evidence that the Laplace-shaped proposal tracks the
posterior well even though it is, after all, only a second-order
approximation to it.

### Why the proposal matters: Laplace-shaped versus prior sampling

The prior is a valid proposal for the same target — every posterior draw is
also a prior draw with positive density — so `log_marlik_prior` is a
legitimate importance-sampling estimator too. But the prior is far more
diffuse than the posterior, especially once $n$ is not tiny, so it is a much
poorer match to $f = L\pi$.

```{r}
#| label: tbl-marlik-ess-vs-n
#| tbl-cap: "Effective sample size (of S = 500 draws) for the Laplace-shaped proposal versus prior sampling, as the sample size n of the simulated data grows. Both target the same posterior for (mu, w)."
set.seed(5)
ns <- c(5, 20, 100, 400)
ess_tbl <- do.call(rbind, lapply(ns, function(n) {
    x <- rnorm(n)
    data.frame(n = n,
               ess_laplace_proposal = attr(log_marlik_imps(x, 0, 5, 0, 5, 500), "ess"),
               ess_prior_sampling   = attr(log_marlik_prior(x, 0, 5, 0, 5, 500), "ess"))
}))
knitr::kable(ess_tbl, digits = c(0, 1, 1))
```

As $n$ grows the posterior sharpens around the truth while the prior does
not, so prior sampling's effective sample size deteriorates steadily; the
Laplace-shaped proposal, which re-centers and re-scales itself around the
posterior mode for every data set, keeps pace. This is the marginal-likelihood
instance of the same lesson as the missed-mode example above: a fixed,
data-independent proposal becomes an increasingly bad match to a target that
concentrates as more data arrive, and only a proposal that adapts to the
target avoids the resulting collapse in efficiency.

## Summary

- Importance sampling reweights draws from a proposal $g$ by
  $w(\theta) = f(\theta)/g(\theta)$ instead of accepting or rejecting them,
  using $\widehat{C_f/C_g} = \frac1S\sum_s w_s$ and the self-normalized
  $\hat A = \sum_s a(\theta_s)\tilde w_s$; only the support condition
  $\{g>0\}\supseteq\{f>0\}$ is required, not an envelope $g \ge f$.
- A good proposal shaped like the target on its support gives near-constant
  weights and $S_{\text{eff}} \approx S$ (the truncated-normal and
  Laplace-proposal examples), and a single weighted sample can answer several
  questions about the same target — a probability and an expectation alike.
- A proposal that misses a mode or part of the support fails **silently**: the
  observed weights can look perfectly healthy while both $C_f/C_g$ and
  $E_f[a]$ are badly wrong, and no amount of extra sampling fixes it.
  $S_{\text{eff}}$ diagnoses only the noisy failure (wildly varying observed
  weights), never the silent one.
- For the marginal likelihood, a proposal built from a Laplace approximation
  at the posterior mode remains efficient as the data accumulate and the
  posterior sharpens, while a fixed, data-independent proposal such as the
  prior does not — reinforcing why the mode-finding methods from earlier
  chapters remain useful even once we turn to sampling-based methods.
