---
subtitle: "Introduction to Statistical Computation"
pagetitle: "intro-slides"
author: "Longhai Li"
format: banner_slides-revealjs
---

# Review of Basic Concepts

## Statistical Inference: The Setup

- Data: $x_1, \ldots, x_n$ observed
- Model: $x_i \overset{iid}{\sim} f(x \mid \theta)$, with $\theta$ unknown
- Three basic tasks:
    1. **Point estimation**: find a single value $\hat\theta$ close to $\theta$
    2. **Hypothesis testing**: decide whether the data are compatible with $H_0$
    3. **Bayesian inference**: describe uncertainty about $\theta$ via a distribution

## Point Estimation

An **estimator** is a function of the data:
$$
\hat\theta = \hat\theta(x_1, \ldots, x_n)
$$

Familiar examples:

- Sample mean $\hat\mu = \bar x = \frac{1}{n}\sum_{i=1}^n x_i$
- Sample variance $\hat\sigma^2 = S^2 = \frac{1}{n-1}\sum_{i=1}^n (x_i - \bar x)^2$
- Least squares in regression $\hat\beta = (X'X)^{-1} X' y$

## Evaluating an Estimator

Since $\hat\theta$ depends on random data, it has a **sampling distribution**.

```{r}
#| fig-height: 3.6
#| fig-width: 9
par(mar = c(4, 1, 1, 1))
curve(dnorm(x, 0.4, 1), -3.5, 4, lwd = 3, col = "steelblue", axes = FALSE,
      xlab = expression(hat(theta)), ylab = "", ylim = c(0, 0.75))
curve(dnorm(x, 0, 0.55), add = TRUE, lwd = 3, col = "firebrick")
axis(1, at = c(0, 0.4), labels = c(expression(theta), expression(E(hat(theta)^(1)))))
abline(v = 0, lty = 2); abline(v = 0.4, lty = 3, col = "steelblue")
arrows(0, 0.72, 0.4, 0.72, code = 3, length = 0.08); text(0.2, 0.74, "bias", pos = 3)
legend("topright", bty = "n", lwd = 3, col = c("steelblue", "firebrick"),
       legend = c(expression(hat(theta)^(1)*": biased, large variance"),
                  expression(hat(theta)^(2)*": unbiased, small variance")))
```

- Bias $E(\hat\theta) - \theta$, variance $V(\hat\theta)$, and $\text{MSE}(\hat\theta) = E(\hat\theta - \theta)^2 = \text{Bias}^2 + V(\hat\theta)$
- Comparing $\hat\theta^{(1)}$ and $\hat\theta^{(2)}$ requires knowing (or approximating) both sampling distributions

## Maximum Likelihood Estimation

The log-likelihood of $\theta$:
$$
\ell(\theta) = \log f(x_1, \ldots, x_n \mid \theta) = \log \prod_{i=1}^n f(x_i \mid \theta) = \sum_{i=1}^n \log f(x_i \mid \theta)
$$

The MLE is the maximizer:
$$
\hat\theta_{\text{MLE}} = \arg\max_\theta \ell(\theta)
$$

- Closed form exists in simple models (normal mean, binomial proportion)
- In most models (GLMs, mixed models, survival models) there is no closed form

## Hypothesis Testing

Test $H_0: \mu = \mu_0$ against $H_1: \mu \neq \mu_0$ using
$$
T = \frac{\bar x - \mu_0}{S/\sqrt{n}}
$$

- Need the **null distribution** of $T$ (distribution of $T$ when $H_0$ is true)
- For normal data, $T \sim t_{n-1}$ exactly; for other data, the null distribution is known only approximately or not at all

## $p$-value, Type I Error, and Power

```{r}
#| fig-height: 4.2
#| fig-width: 10
par(mar = c(4, 1, 1, 1))
tobs <- 1.5; tcrit <- qnorm(0.95); mu1 <- 2.5
curve(dnorm(x), -3.5, 6.5, lwd = 3, col = "steelblue", axes = FALSE,
      xlab = "T", ylab = "", ylim = c(0, 0.45))
curve(dnorm(x, mu1), add = TRUE, lwd = 3, col = "firebrick")
shade <- function(from, to, mean, col) {
  xs <- seq(from, to, length.out = 200)
  polygon(c(xs, rev(xs)), c(dnorm(xs, mean), rep(0, 200)), col = col, border = NA)
}
shade(tcrit, 6.5, mu1, adjustcolor("firebrick", 0.25))   # power
shade(tcrit, 6.5, 0,   adjustcolor("steelblue", 0.45))   # alpha
shade(tobs,  6.5, 0,   adjustcolor("gray30", 0.35))      # p-value
axis(1, at = c(0, tobs, tcrit, mu1),
     labels = c("0", expression(t[obs]), expression(t[alpha]), expression(mu[1])))
abline(v = tcrit, lty = 2); abline(v = tobs, lty = 3)
text(-1.6, 0.30, expression(H[0]*": null distribution"), col = "steelblue")
text(4.3, 0.30, expression(H[1]*": alternative"), col = "firebrick")
text(2.0, 0.055, "p-value", pos = 4, cex = 0.9)
text(tcrit, 0.42, expression(alpha*" = type I error (blue tail)"), pos = 4, cex = 0.9)
text(3.3, 0.16, "power (red tail)", pos = 4, cex = 0.9)
```

- $p$-value $= P_{H_0}(T \ge t_{\text{obs}})$: tail area of the null distribution beyond the observed statistic
- Type I error $\alpha = P_{H_0}(T \ge t_\alpha)$: null tail area beyond the critical value
- Power $= P_{H_1}(T \ge t_\alpha)$: alternative tail area beyond the critical value

All three are tail probabilities of a distribution that is often unknown in closed form.

## Bayesian Inference

Combine a prior $\pi(\theta)$ with the likelihood $f(x \mid \theta)$:
$$
p(\theta \mid x) = \frac{f(x \mid \theta)\,\pi(\theta)}{\int f(x \mid \theta)\,\pi(\theta)\, d\theta}
$$

The denominator is an integral over $\theta$; for most models it has no closed form.

## Prior, Likelihood, and Posterior

```{r}
#| fig-height: 4.2
#| fig-width: 10
par(mar = c(4, 1, 1, 1))
# normal mean with known sigma: vague prior N(0, 3^2), n = 5, xbar = 2, sigma = 2
s2p <- 9; xbar <- 2; n <- 5; s2 <- 4
v_post <- 1 / (1 / s2p + n / s2); m_post <- v_post * (n * xbar / s2)
curve(dnorm(x, 0, sqrt(s2p)), -8, 8, lwd = 3, col = "gray40", lty = 2,
      axes = FALSE, xlab = expression(theta), ylab = "", ylim = c(0, 0.5))
curve(dnorm(x, xbar, sqrt(s2 / n)), add = TRUE, lwd = 3, col = "firebrick")
curve(dnorm(x, m_post, sqrt(v_post)), add = TRUE, lwd = 3, col = "steelblue")
axis(1)
legend("topleft", bty = "n", lwd = 3, lty = c(2, 1, 1),
       col = c("gray40", "firebrick", "steelblue"),
       legend = c(expression("vague prior "*pi(theta)),
                  expression("likelihood "*f(x*"|"*theta)),
                  expression("posterior "*p(theta*"|"*x))))
```

- A vague prior lets the likelihood dominate; the posterior is slightly narrower and pulled toward the prior mean
- Inference uses posterior summaries: $E(\theta \mid x)$, $V(\theta \mid x)$, credible intervals, $P(H_0 \mid x)$
- Each summary is again an integral over $\theta$

## Summary of the Three Tasks

| Task | Object needed | Typical difficulty |
|---|---|---|
| Point estimation | Sampling distribution of $\hat\theta$; maximizer of $\ell(\theta)$ | No closed form |
| Hypothesis testing | Null distribution of $T$ | Unknown for general data |
| Bayesian inference | $\int f(x\mid\theta)\pi(\theta)\,d\theta$; $E(\theta \mid x)$ | High-dimensional integral |

Each task reduces to a computational problem: **calculating**, **simulating**, **optimizing**, or **integrating**.

# The Use of Computers in Statistics

## 1. Manipulating Numbers

Even simple calculations are not trivial on a computer:

- $1.1 + 2.1 + \cdots + 100.1$
- $S^2 = \sum_{i=1}^n (x_i - \bar x)^2 / (n-1)$
- $\prod_{i=1}^n f(x_i \mid \theta)$ for large $n$

Computers store numbers with finite precision, which leads to

- **Overflow**: result too large to represent
- **Underflow**: result too small, rounded to zero
- **Rounding error**: accumulated loss of digits

## Example: Rounding Error

```{r}
#| echo: true
x <- c(1e10 + 1, 1e10 + 2, 1e10 + 3)
var(x)                                    # correct answer: 1
sum(x^2)/2 - 3 * mean(x)^2 / 2           # textbook formula
```

```{r}
#| echo: true
prod(dnorm(rnorm(1000)))                  # likelihood underflows to 0
sum(dnorm(rnorm(1000), log = TRUE))       # log-likelihood is fine
```

Lecture on computer arithmetic: how numbers are stored and how to avoid these errors.

## 2. Simulation with Random Numbers

**Monte Carlo**: replace a probability calculation with repeated random experiments.

Generate $X_1, X_2 \overset{iid}{\sim} \text{Unif}(0,1)$. What is the distribution of $X_1 + X_2$?

```{r}
#| echo: true
#| fig-height: 3.5
z <- runif(10000) + runif(10000)
hist(z, breaks = 40, freq = FALSE, main = "", xlab = "X1 + X2")
```

## Monte Carlo for Sampling Distributions

Repeat many times: draw $x_1, \ldots, x_n$, compute the statistic.

$$
\sum_i x_i, \qquad \sum_i (x_i - \bar x)^2, \qquad T = \frac{\bar x - \mu_0}{S/\sqrt n}
$$

This gives empirical approximations to

- The sampling distribution of $\hat\mu^{(1)}$ vs $\hat\mu^{(2)}$ (bias, variance, MSE)
- The null distribution of $T$ for non-normal data
- Coverage of confidence intervals

Lecture on random numbers and Monte Carlo: pseudo-random generation, transformations, LLN/CLT, evaluating estimators and tests.

## 3. Optimization

Find $x_0 = \arg\max_x f(x)$.

```{r}
#| fig-height: 3.5
curve(dgamma(x, 3, 1), 0, 10, ylab = "f(x)", lwd = 2, main = "")
abline(v = 2, lty = 2)
text(2.1, 0.05, expression(x[0]), adj = 0)
```

For MLE, $f = \ell(\theta)$; there is no closed form, so the maximum is found by iterative search.

Lecture on MLE and optimization: Newton–Raphson, Fisher scoring, Nelder–Mead, gradient methods, stochastic gradient descent.

## 4. Advanced Simulation and Integration

Quantities of interest are integrals:
$$
E(X) = \int x f(x)\,dx, \qquad V(X), \qquad \int f(x\mid\theta)\pi(\theta)\,d\theta
$$

Two approaches:

- **Numerical quadrature**: approximate the area under the curve by rectangles or trapezoids; works well in low dimension
- **Monte Carlo integration**: draw from $f$ and average; the only practical route in high dimension

For the posterior $p(\theta \mid x)$, direct sampling is impossible, so **Markov chain Monte Carlo** is used.

## Numerical Quadrature

```{r}
#| fig-height: 4
f <- function(x) dgamma(x, 3, 1)
curve(f, 0, 8, lwd = 2, ylab = "f(x)", main = "")
k <- 12; xs <- seq(0, 8, length.out = k + 1)
rect(xs[-(k+1)], 0, xs[-1], f((xs[-(k+1)] + xs[-1]) / 2),
     col = adjustcolor("red", 0.3), border = "red")
```

$$
\int_a^b f(x)\,dx \approx \sum_{j=1}^k f(m_j)\,\Delta x
$$

## Course Roadmap

| Topic | Statistical task |
|---|---|
| Computer arithmetic | Reliable calculation of likelihoods and variances |
| Random numbers and Monte Carlo | Sampling distributions, evaluating estimators and tests |
| MLE and optimization | Point estimation without closed forms |
| Numerical integration and MCMC | Bayesian inference |

Throughout: R programming, vectorization, and efficient code.
