---
subtitle: "Bayesian Inference: A Brief Introduction"
pagetitle: "bayesian-slides"
author: "Longhai Li"
format: banner_slides-revealjs
number-sections: false
---

```{r}
#| include: false
set.seed(812)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center")
```

# 1. The Bayesian Framework

## Four Steps

1. **Specify a prior** $\pi(\theta)$: rough knowledge about $\theta$, treated as a random variable
2. **Specify a probability model** for the data: $P(y_1,\ldots,y_n\mid\theta)$
3. **Find the posterior** by Bayes' rule
$$P(\theta\mid y_1,\ldots,y_n) = \frac{P(y_1,\ldots,y_n\mid\theta)\,\pi(\theta)}{\displaystyle\int P(y_1,\ldots,y_n\mid\theta)\,\pi(\theta)\,d\theta}$$
4. **Interpret** the posterior: summarize $\theta$, make predictions

Writing $L(\theta) = P(y_{1:n}\mid\theta)$ for the likelihood,
$$P(\theta\mid y_{1:n}) \propto L(\theta)\,\pi(\theta)$$

The denominator $P(y_{1:n}) = \int L(\theta)\pi(\theta)\,d\theta$ is the **marginal likelihood**; it does not depend on $\theta$ and is usually the hard part to compute.

## Four Steps (Continued)

Left: a flat prior and the likelihood. Right: the posterior $\propto L(\theta)\pi(\theta)$, with a $(1-\alpha)$ credible interval $(\theta_{\alpha/2},\theta_{1-\alpha/2})$ cutting off $\alpha/2$ in each tail.

```{r}
#| fig-width: 10
#| fig-height: 4.6
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
L <- function(t) dnorm(t, 3, 1); pri <- function(t) dnorm(t, 2, 4) * 2.5
curve(L, -2, 8, lwd = 3, col = "firebrick", xlab = expression(theta), ylab = "", yaxt = "n",
      main = "prior and likelihood", ylim = c(0, 0.45))
curve(pri, add = TRUE, lwd = 3, col = "gray40", lty = 2)
legend("topleft", bty = "n", lwd = 3, lty = c(2, 1), col = c("gray40", "firebrick"),
       legend = c(expression(pi(theta)), expression(L(theta))))
post <- function(t) L(t) * pri(t); Z <- integrate(post, -10, 15)$value
pd <- function(t) post(t) / Z
curve(pd, -2, 8, lwd = 3, col = "steelblue", xlab = expression(theta), ylab = "", yaxt = "n",
      main = expression("posterior" %prop% L(theta) * pi(theta)), ylim = c(0, 0.45))
tt <- seq(-2, 8, length.out = 1000); cdf <- cumsum(pd(tt)) * diff(tt)[1]
q <- c(tt[which.min(abs(cdf - 0.025))], tt[which.min(abs(cdf - 0.975))])
for (side in 1:2) { xs <- if (side == 1) seq(-2, q[1], length.out = 100) else seq(q[2], 8, length.out = 100)
  polygon(c(xs, rev(xs)), c(pd(xs), rep(0, 100)), col = adjustcolor("steelblue", 0.35), border = NA) }
abline(v = q, lty = 3); text(q, 0.42, c(expression(theta[alpha/2]), expression(theta[1-alpha/2])))
text(q[1] - 0.5, 0.05, expression(alpha/2)); text(q[2] + 0.5, 0.05, expression(alpha/2))
```

## Summarizing $\theta$: Point Estimates from Loss Functions

The Bayes estimate minimizes the posterior expected loss:
$$\hat\theta = \arg\min_{\hat\theta}\; E\big[L(\theta,\hat\theta)\mid y_{1:n}\big]$$

| Loss $L(\theta,\hat\theta)$ | Bayes estimate $\hat\theta$ |
|---|---|
| $(\theta-\hat\theta)^2$ (squared error) | posterior mean $E(\theta\mid y_{1:n})$ |
| $\lvert\theta-\hat\theta\rvert$ (absolute error) | posterior median |
| $I(\theta\ne\hat\theta)$ (0–1 loss) | posterior mode (MAP) |

**Interval estimate**: a $(1-\alpha)$ credible interval $(\theta_{\alpha/2},\theta_{1-\alpha/2})$ contains $\theta$ with posterior probability $1-\alpha$, a direct probability statement about $\theta$ (unlike a confidence interval).

All of these are integrals or optimizations over the posterior, which is why Bayesian inference is computational.

## Prediction {.smaller}

The **posterior predictive distribution** of a new observation:
$$P(y_{n+1}\mid y_{1:n}) = \frac{\int P(y_{n+1},y_{1:n}\mid\theta)\,\pi(\theta)\,d\theta}{\int P(y_{1:n}\mid\theta)\,\pi(\theta)\,d\theta}$$

If $y_1,\ldots,y_n,y_{n+1}$ are conditionally independent given $\theta$, this simplifies to
$$P(y_{n+1}\mid y_{1:n}) = \int P(y_{n+1}\mid\theta)\,P(\theta\mid y_{1:n})\,d\theta$$

an average of the sampling density over the posterior. With posterior draws $\theta_1,\ldots,\theta_J\sim P(\theta\mid y_{1:n})$,
$$P(y_{n+1}\mid y_{1:n}) \approx \frac1J\sum_{j=1}^J P(y_{n+1}\mid\theta_j)$$

The predictive distribution is wider than any single $P(y_{n+1}\mid\theta_j)$: it carries both sampling variability and posterior uncertainty about $\theta$.

## Prediction (Continued)

Red: $P(y_{n+1}\mid\theta_j)$ for ten posterior draws $\theta_j$. Black: their average, the posterior predictive density.

```{r}
#| fig-width: 9
#| fig-height: 4.6
par(mar = c(4, 4, 1, 1)); J <- 10; th <- rnorm(J, 3, 0.8)
curve(dnorm(x, th[1]), -2, 8, col = adjustcolor("firebrick", 0.6), lwd = 1.5, ylim = c(0, 0.42),
      xlab = expression(y[n+1]), ylab = "density")
for (j in 2:J) curve(dnorm(x, th[j]), add = TRUE, col = adjustcolor("firebrick", 0.6), lwd = 1.5)
curve(sapply(x, function(v) mean(dnorm(v, th))), add = TRUE, lwd = 4)
legend("topright", bty = "n", lwd = c(1.5, 4), col = c("firebrick", "black"),
       legend = c(expression(P(y[n+1]*"|"*theta[j])), expression(P(y[n+1]*"|"*y[1:n]))))
```

# 2. Example: Bernoulli Data with a Beta Prior

## Prior

$y_1,\ldots,y_n\mid\theta \overset{iid}{\sim}\text{Bern}(\theta)$, with $\theta\sim\text{Beta}(\alpha_1,\alpha_0)$, $\alpha_1,\alpha_0>0$ given:
$$\pi(\theta) = \frac{\Gamma(\alpha_0+\alpha_1)}{\Gamma(\alpha_0)\Gamma(\alpha_1)}\,\theta^{\alpha_1-1}(1-\theta)^{\alpha_0-1}, \qquad 0<\theta<1$$

$$E(\theta) = \frac{\alpha_1}{\alpha_0+\alpha_1}, \qquad
V(\theta) = \frac{\alpha_0\alpha_1}{(\alpha_0+\alpha_1)^2(\alpha_0+\alpha_1+1)}$$

- $\alpha_1$ acts like a prior count of successes, $\alpha_0$ of failures; $\alpha_0+\alpha_1$ is the prior "sample size"
- $\text{Beta}(1,1) = \text{Unif}(0,1)$; $\text{Beta}(2,1)$ has density $2\theta$, favouring large $\theta$
- Small $\alpha$'s give U-shaped or flat priors (vague); large $\alpha$'s concentrate the prior

## Prior (Continued)

```{r}
#| fig-width: 9
#| fig-height: 4.8
par(mar = c(4, 4, 1, 1))
pars <- list(c(1, 1), c(2, 1), c(0.5, 0.5), c(2, 5), c(10, 10), c(30, 10))
cols <- c("black", "firebrick", "darkorange", "forestgreen", "steelblue", "purple")
plot(NA, xlim = c(0, 1), ylim = c(0, 6), xlab = expression(theta), ylab = expression(pi(theta)))
for (k in seq_along(pars)) curve(dbeta(x, pars[[k]][1], pars[[k]][2]), add = TRUE, lwd = 3, col = cols[k], n = 501)
legend("top", bty = "n", lwd = 3, col = cols, ncol = 3,
       legend = sapply(pars, function(p) sprintf("Beta(%g, %g)", p[1], p[2])))
```

## Likelihood and Posterior

**Likelihood**: with $P(y_i\mid\theta) = \theta^{y_i}(1-\theta)^{1-y_i}$ and $\mathbf y = (y_1,\ldots,y_n)$,
$$P(\mathbf y\mid\theta) = \prod_{i=1}^n \theta^{y_i}(1-\theta)^{1-y_i} = \theta^{n_1}(1-\theta)^{n_0},
\qquad n_1 = \sum_i y_i,\ \ n_0 = n - n_1$$

**Posterior**:
$$
\begin{aligned}
P(\theta\mid\mathbf y) &= \frac{\pi(\theta)\,P(\mathbf y\mid\theta)}{P(\mathbf y)}
\;\propto\; \theta^{\alpha_1-1}(1-\theta)^{\alpha_0-1}\cdot\theta^{n_1}(1-\theta)^{n_0} \\
&= \theta^{(\alpha_1+n_1)-1}(1-\theta)^{(\alpha_0+n_0)-1}
\end{aligned}
$$

This is the Beta kernel with updated parameters $\alpha_1' = \alpha_1+n_1$, $\alpha_0' = \alpha_0+n_0$:
$$\theta\mid\mathbf y \sim \text{Beta}(\alpha_1+n_1,\ \alpha_0+n_0)$$

The Beta prior is **conjugate** to the Bernoulli likelihood: the posterior has the same form, and the normalizing constant $P(\mathbf y)$ never has to be computed explicitly.

## Posterior Summaries

$$E(\theta\mid\mathbf y) = \frac{\alpha_1+n_1}{\alpha_0+\alpha_1+n}
= \underbrace{\frac{\alpha_0+\alpha_1}{\alpha_0+\alpha_1+n}}_{w}\cdot\frac{\alpha_1}{\alpha_0+\alpha_1}
+ (1-w)\cdot\underbrace{\frac{n_1}{n}}_{\hat\theta_{\text{MLE}}}$$

a weighted average of the prior mean and the MLE; the data weight $1-w\to1$ as $n\to\infty$.

$$V(\theta\mid\mathbf y) = \frac{(\alpha_1+n_1)(\alpha_0+n_0)}{(n+\alpha_0+\alpha_1)^2(n+\alpha_0+\alpha_1+1)} = O\!\left(\frac1n\right)$$

- The posterior variance shrinks to zero: the posterior concentrates at the true $\theta$ as $n\to\infty$
- Compare with the sampling distribution of the MLE, $\hat\theta_{\text{MLE}}\ \dot\sim\ N\big(\theta_0,\,1/(nI(\theta_0))\big)$, whose variance is also $O(1/n)$; for large $n$ the two approaches agree, with the prior's influence vanishing

## Posterior Summaries (Continued)

Prior $\text{Beta}(2,2)$ and posteriors after $n = 10$, $50$, $200$ observations with $\bar y = 0.7$. The posterior narrows at rate $1/\sqrt n$ and centres on the MLE.

```{r}
#| fig-width: 9
#| fig-height: 4.6
par(mar = c(4, 4, 1, 1)); a1 <- 2; a0 <- 2; ybar <- 0.7
ns <- c(10, 50, 200); cols <- c("firebrick", "darkorange", "steelblue")
plot(NA, xlim = c(0, 1), ylim = c(0, 13), xlab = expression(theta), ylab = "density")
curve(dbeta(x, a1, a0), add = TRUE, lwd = 3, lty = 2, col = "gray40")
for (k in seq_along(ns)) curve(dbeta(x, a1 + ns[k] * ybar, a0 + ns[k] * (1 - ybar)), add = TRUE, lwd = 3, col = cols[k])
abline(v = ybar, lty = 3)
legend("topleft", bty = "n", lwd = 3, lty = c(2, 1, 1, 1), col = c("gray40", cols),
       legend = c("prior Beta(2, 2)", paste0("posterior, n = ", ns)))
text(ybar, 12.5, expression(hat(theta)[MLE]), pos = 4)
```

## Where Computation Enters

For the Beta–Bernoulli model everything is closed-form. In general:

| Quantity | Requires |
|---|---|
| Normalizing constant $P(\mathbf y) = \int L(\theta)\pi(\theta)\,d\theta$ | integration over $\theta$ |
| Posterior mean, variance, probabilities | integration over $\theta$ |
| Posterior mode | optimization |
| Posterior predictive $\int P(y_{n+1}\mid\theta)P(\theta\mid\mathbf y)\,d\theta$ | integration over $\theta$ |
| Credible intervals | posterior quantiles |

- One-dimensional $\theta$: numerical quadrature
- Multi-dimensional $\theta$: Monte Carlo with posterior draws $\theta_1,\ldots,\theta_J$, obtained by **Markov chain Monte Carlo** (next lectures), turning every integral into a sample average
